### 2017-11-08 LandSurfaceEffects_LocalRainfall_AGRFORMETsubmit.Rnw
add files

\documentclass[authoryear,preprint,review,12pt]{elsarticle}

\journal{Agriculture and Forest Meteorology}

\usepackage{url}
% Elsevier suggests
\usepackage{amssymb}
 \usepackage{lineno}

% older suggested by Chun
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\usepackage{rotating}
\usepackage{bm}
% french characters
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
%switch on line numbers
\linenumbers

\begin{document}
\SweaveOpts{concordance=TRUE}
\begin{frontmatter}

\title{Observational data analysis of land surface effects on local rainfall}

\author{C. Liang}
\author{R.W. Vervoort \corref{cor1}}
\ead{willem.vervoort@sydney.edu.au}
\author{F.F. Van Ogtrop}

\address{Sydney Institute of Agriculture, School of Life and Environmental Sciences, The University of Sydney}
\cortext[cor1]{}

\begin{itemize}
\item There are limited studies that analyse observational data of vegetation feedback on rainfall
\item Two sites with major step trend landuse change were analysed for changes in rainfall
\item Statistical change in rainfall detected after bushfire, but not after land clearing
\end{itemize}

\begin{abstract}
Analysis of observational data to identify relationships between rainfall and land cover change are scarce due to multiple environmental factors that cannot be controlled. In this study we investigated the relationship using statistical methods on data from best available sources at two sites in Australia. Gridded data of rainfall and tree cover were used as spatially corresponding local conditions. Large scale effects were represented by climatic indicators, such as SOI and IOD. Regression analysis and step trend tests were used to assess the effect of abrupt land surface intervention. At a Queensland site, dramatic tree cover change between 2002 - 2005 did not result in strong statistically significant precipitation changes. On the other hand, results from step trend test on 1998 - 2008 rainfall data in a bushfire affected NSW/VIC region has a better match with the tree cover change map. This indicates the method works better when a abrupt change in the data can be clearly identified. The results from the step trend test implied a positive relationship between the tree cover and the rainfall at 0.1 significance level. 
\end{abstract}

\begin{keyword}
Observed data analysis \sep Rainfall change \sep Landcover feedback
\end{keyword}

\end{frontmatter}

\section{Introduction}

Land use and land cover changes can lead to changes in the local climate. Some observational and many modelling studies have found cloud types and rainfall are correlated to large scale vegetation cover changes, such as deforestation in the Amazon and in the Sahel \citep{Chagnon2005,Pinto2009, Wang2009, Mei2010, kucharski_further_2013,pitman_scale_2016} and afforestation in south Israel \citep{Otterman1990, Ben-Gai1998}. In terms of observational studies, using airborne measurement in Western Australia, \citet{Junkermann2009} showed a significantly higher level of aerosols over an agricultural area compared to an adjacent natural vegetation. They suggested that a modification of aerosol concentrations due to deforestation could have contributed to a reduction of local rainfall, as more but smaller rain droplets were observed. \citet{Nair2011} reported from the Bunny Fence Experiment in Western Australia that local land use change altered the synoptic west coast trough dynamics and surface roughness, and this resulted in the observed rainfall decrease. Climate sensitivity to land cover change is also found in eastern Australia \citep{McAlpine2007}.

Rainfall over land is generally influenced by multiple factors as has been shown from multiple simulation studies. Locally, there are two main sources: moisture from advective atmospheric transport; and local evapotranspiration \citep{Eltahir1996,Bosilovich2006,Dirmeyer2009,Gimeno2010}. According to \citet{Trenberth1999}, the contribution of advective moisture partially depends on the availability of external moisture and atmospheric transport. On the longer time scale, such as monthly and annually, large scale atmospheric dynamics are affected by large scale climate drivers. Many studies have reported significant relationships between rainfall in large parts of Australia and the El Ni\~{n}o-Southern Oscillation (ENSO) \citep{Verdon2004,Risbey2009,Speer2011}. ENSO can be used to represent longer term cycles in rainfall data such as drought. On the other hand, local ET is determined by local land surface characteristics. Local land surface characteristics further influence local scale atmospheric dynamics and hence the amount of rainfall, including contribution from both sources. Therefore land surface plays an important role in local rainfall.

Conventionally, large scale climate drivers are used as predictors for seasonal rainfall forecasts. According to the Australian Bureau of Meteorology (BoM), probablistic seasonal outlooks are developed based on Australian rainfall and temperature as well as sea surface temperature records from the tropical Pacific and Indian Oceans \citep{BoM2012c}.

Different parts of the Australian continent can be more or less influenced by different climatic drivers \citep{Mla2008}. Using both SOI and PDO in a prediction model, \citet{Kamruzzaman2011} reported that PDO was rarely significant for rainfall stations in the MDB and on the southeast coast of Australia, while SOI was at least significant at the 1\% level. In addition to SOI, \citet{Speer2011} found that the observed rainfall decrease in the southeast of NSW was linked to an increasingly positive SAM in 1976 - 1998. In southwest WA, the influence of any oceanic indices is small \citep{Smith2012}.

Although climate drivers demonstrate some capability to predict Australian rainfall, there is still a large amount of unexplained variance. \citet{Westra2010} pointed out that models based on global sea surface temperature anomalies can only predict up to 14.7\% of precipitation variance. Recent studies suggest that land surface processes are important for predicting local rainfall \citep[e.g.][]{Ma2011,Zeng2012, pitman_scale_2016, saha_investigating_2016}. However, they are mostly based on modelling experiments and little evidence was reported from observations. \citet{Pitman2004} found a good match between observations and simulated rainfall changes in southwest Western Australia ,forced by land cover change. \citet{Timbal2006} were able to reproduce the rainfall decline in south west Australia by including land cover influence. Local land use change might not be a primary, but is likely to be a secondary cause of rainfall change \citep{Nicholls2006}. Therefore, land surface modification has, at least partially, contributed to local rainfall variability.

Overall the number of observational data studies related to changes to rainfall due to land cover change is limited. This is because there are some fundamental experimental difficulties in both space (where does evaporated water reappear as rainfall?) and in time (how much time does it take for land cover change effects to appear or disappear?). Therefore, the aim of this study is to use observational data at regional scales to investigate the cause and effect relationship between land cover change and local rainfall using empirical evidence. More specifically, we hypothesize that a step change on the land cover on the surface will cause a step change in the rainfall and that this can be identified statistically. To demonstrate the approach and this effect we studied step changes in rainfall at two locations in Queensland and NSW/Victoria where there are possible step changes in land cover change due to land clearing and bushfires based on tree cover data. More specifically,  step changes in rainfall were identified statistically, which were subsequently associated with land cover change through spatial comparison.

In this paper, after this section (the introduction), section 2 covers the case study areas and the observed land use changed. Section 3 describes the data used in the study in more detail. Section 4 details the statistical methods and the underlying assumptions related to the modelling approach, Section 5 gives the results, which are further discussed in section 6 and finally section 7 offers the conclusions.

\section{Study regions and tree cover change}

In Australia, significant tree cover change has mainly occurred in the north east of the continent and on the southeast coast, as well as in the southwest of Western Australia. According to the National Dynamic Land Cover Dataset (DLCD) \citep{Lymburner2010}, most of these areas have experienced decreasing EVI between 2000 - 2008. Being a vegetation greenness index, the decreasing EVI values indicate lower biomass over time in the tree cover regions. The EVI reduction is possibly due to land clearing, bushfires or drought.

%figure 1
\begin{figure}[ht!]
  \centerline{\includegraphics[height = 4in,scale=0.5]{map_selreg.jpg}}
  \caption{Selected study regions are highlighted by red rectangles in the main map (the red rectangle in the insert indicates the location of the main map). The types of tree cover in 2008 from the DLCD product is shown at the background. In site 1 (the QLD region), the tree cover is mostly sparse. In site 2 (the NSW/VIC region), many areas have open or close forest in which tree cover is denser.}
  \label{fig:sel_reg}
\end{figure}

Two regions were selected where significant tree cover change was present. One region is located in south central Queensland to the north of the Murray Darling Basin (MDB) (site 1 in Figure~\ref{fig:sel_reg}). High rates of land clearing have been reported in this region during the early 2000s \citep{SLATS2004}. The second study region is located at the border of New South Wales and Victoria, including the Snowy Mountains range (site 2 in Figure~\ref{fig:sel_reg}). Severe bushfires occurred in this area and the surroundings in early 2003 (see Figure~\ref{fig:bushfire}). The 2003 bushfires were the largest and the worst in this area for the last 60 years \citep{Fire2011}. Two thirds of Kosciuszko national park was heavily burned and regrowth was reported to be slow due to drought and cold conditions \citep{ABC2003} and the type of species in this region. In both study regions, significant tree cover loss has happened in the last decade, either permanently or temporarily.

%figure 2
\begin{figure}[ht!]
  \centerline{\includegraphics[height=4in, clip, scale=0.6]{bushfire_nswvic.pdf}}
  \caption{Location of bushfires occurring in January 2003, in and around the NSW/VIC study region, as shown by the red pixels. The map shows large area in the Kosciuszko national park has been burned. Some locations in the southwest of ACT have also experienced intensive bushfires.}
  \label{fig:bushfire}
\end{figure}

The two regions have different climate characteristics. The QLD region is partially grassland and partially subtropical, while the NSW/VIC region is mainly within the temperate zone, under the K\"{o}ppen classification. According to Australian Bureau of Meteorology (BoM), the NSW/VIC region receives 1000 - 2000 mm rainfall annually, which is more than double of the rainfall in the QLD region. Evapotranspiration is similar in both regions. Marine moisture and orographic effects are likely to be the main contributors to rainfall in the southeast mountain areas of the NSW/VIC region.

The land use and land cover characteristics in the two regions are also different. In the Queensland region, the tree cover is sparse over most of the area. The MODIS satellite tree cover data (discussed in more detail in section 3) shows that tree cover in this region is generally below 20\% of total ground area. Grazing is the main activity in this region, with over 90\% of land used by the grazing industry \citep{ABARES2010}. Our starting assumption is that the main cause of the EVI  decline over large part of the region is due to land clearing. Tree cover has been cleared at a massive scale over the last decade, especially during 2002 - 2004.

The Kosciuszko national park is within the NSW/VIC region. Here tree cover is denser with open or closed forest (the tree cover distribution is bimodal at 10 - 20\% and 60 - 70\%). The dominant species in the alpine area are Snow Gum and large stand species such as Alpine Ash and Mountain Gum in the sub-alpine area. These trees can reach a great height but they take long time to grow. For example, Alpine Ash would need about 20 years to mature. Although land clearing is not a major issue in this region, it is vulnerable to fires and drought.

Therefore two types of land cover changes were studied. The reports from the Queensland Statewide Land Cover and Trees Study (SLATS) \citep[e.g.][]{SLATS2001,SLATS2003} were used to investigate the time and location of land clearing in the QLD region. The MODIS burned area product, MCD45A1 \citep{Roy2002,Roy2005,Roy2008}, was used to locate bushfires areas in the NSW/VIC region, with a grid resolution of 500 m. MCD45A1 provides monthly burning information on all pixels, which helps to pinpoint an abrupt event. Due to the nature of the different land cover change, the post-change vegetation status in the two regions is expected to be different (see Figure~\ref{fig:tc_simple}).

%figure 3
\begin{figure}[ht!]
  \centerline{\includegraphics[scale=0.65]{tc_simple.pdf}}
  \caption{The expected evolution of the land surface after trees have been removed in (a) the QLD region and (b) the NSW/VIC region.}
  \label{fig:tc_simple}
\end{figure}

Specifically, this study focused on the effect of 2003 - 2004 land clearings in the QLD region and the effect of 2003 bushfires in the NSW/VIC region. These events are expected to cause a step change in the local rainfall. The actual tree cover change at the pixel level during this time was derived from the 11-year MODIS data (discussed below). The difference of tree cover before and after the land disturbance was tested using a Student's t-test. As the length of the tree cover data is shorter than the length of the rainfall data, earlier land clearings in the QLD region cannot be identified spatially, hence they are excluded from the analysis.

\section{Data}
\label{sec:Data}

Two land surface data sets were used in this study. The main one was the MOD44B product Global Vegetation Continuous Field dataset (version 5). This dataset provides estimates of percent tree cover (percentage of ground surface covered by trees) at a grid resolution of 250 m \citep{Townshend2011}, which is finer then the earlier mentioned burned product MCD45A1. The dataset is available on an annual time interval for the period of 2000 - 2010. The National Dynamic Land Cover Dataset (DLCD) \citep{Lymburner2010} from the Australian Collaborative Land Use Mapping Program (ACLUMP) was used to verify the trend of vegetation cover change calculated from the previous dataset. This dataset, developed by Geoscience Australia and Australian Bureau of Agricultural and Resource Economics and Sciences (ABARES), is the first nationally consistent and thematically comprehensive land cover reference for Australia. The DLCD is based on the 16-day Enhanced Vegetation Index (EVI), again from the MODIS satellite, between April 2000 and April 2008. It also has a grid resolution of 250 m. The dataset provides information on the final land cover types (as in 2008) and estimated trend of EVI statistics (annual mean, maximum and minimum).

Rainfall data for Australia \citep{Jones2009} was obtained from BoM. The data has been projected onto a national 0.05\textdegree $\times$ 0.05\textdegree grid (approximately 5 km $\times$ 5 km). This gridded dataset was generated from station observations using an optimised Barnes successive correction technique. The Barnes technique combines a weighted averaging process and defined topographical information to estimate rainfall values between spatial points \citep{BoM2009}. The resulting dataset provides additional information for data-sparse areas like central Australia but reduces information in the data-rich areas, such as southeast Australia where station density is up to 20 per 100 km$^2$. The data is available on a monthly basis from 1900 to current. Here a subset of 30 years (1979 - 2008) was used. The study was conducted on monthly data, as a land cover change effect on annual rainfall might be negligible but can often found to be significant in particular months or seasons \citep[e.g.][]{Otterman1990,Gaertner2001,Semazzi2001,Oleson2004,Deo2009}.

Large scale climate drivers are represented by various climatic indices. The Southern Oscillation Index (SOI) is generally regarded as a good predictor for Australian rainfall \citep{Risbey2009,Chowdhury2010,Westra2010}, but its skill is weaker in some parts of Australia. For example the Southern Annular Mode (SAM) is found to be more important than ENSO in south Western Australia \citep{Meneghini2007}. The suitability of each index for the study regions is analysed in detail in section \ref{sec:reg_model}. The following climate indices were tested as predictors for local rainfall. 
\begin{itemize}
  \item Southern Oscillation Index (SOI). The Troup version of the monthly SOI series used in this study was obtained from BoM (available online at \url{http://www.bom.gov.au/climate/current/soihtm1.shtml}).
  \item Eastern, East Central and Central Tropical Pacific Sea Surface Temperatures (NINO 3, NINO 3.4 and NINO 4). Monthly SST anomalies are available from IRI/LDEO data library and the extended NINO dataset is used (available online at \url{http://iridl.ldeo.columbia.edu/SOURCES/.Indices/.nino/.EXTENDED/}).
  \item Pacific Decadal Oscillation (PDO). The Pacific Decadal Oscillation is the leading principal component of monthly SST anomaly in the North Pacific Ocean.. The monthly PDO series was provided by JISAO (Joint Institute for the Study of the Atmosphere and Ocean, University of Washington) (available online at \url{http://jisao.washington.edu/pdo/PDO.latest}).
  \item The interaction of PDO and SOI (PDO$\times$SOI) \citep{Kamruzzaman2011}.
  \item Indian Ocean Dipole (IOD). The Indian Ocean dipole is commonly measured by the difference between SST anomaly in the western (50 - 70\textdegree E and 10\textdegree S-10\textdegree N) and eastern (90 - 110\textdegree E and 0 - 10\textdegree S) equatorial India Ocean \citep{Saji1999}. Monthly IOD was obtained from JAMSTEC (the Japan Agency for Marine-Earth Science and Technology) (available online at \url{http://www.jamstec.go.jp/frcgc/research/d1/iod/DATA/dmi.monthly.txt}).
\end{itemize}

\section{Statistical method}

As shown in Figure~\ref{fig:ts_mean}, a step change is not obvious in the time series data, even though the data is deseasonalised and detrended. Hence, the step changes in the rainfall were analysed using two different statistical methods to provide a comparison. Both methods make use of a regression model to remove variability in rainfall hypothesised to be caused by factors other than vegetation cover change. In the first method, the tree cover change was implemented as a factor variable in the regression model. In the second method, a rank sum test (step trend test), was applied to the model residuals after effects of other major factors have been removed, assuming that vegetation cover change is the only factor explaining the non-random pattern in the rainfall residuals.

%figure 4
\begin{figure}[ht!]
  \centering
  \includegraphics{mean_ts.pdf}
  \caption{The deseasonalised and detrended rainfall over the 30 years period in (a) the QLD region and (b) the NSW/VIC region. The vertical red lines indicate the year of 2003, in which the studied land cover changes occurred. A change in the time series data is not obvious before and after the land cover changes.}
  \label{fig:ts_mean}
\end{figure}

\subsection{Regression model}
\label{sec:reg_model}

Rainfall is generally affected by several factors. Many studies suggest the large-scale climate drivers, which are related to the global or regional atmospheric air mass movement, are the dominant control on the overall rainfall \citep{Maynard2003,DeAngelis2010,Holper2011,Smith2012}. Long term trends are shown in Australian rainfall, and this has been linked to global climate change \citep{cai_did_2014,Delworth_regional_2014, Holper2011, karoly_climate_2014}.
The difference in available radiation and hence ET in different seasons increase the variability in rainfall. The land surface effect on rainfall can be confused by changes in these factors. Hence a regression model was used to estimate the amount of variability in rainfall that is due to these important factors, and isolate changes resulting from local vegetation cover change.
As highlighted in the introduction, the Australian climate is influenced by sea surface temperature in the tropical Pacific and Indian Oceans, as well as pressure systems in the Southern Ocean \citep{BoM2012}.  \citet{Risbey2009} compared five large-scale drivers, including ENSO (measured by SOI and the Tropical Pacific SSTs), IOD, SAM, MJO\footnote{MJO is a large scale eastward-propagating wave-like disturbance in equatorial latitudes \citep{Risbey2009}.} (Madden-Julian oscillation) and blocking, in relation to Australian rainfall variability. They identified SOI as the most important index among all indices tested for broad parts of Australia (including QLD and NSW/VIC) in almost any season. In this study, up to two important indices identified from the seven climatic indicators (see section \ref{sec:Data}) were used, for each study region, as the explanatory variables in the model .

As a preliminary step, the correlations between rainfall and each climatic index were analysed. Rainfall in each study region was first deseasonalised and detrended using the seasonal decomposition function "stl" in R \citep{Rstats2011}. Detrending is needed to remove the correllations between trends in the data as discussed in \citet{Smith2012}. 
The cross-correlations between the deseasonalised and detrended rainfall and the climatic indices were tested using the Pearson's product moment correlation method, assuming the relationships are linear. Although the optimal technique for exploring the correlation with each index could be different as described in \citet{Risbey2009}, the Pearson's method was applied to all indices for consistency. Because the PDO is a low frequency descriptor of the multi-decadal SST  \citep{MacDonald2005,Zanchettin2008,Kamruzzaman2011} a longer period (108 years, from 1900 to 2008), instead of 30-year rainfall data, was used to estimate the correlation with PDO, up to lag 24. For the other indices, the 30-year rainfall data set was used.

%figure 5
\begin{figure}[ht!]
\centering
\includegraphics[scale=0.6]{cor_qld.pdf}
 \caption{Cross-correlation of the six indices and rainfall in QLD study region. Where the correlations with PDO are sought, 108-year rainfall data (1900 - 2008) are used. Otherwise, 30-year rainfall data are used. The correlation with NINO 3 is not shown as it is very similar to but weaker than the case of NINO 3.4. The blue dashed lines indicate the 95\% confidence interval.}
  \label{fig:cor_rain_qld}
\end{figure}

%figure 6
\begin{figure}[ht!]
\centering
\includegraphics[scale=0.6]{cor_nswvic.pdf}
 \caption{Cross-correlation of the six indices and rainfall in NSW/VIC study region.}
  \label{fig:cor_rain_nsw}
\end{figure}

Based on the correlation between the climatic indices and rainfall (as shown in Figure~\ref{fig:cor_rain_qld} and Figure~\ref{fig:cor_rain_nsw}), the following results were found:
\begin{itemize}
  \setlength{\itemsep}{0cm}
  \setlength{\parskip}{0cm}
  \item In QLD, the correlation between rainfall and SOI at zero time lags is the highest across all indices, outweighing the other ENSO indicators. 
  \item In NSW/VIC, again the SOI has the highest correlation with rainfall, closely followed by the IOD. Both occur at the zero time lags. 
\end{itemize}
The above findings are consistent with previous studies. Although some indices are serially correlated with rainfall up to several months, the lag zero events have the most significant correlation coefficients. Concurrent climatic index series were generally found most useful in rainfall prediction \citep[e.g.][]{Risbey2009,Kamruzzaman2011}. The correlations between the climatic indices and rainfall for each individual season have also been tested. SOI was used in the Qld prediction, SOI and IOD were used in the NSW/VIC prediction.

Rainfall in Australia shows strong seasonal patterns \citep{Holper2011,ABS2012}. For example, the north part of the country is summer rainfall dominant with dry winter, while most of the southern part has a winter rainfall regime. This character is given by the movement of subtropical high pressure system which dominates the Australian climate \citep{BoM2012a}. The seasonal component of rainfall has a periodic pattern so it is better modelled as a smooth term. A spline function was applied on the months to define a smooth seasonal pattern \citep{Wood2011}. %The spline function ``s" in R was used .

Long term trends in the regional rainfall in some parts of Australia are significant \citep{Hughes2003,Gallant2007,Chowdhury2010}. In the northern and eastern parts of the continent, increasing rainfall is reported over the last century \citep{Hughes2003}. The presence of such long term trends may be confused with the outcome of a step change in rainfall. A linear trend term was implemented in the model to remove any long term trend effect.

We assumed all the factors are additive components in determining rainfall as in \citet{Kamruzzaman2011}. Generally, monthly rainfall has a skewed distribution so the normality assumption of the residuals in a general linear model could be violated. In this case, the rainfall model is expressed as a generalised additive model (GAM) \citep{Hastie1986} with a log link function g(), assuming the residuals are gamma distributed (see Figure~\ref{fig:hist_rain}).
\vspace{0.5cm}
\begin{linenomath*}
  \begin{equation}
  g(E(\mathbf{R}_r)) = \beta_0 + s_1(D_{1,r}\mathbf{SOI}) + s_2(D_{2,r}\mathbf{IOD}) + s_3(\mathbf{Season}) + \beta_1\mathbf{Trend} + \boldsymbol{\epsilon}_r
    \label{eqn:model}
  \end{equation}
\end{linenomath*}
The bold letters represent the time series vectors. The subscript $r$ denotes the region, $r=$qld or nswvic. $\beta_u$ ($_u$=0, 1) are the fitted coefficients in the model. $s_v$ ($_v$=1, 2, 3) are the smooth functions on the climatic indices and the season. $D_{1,r}$ and $D_{2,r}$ switch on/off the corresponding climatic index in the model as discussed previously. 
\begin{linenomath*}
  \begin{equation}
    D_{1,r} = 1 
  \end{equation}
\end{linenomath*}
as SOI was used in both regions as rainfall predictor.
\begin{linenomath*}
  \begin{equation}
    D_{2,r} = \left\{ \begin{array}{ll}
     1 &\mbox{for NSW/VIC } \\
     0 &\mbox{for QLD}
    \end{array} \right.
    \label{eqn:dummyD}
  \end{equation}
\end{linenomath*}
The linear long term trend in the rainfall data is modelled by \textbf{Trend}=1,2,3...n, where n is the total number of months in the time series. \textbf{Season} is the seasonal component which is represented by applying a smooth spline function on the month values (1-12. The monthly rainfalls in both regions appear to have a lag 1 autocorrelation, being 0.26 for QLD and 0.33 for NSW/VIC. 
The SOI and IOD terms are also modelled with spline functions. As the effect of large scale drivers on Australian rainfall is more likely to be seasonal \citep{Murphy2008,Schepen2012}, the spline function can closely reproduce the high and low impacts of the climatic indices.

%figure 7
\begin{figure}[ht!]
  \centering
  \includegraphics[scale=0.6]{hist_rain.pdf}
  \caption{Distribution of monthly rainfall in (a) QLD and (b) NSW/VIC. By using a Kolmogorov-Smirnov test with shape = 1 and 2.4 respectively, rainfalls in both regions are shown to have a gamma distribution.}
  \label{fig:hist_rain}
\end{figure}

\subsection{Tree cover change as factor variable}

One of the main difficulties in empirical observation studies on the effect of land cover change on rainfall is the lack of continuous monitoring of land surface variables, or even, no specific variable can be defined to clearly represent the land surface process. Given the lack of a full picture of the land surface process, a factor variable was used in this study to represent the abrupt land surface change (see Equation~\ref{eqn:model2}). The change could be a result of either land clearing or bushfires as long as it is permanent or takes a long time to recover. Here we approached the problem with two different models.

In the first method, the tree cover change was used as a predictor in the regression model, represented by a factor variable \textbf{LC}. The significance of the coefficient of \textbf{LC}, denoted as $\beta'_2$ in Equation~\ref{eqn:model2}, can be determined by a ratio test.
\begin{linenomath*}
  \begin{equation}
    \mbox{LC} = \left\{ \begin{array}{ll}
       \mbox{Trees} \\
       \mbox{Removed}
    \end{array} \right.
    \label{eqn:dummyC}
  \end{equation}
\end{linenomath*}
Therefore in both regions, land cover is ``trees" for the period before land cover change and ``removed" for the period after the change. Here we simply assumed that vegetation cover change has occurred on every pixel. The remaining term $\epsilon_r$ is the amount of rainfall that is attributed to other unspecified factors and random errors. Hence the regression model becomes
\vspace{0.5cm}
%\begin{dmath}
\begin{linenomath*}
\begin{equation}
g(E(\mathbf{R}_r)) = \beta'_0 + s'_1(D_{1,r}\mathbf{SOI}) + s'_2(D_{2,r}\mathbf{IOD}) + s'_3\mathbf{Season} + \beta'_1\mathbf{Trend} + \beta'_2\mathbf{LC} + \boldsymbol{\epsilon'}_r
  \label{eqn:model2}
\end{equation}
\end{linenomath*}

Vegetation cover changes occurred at different times in the two regions. In the QLD region, there no exact time can be assigned to the land clearing. Clearing has occurred between 2003 - 2004 according to SLATS reports. The information on the change in type of land cover during this time period is missing. Therefore, four scenarios were tested in the analysis. In these scenarios the ``after change" period started from: (1) June 2003, (2) January 2004, (3) June 2004 and (4) January 2005. In the NSW/VIC region, severe bushfires were reported in early January 2003. Hence the ``tree" cover state was up to December 2002 then it was changed to ``removed" state from January 2003. The regression model was run from 1979 for both regions.

\subsection{Step trend test}

As a second method, a step trend test was used to detect changes in rainfall after the vegetation cover change. This nonparametric statistical test was modified from the Mann-Whitney Rank-Sum test by \citet{Hirsch1985}. The test was developed to look for a step change in data which is cross correlated. As the gridded rainfall dataset is based on an interpolation method, this results in a high spatial correlation between neighbouring pixels. The step trend test is suitable for the analysis of such a dataset. The advantages of this test are: (1) it does not depend on assumptions of the data distribution; (2) it is not restricted to datasets with no missing data; (3) it is robust and not as easily influenced by outliers and negative numbers \citep{Hirsch1985}.

\subsubsection{Rainfall residuals} %% do we need this heading?

The rainfall residuals from the regression model in Equation~\ref{eqn:model} were used for this test. According to \citet{Hirsch1985}, using deseasonalised and detrended data is important in a test to detect step change. Furthermore, since rainfall is only partially attributed to local sources and conditions, noise can be introduced by large scale dynamics and changes in other climatic factors. The assumption is that the regression model should remove this noise. Hence the model residuals $\epsilon'_r$ exclude the effects due to climate drivers, seasonality and the long term trend. As a result local effects are assumed to explain at least part of the variation in the residuals. The  test, described in the following section, associates changes in rainfall with the tree cover changes.

\subsubsection{Mann-Whitney rank-sum statistic}

The step trend test is a modified version of the Mann-Whitney rank-sum statistic \citep{Hirsch1985}. As a nonparametric rank-based test, the Mann-Whitney test does not use the exact values of rainfall but depends on the ranks of the data. For each month, rainfall residuals of each year were ranked in an ascending order. The ranking of January rainfall in a sample pixel k in QLD is illustrated below:

\begin{table}[h]
  \centering
  \begin{tabular}{lccccccc}
  & 1998 & 1999 & 2000 & 2001 & 2002 & 2005 & 2006\\
  \textit Rainfall residuals: & -0.3 & -60.9 & -16.1 & -71.7 & 111.1 & -7.2 & -60.5 \\
  \textit Rank ($R'_{1k}$): & 6 & 2 & 4 & 1 & 7 & 5 & 3\\
  \end{tabular}
\end{table}
\noindent

Therefore, the smallest value (including negative values) has rank 1 and the largest value has the maximum rank.

The before period and the after period formed two groups of samples. The split point of the two periods was based on the timing of the vegetation cover changes. In the QLD region, changes occurred anytime during 2003 and 2004. In contrast to the previous method, the time period of where the land cover chang occurred was excluded, as the nonparametric test allows missing data. \citet{Hirsch1985} also pointed out that the power of the test is higher if the data of the change period is ignored. Hence 2003 and 2004 were excluded from the analysis. As a result, the after-change period was 2005 - 2007.

In the case of NSW/VIC, the bushfires broke out in early January 2003. The change was within a relatively short period of the year. Therefore the after-change period in this region still started in January 2003. The before period was set to five years (1998 - 2002) in both regions.

The rank of rainfall in month j year i in pixel k is denoted as $R'_{ijk}$. The sum of ranks of rainfall in month j in pixel k before the known intervention is:  
\begin{linenomath*}
  \begin{equation}
    W_{jk} = \sum_{i=1}^{n_1}R'_{ijk}.
    \label{eqn:Wj}
  \end{equation}
\end{linenomath*}
$n_1$ is the number of years before the land cover change. The expected value of $W_{jk}$ is 
\begin{linenomath*}
  \begin{equation}
    \mu_w=n_1(n_1+n_2+1)/2
  \end{equation}
\end{linenomath*}
$n_2$ is the number of years after the change. Hence the expected value of the rank sum before the intervention is the same for all months and all pixels. The sum of ranks for the whole time period is fixed, as $(n_1+n_2)(n_1+n_2+1)/2$. Knowing the rank-sum of one group the rank-sum of the other group can be easily derived. If the rainfall data is temporally and spatially independent, the variance of $W_{jk}$ is 
\begin{linenomath*}
  \begin{equation}
    \sigma^2_w = n_1\cdot n_2(n_1+n_2+1)/m
  \end{equation}
\end{linenomath*}
where m is the number of months which is 12 in the case of a full year.

\subsubsection{Step trend test}

Instead of completing the Mann-Whitney U-test, \citet{Hirsch1985} applied the rank-sum statistics in a standard normal Z test. The modified test can be used to detect step change and it accounts for serial and cross correlation in the data. In the case here, the deseasonalised and detrended data shows little autocorrelation in the time series but possesses strong cross correlation between neighbouring pixels, i.e. $R>0.99$. Hence the covariance between pixels would need to be considered in the test.

The sum of $W_{jk}$ for a block of $ns$ pixels over the whole year, $\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk}$, has a mean value of 
\begin{linenomath*}
  \begin{equation}
    E(\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk})=12\cdot ns\cdot\mu_W
  \end{equation}
\end{linenomath*}
and variance
\begin{linenomath*}
  \begin{equation}
    Var(\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk})=\sum_{j=1}^{12}\sum_{k=1}^{ns}\sum_{h=1}^{ns}C(W_{jk},W_{jh}).
  \end{equation}
\end{linenomath*}
$C(W_{jk},W_{jh})$ is the covariance of the W statistics between pixel k and pixel h in month j. When $k=h$, $C(W_{jk},W_{jh})=\sigma^2_w$. When $k\neq h$, 
\begin{linenomath*}
  \begin{equation}
    C(W_{jk},W_{jh})=\sigma^2_w r(R_k,R_h)
  \end{equation}
\end{linenomath*}
where $r(R_k,R_h)$ is the product moment correlation coefficient of the concurrent ranks in pixel k and h. Here $r$ is calculated on the full time series in each pixel. In this analysis, the test was applied to a square block of four pixels each time. As argued by \citet{Hirsch1985}, $ns=4$ is the most optimal solution to balance the cost and the gain in the test power.

The statistic of the step trend test is then defined as 
\begin{linenomath*}
  \begin{equation}
    Z'=\frac{\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk}-12\cdot ns\cdot\mu_w}{\sqrt{Var(\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk})}}.
    \label{eqn:Z}
  \end{equation}  
\end{linenomath*}
The above statistic is written for a 12 month period. The statistic can also be used to test seasonal rainfall change or for other customized periods, by changing the value 12.

The null hypothesis ($H_0$) in this study is that there was no change in rainfall due to land surface intervention. The results of the step trend test can be interpreted according to the sign of Z' score (see Table~\ref{tab:Zscore}). Z' is normally distributed similar to the standard normal statistics Z. Hence it can be compared to a standard normal distribution to determine the p value.

\begin{table}
  \caption[The interpretation of Z' score in the step trend test.]{The interpretation of Z' score in the step trend test. Following \citet[][Chapter 23, P887]{Hipel1994}.}
  \label{tab:Zscore}
  \centering
  \begin{tabular}{ll} \hline
    Z' $>$ 0 & rainfall decreases after change \\
    Z' $<$ 0 & rainfall increases after change \\ 
    Z' = 0 & rainfall does not change \\ \hline
  \end{tabular}
\end{table}

\section{Results}

\subsection{Tree cover changes}

The pixels with significant tree cover change in each study region are shown in Figure~\ref{fig:tc_trend}, at the p $<$ 0.1 significance level. Despite the limitations of the length of this data series, it still indicates a large change in the NSW/VIC region. In the NSW/VIC region, much of the tree loss between 2002 and 2003 was in the Snowy Mountains which are at the border of NSW and VIC. Tree cover loss occurred in large parts of the QLD study region between 2002 and 2005. Most of the clearings appear in the centre of this region. The tree cover change map is consistent with the annual mean EVI trend map (based on DLCD data, map not shown here), which confirms these changes are significant and persistent over the study period.

%figure 8
\begin{figure}[ht!]
\centering
\includegraphics[scale=0.65]{tc.pdf}
  \caption{The maps show the areas with significant changes in tree cover (at p $<$ 0.1) in (a) the QLD region and (b) the NSW/VIC region. The amount of change was calculated as the difference in tree covers before and after the specified land cover intervention and it is shown as the percentage of the ground area. Green colour indicates an increase in tree cover, while red colour indicates a decrease in tree cover. }
  \label{fig:tc_trend}
\end{figure}

\subsection{Regression Model \& Significance of Vegetation Cover Changes}

The regression model does not explain much of the rainfall variability. The model in Equation~\ref{eqn:model2} accounts for around 13\%\footnote{Here the adjusted $R^2$ was reported. Adjusted $R^2$ is the coefficient of determination, a measurement of the amount of variability predicted by the model adjusting for the number of explanatory terms} of the rainfall variations in the QLD region and 19\% in the NSW/VIC region on average. The residual analysis shows that the assumptions of the regression model are generally met. The standardised residual plots, however, show some funnelling for the NSW/VIC regions, suggesting non constant variance. The residual analysis, based on one sample pixel from each region, in Figure~\ref{fig:residuals_qld} and Figure~\ref{fig:residuals_nswvic} illustrates this. The residual patterns are consistent within each region.

The model confirms the importance of the climate drivers and the seasonality in Australian rainfall. Even at the grid level, the seasons and the climatic indices were significant ($p < 0.05$)  in both regions. The explaining power of the model is mostly due to these variables. The climate drivers (at lag zero) accounted, on average, for 6\% of the rainfall variability in both the QLD region and the NSW/VIC region (see Figure~\ref{fig:rsq} for the distribution of $R^2$ in these two regions). These figures are within the upper bound of seasonal rainfall predictability by SST anomaly field reported by \citet{Westra2010}.

%figure 9
\begin{figure}[ht!]
  \centering
  \includegraphics[scale=0.8]{qld_check.pdf}
  \caption{The residual analysis of a sample pixel in the QLD region.}
  \label{fig:residuals_qld}
\end{figure}

%figure 10
\begin{figure}[ht!]
  \centering
  \includegraphics[ scale=0.8]{nswvic_check.pdf}
  \caption{The residual analysis of a sample pixel in the NSW/VIC region.}
  \label{fig:residuals_nswvic}
\end{figure}

%figure 11
\begin{figure}[ht!]
  \centering
  \includegraphics[scale=0.65]{soi.pdf}
  \caption{The distribution of $R^2$ when rainfall is only modelled by the climate drivers. It shows the percentage of rainfall variability that can be explained by the climate drivers. In the case of (a) the QLD region, SOI was the only climatic index considered. In (b) the NSW/VIC region, SOI and IOD were used.}
  \label{fig:rsq}
\end{figure}

A statistically significant long term trend was not observed in either of the two regions. However, this result does not disprove the importance of a long term trend in rainfall. The overall time period is fairly short and more pixels in NSW/VIC would have a significant step change if the long term trend effect is not removed by the model. As trend free data is an important requirement for the step trend test, the trend term was kept in the regression model to ensure the detection of step change was not due to a possible long term trend.

%figure 12
\begin{figure}[ht!]
\centering
\includegraphics[ scale=0.6]{Cp_30yrs.pdf}
  \caption{The spatial distribution of significance of vegetation cover changes in the two study regions. The significance of vegetation cover changes is assessed in the regression model. The figures report the p values of the coefficient of the land cover variable in the model at different significance level. The plot of the QLD region is for the scenario that the after change period started from January 2005, as the other scenarios don't have any p value less than 0.2.}
  \label{fig:LC_p}
\end{figure}

The land cover variable implies a step change with different values before and after the land cover intervention. This variable was only significant ($p=0.05$) for the rainfall estimates in some areas in NSW/VIC, as shown in Figure~\ref{fig:LC_p}~(b). The effect of tree removal in the NSW/VIC region was highly significant in the area from the Snowy Mountains range in the south to the west side of Australian Capital Territory (ACT), highlighted by the red colour in Figure~\ref{fig:LC_p}~(b). However no significant step change (at p $<$ 5\%) due to the land cover changes was found in rainfall of the QLD region in any of the four scenarios.

In the NSW/VIC region, the locations of rainfall change show some agreement with the locations of vegetation cover change. Comparing the results in Figure~\ref{fig:LC_p}~(b) to Figure~\ref{fig:tc_trend}~(b), the alpine area with massive tree cover loss is located inside the highlighted region. The results show that the step change in rainfall also occurred in the cotter river catchment which was heavily burned in 2003 bushfires (refer to Figure~\ref{fig:bushfire}). However, Figure~\ref{fig:LC_p}~(b) also shows that a significant step change in rainfall was found in an area larger than where the bushfires has occurred. From this point of view, the results might be showing a large scale effect that extends beyond the vegetation cover change effect.

More generally, the model also shows that the tree cover has a positive impact on rainfall. The fitted coefficients for the reference state "trees" were consistently positive for the pixels whose step changes of rainfall were significant. It implies that rainfall was higher when the surface was covered by trees. This was confirmed by applying a two sample Student's t-test on the annual rainfall of the before-change and after-change periods.

\subsection{Step Trend Test}

%figure 13
\begin{figure}[ht!]
  \centering
 \includegraphics[ scale=0.7]{step_new.pdf}
  \caption{Spatial distribution of the step trend test Z' statistics in the two study sites. The test was conducted on the 30-year rainfall data from 1979 to 2009. Warm colours (yellow, orange and red) are for positive Z' values which indicate decreasing rainfall trend due to the land surface intervention. Cold colours (light blue to blue) are for negative Z' values which indicate increasing rainfall trend. The deeper the colour, the more significant the statistic.}
  \label{fig:steptest30}
\end{figure}

The spatial step trend test Z' scores are shown in Figure~\ref{fig:steptest30}. This figure provides two types of information: the sign and the significance level. The sign indicates the direction of the step change, as listed in Table~\ref{tab:Zscore}. In each region, there is a broad area of positive Z' values which implies a decrease in rainfall. In QLD, the areas of positive Z', especially when p is smaller than 0.1, almost align with the locations reported to have heavy land clearing (see Figure~\ref{fig:slat}). However the vegetation cover change map in Figure~\ref{fig:tc_trend}~(a) does not indicate a clear similar pattern. In the Snowy Mountains area and the west of ACT, where severe bushfires occurred in 2003, positive Z' values were also found. It again indicates that loss of tree cover was related to a rainfall decrease. Nevertheless the observed step change in rainfall was not strong. At the 5\% significance level, changes in a number of pixels in the alpine area are possibly where the tree recovery is slow. In the QLD region, 35 pixels obtained a Z' score with p $<$ 5\%. In the NSW/VIC region, there are only 18 pixels in the alpine area having a Z' score with p $<$ 5\%. They are a very small proportion of both study regions.
As the variability of the monthly rainfall, and the variability of rainfall in Australia is considerable \citep{Delworth_regional_2014}, detecting significant trends at p $<$ 5\% could be difficult. Therefore, an analysis based on p $<$ 10\% is presented in the figures. At p = 10\%, 278 pixels in the QLD region had a significant positive Z' score, while 142 pixels in the NSW/VIC region were significant. In both cases this was a bit over 10\% of the total area.

%figure 14
\begin{figure}[ht!]
  \centering
  \includegraphics[scale=0.6]{slat.jpg}
  \caption{Woody vegetation clearing rate in QLD. The maps and figures are obtained from \citet{SLATS2004}. The red rectangle is the boundary of the QLD study region.}
  \label{fig:slat}
\end{figure}

%figure 15
\begin{figure}[ht!]
  \centering
  \includegraphics[scale=0.8]{compare_mean.pdf}
  \caption[Boxplots of annual rainfall (estimated based on rainfall residuals from Equation~\ref{eqn:model}) before and after the land cover intervention during 1998 - 2008 in the study regions.]{Boxplots of annual rainfall (estimated based on rainfall residuals from Equation~\ref{eqn:model}) before and after the land cover intervention during 1998 - 2008 in the study regions. On average, the after period has a lower annual rainfall amount and with outliers of small values.}
  \label{fig:mean_diff}
\end{figure}

The rainfall residuals of the two periods (before-change since 1998 and after-change) were also compared (Figure~\ref{fig:mean_diff}). The annual values of the 11-year period (1998 - 2008) were calculated for each pixel. The boxplots indicate that there were different mean values between the ``before" and ``after" periods in both regions. There are small outliers in the after period in NSW/VIC. A close look at the rainfall data revealed that rainfall was consistently very low in 2006 for the whole region. The low rainfall in 2006 was due to a weak El Ni\~{n}o and memory effects of the previous drought. While the regression model has removed most of the effect of ENSO, the  cumulative drought effect could still be visible in the residuals if the response of the rainfall to the ENSO effect is non-linear, and the memory of the past drought is persistent. This can also explain the significant results in the previous method. To remove the outlier effects, the two periods were compared excluding the 2006 rainfall. An F test confirmed that the variances were different between the before and after periods. Hence an unpaired unequal variance two sample t-test was applied to test whether the after period has lower mean rainfall than the before period, for the group of pixels showing a negative step change in rainfall ($p = 0.05$) and the group of pixels with no change. The rainfall in the after period was lower than the before period (p $<$ 0.05) for pixels showing a negative step change. For those pixels without a step change, there was no statistical evidence that the two periods have different means. Hence the t-test results are consistent with the step trend test results.

The choice of $ns$ has some impact on the test results, as shown by \citet{Hirsch1985}. The cases of $ns = 1$ and $ns = 9$ were also tested. When $ns = 1$, 18 pixels in the QLD region obtained a Z' score at the 5\% significance level. In the NSW/VIC region, the detection of negative step change ($p = 0.05$) reduced to 16 pixels. On the other hand, when $ns = 9$, the results were similar to the case of $ns = 4$. In this case, the changes ($p = 0.05$) were detected in 35 pixels in the QLD region and 18 pixels in the NSW/VIC region. The power of the test does not change much after $ns = 4$, as shown by \citet{Hirsch1985}. The cause of the somewhat inconsistent results in the QLD region between $ns = 1$ and $ns = 4$ is unclear. This could possibly be because the higher $ns$ values smooth the individual pixel results.

The "field significance" of the test is considered to make inferences about the step change at regional scales from multiple local tests \citep{Wilks2006,Westra2013}. Here, the bootstrapping resampling method from \citet{Westra2013} was modified to evaluate the field significance. The spatial structure of the pixels was maintained, while the order of the years and months, were changed by random resampling. The test statistic identifies the percentage of the pixels with significant step change, positive and negative respectively, for the step trend test. The test statistics on 1000 resampled replicates were used to develop the distribution of these percentage values under the local null hypothesis that there was no step change.

The bootstrapping resampling technique was applied on both $ns=1$ and $ns=4$ cases. In the $ns=4$ case, a spatial moving block bootstrapping was used, in which the change of time sequence within the 2$\times$2 block was consistent. The distribution of the test statistics is highly concentrated at zero with a skew to the right. In the outputs, negative step changes in the cases of $ns=1$ and $ns=4$, the test statistics on the observed time series generally fall within the null hypothesis distribution (the cases of $ns=4$ are shown in Figure~\ref{fig:boot_qld} and Figure~\ref{fig:boot_nsw}). However, the bootstrap distribution is very wide and highly skewed, with step change values close to 0. The result for both the positive Queensland and positive NSW/VIC $ns=4$ step change tests are located much further to the right on the tail of the disctribution, away from 0 and give a stronger indication of a change. The results showed that overall the chance of detecting a rainfall change was small, which is related to the strong natural variability of the rainfall in Australia. However, the location of the positive tests on the tail of the distribution suggests some support for the hypothesis that vegetation cover change affects local rainfall.

%figure 16
\begin{figure}[ht!]
\centering
\includegraphics[scale=0.7]{bt4_qld.pdf}
\caption{Percentage of pixels showing statistically significant positive Z' values (left) and negative Z' values (right) in QLD at  p $<$ 0.1. The histogram shows the distribution of results from 1000 bootstrap resampling of the rainfall time series. The red dot represents the results from the observed data.}
\label{fig:boot_qld}
\end{figure}

%figure 17
\begin{figure}[ht!]
\centering
\includegraphics[scale=0.7]{bt4_nsw.pdf}
\caption{Percentage of pixels showing statistically significant positive Z' values (left) and negative Z' values (right) in NSW/VIC at p $<$ 0.1. The histogram shows the distribution of results from 1000 bootstrap resampling of the rainfall time series. The red dot represents the results from the observed data.}
\label{fig:boot_nsw}
\end{figure}

\section{Summary and Discussion}

Generally, empirical studies on LCC-precipitation interaction are conducted within an area with known land surface intervention \citep[e.g.][]{Otterman1990,Durieux2003,Negri2004,Sato2007}. However, these locations are rare and difficult to isolate from real landscape change. In this study we therefore tested the effect of land cover change across a broad region, rather than only for locations where changes were known to occur or have occured. The advantage of this approach is that it does not require a long time series of land cover data which is usually unavailable. Furthermore, it does not assume a specific relationship between vegetation cover change and rainfall but allows the data to show this relationship, by applying the analysis to a broader area outside the boundary of the vegetation cover change. This approach is expected to provide a way to reduce the risk of a false positive paradox, by comparing results between areas with and without vegetation cover change.

Parametric tests are generally more powerful than nonparametric test in detecting a trend, when the data is normally distributed \citep{Onoz2003,Kundzewicz2004}. As a non-parametric test, the step trend test has the advantages of distribution free and having no restriction on missing data \citep{Hirsch1985}. This is particularly useful in rainfall analysis since rainfall data is usually skewed. On the other hand, the disadvantages of non-parametric tests, such as being limited to hypothesis testing and weaker in power, also hold for the step trend test \citep{Whitley2002}.

The regression model used here is a simple model. We only consider the important effects of historical trend, seasonality and climate drivers. Furthermore, no more than two large scale climatic indices were used in order to avoid over-parameterisation and multiple cross-correlations between climatic indices. A purpose of the regression model is to remove the variability in rainfall that is due to these known important factors. The model shows that seasonality, ENSO and IOD together explain no more than 20\% of the rainfall variability, and around 6 - 8\% on average is due to the climate drivers. This is consistent with the literature \citep[e.g.][]{Westra2010}. A large amount of variation is left in the model residuals, which is likely due to random factors. Rainfall is generally considered a stochastic process \citep[e.g.][]{Fowler2005,Cowpertwait2009,Burton2010}. The high variability increases the difficulty to detect a change in rainfall.

There was some evidence that a step change has occurred in the QLD region. The semi-parametric model could not identify any step change in the rainfall data. On the other hand, the step trend test suggested a couple of locations (300 - 500 km$^2$) might have experienced rainfall change ($p = 0.05$). The results from the step trend test also indicated a possible widespread change, but the spatial pattern of change is not consistent with the tree cover change map. Although land clearings in QLD have occurred at a high rate and broad scale \citep{SLATS2005}, this study cannot confirm its impact on the local rainfall. At this location, land cover change was not found to change the mean rainfall significantly in some other studies \citep[e.g.][]{Narisma2003,McAlpine2007}. The characteristics of vegetation cover changes in QLD might increase the difficulty in detecting a step change. QLD has a long history of land clearing. According to the series of SLATS reports on land cover changes in QLD released by the Queensland government, land clearing continued in and around the study region between 1988 - 2008. Major broad scale and high rate clearings occurred in 1999 - 2000 and 2002 - 2004. It is therefore difficult to define a clear cut change in this region. The more continuous ongoing land clearing could have reduced the significance of a step change. 
The two methods indicate quite different results on the level of change in the NSW/VIC region. The semi-parametric model showed that a large area in the NSW/VIC region has experienced significant step change ($p = 0.05$) in rainfall after 2003. But this result is likely to be affected by the low rainfall in 2006 as the area is larger than the bushfires region. The step trend test was able to detect significant changes ($p = 0.05$) in about 400 km$^2$ areas in the Snowy Mt. and some possible changes along the bushfires region ($p = 0.1$). This might be due to the possible small size of the step change and/or a short after-fire period \citep{Hirsch1985}. The vegetation cover change due to bushfires might have possibly changed the rainfall but the evidence was not strong either.

The bushfires locations highlighted in the analysis results is an interesting outcome. The results might be due to a drought effect. The 2002 - 2003 drought has affected rainfall in a large part of Australia, particularly the MDB \citep{Nicholls2004}. The severe bushfires in 2003 were also triggered by the extreme drought conditions. Although the dry episode effect on rainfall has been accounted for in part by the SOI, further impact of drought could be passed through the local land-atmosphere interaction. However, this impact might not be statistically significant, even under an extreme event such as bushfires. Overall, the rainfall feedback to the vegetation cover change could be weak under dry conditions.

The different causes of vegetation cover change in these two regions could lead to different post-change characteristics. The magnitude of EVI decreasing trends in the QLD region are less than in NSW region, as reported in the DLCD data. This is due to the lower tree density in the QLD region than in the NSW/VIC region before land surface interventions. Wild fires might have totally damaged the vegetation cover and recovery was very slow in some areas within the Snowy Mt. The persistent drought in the 2000s \citep{Howden2012} has delayed the regrowth of trees. On the other hand, replacing tree cover with pasture and crops might have a relatively subtle impact on the EVI. On the other hand, regrowth of trees can be expected after fires or as drought condition is relieved; but clearings for agricultural purpose impose permanent or semi-permanent changes to the land surface.

The rainfall data used in this study is a gridded data set. This data set is robust and consistent over a long time series (from 1900 to current) and has a broad national wide coverage which can provide more information spatially. However, high cross correlation between pixels due to the interpolation method generating this data set can also introduce spatial noise. Here the cross correlation has been account for in the step trend test. Some other methods are also available which can be used to perform a comparative trial. For example, \citet{Narisma2007} applied a spatial Gaussian filter on a similar data set and used wavelet analysis to detect step change in rainfall. High quality station data is another option to test whether the observed spatial pattern in the step trend test results was not due to the gridded data itself. Resampling methods, such as bootstrapping and permutation \citep{Wilks1997,Kundzewicz2004,Westra2013}, can also be used to further assess the strength of significance of results and incorporate spatial and temporal patterns in the analysis. We are also aware that the gridded data set is most useful in regions with sparse rain gauge networks but it actually reduces information where the rain gauge density is high \citep{Jones2009}. In the Snowy Mountains area, the coverage of rainfall stations is intensive but they are mainly located in the valleys. The interpolated data might not best represent the local rainfall.

Overall, the inclusion of the anomalous rainfall year 2006, could have influenced the results, and this means that strong conclusions cannot be drawn.

However, the current study provides some evidence to reject the null hypothesis (no step change in rainfall is due to the tree cover loss). Limited by the available data, the time frame under study was chosen within a long lasting drought period \citep{Holper2011}. The strong impact of this prolonged drought might have suppressed the land-atmosphere interaction and confused the cause and effect relation between rainfall and vegetation cover change. This could be one of the reasons that the LCC effects found in other studies \citep[e.g.]{Gorgen2006,McAlpine2007} are not found to be significant here. So possible future work can be conducted for a non-drought period, when a longer series of land cover data is available. This approach should be further assessed in different scenarios of wet period and/or afforestation. Continuous monitoring of land surface conditions is important for future research. The power of the test can also be improved with the longer length of the after-intervention period \citep{Hirsch1985}. But then the question is what is an appropriate length for land cover change study, when ongoing change is possible, such as tree regrowth or cropping. This will require further research.

\section{Conclusion}

In this study, we found some observationl data based evidence, although not strong, that vegetation cover change has changed local rainfall. The semi-parametric method and non-parametric method did totally agree on detecting a significant step change in rainfall in the hot dry QLD region where land clearing has occurred. On the other hand, the bushfires in the humid, temperate mountain range in the NSW/VIC region has experienced reduced rainfall. But the dry spell also plays an important role in the results.

Drought has had a pronounced impact on the land surface condition during the study period, leading to significant reduction in vegetation and extreme events such as bushfires. The associated lack of rainfall and high temperatures may mask the step change in the vegetation. Hence, the signal of LCC feedback on rainfall is probably weaker under such regional dry conditions, as the impact of LCC on rainfall is mainly through changes in moisture convergence \citep{Gorgen2006,Pitman2007}.

\section{acknowledgments}
CL was supported by an Australian Postgraduate Award for this work.

\appendix

\section{Summary of Data}

 \begin{sidewaystable}
 \caption{Summary of data.}
  \label{tab:ch3Data}
 \begin{tabular}{lllll}
  \hline
  \textbf{Data} & \textbf{Source} & \multicolumn{2}{c}{\textbf{Resolution}} & \textbf{Analysis period} \\\cline{3-4}
  & & Temporal & Spatial & \\\hline
  Percent tree cover & MOD44B & Annual &	250m & 2000-2010\\
  Trend of vegetation cover change  &	DLCD (2009)	& Onetime & 250m	& Trend of Apr 2000 - Apr 2008\\
  Rainfall &	AWAP gridded rainfall data &	Monthly &	0.05\textdegree$\times$0.05\textdegree & Jan 1979- Dec 2008\\
  SOI	& BoM	& Monthly &	N/A &	Jan 1979- Dec 2008\\
  NINO 3, 3.4, 4 &	IRI/LDEO data library	& Monthly	& N/A	& Jan 1979- Dec 2008\\
  PDO	& NOAA & Monthly	& N/A	& Jan 1979- Dec 2008\\
  IOD	& POAMA-2 dataset	& Monthly	& N/A	& Jan 1979- Dec 2008\\
  \hline
  \end{tabular}
\end{sidewaystable}

\newpage

%\bibliography
\bibliographystyle{elsarticle-harv} 
\bibliography{reflist_chap3}

\end{document}

### 2017-11-08 LandSurfaceEffects_LocalRainfall_ClimDynsubmit.Rnw
add files

--- 2017-11-08 LandSurfaceEffects_LocalRainfall_AGRFORMETsubmit.Rnw
@@
-\documentclass[authoryear,preprint,review,12pt]{elsarticle}
+\RequirePackage{fix-cm}
 
-\journal{Agriculture and Forest Meteorology}
+\documentclass[onecolumn,referee]{svjour3}
+\journalname{Climate Dynamics}
 
 \usepackage{url}
-% Elsevier suggests
-\usepackage{amssymb}
+% Springer suggests
+\usepackage{graphicx}
+\usepackage{natbib}
+\bibliographystyle{spbasic} 
 \usepackage{lineno}
@@
 \usepackage[T1]{fontenc}
-%switch on line numbers
+
 \linenumbers
@@
 \SweaveOpts{concordance=TRUE}
-\begin{frontmatter}
 
@@
 
-\author{C. Liang}
-\author{R.W. Vervoort \corref{cor1}}
-\ead{willem.vervoort@sydney.edu.au}
-\author{F.F. Van Ogtrop}
+\author{C. Liang \and R.W. Vervoort \and F.F. Van Ogtrop}
 
-\address{Sydney Institute of Agriculture, School of Life and Environmental Sciences, The University of Sydney}
-\cortext[cor1]{}
+\institute{C.Liang \at Sydney Institute of Agriculture, School of Life and Environmental Sciences, The University of Sydney \\
+  \email{summer.chunliang@gmail.com}
+  \and
+  R.W. Vervoort \at Sydney Institute of Agriculture, School of Life and Environmental Sciences, The University of Sydney \\
+  Tel: +61 86271054 \\
+  \email{willem.vervoort@sydney.edu.au}
+   \and
+   F.F. van Ogtrop \at Sydney Institute of Agriculture, School of Life and Environmental Sciences, The University of Sydney \\
+            \email{F.vanogtrop@sydney.edu.au}}
 
-\begin{itemize}
-\item There are limited studies that analyse observational data of vegetation feedback on rainfall
-\item Two sites with major step trend landuse change were analysed for changes in rainfall
-\item Statistical change in rainfall detected after bushfire, but not after land clearing
-\end{itemize}
+\date{Received: date / Accepted: date}
+% The correct dates will be entered by the editor
+
+\maketitle
 
@@
 Analysis of observational data to identify relationships between rainfall and land cover change are scarce due to multiple environmental factors that cannot be controlled.
-In this study we investigated the relationship using statistical methods on data from best available sources at two sites in Australia.
+In this study we investigated the relationship using statistical methods on data from best available sources at two sites (approx 306 by 274 km in Queensland) and approx 203 by 175 km in NSW) in Australia.
 Gridded data of rainfall and tree cover were used as spatially corresponding local conditions.
-Large scale effects were represented by climatic indicators, such as SOI and IOD.
+Large scale effects affecting the regional space were represented by climatic indicators, such as SOI and IOD.
 Regression analysis and step trend tests were used to assess the effect of abrupt land surface intervention.
@@
 
-\begin{keyword}
-Observed data analysis \sep Rainfall change \sep Landcover feedback
-\end{keyword}
-
-\end{frontmatter}
+\keywords{Observed data analysis \and Rainfall change \and Landcover feedback}
 
@@
 This is because there are some fundamental experimental difficulties in both space (where does evaporated water reappear as rainfall?) and in time (how much time does it take for land cover change effects to appear or disappear?).
-Therefore, the aim of this study is to use observational data at regional scales to investigate the cause and effect relationship between land cover change and local rainfall using empirical evidence.
+This might explain why modelling is mostly used to be able to deal with both the convergent and divergent nature of the regional process \citep{CEBON2000211}.
+Therefore, the aim of this study is to use observational data at regional scales to investigate the cause and effect relationship between land cover change and local rainfall.
 More specifically, we hypothesize that a step change on the land cover on the surface will cause a step change in the rainfall and that this can be identified statistically.
@@
 \end{linenomath*}
-
 Vegetation cover changes occurred at different times in the two regions.
@@
 
+In all regional studies, it is difficult to separate the different influences \citep{CEBON2000211}, and this study also does not capture all the different convergent and divergent effects.
 The regression model used here is a simple model.
@@
 \newpage
-
 %\bibliography
-\bibliographystyle{elsarticle-harv} 
 \bibliography{reflist_chap3}

### 2017-11-08 LandSurfaceEffects_LocalRainfall_WRRsubmit.Rnw
add files

--- 2017-11-08 LandSurfaceEffects_LocalRainfall_AGRFORMETsubmit.Rnw
@@
-\documentclass[authoryear,preprint,review,12pt]{elsarticle}
+\documentclass[draft,linenumbers]{agujournal}
+\draftfalse
 
-\journal{Agriculture and Forest Meteorology}
+\journalname{Water Resource Research}
 
 \usepackage{url}
-% Elsevier suggests
-\usepackage{amssymb}
- \usepackage{lineno}
-
-% older suggested by Chun
+%\usepackage[]{color}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 \usepackage{bm}
-% french characters
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-%switch on line numbers
-\linenumbers
 
@@
 \SweaveOpts{concordance=TRUE}
-\begin{frontmatter}
 
-\title{Observational data analysis of land surface effects on local rainfall}
+\title{Empirical study of land surface effects on local rainfall}
 
-\author{C. Liang}
-\author{R.W. Vervoort \corref{cor1}}
-\ead{willem.vervoort@sydney.edu.au}
-\author{F.F. Van Ogtrop}
+\authors{C. Liang \affil{1}, R.W. Vervoort \affil{1}, F.F. Van Ogtrop \affil{1}}
 
-\address{Sydney Institute of Agriculture, School of Life and Environmental Sciences, The University of Sydney}
-\cortext[cor1]{}
+\affiliation{1}{Sydney Institute of Agriculture, School of Life and Environmental Sciences, The University of Sydney}
+\correspondingauthor{R.W. Vervoort}{willem.vervoort@sydney.edu.au}
 
-\begin{itemize}
-\item There are limited studies that analyse observational data of vegetation feedback on rainfall
-\item Two sites with major step trend landuse change were analysed for changes in rainfall
-\item Statistical change in rainfall detected after bushfire, but not after land clearing
-\end{itemize}
+\begin{keypoints}
+\item There are few empirical studies of vegetation feedback on rainfall 
+\item Two sites with major step trend landuse change were analysed
+\item Statistical change in precipitation detected after bushfire
+\end{keypoints}
 
 \begin{abstract}
-Analysis of observational data to identify relationships between rainfall and land cover change are scarce due to multiple environmental factors that cannot be controlled.
+Empirical evidence of correlations between rainfall and land cover change are scarce due to multiple environmental factors that cannot be controlled.
 In this study we investigated the relationship using statistical methods on data from best available sources at two sites in Australia.
@@
 
-\begin{keyword}
-Observed data analysis \sep Rainfall change \sep Landcover feedback
-\end{keyword}
-
-\end{frontmatter}
-
 \section{Introduction}
@@
 Land use and land cover changes can lead to changes in the local climate.
-Some observational and many modelling studies have found cloud types and rainfall are correlated to large scale vegetation cover changes, such as deforestation in the Amazon and in the Sahel \citep{Chagnon2005,Pinto2009, Wang2009, Mei2010, kucharski_further_2013,pitman_scale_2016} and afforestation in south Israel \citep{Otterman1990, Ben-Gai1998}.
-In terms of observational studies, using airborne measurement in Western Australia, \citet{Junkermann2009} showed a significantly higher level of aerosols over an agricultural area compared to an adjacent natural vegetation.
+Both empirical and modelling studies have found cloud types and rainfall are correlated to large scale vegetation cover changes, such as deforestation in the Amazon and in the Sahel \citep{Chagnon2005,Pinto2009, Wang2009, Mei2010, kucharski_further_2013,pitman_scale_2016} and afforestation in south Israel \citep{Otterman1990, Ben-Gai1998}.
+Using airborne measurement in Western Australia, \citet{Junkermann2009} showed a significantly higher level of aerosols over an agricultural area compared to an adjacent natural vegetation.
 They suggested that a modification of aerosol concentrations due to deforestation could have contributed to a reduction of local rainfall, as more but smaller rain droplets were observed.
@@
 
-Rainfall over land is generally influenced by multiple factors as has been shown from multiple simulation studies.
+Rainfall over land is generally influenced by multiple factors.
 Locally, there are two main sources: moisture from advective atmospheric transport; and local evapotranspiration \citep{Eltahir1996,Bosilovich2006,Dirmeyer2009,Gimeno2010}.
@@
 
-Overall the number of observational data studies related to changes to rainfall due to land cover change is limited.
+Overall the number of empirical studies related to changes to rainfall due to land cover change is limited.
 This is because there are some fundamental experimental difficulties in both space (where does evaporated water reappear as rainfall?) and in time (how much time does it take for land cover change effects to appear or disappear?).
-Therefore, the aim of this study is to use observational data at regional scales to investigate the cause and effect relationship between land cover change and local rainfall using empirical evidence.
+Therefore, the aim of this study is to use empirical evidence at regional scales to investigate the cause and effect relationship between land cover change and local rainfall using empirical evidence.
 More specifically, we hypothesize that a step change on the land cover on the surface will cause a step change in the rainfall and that this can be identified statistically.
@@
 \centering
-\includegraphics[scale=0.7]{bt4_qld.pdf}
+\includegraphics[scale=0.8]{bt4_qld.pdf}
 \caption{Percentage of pixels showing statistically significant positive Z' values (left) and negative Z' values (right) in QLD at  p $<$ 0.1.
@@
 \centering
-\includegraphics[scale=0.7]{bt4_nsw.pdf}
+\includegraphics[scale=0.8]{bt4_nsw.pdf}
 \caption{Percentage of pixels showing statistically significant positive Z' values (left) and negative Z' values (right) in NSW/VIC at p $<$ 0.1.
@@
 
-In this study, we found some observationl data based evidence, although not strong, that vegetation cover change has changed local rainfall.
-The semi-parametric method and non-parametric method did totally agree on detecting a significant step change in rainfall in the hot dry QLD region where land clearing has occurred.
-On the other hand, the bushfires in the humid, temperate mountain range in the NSW/VIC region has experienced reduced rainfall.
+In this study, we found some evidence, although not strong, that vegetation cover change has changed local rainfall.
+The semi-parametric method and non-parametric method did not reach an agreement on detecting significant step change in rainfall in the hot dry QLD region where land clearing has occurred.
+On the other hand, the bushfires in the humid temperate mountain range in the NSW/VIC region could have reduced rainfall.
 But the dry spell also plays an important role in the results.
@@
 Drought has had a pronounced impact on the land surface condition during the study period, leading to significant reduction in vegetation and extreme events such as bushfires.
-The associated lack of rainfall and high temperatures may mask the step change in the vegetation.
-Hence, the signal of LCC feedback on rainfall is probably weaker under such regional dry conditions, as the impact of LCC on rainfall is mainly through changes in moisture convergence \citep{Gorgen2006,Pitman2007}.
-
-\section{acknowledgments}
-CL was supported by an Australian Postgraduate Award for this work.
+The lack of rainfall associated and high temperature may cause a step change in the vegetation.
+Hence, the signal of LCC feedback on rainfall is probably weaker under such regional dry condition, as the impact of LCC on rainfall is mainly through changes in moisture convergence \citep{Gorgen2006,Pitman2007}.
 
@@
 \newpage
+\begin{acknowledgments}
+CL was supported by an Australian Postgraduate Award for this work.
+\end{acknowledgments}
 
 %\bibliography
-\bibliographystyle{elsarticle-harv} 
 \bibliography{reflist_chap3}

### 2017-11-08 chap3_paper_WVversion.Rnw
add files

\documentclass[draft,linenumbers]{agujournal}
\draftfalse
\journalname{Water Resource Research}

%\usepackage{agu-ps}
%\usepackage[dvipdf]{graphicx}
%\setkeys{Gin}{draft=false}
%\usepackage{lineno}
%\linenumbers*[1]

\usepackage{url}
%\usepackage[]{color}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\usepackage{rotating}
\usepackage{bm}

%% ------------------------------------------------------------------------ %%
%
%  ENTER PREAMBLE
%
%% ------------------------------------------------------------------------ %%

\authorrunninghead{LIANG ET AL.}
\titlerunninghead{EMPIRICAL LAND COVER-RAINFALL STUDY}

%Corresponding author mailing address and e-mail address:
\authoraddr{Corresponding author: R.W. Vervoort, Sydney Institute of Agriculture, School of Life and Environmental Sciences, The University of Sydney,
 1 Central Avenue, Eveleigh NSW 2015, Australia.
(willem.vervoort@sydney.edu.au)}

\begin{document}
\SweaveOpts{concordance=TRUE}

\title{Empirical study of land surface effects on local rainfall}

\authors{C. Liang, F. Van Ogtrop, \altaffilmark{1} R.W. Vervoort, \altaffilmark{1}}

\altaffiltext{1}{ Sydney Institute of Agriculture, School of Life and Environmental Sciences, The University of Sydney}

%% ------------------------------------------------------------------------ %%
%
%  KEYPOINTS
%
%% ------------------------------------------------------------------------ %%

%\keypoints{(Type in Key Points Here)}

%% ------------------------------------------------------------------------ %%
%
%  ABSTRACT
%
%% ------------------------------------------------------------------------ %%

\begin{abstract}
(text here)
\end{abstract}

%% ------------------------------------------------------------------------ %%
%
%  BEGIN ARTICLE
%
%% ------------------------------------------------------------------------ %%

\begin{article}

\section{Introduction}

Land use and land cover changes can lead to changes in the local climate. Both empirical and modelling studies have found cloud types and rainfall are correlated to large scale vegetation cover changes, such as deforestation in the Amazon and in the Sahel \citep{Chagnon2005,Pinto2009, Wang2009, Mei2010, kucharski_further_2013,pitman_scale_2016} and afforestation in south Israel \citep{Otterman1990, Ben-Gai1998}. Using airborne measurement in Western Australia, \citet{Junkermann2009} showed a significantly higher level of aerosols over an agricultural area compared to an adjacent natural vegetation. They suggested that a modification of aerosol concentrations due to deforestation could have contributed to a reduction of local rainfall, as more but smaller rain droplets were observed. \citet{Nair2011} reported from the Bunny Fence Experiment in Western Australia that local land use change altered the synoptic west coast trough dynamics and surface roughness, and this resulted in the observed rainfall decrease. Climate sensitivity to land cover change is also found in eastern Australia \citep{McAlpine2007}.

Rainfall over land is generally influenced by multiple factors. Locally, there are two main sources: moisture from advective atmospheric transport; and local evapotranspiration \citep{Eltahir1996,Bosilovich2006,Dirmeyer2009,Gimeno2010}. According to \citet{Trenberth1999}, the contribution of advective moisture partially depends on the availability of external moisture and atmospheric transport. On the longer time scale, such as monthly and annually, large scale atmospheric dynamics are affected by large scale climate drivers. Many studies have reported significant relationships between rainfall in large parts of Australia and the El Ni\~{n}o-Southern Oscillation (ENSO) \citep{Verdon2004,Risbey2009,Speer2011}. ENSO can be used to represent longer term cycles in rainfall data such as drought. On the other hand, local ET is determined by local land surface characteristics. Local land surface characteristics further influence local scale atmospheric dynamics and hence the amount of rainfall, including contribution from both sources. Therefore land surface plays an important role in local rainfall.

Conventionally, large scale climate drivers are used as predictors for seasonal rainfall forecasts. According to the Australian Bureau of Meteorology (BoM), probablistic seasonal outlooks are developed based on Australian rainfall and temperature as well as sea surface temperature records from the tropical Pacific and Indian Oceans \citep{BoM2012c}. 
%% comment Expand this section with section from methods (section 4.1)?
Different parts of the Australian continent can be more or less influenced by different climatic drivers \citep{Mla2008}. Using both SOI and PDO in a prediction model, \citet{Kamruzzaman2011} reported that PDO was rarely significant for rainfall stations in the MDB and on the southeast coast of Australia, while SOI was at least significant at the 1\% level. In addition to SOI, \citet{Speer2011} found that the observed rainfall decrease in the southeast of NSW was linked to an increasingly positive SAM in 1976 - 1998. In southwest WA, the influence of any oceanic indices is small \citep{Smith2012}.

Although climate drivers demonstrate some capability to predict Australian rainfall, there is still a large amount of unexplained variance. \citet{Westra2010} pointed out that models based on global sea surface temperature anomalies can only predict up to 14.7\% of precipitation variance. Recent studies suggest that land surface processes are important for predicting local rainfall \citep[e.g.][]{Ma2011,Zeng2012, pitman_scale_2016, saha_investigating_2016}. However, they are mostly based on modelling experiments and little evidence was reported from observations. \citet{Pitman2004} found a good match between observations and simulated rainfall changes in southwest Western Australia ,forced by land cover change. \citet{Timbal2006} were able to reproduce the rainfall decline in south west Australia by including land cover influence. Local land use change might not be a primary, but is likely to be a secondary cause of rainfall change \citep{Nicholls2006}. Therefore, land surface modification has, at least partially, contributed to local rainfall variability.

Overall the number of empirical studies related to changes to rainfall due to land cover change is limited. This is because there are some fundamental experimental difficulties in both space (where does evaporated water reappear as rainfall?) and in time (how much time does it take for land cover change effects to appear or disappear?). Therefore, the aim of this study is to use empirical evidence at regional scales to investigate the cause and effect relationship between land cover change and local rainfall using empirical evidence. More specifically, we hypothesize that a step change on the land cover on the surface will cause a step change in the rainfall. To demonstrate the approach and this effect we studied step changes in rainfall at two locations in Queensland and NSW/Victoria where there are possible step changes in land cover change due to land clearing and bushfires based on tree cover data. Statistical approaches were used to identify the step changes in rainfall, which were subsequently associated with land cover change through spatial comparison.

In this paper, after this section (the introduction), section 2 covers the case study areas and the observed land use changed. Section 3 describes the data used in the study in more detail. Section 4 details the statistical methods and the underlying assumptions related to the modelling approach, Section 5 gives the results, which are further discussed in section 6 and finally section 7 offers the conclusions.

\section{Study regions and tree cover change}

In Australia, significant tree cover change has mainly occurred in the north east of the continent and on the southeast coast, as well as in the southwest of Western Australia. According to the National Dynamic Land Cover Dataset (DLCD) \citep{Lymburner2010}, most of these areas have experienced decreasing EVI between 2000 - 2008. As an index for vegetation greenness, the decreasing values indicate lower biomass over time in the tree cover regions. The possible EVI reduction might be due to land clearing, bushfires or drought.

%figure 1
\begin{figure}[htp]
  \centering
  \includegraphics[bb = 8 1 800 600,scale=0.5]{map_selreg.jpg}
  \caption{Selected study regions are highlighted by red rectangles in the main map (the red rectangle in the insert indicates the location of the main map). The types of tree cover in 2008 from the DLCD product is shown at the background. In site 1 (the QLD region), the tree cover is mostly sparse. In site 2 (the NSW/VIC region), many areas have open or close forest in which tree cover is denser.}
  \label{fig:sel_reg}
\end{figure}

Two regions were selected where significant tree cover change was present. The first region is located in south central Queensland to the north of the Murray Darling Basin (MDB) (site 1 in Figure~\ref{fig:sel_reg}). High rates of land clearing have been reported in this region during the early 2000s \citep{SLATS2004}. The second study region is located at the border of New South Wales and Victoria, and includes the Snowy Mountains ranges (site 2 in Figure~\ref{fig:sel_reg}). Severe bushfires occurred in this area and the surroundings in early 2003 (see Figure~\ref{fig:bushfire}). The 2003 bushfires were the largest and the worst in this area for the last 60 years \citep{Fire2011}. Two thirds of Kosciuszko national park was heavily burned and regrowth was reported to be slow due to drought and cold conditions \citep{ABC2003} and the type of species in this region. In both study regions, significant tree cover loss has happened within the last decade, either permanently or temporarily. It makes the selected areas suitable for this land cover change study. 
%\footnote{Bushfire history, \url{http://www.dse.vic.gov.au/fire-and-other-emergencies/major-bushfires-in-victoria}, viewed on 12/12/2011}
%\footnote{Kosciuszko slow to recover from bushfires, \url{http://www.abc.net.au/news/2003-10-30/kosciuszko-slow-to-recover-from-bushfires/1501722}, viewed on 06/06/2012}

%figure 2
\begin{figure}[htp]
  \centering
  \includegraphics[bb = 1 1 500 500,clip,scale=0.6]{bushfire_nswvic.jpg}
  \caption{Location of bushfires occurring in January 2003, in and around the NSW/VIC study region, as shown by the red pixels. The map shows large area in the Kosciuszko national park has been burned. Some locations in the southwest of ACT have also experienced intensive bushfires.}
  \label{fig:bushfire}
\end{figure}

The two regions have different climate characteristics. The QLD region is partially grassland and partially subtropical, while the NSW/VIC region is mainly within the temperate zone, under the K\"{o}ppen classification. According to Australian Bureau of Meteorology (BoM), the NSW/VIC region receives 1000 - 2000 mm rainfall annually, which is more than double of the rainfall in the QLD region. Evapotranspiration is similar in both regions. Marine moisture and orographic effects are likely to be the main contributors to rainfall in the southeast mountain areas of the NSW/VIC region.

The land use and land cover characteristics in the two regions are also different. In the Queensland region, the tree cover is sparse over most of the area. The MODIS satellite tree cover data (discussed in more detail in section 3) shows that tree cover in this region is generally below 20\% of total ground area. Grazing is the main activity in this region, with over 90\% of land used by the grazing industry \citep{ABARES2010}. Our starting assumption is that the main cause of the EVI  decline over large part of the region is due to land clearing. Tree cover has been cleared at a massive scale over the last decade, especially during 2002 - 2004. The Kosciuszko national park is within the NSW/VIC region. Here tree cover is denser with open or even closed forest (the tree cover distribution is bimodal at 10 - 20\% and 60 - 70\%). The dominant species in the alpine area are Snow Gum and large stand species such as Alpine Ash and Mountain Gum in the sub-alpine area. These trees can reach a great height but they take long time to grow. For example, Alpine Ash would need about 20 years to mature. Although land clearing is not a major issue in this region, it is vulnerable to fires and drought.

Therefore two types of land cover changes were studied. The reports from the Queensland Statewide Land Cover and Trees Study (SLATS) \citep[e.g.][]{SLATS2001,SLATS2003} were used to investigate the time and location of land clearing in the QLD region. The MODIS burned area product, MCD45A1 \citep{Roy2002,Roy2005,Roy2008}, was used to locate bushfires areas in the NSW/VIC region, with a grid resolution of 500 m. MCD45A1 provides monthly burning information on all pixels, which helps to pinpoint an abrupt event. Due to the nature of the different land cover change, the post-change vegetation status in the two regions is expected to be different (see Figure~\ref{fig:tc_simple}).

%figure 3
\begin{figure}[htp]
  \centering
  \includegraphics[bb=10 120 1000 400,scale=0.65]{tc_simple.jpg} 
  \caption{The expected evolution of the land surface after trees have been removed in (a) the QLD region and (b) the NSW/VIC region.}
  \label{fig:tc_simple}
\end{figure}

<<mapping,eval=F, echo = F, results = hide>>=

library(plotKML)
library(raster)
library(rgeos)
library(maptools)
library(fields)
library(RColorBrewer)

## define the boundary of sub-regions
  e1 <- extent(c(1324304.091,1723901.773,-222556.668,80451.265))
  e2 <- extent(c(1346395.141,1534425.527,-1169402.246,-607644.171))

  e11 <- extent(c(145.425, 149.025, -28.475, -25.175))
  e22 <- extent(c(146.675, 149.325, -37.075, -34.675))

  e111 <- extent(c(145.825, 148.625, -28.075, -25.575)) #as(tc_r1,"SpatialPolygons")
  e222 <- extent(c(147.075, 148.925, -36.675, -35.075)) #as(tc_r2,"SpatialPolygons")

## the selected regions for mapping

  aus_reg <- readShapePoly("Data\\Geo\\aus_state2011\\simplify_state11_uproj.shp",proj4string=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0"))
  aus_reg2 <- gSimplify(aus_reg,0.01,topologyPreserve=T)

  mdb_reg <- readShapePoly("Data\\Geo\\MDB\\mdb.shp",proj4string=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0"))
mdb_reg2 <- gSimplify(mdb_reg,0.01,topologyPreserve=TRUE)

  remove(list=c("aus_reg","mdb_reg"))

## Burned area map

fire_0301 <- raster("Data\\Other\\bushfires\\Burnt_Area\\tmpA2003001.burndate.tif")
fire_0301nsw <- crop(fire_0301,e2)
fire_loc <- crop(reproject(fire_0301nsw,CRS=CRS("+proj=longlat +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84")),e222)
remove(list=c("fire_0301","fire_0301nsw"))

  setEPS()
  postscript("bushfire_nswvic.eps")
#  jpg("bushfire_nswvic.jpg")
  par(mar=c(3,3,1,6),xpd=T)
  plot(e222,border="white")  
  image(fire_loc,add=T, breaks=c(0,0.99,366.5,10000),col=c("white","red","white"))
  plot(e222,add=T)
  plot(aus_reg2,border="grey",add=T)
  plot(mdb_reg2,border="grey",lwd=3,add=T)
#   text(1430000,-1000000,"study region",cex=1.2)
  text(149.1, -35.3,"ACT",font=2,cex=1.2)
  text(148.4, -36,"Kosciuszko",font=2,cex=0.9)#,srt=90)
  text(147.5, -36.4,"VIC",font=2,cex=1.2)
  text(148, -35.5,"NSW",font=2,cex=1.2)
  image.plot(zlim=c(0,1),breaks=c(0,0.5,1),lab.breaks=c("","burned \narea",""),legend.only=T,horizontal=F,col=c("red","red"),legend.shrink=0.03,legend.width=1.3,legend.args=list(text="",line=0),lwd.poly=0.1)
legend(149,-36.1,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=0.9)
 dev.off()

@

<<treeCover,eval=F , echo = F, results = hide>>=

## trend of evi (MODIS product) from GA and ABARES

## tree cover data
  mod44_tc <- brick("Data\\Satellite\\MOD44\\B\\mod44_tc_aus.tif")
  tm <- seq(as.Date("2000-03-06"), as.Date("2010-03-06"), "year")
  mod44_tc <- setZ(mod44_tc, tm, "year")

## tree cover in sub-region
  mod_tc_r11 <- crop(mod44_tc,e1)
  mod_tc_r111 <- reproject(mod_tc_r11,CRS=CRS("+proj=longlat +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84"))
  mod_tc_r1 <- crop(mod_tc_r111,e111)

  mod_tc_r22 <- crop(mod44_tc,e2)
  mod_tc_r222 <- reproject(mod_tc_r22,CRS=CRS("+proj=longlat +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84"))
 mod_tc_r2 <- crop(mod_tc_r222,e222)

 remove(list=c("mod44_tc","mod_tc_r11","mod_tc_r111","mod_tc_r22","mod_tc_r222"))

## apply t-test to tree cover data
  fun <- function(x,m) {
    var_x <- sd(x)
    mu_1 <- mean(x[1:3])
    if (m==1) {y <- x[6:11]} else {y <- x[4:11]}   
    mu_2 <- mean(y)
    z <- (mu_2-mu_1)/var_x
    if (!is.na(z) & abs(z)>1.645) {return(y[1]-x[3])} else {return(NA)}
  }

  mod_tc_r1.z <- calc(mod_tc_r1,function(x) fun(x,1))
  mod_tc_r2.z <- calc(mod_tc_r2,function(x) fun(x,2))

  remove(list=c("mod_tc_r1","mod_tc_r2"))

## create plot
  plotclr <- brewer.pal(6,"PiYG")
#"#C51B7D" "#E9A3C9" "#FDE0EF" "#E6F5D0" "#A1D76A" "#4D9221"

  setEPS()
  postscript("tc_qld.eps")
  par(mar=c(3,3,1,7))
  plot(e111)
  image(mod_tc_r1.z,col=plotclr,breaks=c(-70,-50,-20,0,20,50,70),legend=F,add=T)
  plot(e111,add=T)
  plot(mdb_reg2,border="grey",lwd=3,add=T)
  image.plot(zlim=c(-3,3),breaks=c(-3,-2,-1,-0.2,0.2,1,2,3),lab.breaks=c(-70,-50,-20,"no tree","0 or",20,50,70),legend.only=T,horizontal=F,col=c("#C51B7D","#E9A3C9","#FDE0EF","white","#E6F5D0","#A1D76A","#4D9221"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text=expression(paste(Delta,"tree (% ground)",sep="")),line=0.5,cex=0.9,font=2),axis.args=list(cex.axis=0.8),graphics.reset=F)
  dev.off()

  setEPS()
  postscript("tc_nswvic.eps")
  par(mar=c(3,3,1,7))
  plot(e222)
  image(mod_tc_r2.z,col=plotclr,breaks=c(-70,-50,-20,0,20,50,70),legend=F,add=T)
  plot(e222,add=T)
  plot(aus_reg2,border="grey",add=T)
  plot(mdb_reg2,border="grey",lwd=3,add=T)
#   image.plot(zlim=c(-3,3),breaks=c(-3,-2,-1,-0.2,0.2,1,2,3),lab.breaks=c(-70,-50,-20,"no tree","0 or",20,50,70),legend.only=T,horizontal=F,col=c("#C51B7D","#E9A3C9","#FDE0EF","white","#E6F5D0","#A1D76A","#4D9221"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text=expression(paste(Delta,"tree (% ground)",sep="")),line=0.5,cex=0.7,font=2),axis.args=list(cex.axis=0.8))
# legend(149,-36.5,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=0.9)
  dev.off()

setEPS()
postscript("map.eps")
par(mar=c(1.5,1.5,1.5,1.5))
plot(mdb_reg2,border="grey",lwd=3)
plot(aus_reg2,border="grey",add=T)
legend(150.2,-36,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=0.9)
#invisible(text(getSpPPolygonsLabptSlots(aus_reg2), labels=as.character(aus_reg2$NAME), cex=0.7,adj=0.5))
box(lty=1)
plot(e111,col="red",add=T)
plot(e222,col="red",add=T)
text(147.2,-26.8,"QLD",cex=1.5)
text(148,-34.7,"NSW/VIC",cex=1.5)
dev.off()

@

Specifically, this study focused on the effect of 2003 - 2004 land clearings in the QLD region and the effect of 2003 bushfires in the NSW/VIC region. These events are expected to cause a step change in the local rainfall. The actual tree cover change at the pixel level during this time was derived from the 11-year MODIS data (discussed below). The difference of tree cover before and after the land disturbance was tested using a Student's t-test. As the length of the tree cover data is shorter than the length of the rainfall data, earlier land clearings in the QLD region cannot be identified spatially, hence they are excluded from the analysis.

\section{Data}
\label{sec:Data}

Several land surface data sets were used in this study. The main one was the MOD44B product Global Vegetation Continuous Field dataset (version 5). This dataset provides estimates of percent tree cover (percentage of ground surface covered by trees) at a grid resolution of 250 m \citep{Townshend2011}, which is finer then the earlier mentioned burned product MCD45A1. The dataset is available on an annual time interval for the period of 2000 - 2010. The tree cover data was produced from 16-day Terra MODIS Land Surface Reflectance data and Land Surface Temperature \citep{Townshend2011}. The National Dynamic Land Cover Dataset (DLCD) \citep{Lymburner2010} from the Australian Collaborative Land Use Mapping Program (ACLUMP) was used to verify the trend of vegetation cover change calculated from the previous dataset. This dataset, developed by Geoscience Australia and Australian Bureau of Agricultural and Resource Economics and Sciences (ABARES), is the first nationally consistent and thematically comprehensive land cover reference for Australia. The DLCD is based on the 16-day Enhanced Vegetation Index (EVI), again from the MODIS satellite, between April 2000 and April 2008. It also has a grid resolution of 250 m. The dataset provides information on the final land cover types (as in 2008) and estimated trend of EVI statistics (annual mean, maximum and minimum).

% MOD 44 - Vegetation Cover Conversion
% 
% Description:
% The MOD44 suite provides a database of land cover and change. Products are derived from 16-day composites of MODIS reflective and emissive bands at their nominal resolu-tions. The MOD44B Vegetation Continuous Fields provides 250 m sub-pixel estimates of percent woody, herbaceous and bare cover along with leaf type and longevity. The user is able to avoid constraints present in ordinal land cover maps which make a priori decisions regarding class definitions. The MOD44A Vegetative Cover Conversion pro-vides a global alarm of land cover change at a 250 m spatial resolution. Included are per-sistent change via deforestation, burning and inundation

% MOD13A3 - monthly 1 km EVI data

Rainfall data for Australia \citep{Jones2009} was obtained from BoM. The data has been projected onto a national 0.05\textdegree $\times$ 0.05\textdegree grid (approximately 5 km$\times$5 km). This gridded dataset was generated from station observations using an optimised Barnes successive correction technique. The Barnes technique combines a weighted averaging process and defined topographical information to estimate rainfall values between spatial points \citep{BoM2009}. The resulting dataset provides additional information for data-sparse areas like central Australia but reduces information in the data-rich areas, such as southeast Australia where station density is up to 20 per 100 km$^2$. The data is available on a monthly basis from 1900 to current. Here a subset of 30 years (1979 - 2008) was used. The study was conducted on monthly data, as a land cover change effect on annual rainfall might be negligible but can often found to be significant in particular months or seasons \citep[e.g.][]{Otterman1990,Gaertner2001,Semazzi2001,Oleson2004,Deo2009}.

%% comment: in Lo2007 stated the same dataset but daily and refer the technique to Mills (1997)

Large scale climate drivers are represented by various climatic indices. The Southern Oscillation Index (SOI) is generally regarded as a good predictor of Australian rainfall \citep{Risbey2009,Chowdhury2010,Westra2010}, but its skill is weaker in some parts of Australia. For example the Southern Annular Mode (SAM) is found to be more important than ENSO in south Western Australia \citep{Meneghini2007}. The suitability of each index for the regions of interest was tested in section \ref{sec:reg_model}. The following climate indices were used as candidate predictors for local rainfall. 
\begin{itemize}
  \item Southern Oscillation Index (SOI). The Troup version of the monthly SOI series used in this study was obtained from BoM (available online at \url{http://www.bom.gov.au/climate/current/soihtm1.shtml}).
  \item Eastern, East Central and Central Tropical Pacific Sea Surface Temperatures (NINO 3, NINO 3.4 and NINO 4). Monthly SST anomalies are available from IRI/LDEO data library and the extended NINO dataset is used (available online at \url{http://iridl.ldeo.columbia.edu/SOURCES/.Indices/.nino/.EXTENDED/}).
  \item Pacific Decadal Oscillation (PDO)\footnote{Pacific Decadal Oscillation is the leading principal component of monthly SST anomaly in the North Pacific Ocean.}. The monthly PDO series was provided by JISAO (Joint Institute for the Study of the Atmosphere and Ocean, University of Washington) (available online at \url{http://jisao.washington.edu/pdo/PDO.latest}).
  \item The interaction of PDO and SOI (PDO$\times$SOI) \citep{Kamruzzaman2011}.
  %% comment: should we cite Cai et al here?
  \item Indian Ocean Dipole (IOD)\footnote{Indian Ocean dipole is commonly measured by the difference between SST anomaly in the western (50 - 70\textdegree E and 10\textdegree S-10\textdegree N) and eastern (90 - 110\textdegree E and 0 - 10\textdegree S) equatorial India Ocean \citep{Saji1999}.}. Monthly IOD was obtained from JAMSTEC (the Japan Agency for Marine-Earth Science and Technology) (available online at \url{http://www.jamstec.go.jp/frcgc/research/d1/iod/DATA/dmi.monthly.txt}).
%     \item Antarctic Oscillation (AAO). Monthly AAO series was also available from NOAA (available online at \url{http://www.esrl.noaa.gov/psd/data/climateindices/list/}).
\end{itemize}
%\footnote{BoM, 2011, \url{http://poama.bom.gov.au/data_server.shtml}, viewed on 3/1/2012}
% IOD: Saji1999

<<climaticIndex,eval=F , echo = F, results = hide>>=

# setwd("R:\\AVSproject\\WrittenWork\\PhD_thesis\\chap3")
# setwd("B:/My Study/PhD/AVSproject_laptop/WrittenWork/PhD_thesis/chap3")
#setwd("B:/My Study/PhD/AVSproject_PhD/WrittenWork/Paper/Paper 1 - chapter 3")

library(reshape)
library(chron)
library(fields)
library(rgdal)
library(raster)
library(maptools)
library(graphics)
library(zoo)
library(RColorBrewer) # creates nice color schemes
# library(classInt)     # finds class intervals for continuous variables

library(mgcv)
library(ncdf)
library(nnet)

## climatic indicator data

file1 <- "Data\\BoM\\SOI\\SOI.csv"
SOI <- read.csv(file1,sep=",",skip=6)
SOI <- melt(SOI,id="Year")
names(SOI) <- c("Year","Month","SOI")
SOI <- SOI[order(SOI$Year,SOI$Month),]
SOI_ts <- ts(SOI$SOI,start=c(1876,1),end=c(2014,4),frequency=12)
SOI_100a <- window(SOI_ts,start=c(1900,1),end=c(2008,12))
SOI_30a <- window(SOI_ts,start=c(1979,1),end=c(2008,12))
## deseasonalised SOI
SOI_100 <- stl(SOI_100a,"period")$time.series[,3]+stl(SOI_100a,"period")$time.series[,2]
SOI_30 <- stl(SOI_30a,"period")$time.series[,3]+stl(SOI_30a,"period")$time.series[,2]

file2 <- "Data\\BoM\\SST\\SST34.cdf"
SST34 <- open.ncdf(file2)
SST34 <- get.var.ncdf(SST34,SST34$var[[1]])
SST34_ts <- ts(SST34,start=c(1856,1),end=c(2014,4),frequency=12)
SST34_30a <- window(SST34_ts,start=c(1979,1),end=c(2008,12))
## deseasonalised SST34
SST34_30 <- stl(SST34_30a,"period")$time.series[,3]+stl(SST34_30a,"period")$time.series[,2]

file3 <- "Data\\BoM\\SST\\SST3.cdf"
SST3 <- open.ncdf(file3)
SST3 <- get.var.ncdf(SST3,SST3$var[[1]])
SST3_ts <- ts(SST3,start=c(1856,1),end=c(2014,4),frequency=12)
SST3_30a <- window(SST3_ts,start=c(1979,1),end=c(2008,12))
## deseasonalised SST34
SST3_30 <- stl(SST3_30a,"period")$time.series[,3]+stl(SST3_30a,"period")$time.series[,2]

file4 <- "Data\\BoM\\SST\\SST4.cdf"
SST4 <- open.ncdf(file4)
SST4 <- get.var.ncdf(SST4,SST4$var[[1]])
SST4_ts <- ts(SST4,start=c(1856,1),end=c(2014,5),frequency=12)
SST4_30a <- window(SST4_ts,start=c(1979,1),end=c(2008,12))
## deseasonalised SST4
SST4_30 <- stl(SST4_30a,"period")$time.series[,3]+stl(SST4_30a,"period")$time.series[,2]

file5 <- "Data\\BoM\\PDO\\PDO.latest_1900-.txt"
pdo_mth <- read.csv(file5,sep="",header=T)
pdo_mth <- melt.data.frame(pdo_mth,id="YEAR")
names(pdo_mth) <- c("Year","Month","PDO")
pdo_mth <- pdo_mth[order(pdo_mth$Year,pdo_mth$Month),]
PDO_ts <- ts(pdo_mth$PDO,start=c(1900,1),end=c(2014,4),frequency=12)
PDO_100a <- window(PDO_ts,start=c(1900,1),end=c(2008,12))
## deseasonalised PDO
PDO_100 <- stl(PDO_100a,"period")$time.series[,3]+stl(PDO_100a,"period")$time.series[,2]

file7 <- "Data\\BoM\\IOD\\iod_jamstec.txt"
IOD <- read.csv(file7,sep="",header=T)
IOD <- IOD[,4]
IOD_ts <- ts(IOD,start=c(1870,1),frequency=12)
IOD_30a <- window(IOD_ts,start=c(1979,1),end=c(2008,12))
## deseasonalised IOD
IOD_30 <- stl(IOD_30a,"period")$time.series[,3]+stl(IOD_30a,"period")$time.series[,2]
# end of read data

@

\section{Statistical method}

As shown in Figure~\ref{fig:ts_mean}, a step change is not obvious in the time series data, even though the data is deseasonalised and detrended. Hence the step changes on rainfall were analysed by two different statistical methods to provide a comparison. Both methods make use of a regression model to remove variability in rainfall hypothesised to be caused by factors other than vegetation cover change. In the first method, the tree cover change was implemented as a factor variable in the regression model. In the second method, a rank sum test (step trend test), was applied to the model residuals after effects of other major factors have been removed, assuming that vegetation cover change is the only factor explaining the non-random pattern in the rainfall residuals.

%figure 4
\begin{figure}[htp]
  \centering
  \includegraphics[scale=0.6]{mean_ts.jpg}
  \caption{The deseasonalised and detrended rainfall over the 30 years period in (a) the QLD region and (b) the NSW/VIC region. The vertical red lines indicate the year of 2003, in which the studied land cover changes occurred. A change in the time series data is not obvious before and after the land cover changes.}
  \label{fig:ts_mean}
\end{figure}

\subsection{Regression model}
\label{sec:reg_model}

Rainfall, even a single event, is generally affected by several factors. Many studies suggest the large-scale climate drivers, which are related to the global or regional atmospheric air mass movement, are the dominant control on the overall rainfall \citep{Maynard2003,DeAngelis2010,Holper2011,Smith2012}. Long term trends are shown in Australian rainfall, although whether they are due to global warming is still under discussion \citep{Wardle2004,Nicholls2006,Holper2011}.
%% comment: update references?
The difference in available radiation and hence ET in different seasons increase the variability in rainfall. The land surface effect on rainfall can be confused by changes in these factors. Hence a regression model was used to estimate the amount of variability in rainfall that is due to these important factors, and isolate changes resulting from vegetation cover change.
As highlighted in the introduction, the Australian climate is influenced by sea surface temperature in the tropical Pacific and Indian Oceans, as well as pressure systems in the Southern Ocean \citep{BoM2012}.  \citet{Risbey2009} compared five large-scale drivers, including ENSO (measured by SOI and the Tropical Pacific SSTs), IOD, SAM, MJO\footnote{MJO is a large scale eastward-propagating wave-like disturbance in equatorial latitudes \citep{Risbey2009}.} (Madden-Julian oscillation) and blocking, in relation to Australian rainfall variability. They identified SOI as the most important index among all indices tested for broad parts of Australia (including QLD and NSW/VIC) in almost any season. In this study, up to two important indices from the seven climatic indicators (see section \ref{sec:Data}) were used as the explanatory variables in the model for each study region.

<<rainData,eval=F , echo = F, results = hide>>=

## 100 years awap data & 30 years
qld_100 <- read.csv("Data\\rainfall\\QLD_month_rf.csv",header=T)
qld_30 <- subset(qld_100,as.Date(qld_100[,1],"%d/%m/%Y")>="1979-01-01" & as.Date(qld_100[,1],"%d/%m/%Y")<="2008-12-01")
qld100_mean <- ts(rowMeans(qld_100[,2:2801],na.rm=T,dims=1),start=c(1900,1),frequency=12)
qld30_mean <- window(qld100_mean,start=c(1979,1),end=c(2008,12))

nsw_100 <- read.csv("Data\\rainfall\\NSW_month_rf.csv",header=T)
nsw_30 <- subset(nsw_100,as.Date(nsw_100[,1],"%d/%m/%Y")>="1979-01-01" & as.Date(nsw_100[,1],"%d/%m/%Y")<="2008-12-01")
nsw100_mean <- ts(rowMeans(nsw_100[,2:1185],na.rm=T,dims=1),start=c(1900,1),frequency=12)
nsw30_mean <- window(nsw100_mean, start=c(1979,1),end=c(2008,12))

require(MASS)
fitdistr(qld30_mean,"gamma")  ## assume shape=1
fitdistr(nsw30_mean,"gamma")  ## assume shape =8
ks.test(qld30_mean,1,0.03)
ks.test(nsw30_mean,2.4,0.03)

setEPS()
postscript("hist_qld.eps")
 hist(qld30_mean,main="",xlab=expression(paste("rainfall (mm mont",h^-1,")",sep="")))
dev.off()

setEPS()
postscript("hist_nswvic.eps")
 hist(nsw30_mean,main="",xlab=expression(paste("rainfall (mm mont",h^-1,")",sep="")))
dev.off()

## autocorrelation in rainfall
setEPS()
postscript("autocorrelation.eps")
par(mfrow=c(2,2))
  acf(qld30_mean,lag.max=12,main="QLD")
  acf(nsw30_mean,lag.max=12,main="NSW & VIC")

  pacf(qld30_mean,lag.max=12,main="")
  pacf(nsw30_mean,lag.max=12,main="")  
dev.off()

## deseasonalise & detrend rainfall: assume the deseasonalised monthly zonal rainfall is normal distributed and can be used in pearson cross-correlation function
## use stl function in R
  qld30_deseason <- stl(qld30_mean,s.window=length(qld30_mean),t.window=length(qld30_mean))$time.series[,3]
  nsw30_deseason <- stl(nsw30_mean,s.window=length(nsw30_mean),t.window=length(nsw30_mean))$time.series[,3]
  qld100_deseason <- stl(qld100_mean,s.window=length(qld100_mean),t.window=length(qld100_mean))$time.series[,3]
  nsw100_deseason <- stl(nsw100_mean,s.window=length(nsw100_mean),t.window=length(nsw100_mean))$time.series[,3]
# 
# setEPS()
# postscript("qld_mean_ts.eps")
# par(cex.lab=1.3,cex.axis=1.3)
# plot(qld30_deseason,xlab="Year",ylab="monthly rainfall")
# abline(v=2003,col="red",lwd=2)
# text(2004.6,150,"2003",cex=1.5,col="red")
# dev.off()
# 
# setEPS()
# postscript("nswvic_mean_ts.eps")
# par(cex.lab=1.3,cex.axis=1.3)
# plot(nsw30_deseason,xlab="Year",ylab="monthly rainfall")
# abline(v=2003,col="red",lwd=2)
# text(2004.6,120,"2003",cex=1.5,col="red")
# dev.off()
#   
# setEPS()
# postscript("cor_qld.eps")
# par(mfrow=c(2,3),mar=c(5,4,1,1))
# ccf(as.vector(SOI_30),as.vector(qld30_deseason),lag.max=12,xlim=c(-12,0),xlab="SOI  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
# ccf(as.vector(SST34_30),as.vector(qld30_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO3.4  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
# ccf(as.vector(SST4_30),as.vector(qld30_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO4  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
# ccf(as.vector(PDO_100),as.vector(qld100_deseason),lag.max=24,xlim=c(-24,0),xlab="PDO Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
# ccf(as.vector(PDO_100*SOI_100),as.vector(qld100_deseason),lag.max=12,xlim=c(-12,0),xlab="PDO*SOI  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
# ccf(as.vector(IOD_30),as.vector(qld30_deseason),lag.max=12,xlim=c(-12,0),xlab="IOD  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
# dev.off()
# 
# # ccf(subset(SOI_30,cycle(SOI_30)==10|cycle(SOI_30)==11|cycle(SOI_30)==12),subset(qld_deseason,cycle(qld_deseason)==10|cycle(qld_deseason)==11|cycle(qld_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="SOI  Lag",ylab="CCF") ## SOI
# # ccf(subset(SST34_30,cycle(SST34_30)==10|cycle(SST34_30)==11|cycle(SST34_30)==12),subset(qld_deseason,cycle(qld_deseason)==10|cycle(qld_deseason)==11|cycle(qld_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="SST34  Lag",ylab="CCF") ## SST34
# # ccf(subset(SST3_30,cycle(SST3_30)==10|cycle(SST3_30)==11|cycle(SST3_30)==12),subset(qld_deseason,cycle(qld_deseason)==10|cycle(qld_deseason)==11|cycle(qld_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="SST3  Lag",ylab="CCF") ## SST3
# # ccf(subset(PDO_30,cycle(PDO_30)==10|cycle(PDO_30)==11|cycle(PDO_30)==12),subset(qld_deseason,cycle(qld_deseason)==10|cycle(qld_deseason)==11|cycle(qld_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="PDO  Lag",ylab="CCF") ## PDO
# # ccf(subset(PDO_30,cycle(PDO_30)==10|cycle(PDO_30)==11|cycle(PDO_30)==12)* subset(SOI_30,cycle(SOI_30)==10|cycle(SOI_30)==11|cycle(SOI_30)==12),subset(qld_deseason,cycle(qld_deseason)==10|cycle(qld_deseason)==11|cycle(qld_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="PDO  Lag",ylab="CCF") ## PDO
# # ccf(subset(IOD_30,cycle(IOD_30)==10|cycle(IOD_30)==11|cycle(IOD_30)==12),subset(qld_deseason,cycle(qld_deseason)==10|cycle(qld_deseason)==11|cycle(qld_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="IOD  Lag",ylab="CCF") ## IOD
# # 
# # 
# # 
# # ## 1,2,3: SOI
# # ## 4,5,6: SOI
# # ## 7,8,9: SOI, IOD
# # ## 10,11,12: SOI, SST34, SST3, PDO
# # 
# 
# setEPS()
# postscript("cor_nswvic.eps")
# par(mfrow=c(2,3),mar=c(5,4,1,1))
# ccf(as.vector(SOI_30),as.vector(nsw30_deseason),lag.max=12,xlim=c(-12,0),xlab="SOI  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
# ccf(as.vector(SST34_30),as.vector(nsw30_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO3.4  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
# ccf(as.vector(SST4_30),as.vector(nsw30_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO4  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
# ccf(as.vector(PDO_100),as.vector(nsw100_deseason),lag.max=24,xlim=c(-24,0),xlab="PDO Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
# ccf(as.vector(PDO_100*SOI_100),as.vector(nsw100_deseason),lag.max=12,xlim=c(-12,0),xlab="PDO*SOI  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
# ccf(as.vector(IOD_30),as.vector(nsw30_deseason),lag.max=12,xlim=c(-12,0),xlab="IOD  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
# dev.off()
# 
# ccf(subset(SOI_30,cycle(SOI_30)==7|cycle(SOI_30)==8|cycle(SOI_30)==9),subset(nswvic_deseason,cycle(nswvic_deseason)==7|cycle(nswvic_deseason)==8|cycle(nswvic_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="SOI  Lag",ylab="CCF") ## SOI
# ccf(subset(SST34_30,cycle(SST34_30)==7|cycle(SST34_30)==8|cycle(SST34_30)==9),subset(nswvic_deseason,cycle(nswvic_deseason)==7|cycle(nswvic_deseason)==8|cycle(nswvic_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="SST34  Lag",ylab="CCF") ## SST34
# ccf(subset(SST3_30,cycle(SST3_30)==7|cycle(SST3_30)==8|cycle(SST3_30)==9),subset(nswvic_deseason,cycle(nswvic_deseason)==7|cycle(nswvic_deseason)==8|cycle(nswvic_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="SST3  Lag",ylab="CCF") ## SST3
# ccf(subset(PDO_30,cycle(PDO_30)==7|cycle(PDO_30)==8|cycle(PDO_30)==9),subset(nswvic_deseason,cycle(nswvic_deseason)==7|cycle(nswvic_deseason)==8|cycle(nswvic_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="PDO  Lag",ylab="CCF") ## PDO
# ccf(subset(PDO_30,cycle(PDO_30)==7|cycle(PDO_30)==8|cycle(PDO_30)==9)* subset(SOI_30,cycle(SOI_30)==7|cycle(SOI_30)==8|cycle(SOI_30)==9),subset(nswvic_deseason,cycle(nswvic_deseason)==7|cycle(nswvic_deseason)==8|cycle(nswvic_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="PDO  Lag",ylab="CCF") ## PDO
# ccf(subset(IOD_30,cycle(IOD_30)==7|cycle(IOD_30)==8|cycle(IOD_30)==9),subset(nswvic_deseason,cycle(nswvic_deseason)==7|cycle(nswvic_deseason)==8|cycle(nswvic_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="IOD  Lag",ylab="CCF") ## IOD
# 
# 
# 
# ## 1,2,3: NONE
# #4,5,6: SOI
# # 7, 8, 9: SOI, IOD
# ## 10, 11, 12: SOI, SST34, SST3, IOD

@

As a preliminary step, the correlations between rainfall and each climatic index were analysed. Rainfall in each study region was first deseasonalised and detrended using the seasonal decomposition function ``stl" in R \citep{Rstats2011}. The reasons for using detrended data was discussed in \citet{Smith2012}. %% explain what they said
The cross-correlations between the deseasonalised and detrended rainfall and the climatic indices were tested using the Pearson's product moment correlation method, assuming the relationships are linear. Although the optimal technique for exploring the correlation with each index could be different as described in \citet{Risbey2009}, the Pearson's method was applied to all indices for consistency. Because the PDO describes the multi-decadal SST with lower frequency \citep{MacDonald2005,Zanchettin2008,Kamruzzaman2011}, instead of 30-year rainfall data, a longer period (108 years, from 1900 to 2008) was used to estimate the correlation with PDO, up to lag 24. For the other indices, the 30-year data was used. %Here lags up to 12 were applied except for PDO, for which the maximum lag was 24.

%figure 5
\begin{figure}[h]
\centering
\includegraphics[bb=1 1 500 350, scale=0.6]{cor_qld.jpg}
 \caption{Cross-correlation of the six indices and rainfall in QLD study region. Where the correlations with PDO are sought, 108-year rainfall data (1900 - 2008) are used. Otherwise, 30-year rainfall data are used. The correlation with NINO 3 is not shown as it is very similar to but weaker than the case of NINO 3.4. The blue dashed lines indicate the 95\% confidence interval.}
  \label{fig:cor_rain_qld}
\end{figure}

%figure 6
\begin{figure}[htp]
\centering
\includegraphics[bb=1 1 500 550, scale=0.6]{cor_nswvic.jpg}
 \caption{Cross-correlation of the six indices and rainfall in NSW/VIC study region.}
  \label{fig:cor_rain_nsw}
\end{figure}

Based on the correlation between the climatic indices and rainfall (as shown in Figure~\ref{fig:cor_rain_qld} and Figure~\ref{fig:cor_rain_nsw}), the following results were found:
\begin{itemize}
  \setlength{\itemsep}{0cm}
  \setlength{\parskip}{0cm}
  \item In QLD, the correlation between rainfall and SOI at zero time lags is the highest across all indices, outweighing the other ENSO indicators. 
  \item In NSW/VIC, again the SOI has the highest correlation with rainfall, followed by the IOD. Both occur at the zero time lags. 
\end{itemize}
The above findings are consistent with previous studies. Although some indices are serially correlated with rainfall up to several months, the lag zero events have the most significant correlation coefficients. Concurrent climatic index series were generally found most useful in rainfall prediction \citep[e.g.][]{Risbey2009,Kamruzzaman2011}. The correlations between the climatic indices and rainfall for each individual season have also been tested. The results were similar. So there was no change in the results as we were interested at in using no more than two indices in the model.

%% check chamber2002 article later

Rainfall in Australia shows strong seasonal patterns \citep{Holper2011,ABS2012}. For example, the north part of the country is summer rainfall dominant with dry winter, while most of the southern part has a winter rainfall regime. This character is given by the movement of subtropical high pressure system which dominates Australian climate \citep{BoM2012a}. The seasonal component of rainfall has a periodic pattern so it is better modelled as a smooth term. A spline function was applied on the months to define a smooth seasonal pattern \citep{Wood2011}. %The spline function ``s" in R was used .

Long term trends in the regional rainfall in some parts of Australia are significant \citep{Hughes2003,Gallant2007,Chowdhury2010}. In the northern and eastern parts of the continent, increasing rainfall is reported over the last century \citep{Hughes2003}. The presence of such long term trends may be confused with the outcome of a step change in rainfall. A linear trend term was implemented in the model to remove any long term trend effect.

We assumed all the factors are additive components in determining rainfall as in \citet{Kamruzzaman2011}. Generally, monthly rainfall has a skewed distribution so the normality assumption of the residuals in a general linear model could be violated. In this case, the rainfall model is expressed as a generalised additive model (GAM) \citep{Hastie1986} with a log link function g(), assuming the residuals are gamma distributed (see Figure~\ref{fig:hist_rain}).
\vspace{0.5cm}
%\begin{dmath}
\begin{equation}
g(E(\mathbf{R}_r)) = \beta_0 + s_1(D_{1,r}\mathbf{SOI}) + s_2(D_{2,r}\mathbf{IOD}) + s_3(\mathbf{Season}) + \beta_1\mathbf{Trend} + \boldsymbol{\epsilon}_r
  \label{eqn:model}
\end{equation}
%\end{dmath}
The bold letters represent the time series vectors. The subscript $r$ denotes the region, $r=$qld or nswvic. $\beta_u$ ($_u$=0, 1) are the fitted coefficients in the model. $s_v$ ($_v$=1, 2, 3) are the smooth functions on the climatic indices and the season. $D_{1,r}$ and $D_{2,r}$ switch on/off the corresponding climatic index in the model as discussed previously. 
\begin{equation}
  D_{1,r} = 1 
\end{equation}
as SOI was used in both regions as rainfall predictor.
\begin{equation}
  D_{2,r} = \left\{ \begin{array}{ll}
   1 &\mbox{for NSW/VIC } \\
   0 &\mbox{for QLD}
  \end{array} \right.
  \label{eqn:dummyD}
\end{equation}
The linear long term trend in the rainfall data is modelled by \textbf{Trend}=1,2,3...n, where n is the total number of months in the time series. \textbf{Season} is the seasonal component which is represented by applying a smooth spline function on the month values (1-12. The monthly rainfalls in both regions appear to have a lag 1 autocorrelation, being 0.26 for QLD and 0.33 for NSW/VIC. 
%% comment: Do we have to explain why we don't include lag 1 autocorrelation in the model, or use mixed modelling?
The SOI and IOD terms are also modelled with spline functions. As the effect of large scale drivers on Australian rainfall is more likely to be seasonal \citep{Murphy2008,Schepen2012}, the spline function can closely reproduce the high and low impacts of the climatic indices.

%figure 7
\begin{figure}[htp]
  \centering
  \includegraphics[bb=50 80 650 400, scale=0.6]{hist_rain.jpg}
  \caption{Distribution of monthly rainfall in (a) QLD and (b) NSW/VIC. By using a Kolmogorov-Smirnov test with shape = 1 and 2.4 respectively, rainfalls in both regions are shown to have a gamma distribution.}
  \label{fig:hist_rain}
\end{figure}

\subsection{Tree cover change as factor variable}

One of the main difficulties in empirical observation studies on the effect of land cover change on rainfall is the lack of continuous monitoring of land surface variables, or even, that no specific variable can be defined that can clearly represent the land surface process. Given the lack of a full picture of the land surface process, a factor variable was used in this study to represent the abrupt land surface change (see Equation~\ref{eqn:model2}). The change could be a result of either land clearing or bushfires as long as it is permanent or takes a long time to recover. Here we approached the problem with two different models.

In the first method, the tree cover change was used as a predictor in the regression model, represented by a factor variable \textbf{LC}. The significance of the coefficient of \textbf{LC}, denoted as $\beta'_5$ in Equation~\ref{eqn:model2}, can be determined by a ratio test. 
\begin{equation}
  \mbox{LC} = \left\{ \begin{array}{ll}
     \mbox{Trees} \\
     \mbox{Removed}
  \end{array} \right.
  \label{eqn:dummyC}
\end{equation}
Therefore in both regions, land cover is ``trees" for the period before land cover change and ``removed" for the period after the change. Here we simply assumed that vegetation cover change has occurred on every pixel. The remaining term $\epsilon_r$ is the amount of rainfall that is attributed to other unspecified factors and random errors. Hence the regression model becomes
\vspace{0.5cm}
%\begin{dmath}
\begin{equation}
g(E(\mathbf{R}_r)) = \beta'_0 + s'_1(D_{1,r}\mathbf{SOI}) + s'_2(D_{2,r}\mathbf{IOD}) + s'_3\mathbf{Season} + \beta'_1\mathbf{Trend} + \beta'_2\mathbf{LC} + \boldsymbol{\epsilon'}_r
  \label{eqn:model2}
\end{equation}
%\end{dmath}

%The valid time lengths are different in the two regions. In the QLD region, the exact timing of tree clearing is unknown. The information given in SLATS showed that clearings had occurred within 2003 - 2004. Hence in the analysis, the time series from 2003 to 2004 were discarded as the \textbf{LC} value for these two years cannot be determined. In the NSW/VIC region, severe bushfires were reported in January 2003. Storms and flooding were also reported immediately after the fires up to March in the Alpine Shire, Victoria, which could be an aftermath effect of the bushfires \citep{Tryhorn2008}. In this case, regression was performed on the time series excluding January - March 2003 to reduce the noise of storm events. The regression model was run starting from 1979 in both regions, by assuming the land covers before 2000 were ``trees".

Vegetation cover changes occurred at different times in the two regions. In the QLD region, there no exact time can be assigned to the land clearing. Clearing has occurred between 2003 - 2004 according to SLATS reports. The information on the change in type of land cover during this time period is missing. Therefore, four scenarios were tested in the analysis. In these scenarios the ``after change" period started from: (1) June 2003, (2) January 2004, (3) June 2004 and (4) January 2005. In the NSW/VIC region, severe bushfires were reported in early January 2003. Hence the ``tree" cover state was up to December 2002 then it was changed to ``removed" state from January 2003. The regression model was run from 1979 for both regions.
% Storms and flooding were also reported immediately after the fires up to March in the Alpine Shire, Victoria, which could be an aftermath effect of the bushfires \citep{Tryhorn2008}. In this case, two scenarios were tested, setting January 2003 and April 2003 as the beginning of the after-change period.

% comment: missing values and gaps in data are complicating issues and require careful consideration \citep{Kundzewicz2004}
%comment: Hence it was assumed that clearing was at the regional scale. The land surface cover in each individual pixel is homogeneous.

<<simpleMode,eval=F, echo = F, results = hide>>=

## plotting a hypothesed land cover change model
data1 <- data.frame(year=c(1,2,2,3),tc=c(2,2,1,1))
data2 <- data.frame(year=c(1,2,2,3,4),tc=c(2,2,1,1.5,2))

setEPS()
postscript("tc_simple_qld.eps")
par(mar=c(5,6,1,1))
plot(data1$year,data1$tc,type="l",ylim=c(0.5,2.5),xlab="",ylab="",xaxt="n",yaxt="n")
axis(1,at=c(1,2,3),labels=c("before","LCC","after"),cex.axis=1.3)
axis(2,at=c(1,2),label=c("removed","tree"),cex.axis=1.3,las=2)
dev.off()

setEPS()
postscript("tc_simple_nswvic.eps")
par(mar=c(5,6,1,1))
plot(data2$year,data2$tc,type="l",ylim=c(0.5,2.5),xlab="",ylab="",xaxt="n",yaxt="n")
axis(1,at=c(1,2,4),labels=c("before","LCC","after"),cex.axis=1.3)
axis(2,at=c(1,2),label=c("removed","tree"),cex.axis=1.3,las=2)
dev.off()

@

<<fullRegression,eval=F , echo = F, results = hide>>=

## predicting model: land cover is category variable
rain.trend <- function(x,y,m,var.data) {

  var.data$rain <- as.vector(unlist(y[x]))+0.0001

  if (!is.na(mean(var.data$rain))) {
    if (grepl("qld",substitute(y))) {
      var.data <- na.omit(var.data)    
      rain_fit <- gam(rain ~ s(month, k=3)
                  + s(SOI, k=3)
                  + cover  
                  + trd
                  , data=var.data,family=Gamma(link=log))
    } else
     if (grepl("nsw",substitute(y))) {  
      var.data <- na.omit(var.data)    
      rain_fit <- gam(rain ~ s(month, k=3)
                  + s(SOI, k=3)
                  + s(IOD, k=3)
                  + cover  
                  + trd
                  , data=var.data,family=Gamma(link=log))
    }

  return(rain_fit)
  } else {NA}

}

## note: link=log generally has higher r.sq value than link=reverse

## generate variables for the model
trd <- seq(1:360)
month <-rep(1:12,times=30)
mth <- cycle(ts(month,start=c(1979,1),frequency=12))

LC <- c(rep("tree",288),rep("clear",72))
# LC_qld1 <- c(rep("tree",294),rep("clear",66)) ## after-change: from Jun 2003
# LC_qld2 <- c(rep("tree",300),rep("clear",60)) ## after- : from Jan 2004
# LC_qld3 <- c(rep("tree",306),rep("clear",54)) ## after- : from Jun 2004
LC_qld4 <- c(rep("tree",312),rep("clear",48)) ## after- : from Jan 2005

var.data <- data.frame(month, trd, SOI = SOI_30a, IOD = IOD_30a, cover = LC)
# var.qld1 <- data.frame(month, trd, SOI = SOI_30a, IOD = IOD_30a, cover = LC_qld1)
# var.qld2 <- data.frame(month, trd, SOI = SOI_30a, IOD = IOD_30a, cover = LC_qld2)
# var.qld3 <- data.frame(month, trd, SOI = SOI_30a, IOD = IOD_30a, cover = LC_qld3)
var.qld4 <- data.frame(month, trd, SOI = SOI_30a, IOD = IOD_30a, cover = LC_qld4)

## run model for 30 years
#   coef.rain30.qld1 <- lapply(2:ncol(qld_30), function(x) rain.trend(x,qld_30,30,var.qld1))
#   coef.rain30.qld2 <- lapply(2:ncol(qld_30), function(x) rain.trend(x,qld_30,30,var.qld2))
#   coef.rain30.qld3 <- lapply(2:ncol(qld_30), function(x) rain.trend(x,qld_30,30,var.qld3))
  coef.rain30.qld4 <- lapply(2:ncol(qld_30), function(x) rain.trend(x,qld_30,30,var.qld4))
  coef.rain30.nswvic <- lapply(2:ncol(nsw_30), function(x) rain.trend(x,nsw_30,30,var.data))

## extract sign of C
  C30_qld <- unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$p.coeff[2],NA)))
  C30_nswvic <- unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$p.coeff[2],NA)))

## extract p value of C
#   Cp30_qld1 <- unlist(lapply(1:length(coef.rain30.qld1),function(x) ifelse(summary(coef.rain30.qld1)[x,2]=="gam",summary(coef.rain30.qld1[[x]])$p.pv[2],NA)))
# Cp30_qld2 <- unlist(lapply(1:length(coef.rain30.qld2),function(x) ifelse(summary(coef.rain30.qld2)[x,2]=="gam",summary(coef.rain30.qld2[[x]])$p.pv[2],NA)))
# Cp30_qld3 <- unlist(lapply(1:length(coef.rain30.qld3),function(x) ifelse(summary(coef.rain30.qld3)[x,2]=="gam",summary(coef.rain30.qld3[[x]])$p.pv[2],NA)))
Cp30_qld4 <- unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$p.pv[2],NA)))
Cp30_nswvic <- unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$p.pv[2],NA)))

## p value of seasons and climate indices
p_qld <- unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$s.pv[1],NA)))
p_nsw <- unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$s.pv[1],NA)))

## extract r.sq of the model
rsq_qld <- unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$r.sq,NA)))
rsq_nsw <- unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$r.sq,NA)))

## frame of the two regions (100 years data)
qld_tif <- raster("Data\\BoM\\rainfall\\AWAP\\from1900_FVG\\r19000101_qld.tif")
nsw_tif <- raster("Data\\BoM\\rainfall\\AWAP\\from1900_FVG\\r19000101_nsw.tif")

  dist_Cp30_qld4 <- raster(qld_tif)
  dist_Cp30_qld4[] <- Cp30_qld4
#   dist_Cp30_qld <- crop(reproject(dist_Cp100_qld,CRS=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84"),program="raster"),e1)
  dist_Cp30_nsw <- raster(nsw_tif)
  dist_Cp30_nsw[] <- Cp30_nswvic
#   dist_Cp30_nsw <- crop(reproject(dist_Cp30_nsw,CRS=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84"),program="raster"),e2)
# 
#   dist_C30_qld <- raster(reg_qld)
#   dist_C30_qld[] <- C30_qld
#   dist_C30_nswvic <- raster(reg_nswvic)
#   dist_C30_nswvic[] <- C30_nswvic
# 
# ## range of predicted values
  predictmax_qld30 <- unlist(lapply(1:length(coef.rain30.qld4), function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam", max(coef.rain30.qld4[[x]]$fitted.values),NA)))
  predictmin_qld30 <- unlist(lapply(1:length(coef.rain30.qld4), function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam", min(coef.rain30.qld4[[x]]$fitted.values),NA)))
# 
mean_qld <- lapply(1:length(coef.rain30.qld4), function(x) if(summary(coef.rain30.qld4)[x,2]=="gam") {fitted(coef.rain30.qld4[[x]])} else{NA})

## calculate mean residual
  cc <- 1
  qld.sum <- mean_qld[[1]]
  for (i in 2:length(mean_qld)) {
    if(!is.na(mean_qld[[i]])) {
      qld.sum <- qld.sum + mean_qld[[i]]
      cc <- cc+1
    }
  }
  qld.mean <- qld.sum/cc
  qld.mean <- ts(qld.mean,start=c(1979,1),frequency=12)

predictmax_nsw30 <- unlist(lapply(1:length(coef.rain30.nswvic), function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam", max(coef.rain30.nswvic[[x]]$fitted.values),NA)))
# 70 - 254.7
  predictmin_nsw30 <- unlist(lapply(1:length(coef.rain30.nswvic), function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam", min(coef.rain30.nswvic[[x]]$fitted.values),NA)))
# 18.4 - 62.2

mean_nswvic <- lapply(1:length(coef.rain30.nswvic), function(x) if(summary(coef.rain30.nswvic)[x,2]=="gam") {fitted(coef.rain30.nswvic[[x]])} else{NA})

## calculate mean residual
  cc <- 1
  nswvic.sum <- mean_nswvic[[1]]
  for (i in 2:length(mean_nswvic)) {
    if(!is.na(mean_nswvic[[i]])) {
      nswvic.sum <- nswvic.sum + mean_nswvic[[i]]
      cc <- cc+1
    }
  }
  nswvic.mean <- nswvic.sum/cc
  nswvic.mean <- ts(nswvic.mean,start=c(1979,1),frequency=12)

## plots
  setEPS()
  postscript("Cp_30yrs_qld.eps")
  par(mar=c(3,3,1,6),xpd=T)
  plot(e111,xlab="",ylab="")
  image(dist_Cp30_qld4,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
#   plot(aus_reg2,border="grey",add=T)
  plot(mdb_reg2,border="grey",lwd=3,add=T)
  plot(e111,add=T)
image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.7,legend.width=0.9,legend.args=list(text="p value",line=0.5,cex=1.2,font=1),axis.args=list(cex.axis=0.8))
legend(148.7,-27.9,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=1)
  dev.off()

  setEPS()
  postscript("Cp_30yrs_nswvic.eps")
  par(mar=c(3,3,1,6),xpd=T)  
  plot(e222,xlab="",ylab="") #,border="white")
  image(dist_Cp30_nsw,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
  plot(aus_reg2,border="grey",add=T)
  plot(mdb_reg2,border="grey",lwd=3,add=T)
  plot(e222,add=T)
  #image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.8,legend.width=0.7,legend.args=list(text="p value",line=0.5,cex=1,font=2),axis.args=list(cex.axis=0.8))
  dev.off()
# 
# ## R^2 of the regression
#     rsq_qld30 <- unlist(lapply(1:length(coef.rain30.qld), function(x) ifelse(summary(coef.rain30.qld)[x,2]=="gam", summary(coef.rain30.qld[[x]])$r.sq,NA)))
#     rsq_nswvic30 <- unlist(lapply(1:length(coef.rain30.nswvic), function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam", summary(coef.rain30.nswvic[[x]])$r.sq,NA)))
# 
# ## significant of the other two variables
#     v1_qld30 <- unlist(lapply(1:length(coef.rain30.qld), function(x) ifelse(summary(coef.rain30.qld)[x,2]=="gam", summary(coef.rain30.qld[[x]])$s.pv[1],NA)))
#     v1_nswvic30 <- unlist(lapply(1:length(coef.rain30.nswvic), function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam", summary(coef.rain30.nswvic[[x]])$s.pv[1],NA)))
# 
## plot gam.check
  setEPS()
  postscript("qld_check.eps")
  gam.check(coef.rain30.qld4[[1]])
  dev.off()

  setEPS()
  postscript("nswvic_check.eps")
  gam.check(coef.rain30.nswvic[[which(values(dist_Cp30_nsw)<0.05)[6]]])
  dev.off()

@

\subsection{Step trend test}

A step trend test was used to detect changes in rainfall after the vegetation cover change. This nonparametric statistical test was modified from the Mann-Whitney Rank-Sum test by \citet{Hirsch1985}. The test was developed to look for a step change in data which is cross correlated. The gridded rainfall dataset was built by an interpolation method and as a result there is high spatial correlation between neighbouring pixels. Hence the step trend test is suitable for the analysis of such a dataset. The advantages of this test are: (1) it does not depend on assumptions of the data distribution; (2) it is not restricted to datasets with no missing data; (3) it is robust and not as easily influenced by outliers and negative numbers \citep{Hirsch1985}.

\subsubsection{Rainfall residuals} %% do we need this heading?

The rainfall residuals from the regression model in Equation~\ref{eqn:model} were used for this test. According to \citet{Hirsch1985}, using deseasonalised and detrended data is important in a test to detect step change. Furthermore, since rainfall is only partially attributed to local sources and conditions, noise can be introduced by large scale dynamics and changes in other climatic factors. The assumption is that the regression model should remove this noise. Hence the model residuals $\epsilon'_r$ excludes the effects due to climate drivers, seasonality and the long term trend. As a result local effects are assumed to explain at least part of the variation in the residuals. The  test, described in the following sub section, associates changes in rainfall trends with the tree cover changes. 
%The step change of tree cover were expected to be identified from the rainfall residuals. 
%However deseasonalisation is unnecessary as the test was set up to count for the difference between months. Here the season term was not used in the model to reduce the risk of overfitting \citep{Wood2011}.

%chunk 5: note code has changed
<<chunk5,eval=F , echo = F, results = hide>>=

## without land cover variable for step rend test
rain.trend3 <- function(x,y,m,var.data) {

  var.data$rain <- as.vector(unlist(y[x]))+0.0001

  if (!is.na(mean(var.data$rain))) {
    if (grepl("qld",substitute(y))) {
      rain_fit <- gam(rain ~ trd  # 
                  + s(month,k=3)
                  + s(SOI, k=3)
                  , data=var.data,family=Gamma(link=log))
    } else
    if (grepl("nsw",substitute(y))) {
      rain_fit <- gam(rain ~ trd
                  + s(month,k=3)
                  + s(SOI, k=3)
                  + s(IOD, k=3) 
                  , data=var.data,family=Gamma(link=log))
    }

  return(rain_fit)
  } else {NA}

}
# 
# ## change LC variable to be null
#   var.data <- data.frame(month, trd, SOI = SOI_30a, IOD = IOD_30a)
# 
## run model for 30 years
  coef.rain.qld <- lapply(2:ncol(qld_30), function(x) rain.trend3(x,qld_30,30,var.data))
  coef.rain.nswvic <- lapply(2:ncol(nsw_30), function(x) rain.trend3(x,nsw_30,30,var.data))

## explaining power of the partial model
## QLD
  rsq_qld <- unlist(lapply(1:length(coef.rain.qld), function(x) ifelse(summary(coef.rain.qld)[x,2]=="gam", summary(coef.rain.qld[[x]])$r.sq,NA)))
# 
#   predictmax_qld <- unlist(lapply(1:length(coef.rain.qld), function(x) ifelse(summary(coef.rain.qld)[x,2]=="gam", max(coef.rain.qld[[x]]$fitted.values),NA)))
#   predictmin_qld <- unlist(lapply(1:length(coef.rain.qld), function(x) ifelse(summary(coef.rain.qld)[x,2]=="gam", min(coef.rain.qld[[x]]$fitted.values),NA)))
# 
## extract residual
  residual_qld <- lapply(1:length(coef.rain.qld), function(x) if(summary(coef.rain.qld)[x,2]=="gam") {coef.rain.qld[[x]]$y-fitted(coef.rain.qld[[x]])} else {NA})

# ## calculate mean residual
#   cc <- 1
#   qld.rsum <- residual_qld[[1]]
#   for (i in 2:length(residual_qld)) {
#     if(!is.na(residual_qld[[i]])) {
#       qld.rsum <- qld.rsum + residual_qld[[i]]
#       cc <- cc+1
#     }
#   }
#   qld.rm <- qld.rsum/cc
#   qld.rm <- ts(qld.rm,start=c(1979,1),frequency=12)
# 
# ## NSW/VIC
# 
  rsq_nswvic <- unlist(lapply(1:length(coef.rain.nswvic), function(x) ifelse(summary(coef.rain.nswvic)[x,2]=="gam", summary(coef.rain.nswvic[[x]])$r.sq,NA)))
# 
#   predictmax_nswvic <- unlist(lapply(1:length(coef.rain.nswvic), function(x) ifelse(summary(coef.rain.nswvic)[x,2]=="gam", max(coef.rain.nswvic[[x]]$fitted.values),NA)))
#   predictmin_nswvic <- unlist(lapply(1:length(coef.rain.nswvic), function(x) ifelse(summary(coef.rain.nswvic)[x,2]=="gam", min(coef.rain.nswvic[[x]]$fitted.values),NA)))

# extract residual
residual_nswvic <- lapply(1:length(coef.rain.nswvic), function(x) if(summary(coef.rain.nswvic)[x,2]=="gam") {coef.rain.nswvic[[x]]$y-fitted(coef.rain.nswvic[[x]])} else {NA})

# ## calculate mean residual
# cc <- 1
# nswvic.rsum <- residual_nswvic[[1]]
# for (i in 2:length(residual_nswvic)) {
#   if(!is.na(residual_nswvic[[i]])) {
#     nswvic.rsum <- nswvic.rsum + residual_nswvic[[i]]
#     cc <- cc+1
#   }
# }
# 
# nswvic.rm <- nswvic.rsum/cc
# nswvic.rm <- ts(nswvic.rm,start=c(1979,1),frequency=12)
# 
# setEPS()
# postscript("res_qld.eps")
# hist(qld.rm,main="",xlab="rainfall residuals")
# dev.off()
# 
# setEPS()
# postscript("res_nswvic.eps")
# hist(nswvic.rm,main="",xlab="rainfall residuals")
# dev.off()
# 
# setEPS()
# postscript("cor_res_nswvic.eps")
# par(mfrow=c(1,2),mar=c(5,4,5,1))
# ccf(SOI_30,nswvic.rm,lag.max=12,xlim=c(-1,0),xlab="SOI  Lag",main="",ylab="CCF")
# ccf(IOD_30,nswvic.rm,lag.max=12,xlim=c(-1,0),xlab="IOD  Lag",main="",ylab="CCF")
# dev.off()

## explaining power of the partial model
rain.trend4 <- function(x,y,m,var.data) {

  var.data$rain <- as.vector(unlist(y[x]))+0.0001

  if (!is.na(mean(var.data$rain))) {
    if (grepl("qld",substitute(y))) {
      rain_fit <- gam(rain ~ s(SOI, k=3)
                  , data=var.data,family=Gamma(link=log))
    } else
    if (grepl("nsw",substitute(y))) {
      rain_fit <- gam(rain ~ s(SOI, k=3)
                  + s(IOD, k=3) 
                  , data=var.data,family=Gamma(link=log))
    }

  return(rain_fit)
  } else {NA}

}

## explaining power of index
  coef.soi.qld <- lapply(2:ncol(qld_30), function(x) rain.trend4(x,qld_30,30,var.data))
  coef.soi.nswvic <- lapply(2:ncol(nsw_30), function(x) rain.trend4(x,nsw_30,30,var.data))
  rsq_soi_qld <- unlist(lapply(1:length(coef.soi.qld), function(x) ifelse(summary(coef.soi.qld)[x,2]=="gam", summary(coef.soi.qld[[x]])$r.sq,NA)))
  rsq_soi_nswvic <- unlist(lapply(1:length(coef.soi.nswvic), function(x) ifelse(summary(coef.soi.nswvic)[x,2]=="gam", summary(coef.soi.nswvic[[x]])$r.sq,NA)))

setEPS()
postscript("soi_qld.eps")
 hist(rsq_soi_qld*100,main="",xlab=expression(paste("% of variability explained (",R^2,")")),cex.lab=1.2)
 dev.off()

 setEPS()
postscript("soi_nswvic.eps")
 hist(rsq_soi_nswvic*100,main="",xlab=expression(paste("% of variability explained (",R^2,")")),cex.lab=1.2)
 dev.off()

@

\subsubsection{Mann-Whitney rank-sum statistic}

The step trend test is a modified version of the Mann-Whitney rank-sum statistic \citep{Hirsch1985}. As a nonparametric rank-based test, the Mann-Whitney test does not use the exact values of rainfall but depends on the ranks of the data. For each month, rainfall residuals of each year were ranked in an ascending order. The ranking of January rainfall in a sample pixel k in QLD is illustrated below:\\

%\begin{table}[h]
%  \centering
  \begin{tabular}{lccccccc}
  & 1998 & 1999 & 2000 & 2001 & 2002 & 2005 & 2006\\
  \textit Rainfall residuals: & -0.3 & -60.9 & -16.1 & -71.7 & 111.1 & -7.2 & -60.5 \\
  \textit Rank ($R'_{1k}$): & 6 & 2 & 4 & 1 & 7 & 5 & 3\\
  \end{tabular}
%\end{table}
\noindent Therefore, the smallest or most negative value has rank 1 and the largest value has the maximum rank.

The before period and the after period formed two groups of samples. The split point of the two periods was based on the timing of the vegetation cover changes. In the QLD region, changes occurred anytime during 2003 and 2004. In contrast to the previous method, the time period of where the land cover chang occurred was excluded, as the nonparametric test allows missing data. \citet{Hirsch1985} also pointed out that the power of the test is higher if the data of the change period is ignored. Hence 2003 and 2004 were excluded from the analysis. As a result, the after-change period was 2005 - 2007.

In the case of NSW/VIC, the bushfires broke out in early January 2003. The change was within a relatively short period of the year. Therefore the after-change period in this region still started in January 2003. Following \citet{Hirsch1985}, the before period was set to five years (1998 - 2002) in both regions. %The test does not require a long time series before and after the intervention as shown in \citet{Hirsch1985}.

The rank of rainfall in month j year i in pixel k is denoted as $R'_{ijk}$. The sum of ranks of rainfall in month j in pixel k before the known intervention is:  
\begin{equation}
  W_{jk} = \sum_{i=1}^{n_1}R'_{ijk}.
  \label{eqn:Wj}
\end{equation}
$n_1$ is the number of years before the land cover change. The expected value of $W_{jk}$ is 
\begin{equation}
  \mu_w=n_1(n_1+n_2+1)/2
\end{equation}
$n_2$ is the number of years after the change. Hence the expected value of the rank sum before the intervention is the same for all months and all pixels. The sum of ranks for the whole time period is fixed, as $(n_1+n_2)(n_1+n_2+1)/2$. In this study, since there are only two groups (before and after), knowing the rank-sum of one group is the same as knowing the rank-sum of the other group. If the rainfall data is temporally and spatially independent, the variance of $W_{jk}$ is 
\begin{equation}
  \sigma^2_w = n_1\cdot n_2(n_1+n_2+1)/m
\end{equation}
where m is the number of months which is 12 in the case of a full year.

\subsubsection{Step trend test}

Instead of completing the Mann-Whitney U-test, \citet{Hirsch1985} applied the rank-sum statistics in a standard normal Z test. The modified test can be used to detect step change and it accounts for serial and cross correlation in the data. In the case here, the deseasonalised and detrended data shows little autocorrelation in the time series but possesses strong cross correlation between neighbouring pixels, i.e. $R>0.99$. Hence the covariance between pixels would need to be considered in the test.

% Assuming the distribution of $W_{jk}$ is normal and independent,

The sum of $W_{jk}$ for a block of $ns$ pixels over the whole year, $\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk}$, has mean 
\begin{equation}
  E(\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk})=12\cdot ns\cdot\mu_W
\end{equation}
and variance
\begin{equation}
  Var(\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk})=\sum_{j=1}^{12}\sum_{k=1}^{ns}\sum_{h=1}^{ns}C(W_{jk},W_{jh}).
\end{equation}
$C(W_{jk},W_{jh})$ is the covariance of the W statistics between pixel k and pixel h in month j. When $k=h$, $C(W_{jk},W_{jh})=\sigma^2_w$. When $k\neq h$, 
\begin{equation}
  C(W_{jk},W_{jh})=\sigma^2_w r(R_k,R_h)
\end{equation}
where $r(R_k,R_h)$ is the product moment correlation coefficient of the concurrent ranks in pixel k and h. Here $r$ is calculated on the full time series in each pixel. In this analysis, the test was applied to a square block of four pixels each time. As argued by \citet{Hirsch1985}, $ns=4$ is the most optimal solution to balance the cost and the gain in the test power.
% \begin{equation}
%   \sigma^2_{w12} = 12\cdot\sigma^2_W = n_1\cdot n_2(n_1+n_2+1).
% \end{equation}

The statistic of the step trend test is then defined as 
\begin{equation}
  Z'=\frac{\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk}-12\cdot ns\cdot\mu_w}{\sqrt{Var(\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk})}}.
  \label{eqn:Z}
\end{equation}  
The above statistic is written for a 12 month period. By changing the value 12, it can also be used to test seasonal rainfall change or for other customized periods.

%The Z' test is based on the assumption that the sum of $W_j$ is normally distributed. Small numbers of $n_1$ and $n_2$ might weaken this assumption. On the other hand, because there is a much longer period before the intervention in the rainfall data, the possible changes at some other time could increase the noise in the analysis. Given the rainfall time series data was only available up to 2008 and the specified land intervention was closer to the end of this time series, a subset of the rainfall time series was selected from 1998 to 2008. This 11-year rainfall data analysis was compared to the time series analysis for the full period.

The null hypothesis ($H_0$) in this study is that there was no change in rainfall due to land surface intervention. The results of the step trend test can be interpreted according to the sign of Z' score (see Table~\ref{tab:Zscore}). Z' is normally distributed similar to the standard normal statistics Z. Hence it can be compared to a standard normal distribution to determine the p value.

\begin{table}
  \caption[The interpretation of Z' score in the step trend test.]{The interpretation of Z' score in the step trend test. Following \citet[][Chapter 23, P887]{Hipel1994}.}
  \label{tab:Zscore}
  \centering
  \begin{tabular}{|l|l|} \hline
    Z' $>$ 0 & rainfall decreases after change\\ \hline
    Z' $<$ 0 & rainfall increases after change \\ \hline
    Z' = 0 & rainfall does not change \\ \hline
  \end{tabular}
\end{table}

%chunk 6
<<chunk6,eval=F , echo = F, results = hide>>=

## perform step test on the data
## function to calculate the step trend statistic (Hirsch, 1985)
stepTrend2 <- function(x,y,i){

    n1 <- 2002-1997

    if (i==1){

      z <- 56  # no. of column

    } else {

    z <- 37 }

if (x%%z!=0){ #& x%%z!=(z-1)) {
    y.1 <- y[[x]]
    y.2 <- y[[x+1]]
#     y.3 <- y[[x+2]]
    y.3 <- y[[x+z]]
    y.4 <- y[[x+z+1]]
#     y.6 <- y[[x+z+2]]
#     y.7 <- y[[x+2*z]]
#     y.8 <- y[[x+2*z+1]]
#     y.9 <- y[[x+2*z+2]]

    if(!is.na(y.1[1])&!is.na(y.2[1])&!is.na(y.3[1])&!is.na(y.4[1])) { #|!is.na(y.5[1])|!is.na(y.6[1])|!is.na(y.7[1])|!is.na(y.8[1])|!is.na(y.9[1])) {

    if (i == 1) {
      n2 <- 2008-2004    
      v.1 <- ts(y.1[c(229:288,313:360)],start=c(1998,1),frequency=12)
      v.2 <- ts(y.2[c(229:288,313:360)],start=c(1998,1),frequency=12)
      v.3 <- ts(y.3[c(229:288,313:360)],start=c(1998,1),frequency=12)
      v.4 <- ts(y.4[c(229:288,313:360)],start=c(1998,1),frequency=12)
#       v.5 <- ts(y.5[c(229:288,313:348)],start=c(1998,1),frequency=12)
#       v.6 <- ts(y.6[c(229:288,313:348)],start=c(1998,1),frequency=12)
#       v.7 <- ts(y.7[c(229:288,313:348)],start=c(1998,1),frequency=12)
#       v.8 <- ts(y.8[c(229:288,313:348)],start=c(1998,1),frequency=12)
#       v.9 <- ts(y.9[c(229:288,313:348)],start=c(1998,1),frequency=12)
#       v <- (x$y-x$fitted.values)[c(1:288,313:357)]
#       v.s <- ts(v,start=c(1998,1),frequency=12)

    } else {

      n2 <- 2008-2002

      v.1 <- ts(y.1[c(229:360)],start=c(1998,1),frequency=12)
      v.2 <- ts(y.2[c(229:360)],start=c(1998,1),frequency=12)
      v.3 <- ts(y.3[c(229:360)],start=c(1998,1),frequency=12)
      v.4 <- ts(y.4[c(229:360)],start=c(1998,1),frequency=12)
#       v.5 <- ts(y.5[c(229:348)],start=c(1998,1),frequency=12)
#       v.6 <- ts(y.6[c(229:348)],start=c(1998,1),frequency=12)
#       v.7 <- ts(y.7[c(229:348)],start=c(1998,1),frequency=12)
#       v.8 <- ts(y.8[c(229:348)],start=c(1998,1),frequency=12)
#       v.9 <- ts(y.9[c(229:348)],start=c(1998,1),frequency=12)      
#       v <- (x$y-x$fitted.values)[c(1:288, 301:357)]
#       v.s <- ts(v,start=c(1998,1),frequency=12)    
}

## rank by month
   for (j in 1:12) {    
    assign(paste("r1",j,sep=""),rank(subset(v.1,cycle(v.1)==j)))
    assign(paste("r2",j,sep=""),rank(subset(v.2,cycle(v.2)==j)))
    assign(paste("r3",j,sep=""),rank(subset(v.3,cycle(v.3)==j)))
    assign(paste("r4",j,sep=""),rank(subset(v.4,cycle(v.4)==j))) 
#     assign(paste("r5",j,sep=""),rank(subset(v.5,cycle(v.5)==j)))
#     assign(paste("r6",j,sep=""),rank(subset(v.6,cycle(v.6)==j)))
#     assign(paste("r7",j,sep=""),rank(subset(v.7,cycle(v.7)==j)))
#     assign(paste("r8",j,sep=""),rank(subset(v.8,cycle(v.8)==j)))
#     assign(paste("r9",j,sep=""),rank(subset(v.9,cycle(v.9)==j)))

    }

## mann-whitney rank sum statistic
    rlist.1 <- list(r11,r12,r13,r14,r15,r16,r17,r18,r19,r110,r111,r112) ## full year
    rlist.2 <- list(r21,r22,r23,r24,r25,r26,r27,r28,r29,r210,r211,r212)
    rlist.3 <- list(r31,r32,r33,r34,r35,r36,r37,r38,r39,r310,r311,r312)
    rlist.4 <- list(r41,r42,r43,r44,r45,r46,r47,r48,r49,r410,r411,r412)
#     rlist.5 <- list(r51,r52,r53,r54,r55,r56,r57,r58,r59,r510,r511,r512)
#     rlist.6 <- list(r61,r62,r63,r64,r65,r66,r67,r68,r69,r610,r611,r612)    
#     rlist.7 <- list(r71,r72,r73,r74,r75,r76,r77,r78,r79,r710,r711,r712)    
#     rlist.8 <- list(r81,r82,r83,r84,r85,r86,r87,r88,r89,r810,r811,r812)
#     rlist.9 <- list(r91,r92,r93,r94,r95,r96,r97,r98,r99,r910,r911,r912)    
    rlist <- list(unlist(rlist.1),unlist(rlist.2),unlist(rlist.3),unlist(rlist.4)) #,unlist(rlist.5),unlist(rlist.6),unlist(rlist.7),unlist(rlist.8),unlist(rlist.9))
## seasonal
#     rlist <- list(unlist(rlist.1[c(9:11)]),unlist(rlist.2[c(9:11)]),unlist(rlist.3[c(9:11)]),unlist(rlist.4[c(9:11)]))
#     rlist <- list(r1,r2,r12) ## summer
#     rlist <- list(r3,r4,r5)  ## autumn
#     rlist <- list(r6,r7,r8)  ## winter
#     rlist <- list(r9,r10,r11) ## spring
    w.s1 <- 0
    w.s2 <- 0
    w.s3 <- 0
    w.s4 <- 0
#     w.s5 <- 0
#     w.s6 <- 0
#     w.s7 <- 0
#     w.s8 <- 0
#     w.s9 <- 0
  for (k in 1:length(rlist.1)) {
#   for (k in c(9:11)) {
    w.s1 <- sum(rlist.1[[k]][1:n1])+w.s1
    w.s2 <- sum(rlist.2[[k]][1:n1])+w.s2
    w.s3 <- sum(rlist.3[[k]][1:n1])+w.s3
    w.s4 <- sum(rlist.4[[k]][1:n1])+w.s4
#     w.s5 <- sum(rlist.5[[k]][1:n1])+w.s5
#     w.s6 <- sum(rlist.6[[k]][1:n1])+w.s6
#     w.s7 <- sum(rlist.7[[k]][1:n1])+w.s7
#     w.s8 <- sum(rlist.8[[k]][1:n1])+w.s8
#     w.s9 <- sum(rlist.9[[k]][1:n1])+w.s9
    }
    w.s <- sum(w.s1,w.s2,w.s3,w.s4) #,w.s5,w.s6,w.s7,w.s8,w.s9)

## other statistic
# mean for full year
    mu <- n1*(n1+n2+1)/2
    mu.s <- 12*4*mu
# ## mean for one season
#     mu <- n1*(n1+n2+1)/2
#     mu.s <- 3*4*mu

## variance (standard variation). The autocorrelation of the residuals can be assumed to be null (checked, cor.test(), true for NSW, mostly true for WA. QLD? there are some situations in WA require further investigation)
# assume only cross correlation
    ss <- n1*n2*(n1+n2+1)/12
    ss.s <- 0
    for (kk in 1:4) {
      for (kkk in 1:4) {
#         for (kkkk in 1:12) {
          ss.s <- 12*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
#         }
      }
    }
## variance for one season   
#     ss <- n1*n2*(n1+n2+1)/3
#     ss.s <- 0
#     for (kk in 1:4) {
#       for (kkk in 1:4) {
#           ss.s <- 3*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
#       }
#     }

## step trend statistic
    z <- (w.s - mu.s)/sqrt(ss.s)

    return(z) }

     else {return(NA)} }

     else {return(NA)}

  }

## 30 years
## full year, 
## central QLD
  qld.z.new <- unlist(lapply(1:(length(residual_qld)-56),function(x) stepTrend2(x,residual_qld,1)))
  qld.z.new <- c(qld.z.new,rep(NA,56))

  tr_qld.step.new <- raster(qld_tif)
  tr_qld.step.new[] <- qld.z.new
## NSW  
  nswvic.z.new <- unlist(lapply(1:(length(residual_nswvic)-37),function(x) stepTrend2(x,residual_nswvic,2)))
  nswvic.z.new <- c(nswvic.z.new,rep(NA,37))

  tr_nswvic.step.new <- raster(nsw_tif)
  tr_nswvic.step.new[] <- nswvic.z.new

# ## seasonal
#   qld.z.sum <- unlist(lapply(1:(length(residual_qld)-56),function(x) stepTrend2(x,residual_qld,1)))
#   qld.z.sum <- c(qld.z.sum,rep(NA,56))
#   
#   tr_qld.step.sum <- raster(qld_tif)
#   tr_qld.step.sum[] <- qld.z.sum
#   
#   nswvic.z.sum <- unlist(lapply(1:(length(residual_nswvic)-37),function(x) stepTrend2(x,residual_nswvic,2)))
#   nswvic.z.sum <- c(nswvic.z.sum,rep(NA,37))
#   
#   tr_nswvic.step.sum <- raster(nsw_tif)
#   tr_nswvic.step.sum[] <- nswvic.z.sum

  plotclt <- rev(brewer.pal(8,"RdYlBu"))

setEPS()
postscript("step_new_qld.eps")
  par(mar=c(3,3,1,7),xpd=T)
  plot(e111,xlab="",ylab="")#,border="white")
  image(tr_qld.step.new,col=plotclt,breaks=c(-3,-1.645,-1.282,-0.842,0,0.842,1.282,1.645,3),legend=F,add=T)
  plot(mdb_reg2,border="grey",add=T,lwd=3) 
  plot(e111,add=T)
image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,0,1,2,3,4),lab.breaks=c(0.00,0.05,0.1,0.2,1,0.2,0.1,0.05,0.00),legend.only=T,horizontal=F,col=plotclt,legend.shrink=0.7,legend.width=0.9,legend.args=list(text=c(expression(paste("level of significance (",alpha,")")),"positive Z'","negative Z'"),side=c(2,3,1),line=0.5,cex=1,font=2),axis.args=list(cex.axis=0.8))
  dev.off()

setEPS()
  postscript("step_new_nswvic.eps")
  par(mar=c(3,3,1,7),xpd=T)
  plot(e222,xlab="",ylab="")#,border="white")
  image(tr_nswvic.step.new,col=plotclt,breaks=c(-2.3,-1.645,-1.282,-0.842,0,0.842,1.282,1.645,3),legend=F,add=T)
  plot(mdb_reg2,border="grey",add=T,lwd=3)
  plot(aus_reg2,border="grey",add=T) 
  plot(e222,add=T)
image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,0,1,2,3,4),lab.breaks=c(0.00,0.05,0.1,0.2,1,0.2,0.1,0.05,0.00),legend.only=T,horizontal=F,col=plotclt,legend.shrink=0.7,legend.width=0.9,legend.args=list(text=c(expression(paste("level of significance (",alpha,")")),"positive Z'","negative Z'"),side=c(2,3,1),line=0.5,cex=1,font=2),axis.args=list(cex.axis=0.8))
legend(149,-36.7,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=0.9)
  dev.off()

@

% chunk 7: bootstrapping (do not keep spatial structure, not in use)
<<chunk7,eval=F , echo = F, results = hide>>=

## bootstrapping
# ## the % of pixels with significance level < 0.05 in the original data set

qld_per_origp <- length(subset(qld.z.new,qld.z.new>=1.645))/length(qld.z.new)
# 0.0125
qld_per_orign <- length(subset(qld.z.new,qld.z.new<=-1.645))/length(qld.z.new)
# 0
# 
nsw_per_origp <- length(subset(nswvic.z.new,nswvic.z.new>=1.645))/length(nswvic.z.new)
# 0.015
nsw_per_orign <- length(subset(nswvic.z.new,nswvic.z.new<=-1.645))/length(nswvic.z.new)
# 0

# new step trend test: do not keep spatial structure
# NstepTrend <- function(x){  
#   
#   n1 <- 2002-1997
#     
#   n2 <- 2008-2002
# 
#   if(!is.na(x[1])) {
#     
#   v <- x[c(229:357)]
#     
#   y<-ts(v,start=c(1998,1),frequency=12)
#   
#   for(i in 1:12) { 
#   y.mth <- sample(window(y,start=c(1998,i),frequency=1),replace=F)
#   
#   ## rank by month   
#     assign(paste("r",i,sep=""),rank(y.mth))
#   } 
#   
#   
# ## mann-whitney rank sum statistic
#     rlist <- list(r1,r2,r3,r4,r5,r6,r7,r8,r9,r10,r11,r12) ## full year
# 
#     w.s <- 0
#     for (k in 1:length(rlist)) {
#     w.s <- sum(rlist[[k]][1:n1])+w.s
#     }
#   
# ## other statistic
# ## mean for full year
#     mu <- n1*(n1+n2+1)/2
#     mu.s <- length(rlist)*mu
# 
#   
# ## variance (standard variation). The autocorrelation of the residuals can be assumed to be null (checked, true for QLD and NSW, mostly true for WA. there are some situations in WA require further investigation)
#     ss <- n1*n2*(n1+n2+1)/length(rlist)
#     ss.s <- length(rlist)*ss
# 
# ## step trend statistic
#     z <- (w.s - mu.s)/sqrt(ss.s)
#   
#     return(z) 
#     }
#     
#      else {return(NA)}
#   
#   }
# 
# ## conduct bootstrapping
# R <- 1000
# zp.values = numeric(R)
# zn.values = numeric(R)
# 
# ptm <- proc.time()
# for (i in 1:R) {
#   temp <- unlist(lapply(residual_nswvic,function(x) NstepTrend(x)))
#   if(i==1) {write.table(temp,file="R:\\AVSproject\\WrittenWork\\PhD_thesis\\chap3\\bt_zvalues.txt",row.names=F)
#             } else
#               {write.table(temp,file="R:\\AVSproject\\WrittenWork\\PhD_thesis\\chap3\\bt_zvalues.txt",append=T,row.names=F,col.names=F)}
#    zp.values[i] <- length(subset(temp,temp>=1.28))/length(temp)
#   zn.values[i] <- length(subset(temp,temp<=-1.28))/length(temp)
#   
# }
# proc.time()-ptm
# 
# write.table(zp.values,file="R:\\AVSproject\\WrittenWork\\PhD_thesis\\chap3\\bt_outp.txt",row.names=F)
# write.table(zn.values,file="R:\\AVSproject\\WrittenWork\\PhD_thesis\\chap3\\bt_outn.txt",row.names=F)
@

% chunk 8: bootstrapping (spatial structure is kept, in used)
<<chunk8,eval=F , echo = F, results = hide>>=

## one cell step trend
## perform step test on the data for the bootstrapping test (spatial structure was kept)
## function to calculate the step trend statistic (Hirsch, 1985)
stepTrend <- function(x,i){

    n1 <- 2002-1997

  if(!is.na(x[1])) {

    if(i==1) {
      n2 <- 2008-2004    
      v <- x[c(229:288,313:360)]
#       v <- (x$y-x$fitted.values)[c(1:288,313:357)]
      v.s <- ts(v,start=c(1998,1),frequency=12)    
    } else
      if(i==2) {
      n2 <- 2008-2002
      v <- x[c(229:360)]
#       v <- (x$y-x$fitted.values)[c(1:288, 301:357)]
      v.s <- ts(v,start=c(1998,1),frequency=12)      
    } #else 
#       {n2 <- 2008-2002
#       v <- (x$y-x$fitted.values)[229:357]
# #       v <- (x$y-x$fitted.values)
#       v.s <- ts(v,start=c(1998,1),frequency=12)}

## rank by month
   for (j in 1:12) {    
    assign(paste("r",j,sep=""),rank(subset(v.s,cycle(v.s)==j)))
    }

## mann-whitney rank sum statistic
    rlist <- list(r1,r2,r3,r4,r5,r6,r7,r8,r9,r10,r11,r12) ## full year
#     rlist <- list(r1,r2,r12) ## summer
#     rlist <- list(r3,r4,r5)  ## autumn
#     rlist <- list(r6,r7,r8)  ## winter
#     rlist <- list(r9,r10,r11) ## spring
    w.s <- 0
    for (k in 1:length(rlist)) {
    w.s <- sum(rlist[[k]][1:n1])+w.s
    }

## other statistic
## mean for full year
    mu <- n1*(n1+n2+1)/2
    mu.s <- length(rlist)*mu
## mean for one season
#     mu <- n1*(n1+n2+1)/2
#     mu.s <- 3*mu

## variance (standard variation). The autocorrelation of the residuals can be assumed to be null (checked, cor.test(), true for NSW, mostly true for WA. QLD? there are some situations in WA require further investigation)
## assume only cross correlation
    ss <- n1*n2*(n1+n2+1)/length(rlist)
    ss.s <- length(rlist)*ss
## variance for one season   
#     ss <- n1*n2*(n1+n2+1)/3
#     ss.s <- 3*ss

## step trend statistic
    z <- (w.s - mu.s)/sqrt(ss.s)

    return(z) }

     else {return(NA)}

  }

## full year
## central QLD
  qld.z <- unlist(lapply(residual_qld,function(x) stepTrend(x,1)))
## NSW/VIC
## treatment 1: excluding 2003
  nswvic.z <- unlist(lapply(residual_nswvic,function(x) stepTrend(x,2)))

  qld_origp <- length(subset(qld.z,qld.z>=1.645))/length(qld.z)
# 0.01
  qld_orign <- length(subset(qld.z,qld.z<=-1.645))/length(qld.z)
# 0

  nsw_origp <- length(subset(nswvic.z,nswvic.z>=1.645))/length(nswvic.z)
# 0.022
  nsw_orign <- length(subset(nswvic.z,nswvic.z<=-1.645))/length(nswvic.z)

# new step trend test 2: keep spatial structure
NstepTrend2 <- function(x,y,ii){

  n1 <- 2002-1997

  if(ii==1) {
      n2 <- 2008-2004    
      v <- x[c(229:288,313:360)]    
    } else
      if(ii==2) {
      n2 <- 2008-2002
      v <- x[c(229:360)]
    }

  if(!is.na(unlist(x)[1])) {

  yy<-ts(v,start=c(1998,1),frequency=12)

  for(i in 1:12) { 
  y.mth <- data.frame(mth=as.vector(window(yy,start=c(1998,i),frequency=1)),o=y[[i]])[order(y[[i]]),]

  ## rank by month   
    assign(paste("r",i,sep=""),rank(y.mth$mth) )
  }

## mann-whitney rank sum statistic
    rlist <- list(r1,r2,r3,r4,r5,r6,r7,r8,r9,r10,r11,r12) ## full year

    w.s <- 0
    for (k in 1:length(rlist)) {
    w.s <- sum(rlist[[k]][1:n1])+w.s
    }

## other statistic
## mean for full year
    mu <- n1*(n1+n2+1)/2
    mu.s <- length(rlist)*mu

## variance (standard variation). The autocorrelation of the residuals can be assumed to be null (checked, true for QLD and NSW)
    ss <- n1*n2*(n1+n2+1)/length(rlist)
    ss.s <- length(rlist)*ss

## step trend statistic
    z <- (w.s - mu.s)/sqrt(ss.s)

    return(z) 
    }

     else {return(NA)}

  }

## conduct bootstrapping 2
R <- 1000
z2pq.values = numeric(R)
z2nq.values = numeric(R)
## sample years without replacement. note that with this method, only the order of years has changed, but months within year do not change
new.order <- list()
# mth.l <- rep(11,12)
mth.l <- rep(9,12)

ptm <- proc.time()
for (i in 1:R) {

  for (j in 1:12) {
    new.order[[j]]<-sample(1:mth.l[j])
  }

  temp <- unlist(lapply(residual_qld,function(x,y) NstepTrend2(x,y,1),new.order))
  if(i==1) {write.table(temp,file="bt_zvalues2_qld.txt",row.names=F)
            } else
              {write.table(temp,file="bt_zvalues2_qld.txt",append=T,row.names=F,col.names=F)}
  z2pq.values[i] <- length(subset(temp,temp>=1.645))/length(temp)
  z2nq.values[i] <- length(subset(temp,temp<=-1.645))/length(temp)
}
proc.time()-ptm

write.table(z2pq.values,file="bt_out2p_qld.txt",row.names=F)
write.table(z2nq.values,file="bt_out2n_qld.txt",row.names=F)

# zp.values <- unlist(read.table("B:/My Study\\AVSproject\\WrittenWork\\PhD_thesis\\chap3\\bt_outp.txt",header=T,sep=""))*100
# zn.values <- unlist(read.table("B:/My Study\\AVSproject\\WrittenWork\\PhD_thesis\\chap3\\bt_outn.txt",header=T,sep=""))*100
# 
bt1_nswp.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_out2p_nsw.txt",header=T,sep=""))*100
bt1_nswn.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_out2n_nsw.txt",header=T,sep=""))*100

bt1_qldp.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_out2p_qld.txt",header=T,sep=""))*100
bt1_qldn.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_out2n_qld.txt",header=T,sep=""))*100
# # 
# # setEPS()
# # postscript("boot1.eps")
# # hist(zp.values,xlim=c(0,9),xlab="% positive z",main="no spatial dependency")
# # points(per_origp*100,0,pch=16)
# # dev.off()
# # 
# # setEPS()
# # postscript("boot2.eps")
# # hist(zn.values,xlim=c(0,25),xlab="% negative z",main="no spatial dependency")
# # points(per_orign*100,0,pch=16)
# # dev.off()
# # # 
# setEPS()
# postscript("bt1_qldp.eps")
# hist(bt1_qldp.values,nclass=100,xlab="% positive Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
# points(qld_origp*100,0,pch=16,col="red",cex=1.8)
# dev.off()
# 
# setEPS()
# postscript("bt1_qldn.eps")
# hist(bt1_qldn.values,nclass=100,xlab="% negative Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
# points(qld_orign*100,0,pch=16,col="red",cex=1.8)
# dev.off()
# 
# setEPS()
# postscript("bt1_nswp.eps")
# hist(bt1_nswp.values,nclass=100,xlab="% positive Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
# points(nsw_origp*100,0,pch=16,col="red",cex=1.8)
# dev.off()
# 
# setEPS()
# postscript("bt1_nswn.eps")
# hist(bt1_nswn.values,nclass=100,xlab="% negative Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
# points(nsw_orign*100,0,pch=16,col="red",cex=1.8)
# dev.off()
# 
# ## multiple cells
# NstepTrend <- function(x,y,m,i){  
# 
#     n1 <- 2002-1997
#     
#     if (i==1){
#       
#       z <- 56  # no. of column
#       
#     } else {
#     
#     z <- 37 }
#     
# if (x%%z!=0){ #& x%%z!=(z-1)) {
#     y.1 <- m[[x]]
#     y.2 <- m[[x+1]]
# #     y.3 <- y[[x+2]]
#     y.3 <- m[[x+z]]
#     y.4 <- m[[x+z+1]]
# #     y.6 <- y[[x+z+2]]
# #     y.7 <- y[[x+2*z]]
# #     y.8 <- y[[x+2*z+1]]
# #     y.9 <- y[[x+2*z+2]]
#     
#     if(!is.na(y.1[1])&!is.na(y.2[1])&!is.na(y.3[1])&!is.na(y.4[1])) { #|!is.na(y.5[1])|!is.na(y.6[1])|!is.na(y.7[1])|!is.na(y.8[1])|!is.na(y.9[1])) {
#         
#     if (i == 1) {
#       n2 <- 2008-2004    
#       v.1 <- ts(y.1[c(229:288,313:360)],start=c(1998,1),frequency=12)
#       v.2 <- ts(y.2[c(229:288,313:360)],start=c(1998,1),frequency=12)
#       v.3 <- ts(y.3[c(229:288,313:360)],start=c(1998,1),frequency=12)
#       v.4 <- ts(y.4[c(229:288,313:360)],start=c(1998,1),frequency=12)
# #       v.5 <- ts(y.5[c(229:288,313:348)],start=c(1998,1),frequency=12)
# #       v.6 <- ts(y.6[c(229:288,313:348)],start=c(1998,1),frequency=12)
# #       v.7 <- ts(y.7[c(229:288,313:348)],start=c(1998,1),frequency=12)
# #       v.8 <- ts(y.8[c(229:288,313:348)],start=c(1998,1),frequency=12)
# #       v.9 <- ts(y.9[c(229:288,313:348)],start=c(1998,1),frequency=12)
# #       v <- (x$y-x$fitted.values)[c(1:288,313:357)]
# #       v.s <- ts(v,start=c(1998,1),frequency=12)
#       
#     } else {
#     
#       n2 <- 2008-2002
#   
#       v.1 <- ts(y.1[c(229:360)],start=c(1998,1),frequency=12)
#       v.2 <- ts(y.2[c(229:360)],start=c(1998,1),frequency=12)
#       v.3 <- ts(y.3[c(229:360)],start=c(1998,1),frequency=12)
#       v.4 <- ts(y.4[c(229:360)],start=c(1998,1),frequency=12)
# #       v.5 <- ts(y.5[c(229:348)],start=c(1998,1),frequency=12)
# #       v.6 <- ts(y.6[c(229:348)],start=c(1998,1),frequency=12)
# #       v.7 <- ts(y.7[c(229:348)],start=c(1998,1),frequency=12)
# #       v.8 <- ts(y.8[c(229:348)],start=c(1998,1),frequency=12)
# #       v.9 <- ts(y.9[c(229:348)],start=c(1998,1),frequency=12)      
# #       v <- (x$y-x$fitted.values)[c(1:288, 301:357)]
# #       v.s <- ts(v,start=c(1998,1),frequency=12)    
# }
#   
# ## rank by month
#    for (j in 1:12) {  
#     y1.mth <- data.frame(mth=as.vector(window(v.1,start=c(1998,i),frequency=1)),o=y[[j]])[order(y[[j]]),]
#     y2.mth <- data.frame(mth=as.vector(window(v.2,start=c(1998,i),frequency=1)),o=y[[j]])[order(y[[j]]),]
#     y3.mth <- data.frame(mth=as.vector(window(v.3,start=c(1998,i),frequency=1)),o=y[[j]])[order(y[[j]]),]
#     y4.mth <- data.frame(mth=as.vector(window(v.4,start=c(1998,i),frequency=1)),o=y[[j]])[order(y[[j]]),]
#     
#     assign(paste("r1",j,sep=""),rank(y1.mth$mth))
#     assign(paste("r2",j,sep=""),rank(y2.mth$mth))
#     assign(paste("r3",j,sep=""),rank(y3.mth$mth))
#     assign(paste("r4",j,sep=""),rank(y4.mth$mth)) 
# #     assign(paste("r5",j,sep=""),rank(subset(v.5,cycle(v.5)==j)))
# #     assign(paste("r6",j,sep=""),rank(subset(v.6,cycle(v.6)==j)))
# #     assign(paste("r7",j,sep=""),rank(subset(v.7,cycle(v.7)==j)))
# #     assign(paste("r8",j,sep=""),rank(subset(v.8,cycle(v.8)==j)))
# #     assign(paste("r9",j,sep=""),rank(subset(v.9,cycle(v.9)==j)))
#     
#     }
#   
# ## mann-whitney rank sum statistic
#     rlist.1 <- list(r11,r12,r13,r14,r15,r16,r17,r18,r19,r110,r111,r112) ## full year
#     rlist.2 <- list(r21,r22,r23,r24,r25,r26,r27,r28,r29,r210,r211,r212)
#     rlist.3 <- list(r31,r32,r33,r34,r35,r36,r37,r38,r39,r310,r311,r312)
#     rlist.4 <- list(r41,r42,r43,r44,r45,r46,r47,r48,r49,r410,r411,r412)
# #     rlist.5 <- list(r51,r52,r53,r54,r55,r56,r57,r58,r59,r510,r511,r512)
# #     rlist.6 <- list(r61,r62,r63,r64,r65,r66,r67,r68,r69,r610,r611,r612)    
# #     rlist.7 <- list(r71,r72,r73,r74,r75,r76,r77,r78,r79,r710,r711,r712)    
# #     rlist.8 <- list(r81,r82,r83,r84,r85,r86,r87,r88,r89,r810,r811,r812)
# #     rlist.9 <- list(r91,r92,r93,r94,r95,r96,r97,r98,r99,r910,r911,r912)    
#     rlist <- list(unlist(rlist.1),unlist(rlist.2),unlist(rlist.3),unlist(rlist.4)) #,unlist(rlist.5),unlist(rlist.6),unlist(rlist.7),unlist(rlist.8),unlist(rlist.9))
# ## seasonal
# #     rlist <- list(unlist(rlist.1[c(9:11)]),unlist(rlist.2[c(9:11)]),unlist(rlist.3[c(9:11)]),unlist(rlist.4[c(9:11)]))
# #     rlist <- list(r1,r2,r12) ## summer
# #     rlist <- list(r3,r4,r5)  ## autumn
# #     rlist <- list(r6,r7,r8)  ## winter
# #     rlist <- list(r9,r10,r11) ## spring
#     w.s1 <- 0
#     w.s2 <- 0
#     w.s3 <- 0
#     w.s4 <- 0
# #     w.s5 <- 0
# #     w.s6 <- 0
# #     w.s7 <- 0
# #     w.s8 <- 0
# #     w.s9 <- 0
#   for (k in 1:length(rlist.1)) {
# #   for (k in c(9:11)) {
#     w.s1 <- sum(rlist.1[[k]][1:n1])+w.s1
#     w.s2 <- sum(rlist.2[[k]][1:n1])+w.s2
#     w.s3 <- sum(rlist.3[[k]][1:n1])+w.s3
#     w.s4 <- sum(rlist.4[[k]][1:n1])+w.s4
# #     w.s5 <- sum(rlist.5[[k]][1:n1])+w.s5
# #     w.s6 <- sum(rlist.6[[k]][1:n1])+w.s6
# #     w.s7 <- sum(rlist.7[[k]][1:n1])+w.s7
# #     w.s8 <- sum(rlist.8[[k]][1:n1])+w.s8
# #     w.s9 <- sum(rlist.9[[k]][1:n1])+w.s9
#     }
#     w.s <- sum(w.s1,w.s2,w.s3,w.s4) #,w.s5,w.s6,w.s7,w.s8,w.s9)
#   
# ## other statistic
# # mean for full year
#     mu <- n1*(n1+n2+1)/2
#     mu.s <- 12*4*mu
# # ## mean for one season
# #     mu <- n1*(n1+n2+1)/2
# #     mu.s <- 3*4*mu
#   
# ## variance (standard variation). The autocorrelation of the residuals can be assumed to be null (checked, cor.test(), true for NSW, mostly true for WA. QLD? there are some situations in WA require further investigation)
# # assume only cross correlation
#     ss <- n1*n2*(n1+n2+1)/12
#     ss.s <- 0
#     for (kk in 1:4) {
#       for (kkk in 1:4) {
# #         for (kkkk in 1:12) {
#           ss.s <- 12*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
# #         }
#       }
#     }
# ## variance for one season   
# #     ss <- n1*n2*(n1+n2+1)/3
# #     ss.s <- 0
# #     for (kk in 1:4) {
# #       for (kkk in 1:4) {
# #           ss.s <- 3*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
# #       }
# #     }
#   
# ## step trend statistic
#     z <- (w.s - mu.s)/sqrt(ss.s)
#   
#     return(z) }
#     
#      else {return(NA)} }
#      
#      else {return(NA)}
#   
#   }
# 
# 
# zpq.values = numeric(R)
# znq.values = numeric(R)
# mth.lnsw <- rep(11,12)
# mth.lqld <- rep(9,12)
# 
# ptm <- proc.time()
# for (i in 1:R) {
#   
#   for (j in 1:12) {
#     new.order[[j]]<-sample(1:mth.lqld[j])
#   }
#   
#   temp <- unlist(lapply(1:(length(residual_qld)-56),function(x,y) NstepTrend(x,y,residual_qld,1),new.order))
#   if(i==1) {write.table(temp,file="bt_zvalues_qld.txt",row.names=F)
#             } else
#               {write.table(temp,file="bt_zvalues_qld.txt",append=T,row.names=F,col.names=F)}
#   zpq.values[i] <- length(subset(temp,temp>=1.645))/length(temp)
#   znq.values[i] <- length(subset(temp,temp<=-1.645))/length(temp)
# }
# proc.time()-ptm
# 
# write.table(zpq.values,file="bt_outp_qld.txt",row.names=F)
# write.table(znq.values,file="bt_outn_qld.txt",row.names=F)
# 
bt4_nswp.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_outp_nsw.txt",header=T,sep=""))*100
bt4_nswn.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_outn_nsw.txt",header=T,sep=""))*100

bt4_qldp.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_outp_qld.txt",header=T,sep=""))*100
bt4_qldn.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_outn_qld.txt",header=T,sep=""))*100

setEPS()
postscript("bt4_nswp.eps")
hist(bt4_nswp.values,nclass=100,xlab="% positive Z'",main="4-pixel step trend",cex.lab=1.4,cex.main=1.3)
points(nsw_origp*100,0,pch=16,col="red",cex=1.8)
dev.off()

setEPS()
postscript("bt4_nswn.eps")
hist(bt1_nswn.values,nclass=100,xlab="% negative Z'",main="4-pixel step trend",cex.lab=1.4,cex.main=1.3)
points(nsw_orign*100,0,pch=16,col="red",cex=1.8)
dev.off()

setEPS()
postscript("bt4_qldp.eps")
hist(bt4_qldp.values,nclass=100,xlab="% positive Z'",main="4-pixel step trend",cex.lab=1.4,cex.main=1.3)
points(qld_origp*100,0,pch=16,col="red",cex=1.8)
dev.off()

setEPS()
postscript("bt4_qldn.eps")
hist(bt4_qldn.values,nclass=100,xlab="% negative Z'",main="4-pixel step trend",cex.lab=1.4,cex.main=1.3)
points(qld_orign*100,0,pch=16,col="red",cex=1.8)
dev.off()

@

% chunk 9: bootstrapping (spatial structure is kept, remove short term trend, randomly select after change period)
<<chunk9,eval=F , echo = F, results = hide>>=

## x: rainfall series; y: new order; ii: selection of region
NstepTrend3 <- function(x,y,ii){

  if(!is.na(unlist(x)[1])) {

  v <- x[c(229:360)][order(y)]
  ttrd <- c(1:132)
  v <- residuals(lm(v~ttrd))

  n1 <- sample(c(4:8),1)

  if(ii==1) {    
      yy <- ts(v[c(1:60,85:132)],start=c(1998,1),frequency=12)    
      n2 <- 9-n1
    } else
      if(ii==2) {
      yy <- ts(v,start=c(1998,1),frequency=12)
      n2 <- 11-n1
    }

  for(i in 1:12) { 
  y.mth <- as.vector(window(yy,start=c(1998,i),frequency=1))

  ## rank by month   
    assign(paste("r",i,sep=""),rank(y.mth) )
  }

## mann-whitney rank sum statistic
    rlist <- list(r1,r2,r3,r4,r5,r6,r7,r8,r9,r10,r11,r12) ## full year

    w.s <- 0
    for (k in 1:length(rlist)) {
    w.s <- sum(rlist[[k]][1:n1])+w.s
    }

## other statistic
## mean for full year
    mu <- n1*(n1+n2+1)/2
    mu.s <- length(rlist)*mu

## variance (standard variation). The autocorrelation of the residuals can be assumed to be null (checked, true for QLD and NSW)
    ss <- n1*n2*(n1+n2+1)/length(rlist)
    ss.s <- length(rlist)*ss

## step trend statistic
    z <- (w.s - mu.s)/sqrt(ss.s)

    return(z) 
    }

     else {return(NA)}

  }

R <- 1000
z3pq.values = numeric(R)
z3nq.values = numeric(R)
## random sample with replacement. Both the order of the months and the order of years change

ptm <- proc.time()
for (i in 1:R) {

  new.order <- sample(c(1:132),replace=T)
  temp <- unlist(lapply(residual_qld,function(x,y) NstepTrend3(x,y,1),new.order))
#   if(i==1) {write.table(temp,file="bt_zvalues3_qld.txt",row.names=F)
#             } else
#               {write.table(temp,file="bt_zvalues3_qld.txt",append=T,row.names=F,col.names=F)}
  z3pq.values[i] <- length(subset(temp,temp>=1.645))/length(temp)
  z3nq.values[i] <- length(subset(temp,temp<=-1.645))/length(temp)
}
proc.time()-ptm

write.table(z3pq.values,file="bt_out3p_qld.txt",row.names=F)
write.table(z3nq.values,file="bt_out3n_qld.txt",row.names=F)

#NSW
z3pn.values = numeric(R)
z3nn.values = numeric(R)
## random sample with replacement. Both the order of the months and the order of years change

ptm <- proc.time()
for (i in 1:R) {

  new.order <- sample(c(1:132),replace=T)
  temp <- unlist(lapply(residual_nswvic,function(x,y) NstepTrend3(x,y,2),new.order))
#   if(i==1) {write.table(temp,file="bt_zvalues3_qld.txt",row.names=F)
#             } else
#               {write.table(temp,file="bt_zvalues3_qld.txt",append=T,row.names=F,col.names=F)}
  z3pn.values[i] <- length(subset(temp,temp>=1.645))/length(temp)
  z3nn.values[i] <- length(subset(temp,temp<=-1.645))/length(temp)
}
proc.time()-ptm

write.table(z3pn.values,file="bt_out3p_nsw.txt",row.names=F)
write.table(z3nn.values,file="bt_out3n_nsw.txt",row.names=F)
# 
# setEPS()
# postscript("bt3_qldp.eps")
# hist(z3pq.values,nclass=100,xlab="% positive Z'",main="QLD",cex.lab=1.4,cex.main=1.3)
# points(qld_origp,0,pch=16,col="red",cex=1.8)
# dev.off()
# 
# setEPS()
# postscript("bt3_qldn.eps")
# hist(z3nq.values,nclass=100,xlab="% negative Z'",main="QLD",cex.lab=1.4,cex.main=1.3)
# points(qld_orign,0,pch=16,col="red",cex=1.8)
# dev.off()
# 
# setEPS()
# postscript("bt3_nswp.eps")
# hist(z3pn.values,nclass=100,xlab="% positive Z'",main="NSW",cex.lab=1.4,cex.main=1.3)
# points(nsw_origp,0,pch=16,col="red",cex=1.8)
# dev.off()
# 
# setEPS()
# postscript("bt3_nswn.eps")
# hist(z3nn.values,nclass=100,xlab="% negative Z'",main="NSW",cex.lab=1.4,cex.main=1.3)
# points(nsw_orign,0,pch=16,col="red",cex=1.8)
# dev.off()

@
%% apply the LocSig function in SpatialVx for block bootstrapping
<<chunk91,eval=F , echo = F, results = hide>>=

require(SpatialVx)

## reconstruct the data into format required in LocSig
qld_res_mat <- matrix(unlist(residual_qld),nrow=360)
nsw_res_mat <- matrix(unlist(residual_nswvic),nrow=360)

## the step trend function for LocSig
NstepTrend4 <- function(x,ii){

  n1 <- sample(c(4:8),1)
  zz <- vector()

  for(m in 1:ncol(x)) {
    if(!is.na(x[1,m])) {

  v <- x[c(229:360),m]
  ttrd <- c(1:132)
  v <- residuals(lm(v~ttrd))

  if(ii==1) {    
      yy <- ts(v[c(1:60,85:132)],start=c(1998,1),frequency=12)    
      n2 <- 9-n1
    } else
      if(ii==2) {
      yy <- ts(v,start=c(1998,1),frequency=12)
      n2 <- 11-n1
    }

  for(i in 1:12) { 
  y.mth <- as.vector(window(yy,start=c(1998,i),frequency=1))

  ## rank by month   
    assign(paste("r",i,sep=""),rank(y.mth) )
  }

## mann-whitney rank sum statistic
    rlist <- list(r1,r2,r3,r4,r5,r6,r7,r8,r9,r10,r11,r12) ## full year

    w.s <- 0
    for (k in 1:length(rlist)) {
    w.s <- sum(rlist[[k]][1:n1])+w.s
    }

## other statistic
## mean for full year
    mu <- n1*(n1+n2+1)/2
    mu.s <- length(rlist)*mu

## variance (standard variation). The autocorrelation of the residuals can be assumed to be null (checked, true for QLD and NSW)
    ss <- n1*n2*(n1+n2+1)/length(rlist)
    ss.s <- length(rlist)*ss

## step trend statistic
    z <- (w.s - mu.s)/sqrt(ss.s)
    zz[m] <- z
    }

     else {zz[m]<-NA}
  }
  return(zz)

  }
@

\section{Result}

\subsection{Tree cover changes}

The pixels with significant tree cover change in each study region are shown in Figure~\ref{fig:tc_trend}, at the 0.05 significance level. Given the limitation of the length of this data series, it still indicates a large change in the NSW/VIC region. In the NSW/VIC region, much of the tree loss between 2002 and 2003 was concentrated in the Snowy Mountains which are at the border of NSW and VIC. Tree cover loss occurred in large parts of the QLD region between 2002 and 2005. Most of the clearings appear in the centre of this region. The tree cover change map is consistent with the annual mean EVI trend map (based on DLCD data, map not shown here), which confirms these changes are significant and persistent over the study period.

%figure 8
\begin{figure}[htp]
\centering
\includegraphics[scale=0.65]{tc.jpg}
  \caption{The maps show the areas with significant changes in tree covers in (a) the QLD region and (b) the NSW/VIC region. The amount of changes was calculated as the difference in tree covers before and after the specified land cover intervention and it is shown as the percentage of the ground area. Green colour indicates an increase in tree cover, while red colour indicates a decrease in tree cover. }
  \label{fig:tc_trend}
\end{figure}

\subsection{Regression Model \& Significance of Vegetation Cover Changes}

The regression model does not explain much of the rainfall variability. The model in Equation~\ref{eqn:model2} accounts for around 13\%\footnote{Here the adjusted $R^2$ was reported. Adjusted $R^2$ is the coefficient of determination, a measurement of the amount of variability predicted by the model adjusting for the number of explanatory terms} of the rainfall variations in the QLD region and 19\% in the NSW/VIC region on average. The residual analysis shows that the assumptions of the regression model are generally met. The standardised residual plots, however, show some funnelling for the NSW/VIC regions, suggesting non constant variance. The residual analysis, based on one sample pixel from each region, in Figure~\ref{fig:residuals_qld} and Figure~\ref{fig:residuals_nswvic} illustrates this. The residual patterns are consistent within each region.
% Can Fig9 and Fig10 be combined?
%Later when the land cover variable was excluded the results were similar as this variable can only represent a small variation in the data.
%Furthermore, the model residuals in NSW/VIC is random and normally distributed, while it is close to normal in QLD.

The model confirms the importance of the climate drivers and the seasonality in Australian rainfall. Even at the grid level, the seasons and the climatic indices were significant ($\alpha = 0.05$) everywhere in both regions. The explaining power of the model is mostly due to these variables. The climate drivers (at lag zero) accounted, on average, for 6\% of the rainfall variability in both the QLD region and the NSW/VIC region (see Figure~\ref{fig:rsq} for the distribution of $R^2$ in these two regions). These figures are within the upper bound of seasonal rainfall predictability by SST anomaly field reported by \citet{Westra2010}.

%figure 9
\begin{figure}[htp]
  \centering
  \includegraphics[bb=0 0 650 500, scale=0.8]{qld_check.jpg}
  \caption{The residual analysis of a sample pixel in the QLD region.}
  \label{fig:residuals_qld}
\end{figure}

%figure 10
\begin{figure}[htp]
  \centering
  \includegraphics[bb=0 0 650 500, scale=0.8]{nswvic_check.jpg}
  \caption{The residual analysis of a sample pixel in the NSW/VIC region.}
  \label{fig:residuals_nswvic}
\end{figure}

%figure 11
\begin{figure}[htp]
  \centering
  \includegraphics[bb=0 100 960 500,scale=0.65]{soi.jpg}
  \caption{The distribution of $R^2$ when rainfall is only modelled by the climate drivers. It shows the percentage of rainfall variability that can be explained by the climate drivers. In the case of (a) the QLD region, SOI was the only climatic index considered. In (b) the NSW/VIC region, SOI and IOD were used.}
  \label{fig:rsq}
\end{figure}

A statistically significant long term trend was not observed in either of the two regions. However, this result might not disprove the importance of a long term trend in rainfall. The overall time period is farily short and more pixels in NSW/VIC would have significant step change if the long term trend effect is not removed by the model. As trend free data is an important requirement for the step trend test, the trend term was kept in the regression model to ensure the detection of step change was not due to a possible long term trend.

%figure 12
\begin{figure}[htp]
\includegraphics[bb=10 50 960 500, scale=0.6]{Cp_30yrs.jpg}
  \caption{The spatial distribution of significance of vegetation cover changes in the two study regions. The significance of vegetation cover changes is assessed in the regression model. The figures report the p values of the coefficient of the land cover variable in the model at different significance level. The plot of the QLD region is for the scenario that the after change period started from January 2005, as the other scenarios have no p value less than 0.2.}
  \label{fig:LC_p}
\end{figure}

The land cover variable implies a step change with different values before and after the land cover intervention. This variable was only significant ($\alpha=0.05$) for the rainfall estimates in some areas in NSW/VIC, as shown in Figure~\ref{fig:LC_p}~(b). The effect of tree removal in the NSW/VIC region was highly significant in the area from the Snowy mountains ranges in the south to the west side of Australian Capital Territory (ACT), highlighted by the red colour in Figure~\ref{fig:LC_p}~(b). However no significant step change due to the land cover changes was found in rainfall of the QLD region in any of the four scenarios.

In the NSW/VIC region, the locations of rainfall change show some agreement with the locations of vegetation cover change. Comparing the results in Figure~\ref{fig:LC_p}~(b) to Figure~\ref{fig:tc_trend}~(b), the alpine area with massive tree cover loss is located inside the highlighted region. The results show that the step change in rainfall also occurred in the cotter river catchment which was heavily burned in 2003 bushfires (refer to Figure~\ref{fig:bushfire}). However, Figure~\ref{fig:LC_p}~(b) also shows that significant step change in rainfall was found in an area larger than where the bushfires has occurred. From this point of view, the results might be showing a large scale effect that reaches beyond the vegetation cover change effect. %A close look at the rainfall data revealed that it was very low in 2006, which corresponds to a severe drought due to a weak El Ni\~{n}o and unrecovered condition from the previous drought. %this area does not show up in the tree cover change map. The post-fire revegetation and restoration activities in ACT could have reduced the significance of the difference in tree cover before and after the bushfires. It is an indication that deforestation might cause a step change in rainfall but reforestation could take a long time to show effects.

%Due to the uncertainty in the timing of the land clearing, in this experiment different time windows for land clearing have also been tested. Little influence of land cover change effect was found if the time window was changed to 1990 - 2004. If the time window was between 2000 - 2004, the spatial distribution of the p value was similar to Figure~\ref{fig:site1_p}. If the time window included 2005, the results suggested that rainfall has been affected by land cover change over an even larger area on the east. This area also corresponds to the area where a high rate of clearing in 2005 - 2006 was reported \citep{NaturalResources2008}. However the confidence level of the results also decreased as this change further reduced the length of years after the land intervention in the analysis.

More generally, the model also shows that the tree cover has a positive impact on rainfall. The fitted coefficients for the reference state ``trees" were consistently positive for the pixels whose step changes of rainfall were significant. It implies that rainfall was higher when the surface was covered by trees. This was confirmed by applying a two sample Student's t-test on the annual rainfall of the before-change and after-change periods. %With a maximum coefficient value of 0.36, the magnitude of difference between the two states appears to be small.

\subsection{Step Trend Test}

%figure 13
\begin{figure}[hpt]
  \centering
 \includegraphics[bb=0 80 960 500, scale=0.7]{step_new.jpg}
  \caption{Spatial distribution of the step trend test Z' statistics in the two study sites. The test was conducted on the 30-year rainfall data from 1979 to 2009. Warm colours (yellow, orange and red) are for positive Z' values which indicate decreasing rainfall trend due to the land surface intervention. Cold colours (light blue to blue) are for negative Z' values which indicate increasing rainfall trend. The deeper the colour, the more significant the statistic.}
  \label{fig:steptest30}
\end{figure}

The spatial step trend test Z' scores are shown in Figure~\ref{fig:steptest30}. This figure provides two types of information: the sign and the significance level. The sign indicates the direction of the step change, as listed in Table~\ref{tab:Zscore}. In each region, there is a broad area of positive Z' values which implies a decrease in rainfall. In QLD, the areas of positive Z', especially when $\alpha$ is smaller than 0.1, almost align with the locations reported to have heavy land clearing (see Figure~\ref{fig:slat}). However the vegetation cover change map in Figure~\ref{fig:tc_trend}~(a) does not clearly show a similar pattern. In the Snowy Mt area and the west of ACT, where severe bushfires occurred in 2003, positive Z' values were found. It again indicates that loss of tree cover appeared to be related with a rainfall decrease. Nevertheless the observed step change in rainfall was not strong. At the 5\% significance level, changes in a number of pixels in the alpine area are possibly where the tree recovery is slow. In the QLD region, 35 pixels obtained a Z' score with p $<$ 5\%. In the NSW/VIC region, there are only 18 pixels in the alpine area having a Z' score with p $<$ 5\%. They are a very small proportion of both study regions.
% should we also report p $<$ 10\%?
%Although the rainfall changes were not significant, the spatial patterns of the positive Z' scores agree with the nature of the tree cover change in these two regions. This is again showing the possible regional effect of the land cover changes, which is consistent with the results from the regression analysis.

%figure 14
\begin{figure}[htp]
  \centering
  \includegraphics[scale=0.6]{slat.jpg}
  \caption{Woody vegetation clearing rate in QLD. The maps and figures are obtained from \citet{SLATS2004}. The red rectangle is the boundary of the QLD study region.}
  \label{fig:slat}
\end{figure}

%figure 15
\begin{figure}[htp]
  \centering
  \includegraphics[bb=0 0 650 500, scale=0.8]{compare_mean.jpg}
  \caption[Boxplots of annual rainfall (estimated based on rainfall residuals from Equation~\ref{eqn:model}) before and after the land cover intervention during 1998 - 2008 in the study regions.]{Boxplots of annual rainfall (estimated based on rainfall residuals from Equation~\ref{eqn:model}) before and after the land cover intervention during 1998 - 2008 in the study regions. On average, the after period has a lower annual rainfall amount and with outliers of small values.}
  \label{fig:mean_diff}
\end{figure}

The rainfall residuals of the two periods (before-change since 1998 and after-change) were also compared (Figure~\ref{fig:mean_diff}). The annual values of the 11-year period (1998 - 2008) were calculated for each pixel. The boxplots indicate that there were different mean values between the ``before" and ``after" periods in both regions. There are small outliers in the after period in NSW/VIC. A close look at the rainfall data revealed that rainfall was consistently very low in 2006 for the whole region. The low rainfall in 2006 was due to a weak El Ni\~{n}o and unrecovered conditions from the previous drought. While the regression model has removed most of the effect of ENSO, the  cumulative drought effect could still be visible in the residuals if the response of the rainfall to the ENSO effect is non-linear, and the memory of the past drought is persistent. This can also explain the significant results in the previous method. To remove the outlier effects, the two periods were compared excluding the 2006 rainfall. An F test confirmed that the variances were different between the before and after periods. Hence an unpaired unequal variance two sample t-test was applied to test whether the after period has lower mean rainfall than the before period, for the group of pixels showing a negative step change in rainfall ($\alpha = 0.05$) and the group of pixels with no change. The rainfall in the after period was lower than the before period (p $<$ 0.05) for pixels showing a negative step change. For those pixels without a step change, there was no statistical evidence that the two periods have different means. Hence the t-test results are consistent with the step trend test results.
%  that showed a negative step change in rainfall ($\alpha$ = 0.1) and compared to pixels without change
%The boxplots in Figure~\ref{fig:mean_diff} indicate that the rainfall distributions in the two periods are different.

The choice of $ns$ has some impact on the test results, as shown by \citet{Hirsch1985}. The cases of $ns = 1$ and $ns = 9$ were also tested. When $ns = 1$, 18 pixels in the QLD region obtained a Z' score at the 5\% significance level. In the NSW/VIC region, the detection of negative step change ($\alpha = 0.05$) reduced to 16 pixels. On the other hand, when $ns = 9$, the results were similar to the case of $ns = 4$. The changes ($\alpha = 0.05$) were detected in 35 pixels in the QLD region and 18 pixels in the NSW/VIC region. The power of the test does not change much after $ns = 4$, as shown by \citet{Hirsch1985}. The cause of the highly inconsistent results in the QLD region between $ns = 1$ and $ns = 4$ is unclear. This could possibly be because the higher $ns$ values smooth the individual pixel results.

The "field significance" of the test is considered to make inferences about the step change at regional scales from multiple local tests \citep{Wilks2006,Westra2013}. Here, the bootstrapping resampling method from \citet{Westra2013} was adapted , modified to evaluate the field significance. The spatial structure of the pixels was maintained, while the order of the years and months, as well as the length of the before/after windows were changed by random resampling. The test statistics identifies the percentage of the pixels with significant step change, positive and negative respectively, for the step trend test. The test statistics on 1000 resampled replicates were used to develop the distribution of these percentage values under the local null hypothesis that there was no step change.

The bootstrapping resampling technique was applied on both $ns=1$ and $ns=4$ cases. In the $ns=4$ case, a spatial moving block bootstrapping was used, in which the change of time sequence within the 2$\times$2 block was consistent. The distribution of the test statistics is highly concentrated at zero with a skew to the right. In all four outputs, positive step changes and negative step changes in the cases of $ns=1$ and $ns=4$, the test statistics on the observed time series fall within the null hypothesis distribution (the cases of $ns=1$ are shown in Figure~\ref{fig:boot_qld} and Figure~\ref{fig:boot_nsw}). The results showed that the chance of detecting a rainfall change was small. Again there was no strong evidence to support the hypothesis that vegetation cover change can affect local rainfall.
%% I think this shows that the results we got could just be by chance

%figure 16
\begin{figure}[ht]
\centering
\includegraphics[width=6.5in]{bt1_qld.jpg}
\caption{Percentage of pixels showing statistically significant positive Z' values (left) and negative Z' values (right) in QLD. The histogram shows the distribution of results from 1000 bootstrap resampling of the rainfall time series. The red dot represents the results from the observed data.}
\label{fig:boot_qld}
\end{figure}

%figure 17
\begin{figure}[ht]
\centering
\includegraphics[width=6.5in]{bt1_nsw.jpg}
\caption{Percentage of pixels showing statistically significant positive Z' values (left) and negative Z' values (right) in NSW/VIC. The histogram shows the distribution of results from 1000 bootstrap resampling of the rainfall time series. The red dot represents the results from the observed data.}
\label{fig:boot_nsw}
\end{figure}

% chunk 8
<<chunk81,eval=F , echo = F, results = hide>>=

# ## calculate mean rainfall difference in two period for 11-year
# 
#   sel_neg.qld <- residual_qld[which(values(tr_qld.step.new)>=1.645)]
# #   sel_pos.qld <- coef.rain.qld[which(values(tr_qld.step)<(-1.645))]
#   
#   sel_neg.nswvic <- residual_nswvic[which(values(tr_nswvic.step.new)>=1.645)]
# 
# ## compare mean of before and after
# ## boxplot
#   data.qld <- data.frame(year=seq(1998,2008,1))
#   for (i in 1:length(sel_neg.qld)) {
#     data.qld[,i+1] <- as.vector(rollapply(as.zoo(sel_neg.qld[[i]])[229:360],width=12,sum,by=12))
#   }
#   data2.qld <- melt(data.qld,id="year")
#   data2.qld$group[data2.qld$year<2003] <- "1before"
#   data2.qld$group[data2.qld$year>2004] <- "2after"
# 
# 
#   data.nswvic <- data.frame(year=seq(1998,2008,1))
#   for (i in 1:length(sel_neg.nswvic)) {
#     data.nswvic[,i+1] <- as.vector(rollapply(as.zoo(sel_neg.nswvic[[i]])[229:360],width=12,sum,by=12))
#   }
#   data2.nswvic <- melt(data.nswvic,id="year")
#   data2.nswvic$group[data2.nswvic$year<2003] <- "1before"
#   data2.nswvic$group[data2.nswvic$year>2002] <- "2after"
# 
# #   nsw_sig <- nsw_30[which(Cp30_nswvic<=0.05)]
# #   data.nswvic <- data.frame(year=seq(1979,2008,1))
# #   for (i in 1:length(nsw_sig)) {
# #     data.nswvic[,i+1] <- as.vector(rollapply(as.zoo(nsw_sig[i]),width=12,sum,by=12))
# #   }
# #   data2.nswvic <- melt(data.nswvic,id="year")
# #   data2.nswvic$group[data2.nswvic$year<2003] <- "1before"
# #   data2.nswvic$group[data2.nswvic$year>2002] <- "2after"
# 
# 
# ## perform a 2 sample t test
#   var.test(data2.qld[data2.qld$year<2003,3],data2.qld[data2.qld$year>2004 & data2.qld$year!=2006,3],alternative="two.sided")
# ## unequal variance
#   t.test(data2.qld[data2.qld$year<2003,3],data2.qld[data2.qld$year>2004 & data2.qld$year!=2006,3],alternative="greater")
# ## p = 2.2e-16
# 
#   var.test(data2.nswvic[data2.nswvic$year<2003,3],data2.nswvic[data2.nswvic$year>2002& data2.nswvic$year!=2006,3],alternative="two.sided")
# ## unequal variance
#   t.test(data2.nswvic[data2.nswvic$year<2003,3],data2.nswvic[data2.nswvic$year>2002& data2.nswvic$year!=2006,3],alternative="greater")
# ## p < 1.4e-13
# 
# 
#   setEPS()
#   postscript("compare_mean.eps")
#   par(mfrow=c(1,2))
#   boxplot(value~group,data=data2.qld,names=c("before","after"),boxwex=0.4,main="QLD site",ylab="rainfall (mm)")
# 
#   boxplot(value~group,data=data2.nswvic,names=c("before","after"),boxwex=0.4,main="NSW/VIC site")
#   dev.off()

@

\section{Summary and Discussion}

Generally, empirical studies on LCC-precipitation interaction are conducted within an area with known land surface intervention \citep[e.g.][]{Otterman1990,Durieux2003,Negri2004,Sato2007}. However, these locations are rare and difficult to isolate from real landscape change. In this study we therefore tested the effect of land cover change across a broad region, rather than only for locations where changes were known to occur or have occured. The advantage of this approach is that it does not require a long time series of land cover data which is usually unavailable. Furthermore, it does not assume a specific relationship between vegetation cover change and rainfall but allows the data to show this relationship, by applying the analysis to a broader area outside the boundary of the vegetation cover change. This approach is expected to provide a way to reduce the risk of a false positive paradox, by comparing results between areas with and without vegetation cover change. %On the other hand, this method is limited to step change analysis and cannot show a long term relationship.

Parametric tests are generally more powerful than nonparametric test in detecting a trend, when the data is normally distributed \citep{Onoz2003,Kundzewicz2004}. As a non-parametric test, the step trend test has the advantages of distribution free and having no restriction on missing data \citep{Hirsch1985}. This is particularly useful in rainfall analysis since rainfall data is usually skewed. On the other hand, the disadvantages of non-parametric tests, such as being limited to hypothesis testing and weaker in power, also hold for the step trend test \citep{Whitley2002}.

The regression model used here is a simple model. We only consider the important effects of historical trend, seasonality and climate drivers. Furthermore, no more than two large scale climatic indices were used in order to avoid over-parameterisation and multiple cross-correlations between climatic indices. A purpose of the regression model is to remove the variability in rainfall that is due to these known important factors. The model shows that seasonality, ENSO and IOD together explain no more than 20\% of the rainfall variability, and around 6 - 8\% on average is due to the climate drivers. This is consistent with the literature \citep[e.g.][]{Westra2010}. A large amount of variation is left in the model residuals, which is likely due to random factors. Rainfall is generally considered a stochastic process \citep[e.g.][]{Fowler2005,Cowpertwait2009,Burton2010}. The high variability increases the difficulty to detect a change in rainfall.

There was some, but no strong evidence that a step change has occurred in the QLD region. The semi-parametric model could not identify any step change in the rainfall data. On the other hand, the step trend test suggested a couple of locations (300 - 500 km$^2$) might have experienced rainfall change ($\alpha = 0.05$). The results from the step trend test also indicated a possible widespread change, but the spatial pattern of change is not consistent with the tree cover change map. Although land clearings in QLD have occurred at a high rate and broad scale \citep{SLATS2005}, this study cannot confirm its impact on the local rainfall. At this location, land cover change was not found to change the mean rainfall significantly in some other studies \citep[e.g.][]{Narisma2003,McAlpine2007}. The characteristics of vegetation cover changes in QLD might increase the difficulty in detecting a step change. QLD has a long history of land clearing. According to the series of SLATS reports on land cover changes in QLD released by the Queensland government, land clearing continued in and around the study region between 1988 - 2008. Major broad scale and high rate clearings occurred in 1999 - 2000 and 2002 - 2004. It is therefore difficult to define a clear cut change in this region. The more continuous ongoing land clearing could have reduced the significance of a step change. %Furthermore, woodlands were mainly replaced by pasture after the clearing 
%% comment: Can we show that 2006 is an anomaly in the residuals?
The two methods indicate quite different results on the level of change in the NSW/VIC region. The semi-parametric model showed that a large area in the NSW/VIC region has experienced significant step change ($\alpha = 0.05$) in rainfall after 2003. But this result is likely to be affected by the low rainfall in 2006 as the area is larger than the bushfires region. The step trend test was able to detect significant changes ($\alpha = 0.05$) in about 400 km$^2$ areas in the Snowy Mt. and some possible changes along the bushfires region ($\alpha = 0.1$). This might be due to the possible small size of the step change and/or a short after-fire period \citep{Hirsch1985}. The vegetation cover change due to bushfires might have possibly changed the rainfall but the evidence was not strong either.

%Both methods have highlighted the bushfires sites in the Snowy Mt. ranges. Especially in the step trend test results, a step change in rainfall in the bushfires locations is generally more significant than its surrounding regions. The step trend test results also suggested that the step change in rainfall, if any, had a  sign to the west of the bushfires area. This area corresponds to around 30\% of tree growth in the tree cover. This is evidence that the result of the step trend test is robust.

The bushfires locations highlighted in the analysis results is an interesting outcome. The results might be due to a drought effect. The 2002 - 2003 drought has affected rainfall in a large part of Australia, particularly the MDB \citep{Nicholls2004}. The severe bushfires in 2003 were also triggered by the extreme drought conditions. Although the dry episode effect on rainfall has been accounted for in part by the SOI, further impact of drought could be passed through the local land-atmosphere interaction. However, this impact might not be statistically significant, even under an extreme event such as bushfires. Overall, the rainfall feedback to the vegetation cover change could be weak under the dry conditions. We found similar results in the later chapters.

The different causes of vegetation cover change in these two regions could lead to different post-change characteristics. The magnitude of EVI decreasing trends in the QLD region are less than in NSW region, as reported in the DLCD data. This is due to the lower tree density in the QLD region than in the NSW/VIC region before land surface interventions. Wild fires might have totally damaged the vegetation cover and recovery was very slow in some areas within the Snowy Mt. The persistent drought in the 2000s \citep{Howden2012} has delayed the regrowth of trees. On the other hand, replacing tree cover with pasture and crops might have a relatively subtle impact on the EVI. On the other hand, regrowth of trees can be expected after fires or as drought condition is relieved; but clearings for agricultural purpose impose permanent or semi-permanent changes to the land surface.

The rainfall data used in this study is a gridded data set. This data set is robust and consistent over a long time series (from 1900 to current) and has a broad national wide coverage which can provide more information spatially. However, high cross correlation between pixels due to the interpolation method generating this data set can also introduce spatial noise. Here the cross correlation has been account for in the step trend test. Some other methods are also available which can be used to perform a comparative trial. For example, \citet{Narisma2007} applied a spatial Gaussian filter on a similar data set and used wavelet analysis to detect step change in rainfall. High quality station data is another option to test whether the observed spatial pattern in the step trend test results was not due to the gridded data itself. Resampling methods, such as bootstrapping and permutation \citep{Wilks1997,Kundzewicz2004,Westra2013}, can also be used to further assess the strength of significance of results and incorporate spatial and temporal patterns in the analysis. We are also aware that the gridded data set is most useful in regions with sparse rain gauge networks but it actually reduces information where the rain gauge density is high \citep{Jones2009}. In the Snowy Mt. areas, the coverage of rainfall stations is intensive but they are mainly located in the valleys. The interpolated data might not best represent the local rainfall.

The current study cannot provide strong evidence to reject the null hypothesis (no step change in rainfall is due to the tree cover loss). Limiting by the available data, the time frame under study has been chosen within a long lasting drought period \citep{Holper2011}. The strong impact of this prolong drought might have suppressed the land-atmosphere interaction and confused the cause-effect relation between rainfall and vegetation cover change. This could be one of the reasons that the LCC effects found in other studies \citep[e.g.]{Gorgen2006,McAlpine2007} are not found to be significant here. So possible future work can be conducted for non-drought period, when longer series of land cover data is available. This approach should be further assessed in different scenarios of wet period and/or afforestation. Continuous monitoring of land surface conditions is important for future research. The power of the test can also be improved with the longer length of the after-intervention period \citep{Hirsch1985}. But then the question is what is an appropriate length for land cover change study, when ongoing change is possible, such as tree regrowth or cropping. This will require further research.

% \citet{Zhang2007} discover the relation in the way of map contras

% \citet{Webb2005} do not find a strong relationship between forest cover and total rainfall, which appears to be influenced primarily by factors such as distance to the coast; but significant positive relationships between tree cover and the number of rain days consistently emerge.

%Both methods have identified some areas with rainfall change after the specified time of land cover change. This change in QLD appears to be less significant. Furthermore, the spatial distribution of significance of the changes do not well match with the tree cover change map. The results from the step trend test implied that a large scale change could have occurred, possibly due to the land clearing. The tree cover change map only shows a low level of changes (generally less than 20\%) spreading around the region between 2002 - 2005. The ongoing clearing activities in this region could be the reason that a step change cannot be detected. Both methods have better results in the NSW/VIC region, showing rainfall change in an area from the mountain range in the south to the west of ACT. The dummy variable is a significant predictor for rainfall in this area at the 0.05 level. In the step trend test the locations of bushfires is highlighted but the significance level is reduced. The two methods agree on the type of relationship. The regression analysis with dummy variable indicated that tree cover has a positive relationship with rainfall, while the step trend test showed that the tree cover loss might have decreased rainfall. However, the estimated relationship between land cover changes and the rainfall might not be highly significant. There is no strong evidence that the rainfall change is solely due to land cover change.

\section{Conclusion}

In this study, we found some evidence, although not strong, that vegetation cover change has changed local rainfall. The semi-parametric method and non-parametric method did not reach an agreement on detecting significant step change in rainfall in the hot dry QLD region where land clearing has occurred. On the other hand, the bushfires in the humid temperate mountain range in the NSW/VIC region could have reduced rainfall. But the dry spell also plays an important role in the results.

Drought has had a pronounced impact on the land surface condition during the study period, leading to significant reduction in vegetation and extreme events such as bushfires. The lack of rainfall associated and high temperature may cause a step change in the vegetation. Hence, the signal of LCC feedback on rainfall is probably weaker under such regional dry condition, as the impact of LCC on rainfall is mainly through changes in moisture convergence \citep{Gorgen2006,Pitman2007}.

%Empirical evidence of correlations between rainfall and land cover change are scarce due to multiple environmental factors that cannot be controlled. In this study we investigated the relationship by applying statistical methods on data from best available sources. Gridded data of rainfall and tree cover were used as spatially corresponding local conditions. Large scale effects were represented by climatic indicators, such as SOI and IOD. Regression analysis and step trend tests were used to assess the effect of abrupt land surface intervention.

%The step trend test is a better method than the regression method in this study. In order to obtain better results, the regression model requires fine details of the land covers. The defined land cover variable in the regression model does not significantly improve the model's reliability as the signal of change is smaller than the noise. Furthermore, the decrease of tree cover does not necessarily cause an immediate decrease in rainfall but it might alter the rainfall patterns or rainfall trend over the long term. While the regression analysis requires a long data record, the step trend test itself can work on a shorter data record. The step trend test can also achieve a better result than the regression analysis. Hence the step trend test is an easy and relatively reliable method to investigate the relationship between land cover changes and the rainfall when an abrupt change can be clearly identified.

%The general information from the tree cover data can not support the assumption  In QLD, the continuous land clearing over more than a decade might provide some explanations to the results. The selected time window of land cover change could be unrealistic and the intensity of tree cover change in 2002 - 2005 is not climatically significant. On the other hand, results from step trend test on 1998 - 2008 rainfall data in NSW/VIC region has a better visual match with the tree cover change map, which indicates the method works better when a abrupt change can be clearly identified. The results from the step trend test on 11-year data implies a positive relationship between the tree covers and the rainfall at 0.1 significance level. This is a weak evidence that the loss of tree covers has led to rainfall decreasing.

%This study is a new attempt to use the step trend test in detecting rainfall change and attribute the result to specified land cover change. Comparing with the generalized additive regression method, this test does not require detailed information on land cover change but still achieves similar or even improved results. However, the relatively short length of rainfall data, especially there were only four to five years data after the specified intervention, has limited the performance of the test. Furthermore, equivalent long time series of good quality land cover data is also neccessary to investigate the history of vegetation cover change and identify the most significant events.

%----------------------------------------------------------------------------------------
%	APPENDICES (OPTIONAL)
%----------------------------------------------------------------------------------------

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\appendix

\section{Summary of Data}

 \begin{sidewaystable}
 \caption{Summary of data.}
  \label{tab:ch3Data}
 \begin{tabular}{lllll}
  \hline
  \textbf{Data} & \textbf{Source} & \multicolumn{2}{c}{\textbf{Resolution}} & \textbf{Analysis period} \\\cline{3-4}
  & & Temporal & Spatial & \\\hline
  Percent tree cover & MOD44B & Annual &	250m & 2000-2010\\
  Trend of vegetation cover change  &	DLCD (2009)	& Onetime & 250m	& Trend of Apr 2000 - Apr 2008\\
  Rainfall &	AWAP gridded rainfall data &	Monthly &	0.05\textdegree$\times$0.05\textdegree & Jan 1979- Dec 2008\\
  SOI	& BoM	& Monthly &	N/A &	Jan 1979- Dec 2008\\
  NINO 3, 3.4, 4 &	IRI/LDEO data library	& Monthly	& N/A	& Jan 1979- Dec 2008\\
  PDO	& NOAA & Monthly	& N/A	& Jan 1979- Dec 2008\\
  IOD	& POAMA-2 dataset	& Monthly	& N/A	& Jan 1979- Dec 2008\\
  \hline
  \end{tabular}
\end{sidewaystable}

%----------------------------------------------------------------------------------------
%	ACKNOWLEDGEMENTS
%----------------------------------------------------------------------------------------

\begin{acknowledgments}
CL was supported by an Australian Postgraduate Award for this work.
\end{acknowledgments}

%----------------------------------------------------------------------------------------
%	BIBLIOGRAPHY
%----------------------------------------------------------------------------------------

\bibliographystyle{agufull08}
\bibliography{reflist_chap3}

\end{article}

\end{document}

%% backup old code

% chunk 7
<<chunk73,eval=F ,echo=F,results = hide>>=

#   tcr_r1 <- resample(aggregate(mod_tc_r1,fact=18,expand=F),reg_qld[[1]])
#   tcr_r2 <- resample(aggregate(mod_tc_r2,fact=18,expand=F),reg_nswvic[[1]])
# 
#   tc_r1 <- tcr_r1[[6]]-tcr_r1[[3]]
#   tc_r2 <- tcr_r2[[4]]-tcr_r2[[3]]
#   
#   
#   plotclr <- brewer.pal(6,"PiYG")
# 
#   setEPS()
#   postscript("R:\\AVSproject\\WrittenWork\\Paper\\Paper 11 - Australian data regression\\tc_qld.eps")
#   
#   par(mar=c(1,1,1,6))
#   plot(e111,border="white")
#   image(tc_r1,col=plotclr,breaks=c(-30,-20,-10,0,10,20,30),legend=F,add=T)
#   plot(mdb_reg,border="grey",add=T)
#   image.plot(zlim=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3),lab.breaks=c(-30,-20,-10,0,10,20,30),legend.only=T,horizontal=F,col=plotclr,legend.shrink=0.4,legend.width=0.7,legend.args=list(text="change (% ground)",line=0.5,cex=0.7,font=2),axis.args=list(cex.axis=0.8))
# 
#   dev.off()
#   
#   setEPS()
#   postscript("R:\\AVSproject\\WrittenWork\\Paper\\Paper 11 - Australian data regression\\tc_nswvic.eps")
#   
#   par(mar=c(1,1,1,6))
#   plot(e222,border="white")
#   image(tc_r2,col=plotclr,breaks=c(-30,-20,-10,0,10,20,30),legend=F,add=T)
#   plot(mdb_reg,border="grey",add=T)
#   plot(aus_reg,border="grey",add=T)  
#   image.plot(zlim=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3),lab.breaks=c(-30,-20,-10,0,10,20,30),legend.only=T,horizontal=F,col=plotclr,legend.shrink=0.4,legend.width=0.7,legend.args=list(text="change (% ground)",line=0.5,cex=0.7,font=2),axis.args=list(cex.axis=0.8))
# 
#   dev.off()

@

%The tree cover data is upscaled to match the resolution of the rainfall data. The upscaling process, taking the mean of pixel values, reduces the intensity of tree cover change, especially in the area where the change concentrates within small scale. Comparing the tree cover (as percentage of the ground area) in the QLD region between 2005 and 2002, light reduction of trees at the 5 km $\times$ 5 km resolution has been observed in the west and southwest of the region (Figure~\ref{fig:site1_tc}). In the NSW/VIC region, moderate to high tree cover change has shown along the MDB boundary (Figure~\ref{fig:site2_tc}). In the Snowy Mt. range, tree cover decreased between 2002 - 2003, while close to the southwest of ACT border there is increasing tree cover within the same period.

%\begin{figure}[htp]
%  \centering
%  \mbox{\subfigure[site 1: tree cover change between 2002 & 2005]{\label{fig:site1_tc}\includegraphics[width=2.5in]{chap3/tc_qld.eps}}
%  \quad
%  \subfigure[site 2: tree cover change between 2002 & 2003 ]{\label{fig:site2_tc}\includegraphics[width=2.5in]{chap3/tc_nswvic.eps}}}
%  \caption{Tree cover change as percentage of ground area in the two study regions. Green colours indicate tree cover increase and pink colours indicate tree cover decrease.}
%  \label{fig:tc}
%\end{figure}

%The tree cover change map and the rainfall change map show some similarity in the NSW/VIC region. Positive values of Z' statistics (decreasing rainfall) in the Snowy Mountains correspond to tree cover reduction in 2002 - 2003. Around the higher Z' values (positive, significant at 10\%), the tree cover reduction is relatively more intense. However, where the increasing rainfall (significant at 10\%) is seen, the tree cover appears to be a mixture of small postive and negative changes which are less than 10\% of the ground area.

% chunk 7
<<chunk72,eval=F ,echo=F,results = hide>>=

## function to determine whether the tree cover change is permanent
#   fun1 <- function(x,m) {
#     m1 <- mean(x[1:3])
#     if (m==1) {
#       m2 <- mean(x[6:11])
#     } else { m2 <- mean(x[4:11])}
#     if(!is.na(m2-m1) & abs(m2-m1)>5) {return (m2-m1)} else {return(NA)}
#   }
#   
#   tc_m_r1 <- calc(tcr_r1, function(x) fun1(x,1))
#   tc_m_r2 <- calc(tcr_r2, function(x) fun1(x,2))
#   
#   fun2 <- function(x,m) {
#     d1 <- sd(x[1:3])
#     if (m==1) {
#       d2 <- sd(x[6:11])
#     } else { d2 <- sd(x[4:11])}
#     if(!is.na(d2/d1) & d2/d1>1) {return(NA)} else {return (d2/d1)}
#   }
#   
#   tc_d_r1 <- calc(tcr_r1, function(x) fun2(x,1))
#   tc_d_r2 <- calc(tcr_r2, function(x) fun2(x,2))
#   
#   tc_m_r1 <- mask(tc_m_r1,tc_d_r1)
#   tc_m_r2 <- mask(tc_m_r2,tc_d_r2)
#   
# ## find out whether the trend and the change have the same sign  
#   temp_r1 <- tc_r1*tc_m_r1
#   temp_r1[temp_r1<0] <- NA
#   tc_r1 <- mask(tc_r1,temp_r1)
#   
#   temp_r2 <- tc_r2*tc_m_r2
#   temp_r2[temp_r2<0] <- NA
#   tc_r2 <- mask(tc_r2,temp_r2)
#   
# ## find out the relationship between Z' and tc
#   sigtc_z.qld <- tr_qld.step*tc_r1
#   sigtc_z.nswvic <- tr_nswvic.step*tc_r2
# 
#   setEPS()
#   postscript("R:\\AVSproject\\WrittenWork\\Paper\\Paper 11 - Australian data regression\\tc_z_qld.eps")
#   
#   par(mar=c(1,1,1,6))
#   plot(e111,border="grey")
#   image(sigtc_z.qld,col=c("blue","red"),breaks=c(-10,0,10),legend=F,add=T)
#   plot(e111,border="grey",add=T)
#   plot(mdb_reg,border="grey",add=T)
#   image.plot(zlim=c(-1,1),breaks=c(-1,0,1),lab.breaks=c("negative","","positive"),legend.only=T,horizontal=F,col=c("red","blue"),legend.shrink=0.2,legend.width=0.7,axis.args=list(cex.axis=0.8))
# 
#   dev.off()
#   
#   setEPS()
#   postscript("R:\\AVSproject\\WrittenWork\\Paper\\Paper 11 - Australian data regression\\tc_z_nswvic.eps")
#   
#   par(mar=c(1,1,1,6))
#   plot(e222,border="grey")
#   image(sigtc_z.nswvic,col=c("blue","red"),breaks=c(-20,0,5),legend=F,add=T)
#   plot(e222,border="grey",add=T)
#   plot(mdb_reg,border="grey",add=T)
#   plot(aus_reg,border="grey",add=T)
#   image.plot(zlim=c(-1,1),breaks=c(-1,0,1),lab.breaks=c("negative","","positive"),legend.only=T,horizontal=F,col=c("red","blue"),legend.shrink=0.2,legend.width=0.7,axis.args=list(cex.axis=0.8))
# 
#   dev.off()

@

%Some of the land cover changes shown in Figure~\ref{fig:tc} are not significant, which means it might not be the largest land intervention or the change is not persistent. In Figure~\ref{fig:tc_z}, the relationship between the significant tree cover change and Z' value is shown. Here ''significant tree cover change" is defined as: difference of average tree cover between periods before and after intervention is over 5\% of ground area; and the variance in the after period is at most as much as the before period. A positive relationship refers to increasing (decreasing) rainfall (as represent by negative (positive) Z' value) associated with increasing (decreasing) land cover change. Otherwise it is negative relationship. There is limited number of pixels that have significant land cover changes in each region respectively. In the QLD region, these pixels all indicate tree cover reduction. there is about equal chance of positive relationship and negative relationship. In the NSW/VIC region, positive relationship is dominant. Around the Snowy Mt. range where tree cover was reduced, rainfall probably decrease too. On the other hand, in the middle of the region above Tumbarumba, tree cover increase is associated with possible increase trend in rainfall.

%\begin{figure}[htp]
%  \centering
%  \mbox{\subfigure[site 1: relationship between Z' value and sig. tree cover change]{\label{fig:site1_tcz}\includegraphics[width=2.5in]{tc_z_qld.eps}}
%  \quad
%  \subfigure[site 2: relationship between Z' value and sig. tree cover change ]{\label{fig:site2_tcz}\includegraphics[width=2.5in]{tc_z_nswvic.eps}}}
%  \caption{The maps show location where tree cover change is significant. Blue colour indicates the relationship between tree cover change and rainfall is positive (i.e. increasing tree cover and increasing rainfall or vice versa). Red colour indicates the relationship is negative.}
%  \label{fig:tc_z}
%\end{figure}

%\begin{figure}[htp]
%  \centering
%  \subfigure[QLD]{\label{fig:trend_qld}\includegraphics[height=2.5in]{"cv_trend1"}}
%  \subfigure[NSW/VIC]{\label{fig:trend_nsw}\includegraphics[height=1in,trim=0cm 5cm 0cm 5cm, clip=T]{"cv_trend2"}}
%  \subfigure[WA]{\label{fig:trend_wa}\includegraphics[height=1in,trim=0cm 11.5cm 0cm 1.5cm,clip=T]{"cv_trend3"}}  
%  \caption{Average annual rainfall (left), coefficient of variance (CV) of monthly rainfall (middle) and decadal rainfall change due to trend (right) for the three study regions. The area with significant trend (at $\alpha=0.05$ and $\alpha=0.1$) are also shown as an orange transparent patch on all the plots.}%
%  \label{fig:trend_rain}
%\end{figure}

%% supplement R code for other purpose
<<other,eval=F ,echo=F,results = hide>>=

## mod_tc_e.l are the raw tree cover data for each region
# mod_tc_e2.l
# mod_tc_e4.l
# mod_tc_e5.l
# 
# LCtrend_qld2 <- addLayer(mod_tc_e2.l,tr_qld2.step)
# LCtrend_nswvic <- addLayer(mod_tc_e4.l,tr_nswvic.step)
# LCtrend_wa <- addLayer(mod_tc_e5.l,tr_wa.step)
# 
# cover.trend <- function(x,y,m) {
#   yr <- seq(1:11)
# #   if(is.na(y[x][12])) {
# #   return(NA) } else {
#     tc <- y[x][1:11]
#     fit.tr <- lm(tc~yr)
#     if(m==1) return(coef(summary.lm(fit.tr))[2,1])
#     if(m==2) return(coef(summary.lm(fit.tr))[2,4])
# #   }
# }
# 
# LCtr_qld2 <- lapply(1:ncell(mod_tc_e2.l),cover.trend,mod_tc_e2.l,1)
# LCtrq_qld2 <- lapply(1:ncell(mod_tc_e2.l),cover.trend,mod_tc_e2.l,2)
# LCtr.qld2 <- raster(tr_qld2.step)
# LCtr.qld2[1:ncell(tr_qld2.step)] <- LCtr_qld2
# LCtrq.qld2 <- raster(tr_qld2.step)
# LCtrq.qld2[1:ncell(tr_qld2.step)] <- LCtrq_qld2
# LCtrend_qld2 <- addLayer(LCtrend_qld2,LCtr.qld2,LCtrq.qld2)
# 
# LCtr_nswvic <- lapply(1:ncell(LCtrend_nswvic),cover.trend,LCtrend_nswvic,1)
# LCtrq_nswvic <- lapply(1:ncell(LCtrend_nswvic),cover.trend,LCtrend_nswvic,2)
# LCtr.nswvic <- raster(tr_nswvic.step)
# LCtr.nswvic[1:ncell(tr_nswvic.step)] <- LCtr_nswvic
# LCtrq.nswvic <- raster(tr_nswvic.step)
# LCtrq.nswvic[1:ncell(tr_nswvic.step)] <- LCtrq_nswvic
# LCtrend_nswvic <- addLayer(LCtrend_nswvic,LCtr.nswvic,LCtrq.nswvic)
# 
# LCtr_wa <- lapply(1:ncell(LCtrend_wa),cover.trend,LCtrend_wa,1)
# LCtrq_wa <- lapply(1:ncell(LCtrend_wa),cover.trend,LCtrend_wa,2)
# LCtr.wa <- raster(tr_wa.step)
# LCtr.wa[1:ncell(tr_wa.step)] <- LCtr_wa
# LCtrq.wa <- raster(tr_wa.step)
# LCtrq.wa[1:ncell(tr_wa.step)] <- LCtrq_wa
# LCtrend_wa <- addLayer(LCtrend_wa,LCtr.wa,LCtrq.wa)
# 
# ##----------------------------------------------------------------------------------------------
# ## central QLD
# plot(e12,border="grey",lwd=3)
# image(LCtr.qld2,col=c(plotclt[8],plotclt[3]),breaks=c(-3,0,2),add=T)
# image.plot(zlim=c(0,2),breaks=c(0,1,2),lab.breaks=c(-3,0,2),legend.only=T,horizontal=F,col=c(plotclt[8],plotclt[3]),legend.shrink=0.4,legend.width=0.7,legend.args=list(text="trend",line=0.5,cex=0.7,font=2),axis.args=list(cex.axis=0.8))
# 
# plot(e12,border="grey",lwd=3)
# image(LCtrq.qld2,col=rev(plotclt[6:10]),breaks=c(0,0.01,0.05,0.1,0.2,1),add=T)
# image.plot(zlim=c(0,5),breaks=c(0,1,2,3,4,5),lab.breaks=c(0,0.01,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=rev(plotclt[6:10]),legend.shrink=0.4,legend.width=0.7,legend.args=list(text="p",line=0.5,cex=0.7,font=2),axis.args=list(cex.axis=0.8))
# 
# ##----------------------------------------------------------------------------------------------
# ## NSWVIC
# plot(e22,border="grey",lwd=3)
# image(LCtrend_nswvic[[13]],col=c(plotclt[8],plotclt[3]),breaks=c(-6,0,3),add=T)
# image.plot(zlim=c(0,2),breaks=c(0,1,2),lab.breaks=c(-6,0,3),legend.only=T,horizontal=F,col=c(plotclt[8],plotclt[3]),legend.shrink=0.4,legend.width=0.7,legend.args=list(text="trend",line=0.5,cex=0.7,font=2),axis.args=list(cex.axis=0.8))
# 
# plot(e22,border="grey",lwd=3)
# image(LCtrend_nswvic[[14]],col=rev(plotclt[6:10]),breaks=c(0,0.01,0.05,0.1,0.2,1),add=T)
# image.plot(zlim=c(0,5),breaks=c(0,1,2,3,4,5),lab.breaks=c(0,0.01,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=rev(plotclt[6:10]),legend.shrink=0.4,legend.width=0.7,legend.args=list(text="p",line=0.5,cex=0.7,font=2),axis.args=list(cex.axis=0.8))
# 
# ##----------------------------------------------------------------------------------------------
# ## WA
# plot(e31,border="grey",lwd=3)
# image(LCtrend_wa[[13]],col=c(plotclt[8],plotclt[3]),breaks=c(-5,0,3),add=T)
# image.plot(zlim=c(0,2),breaks=c(0,1,2),lab.breaks=c(-5,0,3),legend.only=T,horizontal=F,col=c(plotclt[8],plotclt[3]),legend.shrink=0.4,legend.width=0.7,legend.args=list(text="trend",line=0.5,cex=0.7,font=2),axis.args=list(cex.axis=0.8))
# 
# plot(e31,border="grey",lwd=3)
# image(LCtrend_wa[[14]],col=rev(plotclt[6:10]),breaks=c(0,0.01,0.05,0.1,0.2,1),add=T)
# image.plot(zlim=c(0,5),breaks=c(0,1,2,3,4,5),lab.breaks=c(0,0.01,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=rev(plotclt[6:10]),legend.shrink=0.4,legend.width=0.7,legend.args=list(text="p",line=0.5,cex=0.7,font=2),axis.args=list(cex.axis=0.8))
# 
# 
# ## calculate difference between any two years  
#   fun <- function(x) {
#     tc_diff <- diff(x)
#     y <- x[1:10]
#     tc_change <- tc_diff/y
#     return(tc_change)
#   }
#   mod_tc_e1 <- calc(mod_tc_r1,fun)
#   mod_tc_e2 <- calc(mod_tc_r2,fun)
#   
# ## set pixels with max change <0.5 or >100 to NA  
#   temp_qld <- stackApply(abs(mod_tc_e1),indices=rep(1,10),fun=max)
#   temp_nsw <- stackApply(abs(mod_tc_e2),indices=rep(1,10),fun=max)
#   
#   fun <- function(x) {
#     x[x<0.5 | x>100] <- NA
#     return(x)
#   }
#   temp_qld <- calc(temp_qld,fun)
#   mod_tc_r1 <- mask(mod_tc_r1,temp_qld)
#   mod_tc_e1 <- mask(mod_tc_e1,temp_qld)
#   
#   temp_nsw <- calc(temp_nsw,fun)
#   mod_tc_r2 <- mask(mod_tc_r2,temp_nsw)
#   mod_tc_e2 <- mask(mod_tc_e2,temp_nsw)  
#   
# ## calculate average tree cover over 11 years. discard pixels with <5% tree cover
#   mod44_tc_mean <- calc(mod44_tc_aus,mean)
#   fun <- function(x) {x[x<5 | x==-1 | x==253]<-NA; return(x)}
#   mod44_tc_mean <- calc(mod44_tc_mean,fun)  
#   
# ## analysis change of tree cover change
# ## 1. cover trend and significant
#   cover.trend <- function(x,y) {
#     yr <- seq(1:length(y[x]))
#     tc <- y[x]
#     if (is.na(mean(tc))) {
#       return(NA)
#     } else {
#       fit.tr <- lm(tc~yr)
#       return(coef(summary.lm(fit.tr)))
#     }
#   }
# 
#   LC_qld <- lapply(1:ncell(mod_tc_r1),cover.trend,mod_tc_r1)
#   
#   LCtr_qld <- raster(mod_tc_r1)
#   LCtr_qld[] <- sapply(LC_qld,function(x) if(!is.na(x)) {x[2,1]} else {x})
#   
#   LCtrp_qld <- raster(mod_tc_r1)
#   LCtrp_qld[] <- sapply(LC_qld,function(x) if(!is.na(x)) {x[2,4]} else {x})
#   
#   LC_nswvic <- lapply(1:ncell(mod_tc_r2),cover.trend,mod_tc_r2)
#   
#   LCtr_nswvic <- raster(mod_tc_r2)
#   LCtr_nswvic[] <- sapply(LC_nswvic,function(x) if(!is.na(x)) {x[2,1]} else {x})
#   
#   LCtrp_nswvic <- raster(mod_tc_r2)
#   LCtrp_nswvic[] <- sapply(LC_nswvic,function(x) if(!is.na(x)) {x[2,4]} else {x})
#   
# ## note: in VIC, the two data sets match well in terms of the large trend  
#   
# ## 2. maximum jump and its position of tree cover change
#   cover_change <- function(x,y) {
#     temp <- y[x]
#     if(is.na(mean(temp))) {
#       return(NA)
#     } else {
#       jump <- temp[which.is.max(abs(temp))]
#       jump.position <- which.is.max(abs(temp))+1   # the position of after jump
#       bef.m <- mean(temp[1:(jump.position-1)])
#       aft.m <- mean(temp[jump.position:length(temp)])
#       bef.v <- sd(temp[1:(jump.position-1)])
#       aft.v <- sd(temp[jump.position:length(temp)])
#           }
#     return(data.frame(jump=jump,jump.pos=jump.position,bef.m=bef.m,bef.v=bef.v,aft.m=aft.m,aft.v=aft.v))
#   }
# 
#   LC1_qld <- lapply(1:ncell(mod_tc_e1),cover_change,mod_tc_e1)
#   
#   qld.jump <- raster(mod_tc_e1)
#   qld.jump[] <- sapply(LC1_qld,function(x) if(!is.na(x)) {x$jump} else {x})
# 
#   qld.pos <- raster(mod_tc_e1)
#   qld.pos[] <- sapply(LC1_qld,function(x) if(!is.na(x)) {x$jump.pos} else {x})
#   
#   qld.befm <- raster(mod_tc_e1)
#   qld.befm[] <- sapply(LC1_qld,function(x) if(!is.na(x)) {x$bef.m} else {x})  
#   
#   qld.befv <- raster(mod_tc_e1)
#   qld.befv[] <- sapply(LC1_qld,function(x) if(!is.na(x)) {x$bef.v} else {x})  
#   
#   qld.aftm <- raster(mod_tc_e1)
#   qld.aftm[] <- sapply(LC1_qld,function(x) if(!is.na(x)) {x$aft.m} else {x}) 
#   
#   qld.aftv <- raster(mod_tc_e1)
#   qld.aftv[] <- sapply(LC1_qld,function(x) if(!is.na(x)) {x$aft.v} else {x})  
#   
#   LC1_nswvic <- lapply(1:ncell(mod_tc_e2),cover_change,mod_tc_e2)
#   
#   nswvic.jump <- raster(mod_tc_e2)
#   nswvic.jump[] <- sapply(LC1_nswvic,function(x) if(!is.na(x)) {x$jump} else {x})  
#   
#   nswvic.pos <- raster(mod_tc_e2)
#   nswvic.pos[] <- sapply(LC1_nswvic,function(x) if(!is.na(x)) {x$jump.pos} else {x}) 
#   
#   nswvic.befm <- raster(mod_tc_e2)
#   nswvic.befm[] <- sapply(LC1_nswvic,function(x) if(!is.na(x)) {x$bef.m} else {x})  
#   
#   nswvic.befv <- raster(mod_tc_e2)
#   nswvic.befv[] <- sapply(LC1_nswvic,function(x) if(!is.na(x)) {x$bef.v} else {x})  
#   
#   nswvic.aftm <- raster(mod_tc_e2)
#   nswvic.aftm[] <- sapply(LC1_nswvic,function(x) if(!is.na(x)) {x$aft.m} else {x})  
#   
#   nswvic.aftv <- raster(mod_tc_e2)
#   nswvic.aftv[] <- sapply(LC1_nswvic,function(x) if(!is.na(x)) {x$aft.v} else {x})  
#   
# ## find jumps that occur within the first 5 years (2000 - 2004)
# 
#   LC2_qld <- brick(qld.jump,qld.pos,qld.befm,qld.befv,qld.aftm,qld.aftv)
#   qld.pos[qld.pos>5] <- NA
#   LC2_qld <- mask(LC2_qld,qld.pos)
#   
#   LC2_nswvic <- brick(nswvic.jump,nswvic.pos,nswvic.befm,nswvic.befv,nswvic.aftm,nswvic.aftv)
#   nswvic.pos[nswvic.pos>5] <- NA
#   LC2_nswvic <- mask(LC2_nswvic,nswvic.pos)

@

%% chunk 10
<<chunk10,eval=F ,echo=F,results = hide>>=

## climatic indicator data, 100 years

# file1 <- "R:\\Data\\BoM\\SOI\\SOI.csv"
# SOI <- read.csv(file1,sep=",",skip=6)
# SOI <- melt(SOI,id="Year")
# names(SOI) <- c("Year","Month","SOI")
# SOI <- SOI[order(SOI$Year,SOI$Month),]
# SOI_ts <- ts(SOI$SOI,start=c(1876,1),end=c(2014,4),frequency=12)
# SOI_100a <- window(SOI_ts,start=c(1900,1),end=c(2013,12))
# ## deseasonalised SOI
# SOI_100 <- stl(SOI_100a,"period")$time.series[,3]+stl(SOI_100a,"period")$time.series[,2]
# 
# file2 <- "R:\\Data\\BoM\\SST\\SST34.cdf"
# SST34 <- open.ncdf(file2)
# SST34 <- get.var.ncdf(SST34,SST34$var[[1]])
# SST34_ts <- ts(SST34,start=c(1856,1),end=c(2014,4),frequency=12)
# SST34_100a <- window(SST34_ts,start=c(1900,1),end=c(2013,12))
# ## deseasonalised SST34
# SST34_100 <- stl(SST34_100a,"period")$time.series[,3]+stl(SST34_100a,"period")$time.series[,2]
# 
# file3 <- "R:\\Data\\BoM\\SST\\SST3.cdf"
# SST3 <- open.ncdf(file2)
# SST3 <- get.var.ncdf(SST3,SST3$var[[1]])
# SST3_ts <- ts(SST3,start=c(1856,1),end=c(2014,4),frequency=12)
# SST3_100a <- window(SST3_ts,start=c(1900,1),end=c(2013,12))
# ## deseasonalised SST34
# SST3_100 <- stl(SST3_100a,"period")$time.series[,3]+stl(SST3_100a,"period")$time.series[,2]
# 
# file4 <- "R:\\Data\\BoM\\SST\\SST4.cdf"
# SST4 <- open.ncdf(file4)
# SST4 <- get.var.ncdf(SST4,SST4$var[[1]])
# SST4_ts <- ts(SST4,start=c(1856,1),end=c(2014,5),frequency=12)
# SST4_100a <- window(SST4_ts,start=c(1900,1),end=c(2013,12))
# ## deseasonalised SST4
# SST4_100 <- stl(SST4_100a,"period")$time.series[,3]+stl(SST4_100a,"period")$time.series[,2]
# 
# file5 <- "R:\\Data\\BoM\\PDO\\PDO.latest_1900-.txt"
# pdo_mth <- read.csv(file5,sep="",header=T)
# #names(pdo_mth) <- c("year","Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")
# pdo_mth <- melt.data.frame(pdo_mth,id="YEAR")
# names(pdo_mth) <- c("Year","Month","PDO")
# pdo_mth <- pdo_mth[order(pdo_mth$Year,pdo_mth$Month),]
# PDO_ts <- ts(pdo_mth$PDO,start=c(1900,1),end=c(2014,4),frequency=12)
# PDO_100a <- window(PDO_ts,start=c(1900,1),end=c(2013,12))
# 
# file7 <- "R:\\Data\\BoM\\IOD\\iod_jamstec.txt"
# IOD <- read.csv(file7,sep="",header=T)
# IOD <- IOD[,4]
# IOD_ts <- ts(IOD,start=c(1870,1),frequency=12)
# IOD_100a <- window(IOD_ts,start=c(1900,1),end=c(2013,12))
# 
# ## 100 years awap data
# nsw_100 <- read.table("C:\\Users\\Sue\\Desktop\\thesis\\NSW_month_rf.csv",header=T)
# nsw100_mean <- ts(rowMeans(nsw_100[,2:1185],na.rm=T,dims=1),start=c(1900,1),frequency=12)
# 
# qld_100 <- read.csv("C:\\Users\\Sue\\Desktop\\thesis\\QLD_month_rf.csv",header=T)
# qld100_mean <- ts(rowMeans(qld_100[,2:2801],na.rm=T,dims=1),start=c(1900,1),frequency=12)
# 
# require(MASS)
# fitdistr(qld100_mean,"gamma")  ## assume shape=1
# fitdistr(nsw100_mean,"gamma")  ## assume shape =8
# ks.test(qld100_mean,1,0.03)
# ks.test(nsw100_mean,8,0.03)
#   
# ## autocorrelation in rainfall
# setEPS()
# postscript("autocorrelation100.eps")
# par(mfrow=c(2,2))
#   acf(qld100_mean,lag.max=12,main="QLD")
#   acf(nsw100_mean,lag.max=12,main="NSW & VIC")
# 
#   pacf(qld100_mean,lag.max=12,main="")
#   pacf(nsw100_mean,lag.max=12,main="")  
# dev.off()
#   
# ## deseasonalise & detrend rainfall: assume the deseasonalised monthly zonal rainfall is normal distributed and can be used in pearson cross-correlation function
# ## use stl function in R
#   qld100_deseason <- stl(qld100_mean,s.window=length(qld100_mean),t.window=length(qld100_mean))$time.series[,3]
#   nsw100_deseason <- stl(nsw100_mean,s.window=length(nsw100_mean),t.window=length(nsw100_mean))$time.series[,3]
# 
# setEPS()
# postscript("qld100_mean_ts.eps")
# plot(qld100_deseason,xlab="Year",ylab="monthly rainfall")
# abline(v=2003,col="red",lwd=2)
# text(2004,150,"2003",cex=0.9,col="red")
# dev.off()
# 
# setEPS()
# postscript("nsw100_mean_ts.eps")
# plot(nsw100_deseason,xlab="Year",ylab="monthly rainfall")
# abline(v=2003,col="red",lwd=2)
# text(2004,150,"2003",cex=0.9,col="red")
# dev.off()
#   
# setEPS()
# postscript("cor_qld100.eps")
# par(mfrow=c(2,3),mar=c(5,4,1,1))
# ccf(as.vector(SOI_100a),as.vector(qld100_deseason),lag.max=12,xlim=c(-12,0),xlab="SOI  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
# ccf(as.vector(SST34_100a),as.vector(qld100_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO3.4  Lag",ylab="CCF")
# ccf(as.vector(SST4_100a),as.vector(qld100_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO4  Lag",ylab="CCF")
# ccf(as.vector(PDO_100a),as.vector(qld100_deseason),lag.max=24,xlim=c(-24,0),xlab="PDO Lag",ylab="CCF")
# ccf(as.vector(PDO_100a*SOI_100a),as.vector(qld100_deseason),lag.max=12,xlim=c(-12,0),xlab="PDO*SOI  Lag",ylab="CCF")
# ccf(as.vector(IOD_100a),as.vector(qld100_deseason),lag.max=12,xlim=c(-12,0),xlab="IOD  Lag",ylab="CCF")
# dev.off()
# 
# ccf(subset(SOI_100a,cycle(SOI_100a)==10|cycle(SOI_100a)==11|cycle(SOI_100a)==12),subset(qld100_deseason,cycle(qld100_deseason)==10|cycle(qld100_deseason)==11|cycle(qld100_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="SOI  Lag",ylab="CCF") ## SOI
# ccf(subset(SST34_100a,cycle(SST34_100a)==10|cycle(SST34_100a)==11|cycle(SST34_100a)==12),subset(qld100_deseason,cycle(qld100_deseason)==10|cycle(qld100_deseason)==11|cycle(qld100_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="SST34  Lag",ylab="CCF") ## SST34
# ccf(subset(SST3_100a,cycle(SST3_100a)==10|cycle(SST3_100a)==11|cycle(SST3_100a)==12),subset(qld100_deseason,cycle(qld100_deseason)==10|cycle(qld100_deseason)==11|cycle(qld100_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="SST3  Lag",ylab="CCF") ## SST3
# ccf(subset(PDO_100a,cycle(PDO_100a)==10|cycle(PDO_100a)==11|cycle(PDO_100a)==12),subset(qld100_deseason,cycle(qld100_deseason)==10|cycle(qld100_deseason)==11|cycle(qld100_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="PDO  Lag",ylab="CCF") ## PDO
# ccf(subset(PDO_100a,cycle(PDO_100a)==10|cycle(PDO_100a)==11|cycle(PDO_100a)==12)* subset(SOI_100a,cycle(SOI_100a)==10|cycle(SOI_100a)==11|cycle(SOI_100a)==12),subset(qld100_deseason,cycle(qld100_deseason)==10|cycle(qld100_deseason)==11|cycle(qld100_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="PDO  Lag",ylab="CCF") ## PDO
# ccf(subset(IOD_100a,cycle(IOD_100a)==10|cycle(IOD_100a)==11|cycle(IOD_100a)==12),subset(qld100_deseason,cycle(qld100_deseason)==10|cycle(qld100_deseason)==11|cycle(qld100_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="IOD  Lag",ylab="CCF") ## IOD
# 
# 
# 
# # ## 1,2,3: SOI
# # ## 4,5,6: SOI
# # ## 7,8,9: SOI, IOD
# # ## 10,11,12: SOI, SST34, SST3, PDO
# # 
# # 
# setEPS()
# postscript("cor_nsw100.eps")
# par(mfrow=c(2,3),mar=c(5,4,1,1))
# ccf(as.vector(SOI_100a),as.vector(nsw100_deseason),lag.max=12,xlim=c(-12,0),xlab="SOI  Lag",ylab="CCF")
# ccf(as.vector(SST34_100a),as.vector(nsw100_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO3.4  Lag",ylab="CCF")
# ccf(as.vector(SST4_100a),as.vector(nsw100_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO4  Lag",ylab="CCF")
# ccf(as.vector(PDO_100a),as.vector(nsw100_deseason),lag.max=24,xlim=c(-24,0),xlab="PDO Lag",ylab="CCF")
# ccf(as.vector(PDO_100a*SOI_100a),as.vector(nsw100_deseason),lag.max=12,xlim=c(-12,0),xlab="PDO*SOI  Lag",ylab="CCF")
# ccf(as.vector(IOD_100a),as.vector(nsw100_deseason),lag.max=12,xlim=c(-12,0),xlab="IOD  Lag",ylab="CCF")
# dev.off()
# 
# ccf(subset(SOI_100a,cycle(SOI_100a)==7|cycle(SOI_100a)==8|cycle(SOI_100a)==9),subset(nsw100_deseason,cycle(nsw100_deseason)==7|cycle(nsw100_deseason)==8|cycle(nsw100_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="SOI  Lag",ylab="CCF") ## SOI
# ccf(subset(SST34_100a,cycle(SST34_100a)==7|cycle(SST34_100a)==8|cycle(SST34_100a)==9),subset(nsw100_deseason,cycle(nsw100_deseason)==7|cycle(nsw100_deseason)==8|cycle(nsw100_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="SST34  Lag",ylab="CCF") ## SST34
# ccf(subset(SST3_100a,cycle(SST3_100a)==7|cycle(SST3_100a)==8|cycle(SST3_100a)==9),subset(nsw100_deseason,cycle(nsw100_deseason)==7|cycle(nsw100_deseason)==8|cycle(nsw100_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="SST3  Lag",ylab="CCF") ## SST3
# ccf(subset(PDO_100a,cycle(PDO_100a)==7|cycle(PDO_100a)==8|cycle(PDO_100a)==9),subset(nsw100_deseason,cycle(nsw100_deseason)==7|cycle(nsw100_deseason)==8|cycle(nsw100_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="PDO  Lag",ylab="CCF") ## PDO
# ccf(subset(PDO_100a,cycle(PDO_100a)==7|cycle(PDO_100a)==8|cycle(PDO_100a)==9)* subset(SOI_100a,cycle(SOI_100a)==7|cycle(SOI_100a)==8|cycle(SOI_100a)==9),subset(nsw100_deseason,cycle(nsw100_deseason)==7|cycle(nsw100_deseason)==8|cycle(nsw100_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="PDO  Lag",ylab="CCF") ## PDO
# ccf(subset(IOD_100a,cycle(IOD_100a)==7|cycle(IOD_100a)==8|cycle(IOD_100a)==9),subset(nsw100_deseason,cycle(nsw100_deseason)==7|cycle(nsw100_deseason)==8|cycle(nsw100_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="IOD  Lag",ylab="CCF") ## IOD

## 1,2,3: NONE
#4,5,6: SOI
# 7, 8, 9: SOI, IOD
## 10, 11, 12: SOI, SST34, SST3, IOD

### 2017-11-08 chap3_paper_WVversion.tex
add files

--- 2017-11-08 chap3_paper_WVversion.Rnw
@@
 
+\usepackage{Sweave}
 \begin{document}
-\SweaveOpts{concordance=TRUE}
+\input{chap3_paper_WVversion-concordance}
 
@@
 \end{figure}
-
-<<mapping,eval=F, echo = F, results = hide>>=
-
-library(plotKML)
-library(raster)
-library(rgeos)
-library(maptools)
-library(fields)
-library(RColorBrewer)
-
-## define the boundary of sub-regions
-  e1 <- extent(c(1324304.091,1723901.773,-222556.668,80451.265))
-  e2 <- extent(c(1346395.141,1534425.527,-1169402.246,-607644.171))
-
-  e11 <- extent(c(145.425, 149.025, -28.475, -25.175))
-  e22 <- extent(c(146.675, 149.325, -37.075, -34.675))
-
-  e111 <- extent(c(145.825, 148.625, -28.075, -25.575)) #as(tc_r1,"SpatialPolygons")
-  e222 <- extent(c(147.075, 148.925, -36.675, -35.075)) #as(tc_r2,"SpatialPolygons")
-
-## the selected regions for mapping
-
-  aus_reg <- readShapePoly("Data\\Geo\\aus_state2011\\simplify_state11_uproj.shp",proj4string=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0"))
-  aus_reg2 <- gSimplify(aus_reg,0.01,topologyPreserve=T)
-
-  mdb_reg <- readShapePoly("Data\\Geo\\MDB\\mdb.shp",proj4string=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0"))
-mdb_reg2 <- gSimplify(mdb_reg,0.01,topologyPreserve=TRUE)
-
-  remove(list=c("aus_reg","mdb_reg"))
-
-## Burned area map
-
-fire_0301 <- raster("Data\\Other\\bushfires\\Burnt_Area\\tmpA2003001.burndate.tif")
-fire_0301nsw <- crop(fire_0301,e2)
-fire_loc <- crop(reproject(fire_0301nsw,CRS=CRS("+proj=longlat +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84")),e222)
-remove(list=c("fire_0301","fire_0301nsw"))
-
-  setEPS()
-  postscript("bushfire_nswvic.eps")
-#  jpg("bushfire_nswvic.jpg")
-  par(mar=c(3,3,1,6),xpd=T)
-  plot(e222,border="white")  
-  image(fire_loc,add=T, breaks=c(0,0.99,366.5,10000),col=c("white","red","white"))
-  plot(e222,add=T)
-  plot(aus_reg2,border="grey",add=T)
-  plot(mdb_reg2,border="grey",lwd=3,add=T)
-#   text(1430000,-1000000,"study region",cex=1.2)
-  text(149.1, -35.3,"ACT",font=2,cex=1.2)
-  text(148.4, -36,"Kosciuszko",font=2,cex=0.9)#,srt=90)
-  text(147.5, -36.4,"VIC",font=2,cex=1.2)
-  text(148, -35.5,"NSW",font=2,cex=1.2)
-  image.plot(zlim=c(0,1),breaks=c(0,0.5,1),lab.breaks=c("","burned \narea",""),legend.only=T,horizontal=F,col=c("red","red"),legend.shrink=0.03,legend.width=1.3,legend.args=list(text="",line=0),lwd.poly=0.1)
-legend(149,-36.1,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=0.9)
- dev.off()
-
-@
-
-<<treeCover,eval=F , echo = F, results = hide>>=
-
-## trend of evi (MODIS product) from GA and ABARES
-
-## tree cover data
-  mod44_tc <- brick("Data\\Satellite\\MOD44\\B\\mod44_tc_aus.tif")
-  tm <- seq(as.Date("2000-03-06"), as.Date("2010-03-06"), "year")
-  mod44_tc <- setZ(mod44_tc, tm, "year")
-
-## tree cover in sub-region
-  mod_tc_r11 <- crop(mod44_tc,e1)
-  mod_tc_r111 <- reproject(mod_tc_r11,CRS=CRS("+proj=longlat +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84"))
-  mod_tc_r1 <- crop(mod_tc_r111,e111)
-
-  mod_tc_r22 <- crop(mod44_tc,e2)
-  mod_tc_r222 <- reproject(mod_tc_r22,CRS=CRS("+proj=longlat +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84"))
- mod_tc_r2 <- crop(mod_tc_r222,e222)
-
- remove(list=c("mod44_tc","mod_tc_r11","mod_tc_r111","mod_tc_r22","mod_tc_r222"))
-
-## apply t-test to tree cover data
-  fun <- function(x,m) {
-    var_x <- sd(x)
-    mu_1 <- mean(x[1:3])
-    if (m==1) {y <- x[6:11]} else {y <- x[4:11]}   
-    mu_2 <- mean(y)
-    z <- (mu_2-mu_1)/var_x
-    if (!is.na(z) & abs(z)>1.645) {return(y[1]-x[3])} else {return(NA)}
-  }
-
-  mod_tc_r1.z <- calc(mod_tc_r1,function(x) fun(x,1))
-  mod_tc_r2.z <- calc(mod_tc_r2,function(x) fun(x,2))
-
-  remove(list=c("mod_tc_r1","mod_tc_r2"))
-
-## create plot
-  plotclr <- brewer.pal(6,"PiYG")
-#"#C51B7D" "#E9A3C9" "#FDE0EF" "#E6F5D0" "#A1D76A" "#4D9221"
-
-  setEPS()
-  postscript("tc_qld.eps")
-  par(mar=c(3,3,1,7))
-  plot(e111)
-  image(mod_tc_r1.z,col=plotclr,breaks=c(-70,-50,-20,0,20,50,70),legend=F,add=T)
-  plot(e111,add=T)
-  plot(mdb_reg2,border="grey",lwd=3,add=T)
-  image.plot(zlim=c(-3,3),breaks=c(-3,-2,-1,-0.2,0.2,1,2,3),lab.breaks=c(-70,-50,-20,"no tree","0 or",20,50,70),legend.only=T,horizontal=F,col=c("#C51B7D","#E9A3C9","#FDE0EF","white","#E6F5D0","#A1D76A","#4D9221"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text=expression(paste(Delta,"tree (% ground)",sep="")),line=0.5,cex=0.9,font=2),axis.args=list(cex.axis=0.8),graphics.reset=F)
-  dev.off()
-
-  setEPS()
-  postscript("tc_nswvic.eps")
-  par(mar=c(3,3,1,7))
-  plot(e222)
-  image(mod_tc_r2.z,col=plotclr,breaks=c(-70,-50,-20,0,20,50,70),legend=F,add=T)
-  plot(e222,add=T)
-  plot(aus_reg2,border="grey",add=T)
-  plot(mdb_reg2,border="grey",lwd=3,add=T)
-#   image.plot(zlim=c(-3,3),breaks=c(-3,-2,-1,-0.2,0.2,1,2,3),lab.breaks=c(-70,-50,-20,"no tree","0 or",20,50,70),legend.only=T,horizontal=F,col=c("#C51B7D","#E9A3C9","#FDE0EF","white","#E6F5D0","#A1D76A","#4D9221"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text=expression(paste(Delta,"tree (% ground)",sep="")),line=0.5,cex=0.7,font=2),axis.args=list(cex.axis=0.8))
-# legend(149,-36.5,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=0.9)
-  dev.off()
-
-setEPS()
-postscript("map.eps")
-par(mar=c(1.5,1.5,1.5,1.5))
-plot(mdb_reg2,border="grey",lwd=3)
-plot(aus_reg2,border="grey",add=T)
-legend(150.2,-36,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=0.9)
-#invisible(text(getSpPPolygonsLabptSlots(aus_reg2), labels=as.character(aus_reg2$NAME), cex=0.7,adj=0.5))
-box(lty=1)
-plot(e111,col="red",add=T)
-plot(e222,col="red",add=T)
-text(147.2,-26.8,"QLD",cex=1.5)
-text(148,-34.7,"NSW/VIC",cex=1.5)
-dev.off()
-
-@
 
@@
 
-<<climaticIndex,eval=F , echo = F, results = hide>>=
-
-# setwd("R:\\AVSproject\\WrittenWork\\PhD_thesis\\chap3")
-# setwd("B:/My Study/PhD/AVSproject_laptop/WrittenWork/PhD_thesis/chap3")
-#setwd("B:/My Study/PhD/AVSproject_PhD/WrittenWork/Paper/Paper 1 - chapter 3")
-
-library(reshape)
-library(chron)
-library(fields)
-library(rgdal)
-library(raster)
-library(maptools)
-library(graphics)
-library(zoo)
-library(RColorBrewer) # creates nice color schemes
-# library(classInt)     # finds class intervals for continuous variables
-
-library(mgcv)
-library(ncdf)
-library(nnet)
-
-## climatic indicator data
-
-file1 <- "Data\\BoM\\SOI\\SOI.csv"
-SOI <- read.csv(file1,sep=",",skip=6)
-SOI <- melt(SOI,id="Year")
-names(SOI) <- c("Year","Month","SOI")
-SOI <- SOI[order(SOI$Year,SOI$Month),]
-SOI_ts <- ts(SOI$SOI,start=c(1876,1),end=c(2014,4),frequency=12)
-SOI_100a <- window(SOI_ts,start=c(1900,1),end=c(2008,12))
-SOI_30a <- window(SOI_ts,start=c(1979,1),end=c(2008,12))
-## deseasonalised SOI
-SOI_100 <- stl(SOI_100a,"period")$time.series[,3]+stl(SOI_100a,"period")$time.series[,2]
-SOI_30 <- stl(SOI_30a,"period")$time.series[,3]+stl(SOI_30a,"period")$time.series[,2]
-
-file2 <- "Data\\BoM\\SST\\SST34.cdf"
-SST34 <- open.ncdf(file2)
-SST34 <- get.var.ncdf(SST34,SST34$var[[1]])
-SST34_ts <- ts(SST34,start=c(1856,1),end=c(2014,4),frequency=12)
-SST34_30a <- window(SST34_ts,start=c(1979,1),end=c(2008,12))
-## deseasonalised SST34
-SST34_30 <- stl(SST34_30a,"period")$time.series[,3]+stl(SST34_30a,"period")$time.series[,2]
-
-file3 <- "Data\\BoM\\SST\\SST3.cdf"
-SST3 <- open.ncdf(file3)
-SST3 <- get.var.ncdf(SST3,SST3$var[[1]])
-SST3_ts <- ts(SST3,start=c(1856,1),end=c(2014,4),frequency=12)
-SST3_30a <- window(SST3_ts,start=c(1979,1),end=c(2008,12))
-## deseasonalised SST34
-SST3_30 <- stl(SST3_30a,"period")$time.series[,3]+stl(SST3_30a,"period")$time.series[,2]
-
-file4 <- "Data\\BoM\\SST\\SST4.cdf"
-SST4 <- open.ncdf(file4)
-SST4 <- get.var.ncdf(SST4,SST4$var[[1]])
-SST4_ts <- ts(SST4,start=c(1856,1),end=c(2014,5),frequency=12)
-SST4_30a <- window(SST4_ts,start=c(1979,1),end=c(2008,12))
-## deseasonalised SST4
-SST4_30 <- stl(SST4_30a,"period")$time.series[,3]+stl(SST4_30a,"period")$time.series[,2]
-
-file5 <- "Data\\BoM\\PDO\\PDO.latest_1900-.txt"
-pdo_mth <- read.csv(file5,sep="",header=T)
-pdo_mth <- melt.data.frame(pdo_mth,id="YEAR")
-names(pdo_mth) <- c("Year","Month","PDO")
-pdo_mth <- pdo_mth[order(pdo_mth$Year,pdo_mth$Month),]
-PDO_ts <- ts(pdo_mth$PDO,start=c(1900,1),end=c(2014,4),frequency=12)
-PDO_100a <- window(PDO_ts,start=c(1900,1),end=c(2008,12))
-## deseasonalised PDO
-PDO_100 <- stl(PDO_100a,"period")$time.series[,3]+stl(PDO_100a,"period")$time.series[,2]
-
-file7 <- "Data\\BoM\\IOD\\iod_jamstec.txt"
-IOD <- read.csv(file7,sep="",header=T)
-IOD <- IOD[,4]
-IOD_ts <- ts(IOD,start=c(1870,1),frequency=12)
-IOD_30a <- window(IOD_ts,start=c(1979,1),end=c(2008,12))
-## deseasonalised IOD
-IOD_30 <- stl(IOD_30a,"period")$time.series[,3]+stl(IOD_30a,"period")$time.series[,2]
-# end of read data
-
-@
-
 \section{Statistical method}
@@
 In this study, up to two important indices from the seven climatic indicators (see section \ref{sec:Data}) were used as the explanatory variables in the model for each study region.
-
-<<rainData,eval=F , echo = F, results = hide>>=
-
-## 100 years awap data & 30 years
-qld_100 <- read.csv("Data\\rainfall\\QLD_month_rf.csv",header=T)
-qld_30 <- subset(qld_100,as.Date(qld_100[,1],"%d/%m/%Y")>="1979-01-01" & as.Date(qld_100[,1],"%d/%m/%Y")<="2008-12-01")
-qld100_mean <- ts(rowMeans(qld_100[,2:2801],na.rm=T,dims=1),start=c(1900,1),frequency=12)
-qld30_mean <- window(qld100_mean,start=c(1979,1),end=c(2008,12))
-
-nsw_100 <- read.csv("Data\\rainfall\\NSW_month_rf.csv",header=T)
-nsw_30 <- subset(nsw_100,as.Date(nsw_100[,1],"%d/%m/%Y")>="1979-01-01" & as.Date(nsw_100[,1],"%d/%m/%Y")<="2008-12-01")
-nsw100_mean <- ts(rowMeans(nsw_100[,2:1185],na.rm=T,dims=1),start=c(1900,1),frequency=12)
-nsw30_mean <- window(nsw100_mean, start=c(1979,1),end=c(2008,12))
-
-require(MASS)
-fitdistr(qld30_mean,"gamma")  ## assume shape=1
-fitdistr(nsw30_mean,"gamma")  ## assume shape =8
-ks.test(qld30_mean,1,0.03)
-ks.test(nsw30_mean,2.4,0.03)
-
-setEPS()
-postscript("hist_qld.eps")
- hist(qld30_mean,main="",xlab=expression(paste("rainfall (mm mont",h^-1,")",sep="")))
-dev.off()
-
-setEPS()
-postscript("hist_nswvic.eps")
- hist(nsw30_mean,main="",xlab=expression(paste("rainfall (mm mont",h^-1,")",sep="")))
-dev.off()
-
-## autocorrelation in rainfall
-setEPS()
-postscript("autocorrelation.eps")
-par(mfrow=c(2,2))
-  acf(qld30_mean,lag.max=12,main="QLD")
-  acf(nsw30_mean,lag.max=12,main="NSW & VIC")
-
-  pacf(qld30_mean,lag.max=12,main="")
-  pacf(nsw30_mean,lag.max=12,main="")  
-dev.off()
-
-## deseasonalise & detrend rainfall: assume the deseasonalised monthly zonal rainfall is normal distributed and can be used in pearson cross-correlation function
-## use stl function in R
-  qld30_deseason <- stl(qld30_mean,s.window=length(qld30_mean),t.window=length(qld30_mean))$time.series[,3]
-  nsw30_deseason <- stl(nsw30_mean,s.window=length(nsw30_mean),t.window=length(nsw30_mean))$time.series[,3]
-  qld100_deseason <- stl(qld100_mean,s.window=length(qld100_mean),t.window=length(qld100_mean))$time.series[,3]
-  nsw100_deseason <- stl(nsw100_mean,s.window=length(nsw100_mean),t.window=length(nsw100_mean))$time.series[,3]
-# 
-# setEPS()
-# postscript("qld_mean_ts.eps")
-# par(cex.lab=1.3,cex.axis=1.3)
-# plot(qld30_deseason,xlab="Year",ylab="monthly rainfall")
-# abline(v=2003,col="red",lwd=2)
-# text(2004.6,150,"2003",cex=1.5,col="red")
-# dev.off()
-# 
-# setEPS()
-# postscript("nswvic_mean_ts.eps")
-# par(cex.lab=1.3,cex.axis=1.3)
-# plot(nsw30_deseason,xlab="Year",ylab="monthly rainfall")
-# abline(v=2003,col="red",lwd=2)
-# text(2004.6,120,"2003",cex=1.5,col="red")
-# dev.off()
-#   
-# setEPS()
-# postscript("cor_qld.eps")
-# par(mfrow=c(2,3),mar=c(5,4,1,1))
-# ccf(as.vector(SOI_30),as.vector(qld30_deseason),lag.max=12,xlim=c(-12,0),xlab="SOI  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# ccf(as.vector(SST34_30),as.vector(qld30_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO3.4  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# ccf(as.vector(SST4_30),as.vector(qld30_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO4  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# ccf(as.vector(PDO_100),as.vector(qld100_deseason),lag.max=24,xlim=c(-24,0),xlab="PDO Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# ccf(as.vector(PDO_100*SOI_100),as.vector(qld100_deseason),lag.max=12,xlim=c(-12,0),xlab="PDO*SOI  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# ccf(as.vector(IOD_30),as.vector(qld30_deseason),lag.max=12,xlim=c(-12,0),xlab="IOD  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# dev.off()
-# 
-# # ccf(subset(SOI_30,cycle(SOI_30)==10|cycle(SOI_30)==11|cycle(SOI_30)==12),subset(qld_deseason,cycle(qld_deseason)==10|cycle(qld_deseason)==11|cycle(qld_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="SOI  Lag",ylab="CCF") ## SOI
-# # ccf(subset(SST34_30,cycle(SST34_30)==10|cycle(SST34_30)==11|cycle(SST34_30)==12),subset(qld_deseason,cycle(qld_deseason)==10|cycle(qld_deseason)==11|cycle(qld_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="SST34  Lag",ylab="CCF") ## SST34
-# # ccf(subset(SST3_30,cycle(SST3_30)==10|cycle(SST3_30)==11|cycle(SST3_30)==12),subset(qld_deseason,cycle(qld_deseason)==10|cycle(qld_deseason)==11|cycle(qld_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="SST3  Lag",ylab="CCF") ## SST3
-# # ccf(subset(PDO_30,cycle(PDO_30)==10|cycle(PDO_30)==11|cycle(PDO_30)==12),subset(qld_deseason,cycle(qld_deseason)==10|cycle(qld_deseason)==11|cycle(qld_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="PDO  Lag",ylab="CCF") ## PDO
-# # ccf(subset(PDO_30,cycle(PDO_30)==10|cycle(PDO_30)==11|cycle(PDO_30)==12)* subset(SOI_30,cycle(SOI_30)==10|cycle(SOI_30)==11|cycle(SOI_30)==12),subset(qld_deseason,cycle(qld_deseason)==10|cycle(qld_deseason)==11|cycle(qld_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="PDO  Lag",ylab="CCF") ## PDO
-# # ccf(subset(IOD_30,cycle(IOD_30)==10|cycle(IOD_30)==11|cycle(IOD_30)==12),subset(qld_deseason,cycle(qld_deseason)==10|cycle(qld_deseason)==11|cycle(qld_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="IOD  Lag",ylab="CCF") ## IOD
-# # 
-# # 
-# # 
-# # ## 1,2,3: SOI
-# # ## 4,5,6: SOI
-# # ## 7,8,9: SOI, IOD
-# # ## 10,11,12: SOI, SST34, SST3, PDO
-# # 
-# 
-# setEPS()
-# postscript("cor_nswvic.eps")
-# par(mfrow=c(2,3),mar=c(5,4,1,1))
-# ccf(as.vector(SOI_30),as.vector(nsw30_deseason),lag.max=12,xlim=c(-12,0),xlab="SOI  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# ccf(as.vector(SST34_30),as.vector(nsw30_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO3.4  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# ccf(as.vector(SST4_30),as.vector(nsw30_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO4  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# ccf(as.vector(PDO_100),as.vector(nsw100_deseason),lag.max=24,xlim=c(-24,0),xlab="PDO Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# ccf(as.vector(PDO_100*SOI_100),as.vector(nsw100_deseason),lag.max=12,xlim=c(-12,0),xlab="PDO*SOI  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# ccf(as.vector(IOD_30),as.vector(nsw30_deseason),lag.max=12,xlim=c(-12,0),xlab="IOD  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# dev.off()
-# 
-# ccf(subset(SOI_30,cycle(SOI_30)==7|cycle(SOI_30)==8|cycle(SOI_30)==9),subset(nswvic_deseason,cycle(nswvic_deseason)==7|cycle(nswvic_deseason)==8|cycle(nswvic_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="SOI  Lag",ylab="CCF") ## SOI
-# ccf(subset(SST34_30,cycle(SST34_30)==7|cycle(SST34_30)==8|cycle(SST34_30)==9),subset(nswvic_deseason,cycle(nswvic_deseason)==7|cycle(nswvic_deseason)==8|cycle(nswvic_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="SST34  Lag",ylab="CCF") ## SST34
-# ccf(subset(SST3_30,cycle(SST3_30)==7|cycle(SST3_30)==8|cycle(SST3_30)==9),subset(nswvic_deseason,cycle(nswvic_deseason)==7|cycle(nswvic_deseason)==8|cycle(nswvic_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="SST3  Lag",ylab="CCF") ## SST3
-# ccf(subset(PDO_30,cycle(PDO_30)==7|cycle(PDO_30)==8|cycle(PDO_30)==9),subset(nswvic_deseason,cycle(nswvic_deseason)==7|cycle(nswvic_deseason)==8|cycle(nswvic_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="PDO  Lag",ylab="CCF") ## PDO
-# ccf(subset(PDO_30,cycle(PDO_30)==7|cycle(PDO_30)==8|cycle(PDO_30)==9)* subset(SOI_30,cycle(SOI_30)==7|cycle(SOI_30)==8|cycle(SOI_30)==9),subset(nswvic_deseason,cycle(nswvic_deseason)==7|cycle(nswvic_deseason)==8|cycle(nswvic_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="PDO  Lag",ylab="CCF") ## PDO
-# ccf(subset(IOD_30,cycle(IOD_30)==7|cycle(IOD_30)==8|cycle(IOD_30)==9),subset(nswvic_deseason,cycle(nswvic_deseason)==7|cycle(nswvic_deseason)==8|cycle(nswvic_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="IOD  Lag",ylab="CCF") ## IOD
-# 
-# 
-# 
-# ## 1,2,3: NONE
-# #4,5,6: SOI
-# # 7, 8, 9: SOI, IOD
-# ## 10, 11, 12: SOI, SST34, SST3, IOD
-
-@
 
@@
 
-<<simpleMode,eval=F, echo = F, results = hide>>=
-
-## plotting a hypothesed land cover change model
-data1 <- data.frame(year=c(1,2,2,3),tc=c(2,2,1,1))
-data2 <- data.frame(year=c(1,2,2,3,4),tc=c(2,2,1,1.5,2))
-
-setEPS()
-postscript("tc_simple_qld.eps")
-par(mar=c(5,6,1,1))
-plot(data1$year,data1$tc,type="l",ylim=c(0.5,2.5),xlab="",ylab="",xaxt="n",yaxt="n")
-axis(1,at=c(1,2,3),labels=c("before","LCC","after"),cex.axis=1.3)
-axis(2,at=c(1,2),label=c("removed","tree"),cex.axis=1.3,las=2)
-dev.off()
-
-setEPS()
-postscript("tc_simple_nswvic.eps")
-par(mar=c(5,6,1,1))
-plot(data2$year,data2$tc,type="l",ylim=c(0.5,2.5),xlab="",ylab="",xaxt="n",yaxt="n")
-axis(1,at=c(1,2,4),labels=c("before","LCC","after"),cex.axis=1.3)
-axis(2,at=c(1,2),label=c("removed","tree"),cex.axis=1.3,las=2)
-dev.off()
-
-@
-
-<<fullRegression,eval=F , echo = F, results = hide>>=
-
-## predicting model: land cover is category variable
-rain.trend <- function(x,y,m,var.data) {
-
-  var.data$rain <- as.vector(unlist(y[x]))+0.0001
-
-  if (!is.na(mean(var.data$rain))) {
-    if (grepl("qld",substitute(y))) {
-      var.data <- na.omit(var.data)    
-      rain_fit <- gam(rain ~ s(month, k=3)
-                  + s(SOI, k=3)
-                  + cover  
-                  + trd
-                  , data=var.data,family=Gamma(link=log))
-    } else
-     if (grepl("nsw",substitute(y))) {  
-      var.data <- na.omit(var.data)    
-      rain_fit <- gam(rain ~ s(month, k=3)
-                  + s(SOI, k=3)
-                  + s(IOD, k=3)
-                  + cover  
-                  + trd
-                  , data=var.data,family=Gamma(link=log))
-    }
-
-  return(rain_fit)
-  } else {NA}
-
-}
-
-## note: link=log generally has higher r.sq value than link=reverse
-
-## generate variables for the model
-trd <- seq(1:360)
-month <-rep(1:12,times=30)
-mth <- cycle(ts(month,start=c(1979,1),frequency=12))
-
-LC <- c(rep("tree",288),rep("clear",72))
-# LC_qld1 <- c(rep("tree",294),rep("clear",66)) ## after-change: from Jun 2003
-# LC_qld2 <- c(rep("tree",300),rep("clear",60)) ## after- : from Jan 2004
-# LC_qld3 <- c(rep("tree",306),rep("clear",54)) ## after- : from Jun 2004
-LC_qld4 <- c(rep("tree",312),rep("clear",48)) ## after- : from Jan 2005
-
-var.data <- data.frame(month, trd, SOI = SOI_30a, IOD = IOD_30a, cover = LC)
-# var.qld1 <- data.frame(month, trd, SOI = SOI_30a, IOD = IOD_30a, cover = LC_qld1)
-# var.qld2 <- data.frame(month, trd, SOI = SOI_30a, IOD = IOD_30a, cover = LC_qld2)
-# var.qld3 <- data.frame(month, trd, SOI = SOI_30a, IOD = IOD_30a, cover = LC_qld3)
-var.qld4 <- data.frame(month, trd, SOI = SOI_30a, IOD = IOD_30a, cover = LC_qld4)
-
-## run model for 30 years
-#   coef.rain30.qld1 <- lapply(2:ncol(qld_30), function(x) rain.trend(x,qld_30,30,var.qld1))
-#   coef.rain30.qld2 <- lapply(2:ncol(qld_30), function(x) rain.trend(x,qld_30,30,var.qld2))
-#   coef.rain30.qld3 <- lapply(2:ncol(qld_30), function(x) rain.trend(x,qld_30,30,var.qld3))
-  coef.rain30.qld4 <- lapply(2:ncol(qld_30), function(x) rain.trend(x,qld_30,30,var.qld4))
-  coef.rain30.nswvic <- lapply(2:ncol(nsw_30), function(x) rain.trend(x,nsw_30,30,var.data))
-
-## extract sign of C
-  C30_qld <- unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$p.coeff[2],NA)))
-  C30_nswvic <- unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$p.coeff[2],NA)))
-
-## extract p value of C
-#   Cp30_qld1 <- unlist(lapply(1:length(coef.rain30.qld1),function(x) ifelse(summary(coef.rain30.qld1)[x,2]=="gam",summary(coef.rain30.qld1[[x]])$p.pv[2],NA)))
-# Cp30_qld2 <- unlist(lapply(1:length(coef.rain30.qld2),function(x) ifelse(summary(coef.rain30.qld2)[x,2]=="gam",summary(coef.rain30.qld2[[x]])$p.pv[2],NA)))
-# Cp30_qld3 <- unlist(lapply(1:length(coef.rain30.qld3),function(x) ifelse(summary(coef.rain30.qld3)[x,2]=="gam",summary(coef.rain30.qld3[[x]])$p.pv[2],NA)))
-Cp30_qld4 <- unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$p.pv[2],NA)))
-Cp30_nswvic <- unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$p.pv[2],NA)))
-
-## p value of seasons and climate indices
-p_qld <- unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$s.pv[1],NA)))
-p_nsw <- unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$s.pv[1],NA)))
-
-## extract r.sq of the model
-rsq_qld <- unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$r.sq,NA)))
-rsq_nsw <- unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$r.sq,NA)))
-
-## frame of the two regions (100 years data)
-qld_tif <- raster("Data\\BoM\\rainfall\\AWAP\\from1900_FVG\\r19000101_qld.tif")
-nsw_tif <- raster("Data\\BoM\\rainfall\\AWAP\\from1900_FVG\\r19000101_nsw.tif")
-
-  dist_Cp30_qld4 <- raster(qld_tif)
-  dist_Cp30_qld4[] <- Cp30_qld4
-#   dist_Cp30_qld <- crop(reproject(dist_Cp100_qld,CRS=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84"),program="raster"),e1)
-  dist_Cp30_nsw <- raster(nsw_tif)
-  dist_Cp30_nsw[] <- Cp30_nswvic
-#   dist_Cp30_nsw <- crop(reproject(dist_Cp30_nsw,CRS=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84"),program="raster"),e2)
-# 
-#   dist_C30_qld <- raster(reg_qld)
-#   dist_C30_qld[] <- C30_qld
-#   dist_C30_nswvic <- raster(reg_nswvic)
-#   dist_C30_nswvic[] <- C30_nswvic
-# 
-# ## range of predicted values
-  predictmax_qld30 <- unlist(lapply(1:length(coef.rain30.qld4), function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam", max(coef.rain30.qld4[[x]]$fitted.values),NA)))
-  predictmin_qld30 <- unlist(lapply(1:length(coef.rain30.qld4), function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam", min(coef.rain30.qld4[[x]]$fitted.values),NA)))
-# 
-mean_qld <- lapply(1:length(coef.rain30.qld4), function(x) if(summary(coef.rain30.qld4)[x,2]=="gam") {fitted(coef.rain30.qld4[[x]])} else{NA})
-
-## calculate mean residual
-  cc <- 1
-  qld.sum <- mean_qld[[1]]
-  for (i in 2:length(mean_qld)) {
-    if(!is.na(mean_qld[[i]])) {
-      qld.sum <- qld.sum + mean_qld[[i]]
-      cc <- cc+1
-    }
-  }
-  qld.mean <- qld.sum/cc
-  qld.mean <- ts(qld.mean,start=c(1979,1),frequency=12)
-
-predictmax_nsw30 <- unlist(lapply(1:length(coef.rain30.nswvic), function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam", max(coef.rain30.nswvic[[x]]$fitted.values),NA)))
-# 70 - 254.7
-  predictmin_nsw30 <- unlist(lapply(1:length(coef.rain30.nswvic), function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam", min(coef.rain30.nswvic[[x]]$fitted.values),NA)))
-# 18.4 - 62.2
-
-mean_nswvic <- lapply(1:length(coef.rain30.nswvic), function(x) if(summary(coef.rain30.nswvic)[x,2]=="gam") {fitted(coef.rain30.nswvic[[x]])} else{NA})
-
-## calculate mean residual
-  cc <- 1
-  nswvic.sum <- mean_nswvic[[1]]
-  for (i in 2:length(mean_nswvic)) {
-    if(!is.na(mean_nswvic[[i]])) {
-      nswvic.sum <- nswvic.sum + mean_nswvic[[i]]
-      cc <- cc+1
-    }
-  }
-  nswvic.mean <- nswvic.sum/cc
-  nswvic.mean <- ts(nswvic.mean,start=c(1979,1),frequency=12)
-
-## plots
-  setEPS()
-  postscript("Cp_30yrs_qld.eps")
-  par(mar=c(3,3,1,6),xpd=T)
-  plot(e111,xlab="",ylab="")
-  image(dist_Cp30_qld4,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
-#   plot(aus_reg2,border="grey",add=T)
-  plot(mdb_reg2,border="grey",lwd=3,add=T)
-  plot(e111,add=T)
-image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.7,legend.width=0.9,legend.args=list(text="p value",line=0.5,cex=1.2,font=1),axis.args=list(cex.axis=0.8))
-legend(148.7,-27.9,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=1)
-  dev.off()
-
-  setEPS()
-  postscript("Cp_30yrs_nswvic.eps")
-  par(mar=c(3,3,1,6),xpd=T)  
-  plot(e222,xlab="",ylab="") #,border="white")
-  image(dist_Cp30_nsw,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
-  plot(aus_reg2,border="grey",add=T)
-  plot(mdb_reg2,border="grey",lwd=3,add=T)
-  plot(e222,add=T)
-  #image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.8,legend.width=0.7,legend.args=list(text="p value",line=0.5,cex=1,font=2),axis.args=list(cex.axis=0.8))
-  dev.off()
-# 
-# ## R^2 of the regression
-#     rsq_qld30 <- unlist(lapply(1:length(coef.rain30.qld), function(x) ifelse(summary(coef.rain30.qld)[x,2]=="gam", summary(coef.rain30.qld[[x]])$r.sq,NA)))
-#     rsq_nswvic30 <- unlist(lapply(1:length(coef.rain30.nswvic), function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam", summary(coef.rain30.nswvic[[x]])$r.sq,NA)))
-# 
-# ## significant of the other two variables
-#     v1_qld30 <- unlist(lapply(1:length(coef.rain30.qld), function(x) ifelse(summary(coef.rain30.qld)[x,2]=="gam", summary(coef.rain30.qld[[x]])$s.pv[1],NA)))
-#     v1_nswvic30 <- unlist(lapply(1:length(coef.rain30.nswvic), function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam", summary(coef.rain30.nswvic[[x]])$s.pv[1],NA)))
-# 
-## plot gam.check
-  setEPS()
-  postscript("qld_check.eps")
-  gam.check(coef.rain30.qld4[[1]])
-  dev.off()
-
-  setEPS()
-  postscript("nswvic_check.eps")
-  gam.check(coef.rain30.nswvic[[which(values(dist_Cp30_nsw)<0.05)[6]]])
-  dev.off()
-
-@
-
 \subsection{Step trend test}
@@
 %chunk 5: note code has changed
-<<chunk5,eval=F , echo = F, results = hide>>=
-
-## without land cover variable for step rend test
-rain.trend3 <- function(x,y,m,var.data) {
-
-  var.data$rain <- as.vector(unlist(y[x]))+0.0001
-
-  if (!is.na(mean(var.data$rain))) {
-    if (grepl("qld",substitute(y))) {
-      rain_fit <- gam(rain ~ trd  # 
-                  + s(month,k=3)
-                  + s(SOI, k=3)
-                  , data=var.data,family=Gamma(link=log))
-    } else
-    if (grepl("nsw",substitute(y))) {
-      rain_fit <- gam(rain ~ trd
-                  + s(month,k=3)
-                  + s(SOI, k=3)
-                  + s(IOD, k=3) 
-                  , data=var.data,family=Gamma(link=log))
-    }
-
-  return(rain_fit)
-  } else {NA}
-
-}
-# 
-# ## change LC variable to be null
-#   var.data <- data.frame(month, trd, SOI = SOI_30a, IOD = IOD_30a)
-# 
-## run model for 30 years
-  coef.rain.qld <- lapply(2:ncol(qld_30), function(x) rain.trend3(x,qld_30,30,var.data))
-  coef.rain.nswvic <- lapply(2:ncol(nsw_30), function(x) rain.trend3(x,nsw_30,30,var.data))
-
-## explaining power of the partial model
-## QLD
-  rsq_qld <- unlist(lapply(1:length(coef.rain.qld), function(x) ifelse(summary(coef.rain.qld)[x,2]=="gam", summary(coef.rain.qld[[x]])$r.sq,NA)))
-# 
-#   predictmax_qld <- unlist(lapply(1:length(coef.rain.qld), function(x) ifelse(summary(coef.rain.qld)[x,2]=="gam", max(coef.rain.qld[[x]]$fitted.values),NA)))
-#   predictmin_qld <- unlist(lapply(1:length(coef.rain.qld), function(x) ifelse(summary(coef.rain.qld)[x,2]=="gam", min(coef.rain.qld[[x]]$fitted.values),NA)))
-# 
-## extract residual
-  residual_qld <- lapply(1:length(coef.rain.qld), function(x) if(summary(coef.rain.qld)[x,2]=="gam") {coef.rain.qld[[x]]$y-fitted(coef.rain.qld[[x]])} else {NA})
-
-# ## calculate mean residual
-#   cc <- 1
-#   qld.rsum <- residual_qld[[1]]
-#   for (i in 2:length(residual_qld)) {
-#     if(!is.na(residual_qld[[i]])) {
-#       qld.rsum <- qld.rsum + residual_qld[[i]]
-#       cc <- cc+1
-#     }
-#   }
-#   qld.rm <- qld.rsum/cc
-#   qld.rm <- ts(qld.rm,start=c(1979,1),frequency=12)
-# 
-# ## NSW/VIC
-# 
-  rsq_nswvic <- unlist(lapply(1:length(coef.rain.nswvic), function(x) ifelse(summary(coef.rain.nswvic)[x,2]=="gam", summary(coef.rain.nswvic[[x]])$r.sq,NA)))
-# 
-#   predictmax_nswvic <- unlist(lapply(1:length(coef.rain.nswvic), function(x) ifelse(summary(coef.rain.nswvic)[x,2]=="gam", max(coef.rain.nswvic[[x]]$fitted.values),NA)))
-#   predictmin_nswvic <- unlist(lapply(1:length(coef.rain.nswvic), function(x) ifelse(summary(coef.rain.nswvic)[x,2]=="gam", min(coef.rain.nswvic[[x]]$fitted.values),NA)))
-
-# extract residual
-residual_nswvic <- lapply(1:length(coef.rain.nswvic), function(x) if(summary(coef.rain.nswvic)[x,2]=="gam") {coef.rain.nswvic[[x]]$y-fitted(coef.rain.nswvic[[x]])} else {NA})
-
-# ## calculate mean residual
-# cc <- 1
-# nswvic.rsum <- residual_nswvic[[1]]
-# for (i in 2:length(residual_nswvic)) {
-#   if(!is.na(residual_nswvic[[i]])) {
-#     nswvic.rsum <- nswvic.rsum + residual_nswvic[[i]]
-#     cc <- cc+1
-#   }
-# }
-# 
-# nswvic.rm <- nswvic.rsum/cc
-# nswvic.rm <- ts(nswvic.rm,start=c(1979,1),frequency=12)
-# 
-# setEPS()
-# postscript("res_qld.eps")
-# hist(qld.rm,main="",xlab="rainfall residuals")
-# dev.off()
-# 
-# setEPS()
-# postscript("res_nswvic.eps")
-# hist(nswvic.rm,main="",xlab="rainfall residuals")
-# dev.off()
-# 
-# setEPS()
-# postscript("cor_res_nswvic.eps")
-# par(mfrow=c(1,2),mar=c(5,4,5,1))
-# ccf(SOI_30,nswvic.rm,lag.max=12,xlim=c(-1,0),xlab="SOI  Lag",main="",ylab="CCF")
-# ccf(IOD_30,nswvic.rm,lag.max=12,xlim=c(-1,0),xlab="IOD  Lag",main="",ylab="CCF")
-# dev.off()
-
-## explaining power of the partial model
-rain.trend4 <- function(x,y,m,var.data) {
-
-  var.data$rain <- as.vector(unlist(y[x]))+0.0001
-
-  if (!is.na(mean(var.data$rain))) {
-    if (grepl("qld",substitute(y))) {
-      rain_fit <- gam(rain ~ s(SOI, k=3)
-                  , data=var.data,family=Gamma(link=log))
-    } else
-    if (grepl("nsw",substitute(y))) {
-      rain_fit <- gam(rain ~ s(SOI, k=3)
-                  + s(IOD, k=3) 
-                  , data=var.data,family=Gamma(link=log))
-    }
-
-  return(rain_fit)
-  } else {NA}
-
-}
-
-## explaining power of index
-  coef.soi.qld <- lapply(2:ncol(qld_30), function(x) rain.trend4(x,qld_30,30,var.data))
-  coef.soi.nswvic <- lapply(2:ncol(nsw_30), function(x) rain.trend4(x,nsw_30,30,var.data))
-  rsq_soi_qld <- unlist(lapply(1:length(coef.soi.qld), function(x) ifelse(summary(coef.soi.qld)[x,2]=="gam", summary(coef.soi.qld[[x]])$r.sq,NA)))
-  rsq_soi_nswvic <- unlist(lapply(1:length(coef.soi.nswvic), function(x) ifelse(summary(coef.soi.nswvic)[x,2]=="gam", summary(coef.soi.nswvic[[x]])$r.sq,NA)))
-
-setEPS()
-postscript("soi_qld.eps")
- hist(rsq_soi_qld*100,main="",xlab=expression(paste("% of variability explained (",R^2,")")),cex.lab=1.2)
- dev.off()
-
- setEPS()
-postscript("soi_nswvic.eps")
- hist(rsq_soi_nswvic*100,main="",xlab=expression(paste("% of variability explained (",R^2,")")),cex.lab=1.2)
- dev.off()
-
-@
 
@@
 %chunk 6
-<<chunk6,eval=F , echo = F, results = hide>>=
-
-## perform step test on the data
-## function to calculate the step trend statistic (Hirsch, 1985)
-stepTrend2 <- function(x,y,i){
-
-    n1 <- 2002-1997
-
-    if (i==1){
-
-      z <- 56  # no. of column
-
-    } else {
-
-    z <- 37 }
-
-if (x%%z!=0){ #& x%%z!=(z-1)) {
-    y.1 <- y[[x]]
-    y.2 <- y[[x+1]]
-#     y.3 <- y[[x+2]]
-    y.3 <- y[[x+z]]
-    y.4 <- y[[x+z+1]]
-#     y.6 <- y[[x+z+2]]
-#     y.7 <- y[[x+2*z]]
-#     y.8 <- y[[x+2*z+1]]
-#     y.9 <- y[[x+2*z+2]]
-
-    if(!is.na(y.1[1])&!is.na(y.2[1])&!is.na(y.3[1])&!is.na(y.4[1])) { #|!is.na(y.5[1])|!is.na(y.6[1])|!is.na(y.7[1])|!is.na(y.8[1])|!is.na(y.9[1])) {
-
-    if (i == 1) {
-      n2 <- 2008-2004    
-      v.1 <- ts(y.1[c(229:288,313:360)],start=c(1998,1),frequency=12)
-      v.2 <- ts(y.2[c(229:288,313:360)],start=c(1998,1),frequency=12)
-      v.3 <- ts(y.3[c(229:288,313:360)],start=c(1998,1),frequency=12)
-      v.4 <- ts(y.4[c(229:288,313:360)],start=c(1998,1),frequency=12)
-#       v.5 <- ts(y.5[c(229:288,313:348)],start=c(1998,1),frequency=12)
-#       v.6 <- ts(y.6[c(229:288,313:348)],start=c(1998,1),frequency=12)
-#       v.7 <- ts(y.7[c(229:288,313:348)],start=c(1998,1),frequency=12)
-#       v.8 <- ts(y.8[c(229:288,313:348)],start=c(1998,1),frequency=12)
-#       v.9 <- ts(y.9[c(229:288,313:348)],start=c(1998,1),frequency=12)
-#       v <- (x$y-x$fitted.values)[c(1:288,313:357)]
-#       v.s <- ts(v,start=c(1998,1),frequency=12)
-
-    } else {
-
-      n2 <- 2008-2002
-
-      v.1 <- ts(y.1[c(229:360)],start=c(1998,1),frequency=12)
-      v.2 <- ts(y.2[c(229:360)],start=c(1998,1),frequency=12)
-      v.3 <- ts(y.3[c(229:360)],start=c(1998,1),frequency=12)
-      v.4 <- ts(y.4[c(229:360)],start=c(1998,1),frequency=12)
-#       v.5 <- ts(y.5[c(229:348)],start=c(1998,1),frequency=12)
-#       v.6 <- ts(y.6[c(229:348)],start=c(1998,1),frequency=12)
-#       v.7 <- ts(y.7[c(229:348)],start=c(1998,1),frequency=12)
-#       v.8 <- ts(y.8[c(229:348)],start=c(1998,1),frequency=12)
-#       v.9 <- ts(y.9[c(229:348)],start=c(1998,1),frequency=12)      
-#       v <- (x$y-x$fitted.values)[c(1:288, 301:357)]
-#       v.s <- ts(v,start=c(1998,1),frequency=12)    
-}
-
-## rank by month
-   for (j in 1:12) {    
-    assign(paste("r1",j,sep=""),rank(subset(v.1,cycle(v.1)==j)))
-    assign(paste("r2",j,sep=""),rank(subset(v.2,cycle(v.2)==j)))
-    assign(paste("r3",j,sep=""),rank(subset(v.3,cycle(v.3)==j)))
-    assign(paste("r4",j,sep=""),rank(subset(v.4,cycle(v.4)==j))) 
-#     assign(paste("r5",j,sep=""),rank(subset(v.5,cycle(v.5)==j)))
-#     assign(paste("r6",j,sep=""),rank(subset(v.6,cycle(v.6)==j)))
-#     assign(paste("r7",j,sep=""),rank(subset(v.7,cycle(v.7)==j)))
-#     assign(paste("r8",j,sep=""),rank(subset(v.8,cycle(v.8)==j)))
-#     assign(paste("r9",j,sep=""),rank(subset(v.9,cycle(v.9)==j)))
-
-    }
-
-## mann-whitney rank sum statistic
-    rlist.1 <- list(r11,r12,r13,r14,r15,r16,r17,r18,r19,r110,r111,r112) ## full year
-    rlist.2 <- list(r21,r22,r23,r24,r25,r26,r27,r28,r29,r210,r211,r212)
-    rlist.3 <- list(r31,r32,r33,r34,r35,r36,r37,r38,r39,r310,r311,r312)
-    rlist.4 <- list(r41,r42,r43,r44,r45,r46,r47,r48,r49,r410,r411,r412)
-#     rlist.5 <- list(r51,r52,r53,r54,r55,r56,r57,r58,r59,r510,r511,r512)
-#     rlist.6 <- list(r61,r62,r63,r64,r65,r66,r67,r68,r69,r610,r611,r612)    
-#     rlist.7 <- list(r71,r72,r73,r74,r75,r76,r77,r78,r79,r710,r711,r712)    
-#     rlist.8 <- list(r81,r82,r83,r84,r85,r86,r87,r88,r89,r810,r811,r812)
-#     rlist.9 <- list(r91,r92,r93,r94,r95,r96,r97,r98,r99,r910,r911,r912)    
-    rlist <- list(unlist(rlist.1),unlist(rlist.2),unlist(rlist.3),unlist(rlist.4)) #,unlist(rlist.5),unlist(rlist.6),unlist(rlist.7),unlist(rlist.8),unlist(rlist.9))
-## seasonal
-#     rlist <- list(unlist(rlist.1[c(9:11)]),unlist(rlist.2[c(9:11)]),unlist(rlist.3[c(9:11)]),unlist(rlist.4[c(9:11)]))
-#     rlist <- list(r1,r2,r12) ## summer
-#     rlist <- list(r3,r4,r5)  ## autumn
-#     rlist <- list(r6,r7,r8)  ## winter
-#     rlist <- list(r9,r10,r11) ## spring
-    w.s1 <- 0
-    w.s2 <- 0
-    w.s3 <- 0
-    w.s4 <- 0
-#     w.s5 <- 0
-#     w.s6 <- 0
-#     w.s7 <- 0
-#     w.s8 <- 0
-#     w.s9 <- 0
-  for (k in 1:length(rlist.1)) {
-#   for (k in c(9:11)) {
-    w.s1 <- sum(rlist.1[[k]][1:n1])+w.s1
-    w.s2 <- sum(rlist.2[[k]][1:n1])+w.s2
-    w.s3 <- sum(rlist.3[[k]][1:n1])+w.s3
-    w.s4 <- sum(rlist.4[[k]][1:n1])+w.s4
-#     w.s5 <- sum(rlist.5[[k]][1:n1])+w.s5
-#     w.s6 <- sum(rlist.6[[k]][1:n1])+w.s6
-#     w.s7 <- sum(rlist.7[[k]][1:n1])+w.s7
-#     w.s8 <- sum(rlist.8[[k]][1:n1])+w.s8
-#     w.s9 <- sum(rlist.9[[k]][1:n1])+w.s9
-    }
-    w.s <- sum(w.s1,w.s2,w.s3,w.s4) #,w.s5,w.s6,w.s7,w.s8,w.s9)
-
-## other statistic
-# mean for full year
-    mu <- n1*(n1+n2+1)/2
-    mu.s <- 12*4*mu
-# ## mean for one season
-#     mu <- n1*(n1+n2+1)/2
-#     mu.s <- 3*4*mu
-
-## variance (standard variation).
-The autocorrelation of the residuals can be assumed to be null (checked, cor.test(), true for NSW, mostly true for WA.
-QLD?
-there are some situations in WA require further investigation)
-# assume only cross correlation
-    ss <- n1*n2*(n1+n2+1)/12
-    ss.s <- 0
-    for (kk in 1:4) {
-      for (kkk in 1:4) {
-#         for (kkkk in 1:12) {
-          ss.s <- 12*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
-#         }
-      }
-    }
-## variance for one season   
-#     ss <- n1*n2*(n1+n2+1)/3
-#     ss.s <- 0
-#     for (kk in 1:4) {
-#       for (kkk in 1:4) {
-#           ss.s <- 3*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
-#       }
-#     }
-
-## step trend statistic
-    z <- (w.s - mu.s)/sqrt(ss.s)
-
-    return(z) }
-
-     else {return(NA)} }
-
-     else {return(NA)}
-
-  }
-
-## 30 years
-## full year, 
-## central QLD
-  qld.z.new <- unlist(lapply(1:(length(residual_qld)-56),function(x) stepTrend2(x,residual_qld,1)))
-  qld.z.new <- c(qld.z.new,rep(NA,56))
-
-  tr_qld.step.new <- raster(qld_tif)
-  tr_qld.step.new[] <- qld.z.new
-## NSW  
-  nswvic.z.new <- unlist(lapply(1:(length(residual_nswvic)-37),function(x) stepTrend2(x,residual_nswvic,2)))
-  nswvic.z.new <- c(nswvic.z.new,rep(NA,37))
-
-  tr_nswvic.step.new <- raster(nsw_tif)
-  tr_nswvic.step.new[] <- nswvic.z.new
-
-# ## seasonal
-#   qld.z.sum <- unlist(lapply(1:(length(residual_qld)-56),function(x) stepTrend2(x,residual_qld,1)))
-#   qld.z.sum <- c(qld.z.sum,rep(NA,56))
-#   
-#   tr_qld.step.sum <- raster(qld_tif)
-#   tr_qld.step.sum[] <- qld.z.sum
-#   
-#   nswvic.z.sum <- unlist(lapply(1:(length(residual_nswvic)-37),function(x) stepTrend2(x,residual_nswvic,2)))
-#   nswvic.z.sum <- c(nswvic.z.sum,rep(NA,37))
-#   
-#   tr_nswvic.step.sum <- raster(nsw_tif)
-#   tr_nswvic.step.sum[] <- nswvic.z.sum
-
-  plotclt <- rev(brewer.pal(8,"RdYlBu"))
-
-setEPS()
-postscript("step_new_qld.eps")
-  par(mar=c(3,3,1,7),xpd=T)
-  plot(e111,xlab="",ylab="")#,border="white")
-  image(tr_qld.step.new,col=plotclt,breaks=c(-3,-1.645,-1.282,-0.842,0,0.842,1.282,1.645,3),legend=F,add=T)
-  plot(mdb_reg2,border="grey",add=T,lwd=3) 
-  plot(e111,add=T)
-image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,0,1,2,3,4),lab.breaks=c(0.00,0.05,0.1,0.2,1,0.2,0.1,0.05,0.00),legend.only=T,horizontal=F,col=plotclt,legend.shrink=0.7,legend.width=0.9,legend.args=list(text=c(expression(paste("level of significance (",alpha,")")),"positive Z'","negative Z'"),side=c(2,3,1),line=0.5,cex=1,font=2),axis.args=list(cex.axis=0.8))
-  dev.off()
-
-setEPS()
-  postscript("step_new_nswvic.eps")
-  par(mar=c(3,3,1,7),xpd=T)
-  plot(e222,xlab="",ylab="")#,border="white")
-  image(tr_nswvic.step.new,col=plotclt,breaks=c(-2.3,-1.645,-1.282,-0.842,0,0.842,1.282,1.645,3),legend=F,add=T)
-  plot(mdb_reg2,border="grey",add=T,lwd=3)
-  plot(aus_reg2,border="grey",add=T) 
-  plot(e222,add=T)
-image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,0,1,2,3,4),lab.breaks=c(0.00,0.05,0.1,0.2,1,0.2,0.1,0.05,0.00),legend.only=T,horizontal=F,col=plotclt,legend.shrink=0.7,legend.width=0.9,legend.args=list(text=c(expression(paste("level of significance (",alpha,")")),"positive Z'","negative Z'"),side=c(2,3,1),line=0.5,cex=1,font=2),axis.args=list(cex.axis=0.8))
-legend(149,-36.7,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=0.9)
-  dev.off()
-
-@
 
 % chunk 7: bootstrapping (do not keep spatial structure, not in use)
-<<chunk7,eval=F , echo = F, results = hide>>=
-
-## bootstrapping
-# ## the % of pixels with significance level < 0.05 in the original data set
-
-qld_per_origp <- length(subset(qld.z.new,qld.z.new>=1.645))/length(qld.z.new)
-# 0.0125
-qld_per_orign <- length(subset(qld.z.new,qld.z.new<=-1.645))/length(qld.z.new)
-# 0
-# 
-nsw_per_origp <- length(subset(nswvic.z.new,nswvic.z.new>=1.645))/length(nswvic.z.new)
-# 0.015
-nsw_per_orign <- length(subset(nswvic.z.new,nswvic.z.new<=-1.645))/length(nswvic.z.new)
-# 0
-
-# new step trend test: do not keep spatial structure
-# NstepTrend <- function(x){  
-#   
-#   n1 <- 2002-1997
-#     
-#   n2 <- 2008-2002
-# 
-#   if(!is.na(x[1])) {
-#     
-#   v <- x[c(229:357)]
-#     
-#   y<-ts(v,start=c(1998,1),frequency=12)
-#   
-#   for(i in 1:12) { 
-#   y.mth <- sample(window(y,start=c(1998,i),frequency=1),replace=F)
-#   
-#   ## rank by month   
-#     assign(paste("r",i,sep=""),rank(y.mth))
-#   } 
-#   
-#   
-# ## mann-whitney rank sum statistic
-#     rlist <- list(r1,r2,r3,r4,r5,r6,r7,r8,r9,r10,r11,r12) ## full year
-# 
-#     w.s <- 0
-#     for (k in 1:length(rlist)) {
-#     w.s <- sum(rlist[[k]][1:n1])+w.s
-#     }
-#   
-# ## other statistic
-# ## mean for full year
-#     mu <- n1*(n1+n2+1)/2
-#     mu.s <- length(rlist)*mu
-# 
-#   
-# ## variance (standard variation).
-The autocorrelation of the residuals can be assumed to be null (checked, true for QLD and NSW, mostly true for WA.
-there are some situations in WA require further investigation)
-#     ss <- n1*n2*(n1+n2+1)/length(rlist)
-#     ss.s <- length(rlist)*ss
-# 
-# ## step trend statistic
-#     z <- (w.s - mu.s)/sqrt(ss.s)
-#   
-#     return(z) 
-#     }
-#     
-#      else {return(NA)}
-#   
-#   }
-# 
-# ## conduct bootstrapping
-# R <- 1000
-# zp.values = numeric(R)
-# zn.values = numeric(R)
-# 
-# ptm <- proc.time()
-# for (i in 1:R) {
-#   temp <- unlist(lapply(residual_nswvic,function(x) NstepTrend(x)))
-#   if(i==1) {write.table(temp,file="R:\\AVSproject\\WrittenWork\\PhD_thesis\\chap3\\bt_zvalues.txt",row.names=F)
-#             } else
-#               {write.table(temp,file="R:\\AVSproject\\WrittenWork\\PhD_thesis\\chap3\\bt_zvalues.txt",append=T,row.names=F,col.names=F)}
-#    zp.values[i] <- length(subset(temp,temp>=1.28))/length(temp)
-#   zn.values[i] <- length(subset(temp,temp<=-1.28))/length(temp)
-#   
-# }
-# proc.time()-ptm
-# 
-# write.table(zp.values,file="R:\\AVSproject\\WrittenWork\\PhD_thesis\\chap3\\bt_outp.txt",row.names=F)
-# write.table(zn.values,file="R:\\AVSproject\\WrittenWork\\PhD_thesis\\chap3\\bt_outn.txt",row.names=F)
-@
 
 % chunk 8: bootstrapping (spatial structure is kept, in used)
-<<chunk8,eval=F , echo = F, results = hide>>=
-
-## one cell step trend
-## perform step test on the data for the bootstrapping test (spatial structure was kept)
-## function to calculate the step trend statistic (Hirsch, 1985)
-stepTrend <- function(x,i){
-
-    n1 <- 2002-1997
-
-  if(!is.na(x[1])) {
-
-    if(i==1) {
-      n2 <- 2008-2004    
-      v <- x[c(229:288,313:360)]
-#       v <- (x$y-x$fitted.values)[c(1:288,313:357)]
-      v.s <- ts(v,start=c(1998,1),frequency=12)    
-    } else
-      if(i==2) {
-      n2 <- 2008-2002
-      v <- x[c(229:360)]
-#       v <- (x$y-x$fitted.values)[c(1:288, 301:357)]
-      v.s <- ts(v,start=c(1998,1),frequency=12)      
-    } #else 
-#       {n2 <- 2008-2002
-#       v <- (x$y-x$fitted.values)[229:357]
-# #       v <- (x$y-x$fitted.values)
-#       v.s <- ts(v,start=c(1998,1),frequency=12)}
-
-## rank by month
-   for (j in 1:12) {    
-    assign(paste("r",j,sep=""),rank(subset(v.s,cycle(v.s)==j)))
-    }
-
-## mann-whitney rank sum statistic
-    rlist <- list(r1,r2,r3,r4,r5,r6,r7,r8,r9,r10,r11,r12) ## full year
-#     rlist <- list(r1,r2,r12) ## summer
-#     rlist <- list(r3,r4,r5)  ## autumn
-#     rlist <- list(r6,r7,r8)  ## winter
-#     rlist <- list(r9,r10,r11) ## spring
-    w.s <- 0
-    for (k in 1:length(rlist)) {
-    w.s <- sum(rlist[[k]][1:n1])+w.s
-    }
-
-## other statistic
-## mean for full year
-    mu <- n1*(n1+n2+1)/2
-    mu.s <- length(rlist)*mu
-## mean for one season
-#     mu <- n1*(n1+n2+1)/2
-#     mu.s <- 3*mu
-
-## variance (standard variation).
-The autocorrelation of the residuals can be assumed to be null (checked, cor.test(), true for NSW, mostly true for WA.
-QLD?
-there are some situations in WA require further investigation)
-## assume only cross correlation
-    ss <- n1*n2*(n1+n2+1)/length(rlist)
-    ss.s <- length(rlist)*ss
-## variance for one season   
-#     ss <- n1*n2*(n1+n2+1)/3
-#     ss.s <- 3*ss
-
-## step trend statistic
-    z <- (w.s - mu.s)/sqrt(ss.s)
-
-    return(z) }
-
-     else {return(NA)}
-
-  }
-
-## full year
-## central QLD
-  qld.z <- unlist(lapply(residual_qld,function(x) stepTrend(x,1)))
-## NSW/VIC
-## treatment 1: excluding 2003
-  nswvic.z <- unlist(lapply(residual_nswvic,function(x) stepTrend(x,2)))
-
-  qld_origp <- length(subset(qld.z,qld.z>=1.645))/length(qld.z)
-# 0.01
-  qld_orign <- length(subset(qld.z,qld.z<=-1.645))/length(qld.z)
-# 0
-
-  nsw_origp <- length(subset(nswvic.z,nswvic.z>=1.645))/length(nswvic.z)
-# 0.022
-  nsw_orign <- length(subset(nswvic.z,nswvic.z<=-1.645))/length(nswvic.z)
-
-# new step trend test 2: keep spatial structure
-NstepTrend2 <- function(x,y,ii){
-
-  n1 <- 2002-1997
-
-  if(ii==1) {
-      n2 <- 2008-2004    
-      v <- x[c(229:288,313:360)]    
-    } else
-      if(ii==2) {
-      n2 <- 2008-2002
-      v <- x[c(229:360)]
-    }
-
-  if(!is.na(unlist(x)[1])) {
-
-  yy<-ts(v,start=c(1998,1),frequency=12)
-
-  for(i in 1:12) { 
-  y.mth <- data.frame(mth=as.vector(window(yy,start=c(1998,i),frequency=1)),o=y[[i]])[order(y[[i]]),]
-
-  ## rank by month   
-    assign(paste("r",i,sep=""),rank(y.mth$mth) )
-  }
-
-## mann-whitney rank sum statistic
-    rlist <- list(r1,r2,r3,r4,r5,r6,r7,r8,r9,r10,r11,r12) ## full year
-
-    w.s <- 0
-    for (k in 1:length(rlist)) {
-    w.s <- sum(rlist[[k]][1:n1])+w.s
-    }
-
-## other statistic
-## mean for full year
-    mu <- n1*(n1+n2+1)/2
-    mu.s <- length(rlist)*mu
-
-## variance (standard variation). The autocorrelation of the residuals can be assumed to be null (checked, true for QLD and NSW)
-    ss <- n1*n2*(n1+n2+1)/length(rlist)
-    ss.s <- length(rlist)*ss
-
-## step trend statistic
-    z <- (w.s - mu.s)/sqrt(ss.s)
-
-    return(z) 
-    }
-
-     else {return(NA)}
-
-  }
-
-## conduct bootstrapping 2
-R <- 1000
-z2pq.values = numeric(R)
-z2nq.values = numeric(R)
-## sample years without replacement. note that with this method, only the order of years has changed, but months within year do not change
-new.order <- list()
-# mth.l <- rep(11,12)
-mth.l <- rep(9,12)
-
-ptm <- proc.time()
-for (i in 1:R) {
-
-  for (j in 1:12) {
-    new.order[[j]]<-sample(1:mth.l[j])
-  }
-
-  temp <- unlist(lapply(residual_qld,function(x,y) NstepTrend2(x,y,1),new.order))
-  if(i==1) {write.table(temp,file="bt_zvalues2_qld.txt",row.names=F)
-            } else
-              {write.table(temp,file="bt_zvalues2_qld.txt",append=T,row.names=F,col.names=F)}
-  z2pq.values[i] <- length(subset(temp,temp>=1.645))/length(temp)
-  z2nq.values[i] <- length(subset(temp,temp<=-1.645))/length(temp)
-}
-proc.time()-ptm
-
-write.table(z2pq.values,file="bt_out2p_qld.txt",row.names=F)
-write.table(z2nq.values,file="bt_out2n_qld.txt",row.names=F)
-
-# zp.values <- unlist(read.table("B:/My Study\\AVSproject\\WrittenWork\\PhD_thesis\\chap3\\bt_outp.txt",header=T,sep=""))*100
-# zn.values <- unlist(read.table("B:/My Study\\AVSproject\\WrittenWork\\PhD_thesis\\chap3\\bt_outn.txt",header=T,sep=""))*100
-# 
-bt1_nswp.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_out2p_nsw.txt",header=T,sep=""))*100
-bt1_nswn.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_out2n_nsw.txt",header=T,sep=""))*100
-
-bt1_qldp.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_out2p_qld.txt",header=T,sep=""))*100
-bt1_qldn.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_out2n_qld.txt",header=T,sep=""))*100
-# # 
-# # setEPS()
-# # postscript("boot1.eps")
-# # hist(zp.values,xlim=c(0,9),xlab="% positive z",main="no spatial dependency")
-# # points(per_origp*100,0,pch=16)
-# # dev.off()
-# # 
-# # setEPS()
-# # postscript("boot2.eps")
-# # hist(zn.values,xlim=c(0,25),xlab="% negative z",main="no spatial dependency")
-# # points(per_orign*100,0,pch=16)
-# # dev.off()
-# # # 
-# setEPS()
-# postscript("bt1_qldp.eps")
-# hist(bt1_qldp.values,nclass=100,xlab="% positive Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
-# points(qld_origp*100,0,pch=16,col="red",cex=1.8)
-# dev.off()
-# 
-# setEPS()
-# postscript("bt1_qldn.eps")
-# hist(bt1_qldn.values,nclass=100,xlab="% negative Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
-# points(qld_orign*100,0,pch=16,col="red",cex=1.8)
-# dev.off()
-# 
-# setEPS()
-# postscript("bt1_nswp.eps")
-# hist(bt1_nswp.values,nclass=100,xlab="% positive Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
-# points(nsw_origp*100,0,pch=16,col="red",cex=1.8)
-# dev.off()
-# 
-# setEPS()
-# postscript("bt1_nswn.eps")
-# hist(bt1_nswn.values,nclass=100,xlab="% negative Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
-# points(nsw_orign*100,0,pch=16,col="red",cex=1.8)
-# dev.off()
-# 
-# ## multiple cells
-# NstepTrend <- function(x,y,m,i){  
-# 
-#     n1 <- 2002-1997
-#     
-#     if (i==1){
-#       
-#       z <- 56  # no. of column
-#       
-#     } else {
-#     
-#     z <- 37 }
-#     
-# if (x%%z!=0){ #& x%%z!=(z-1)) {
-#     y.1 <- m[[x]]
-#     y.2 <- m[[x+1]]
-# #     y.3 <- y[[x+2]]
-#     y.3 <- m[[x+z]]
-#     y.4 <- m[[x+z+1]]
-# #     y.6 <- y[[x+z+2]]
-# #     y.7 <- y[[x+2*z]]
-# #     y.8 <- y[[x+2*z+1]]
-# #     y.9 <- y[[x+2*z+2]]
-#     
-#     if(!is.na(y.1[1])&!is.na(y.2[1])&!is.na(y.3[1])&!is.na(y.4[1])) { #|!is.na(y.5[1])|!is.na(y.6[1])|!is.na(y.7[1])|!is.na(y.8[1])|!is.na(y.9[1])) {
-#         
-#     if (i == 1) {
-#       n2 <- 2008-2004    
-#       v.1 <- ts(y.1[c(229:288,313:360)],start=c(1998,1),frequency=12)
-#       v.2 <- ts(y.2[c(229:288,313:360)],start=c(1998,1),frequency=12)
-#       v.3 <- ts(y.3[c(229:288,313:360)],start=c(1998,1),frequency=12)
-#       v.4 <- ts(y.4[c(229:288,313:360)],start=c(1998,1),frequency=12)
-# #       v.5 <- ts(y.5[c(229:288,313:348)],start=c(1998,1),frequency=12)
-# #       v.6 <- ts(y.6[c(229:288,313:348)],start=c(1998,1),frequency=12)
-# #       v.7 <- ts(y.7[c(229:288,313:348)],start=c(1998,1),frequency=12)
-# #       v.8 <- ts(y.8[c(229:288,313:348)],start=c(1998,1),frequency=12)
-# #       v.9 <- ts(y.9[c(229:288,313:348)],start=c(1998,1),frequency=12)
-# #       v <- (x$y-x$fitted.values)[c(1:288,313:357)]
-# #       v.s <- ts(v,start=c(1998,1),frequency=12)
-#       
-#     } else {
-#     
-#       n2 <- 2008-2002
-#   
-#       v.1 <- ts(y.1[c(229:360)],start=c(1998,1),frequency=12)
-#       v.2 <- ts(y.2[c(229:360)],start=c(1998,1),frequency=12)
-#       v.3 <- ts(y.3[c(229:360)],start=c(1998,1),frequency=12)
-#       v.4 <- ts(y.4[c(229:360)],start=c(1998,1),frequency=12)
-# #       v.5 <- ts(y.5[c(229:348)],start=c(1998,1),frequency=12)
-# #       v.6 <- ts(y.6[c(229:348)],start=c(1998,1),frequency=12)
-# #       v.7 <- ts(y.7[c(229:348)],start=c(1998,1),frequency=12)
-# #       v.8 <- ts(y.8[c(229:348)],start=c(1998,1),frequency=12)
-# #       v.9 <- ts(y.9[c(229:348)],start=c(1998,1),frequency=12)      
-# #       v <- (x$y-x$fitted.values)[c(1:288, 301:357)]
-# #       v.s <- ts(v,start=c(1998,1),frequency=12)    
-# }
-#   
-# ## rank by month
-#    for (j in 1:12) {  
-#     y1.mth <- data.frame(mth=as.vector(window(v.1,start=c(1998,i),frequency=1)),o=y[[j]])[order(y[[j]]),]
-#     y2.mth <- data.frame(mth=as.vector(window(v.2,start=c(1998,i),frequency=1)),o=y[[j]])[order(y[[j]]),]
-#     y3.mth <- data.frame(mth=as.vector(window(v.3,start=c(1998,i),frequency=1)),o=y[[j]])[order(y[[j]]),]
-#     y4.mth <- data.frame(mth=as.vector(window(v.4,start=c(1998,i),frequency=1)),o=y[[j]])[order(y[[j]]),]
-#     
-#     assign(paste("r1",j,sep=""),rank(y1.mth$mth))
-#     assign(paste("r2",j,sep=""),rank(y2.mth$mth))
-#     assign(paste("r3",j,sep=""),rank(y3.mth$mth))
-#     assign(paste("r4",j,sep=""),rank(y4.mth$mth)) 
-# #     assign(paste("r5",j,sep=""),rank(subset(v.5,cycle(v.5)==j)))
-# #     assign(paste("r6",j,sep=""),rank(subset(v.6,cycle(v.6)==j)))
-# #     assign(paste("r7",j,sep=""),rank(subset(v.7,cycle(v.7)==j)))
-# #     assign(paste("r8",j,sep=""),rank(subset(v.8,cycle(v.8)==j)))
-# #     assign(paste("r9",j,sep=""),rank(subset(v.9,cycle(v.9)==j)))
-#     
-#     }
-#   
-# ## mann-whitney rank sum statistic
-#     rlist.1 <- list(r11,r12,r13,r14,r15,r16,r17,r18,r19,r110,r111,r112) ## full year
-#     rlist.2 <- list(r21,r22,r23,r24,r25,r26,r27,r28,r29,r210,r211,r212)
-#     rlist.3 <- list(r31,r32,r33,r34,r35,r36,r37,r38,r39,r310,r311,r312)
-#     rlist.4 <- list(r41,r42,r43,r44,r45,r46,r47,r48,r49,r410,r411,r412)
-# #     rlist.5 <- list(r51,r52,r53,r54,r55,r56,r57,r58,r59,r510,r511,r512)
-# #     rlist.6 <- list(r61,r62,r63,r64,r65,r66,r67,r68,r69,r610,r611,r612)    
-# #     rlist.7 <- list(r71,r72,r73,r74,r75,r76,r77,r78,r79,r710,r711,r712)    
-# #     rlist.8 <- list(r81,r82,r83,r84,r85,r86,r87,r88,r89,r810,r811,r812)
-# #     rlist.9 <- list(r91,r92,r93,r94,r95,r96,r97,r98,r99,r910,r911,r912)    
-#     rlist <- list(unlist(rlist.1),unlist(rlist.2),unlist(rlist.3),unlist(rlist.4)) #,unlist(rlist.5),unlist(rlist.6),unlist(rlist.7),unlist(rlist.8),unlist(rlist.9))
-# ## seasonal
-# #     rlist <- list(unlist(rlist.1[c(9:11)]),unlist(rlist.2[c(9:11)]),unlist(rlist.3[c(9:11)]),unlist(rlist.4[c(9:11)]))
-# #     rlist <- list(r1,r2,r12) ## summer
-# #     rlist <- list(r3,r4,r5)  ## autumn
-# #     rlist <- list(r6,r7,r8)  ## winter
-# #     rlist <- list(r9,r10,r11) ## spring
-#     w.s1 <- 0
-#     w.s2 <- 0
-#     w.s3 <- 0
-#     w.s4 <- 0
-# #     w.s5 <- 0
-# #     w.s6 <- 0
-# #     w.s7 <- 0
-# #     w.s8 <- 0
-# #     w.s9 <- 0
-#   for (k in 1:length(rlist.1)) {
-# #   for (k in c(9:11)) {
-#     w.s1 <- sum(rlist.1[[k]][1:n1])+w.s1
-#     w.s2 <- sum(rlist.2[[k]][1:n1])+w.s2
-#     w.s3 <- sum(rlist.3[[k]][1:n1])+w.s3
-#     w.s4 <- sum(rlist.4[[k]][1:n1])+w.s4
-# #     w.s5 <- sum(rlist.5[[k]][1:n1])+w.s5
-# #     w.s6 <- sum(rlist.6[[k]][1:n1])+w.s6
-# #     w.s7 <- sum(rlist.7[[k]][1:n1])+w.s7
-# #     w.s8 <- sum(rlist.8[[k]][1:n1])+w.s8
-# #     w.s9 <- sum(rlist.9[[k]][1:n1])+w.s9
-#     }
-#     w.s <- sum(w.s1,w.s2,w.s3,w.s4) #,w.s5,w.s6,w.s7,w.s8,w.s9)
-#   
-# ## other statistic
-# # mean for full year
-#     mu <- n1*(n1+n2+1)/2
-#     mu.s <- 12*4*mu
-# # ## mean for one season
-# #     mu <- n1*(n1+n2+1)/2
-# #     mu.s <- 3*4*mu
-#   
-# ## variance (standard variation).
-The autocorrelation of the residuals can be assumed to be null (checked, cor.test(), true for NSW, mostly true for WA.
-QLD?
-there are some situations in WA require further investigation)
-# # assume only cross correlation
-#     ss <- n1*n2*(n1+n2+1)/12
-#     ss.s <- 0
-#     for (kk in 1:4) {
-#       for (kkk in 1:4) {
-# #         for (kkkk in 1:12) {
-#           ss.s <- 12*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
-# #         }
-#       }
-#     }
-# ## variance for one season   
-# #     ss <- n1*n2*(n1+n2+1)/3
-# #     ss.s <- 0
-# #     for (kk in 1:4) {
-# #       for (kkk in 1:4) {
-# #           ss.s <- 3*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
-# #       }
-# #     }
-#   
-# ## step trend statistic
-#     z <- (w.s - mu.s)/sqrt(ss.s)
-#   
-#     return(z) }
-#     
-#      else {return(NA)} }
-#      
-#      else {return(NA)}
-#   
-#   }
-# 
-# 
-# zpq.values = numeric(R)
-# znq.values = numeric(R)
-# mth.lnsw <- rep(11,12)
-# mth.lqld <- rep(9,12)
-# 
-# ptm <- proc.time()
-# for (i in 1:R) {
-#   
-#   for (j in 1:12) {
-#     new.order[[j]]<-sample(1:mth.lqld[j])
-#   }
-#   
-#   temp <- unlist(lapply(1:(length(residual_qld)-56),function(x,y) NstepTrend(x,y,residual_qld,1),new.order))
-#   if(i==1) {write.table(temp,file="bt_zvalues_qld.txt",row.names=F)
-#             } else
-#               {write.table(temp,file="bt_zvalues_qld.txt",append=T,row.names=F,col.names=F)}
-#   zpq.values[i] <- length(subset(temp,temp>=1.645))/length(temp)
-#   znq.values[i] <- length(subset(temp,temp<=-1.645))/length(temp)
-# }
-# proc.time()-ptm
-# 
-# write.table(zpq.values,file="bt_outp_qld.txt",row.names=F)
-# write.table(znq.values,file="bt_outn_qld.txt",row.names=F)
-# 
-bt4_nswp.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_outp_nsw.txt",header=T,sep=""))*100
-bt4_nswn.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_outn_nsw.txt",header=T,sep=""))*100
-
-bt4_qldp.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_outp_qld.txt",header=T,sep=""))*100
-bt4_qldn.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_outn_qld.txt",header=T,sep=""))*100
-
-setEPS()
-postscript("bt4_nswp.eps")
-hist(bt4_nswp.values,nclass=100,xlab="% positive Z'",main="4-pixel step trend",cex.lab=1.4,cex.main=1.3)
-points(nsw_origp*100,0,pch=16,col="red",cex=1.8)
-dev.off()
-
-setEPS()
-postscript("bt4_nswn.eps")
-hist(bt1_nswn.values,nclass=100,xlab="% negative Z'",main="4-pixel step trend",cex.lab=1.4,cex.main=1.3)
-points(nsw_orign*100,0,pch=16,col="red",cex=1.8)
-dev.off()
-
-setEPS()
-postscript("bt4_qldp.eps")
-hist(bt4_qldp.values,nclass=100,xlab="% positive Z'",main="4-pixel step trend",cex.lab=1.4,cex.main=1.3)
-points(qld_origp*100,0,pch=16,col="red",cex=1.8)
-dev.off()
-
-setEPS()
-postscript("bt4_qldn.eps")
-hist(bt4_qldn.values,nclass=100,xlab="% negative Z'",main="4-pixel step trend",cex.lab=1.4,cex.main=1.3)
-points(qld_orign*100,0,pch=16,col="red",cex=1.8)
-dev.off()
-
-@
 
 % chunk 9: bootstrapping (spatial structure is kept, remove short term trend, randomly select after change period)
-<<chunk9,eval=F , echo = F, results = hide>>=
-
-## x: rainfall series; y: new order; ii: selection of region
-NstepTrend3 <- function(x,y,ii){
-
-  if(!is.na(unlist(x)[1])) {
-
-  v <- x[c(229:360)][order(y)]
-  ttrd <- c(1:132)
-  v <- residuals(lm(v~ttrd))
-
-  n1 <- sample(c(4:8),1)
-
-  if(ii==1) {    
-      yy <- ts(v[c(1:60,85:132)],start=c(1998,1),frequency=12)    
-      n2 <- 9-n1
-    } else
-      if(ii==2) {
-      yy <- ts(v,start=c(1998,1),frequency=12)
-      n2 <- 11-n1
-    }
-
-  for(i in 1:12) { 
-  y.mth <- as.vector(window(yy,start=c(1998,i),frequency=1))
-
-  ## rank by month   
-    assign(paste("r",i,sep=""),rank(y.mth) )
-  }
-
-## mann-whitney rank sum statistic
-    rlist <- list(r1,r2,r3,r4,r5,r6,r7,r8,r9,r10,r11,r12) ## full year
-
-    w.s <- 0
-    for (k in 1:length(rlist)) {
-    w.s <- sum(rlist[[k]][1:n1])+w.s
-    }
-
-## other statistic
-## mean for full year
-    mu <- n1*(n1+n2+1)/2
-    mu.s <- length(rlist)*mu
-
-## variance (standard variation). The autocorrelation of the residuals can be assumed to be null (checked, true for QLD and NSW)
-    ss <- n1*n2*(n1+n2+1)/length(rlist)
-    ss.s <- length(rlist)*ss
-
-## step trend statistic
-    z <- (w.s - mu.s)/sqrt(ss.s)
-
-    return(z) 
-    }
-
-     else {return(NA)}
-
-  }
-
-R <- 1000
-z3pq.values = numeric(R)
-z3nq.values = numeric(R)
-## random sample with replacement. Both the order of the months and the order of years change
-
-ptm <- proc.time()
-for (i in 1:R) {
-
-  new.order <- sample(c(1:132),replace=T)
-  temp <- unlist(lapply(residual_qld,function(x,y) NstepTrend3(x,y,1),new.order))
-#   if(i==1) {write.table(temp,file="bt_zvalues3_qld.txt",row.names=F)
-#             } else
-#               {write.table(temp,file="bt_zvalues3_qld.txt",append=T,row.names=F,col.names=F)}
-  z3pq.values[i] <- length(subset(temp,temp>=1.645))/length(temp)
-  z3nq.values[i] <- length(subset(temp,temp<=-1.645))/length(temp)
-}
-proc.time()-ptm
-
-write.table(z3pq.values,file="bt_out3p_qld.txt",row.names=F)
-write.table(z3nq.values,file="bt_out3n_qld.txt",row.names=F)
-
-#NSW
-z3pn.values = numeric(R)
-z3nn.values = numeric(R)
-## random sample with replacement. Both the order of the months and the order of years change
-
-ptm <- proc.time()
-for (i in 1:R) {
-
-  new.order <- sample(c(1:132),replace=T)
-  temp <- unlist(lapply(residual_nswvic,function(x,y) NstepTrend3(x,y,2),new.order))
-#   if(i==1) {write.table(temp,file="bt_zvalues3_qld.txt",row.names=F)
-#             } else
-#               {write.table(temp,file="bt_zvalues3_qld.txt",append=T,row.names=F,col.names=F)}
-  z3pn.values[i] <- length(subset(temp,temp>=1.645))/length(temp)
-  z3nn.values[i] <- length(subset(temp,temp<=-1.645))/length(temp)
-}
-proc.time()-ptm
-
-write.table(z3pn.values,file="bt_out3p_nsw.txt",row.names=F)
-write.table(z3nn.values,file="bt_out3n_nsw.txt",row.names=F)
-# 
-# setEPS()
-# postscript("bt3_qldp.eps")
-# hist(z3pq.values,nclass=100,xlab="% positive Z'",main="QLD",cex.lab=1.4,cex.main=1.3)
-# points(qld_origp,0,pch=16,col="red",cex=1.8)
-# dev.off()
-# 
-# setEPS()
-# postscript("bt3_qldn.eps")
-# hist(z3nq.values,nclass=100,xlab="% negative Z'",main="QLD",cex.lab=1.4,cex.main=1.3)
-# points(qld_orign,0,pch=16,col="red",cex=1.8)
-# dev.off()
-# 
-# setEPS()
-# postscript("bt3_nswp.eps")
-# hist(z3pn.values,nclass=100,xlab="% positive Z'",main="NSW",cex.lab=1.4,cex.main=1.3)
-# points(nsw_origp,0,pch=16,col="red",cex=1.8)
-# dev.off()
-# 
-# setEPS()
-# postscript("bt3_nswn.eps")
-# hist(z3nn.values,nclass=100,xlab="% negative Z'",main="NSW",cex.lab=1.4,cex.main=1.3)
-# points(nsw_orign,0,pch=16,col="red",cex=1.8)
-# dev.off()
-
-@
 %% apply the LocSig function in SpatialVx for block bootstrapping
-<<chunk91,eval=F , echo = F, results = hide>>=
-
-require(SpatialVx)
-
-## reconstruct the data into format required in LocSig
-qld_res_mat <- matrix(unlist(residual_qld),nrow=360)
-nsw_res_mat <- matrix(unlist(residual_nswvic),nrow=360)
-
-## the step trend function for LocSig
-NstepTrend4 <- function(x,ii){
-
-  n1 <- sample(c(4:8),1)
-  zz <- vector()
-
-  for(m in 1:ncol(x)) {
-    if(!is.na(x[1,m])) {
-
-  v <- x[c(229:360),m]
-  ttrd <- c(1:132)
-  v <- residuals(lm(v~ttrd))
-
-  if(ii==1) {    
-      yy <- ts(v[c(1:60,85:132)],start=c(1998,1),frequency=12)    
-      n2 <- 9-n1
-    } else
-      if(ii==2) {
-      yy <- ts(v,start=c(1998,1),frequency=12)
-      n2 <- 11-n1
-    }
-
-  for(i in 1:12) { 
-  y.mth <- as.vector(window(yy,start=c(1998,i),frequency=1))
-
-  ## rank by month   
-    assign(paste("r",i,sep=""),rank(y.mth) )
-  }
-
-## mann-whitney rank sum statistic
-    rlist <- list(r1,r2,r3,r4,r5,r6,r7,r8,r9,r10,r11,r12) ## full year
-
-    w.s <- 0
-    for (k in 1:length(rlist)) {
-    w.s <- sum(rlist[[k]][1:n1])+w.s
-    }
-
-## other statistic
-## mean for full year
-    mu <- n1*(n1+n2+1)/2
-    mu.s <- length(rlist)*mu
-
-## variance (standard variation). The autocorrelation of the residuals can be assumed to be null (checked, true for QLD and NSW)
-    ss <- n1*n2*(n1+n2+1)/length(rlist)
-    ss.s <- length(rlist)*ss
-
-## step trend statistic
-    z <- (w.s - mu.s)/sqrt(ss.s)
-    zz[m] <- z
-    }
-
-     else {zz[m]<-NA}
-  }
-  return(zz)
-
-  }
-@
 
@@
 % chunk 8
-<<chunk81,eval=F , echo = F, results = hide>>=
-
-# ## calculate mean rainfall difference in two period for 11-year
-# 
-#   sel_neg.qld <- residual_qld[which(values(tr_qld.step.new)>=1.645)]
-# #   sel_pos.qld <- coef.rain.qld[which(values(tr_qld.step)<(-1.645))]
-#   
-#   sel_neg.nswvic <- residual_nswvic[which(values(tr_nswvic.step.new)>=1.645)]
-# 
-# ## compare mean of before and after
-# ## boxplot
-#   data.qld <- data.frame(year=seq(1998,2008,1))
-#   for (i in 1:length(sel_neg.qld)) {
-#     data.qld[,i+1] <- as.vector(rollapply(as.zoo(sel_neg.qld[[i]])[229:360],width=12,sum,by=12))
-#   }
-#   data2.qld <- melt(data.qld,id="year")
-#   data2.qld$group[data2.qld$year<2003] <- "1before"
-#   data2.qld$group[data2.qld$year>2004] <- "2after"
-# 
-# 
-#   data.nswvic <- data.frame(year=seq(1998,2008,1))
-#   for (i in 1:length(sel_neg.nswvic)) {
-#     data.nswvic[,i+1] <- as.vector(rollapply(as.zoo(sel_neg.nswvic[[i]])[229:360],width=12,sum,by=12))
-#   }
-#   data2.nswvic <- melt(data.nswvic,id="year")
-#   data2.nswvic$group[data2.nswvic$year<2003] <- "1before"
-#   data2.nswvic$group[data2.nswvic$year>2002] <- "2after"
-# 
-# #   nsw_sig <- nsw_30[which(Cp30_nswvic<=0.05)]
-# #   data.nswvic <- data.frame(year=seq(1979,2008,1))
-# #   for (i in 1:length(nsw_sig)) {
-# #     data.nswvic[,i+1] <- as.vector(rollapply(as.zoo(nsw_sig[i]),width=12,sum,by=12))
-# #   }
-# #   data2.nswvic <- melt(data.nswvic,id="year")
-# #   data2.nswvic$group[data2.nswvic$year<2003] <- "1before"
-# #   data2.nswvic$group[data2.nswvic$year>2002] <- "2after"
-# 
-# 
-# ## perform a 2 sample t test
-#   var.test(data2.qld[data2.qld$year<2003,3],data2.qld[data2.qld$year>2004 & data2.qld$year!=2006,3],alternative="two.sided")
-# ## unequal variance
-#   t.test(data2.qld[data2.qld$year<2003,3],data2.qld[data2.qld$year>2004 & data2.qld$year!=2006,3],alternative="greater")
-# ## p = 2.2e-16
-# 
-#   var.test(data2.nswvic[data2.nswvic$year<2003,3],data2.nswvic[data2.nswvic$year>2002& data2.nswvic$year!=2006,3],alternative="two.sided")
-# ## unequal variance
-#   t.test(data2.nswvic[data2.nswvic$year<2003,3],data2.nswvic[data2.nswvic$year>2002& data2.nswvic$year!=2006,3],alternative="greater")
-# ## p < 1.4e-13
-# 
-# 
-#   setEPS()
-#   postscript("compare_mean.eps")
-#   par(mfrow=c(1,2))
-#   boxplot(value~group,data=data2.qld,names=c("before","after"),boxwex=0.4,main="QLD site",ylab="rainfall (mm)")
-# 
-#   boxplot(value~group,data=data2.nswvic,names=c("before","after"),boxwex=0.4,main="NSW/VIC site")
-#   dev.off()
-
-@
 
@@
 % chunk 7
-<<chunk73,eval=F ,echo=F,results = hide>>=
-
-#   tcr_r1 <- resample(aggregate(mod_tc_r1,fact=18,expand=F),reg_qld[[1]])
-#   tcr_r2 <- resample(aggregate(mod_tc_r2,fact=18,expand=F),reg_nswvic[[1]])
-# 
-#   tc_r1 <- tcr_r1[[6]]-tcr_r1[[3]]
-#   tc_r2 <- tcr_r2[[4]]-tcr_r2[[3]]
-#   
-#   
-#   plotclr <- brewer.pal(6,"PiYG")
-# 
-#   setEPS()
-#   postscript("R:\\AVSproject\\WrittenWork\\Paper\\Paper 11 - Australian data regression\\tc_qld.eps")
-#   
-#   par(mar=c(1,1,1,6))
-#   plot(e111,border="white")
-#   image(tc_r1,col=plotclr,breaks=c(-30,-20,-10,0,10,20,30),legend=F,add=T)
-#   plot(mdb_reg,border="grey",add=T)
-#   image.plot(zlim=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3),lab.breaks=c(-30,-20,-10,0,10,20,30),legend.only=T,horizontal=F,col=plotclr,legend.shrink=0.4,legend.width=0.7,legend.args=list(text="change (% ground)",line=0.5,cex=0.7,font=2),axis.args=list(cex.axis=0.8))
-# 
-#   dev.off()
-#   
-#   setEPS()
-#   postscript("R:\\AVSproject\\WrittenWork\\Paper\\Paper 11 - Australian data regression\\tc_nswvic.eps")
-#   
-#   par(mar=c(1,1,1,6))
-#   plot(e222,border="white")
-#   image(tc_r2,col=plotclr,breaks=c(-30,-20,-10,0,10,20,30),legend=F,add=T)
-#   plot(mdb_reg,border="grey",add=T)
-#   plot(aus_reg,border="grey",add=T)  
-#   image.plot(zlim=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3),lab.breaks=c(-30,-20,-10,0,10,20,30),legend.only=T,horizontal=F,col=plotclr,legend.shrink=0.4,legend.width=0.7,legend.args=list(text="change (% ground)",line=0.5,cex=0.7,font=2),axis.args=list(cex.axis=0.8))
-# 
-#   dev.off()
-
-@
 
@@
 % chunk 7
-<<chunk72,eval=F ,echo=F,results = hide>>=
-
-## function to determine whether the tree cover change is permanent
-#   fun1 <- function(x,m) {
-#     m1 <- mean(x[1:3])
-#     if (m==1) {
-#       m2 <- mean(x[6:11])
-#     } else { m2 <- mean(x[4:11])}
-#     if(!is.na(m2-m1) & abs(m2-m1)>5) {return (m2-m1)} else {return(NA)}
-#   }
-#   
-#   tc_m_r1 <- calc(tcr_r1, function(x) fun1(x,1))
-#   tc_m_r2 <- calc(tcr_r2, function(x) fun1(x,2))
-#   
-#   fun2 <- function(x,m) {
-#     d1 <- sd(x[1:3])
-#     if (m==1) {
-#       d2 <- sd(x[6:11])
-#     } else { d2 <- sd(x[4:11])}
-#     if(!is.na(d2/d1) & d2/d1>1) {return(NA)} else {return (d2/d1)}
-#   }
-#   
-#   tc_d_r1 <- calc(tcr_r1, function(x) fun2(x,1))
-#   tc_d_r2 <- calc(tcr_r2, function(x) fun2(x,2))
-#   
-#   tc_m_r1 <- mask(tc_m_r1,tc_d_r1)
-#   tc_m_r2 <- mask(tc_m_r2,tc_d_r2)
-#   
-# ## find out whether the trend and the change have the same sign  
-#   temp_r1 <- tc_r1*tc_m_r1
-#   temp_r1[temp_r1<0] <- NA
-#   tc_r1 <- mask(tc_r1,temp_r1)
-#   
-#   temp_r2 <- tc_r2*tc_m_r2
-#   temp_r2[temp_r2<0] <- NA
-#   tc_r2 <- mask(tc_r2,temp_r2)
-#   
-# ## find out the relationship between Z' and tc
-#   sigtc_z.qld <- tr_qld.step*tc_r1
-#   sigtc_z.nswvic <- tr_nswvic.step*tc_r2
-# 
-#   setEPS()
-#   postscript("R:\\AVSproject\\WrittenWork\\Paper\\Paper 11 - Australian data regression\\tc_z_qld.eps")
-#   
-#   par(mar=c(1,1,1,6))
-#   plot(e111,border="grey")
-#   image(sigtc_z.qld,col=c("blue","red"),breaks=c(-10,0,10),legend=F,add=T)
-#   plot(e111,border="grey",add=T)
-#   plot(mdb_reg,border="grey",add=T)
-#   image.plot(zlim=c(-1,1),breaks=c(-1,0,1),lab.breaks=c("negative","","positive"),legend.only=T,horizontal=F,col=c("red","blue"),legend.shrink=0.2,legend.width=0.7,axis.args=list(cex.axis=0.8))
-# 
-#   dev.off()
-#   
-#   setEPS()
-#   postscript("R:\\AVSproject\\WrittenWork\\Paper\\Paper 11 - Australian data regression\\tc_z_nswvic.eps")
-#   
-#   par(mar=c(1,1,1,6))
-#   plot(e222,border="grey")
-#   image(sigtc_z.nswvic,col=c("blue","red"),breaks=c(-20,0,5),legend=F,add=T)
-#   plot(e222,border="grey",add=T)
-#   plot(mdb_reg,border="grey",add=T)
-#   plot(aus_reg,border="grey",add=T)
-#   image.plot(zlim=c(-1,1),breaks=c(-1,0,1),lab.breaks=c("negative","","positive"),legend.only=T,horizontal=F,col=c("red","blue"),legend.shrink=0.2,legend.width=0.7,axis.args=list(cex.axis=0.8))
-# 
-#   dev.off()
-
-@
 
@@
 %% supplement R code for other purpose
-<<other,eval=F ,echo=F,results = hide>>=
-
-## mod_tc_e.l are the raw tree cover data for each region
-# mod_tc_e2.l
-# mod_tc_e4.l
-# mod_tc_e5.l
-# 
-# LCtrend_qld2 <- addLayer(mod_tc_e2.l,tr_qld2.step)
-# LCtrend_nswvic <- addLayer(mod_tc_e4.l,tr_nswvic.step)
-# LCtrend_wa <- addLayer(mod_tc_e5.l,tr_wa.step)
-# 
-# cover.trend <- function(x,y,m) {
-#   yr <- seq(1:11)
-# #   if(is.na(y[x][12])) {
-# #   return(NA) } else {
-#     tc <- y[x][1:11]
-#     fit.tr <- lm(tc~yr)
-#     if(m==1) return(coef(summary.lm(fit.tr))[2,1])
-#     if(m==2) return(coef(summary.lm(fit.tr))[2,4])
-# #   }
-# }
-# 
-# LCtr_qld2 <- lapply(1:ncell(mod_tc_e2.l),cover.trend,mod_tc_e2.l,1)
-# LCtrq_qld2 <- lapply(1:ncell(mod_tc_e2.l),cover.trend,mod_tc_e2.l,2)
-# LCtr.qld2 <- raster(tr_qld2.step)
-# LCtr.qld2[1:ncell(tr_qld2.step)] <- LCtr_qld2
-# LCtrq.qld2 <- raster(tr_qld2.step)
-# LCtrq.qld2[1:ncell(tr_qld2.step)] <- LCtrq_qld2
-# LCtrend_qld2 <- addLayer(LCtrend_qld2,LCtr.qld2,LCtrq.qld2)
-# 
-# LCtr_nswvic <- lapply(1:ncell(LCtrend_nswvic),cover.trend,LCtrend_nswvic,1)
-# LCtrq_nswvic <- lapply(1:ncell(LCtrend_nswvic),cover.trend,LCtrend_nswvic,2)
-# LCtr.nswvic <- raster(tr_nswvic.step)
-# LCtr.nswvic[1:ncell(tr_nswvic.step)] <- LCtr_nswvic
-# LCtrq.nswvic <- raster(tr_nswvic.step)
-# LCtrq.nswvic[1:ncell(tr_nswvic.step)] <- LCtrq_nswvic
-# LCtrend_nswvic <- addLayer(LCtrend_nswvic,LCtr.nswvic,LCtrq.nswvic)
-# 
-# LCtr_wa <- lapply(1:ncell(LCtrend_wa),cover.trend,LCtrend_wa,1)
-# LCtrq_wa <- lapply(1:ncell(LCtrend_wa),cover.trend,LCtrend_wa,2)
-# LCtr.wa <- raster(tr_wa.step)
-# LCtr.wa[1:ncell(tr_wa.step)] <- LCtr_wa
-# LCtrq.wa <- raster(tr_wa.step)
-# LCtrq.wa[1:ncell(tr_wa.step)] <- LCtrq_wa
-# LCtrend_wa <- addLayer(LCtrend_wa,LCtr.wa,LCtrq.wa)
-# 
-# ##----------------------------------------------------------------------------------------------
-# ## central QLD
-# plot(e12,border="grey",lwd=3)
-# image(LCtr.qld2,col=c(plotclt[8],plotclt[3]),breaks=c(-3,0,2),add=T)
-# image.plot(zlim=c(0,2),breaks=c(0,1,2),lab.breaks=c(-3,0,2),legend.only=T,horizontal=F,col=c(plotclt[8],plotclt[3]),legend.shrink=0.4,legend.width=0.7,legend.args=list(text="trend",line=0.5,cex=0.7,font=2),axis.args=list(cex.axis=0.8))
-# 
-# plot(e12,border="grey",lwd=3)
-# image(LCtrq.qld2,col=rev(plotclt[6:10]),breaks=c(0,0.01,0.05,0.1,0.2,1),add=T)
-# image.plot(zlim=c(0,5),breaks=c(0,1,2,3,4,5),lab.breaks=c(0,0.01,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=rev(plotclt[6:10]),legend.shrink=0.4,legend.width=0.7,legend.args=list(text="p",line=0.5,cex=0.7,font=2),axis.args=list(cex.axis=0.8))
-# 
-# ##----------------------------------------------------------------------------------------------
-# ## NSWVIC
-# plot(e22,border="grey",lwd=3)
-# image(LCtrend_nswvic[[13]],col=c(plotclt[8],plotclt[3]),breaks=c(-6,0,3),add=T)
-# image.plot(zlim=c(0,2),breaks=c(0,1,2),lab.breaks=c(-6,0,3),legend.only=T,horizontal=F,col=c(plotclt[8],plotclt[3]),legend.shrink=0.4,legend.width=0.7,legend.args=list(text="trend",line=0.5,cex=0.7,font=2),axis.args=list(cex.axis=0.8))
-# 
-# plot(e22,border="grey",lwd=3)
-# image(LCtrend_nswvic[[14]],col=rev(plotclt[6:10]),breaks=c(0,0.01,0.05,0.1,0.2,1),add=T)
-# image.plot(zlim=c(0,5),breaks=c(0,1,2,3,4,5),lab.breaks=c(0,0.01,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=rev(plotclt[6:10]),legend.shrink=0.4,legend.width=0.7,legend.args=list(text="p",line=0.5,cex=0.7,font=2),axis.args=list(cex.axis=0.8))
-# 
-# ##----------------------------------------------------------------------------------------------
-# ## WA
-# plot(e31,border="grey",lwd=3)
-# image(LCtrend_wa[[13]],col=c(plotclt[8],plotclt[3]),breaks=c(-5,0,3),add=T)
-# image.plot(zlim=c(0,2),breaks=c(0,1,2),lab.breaks=c(-5,0,3),legend.only=T,horizontal=F,col=c(plotclt[8],plotclt[3]),legend.shrink=0.4,legend.width=0.7,legend.args=list(text="trend",line=0.5,cex=0.7,font=2),axis.args=list(cex.axis=0.8))
-# 
-# plot(e31,border="grey",lwd=3)
-# image(LCtrend_wa[[14]],col=rev(plotclt[6:10]),breaks=c(0,0.01,0.05,0.1,0.2,1),add=T)
-# image.plot(zlim=c(0,5),breaks=c(0,1,2,3,4,5),lab.breaks=c(0,0.01,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=rev(plotclt[6:10]),legend.shrink=0.4,legend.width=0.7,legend.args=list(text="p",line=0.5,cex=0.7,font=2),axis.args=list(cex.axis=0.8))
-# 
-# 
-# ## calculate difference between any two years  
-#   fun <- function(x) {
-#     tc_diff <- diff(x)
-#     y <- x[1:10]
-#     tc_change <- tc_diff/y
-#     return(tc_change)
-#   }
-#   mod_tc_e1 <- calc(mod_tc_r1,fun)
-#   mod_tc_e2 <- calc(mod_tc_r2,fun)
-#   
-# ## set pixels with max change <0.5 or >100 to NA  
-#   temp_qld <- stackApply(abs(mod_tc_e1),indices=rep(1,10),fun=max)
-#   temp_nsw <- stackApply(abs(mod_tc_e2),indices=rep(1,10),fun=max)
-#   
-#   fun <- function(x) {
-#     x[x<0.5 | x>100] <- NA
-#     return(x)
-#   }
-#   temp_qld <- calc(temp_qld,fun)
-#   mod_tc_r1 <- mask(mod_tc_r1,temp_qld)
-#   mod_tc_e1 <- mask(mod_tc_e1,temp_qld)
-#   
-#   temp_nsw <- calc(temp_nsw,fun)
-#   mod_tc_r2 <- mask(mod_tc_r2,temp_nsw)
-#   mod_tc_e2 <- mask(mod_tc_e2,temp_nsw)  
-#   
-# ## calculate average tree cover over 11 years. discard pixels with <5% tree cover
-#   mod44_tc_mean <- calc(mod44_tc_aus,mean)
-#   fun <- function(x) {x[x<5 | x==-1 | x==253]<-NA; return(x)}
-#   mod44_tc_mean <- calc(mod44_tc_mean,fun)  
-#   
-# ## analysis change of tree cover change
-# ## 1. cover trend and significant
-#   cover.trend <- function(x,y) {
-#     yr <- seq(1:length(y[x]))
-#     tc <- y[x]
-#     if (is.na(mean(tc))) {
-#       return(NA)
-#     } else {
-#       fit.tr <- lm(tc~yr)
-#       return(coef(summary.lm(fit.tr)))
-#     }
-#   }
-# 
-#   LC_qld <- lapply(1:ncell(mod_tc_r1),cover.trend,mod_tc_r1)
-#   
-#   LCtr_qld <- raster(mod_tc_r1)
-#   LCtr_qld[] <- sapply(LC_qld,function(x) if(!is.na(x)) {x[2,1]} else {x})
-#   
-#   LCtrp_qld <- raster(mod_tc_r1)
-#   LCtrp_qld[] <- sapply(LC_qld,function(x) if(!is.na(x)) {x[2,4]} else {x})
-#   
-#   LC_nswvic <- lapply(1:ncell(mod_tc_r2),cover.trend,mod_tc_r2)
-#   
-#   LCtr_nswvic <- raster(mod_tc_r2)
-#   LCtr_nswvic[] <- sapply(LC_nswvic,function(x) if(!is.na(x)) {x[2,1]} else {x})
-#   
-#   LCtrp_nswvic <- raster(mod_tc_r2)
-#   LCtrp_nswvic[] <- sapply(LC_nswvic,function(x) if(!is.na(x)) {x[2,4]} else {x})
-#   
-# ## note: in VIC, the two data sets match well in terms of the large trend  
-#   
-# ## 2. maximum jump and its position of tree cover change
-#   cover_change <- function(x,y) {
-#     temp <- y[x]
-#     if(is.na(mean(temp))) {
-#       return(NA)
-#     } else {
-#       jump <- temp[which.is.max(abs(temp))]
-#       jump.position <- which.is.max(abs(temp))+1   # the position of after jump
-#       bef.m <- mean(temp[1:(jump.position-1)])
-#       aft.m <- mean(temp[jump.position:length(temp)])
-#       bef.v <- sd(temp[1:(jump.position-1)])
-#       aft.v <- sd(temp[jump.position:length(temp)])
-#           }
-#     return(data.frame(jump=jump,jump.pos=jump.position,bef.m=bef.m,bef.v=bef.v,aft.m=aft.m,aft.v=aft.v))
-#   }
-# 
-#   LC1_qld <- lapply(1:ncell(mod_tc_e1),cover_change,mod_tc_e1)
-#   
-#   qld.jump <- raster(mod_tc_e1)
-#   qld.jump[] <- sapply(LC1_qld,function(x) if(!is.na(x)) {x$jump} else {x})
-# 
-#   qld.pos <- raster(mod_tc_e1)
-#   qld.pos[] <- sapply(LC1_qld,function(x) if(!is.na(x)) {x$jump.pos} else {x})
-#   
-#   qld.befm <- raster(mod_tc_e1)
-#   qld.befm[] <- sapply(LC1_qld,function(x) if(!is.na(x)) {x$bef.m} else {x})  
-#   
-#   qld.befv <- raster(mod_tc_e1)
-#   qld.befv[] <- sapply(LC1_qld,function(x) if(!is.na(x)) {x$bef.v} else {x})  
-#   
-#   qld.aftm <- raster(mod_tc_e1)
-#   qld.aftm[] <- sapply(LC1_qld,function(x) if(!is.na(x)) {x$aft.m} else {x}) 
-#   
-#   qld.aftv <- raster(mod_tc_e1)
-#   qld.aftv[] <- sapply(LC1_qld,function(x) if(!is.na(x)) {x$aft.v} else {x})  
-#   
-#   LC1_nswvic <- lapply(1:ncell(mod_tc_e2),cover_change,mod_tc_e2)
-#   
-#   nswvic.jump <- raster(mod_tc_e2)
-#   nswvic.jump[] <- sapply(LC1_nswvic,function(x) if(!is.na(x)) {x$jump} else {x})  
-#   
-#   nswvic.pos <- raster(mod_tc_e2)
-#   nswvic.pos[] <- sapply(LC1_nswvic,function(x) if(!is.na(x)) {x$jump.pos} else {x}) 
-#   
-#   nswvic.befm <- raster(mod_tc_e2)
-#   nswvic.befm[] <- sapply(LC1_nswvic,function(x) if(!is.na(x)) {x$bef.m} else {x})  
-#   
-#   nswvic.befv <- raster(mod_tc_e2)
-#   nswvic.befv[] <- sapply(LC1_nswvic,function(x) if(!is.na(x)) {x$bef.v} else {x})  
-#   
-#   nswvic.aftm <- raster(mod_tc_e2)
-#   nswvic.aftm[] <- sapply(LC1_nswvic,function(x) if(!is.na(x)) {x$aft.m} else {x})  
-#   
-#   nswvic.aftv <- raster(mod_tc_e2)
-#   nswvic.aftv[] <- sapply(LC1_nswvic,function(x) if(!is.na(x)) {x$aft.v} else {x})  
-#   
-# ## find jumps that occur within the first 5 years (2000 - 2004)
-# 
-#   LC2_qld <- brick(qld.jump,qld.pos,qld.befm,qld.befv,qld.aftm,qld.aftv)
-#   qld.pos[qld.pos>5] <- NA
-#   LC2_qld <- mask(LC2_qld,qld.pos)
-#   
-#   LC2_nswvic <- brick(nswvic.jump,nswvic.pos,nswvic.befm,nswvic.befv,nswvic.aftm,nswvic.aftv)
-#   nswvic.pos[nswvic.pos>5] <- NA
-#   LC2_nswvic <- mask(LC2_nswvic,nswvic.pos)
-
-@
 
 %% chunk 10
-<<chunk10,eval=F ,echo=F,results = hide>>=
-
-## climatic indicator data, 100 years
-
-# file1 <- "R:\\Data\\BoM\\SOI\\SOI.csv"
-# SOI <- read.csv(file1,sep=",",skip=6)
-# SOI <- melt(SOI,id="Year")
-# names(SOI) <- c("Year","Month","SOI")
-# SOI <- SOI[order(SOI$Year,SOI$Month),]
-# SOI_ts <- ts(SOI$SOI,start=c(1876,1),end=c(2014,4),frequency=12)
-# SOI_100a <- window(SOI_ts,start=c(1900,1),end=c(2013,12))
-# ## deseasonalised SOI
-# SOI_100 <- stl(SOI_100a,"period")$time.series[,3]+stl(SOI_100a,"period")$time.series[,2]
-# 
-# file2 <- "R:\\Data\\BoM\\SST\\SST34.cdf"
-# SST34 <- open.ncdf(file2)
-# SST34 <- get.var.ncdf(SST34,SST34$var[[1]])
-# SST34_ts <- ts(SST34,start=c(1856,1),end=c(2014,4),frequency=12)
-# SST34_100a <- window(SST34_ts,start=c(1900,1),end=c(2013,12))
-# ## deseasonalised SST34
-# SST34_100 <- stl(SST34_100a,"period")$time.series[,3]+stl(SST34_100a,"period")$time.series[,2]
-# 
-# file3 <- "R:\\Data\\BoM\\SST\\SST3.cdf"
-# SST3 <- open.ncdf(file2)
-# SST3 <- get.var.ncdf(SST3,SST3$var[[1]])
-# SST3_ts <- ts(SST3,start=c(1856,1),end=c(2014,4),frequency=12)
-# SST3_100a <- window(SST3_ts,start=c(1900,1),end=c(2013,12))
-# ## deseasonalised SST34
-# SST3_100 <- stl(SST3_100a,"period")$time.series[,3]+stl(SST3_100a,"period")$time.series[,2]
-# 
-# file4 <- "R:\\Data\\BoM\\SST\\SST4.cdf"
-# SST4 <- open.ncdf(file4)
-# SST4 <- get.var.ncdf(SST4,SST4$var[[1]])
-# SST4_ts <- ts(SST4,start=c(1856,1),end=c(2014,5),frequency=12)
-# SST4_100a <- window(SST4_ts,start=c(1900,1),end=c(2013,12))
-# ## deseasonalised SST4
-# SST4_100 <- stl(SST4_100a,"period")$time.series[,3]+stl(SST4_100a,"period")$time.series[,2]
-# 
-# file5 <- "R:\\Data\\BoM\\PDO\\PDO.latest_1900-.txt"
-# pdo_mth <- read.csv(file5,sep="",header=T)
-# #names(pdo_mth) <- c("year","Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")
-# pdo_mth <- melt.data.frame(pdo_mth,id="YEAR")
-# names(pdo_mth) <- c("Year","Month","PDO")
-# pdo_mth <- pdo_mth[order(pdo_mth$Year,pdo_mth$Month),]
-# PDO_ts <- ts(pdo_mth$PDO,start=c(1900,1),end=c(2014,4),frequency=12)
-# PDO_100a <- window(PDO_ts,start=c(1900,1),end=c(2013,12))
-# 
-# file7 <- "R:\\Data\\BoM\\IOD\\iod_jamstec.txt"
-# IOD <- read.csv(file7,sep="",header=T)
-# IOD <- IOD[,4]
-# IOD_ts <- ts(IOD,start=c(1870,1),frequency=12)
-# IOD_100a <- window(IOD_ts,start=c(1900,1),end=c(2013,12))
-# 
-# ## 100 years awap data
-# nsw_100 <- read.table("C:\\Users\\Sue\\Desktop\\thesis\\NSW_month_rf.csv",header=T)
-# nsw100_mean <- ts(rowMeans(nsw_100[,2:1185],na.rm=T,dims=1),start=c(1900,1),frequency=12)
-# 
-# qld_100 <- read.csv("C:\\Users\\Sue\\Desktop\\thesis\\QLD_month_rf.csv",header=T)
-# qld100_mean <- ts(rowMeans(qld_100[,2:2801],na.rm=T,dims=1),start=c(1900,1),frequency=12)
-# 
-# require(MASS)
-# fitdistr(qld100_mean,"gamma")  ## assume shape=1
-# fitdistr(nsw100_mean,"gamma")  ## assume shape =8
-# ks.test(qld100_mean,1,0.03)
-# ks.test(nsw100_mean,8,0.03)
-#   
-# ## autocorrelation in rainfall
-# setEPS()
-# postscript("autocorrelation100.eps")
-# par(mfrow=c(2,2))
-#   acf(qld100_mean,lag.max=12,main="QLD")
-#   acf(nsw100_mean,lag.max=12,main="NSW & VIC")
-# 
-#   pacf(qld100_mean,lag.max=12,main="")
-#   pacf(nsw100_mean,lag.max=12,main="")  
-# dev.off()
-#   
-# ## deseasonalise & detrend rainfall: assume the deseasonalised monthly zonal rainfall is normal distributed and can be used in pearson cross-correlation function
-# ## use stl function in R
-#   qld100_deseason <- stl(qld100_mean,s.window=length(qld100_mean),t.window=length(qld100_mean))$time.series[,3]
-#   nsw100_deseason <- stl(nsw100_mean,s.window=length(nsw100_mean),t.window=length(nsw100_mean))$time.series[,3]
-# 
-# setEPS()
-# postscript("qld100_mean_ts.eps")
-# plot(qld100_deseason,xlab="Year",ylab="monthly rainfall")
-# abline(v=2003,col="red",lwd=2)
-# text(2004,150,"2003",cex=0.9,col="red")
-# dev.off()
-# 
-# setEPS()
-# postscript("nsw100_mean_ts.eps")
-# plot(nsw100_deseason,xlab="Year",ylab="monthly rainfall")
-# abline(v=2003,col="red",lwd=2)
-# text(2004,150,"2003",cex=0.9,col="red")
-# dev.off()
-#   
-# setEPS()
-# postscript("cor_qld100.eps")
-# par(mfrow=c(2,3),mar=c(5,4,1,1))
-# ccf(as.vector(SOI_100a),as.vector(qld100_deseason),lag.max=12,xlim=c(-12,0),xlab="SOI  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# ccf(as.vector(SST34_100a),as.vector(qld100_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO3.4  Lag",ylab="CCF")
-# ccf(as.vector(SST4_100a),as.vector(qld100_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO4  Lag",ylab="CCF")
-# ccf(as.vector(PDO_100a),as.vector(qld100_deseason),lag.max=24,xlim=c(-24,0),xlab="PDO Lag",ylab="CCF")
-# ccf(as.vector(PDO_100a*SOI_100a),as.vector(qld100_deseason),lag.max=12,xlim=c(-12,0),xlab="PDO*SOI  Lag",ylab="CCF")
-# ccf(as.vector(IOD_100a),as.vector(qld100_deseason),lag.max=12,xlim=c(-12,0),xlab="IOD  Lag",ylab="CCF")
-# dev.off()
-# 
-# ccf(subset(SOI_100a,cycle(SOI_100a)==10|cycle(SOI_100a)==11|cycle(SOI_100a)==12),subset(qld100_deseason,cycle(qld100_deseason)==10|cycle(qld100_deseason)==11|cycle(qld100_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="SOI  Lag",ylab="CCF") ## SOI
-# ccf(subset(SST34_100a,cycle(SST34_100a)==10|cycle(SST34_100a)==11|cycle(SST34_100a)==12),subset(qld100_deseason,cycle(qld100_deseason)==10|cycle(qld100_deseason)==11|cycle(qld100_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="SST34  Lag",ylab="CCF") ## SST34
-# ccf(subset(SST3_100a,cycle(SST3_100a)==10|cycle(SST3_100a)==11|cycle(SST3_100a)==12),subset(qld100_deseason,cycle(qld100_deseason)==10|cycle(qld100_deseason)==11|cycle(qld100_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="SST3  Lag",ylab="CCF") ## SST3
-# ccf(subset(PDO_100a,cycle(PDO_100a)==10|cycle(PDO_100a)==11|cycle(PDO_100a)==12),subset(qld100_deseason,cycle(qld100_deseason)==10|cycle(qld100_deseason)==11|cycle(qld100_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="PDO  Lag",ylab="CCF") ## PDO
-# ccf(subset(PDO_100a,cycle(PDO_100a)==10|cycle(PDO_100a)==11|cycle(PDO_100a)==12)* subset(SOI_100a,cycle(SOI_100a)==10|cycle(SOI_100a)==11|cycle(SOI_100a)==12),subset(qld100_deseason,cycle(qld100_deseason)==10|cycle(qld100_deseason)==11|cycle(qld100_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="PDO  Lag",ylab="CCF") ## PDO
-# ccf(subset(IOD_100a,cycle(IOD_100a)==10|cycle(IOD_100a)==11|cycle(IOD_100a)==12),subset(qld100_deseason,cycle(qld100_deseason)==10|cycle(qld100_deseason)==11|cycle(qld100_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="IOD  Lag",ylab="CCF") ## IOD
-# 
-# 
-# 
-# # ## 1,2,3: SOI
-# # ## 4,5,6: SOI
-# # ## 7,8,9: SOI, IOD
-# # ## 10,11,12: SOI, SST34, SST3, PDO
-# # 
-# # 
-# setEPS()
-# postscript("cor_nsw100.eps")
-# par(mfrow=c(2,3),mar=c(5,4,1,1))
-# ccf(as.vector(SOI_100a),as.vector(nsw100_deseason),lag.max=12,xlim=c(-12,0),xlab="SOI  Lag",ylab="CCF")
-# ccf(as.vector(SST34_100a),as.vector(nsw100_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO3.4  Lag",ylab="CCF")
-# ccf(as.vector(SST4_100a),as.vector(nsw100_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO4  Lag",ylab="CCF")
-# ccf(as.vector(PDO_100a),as.vector(nsw100_deseason),lag.max=24,xlim=c(-24,0),xlab="PDO Lag",ylab="CCF")
-# ccf(as.vector(PDO_100a*SOI_100a),as.vector(nsw100_deseason),lag.max=12,xlim=c(-12,0),xlab="PDO*SOI  Lag",ylab="CCF")
-# ccf(as.vector(IOD_100a),as.vector(nsw100_deseason),lag.max=12,xlim=c(-12,0),xlab="IOD  Lag",ylab="CCF")
-# dev.off()
-# 
-# ccf(subset(SOI_100a,cycle(SOI_100a)==7|cycle(SOI_100a)==8|cycle(SOI_100a)==9),subset(nsw100_deseason,cycle(nsw100_deseason)==7|cycle(nsw100_deseason)==8|cycle(nsw100_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="SOI  Lag",ylab="CCF") ## SOI
-# ccf(subset(SST34_100a,cycle(SST34_100a)==7|cycle(SST34_100a)==8|cycle(SST34_100a)==9),subset(nsw100_deseason,cycle(nsw100_deseason)==7|cycle(nsw100_deseason)==8|cycle(nsw100_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="SST34  Lag",ylab="CCF") ## SST34
-# ccf(subset(SST3_100a,cycle(SST3_100a)==7|cycle(SST3_100a)==8|cycle(SST3_100a)==9),subset(nsw100_deseason,cycle(nsw100_deseason)==7|cycle(nsw100_deseason)==8|cycle(nsw100_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="SST3  Lag",ylab="CCF") ## SST3
-# ccf(subset(PDO_100a,cycle(PDO_100a)==7|cycle(PDO_100a)==8|cycle(PDO_100a)==9),subset(nsw100_deseason,cycle(nsw100_deseason)==7|cycle(nsw100_deseason)==8|cycle(nsw100_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="PDO  Lag",ylab="CCF") ## PDO
-# ccf(subset(PDO_100a,cycle(PDO_100a)==7|cycle(PDO_100a)==8|cycle(PDO_100a)==9)* subset(SOI_100a,cycle(SOI_100a)==7|cycle(SOI_100a)==8|cycle(SOI_100a)==9),subset(nsw100_deseason,cycle(nsw100_deseason)==7|cycle(nsw100_deseason)==8|cycle(nsw100_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="PDO  Lag",ylab="CCF") ## PDO
-# ccf(subset(IOD_100a,cycle(IOD_100a)==7|cycle(IOD_100a)==8|cycle(IOD_100a)==9),subset(nsw100_deseason,cycle(nsw100_deseason)==7|cycle(nsw100_deseason)==8|cycle(nsw100_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="IOD  Lag",ylab="CCF") ## IOD
-
-## 1,2,3: NONE
-#4,5,6: SOI
-# 7, 8, 9: SOI, IOD
-## 10, 11, 12: SOI, SST34, SST3, IOD

### 2017-11-08 chap3_paper_new.Rnw
add files

--- 2017-11-08 LandSurfaceEffects_LocalRainfall_WRRsubmit.Rnw
@@
 \begin{keypoints}
-\item There are few empirical studies of vegetation feedback on rainfall 
+\item There are few empirical studies of vegetation feedback 
 \item Two sites with major step trend landuse change were analysed
@@
 Regression analysis and step trend tests were used to assess the effect of abrupt land surface intervention.
-At a Queensland site, dramatic tree cover change between 2002 - 2005 did not result in strong statistically significant precipitation changes.
+At a Queensland site, dramatic tree cover change between 2002 - 2005 did not result in statistically significant precipitation change.
 On the other hand, results from step trend test on 1998 - 2008 rainfall data in a bushfire affected NSW/VIC region has a better match with the tree cover change map.
@@
 \end{abstract}
+
+\section{Comments to answer by CL FvO and RWV}
+
+\begin{itemize}
+  \item comment 2: Do we have to explain why we don't include lag 1 autocorrelation in the model, or use mixed modelling?  
+  \item comment 3: should we also report p $<$ 10\%? Now included
+  \item comment 5: we seem to be mixing p $<$ 5\% and $\alpha < 0.05$. Changed, Chun can you please check?
+\end{itemize}
 
@@
 \end{figure}
+
+<<mapping,eval=F, echo = F, results = hide>>=
+
+library(plotKML)
+library(raster)
+library(rgeos)
+library(maptools)
+library(fields)
+library(RColorBrewer)
+
+## define the boundary of sub-regions
+  e1 <- extent(c(1324304.091,1723901.773,-222556.668,80451.265))
+  # 400 km by 303 km Qld
+  e2 <- extent(c(1346395.141,1534425.527,-1169402.246,-607644.171))
+  # 188 km by 561 km NSW
+  e11 <- extent(c(145.425, 149.025, -28.475, -25.175))
+  e22 <- extent(c(146.675, 149.325, -37.075, -34.675))
+
+  e111 <- extent(c(145.825, 148.625, -28.075, -25.575)) #as(tc_r1,"SpatialPolygons")
+  # 306 km by 273.5 km
+  e222 <- extent(c(147.075, 148.925, -36.675, -35.075)) #as(tc_r2,"SpatialPolygons")
+  # 203 km by 175 km
+## the selected regions for mapping
+
+  aus_reg <- readShapePoly("Data\\Geo\\aus_state2011\\simplify_state11_uproj.shp",proj4string=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0"))
+  aus_reg2 <- gSimplify(aus_reg,0.01,topologyPreserve=T)
+
+  mdb_reg <- readShapePoly("Data\\Geo\\MDB\\mdb.shp",proj4string=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0"))
+mdb_reg2 <- gSimplify(mdb_reg,0.01,topologyPreserve=TRUE)
+
+  remove(list=c("aus_reg","mdb_reg"))
+
+## Burned area map
+
+fire_0301 <- raster("Data\\Other\\bushfires\\Burnt_Area\\tmpA2003001.burndate.tif")
+fire_0301nsw <- crop(fire_0301,e2)
+fire_loc <- crop(reproject(fire_0301nsw,CRS=CRS("+proj=longlat +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84")),e222)
+remove(list=c("fire_0301","fire_0301nsw"))
+
+  pdf("bushfire_nswvic.pdf")
+#  jpg("bushfire_nswvic.jpg")
+  par(mar=c(3,3,1,6),xpd=T)
+  plot(e222,border="white")  
+  image(fire_loc,add=T, breaks=c(0,0.99,366.5,10000),col=c("white","red","white"))
+  plot(e222,add=T)
+  plot(aus_reg2,border="grey",add=T)
+  plot(mdb_reg2,border="grey",lwd=3,add=T)
+#   text(1430000,-1000000,"study region",cex=1.2)
+  text(149.1, -35.3,"ACT",font=2,cex=1.2)
+  text(148.4, -36,"Kosciuszko",font=2,cex=0.9)#,srt=90)
+  text(147.5, -36.4,"VIC",font=2,cex=1.2)
+  text(148, -35.5,"NSW",font=2,cex=1.2)
+  image.plot(zlim=c(0,1),breaks=c(0,0.5,1),lab.breaks=c("","burned \narea",""),legend.only=T,horizontal=F,col=c("red","red"),legend.shrink=0.03,legend.width=1.3,legend.args=list(text="",line=0),lwd.poly=0.1)
+legend(149,-36.1,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=0.9)
+ dev.off()
+
+@
+
+<<treeCover,eval=F , echo = F, results = hide>>=
+
+## trend of evi (MODIS product) from GA and ABARES
+
+## tree cover data
+  mod44_tc <- brick("Data\\Satellite\\MOD44\\B\\mod44_tc_aus.tif")
+  tm <- seq(as.Date("2000-03-06"), as.Date("2010-03-06"), "year")
+  mod44_tc <- setZ(mod44_tc, tm, "year")
+
+## tree cover in sub-region
+  mod_tc_r11 <- crop(mod44_tc,e1)
+  mod_tc_r111 <- reproject(mod_tc_r11,CRS=CRS("+proj=longlat +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84"))
+  mod_tc_r1 <- crop(mod_tc_r111,e111)
+
+  mod_tc_r22 <- crop(mod44_tc,e2)
+  mod_tc_r222 <- reproject(mod_tc_r22,CRS=CRS("+proj=longlat +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84"))
+ mod_tc_r2 <- crop(mod_tc_r222,e222)
+
+ remove(list=c("mod44_tc","mod_tc_r11","mod_tc_r111","mod_tc_r22","mod_tc_r222"))
+
+## apply t-test to tree cover data
+  fun <- function(x,m) {
+    var_x <- sd(x)
+    mu_1 <- mean(x[1:3])
+    if (m==1) {y <- x[6:11]} else {y <- x[4:11]}   
+    mu_2 <- mean(y)
+    z <- (mu_2-mu_1)/var_x
+    if (!is.na(z) & abs(z)>qt(0.1,length(z))) {return(y[1]-x[3])} else {return(NA)}
+  }
+
+  mod_tc_r1.z <- calc(mod_tc_r1,function(x) fun(x,1))
+  mod_tc_r2.z <- calc(mod_tc_r2,function(x) fun(x,2))
+
+  remove(list=c("mod_tc_r1","mod_tc_r2"))
+
+## create plots
+  plotclr <- brewer.pal(6,"PiYG")
+#"#C51B7D" "#E9A3C9" "#FDE0EF" "#E6F5D0" "#A1D76A" "#4D9221"
+
+  pdf("tc.pdf", width=10, height= 5)
+  par(mar=c(3,3,1,7), mfrow=c(1,2))
+  plot(e111)
+  image(mod_tc_r1.z,col=plotclr,breaks=c(-70,-50,-20,0,20,50,70),legend=F,add=T)
+  plot(e111,add=T)
+  plot(mdb_reg2,border="grey",lwd=3,add=T)
+  image.plot(zlim=c(-3,3),breaks=c(-3,-2,-1,-0.2,0.2,1,2,3),lab.breaks=c(-70,-50,-20,"no tree","0 or",20,50,70),legend.only=T,horizontal=F,col=c("#C51B7D","#E9A3C9","#FDE0EF","white","#E6F5D0","#A1D76A","#4D9221"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text=expression(paste(Delta,"tree (% ground)",sep="")),line=0.5,cex=0.9,font=2),axis.args=list(cex.axis=0.8),graphics.reset=F)
+  text(146.0, -25.6, "(a)", font=2)
+  plot(e222)
+  image(mod_tc_r2.z,col=plotclr,breaks=c(-70,-50,-20,0,20,50,70),legend=F,add=T)
+  plot(e222,add=T)
+  plot(aus_reg2,border="grey",add=T)
+  plot(mdb_reg2,border="grey",lwd=3,add=T)
+  text(147.2, -35.1, "(b)", font=2)
+  dev.off()
+
+pdf("map.pdf")
+par(mar=c(1.5,1.5,1.5,1.5))
+plot(mdb_reg2,border="grey",lwd=3)
+plot(aus_reg2,border="grey",add=T)
+legend(150.2,-36,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=0.9)
+#invisible(text(getSpPPolygonsLabptSlots(aus_reg2), labels=as.character(aus_reg2$NAME), cex=0.7,adj=0.5))
+box(lty=1)
+plot(e111,col="red",add=T)
+plot(e222,col="red",add=T)
+text(147.2,-26.8,"QLD",cex=1.5)
+text(148,-34.7,"NSW/VIC",cex=1.5)
+dev.off()
+
+@
 
@@
 
+<<climaticIndex,eval=F , echo = F, results = hide>>=
+
+library(reshape)
+library(chron)
+library(fields)
+library(rgdal)
+library(raster)
+library(maptools)
+library(graphics)
+library(zoo)
+library(RColorBrewer) # creates nice color schemes
+# library(classInt)     # finds class intervals for continuous variables
+
+library(mgcv)
+library(ncdf4)
+library(nnet)
+
+## climatic indicator data
+
+file1 <- "Data\\BoM\\SOI\\SOI.csv"
+SOI <- read.csv(file1,sep=",",skip=6)
+SOI <- melt(SOI,id="Year")
+names(SOI) <- c("Year","Month","SOI")
+SOI <- SOI[order(SOI$Year,SOI$Month),]
+SOI_ts <- ts(SOI$SOI,start=c(1876,1),end=c(2014,4),frequency=12)
+SOI_100a <- window(SOI_ts,start=c(1900,1),end=c(2008,12))
+SOI_30a <- window(SOI_ts,start=c(1979,1),end=c(2008,12))
+## deseasonalised SOI
+SOI_100 <- stl(SOI_100a,"period")$time.series[,3]+stl(SOI_100a,"period")$time.series[,2]
+SOI_30 <- stl(SOI_30a,"period")$time.series[,3]+stl(SOI_30a,"period")$time.series[,2]
+
+file2 <- "Data\\BoM\\SST\\SST34.cdf"
+SST34 <- nc_open(file2)
+SST34 <- ncvar_get(SST34,SST34$var[[1]])
+SST34_ts <- ts(SST34,start=c(1856,1),end=c(2014,4),frequency=12)
+SST34_30a <- window(SST34_ts,start=c(1979,1),end=c(2008,12))
+## deseasonalised SST34
+SST34_30 <- stl(SST34_30a,"period")$time.series[,3]+stl(SST34_30a,"period")$time.series[,2]
+
+file3 <- "Data\\BoM\\SST\\SST3.cdf"
+SST3 <- nc_open(file3)
+SST3 <- ncvar_get(SST3,SST3$var[[1]])
+SST3_ts <- ts(SST3,start=c(1856,1),end=c(2014,4),frequency=12)
+SST3_30a <- window(SST3_ts,start=c(1979,1),end=c(2008,12))
+## deseasonalised SST34
+SST3_30 <- stl(SST3_30a,"period")$time.series[,3]+stl(SST3_30a,"period")$time.series[,2]
+
+file4 <- "Data\\BoM\\SST\\SST4.cdf"
+SST4 <- nc_open(file4)
+SST4 <- ncvar_get(SST4,SST4$var[[1]])
+SST4_ts <- ts(SST4,start=c(1856,1),end=c(2014,5),frequency=12)
+SST4_30a <- window(SST4_ts,start=c(1979,1),end=c(2008,12))
+## deseasonalised SST4
+SST4_30 <- stl(SST4_30a,"period")$time.series[,3]+stl(SST4_30a,"period")$time.series[,2]
+
+file5 <- "Data\\BoM\\PDO\\PDO.latest_1900-.txt"
+pdo_mth <- read.csv(file5,sep="",header=T)
+pdo_mth <- melt.data.frame(pdo_mth,id="YEAR")
+names(pdo_mth) <- c("Year","Month","PDO")
+pdo_mth <- pdo_mth[order(pdo_mth$Year,pdo_mth$Month),]
+PDO_ts <- ts(pdo_mth$PDO,start=c(1900,1),end=c(2014,4),frequency=12)
+PDO_100a <- window(PDO_ts,start=c(1900,1),end=c(2008,12))
+## deseasonalised PDO
+PDO_100 <- stl(PDO_100a,"period")$time.series[,3]+stl(PDO_100a,"period")$time.series[,2]
+
+file7 <- "Data\\BoM\\IOD\\iod_jamstec.txt"
+IOD <- read.csv(file7,sep="",header=T)
+IOD <- IOD[,4]
+IOD_ts <- ts(IOD,start=c(1870,1),frequency=12)
+IOD_30a <- window(IOD_ts,start=c(1979,1),end=c(2008,12))
+## deseasonalised IOD
+IOD_30 <- stl(IOD_30a,"period")$time.series[,3]+stl(IOD_30a,"period")$time.series[,2]
+# end of read data
+
+@
+
 \section{Statistical method}
@@
 In this study, up to two important indices identified from the seven climatic indicators (see section \ref{sec:Data}) were used, for each study region, as the explanatory variables in the model .
+
+<<rainData,eval=F , echo = F, results = hide>>=
+
+## 100 years awap data & 30 years
+qld_100 <- read.csv("Data\\BOM\\rainfall\\QLD_month_rf.csv",header=T)
+qld_30 <- subset(qld_100,as.Date(qld_100[,1],"%d/%m/%Y")>="1979-01-01" & as.Date(qld_100[,1],"%d/%m/%Y")<="2008-12-01")
+qld100_mean <- ts(rowMeans(qld_100[,2:2801],na.rm=T,dims=1),start=c(1900,1),frequency=12)
+qld30_mean <- window(qld100_mean,start=c(1979,1),end=c(2008,12))
+
+nsw_100 <- read.csv("Data\\BOM\\rainfall\\NSW_month_rf.csv",header=T)
+nsw_30 <- subset(nsw_100,as.Date(nsw_100[,1],"%d/%m/%Y")>="1979-01-01" & as.Date(nsw_100[,1],"%d/%m/%Y")<="2008-12-01")
+nsw100_mean <- ts(rowMeans(nsw_100[,2:1185],na.rm=T,dims=1),start=c(1900,1),frequency=12)
+nsw30_mean <- window(nsw100_mean, start=c(1979,1),end=c(2008,12))
+
+require(MASS)
+fitdistr(qld30_mean,"gamma")  ## assume shape=1
+fitdistr(nsw30_mean,"gamma")  ## assume shape =8
+ks.test(qld30_mean,1,0.03)
+ks.test(nsw30_mean,2.4,0.03)
+
+pdf("hist_rain.pdf", width=8, height = 5)
+par(mfrow=c(1,2))
+ hist(qld30_mean,main="",xlab=expression(paste("rainfall (mm mont",h^-1,")",sep="")))
+ text(200,120,"(a)", font=2)
+ hist(nsw30_mean,main="",xlab=expression(paste("rainfall (mm mont",h^-1,")",sep="")))
+text(200,60,"(b)", font=2)
+dev.off()
+
+## autocorrelation in rainfall
+pdf("autocorrelation.pdf")
+par(mfrow=c(2,2))
+  acf(qld30_mean,lag.max=12,main="QLD")
+  acf(nsw30_mean,lag.max=12,main="NSW & VIC")
+
+  pacf(qld30_mean,lag.max=12,main="")
+  pacf(nsw30_mean,lag.max=12,main="")  
+dev.off()
+
+## deseasonalise & detrend rainfall: assume the deseasonalised monthly zonal rainfall is normal distributed and can be used in pearson cross-correlation function
+## use stl function in R
+  qld30_deseason <- stl(qld30_mean,s.window=length(qld30_mean),t.window=length(qld30_mean))$time.series[,3]
+  nsw30_deseason <- stl(nsw30_mean,s.window=length(nsw30_mean),t.window=length(nsw30_mean))$time.series[,3]
+  qld100_deseason <- stl(qld100_mean,s.window=length(qld100_mean),t.window=length(qld100_mean))$time.series[,3]
+  nsw100_deseason <- stl(nsw100_mean,s.window=length(nsw100_mean),t.window=length(nsw100_mean))$time.series[,3]
+
+pdf("mean_ts.pdf",width=8,height=4)
+par(cex.lab=1,cex.axis=1,mfrow=c(1,2))
+plot(qld30_deseason,xlab="Year",ylab="monthly rainfall")
+abline(v=2003,col="red",lwd=2)
+text(2004.6,150,"2003",cex=1.2,col="red")
+text(1980.6,200,"(a)",font=2, cex=1.2)
+plot(nsw30_deseason,xlab="Year",ylab="monthly rainfall")
+abline(v=2003,col="red",lwd=2)
+text(2004.6,120,"2003",cex=1.2,col="red")
+text(1980.6,150,"(b)", font=2,cex=1.2)
+dev.off()
+
+# Figure 5
+pdf("cor_qld.pdf", width = 8, height = 8)
+par(mfrow=c(2,3),mar=c(5,4,1,1))
+ccf(as.vector(SOI_30),as.vector(qld30_deseason),lag.max=12,xlim=c(-12,0),xlab="SOI  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
+ccf(as.vector(SST34_30),as.vector(qld30_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO3.4  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
+ccf(as.vector(SST4_30),as.vector(qld30_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO4  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
+ccf(as.vector(PDO_100),as.vector(qld100_deseason),lag.max=24,xlim=c(-24,0),xlab="PDO Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
+ccf(as.vector(PDO_100*SOI_100),as.vector(qld100_deseason),lag.max=12,xlim=c(-12,0),xlab="PDO*SOI  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
+ccf(as.vector(IOD_30),as.vector(qld30_deseason),lag.max=12,xlim=c(-12,0),xlab="IOD  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
+dev.off()
+
+pdf("cor_nswvic.pdf",width = 8, height = 8)
+par(mfrow=c(2,3),mar=c(5,4,1,1))
+ccf(as.vector(SOI_30),as.vector(nsw30_deseason),lag.max=12,xlim=c(-12,0),xlab="SOI  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
+ccf(as.vector(SST34_30),as.vector(nsw30_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO3.4  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
+ccf(as.vector(SST4_30),as.vector(nsw30_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO4  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
+ccf(as.vector(PDO_100),as.vector(nsw100_deseason),lag.max=24,xlim=c(-24,0),xlab="PDO Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
+ccf(as.vector(PDO_100*SOI_100),as.vector(nsw100_deseason),lag.max=12,xlim=c(-12,0),xlab="PDO*SOI  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
+ccf(as.vector(IOD_30),as.vector(nsw30_deseason),lag.max=12,xlim=c(-12,0),xlab="IOD  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
+dev.off()
+
+@
 
@@
 
+<<simpleMode,eval=F, echo = F, results = hide>>=
+
+## plotting a hypothesed land cover change model
+data1 <- data.frame(year=c(1,2,2,3),tc=c(2,2,1,1))
+data2 <- data.frame(year=c(1,2,2,3,4),tc=c(2,2,1,1.5,2))
+
+pdf("tc_simple.pdf", width = 9, height = 4)
+par(mar=c(5,6,1,1), mfrow=c(1,2))
+plot(data1$year,data1$tc,type="l",ylim=c(0.5,2.5),xlab="",ylab="",
+     axes=F, frame.plot=F)
+axis(1,at=c(1,2,3),labels=c("before","LCC","after"),cex.axis=1.3)
+axis(2,at=c(1,2),label=c("removed","tree"),cex.axis=1.3,las=2)
+lines(c(1,4),c(2,2),lty=2)
+lines(c(1,4),c(1,1),lty=2)
+lines(c(2,2),c(0,2.5),lty=2)
+text(2.8,2.1,"(a)", font=2)
+plot(data2$year,data2$tc,type="l",ylim=c(0.5,2.5),xlab="",ylab="",axes=F, frame.plot=F)
+axis(1,at=c(1,2,4),labels=c("before","LCC","after"),cex.axis=1.3)
+axis(2,at=c(1,2),label=c("removed","tree"),cex.axis=1.3,las=2)
+lines(c(1,4),c(2,2),lty=2)
+lines(c(1,4),c(1,1),lty=2)
+lines(c(2,2),c(0,2.5),lty=2)
+text(3.5,2.1,"(b)", font=2)
+dev.off()
+
+@
+
+<<fullRegression,eval=F , echo = F, results = hide>>=
+
+## predicting model: land cover is category variable
+rain.trend <- function(x,y,m,var.data) {
+
+  var.data$rain <- as.vector(unlist(y[x]))+0.0001
+
+  if (!is.na(mean(var.data$rain))) {
+    if (grepl("qld",substitute(y))) {
+      var.data <- na.omit(var.data)    
+      rain_fit <- gam(rain ~ s(month, k=3)
+                  + s(SOI, k=3)
+                  + cover  
+                  + trd
+                  , data=var.data,family=Gamma(link=log))
+    } else
+     if (grepl("nsw",substitute(y))) {  
+      var.data <- na.omit(var.data)    
+      rain_fit <- gam(rain ~ s(month, k=3)
+                  + s(SOI, k=3)
+                  + s(IOD, k=3)
+                  + cover  
+                  + trd
+                  , data=var.data,family=Gamma(link=log))
+    }
+
+  return(rain_fit)
+  } else {NA}
+
+}
+
+## note: link=log generally has higher r.sq value than link=reverse
+
+## generate variables for the model
+trd <- seq(1:360)
+month <-rep(1:12,times=30)
+mth <- cycle(ts(month,start=c(1979,1),frequency=12))
+
+LC <- c(rep("tree",288),rep("clear",72))
+LC_qld4 <- c(rep("tree",312),rep("clear",48)) ## after- : from Jan 2005
+
+var.data <- data.frame(month, trd, SOI = SOI_30a, IOD = IOD_30a, cover = LC)
+var.qld4 <- data.frame(month, trd, SOI = SOI_30a, IOD = IOD_30a, cover = LC_qld4)
+
+  coef.rain30.qld4 <- lapply(2:ncol(qld_30), 
+                             function(x) rain.trend(x,qld_30,30,
+                                                    var.qld4))
+  coef.rain30.nswvic <- lapply(2:ncol(nsw_30), 
+                               function(x) rain.trend(x,nsw_30,
+                                                      30,var.data))
+
+## extract sign of C
+  C30_qld <- unlist(lapply(1:length(coef.rain30.qld4),
+    function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",
+                       summary(coef.rain30.qld4[[x]])$p.coeff[2],
+                       NA)))
+  C30_nswvic <- unlist(lapply(1:length(coef.rain30.nswvic),
+    function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",
+                       summary(coef.rain30.nswvic[[x]])$p.coeff[2],
+                       NA)))
+
+Cp30_qld4 <- unlist(lapply(1:length(coef.rain30.qld4),
+   function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",
+                      summary(coef.rain30.qld4[[x]])$p.pv[2],NA)))
+Cp30_nswvic <- unlist(lapply(1:length(coef.rain30.nswvic),
+    function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",
+                       summary(coef.rain30.nswvic[[x]])$p.pv[2],
+                       NA)))
+
+## p value of seasons and climate indices
+p_qld <- unlist(lapply(1:length(coef.rain30.qld4),
+      function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",
+                         summary(coef.rain30.qld4[[x]])$s.pv[1],
+                         NA)))
+p_nsw <- unlist(lapply(1:length(coef.rain30.nswvic),
+      function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",
+                         summary(coef.rain30.nswvic[[x]])$s.pv[1],
+                         NA)))
+
+## extract r.sq of the model
+rsq_qld <- unlist(lapply(1:length(coef.rain30.qld4),
+     function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",
+                        summary(coef.rain30.qld4[[x]])$r.sq,NA)))
+rsq_nsw <- unlist(lapply(1:length(coef.rain30.nswvic),
+     function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",
+                        summary(coef.rain30.nswvic[[x]])$r.sq,NA)))
+
+## frame of the two regions (100 years data)
+qld_tif <- raster("Data\\BoM\\rainfall\\AWAP\\from1900_FVG\\r19000101_qld.tif")
+nsw_tif <- raster("Data\\BoM\\rainfall\\AWAP\\from1900_FVG\\r19000101_nsw.tif")
+
+  dist_Cp30_qld4 <- raster(qld_tif)
+  dist_Cp30_qld4[] <- Cp30_qld4
+  dist_Cp30_nsw <- raster(nsw_tif)
+  dist_Cp30_nsw[] <- Cp30_nswvic
+# ## range of predicted values
+  predictmax_qld30 <- unlist(lapply(1:length(coef.rain30.qld4),
+        function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",
+                          max(coef.rain30.qld4[[x]]$fitted.values),
+                           NA)))
+  predictmin_qld30 <- unlist(lapply(1:length(coef.rain30.qld4),
+        function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",                 min(coef.rain30.qld4[[x]]$fitted.values),NA)))
+# 
+mean_qld <- lapply(1:length(coef.rain30.qld4), 
+          function(x) if(summary(coef.rain30.qld4)[x,2]=="gam") {fitted(coef.rain30.qld4[[x]])
+            } else{NA})
+
+## calculate mean residual
+  cc <- 1
+  qld.sum <- mean_qld[[1]]
+  for (i in 2:length(mean_qld)) {
+    if(!is.na(mean_qld[[i]])) {
+      qld.sum <- qld.sum + mean_qld[[i]]
+      cc <- cc+1
+    }
+  }
+  qld.mean <- qld.sum/cc
+  qld.mean <- ts(qld.mean,start=c(1979,1),frequency=12)
+
+predictmax_nsw30 <- unlist(lapply(1:length(coef.rain30.nswvic),
+        function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",
+                      max(coef.rain30.nswvic[[x]]$fitted.values),
+                      NA)))
+# 70 - 254.7
+  predictmin_nsw30 <- unlist(lapply(1:length(coef.rain30.nswvic),
+        function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",
+                      min(coef.rain30.nswvic[[x]]$fitted.values),
+                      NA)))
+# 18.4 - 62.2
+
+mean_nswvic <- lapply(1:length(coef.rain30.nswvic), 
+          function(x) if(summary(coef.rain30.nswvic)[x,2]=="gam") {
+              fitted(coef.rain30.nswvic[[x]])} else{NA})
+
+## calculate mean residual
+  cc <- 1
+  nswvic.sum <- mean_nswvic[[1]]
+  for (i in 2:length(mean_nswvic)) {
+    if(!is.na(mean_nswvic[[i]])) {
+      nswvic.sum <- nswvic.sum + mean_nswvic[[i]]
+      cc <- cc+1
+    }
+  }
+  nswvic.mean <- nswvic.sum/cc
+  nswvic.mean <- ts(nswvic.mean,start=c(1979,1),frequency=12)
+
+## plots
+  pdf("Cp_30yrs.pdf", width=10, height=4.5)
+  par(mar=c(3,3,1,6),xpd=T, mfrow=c(1,2))
+  plot(e111,xlab="",ylab="")
+  image(dist_Cp30_qld4,col=c("red","orange","yellow",
+                            "light blue"),
+        breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
+#   plot(aus_reg2,border="grey",add=T)
+  plot(mdb_reg2,border="grey",lwd=3,add=T)
+  plot(e111,add=T)
+text(146, -25.4, "(a)", font=2)
+image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.7,legend.width=0.9,legend.args=list(text="p value",line=0.5,cex=1.2,font=1),axis.args=list(cex.axis=0.8))
+legend(148.7,-27.9,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=1)
+  plot(e222,xlab="",ylab="") #,border="white")
+  image(dist_Cp30_nsw,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
+  plot(aus_reg2,border="grey",add=T)
+  plot(mdb_reg2,border="grey",lwd=3,add=T)
+  plot(e222,add=T)
+  text(147.1, -34.96, "(b)", font=2)
+  dev.off()
+# 
+# 
+## plot gam.check
+  pdf("qld_check.pdf")
+  gam.check(coef.rain30.qld4[[1]])
+  dev.off()
+
+  pdf("nswvic_check.pdf")
+  gam.check(coef.rain30.nswvic[[which(values(dist_Cp30_nsw)<0.05)[6]]])
+  dev.off()
+
+@
+
 \subsection{Step trend test}
@@
 The  test, described in the following section, associates changes in rainfall with the tree cover changes.
+
+%chunk 5: note code has changed
+<<chunk5,eval=F , echo = F, results = hide>>=
+
+## without land cover variable for step rend test
+rain.trend3 <- function(x,y,m,var.data) {
+
+  var.data$rain <- as.vector(unlist(y[x]))+0.0001
+
+  if (!is.na(mean(var.data$rain))) {
+    if (grepl("qld",substitute(y))) {
+      rain_fit <- gam(rain ~ trd  # 
+                  + s(month,k=3)
+                  + s(SOI, k=3)
+                  , data=var.data,family=Gamma(link=log))
+    } else
+    if (grepl("nsw",substitute(y))) {
+      rain_fit <- gam(rain ~ trd
+                  + s(month,k=3)
+                  + s(SOI, k=3)
+                  + s(IOD, k=3) 
+                  , data=var.data,family=Gamma(link=log))
+    }
+
+  return(rain_fit)
+  } else {NA}
+
+}
+# 
+# 
+## run model for 30 years
+  coef.rain.qld <- lapply(2:ncol(qld_30), 
+                function(x) rain.trend3(x,qld_30,30,var.data))
+  coef.rain.nswvic <- lapply(2:ncol(nsw_30), 
+                function(x) rain.trend3(x,nsw_30,30,var.data))
+
+## explaining power of the partial model
+## QLD
+  rsq_qld <- unlist(lapply(1:length(coef.rain.qld), 
+          function(x) ifelse(summary(coef.rain.qld)[x,2]=="gam",
+                             summary(coef.rain.qld[[x]])$r.sq,
+                             NA)))
+# 
+## extract residual
+  residual_qld <- lapply(1:length(coef.rain.qld), 
+            function(x) if(summary(coef.rain.qld)[x,2]=="gam") {
+              coef.rain.qld[[x]]$y-fitted(coef.rain.qld[[x]])
+              } else {NA})
+
+# 
+# ## NSW/VIC
+# 
+  rsq_nswvic <- unlist(lapply(1:length(coef.rain.nswvic), 
+        function(x) ifelse(summary(coef.rain.nswvic)[x,2]=="gam",
+                           summary(coef.rain.nswvic[[x]])$r.sq,
+                           NA)))
+#
+
+# extract residual
+residual_nswvic <- lapply(1:length(coef.rain.nswvic), 
+        function(x) if(summary(coef.rain.nswvic)[x,2]=="gam") {
+          coef.rain.nswvic[[x]]$y-fitted(coef.rain.nswvic[[x]])
+          } else {NA})
+saveRDS(residual_nswvic, file="residual_nswvic.RDS")
+saveRDS(residual_qld, file="residual_qld.RDS")
+
+## explaining power of the partial model
+rain.trend4 <- function(x,y,m,var.data) {
+
+  var.data$rain <- as.vector(unlist(y[x]))+0.0001
+
+  if (!is.na(mean(var.data$rain))) {
+    if (grepl("qld",substitute(y))) {
+      rain_fit <- gam(rain ~ s(SOI, k=3)
+                  , data=var.data,family=Gamma(link=log))
+    } else
+    if (grepl("nsw",substitute(y))) {
+      rain_fit <- gam(rain ~ s(SOI, k=3)
+                  + s(IOD, k=3) 
+                  , data=var.data,family=Gamma(link=log))
+    }
+
+  return(rain_fit)
+  } else {NA}
+
+}
+
+## explaining power of index
+  coef.soi.qld <- lapply(2:ncol(qld_30), 
+                function(x) rain.trend4(x,qld_30,30,var.data))
+  coef.soi.nswvic <- lapply(2:ncol(nsw_30), 
+                  function(x) rain.trend4(x,nsw_30,30,var.data))
+  rsq_soi_qld <- unlist(lapply(1:length(coef.soi.qld), 
+          function(x) ifelse(summary(coef.soi.qld)[x,2]=="gam",
+                             summary(coef.soi.qld[[x]])$r.sq,NA)))
+  rsq_soi_nswvic <- unlist(lapply(1:length(coef.soi.nswvic),
+          function(x) ifelse(summary(coef.soi.nswvic)[x,2]=="gam",
+                             summary(coef.soi.nswvic[[x]])$r.sq,
+                             NA)))
+
+pdf("soi.pdf", width = 10, height = 4)
+par(mfrow = c(1,2))
+ hist(rsq_soi_qld*100,main="",xlab=expression(paste("% of variability explained (",R^2,")")),cex.lab=1.2)
+ text(4.2,400, "(a)", font=2)
+ hist(rsq_soi_nswvic*100,main="",xlab=expression(paste("% of variability explained (",R^2,")")),cex.lab=1.2)
+text(3.2,300, "(b)", font=2)
+dev.off()
+
+@
 
@@
 
+%chunk 6
+<<chunk6,eval=F , echo = F, results = hide>>=
+
+## perform step test on the data
+## function to calculate the step trend statistic (Hirsch, 1985)
+stepTrend2 <- function(x,y,i){  
+    n1 <- 2002-1997
+    if (i==1){
+      z <- 56  # no. of column
+    } else {
+    z <- 37 } # close if
+
+  if (x%%z!=0){ #& x%%z!=(z-1)) {
+    y.1 <- y[[x]]
+    y.2 <- y[[x+1]]
+    y.3 <- y[[x+z]]
+    y.4 <- y[[x+z+1]]
+
+  if(!is.na(y.1[1])&!is.na(y.2[1])&!is.na(y.3[1])&!is.na(y.4[1])) {
+    if (i == 1) {
+      n2 <- 2008-2004    
+      v.1 <- ts(y.1[c(229:288,313:360)],start=c(1998,1),
+                frequency=12)
+      v.2 <- ts(y.2[c(229:288,313:360)],start=c(1998,1),
+                frequency=12)
+      v.3 <- ts(y.3[c(229:288,313:360)],start=c(1998,1),
+                frequency=12)
+      v.4 <- ts(y.4[c(229:288,313:360)],start=c(1998,1),
+                frequency=12)
+    } else {
+      n2 <- 2008-2002
+      v.1 <- ts(y.1[c(229:360)],start=c(1998,1),frequency=12)
+      v.2 <- ts(y.2[c(229:360)],start=c(1998,1),frequency=12)
+      v.3 <- ts(y.3[c(229:360)],start=c(1998,1),frequency=12)
+      v.4 <- ts(y.4[c(229:360)],start=c(1998,1),frequency=12)
+    } # close if (i ==1)
+  #browser()
+## rank by month
+   for (j in 1:12) { 
+    assign(paste("r1",j,sep=""),rank(subset(v.1,cycle(v.1)==j)))
+    assign(paste("r2",j,sep=""),rank(subset(v.2,cycle(v.2)==j)))
+    assign(paste("r3",j,sep=""),rank(subset(v.3,cycle(v.3)==j)))
+    assign(paste("r4",j,sep=""),rank(subset(v.4,cycle(v.4)==j))) 
+   } # close rank by month
+## mann-whitney rank sum statistic
+    rlist.1 <- list(r11,r12,r13,r14,r15,r16,r17,r18,r19,r110,
+                    r111,r112) ## full year
+    rlist.2 <- list(r21,r22,r23,r24,r25,r26,r27,r28,r29,r210,
+                    r211,r212)
+    rlist.3 <- list(r31,r32,r33,r34,r35,r36,r37,r38,r39,r310,
+                    r311,r312)
+    rlist.4 <- list(r41,r42,r43,r44,r45,r46,r47,r48,r49,r410,
+                    r411,r412)
+    rlist <- list(unlist(rlist.1),unlist(rlist.2),unlist(rlist.3),unlist(rlist.4)) 
+    w.s1 <- 0
+    w.s2 <- 0
+    w.s3 <- 0
+    w.s4 <- 0
+  for (k in 1:length(rlist.1)) {
+    w.s1 <- sum(rlist.1[[k]][1:n1])+w.s1
+    w.s2 <- sum(rlist.2[[k]][1:n1])+w.s2
+    w.s3 <- sum(rlist.3[[k]][1:n1])+w.s3
+    w.s4 <- sum(rlist.4[[k]][1:n1])+w.s4
+    }
+    w.s <- sum(w.s1,w.s2,w.s3,w.s4) #,w.s5,w.s6,w.s7,w.s8,w.s9)
+
+## other statistic
+# mean for full year
+    mu <- n1*(n1+n2+1)/2
+    mu.s <- 12*4*mu
+
+# assume only cross correlation
+    ss <- n1*n2*(n1+n2+1)/12
+    ss.s <- 0
+    for (kk in 1:4) {
+      for (kkk in 1:4) {
+          ss.s <- 12*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
+      }
+    }
+## step trend statistic
+    z <- (w.s - mu.s)/sqrt(ss.s)
+    return(z) }
+     else {return(NA)} }
+     else {return(NA)}
+  }
+
+## 30 years
+## full year, 
+## central QLD
+  qld.z.new <- unlist(lapply(1:(length(residual_qld)-56),
+                    function(x) stepTrend2(x,residual_qld,1)))
+  qld.z.new <- c(qld.z.new,rep(NA,56))
+
+  tr_qld.step.new <- raster(qld_tif)
+  tr_qld.step.new[] <- qld.z.new
+## NSW  
+  nswvic.z.new <- unlist(lapply(1:(length(residual_nswvic)-37),
+                  function(x) stepTrend2(x,residual_nswvic,2)))
+
+  nswvic.z.new <- c(nswvic.z.new,rep(NA,37))
+
+  tr_nswvic.step.new <- raster(nsw_tif)
+  tr_nswvic.step.new[] <- nswvic.z.new
+
+  plotclt <- rev(brewer.pal(8,"RdYlBu"))
+
+pdf("step_new.pdf", width=10, height = 4.5)
+  par(mar=c(3,3,1,7),xpd=T, mfrow=c(1,2))
+  plot(e111,xlab="",ylab="")#,border="white")
+  image(tr_qld.step.new,col=plotclt,breaks=c(-3,-1.645,-1.282,-0.842,0,0.842,1.282,1.645,3),legend=F,add=T)
+  plot(mdb_reg2,border="grey",add=T,lwd=3) 
+  plot(e111,add=T)
+image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,0,1,2,3,4),lab.breaks=c(0.00,0.05,0.1,0.2,1,0.2,0.1,0.05,0.00),legend.only=T,horizontal=F,col=plotclt,legend.shrink=0.7,legend.width=0.9,legend.args=list(text=c(expression(paste("level of significance (",alpha,")")),"positive Z'","negative Z'"),side=c(2,3,1),line=0.5,cex=1,font=2),axis.args=list(cex.axis=0.8))
+text(146, -25.4, "(a)", font=2)
+  plot(e222,xlab="",ylab="")#,border="white")
+  image(tr_nswvic.step.new,col=plotclt,breaks=c(-2.3,-1.645,-1.282,-0.842,0,0.842,1.282,1.645,3),legend=F,add=T)
+  plot(mdb_reg2,border="grey",add=T,lwd=3)
+  plot(aus_reg2,border="grey",add=T) 
+  plot(e222,add=T)
+# image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,0,1,2,3,4),lab.breaks=c(0.00,0.05,0.1,0.2,1,0.2,0.1,0.05,0.00),legend.only=T,horizontal=F,col=plotclt,legend.shrink=0.7,legend.width=0.9,legend.args=list(text=c(expression(paste("level of significance (",alpha,")")),"positive Z'","negative Z'"),side=c(2,3,1),line=0.5,cex=1,font=2),axis.args=list(cex.axis=0.8))
+text(147.1, -34.96, "(b)", font=2)
+legend(149,-36.0,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=0.9)
+  dev.off()
+
+@
+
+% chunk 8: bootstrapping (spatial structure is kept, used)
+<<chunk8,eval = F , echo = F, results = hide>>=
+
+## multiple cell step trend
+## perform step test on the data for the bootstrapping test (spatial structure was kept)
+## function to calculate the step trend statistic (Hirsch, 1985)
+NstepTrend <- function(x,m,i){
+
+    n1 <- 2002-1997
+
+    if (i==1){
+
+      z <- 56  # no. of column
+
+    } else {
+
+    z <- 37 }
+#browser()
+   # print(x)
+if (x%%z!=0){ #& x%%z!=(z-1)) {
+    y.1 <- m[,x]
+    y.2 <- m[,(x+1)]
+#     y.3 <- y[,x+2]
+    y.3 <- m[,(x+z)]
+    y.4 <- m[,(x+z+1)]
+#     y.6 <- y[[x+z+2]]
+#     y.7 <- y[[x+2*z]]
+#     y.8 <- y[[x+2*z+1]]
+#     y.9 <- y[[x+2*z+2]]
+
+    if(!is.na(y.1[1])&!is.na(y.2[1])&!is.na(y.3[1])&!is.na(y.4[1])) { #|!is.na(y.5[1])|!is.na(y.6[1])|!is.na(y.7[1])|!is.na(y.8[1])|!is.na(y.9[1])) {
+
+    if (i == 1) {
+      n2 <- 2008-2004
+      v.1 <- ts(y.1[c(229:288,313:360)],start=c(1998,1),frequency=12)
+      v.2 <- ts(y.2[c(229:288,313:360)],start=c(1998,1),frequency=12)
+      v.3 <- ts(y.3[c(229:288,313:360)],start=c(1998,1),frequency=12)
+      v.4 <- ts(y.4[c(229:288,313:360)],start=c(1998,1),frequency=12)
+#       v.5 <- ts(y.5[c(229:288,313:348)],start=c(1998,1),frequency=12)
+#       v.6 <- ts(y.6[c(229:288,313:348)],start=c(1998,1),frequency=12)
+#       v.7 <- ts(y.7[c(229:288,313:348)],start=c(1998,1),frequency=12)
+#       v.8 <- ts(y.8[c(229:288,313:348)],start=c(1998,1),frequency=12)
+#       v.9 <- ts(y.9[c(229:288,313:348)],start=c(1998,1),frequency=12)
+#       v <- (x$y-x$fitted.values)[c(1:288,313:357)]
+#       v.s <- ts(v,start=c(1998,1),frequency=12)
+
+    } else {
+
+      n2 <- 2008-2002
+
+      v.1 <- ts(y.1[c(229:360)],start=c(1998,1),frequency=12)
+      v.2 <- ts(y.2[c(229:360)],start=c(1998,1),frequency=12)
+      v.3 <- ts(y.3[c(229:360)],start=c(1998,1),frequency=12)
+      v.4 <- ts(y.4[c(229:360)],start=c(1998,1),frequency=12)
+#       v.5 <- ts(y.5[c(229:348)],start=c(1998,1),frequency=12)
+#       v.6 <- ts(y.6[c(229:348)],start=c(1998,1),frequency=12)
+#       v.7 <- ts(y.7[c(229:348)],start=c(1998,1),frequency=12)
+#       v.8 <- ts(y.8[c(229:348)],start=c(1998,1),frequency=12)
+#       v.9 <- ts(y.9[c(229:348)],start=c(1998,1),frequency=12)
+#       v <- (x$y-x$fitted.values)[c(1:288, 301:357)]
+#       v.s <- ts(v,start=c(1998,1),frequency=12)
+}
+
+## rank by month
+   for (j in 1:12) {
+    # y1.mth <- data.frame(mth=as.vector(window(v.1,start=c(1998,i),frequency=1)),o=y[[j]])[order(y[[j]]),]
+    # y2.mth <- data.frame(mth=as.vector(window(v.2,start=c(1998,i),frequency=1)),o=y[[j]])[order(y[[j]]),]
+    # y3.mth <- data.frame(mth=as.vector(window(v.3,start=c(1998,i),frequency=1)),o=y[[j]])[order(y[[j]]),]
+    # y4.mth <- data.frame(mth=as.vector(window(v.4,start=c(1998,i),frequency=1)),o=y[[j]])[order(y[[j]]),]
+
+    assign(paste("r1",j,sep=""),rank(subset(v.1,cycle(v.1)==j)))
+    assign(paste("r2",j,sep=""),rank(subset(v.2,cycle(v.2)==j)))
+    assign(paste("r3",j,sep=""),rank(subset(v.3,cycle(v.3)==j)))
+    assign(paste("r4",j,sep=""),rank(subset(v.4,cycle(v.4)==j)))
+#     assign(paste("r5",j,sep=""),rank(subset(v.5,cycle(v.5)==j)))
+#     assign(paste("r6",j,sep=""),rank(subset(v.6,cycle(v.6)==j)))
+#     assign(paste("r7",j,sep=""),rank(subset(v.7,cycle(v.7)==j)))
+#     assign(paste("r8",j,sep=""),rank(subset(v.8,cycle(v.8)==j)))
+#     assign(paste("r9",j,sep=""),rank(subset(v.9,cycle(v.9)==j)))
+
+    }
+
+## mann-whitney rank sum statistic
+    rlist.1 <- list(r11,r12,r13,r14,r15,r16,r17,r18,r19,r110,r111,r112) ## full year
+    rlist.2 <- list(r21,r22,r23,r24,r25,r26,r27,r28,r29,r210,r211,r212)
+    rlist.3 <- list(r31,r32,r33,r34,r35,r36,r37,r38,r39,r310,r311,r312)
+    rlist.4 <- list(r41,r42,r43,r44,r45,r46,r47,r48,r49,r410,r411,r412)
+#     rlist.5 <- list(r51,r52,r53,r54,r55,r56,r57,r58,r59,r510,r511,r512)
+#     rlist.6 <- list(r61,r62,r63,r64,r65,r66,r67,r68,r69,r610,r611,r612)
+#     rlist.7 <- list(r71,r72,r73,r74,r75,r76,r77,r78,r79,r710,r711,r712)
+#     rlist.8 <- list(r81,r82,r83,r84,r85,r86,r87,r88,r89,r810,r811,r812)
+#     rlist.9 <- list(r91,r92,r93,r94,r95,r96,r97,r98,r99,r910,r911,r912)
+    rlist <- list(unlist(rlist.1),unlist(rlist.2),unlist(rlist.3),unlist(rlist.4)) #,unlist(rlist.5),unlist(rlist.6),unlist(rlist.7),unlist(rlist.8),unlist(rlist.9))
+## seasonal
+#     rlist <- list(unlist(rlist.1[c(9:11)]),unlist(rlist.2[c(9:11)]),unlist(rlist.3[c(9:11)]),unlist(rlist.4[c(9:11)]))
+#     rlist <- list(r1,r2,r12) ## summer
+#     rlist <- list(r3,r4,r5)  ## autumn
+#     rlist <- list(r6,r7,r8)  ## winter
+#     rlist <- list(r9,r10,r11) ## spring
+    w.s1 <- 0
+    w.s2 <- 0
+    w.s3 <- 0
+    w.s4 <- 0
+#     w.s5 <- 0
+#     w.s6 <- 0
+#     w.s7 <- 0
+#     w.s8 <- 0
+#     w.s9 <- 0
+  for (k in 1:length(rlist.1)) {
+#   for (k in c(9:11)) {
+    w.s1 <- sum(rlist.1[[k]][1:n1])+w.s1
+    w.s2 <- sum(rlist.2[[k]][1:n1])+w.s2
+    w.s3 <- sum(rlist.3[[k]][1:n1])+w.s3
+    w.s4 <- sum(rlist.4[[k]][1:n1])+w.s4
+#     w.s5 <- sum(rlist.5[[k]][1:n1])+w.s5
+#     w.s6 <- sum(rlist.6[[k]][1:n1])+w.s6
+#     w.s7 <- sum(rlist.7[[k]][1:n1])+w.s7
+#     w.s8 <- sum(rlist.8[[k]][1:n1])+w.s8
+#     w.s9 <- sum(rlist.9[[k]][1:n1])+w.s9
+    }
+    w.s <- sum(w.s1,w.s2,w.s3,w.s4) #,w.s5,w.s6,w.s7,w.s8,w.s9)
+
+## other statistic
+# mean for full year
+    mu <- n1*(n1+n2+1)/2
+    mu.s <- 12*4*mu
+# ## mean for one season
+#     mu <- n1*(n1+n2+1)/2
+#     mu.s <- 3*4*mu
+
+## variance (standard variation).
+The autocorrelation of the residuals can be assumed to be null (checked, cor.test(), true for NSW, mostly true for WA.
+QLD?
+there are some situations in WA require further investigation)
+# assume only cross correlation
+    ss <- n1*n2*(n1+n2+1)/12
+    ss.s <- 0
+    for (kk in 1:4) {
+      for (kkk in 1:4) {
+#         for (kkkk in 1:12) {
+          ss.s <- 12*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
+#         }
+      }
+    }
+## variance for one season
+#     ss <- n1*n2*(n1+n2+1)/3
+#     ss.s <- 0
+#     for (kk in 1:4) {
+#       for (kkk in 1:4) {
+#           ss.s <- 3*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
+#       }
+#     }
+
+## step trend statistic
+    z <- (w.s - mu.s)/sqrt(ss.s)
+
+    return(z) }
+
+     else {return(NA)} }
+
+     else {return(NA)}
+
+  }
+
+## full year
+residual_nswvic <- readRDS(file="residual_nswvic.RDS")
+residual_qld <- readRDS(file="residual_qld.RDS")
+
+## central QLD
+  data_df <-  as.data.frame(t(do.call(rbind, residual_qld)))
+  data_df$years <- rep(1979:2008,each=12)
+
+qld.z <- sapply(1:(ncol(data_df)-58),
+                    function(x) NstepTrend(x,data_df,1))
+## NSW/VIC
+## treatment 1: excluding 2003
+  data_df <-  as.data.frame(t(do.call(rbind, residual_nswvic)))
+  data_df$years <- rep(1979:2008,each=12)
+  nswvic.z <- sapply(1:(ncol(data_df)-39),
+                    function(x) NstepTrend(x,data_df,2))
+
+  qld_origp <- length(qld.z[qld.z >= 
+                      abs(qt(0.1,length(qld.z)))])/length(qld.z)
+# 0.01
+  qld_orign <- length(qld.z[qld.z <= 
+                      qt(0.1,length(qld.z))])/length(qld.z)
+# 0
+
+  nsw_origp <- length(nswvic.z[nswvic.z >= 
+                   abs(qt(0.1,length(nswvic.z)))])/length(nswvic.z)
+# 0.022
+  nsw_orign <- length(nswvic.z[nswvic.z <= 
+                    qt(0.1,length(nswvic.z))])/length(nswvic.z)
+
+# Bootstrapping new (September 2017)  
+# 1000 bootstrap samples, time is in the vector (will become column), xy locations are the list elements (will become rows)
+
+require(doParallel)
+# parallelise  
+Boot_strap_fun <- function(DataList, k, Site) {
+  # k is number of samples
+  # Site is location 1 = qld, 2 = nsw_vic
+  data_df <-  as.data.frame(t(do.call(rbind, DataList)))
+  data_df$years <- rep(1979:2008,each=12)
+  # storage
+  nyears <- nrow(data_df)/12
+#  browser()
+  # run the bootstrap
+  cl <- makeCluster(6) # create a cluster with 6 cores
+  registerDoParallel(cl) # register the cluster
+  # use a foreach loop to calibrate
+  out <- foreach(i = 1:k, .export=c("NstepTrend")) %dopar% {
+
+ # for (i in 1:k) {
+      mangle_years <- sample(unique(data_df$years),nyears)
+    for (j in 1:length(mangle_years)) {
+      if (j==1) { 
+        new_df <- as.data.frame(data_df[data_df$years==mangle_years[j],])
+      } else {
+    # rbind to dataframe
+        new_df <- rbind(new_df,
+                       as.data.frame(data_df[data_df$years==mangle_years[j],]))
+      }
+    }
+    new_df$years <- rep(1979:2008,each=12)
+  # run step_trend on block of cells
+  #  browser()
+    z_score <- sapply(1:(ncol(new_df)-ifelse(Site==1,58,39)),
+                    function(x) NstepTrend(x,new_df,i=Site))
+    z_score
+    }
+
+  return(out)
+}
+system.time({
+qld_z_bs <- Boot_strap_fun(residual_qld,k=1000, Site = 1)
+})
+nswvic_z_bs <- Boot_strap_fun(residual_nswvic,k=1000, Site = 2)
+
+zpq.values <- unlist(lapply(qld_z_bs,function(x)
+  length(x[x >= abs(qt(0.1,length(x)))])/length(x)*100))
+znq.values <- unlist(lapply(qld_z_bs,function(x)
+  length(x[x <= qt(0.1,length(x))])/length(x)*100))
+zpnsw.values <- unlist(lapply(nswvic_z_bs,function(x)
+  length(x[x >= abs(qt(0.1,length(x)))])/length(x)*100))
+znnsw.values <- unlist(lapply(nswvic_z_bs,function(x)
+  length(x[x <= qt(0.1,length(x))])/length(x)*100))
+
+write.table(zpq.values,file="bt_outp_qld.txt",row.names=F)
+write.table(znq.values,file="bt_outn_qld.txt",row.names=F)
+write.table(zpnsw.values,file="bt_outp_nsw.txt",row.names=F)
+write.table(znnsw.values,file="bt_outn_nsw.txt",row.names=F)
+
+# 
+bt4_nswp.values <- read.table("bt_outp_nsw.txt",header=T,sep="")
+bt4_nswn.values <- read.table("bt_outn_nsw.txt",header=T,sep="")
+bt4_qldp.values <- read.table("bt_outp_qld.txt",header=T,sep="")
+bt4_qldn.values <- read.table("bt_outn_qld.txt",header=T,sep="")
+
+pdf("bt4_nsw.pdf",width=8,height=4)
+par(mfrow=c(1,2))
+hist(bt4_nswp.values[,1],nclass=100,xlab="% positive Z'",main="4-pixel step trend",cex.lab=1.4,cex.main=1.3)
+points(nsw_origp*100,0,pch=16,col="red",cex=1.8)
+hist(bt4_nswn.values[,1],nclass=100,xlab="% negative Z'",main="4-pixel step trend",cex.lab=1.4,cex.main=1.3)
+points(nsw_orign*100,0,pch=16,col="red",cex=1.8)
+dev.off()
+
+# pdf("bt4_nswn.pdf")
+# dev.off()
+
+pdf("bt4_qld.pdf",width=8,height=4)
+par(mfrow=c(1,2))
+hist(bt4_qldp.values[,1],nclass=100,xlab="% positive Z'",main="4-pixel step trend",cex.lab=1.4,cex.main=1.3)
+points(qld_origp*100,0,pch=16,col="red",cex=1.8)
+hist(bt4_qldn.values[,1],nclass=100,xlab="% negative Z'",main="4-pixel step trend",cex.lab=1.4,cex.main=1.3)
+points(qld_orign*100,0,pch=16,col="red",cex=1.8)
+dev.off()
+
+# pdf("bt4_qldn.pdf")
+# dev.off()
+
+@
+
+% chunk 9: bootstrapping (spatial structure is kept, remove short term trend, randomly select after change period)
+<<chunk9,eval=F , echo = F, results = hide>>=
+
+## x: rainfall series; y: new order; ii: selection of region
+# NstepTrend3 <- function(x,y,ii){  
+# 
+#   if(!is.na(unlist(x)[1])) {
+#   
+#   v <- x[c(229:360)][order(y)]
+#   ttrd <- c(1:132)
+#   v <- residuals(lm(v~ttrd))
+#   
+#   n1 <- sample(c(4:8),1)
+#   
+#   if(ii==1) {    
+#       yy <- ts(v[c(1:60,85:132)],start=c(1998,1),frequency=12)    
+#       n2 <- 9-n1
+#     } else
+#       if(ii==2) {
+#       yy <- ts(v,start=c(1998,1),frequency=12)
+#       n2 <- 11-n1
+#     } 
+#   
+#   for(i in 1:12) { 
+#   y.mth <- as.vector(window(yy,start=c(1998,i),frequency=1))
+#   
+#   ## rank by month   
+#     assign(paste("r",i,sep=""),rank(y.mth) )
+#   } 
+#   
+#   
+# ## mann-whitney rank sum statistic
+#     rlist <- list(r1,r2,r3,r4,r5,r6,r7,r8,r9,r10,r11,r12) ## full year
+# 
+#     w.s <- 0
+#     for (k in 1:length(rlist)) {
+#     w.s <- sum(rlist[[k]][1:n1])+w.s
+#     }
+#   
+# ## other statistic
+# ## mean for full year
+#     mu <- n1*(n1+n2+1)/2
+#     mu.s <- length(rlist)*mu
+# 
+#   
+# ## variance (standard variation). The autocorrelation of the residuals can be assumed to be null (checked, true for QLD and NSW)
+#     ss <- n1*n2*(n1+n2+1)/length(rlist)
+#     ss.s <- length(rlist)*ss
+# 
+# ## step trend statistic
+#     z <- (w.s - mu.s)/sqrt(ss.s)
+#   
+#     return(z) 
+#     }
+#     
+#      else {return(NA)}
+#   
+#   }
+# 
+# R <- 1000
+# z3pq.values = numeric(R)
+# z3nq.values = numeric(R)
+# ## random sample with replacement. Both the order of the months and the order of years change
+# 
+# ptm <- proc.time()
+# for (i in 1:R) {
+#   
+#   new.order <- sample(c(1:132),replace=T)
+#   temp <- unlist(lapply(residual_qld,function(x,y) NstepTrend3(x,y,1),new.order))
+# #   if(i==1) {write.table(temp,file="bt_zvalues3_qld.txt",row.names=F)
+# #             } else
+# #               {write.table(temp,file="bt_zvalues3_qld.txt",append=T,row.names=F,col.names=F)}
+#   z3pq.values[i] <- length(subset(temp,temp>=1.645))/length(temp)
+#   z3nq.values[i] <- length(subset(temp,temp<=-1.645))/length(temp)
+# }
+# proc.time()-ptm
+# 
+# write.table(z3pq.values,file="bt_out3p_qld.txt",row.names=F)
+# write.table(z3nq.values,file="bt_out3n_qld.txt",row.names=F)
+# 
+# #NSW
+# z3pn.values = numeric(R)
+# z3nn.values = numeric(R)
+# ## random sample with replacement. Both the order of the months and the order of years change
+# 
+# ptm <- proc.time()
+# for (i in 1:R) {
+#   
+#   new.order <- sample(c(1:132),replace=T)
+#   temp <- unlist(lapply(residual_nswvic,function(x,y) NstepTrend3(x,y,2),new.order))
+# #   if(i==1) {write.table(temp,file="bt_zvalues3_qld.txt",row.names=F)
+# #             } else
+# #               {write.table(temp,file="bt_zvalues3_qld.txt",append=T,row.names=F,col.names=F)}
+#   z3pn.values[i] <- length(subset(temp,temp>=1.645))/length(temp)
+#   z3nn.values[i] <- length(subset(temp,temp<=-1.645))/length(temp)
+# }
+# proc.time()-ptm
+# 
+# write.table(z3pn.values,file="bt_out3p_nsw.txt",row.names=F)
+# write.table(z3nn.values,file="bt_out3n_nsw.txt",row.names=F)
+# 
+# setEPS()
+# postscript("bt3_qldp.eps")
+# hist(z3pq.values,nclass=100,xlab="% positive Z'",main="QLD",cex.lab=1.4,cex.main=1.3)
+# points(qld_origp,0,pch=16,col="red",cex=1.8)
+# dev.off()
+# 
+# setEPS()
+# postscript("bt3_qldn.eps")
+# hist(z3nq.values,nclass=100,xlab="% negative Z'",main="QLD",cex.lab=1.4,cex.main=1.3)
+# points(qld_orign,0,pch=16,col="red",cex=1.8)
+# dev.off()
+# 
+# setEPS()
+# postscript("bt3_nswp.eps")
+# hist(z3pn.values,nclass=100,xlab="% positive Z'",main="NSW",cex.lab=1.4,cex.main=1.3)
+# points(nsw_origp,0,pch=16,col="red",cex=1.8)
+# dev.off()
+# 
+# setEPS()
+# postscript("bt3_nswn.eps")
+# hist(z3nn.values,nclass=100,xlab="% negative Z'",main="NSW",cex.lab=1.4,cex.main=1.3)
+# points(nsw_orign,0,pch=16,col="red",cex=1.8)
+# dev.off()
+
+@
+% apply the LocSig function in SpatialVx for block bootstrapping
+<<chunk91,eval=F , echo = F, results = hide>>=
+
+# require(SpatialVx)
+# 
+# ## reconstruct the data into format required in LocSig
+# qld_res_mat <- matrix(unlist(residual_qld),nrow=360)
+# nsw_res_mat <- matrix(unlist(residual_nswvic),nrow=360)
+# 
+# ## the step trend function for LocSig
+# NstepTrend4 <- function(x,ii){  
+#   
+#   n1 <- sample(c(4:8),1)
+#   zz <- vector()
+# 
+#   for(m in 1:ncol(x)) {
+#     if(!is.na(x[1,m])) {
+#   
+#   v <- x[c(229:360),m]
+#   ttrd <- c(1:132)
+#   v <- residuals(lm(v~ttrd))
+#   
+#   if(ii==1) {    
+#       yy <- ts(v[c(1:60,85:132)],start=c(1998,1),frequency=12)    
+#       n2 <- 9-n1
+#     } else
+#       if(ii==2) {
+#       yy <- ts(v,start=c(1998,1),frequency=12)
+#       n2 <- 11-n1
+#     } 
+#   
+#   for(i in 1:12) { 
+#   y.mth <- as.vector(window(yy,start=c(1998,i),frequency=1))
+#   
+#   ## rank by month   
+#     assign(paste("r",i,sep=""),rank(y.mth) )
+#   } 
+#   
+#   
+# ## mann-whitney rank sum statistic
+#     rlist <- list(r1,r2,r3,r4,r5,r6,r7,r8,r9,r10,r11,r12) ## full year
+# 
+#     w.s <- 0
+#     for (k in 1:length(rlist)) {
+#     w.s <- sum(rlist[[k]][1:n1])+w.s
+#     }
+#   
+# ## other statistic
+# ## mean for full year
+#     mu <- n1*(n1+n2+1)/2
+#     mu.s <- length(rlist)*mu
+# 
+#   
+# ## variance (standard variation). The autocorrelation of the residuals can be assumed to be null (checked, true for QLD and NSW)
+#     ss <- n1*n2*(n1+n2+1)/length(rlist)
+#     ss.s <- length(rlist)*ss
+# 
+# ## step trend statistic
+#     z <- (w.s - mu.s)/sqrt(ss.s)
+#     zz[m] <- z
+#     }
+#     
+#      else {zz[m]<-NA}
+#   }
+#   return(zz)
+#   
+#   }
+@
+
 \section{Results}
@@
 
+% chunk 81
+<<chunk81,eval=F , echo = F, results = hide>>=
+
+## calculate mean rainfall difference in two period for 11-year
+
+  sel_neg.qld <- residual_qld[which(values(tr_qld.step.new)>=1.645)]
+
+  sel_neg.nswvic <- residual_nswvic[which(values(tr_nswvic.step.new)>=1.645)]
+
+## compare mean of before and after
+## boxplot
+  data.qld <- data.frame(year=seq(1998,2008,1))
+  for (i in 1:length(sel_neg.qld)) {
+    data.qld[,i+1] <- as.vector(rollapply(as.zoo(sel_neg.qld[[i]])[229:360],width=12,sum,by=12))
+  }
+  data2.qld <- melt(data.qld,id="year")
+  data2.qld$group[data2.qld$year<2003] <- "1before"
+  data2.qld$group[data2.qld$year>2004] <- "2after"
+
+  data.nswvic <- data.frame(year=seq(1998,2008,1))
+  for (i in 1:length(sel_neg.nswvic)) {
+    data.nswvic[,i+1] <- as.vector(rollapply(as.zoo(sel_neg.nswvic[[i]])[229:360],width=12,sum,by=12))
+  }
+  data2.nswvic <- melt(data.nswvic,id="year")
+  data2.nswvic$group[data2.nswvic$year<2003] <- "1before"
+  data2.nswvic$group[data2.nswvic$year>2002] <- "2after"
+
+## perform a 2 sample t test
+  var.test(data2.qld[data2.qld$year<2003,3],data2.qld[data2.qld$year>2004 & data2.qld$year!=2006,3],alternative="two.sided")
+## unequal variance
+  t.test(data2.qld[data2.qld$year<2003,3],data2.qld[data2.qld$year>2004 & data2.qld$year!=2006,3],alternative="greater")
+## p = 2.2e-16
+
+  var.test(data2.nswvic[data2.nswvic$year<2003,3],data2.nswvic[data2.nswvic$year>2002& data2.nswvic$year!=2006,3],alternative="two.sided")
+## unequal variance
+  t.test(data2.nswvic[data2.nswvic$year<2003,3],data2.nswvic[data2.nswvic$year>2002& data2.nswvic$year!=2006,3],alternative="greater")
+## p < 1.4e-13
+
+  pdf("compare_mean.pdf")
+  par(mfrow=c(1,2))
+  boxplot(value~group,data=data2.qld,names=c("before","after"),boxwex=0.4,main="QLD site",ylab="rainfall (mm)")
+
+  boxplot(value~group,data=data2.nswvic,names=c("before","after"),boxwex=0.4,main="NSW/VIC site")
+  dev.off()
+
+@
+
+% chunk 82
+<<chunk82,eval=F , echo = F, results = hide>>=
+# identify whether 2006 is significantly different in the rainfall anomalies
+require(ggplot2)
+
+residual_nswvic <- readRDS(file="residual_nswvic.RDS")
+residual_qld <- readRDS(file="residual_qld.RDS")
+
+## central QLD
+  data_df <-  as.data.frame(t(do.call(rbind, residual_qld)))
+  data_df$years <- rep(1979:2008,each=12)
+  foo <- gather(data_df, key=pixel, value = anomaly,V1:V2800)
+  qld_foo <- foo[foo$years>1996,]
+  pl_qld <- ggplot(qld_foo,aes(x=as.factor(years),y=anomaly)) + geom_boxplot()
+  pdf("QldrainfallAnomalyByYear.pdf")
+  pl_qld
+  dev.off()
+
+  mod_qld <- aov(anomaly~as.factor(years),data=qld_foo)
+  TukeyHSD(mod_qld)
+
+## NSW/VIC
+## treatment 1: excluding 2003
+  data_df <-  as.data.frame(t(do.call(rbind, residual_nswvic)))
+  data_df$years <- rep(1979:2008,each=12)
+  foo <- gather(data_df, key=pixel, value = anomaly,V1:V1184)
+  nsw_foo <- foo[foo$years>1996,]
+  pl <- ggplot(nsw_foo,aes(x=as.factor(years),y=anomaly)) + geom_boxplot()
+  pdf("NSWrainfallAnomalyByYear.pdf")
+  pl
+  dev.off()
+
+  mod_nsw <- aov(anomaly~as.factor(years),data=nsw_foo)
+  TukeyHSD(mod_nsw)
+
+@
 \section{Summary and Discussion}

### 2017-11-08 chap3_paper_new.tex
add files

--- 2017-11-08 LandSurfaceEffects_LocalRainfall_WRRsubmit.Rnw
@@
 
+\usepackage{Sweave}
 \begin{document}
-\SweaveOpts{concordance=TRUE}
+\input{chap3_paper_new-concordance}
 
@@
 
+%chunk 5: note code has changed
+
 \subsubsection{Mann-Whitney rank-sum statistic}
@@
 \end{table}
+
+%chunk 6
+
+% chunk 8: bootstrapping (spatial structure is kept, used)
+
+% chunk 9: bootstrapping (spatial structure is kept, remove short term trend, randomly select after change period)
+% apply the LocSig function in SpatialVx for block bootstrapping
 
@@
 
+% chunk 81
+
+% chunk 82
 \section{Summary and Discussion}

### 2018-09-04 Manuscripts/Rainfall_Landuse_paper/Rainfall_Landuse_paper.Rmd
changes for paper

---
title: Detecting influences of changes in land cover on observed rainfall.
author:
  - name: Chun X. Liang
    email: summer.chunliang@gmail.com 
    affiliation: a
  - name: Floris F. van Ogtrop
    email: floris.vanogtrop@sydney.edu.au 
    affiliation: a
  - name: R. Willem Vervoort
    email: willem.vervoort@sydney.edu.au
    affiliation: a
    footnote: Corresponding Author
address:
  - code: a
    address: Sydney Institute of Agriculture, The University of Sydney, NSW 2006
abstract: |
  Analysis of observational data to identify relationships between rainfall and land cover change are difficult due to multiple environmental factors that cannot be strictly controlled. In this study we present a methodology to investigate the relationship using statistical methods on data from best available sources at two sites in Australia. Gridded data of rainfall and tree cover were used as spatially corresponding local conditions. Large scale effects were represented by climatic indicators, such as SOI and IOD. Regression analysis and step trend tests were used to assess the effect of abrupt land surface intervention. 
  At a Queensland site, significant tree cover change between 2002 - 2005 did not result in strong statistically significant precipitation changes. On the other hand, results from a bushfire affected NSW/VIC region suggests significant changes in the rainfall. This indicates the method works better when a abrupt change in the data can be clearly identified. The results from the step trend test implied a positive relationship between the tree cover and the rainfall at 0.1 significance level in both locations in data up to 2009. However, high rainfall variability and possible regrowth meant that no signifcant changes were observed in alonger time sereies to 2015.

journal: "Journal of Hydrology Regional Studies"
date: "`r Sys.Date()`"
bibliography: reflist_chap3.bib
output: 
 bookdown::pdf_book:
    base_format: rticles::elsevier_article
---
```{r packages, echo=F, warnings=F, message=F, eval=T}
require(knitr)
require(bookdown)
require(pander)
library(fields)
library(graphics)
library(maptools)
#library(plotKML)
library(raster)
library(reshape)
library(rgeos)
library(rgdal)
library(RColorBrewer)
library(tidyverse)
require(lubridate)
library(deseasonalize)
library(mgcv)
library(ncdf4)
require(gridExtra)
#library(nnet)
```

```{r setup, warning=F, message=F, echo =F}
# root dir
knitr::opts_knit$set(root.dir = 
                       "C:/Users/rver4657/ownCloud/chun/chun/manuscripts/Rainfall_Landuse_paper")
```

Introduction
==========

Land use and land cover changes can lead to changes in the local climate. Empirical and modelling studies have found cloud types and rainfall are correlated to large scale vegetation cover changes, such as deforestation in the Amazon and in the Sahel [@Chagnon2005;@Pinto2009; @Wang2009; @Mei2010; @kucharski_further_2013; @pitman_scale_2016] and afforestation in south Israel [@Otterman1990; @Ben-Gai1998]. Using airborne measurement in Western Australia, @Junkermann2009 showed a significantly higher level of aerosols over an agricultural area compared to an adjacent natural vegetation. They suggested that a modification of aerosol concentrations due to deforestation could have contributed to a reduction of local rainfall, as more, but smaller rain droplets were observed. @Nair2011 reported from the Bunny Fence Experiment in Western Australia that local land use change altered the synoptic west coast trough dynamics and surface roughness, and this resulted in an observed rainfall decrease. Maximum temperatures were also found to be sensitive to land cover change in eastern Australia [@McAlpine2007].

Overall the number of empirical studies analyzing changes to rainfall due to land cover change from observational data is limited. Most of the studies mentioned previously were either model simulations, or comparisons of modelled data with observations. This is because there are some fundamental experimental difficulties in both space (where does evaporated water reappear as rainfall?) and in time (how much time does it take for land cover change effects to appear or disappear?). In addition, in many areas across the globe rainfall variability is related to complex set of interactions, of which land use change might only be a minor component.

Locally, there are two main sources that generate rainfall: moisture from advective atmospheric transport; and local evapotranspiration [@Eltahir1996; @Bosilovich2006; @Dirmeyer2009; @Gimeno2010]. The local evapotranspiration component is the component considered to be affected by land use change [@Eltahir1996]. According to @Trenberth1999, the contribution of advective moisture partially depends on the availability of external moisture and atmospheric transport. On the longer time scale, such as monthly and annually, large scale atmospheric dynamics are affected by large scale climate drivers. For example, many studies have reported significant relationships between rainfall in large parts of Australia and the El Niño-Southern Oscillation (ENSO) [@Verdon2004; @Risbey2009; @Speer2011]. In contrast, local ET is determined by local land surface characteristics, which influence local scale atmospheric dynamics and hence the amount of rainfall, including contribution from both main sources. Therefore land surface is suggested to play an important role in local rainfall.

Although climate drivers demonstrate some capability to predict Australian rainfall, there is still a large amount of unexplained variance. @Westra2010 pointed out that models based on global sea surface temperature anomalies can only predict up to 14.7\% of annual precipitation variance. Some of the remaining variance could be due to land surface processes as suggested in studies predicting local rainfall [e.g. @Ma2011; @Zeng2012; @pitman_scale_2016; @saha_investigating_2016]. However, they are mostly based on modelling experiments and little evidence was reported from observations. However, @Pitman2004 found a good match between observations and simulated rainfall changes in southwest Western Australia ,forced by land cover change. @Timbal2006 were able to reproduce the rainfall decline in south west Australia by including land cover influence. In addition, local land use change might not be a primary, but is likely to be a secondary cause of rainfall change [@Nicholls2006].

Therefore, the aim of this study is to expand the available methodology to use empirical evidence at regional scales to investigate the cause and effect relationship between land cover change and local rainfall. More specifically, we hypothesize that a step change on the land cover on the surface will cause a step change in the rainfall. To demonstrate the approach and this effect we applied the methodology to study the changes in rainfall at two locations in Queensland and NSW/Victoria where there are possible step changes in land cover change due to land clearing and bush fires. The methodology is mainly based on statistical approaches to identify changes in rainfall, which were subsequently associated with land cover change through spatial comparison.

In this paper, after this section (the introduction), section 2 covers the case study areas and the observed land use change. Section 3 describes the data used in the study in more detail. Section 4 details the statistical methods and the underlying assumptions related to the modelling approach, Section 5 gives the results, which are further discussed in section 6 and finally section 7 offers the conclusions.

Study regions and tree cover change
===================================

We use two areas in Australia to demonstrate the approach. In Australia, significant tree cover change has mainly occurred in the north east of the continent and on the southeast coast, as well as in the southwest of Western Australia. According to the National Dynamic Land Cover Dataset (DLCD) [@Lymburner2010], most of these areas have experienced decreasing EVI post 2000. As an index for vegetation greenness, the decreasing values indicate lower biomass over time in the tree cover regions. The possible EVI reduction might be due to land clearing, bush fires or drought.

```{r selreg, fig.cap="Selected study regions are highlighted by red rectangles in the main map (the red rectangle in the insert indicates the location of the main map). The types of tree cover in 2008 from the DLCD product is shown at the background. In site 1 (the QLD region), the tree cover is mostly sparse. In site 2 (the NSW/VIC region), many areas have open or close forest in which tree cover is denser.", out.width="90%", echo=F}

include_graphics("figures/map_selreg.jpg")
```

Two regions were selected where significant tree cover change was reported. The first region is located in south central Queensland to the north of the Murray Darling Basin (MDB) (site 1 in Figure \@ref(fig:selreg). High rates of land clearing have been reported in this region during the early 2000s [@SLATS2004]. The second study region is located at the border of New South Wales and Victoria, and includes the Snowy Mountain ranges (site 2 in Figure \@ref(fig:selreg). Severe bush fires occurred in this area and the surroundings in early 2003 (see Figure \@ref(fig:figure2bushfire)). The 2003 bush fires were the largest and the worst in this area for the last 60 years [@Fire2011]. Two thirds of Kosciuszko national park was heavily burned and regrowth was reported to be slow due to drought and cold conditions [@ABC2003] and the type of species in this region. However, in the longer term, after an early high transpiration period a recovery of pre-fire evapotranspiration would be expected [@Kuzcera1987]. As a result , significant tree cover loss has happened in both study areas in the last decade, either permanently or temporarily.

```{r figure2bushfire, fig.cap = "Location of bushfires occurring in January 2003, in and around the NSW/VIC study region, as shown by the red pixels. The map shows large area in the Kosciuszko national park has been burned. Some locations in the southwest of ACT have also experienced intensive bushfires.", out.width="90%", echo=F}
include_graphics("figures/bushfire_nswvic.jpg")
```

The two regions have different climate characteristics. The QLD region is partially grassland and partially subtropical, while the NSW/VIC region is mainly within the temperate zone, under the Köppen classification. According to Australian Bureau of Meteorology (BoM), the NSW/VIC region receives 1000 - 2000 mm rainfall annually, which is more than double of the rainfall in the QLD region. Evapotranspiration is similar in both regions. Marine moisture and orographic effects are likely to be the main contributors to rainfall in the southeast mountain areas of the NSW/VIC region.

The land use and land cover characteristics in the two regions are also different. In the Queensland region, the tree cover is sparse over most of the area. The MODIS satellite tree cover data (discussed in more detail in section 3) shows that tree cover in this region is generally below 20\% of total ground area. Grazing is the main activity in this region, with over 90\% of land used by the grazing industry [@ABARES2010]. Our starting assumption is that the main cause of the EVI  decline over large part of the region is due to land clearing. Tree cover has been cleared at a massive scale over the last decade, especially during 2002 - 2004. The Kosciuszko national park is within the NSW/VIC region. Here tree cover is denser with open or even closed forest (the tree cover distribution is bimodal at 10 - 20\% and 60 - 70\%). The dominant species in the alpine area are Snow Gum and large stand species such as Alpine Ash and Mountain Gum in the sub-alpine area. These trees can reach a great height but they take long time to grow. For example, Alpine Ash would need about 20 years to mature. Although land clearing is not a major issue in this region, it is vulnerable to fires and drought.

Therefore two types of land cover changes were studied. The reports from the Queensland Statewide Land Cover and Trees Study (SLATS) [e.g. @SLATS2001; @SLATS2017] were used to investigate the time and location of land clearing in the QLD region. The MODIS burned area product, MCD45A1 [@Roy2002; @Roy2005; @Roy2008], was used to locate bush fires areas in the NSW/VIC region, with a grid resolution of 500 m. MCD45A1 provides monthly burning information on all pixels, which helps to pinpoint an abrupt event. Due to the nature of the different land cover change, the post-change vegetation status in the two regions is expected to be different (see Figure \@ref(fig:figure3-tc-simple)).

```{r figure3-tc-simple, fig.cap= "The expected evolution of the land surface after trees have been removed in (a) the QLD region and (b) the NSW/VIC region.", out.width="90%", echo=F}
include_graphics("figures/tc_simple.jpg")
```

```{r mapping, echo = F, eval=F}

## define the boundary of sub-regions
   e1 <- extent(c(1324304.091,1723901.773,-222556.668,80451.265))
   e2 <- extent(c(1346395.141,1534425.527,-1169402.246,-607644.171))
  # 
   e11 <- extent(c(145.425, 149.025, -28.475, -25.175))
   e22 <- extent(c(146.675, 149.325, -37.075, -34.675))

  e111 <- extent(c(145.825, 148.625, -28.075, -25.575)) 
  e222 <- extent(c(147.075, 148.925, -36.675, -35.075))

## the selected regions for mapping
root_dir <- "c:/users/rver4657/owncloud/chun/chun/data"
aus_reg <- readShapePoly(paste(root_dir,
                               "Geo\\aus_state2011\\STE11aAust.shp",sep="/"),
                         proj4string=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0"))
aus_reg2 <- gSimplify(aus_reg,0.01,topologyPreserve=T)

mdb_reg <- readShapePoly(paste(root_dir,
                               "Geo\\MDB\\mdb_boundary.shp", sep="/"),
                               proj4string=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0"))
mdb_reg2 <- gSimplify(mdb_reg,0.01,topologyPreserve=TRUE)

#remove(list=c("aus_reg","mdb_reg"))

## Burned area map

fire_0301 <- raster(paste(root_dir,
          "Other\\bushfires\\Burnt_Area\\tmpA2003001.burndate.jpg",
                          sep="/"))
fire_0301nsw <- crop(fire_0301,e2)
fire_loc <- crop(reproject(fire_0301nsw,
                           CRS=CRS("+proj=longlat +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84")),
                 e222)
remove(list=c("fire_0301","fire_0301nsw"))

  jpeg("../../figures/bushfire_nswvic.jpg", width=960, height=960)
  par(mar=c(3,3,1,6),xpd=T)
  plot(e222,border="white")  
  image(fire_loc,add=T, breaks=c(0,0.99,366.5,10000),col=c("white","red","white"))
  plot(e222,add=T)
  plot(aus_reg2,border="grey",add=T)
  plot(mdb_reg2,border="grey",lwd=3,add=T)
#   text(1430000,-1000000,"study region",cex=1.2)
  text(149.1, -35.3,"ACT",font=2,cex=1.2)
  text(148.4, -36,"Kosciuszko",font=2,cex=0.9)#,srt=90)
  text(147.5, -36.4,"VIC",font=2,cex=1.2)
  text(148, -35.5,"NSW",font=2,cex=1.2)
  image.plot(zlim=c(0,1),breaks=c(0,0.5,1),lab.breaks=c("","burned \narea",""),legend.only=T,horizontal=F,col=c("red","red"),legend.shrink=0.03,legend.width=1.3,legend.args=list(text="",line=0),lwd.poly=0.1)
legend(149,-36.1,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=0.9)
 dev.off()

Data <- read_csv(paste(root_dir,
          "/Other/slats/SLATSOverviewClearing20180814.csv",
          sep=""))

jpeg("../../figures/slats.jpg", width=1024, height = 760)
Data %>% 
  gather(key="Year",value="clearing", `1988-1989`:`2014-2015`) %>%
  filter(Bioregion == "Mulga Lands" |
           Bioregion == "Brigalow Belt") %>%
  ggplot(aes(Year,clearing, fill=Bioregion)) + 
  geom_bar(stat="identity") + ylab("Clearing (ha)") +
  scale_fill_grey() +
  theme(axis.text.x = element_text(size = rel(2),
                                   angle = 45, vjust=0.5),
        axis.text.y = element_text(size = rel(2)),
        axis.title = element_text(size = rel(2), face = "bold"),
        legend.text = element_text(size = rel(2)),
        legend.title = element_text(size = rel(2), face = "bold"))
dev.off()

```

```{r treeCover,eval=F , echo = F}

rootdir <-"C:/users/rver4657/owncloud/chun2/data/satellite/MOD44/B"

# read in the file list from the directory
x <- dir(path=rootdir,pattern="Percent_Tree_Cover.tif")
# Identify the different states in the images
# NSW
NSW <- x[grep("NSW",x)]
# Qld
Qld <- x[grep("Qld",x)]

img <- list()
# raster and brick
for(i in 1:length(NSW)) {
  img[[i]] <- raster(paste(rootdir,NSW[i],sep="/"))
}

imgQ <- list()
# raster and brick
for(i in 1:length(Qld)) {
  imgQ[[i]] <- raster(paste(rootdir,Qld[i],sep="/"))
}

NSW_b <- do.call(brick,img)
Qld_b <- do.call(brick, imgQ)

tm <- seq(as.Date("2000-03-06"), as.Date("2016-03-06"), "year")
mod_tc_r22 <- setZ(NSW_b, tm, "year")

mod_tc_r11 <- setZ(Qld_b, tm, "year")

## apply t-test to tree cover data
fun <- function(x,m) {
  var_x <- sd(x,na.rm=T)
  mu_1 <- mean(x[1:3],na.rm=T)
  # m = 1 Queensland, 
  if (m==1) {y <- x[4:16]} else {
    # NSW whole series m = 2
    if (m ==2) {y <- x[4:16]} else {
      # build in other possibilities for length series
      y <- x[4:(16 - m)]
    }  
  mu_2 <- mean(y,na.rm=T)
  z <- (mu_2-mu_1)/var_x
  if (!is.na(z) & abs(z)>1.645) {return(mu_2-mu_1)} else {return(NA)}
  }
}

# Qld
mod_tc_r1.z <- calc(mod_tc_r11,function(x) fun(x,1))
# Qld 2011
mod_tc_r1_7y.z <- calc(mod_tc_r11,function(x) fun(x,5))
# NSW
mod_tc_r2.z <- calc(mod_tc_r22,function(x) fun(x,2))
# # NSW only 5 years after
# mod_tc_r2_5y.z <- calc(mod_tc_r22,function(x) fun(x,7))
# NSW 7 years after (2011)
mod_tc_r2_7y.z <- calc(mod_tc_r22,function(x) fun(x,5))
# # NSW 9 years after
# mod_tc_r2_9y.z <- calc(mod_tc_r22,function(x) fun(x,3))

plotclr <- brewer.pal(8,"PiYG")

jpeg("../../figures/tc_figs.jpg", width=1280, height=960, quality=100)
par(mar=c(3,3,2,11),mfrow=c(2,2))
# Qld 7 year 2011
plot(e111, cex.axis=1.5, font.axis=2, ylab="")
image(mod_tc_r1_7y.z,col=plotclr,breaks=c(-70,-20,-10,-5,0,5,10,20,70),
      legend=F,add=T)
#plot(e111,add=T)
plot(aus_reg2,border="grey",add=T)
plot(mdb_reg2,border="grey",lwd=3,add=T)
legend(146,-27.5,c("State","MDB"),col=c("grey","grey"),
       lwd=c(1,3),bty="n",title="Boundary",cex=1.8)
image.plot(zlim=c(-4,4), breaks=c(-4,-3,-2,-1,-0.2,0.2,1,2,3,4),
           lab.breaks=c(-70,-20,-10,-5,"no tree","0 or",5,10,20,70),
           legend.only=T,horizontal=F,
           col=c("#C51B7D","#DE77AE","#F1B6DA","#FDE0EF","white",
                 "#E6F5D0","#B8E186","#7FBC41","#4D9221"),
           legend.shrink=0.8,
           legend.width=1.5,
           legend.args=list(text=expression(paste(Delta,
                            "tree (% ground)",sep="")),
                            line=1,cex=1.3,font=2),
           axis.args=list(cex.axis=1.5),graphics.reset=F)

par(mar=c(3,9,2,5))
plot(e111, cex.axis=1.5, font.axis=2, ylab="")
image(mod_tc_r1.z,col=plotclr,breaks=c(-70,-20,-10,-5,0,5,10,20,70),
      legend=F,add=T)
#plot(e111,add=T)
plot(aus_reg2,border="grey",add=T)
plot(mdb_reg2,border="grey",lwd=3,add=T)
# image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,-0.2,0.2,1,2,3,4),
#            lab.breaks=c(-70,-20,-10,-5,"no tree","0 or",5,10,20,70),
#            legend.only=T,horizontal=F,
#            col=c("#C51B7D","#DE77AE","#F1B6DA","#FDE0EF","white",
#                  "#E6F5D0","#B8E186","#7FBC41","#4D9221"),
#            legend.shrink=0.7,
#            legend.width=0.7,
#            legend.args=list(text=expression(paste(Delta,
#                             "tree (% ground)",sep="")),
#                             line=0.5,cex=0.9,font=2),
#            axis.args=list(cex.axis=0.8),graphics.reset=F)
# 2011 NSW
par(mar=c(3,3,1,11))
plot(e222, cex.axis=1.5, font.axis=2, ylab="")
image(mod_tc_r2_7y.z,col=plotclr,breaks=c(-70,-20,-10,-5,0,5,10,20,70),
      legend=F,add=T)
#plot(e111,add=T)
plot(aus_reg2,border="grey",add=T)
plot(mdb_reg2,border="grey",lwd=3,add=T)
image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,-0.2,0.2,1,2,3,4),
           lab.breaks=c(-70,-20,-10,-5,"no tree","0 or",5,10,20,70),
           legend.only=T,horizontal=F,
           col=c("#C51B7D","#DE77AE","#F1B6DA","#FDE0EF","white",
                 "#E6F5D0","#B8E186","#7FBC41","#4D9221"),
           legend.shrink=0.8,
           legend.width=1.5,
           legend.args=list(text=expression(paste(Delta,
                            "tree (% ground)",sep="")),
                            line=1,cex=1.3,font=2),
           axis.args=list(cex.axis=1.5),graphics.reset=F)

# FULL PERIOD NSW
par(mar=c(3,9,1,5))
plot(e222, cex.axis=1.5, font.axis=2, ylab="")
image(mod_tc_r2.z,col=plotclr,breaks=c(-70,-20,-10,-5,0,5,10,20,70),
      legend=F,add=T)
#plot(e111,add=T)
plot(aus_reg2,border="grey",add=T)
plot(mdb_reg2,border="grey",lwd=3,add=T)
# image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,-0.2,0.2,1,2,3,4),
#            lab.breaks=c(-70,-20,-10,-5,"no tree","0 or",5,10,20,70),
#            legend.only=T,horizontal=F,
#            col=c("#C51B7D","#DE77AE","#F1B6DA","#FDE0EF","white",
#                  "#E6F5D0","#B8E186","#7FBC41","#4D9221"),
#            legend.shrink=0.7,
#            legend.width=0.7,
#            legend.args=list(text=expression(paste(Delta,
#                             "tree (% ground)",sep="")),
#                             line=0.5,cex=0.9,font=2),
#            axis.args=list(cex.axis=0.8),graphics.reset=F)
dev.off()

jpeg("../../figures/map.jpeg", width=960, height=960)
par(mar=c(1.5,1.5,1.5,1.5))
plot(mdb_reg2,border="grey",lwd=3)
plot(aus_reg2,border="grey",add=T)
legend(150.2,-36,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=1.5)
#invisible(text(getSpPPolygonsLabptSlots(aus_reg2), labels=as.character(aus_reg2$NAME), cex=0.7,adj=0.5))
box(lty=1)
plot(e111,col="red",add=T)
plot(e222,col="red",add=T)
text(147.2,-26.8,"QLD",cex=1.5)
text(148,-34.7,"NSW/VIC",cex=1.5)
dev.off()

```

The hypothesis in this study therefore is that the effect of 2003 - 2004 land clearings in the QLD region and the 2003 bush fires in the NSW/VIC region cause a step change in the local rainfall. The actual tree cover change at the pixel level during this time was derived from the 15-year MODIS data (discussed below). The difference of tree cover before and after the land disturbance was tested using a Student's t-test. As the length of the tree cover data is shorter than available rainfall data, earlier land clearings in the QLD region cannot be identified spatially, hence they are excluded from the analysis.

## Data {#Data}

Several land surface data sets were used in this study. The main one was the MOD44B product Global Vegetation Continuous Field data set (version 5). This data set provides estimates of percent tree cover (percentage of ground surface covered by trees) at a grid resolution of 250 m [@Townshend2011], which is finer then the earlier mentioned burned product MCD45A1. The data set is available on an annual time interval for the study period of 2000 - 2015. The tree cover data was produced from 16-day Terra MODIS Land Surface Reflectance data and Land Surface Temperature [@Townshend2011]. The National Dynamic Land Cover Dataset (DLCD) [@Lymburner2010] from the Australian Collaborative Land Use Mapping Program (ACLUMP) was used to verify the trend of vegetation cover change calculated from the previous data set. This data set, developed by Geoscience Australia and Australian Bureau of Agricultural and Resource Economics and Sciences (ABARES), is the first nationally consistent and thematically comprehensive land cover reference for Australia. The DLCD is based on the 16-day Enhanced Vegetation Index (EVI), again from the MODIS satellite, between April 2000 and April 2015. It also has a grid resolution of 250 m. The data set provides information on the final land cover types (as in 2015) and estimated trend of EVI statistics (annual mean, maximum and minimum).

Rainfall data for Australia [@Jones2009] was obtained from the Bureau of Meteorology. The data has been projected onto a national 0.05\textdegree × 0.05\textdegree grid (approximately 5 km × 5 km). This gridded data set was generated from station observations using an optimised Barnes successive correction technique. The Barnes technique combines a weighted averaging process and defined topographical information to estimate rainfall values between spatial points [@BoM2009]. The resulting data set provides additional information for data-sparse areas like central Australia but reduces information in the data-rich areas, such as southeast Australia where station density is up to 20 per 100 $\text{km}^2$. The data is available on a monthly basis from 1900 to current. Here a subset of 36 years (1979 - 2015) was used. The study was conducted on monthly data, as a land cover change effect on annual rainfall might be negligible but can often found to be significant in particular months or seasons [e.g. @Otterman1990; @Gaertner2001; @Semazzi2001; @Oleson2004; @Deo2009].

Large scale climate drivers are represented by various climatic indices. The Southern Oscillation Index (SOI) is generally regarded as a good predictor of Australian rainfall [@Risbey2009; @Chowdhury2010; @Westra2010], but its skill is weaker in some parts of Australia. For example the Southern Annular Mode (SAM) is found to be more important than ENSO in south Western Australia [@Meneghini2007]. The suitability of each index for the regions of interest was tested in section \@ref(sec:reg_model). The following climate indices were used as candidate predictors for local rainfall.

- Southern Oscillation Index (SOI). The Troup version of the monthly SOI series used in this study was obtained from BoM (available online at \url{http://www.bom.gov.au/climate/current/soihtm1.shtml}).  
- Eastern, East Central and Central Tropical Pacific Sea Surface Temperatures (NINO 3, NINO 3.4 and NINO 4). Monthly SST anomalies are available from IRI/LDEO data library and the extended NINO data set is used (available online at \url{http://iridl.ldeo.columbia.edu/SOURCES/.Indices/.nino/.EXTENDED/}).   
- Pacific Decadal Oscillation (PDO). The Pacific Decadal Oscillation is the leading principal component of monthly SST anomaly in the North Pacific Ocean.. The monthly PDO series was provided by JISAO (Joint Institute for the Study of the Atmosphere and Ocean, University of Washington) (available online at \url{http://jisao.washington.edu/pdo/PDO.latest}).  
- Indian Ocean Dipole (IOD). The Indian Ocean dipole is commonly measured by the difference between SST anomaly in the western (50 - 70\textdegree E and 10\textdegree S-10\textdegree N) and eastern (90 - 110\textdegree E and 0 - 10\textdegree S) equatorial India Ocean [@Saji1999]. Monthly IOD was obtained from JAMSTEC (the Japan Agency for Marine-Earth Science and Technology) (available online at \url{http://www.jamstec.go.jp/frcgc/research/d1/iod/DATA/dmi.monthly.txt}).

```{r climaticIndex,eval=F , echo = F}

# climate indicator data
file1 <- "../..\\Data\\BoM\\SOI\\SOI.csv"
SOI <- read_csv(file1)
SOI <- gather(SOI,key="Month", value="SOI",Jan:Dec)
SOI <- SOI[order(as.Date(paste(SOI$Year,SOI$Month,
                                        "01",sep="-"), "%Y-%b-%d")),]
SOI_100a <- SOI %>%
  filter(Year > 1899 & Year < 2016)
SOI_30a <- SOI %>%
  filter(Year > 1978 & Year < 2016)
## deseasonalised SOI
SOI_100 <- ds(SOI_100a$SOI, ic = "AIC", type = "monthly",
              standardizeQ=F)$z
SOI_30 <- ds(SOI_30a$SOI, ic = "AIC", type = "monthly",
             standardizeQ=F)$z

file2 <- "../..\\Data\\BoM\\SST\\SST34_anomaly.nc"
SST34 <- nc_open(file2)
SST34_var <- ncvar_get(SST34,SST34$var[[1]])
nc_close(SST34)
SST34_df <- tibble(Date = seq.Date(as.Date("1856-01-01"),
                                   as.Date("2018-04-01"),
                                   by="month"),
                   SST34 = SST34_var)
SST34_100a <- SST34_df %>%
  filter(Date >= "1900-01-01" & Date < "2016-01-01")
SST34_30a <- SST34_df %>%
  filter(Date >= "1979-01-01" & Date < "2016-01-01")
## deseasonalised SST34
SST34_30 <- ds(SST34_30a$SST34,ic = "AIC", type="monthly",
               standardizeQ=F)$z
SST34_100 <- ds(SST34_100a$SST34,ic = "AIC", type="monthly",
               standardizeQ=F)$z

file3 <- "../..\\Data\\BoM\\SST\\SST3_anomaly.nc"
SST3 <- nc_open(file3)
SST3_var <- ncvar_get(SST3,SST3$var[[1]])
nc_close(SST3)
SST3_df <- tibble(Date = seq.Date(as.Date("1856-01-01"),
                                   as.Date("2018-04-01"),
                                   by="month"),
                   SST3 = SST3_var)
SST3_100a <- SST3_df %>%
  filter(Date >= "1900-01-01" & Date < "2016-01-01")
SST3_30a <- SST3_df %>%
  filter(Date >= "1979-01-01" & Date < "2016-01-01")
## deseasonalised SST3
SST3_30 <- ds(SST3_30a$SST3,ic = "AIC", type="monthly",
              standardizeQ=F)$z
SST3_100 <- ds(SST3_100a$SST3,ic = "AIC", type="monthly",
              standardizeQ=F)$z

file4 <- "../..\\Data\\BoM\\SST\\SST4_anomaly.nc"
SST4 <- nc_open(file4)
SST4_var <- ncvar_get(SST4,SST4$var[[1]])
nc_close(SST4)
SST4_df <- tibble(Date = seq.Date(as.Date("1856-01-01"),
                                   as.Date("2018-04-01"),
                                   by="month"),
                   SST4 = SST4_var)
SST4_100a <- SST4_df %>%
  filter(Date >= "1900-01-01" & Date < "2016-01-01")
SST4_30a <- SST4_df %>%
  filter(Date >= "1979-01-01" & Date < "2016-01-01")
## deseasonalised SST4
SST4_30 <- ds(SST4_30a$SST4,ic = "AIC", type="monthly",
              standardizeQ=F)$z
SST4_100 <- ds(SST4_100a$SST4,ic = "AIC", type="monthly",
              standardizeQ=F)$z

file5 <- "../..\\Data\\BoM\\PDO\\PDO.latest_1900-.txt"
pdo_mth <- read_table(file5)
pdo_mth <- gather(pdo_mth,key="Month", value="PDO", JAN:DEC)
pdo_mth$YEAR <- substr(pdo_mth$YEAR,1,4)
pdo_mth <- pdo_mth[order(as.Date(paste(pdo_mth$YEAR,pdo_mth$Month,
                                        "01",sep="-"), "%Y-%b-%d")),]
PDO_100a <- pdo_mth %>%
  filter(YEAR > 1899 & YEAR < 2016)
## deseasonalised PDO
PDO_100 <- ds(PDO_100a$PDO,ic = "AIC", type = "monthly",
              standardizeQ=F)$z

file7 <- "../..\\Data\\BoM\\IOD\\iod_jamstec.txt"
IOD <- read_table2(file7)
IOD <- tibble(Date = seq.Date(as.Date("1870-01-16"),
                                   as.Date("2018-03-16"),
                                   by="month"),
              IOD = IOD$`DMI(=West-East)`)
IOD_30a <- IOD %>%
  filter(Date > "1979-01-01" & Date < "2016-01-01")
IOD_100a <- IOD %>%
  filter(Date > "1899-01-01" & Date < "2016-01-01")
## deseasonalised IOD
IOD_30 <- ds(IOD_30a$IOD, ic = "AIC", type = "monthly",
             standardizeQ=F)$z
IOD_100 <- ds(IOD_100a$IOD, ic = "AIC", type = "monthly",
             standardizeQ=F)$z
# end of read data
# should write out useful data
save.image("../../data/data_environment.Rdata")
```

## Statistical method

The step changes on rainfall were analysed using two different statistical methods. A step change is not obvious in the time series data of the rainfall residuals (Figure \@ref(fig:ts-mean)), even though the data is deseasonalised and detrended.  Both methods in this paper make use of a regression model to remove variability in rainfall due to climatic factors to strengthen the tree cover change signal. In the first method, the tree cover change was implemented as a factor variable in the regression model. In the second method, a rank sum test (step trend test), was applied to the model residuals after effects of other major factors were removed. This assumes that after removal of all climate variation, the vegetation cover change is the only factor explaining the non-random pattern in the rainfall residuals.

```{r ts-mean, echo = F, fig.cap = "The deseasonalised and detrended rainfall over the 30 years period in (a) the QLD region and (b) the NSW/VIC region. The vertical red lines indicate the year of 2003, in which the studied land cover changes occurred. A change in the time series data is not obvious before and after the land cover changes.", out.width="90%", echo=F}

include_graphics("figures/Rainfall_resid.jpg")
```

### Regression model {#reg_model}

As highlighted in the introduction, the Australian climate is influenced by sea surface temperatures in the tropical Pacific and Indian Oceans, as well as pressure systems in the Southern Ocean [@BoM2012].  @Risbey2009 compared five large-scale drivers, including ENSO (measured by SOI and the Tropical Pacific SSTs), IOD, SAM, MJO (Madden-Julian oscillation) and blocking, in relation to Australian rainfall variability. The MJO is a large scale eastward-propagating wave-like disturbance in equatorial latitudes [@Risbey2009]. They identified SOI as the most important index among all indices tested for broad parts of Australia (including QLD and NSW/VIC) in almost any season. In this study, four indices were selected from the main climatic indicators (see section [Data]) were used as the explanatory variables in the model for each study region.

```{r climateData, echo=F, eval=F}
load("../../data/data_environment.Rdata")

# read in the BOM gridded rainfall data extracted earlier
rain <- readRDS("../../data/bom/rainfall/Raindata.rds")
# dim(rain)
# 43829 columns (days from 19000101) by 4161 locations (rows)

# split by Qld and NSW
# NSW has first 1254 locations (columns) Qld other 2907
rainNSW <- rain[1:1254,]
rainQld <- rain[1255:4161,]
rm(rain)

# summarise to monthly
# NSW
rainNSW_t <- as.tibble(data.frame(Date = 
                                    seq.Date(as.Date("1900-01-01"),
                                    as.Date("2015-12-31"),by=1),
                    rain=t(rainNSW)))
rm(rainNSW)
rainNSW_t <- rainNSW_t %>% 
  mutate(Month = format(Date,"%Y-%m")) 
rain_monthNSW <- rainNSW_t %>%
  group_by(Month) %>%
  summarise_at(vars(starts_with("rain")), funs(sum = sum(.)))

# Qld
rainQld_t <- as.tibble(data.frame(Date = 
                                    seq.Date(as.Date("1900-01-01"),
                                    as.Date("2015-12-31"),by=1),
                    rain=t(rainQld)))
rm(rainQld)
rainQld_t <- rainQld_t %>% 
  mutate(Month = format(Date,"%Y-%m")) 
rain_monthQld <- rainQld_t %>%
  group_by(Month) %>%
  summarise_at(vars(starts_with("rain")), funs(sum = sum(.)))
rm(rainNSW_t, rainQld_t)

## 100 years data & 30 years
# Qld
qld_100 <- rain_monthQld
qld_30 <- qld_100 %>%
 filter(Month >= "1979-01")

qld100_mean <- qld_100 %>%
  mutate(Means = rowMeans(.[,-1])) %>%
  dplyr::select(Month,Means)

qld30_mean <- qld100_mean %>%
 filter(Month >= "1979-01")

# NSW
nsw_100 <- rain_monthNSW
nsw_30 <- nsw_100 %>%
 filter(Month >= "1979-01")

nsw100_mean <- nsw_100 %>%
  mutate(Means = rowMeans(.[,-1])) %>%
  dplyr::select(Month,Means)

nsw30_mean <- nsw100_mean %>%
 filter(Month >= "1979-01")

require(MASS)
fitdistr(qld30_mean$Means,"gamma")  ## assume shape=0.9
fitdistr(nsw30_mean$Means,"gamma")  ## assume shape =2.19
ks.test(qld30_mean$Means,"pgamma",0.90,0.03)
ks.test(nsw30_mean$Means,"pgamma",2,0.03)

jpeg("../../figures/hist_rainfall.jpg", width=960, height=480)
par(mfrow=c(1,2))
hist(qld30_mean$Means,main="",xlab=expression(paste("rainfall (mm mont",h^-1,")",sep="")))
 hist(nsw30_mean$Means,main="",xlab=expression(paste("rainfall (mm mont",h^-1,")",sep="")))
dev.off()

# autocorrelation in rainfall
acf_df <- data_frame(lag = 1:12, 
                     acf_Qld = acf(qld30_mean$Means,
                                   lag.max=12,main="QLD")$acf[2:13,1,1],
                     acf_NSW = acf(nsw30_mean$Means,
                                   lag.max=12,main="NSW & VIC")$acf[2:13,1,1],
                     pacf_Qld = pacf(qld30_mean$Means,
                                     lag.max=12,main="")$acf[,1,1],
                     pacf_NSW = pacf(nsw30_mean$Means,
                                     lag.max=12,main="")$acf[,1,1])

jpeg("../../figures/autocorrelation.jpg")
acf_df %>%
  gather(key="acf", value="cor", acf_Qld:pacf_NSW) %>%
  ggplot(aes(lag, cor)) + geom_bar(stat="identity", fill = "blue") +
  facet_wrap(~acf) + xlab("Lag") + ylab("correlation")
dev.off()

## deseasonalise & detrend rainfall: assume the deseasonalised monthly zonal rainfall is normal distributed and can be used in pearson cross-correlation function
## use stl function in R

## revision June 2018: change this to only deseasonalise using the 
# deseasonalise package
qld30_deseason <- ds(qld30_mean$Means,ic="AIC", type="monthly",
                     standardizeQ=F)$z
nsw30_deseason <- ds(nsw30_mean$Means,ic="AIC", type="monthly",
                     standardizeQ=F)$z
qld100_deseason <- ds(qld100_mean$Means,ic="AIC", type="monthly",
                      standardizeQ=F)$z
nsw100_deseason <- ds(qld100_mean$Means,ic="AIC", type="monthly",
                      standardizeQ=F)$z

# test whether any linear trends are significant
df_list <- list(qld30 = data.frame(z=qld30_deseason,
                           trend=1:length(qld30_deseason)),
                nsw30 = data.frame(z=nsw30_deseason,
                           trend=1:length(nsw30_deseason)),
                qld100 = data.frame(z=qld100_deseason,
                 trend=1:length(qld100_deseason)),
                nsw100 = data.frame(z=nsw100_deseason,
                 trend=1:length(nsw100_deseason)))
result <- lapply(df_list,
                 function(x) summary(lm(z~trend,
                                        data=x))$coefficients)
# none have a trend

# 
plot_df <- tibble(Dates = rep(ymd(paste(qld30_mean$Month,
                                        "01",sep="_")),2),
                  Resid = c(qld30_deseason,nsw30_deseason),
                  State = c(rep("Qld",length(qld30_deseason)),
                            rep("NSW",length(nsw30_deseason))))

jpeg("../../figures/Rainfall_resid.jpg", width=960, height=960)
plot_df %>%
  ggplot(aes(Dates,Resid)) + geom_line() + 
  geom_vline(xintercept = as.Date("2003-01-01"), 
             col="red", size=1.5) +
  geom_text(aes(x=as.Date("2004-12-01"),y=150),
            label="2003", col="red", size=7) +
  facet_wrap(~State, ncol = 1) + xlab("Date") +
  ylab("Deseasonalised rainfall for the NSW and Qld locations") +
  theme(axis.text=element_text(size=rel(2)),
        axis.title=element_text(size=rel(2.5)),
        strip.text =element_text(size=rel(2), face="bold"))
dev.off()

# Figure 5
plot_df_qld <- tibble(Lag = ccf(as.vector(SOI_30),
                            as.vector(qld30_deseason),lag.max=12, 
                            plot=F)$lag[,,1],
                  SOI = ccf(as.vector(SOI_30),
                            as.vector(qld30_deseason),lag.max=12, 
                            plot=F)$acf[,,1],
                  NINO3.4 = ccf(as.vector(SST34_30),
                                as.vector(qld30_deseason),lag.max=12,
                                plot=F)$acf[,,1],
                  NINO4 = ccf(as.vector(SST4_30),
                              as.vector(qld30_deseason),lag.max=12,
                              plot=F)$acf[,,1],
                  PDO = ccf(as.vector(PDO_100),
                            as.vector(qld100_deseason),lag.max=12,
                            plot=F)$acf[,,1],
                  IOD = ccf(as.vector(IOD_30),
                            as.vector(qld30_deseason),lag.max=12, 
                            plot=F)$acf[,,1])

jpeg("../../figures/cor_qld.jpg", width=960, height=960)
plot_df_qld %>%
  gather(key="SST", value="correlation", SOI:IOD) %>%
  ggplot(aes(Lag,correlation)) + 
  geom_bar(fill="blue", stat="identity") +
  facet_wrap(~SST) + xlab("Lag in months") + ylab("correlation") +
  xlim(c(-12.5,0.5)) +
  theme(axis.text=element_text(size=rel(2)),
        axis.title=element_text(size=rel(2.5)),
        strip.text =element_text(size=rel(2), face="bold"))
dev.off()

# to add and calculate ci: qnorm((1 + 0.95)/2)/sqrt(x$n.used)

plot_df <- tibble(Lag = ccf(as.vector(SOI_30),
                            as.vector(nsw30_deseason),lag.max=12, 
                            plot=F)$lag[,,1],
                  SOI = ccf(as.vector(SOI_30),
                            as.vector(nsw30_deseason),lag.max=12, 
                            plot=F)$acf[,,1],
                  NINO3.4 = ccf(as.vector(SST34_30),
                                as.vector(nsw30_deseason),lag.max=12,
                                plot=F)$acf[,,1],
                  NINO4 = ccf(as.vector(SST4_30),
                              as.vector(nsw30_deseason),lag.max=12,
                              plot=F)$acf[,,1],
                  PDO = ccf(as.vector(PDO_100),
                            as.vector(nsw100_deseason),lag.max=12,
                            plot=F)$acf[,,1],
                  IOD = ccf(as.vector(IOD_30),
                            as.vector(nsw30_deseason),lag.max=12, 
                            plot=F)$acf[,,1])

jpeg("../../figures/cor_nswvic.jpg", width=960, height=960)
plot_df %>%
  gather(key="SST", value="correlation", SOI:IOD) %>%
  ggplot(aes(Lag,correlation)) + 
  geom_bar(fill="blue", stat="identity") +
  facet_wrap(~SST) + xlab("Lag in months") + ylab("correlation") +
  xlim(c(-12.5,0.5)) +
  theme(axis.text=element_text(size=rel(2)),
        axis.title=element_text(size=rel(2.5)),
        strip.text =element_text(size=rel(2), face="bold"))
dev.off()

```

Correlations between rainfall and each climate index were analysed. Rainfall in each study region was first deseasonalised and detrended using the seasonal decomposition function `ds` in the package `deseasonalise` in R [@Rstats2018]. Using detrended data gives a better indication of the underlying correlation rather than the correlation between trends in the data [@Smith2012]. 
The cross-correlations between the deseasonalised and detrended rainfall and the climatic indices were tested using the Pearson's product moment correlation method, assuming the relationships are linear. Although the optimal technique for exploring the correlation with each index could be different as described in @Risbey2009, the Pearson's method was applied to all indices for consistency. Because the PDO describes the multi-decadal SST with lower frequency [@MacDonald2005; @Zanchettin2008; @Kamruzzaman2011], instead of 37-year rainfall data, a longer period (115 years, from 1900 to 2015) was used to estimate the correlation with PDO, up to lag 24. For the other indices, the 37-year data was used.

```{r cor-rain-qld, fig.cap="Cross-correlation of six climate indices and rainfall in QLD study region. For the PDO analysis, 108-year rainfall data (1900 - 2008) are used. Otherwise, 36-year rainfall data are used. The correlation with NINO 3 is not shown as it is very similar to but weaker than for NINO 3.4.", out.width="90%", echo=F}
include_graphics("figures/cor_qld.tif")
```

```{r cor-rain-nsw, fig.cap="Cross-correlation of six climate indices and rainfall in NSW/VIC study region.", out.width="90%", echo=F}
include_graphics("figures/cor_nswvic.tif")
```

Based on the correlation between the climatic indices and rainfall (as shown in Figure \@ref(fig:cor-rain-qld) and Figure \@ref(fig:cor-rain-nsw)), it can be concluded that:
\begin{itemize}
  \setlength{\itemsep}{0cm}
  \setlength{\parskip}{0cm}
  \item In QLD, the correlation between rainfall and SOI at zero time lags is the highest across all indices, outweighing the other ENSO indicators. IOD, had a weak influence in QLD.
  \item In NSW/VIC, again the SOI has the highest correlation with rainfall, followed by the IOD. Both occur at the zero time lags. 
  \item in both cases PDO had the weakest correlations, and this factor was not further considered as a predictor.
\end{itemize}
The above findings are consistent with previous studies. Although some indices are serially correlated with rainfall up to several months, the lag zero events have the most significant correlation coefficients. Concurrent climatic index series were generally found most useful in rainfall prediction [e.g. @Risbey2009; @Kamruzzaman2011]. The correlations between the climatic indices and rainfall for each individual season have also been tested, and the results were similar.

Rainfall in Australia shows strong seasonal patterns [@Holper2011; @ABS2012]. For example, the north part of the country is summer rainfall dominant with a dry winter, while most of the southern part has a winter rainfall regime. This character is driven by the movement of subtropical high pressure systems which dominate the Australian climate [@BoM2012a]. These characteristic summer and winter rainfall patterns mean that the seasonal component of rainfall has a periodic pattern which should be included in the model.

Long term trends in the regional rainfall in some parts of Australia are significant [@Hughes2003; @Gallant2007; @Chowdhury2010]. In the northern and eastern parts of the continent, increasing rainfall is reported over the last century [@Hughes2003]. The presence of such long term trends may be confused with the outcome of a step change in rainfall. As a result a linear trend term was implemented in the model to remove any long term effect.

We assumed all the factors are additive smooth components in determining rainfall following @Kamruzzaman2011. Generally, monthly rainfall has a skewed distribution so the normality assumption of the residuals in a general linear model could be violated. In this case, the rainfall model is a generalised additive model (GAM) [@Hastie1986] with a log link function `g()` and assuming the residuals are gamma distributed (see Figure \@ref(fig:hist-rain)). This means all predictors are modelled as smooth functions, in this case using the shrinkage version of the cubic regression splines [@Wood2011].
\vspace{0.5cm}
\begin{equation}
\begin{array}{lll}
g(E(\mathbf{R}_r)) = &\beta_0 + s_1(\mathbf{SOI}) + s_2(\mathbf{IOD}) + \\                &s_3(\mathbf{Nino3.4}) + s_4(\mathbf{Nino4}) + s_5(\mathbf{Season}) + \\
           &\beta_1\mathbf{Trend} + \boldsymbol{\epsilon}_r
\end{array}
(\#eq:model)
\end{equation}
The bold letters represent the time series vectors.  The region is indicated by $r$, while $\beta_u$ ($_u$=0, 1) are the fitted coefficients in the model. $s_v$ ($_v$=1, 2, 3) are the smooth penalized cubic regression spline functions on the climatic indices and the season. Apart from dropping the PDO as a predictor, all other climate indices were included, allowing the model to select the appropriate predictors.

A possible linear long term trend in the rainfall data is modelled by \textbf{Trend} = 1,2,3...n, where n is the total number of months in the time series. \textbf{Season} is the seasonal component.
The climatic terms are also modelled with smooth functions. The effect of large scale drivers on Australian rainfall is more likely to be seasonal [@Murphy2008; @Schepen2012], and the smooth spline function is more flexible in reproducing the variability in impacts of the climatic indices.

```{r hist-rain, fig.cap="Distribution of monthly rainfall in (a) QLD and (b) NSW/VIC. Using a Kolmogorov-Smirnov test with shape = 1 and 2.4  for QLD and NSW/VIC respectively, rainfall in both regions can be modelled as a gamma distribution.", echo=F, out.width="90%"}
include_graphics("figures/hist_rainfall.jpg")
```

\subsection{Tree cover change as factor variable}

One of the main difficulties in empirical observation studies on the effect of land cover change on rainfall is the lack of continuous monitoring of land surface variables, or even, that no specific variable can be defined that can clearly represent the land surface process. Given the lack of a full picture of the land surface process, a factor variable was used in this study to represent the abrupt land surface change (see Equation \@ref(eq:model)). The change could be a result of either land clearing or bush fires as long as it is permanent or takes a long time to recover. Here we approached the problem with two different models.

In the first method, the tree cover change was used as a predictor in the regression model, represented by a factor variable \textbf{LC}. The significance of the coefficient of \textbf{LC}, denoted as $\beta'_5$ in Equation \@ref(eq:model), can be determined by a ratio test.
\begin{equation}
  \mbox{LC} = \left\{
  \begin{array}{ll}
     \mbox{Trees} \\
     \mbox{Removed}
  \end{array} \right.\\
  (\#eq:dummyC)
\end{equation}
Therefore in both regions, land cover is "trees" for the period before land cover change and "removed" for the period after the change. Here we simply assumed that vegetation cover change has occurred on every pixel. The remaining term $\epsilon_r$ is the amount of rainfall that is attributed to other unspecified factors and random errors. Hence the regression model becomes
\vspace{0.5cm}
\begin{equation}
\begin{array}{lll}
g(E(\mathbf{R}_r)) = &\beta'_0 + s'_1(\mathbf{SOI}) + s'_2(\mathbf{IOD}) + \\
  &s'_3(\mathbf{Nino3.4}) + s'_4(\mathbf{Nino4}) + s'_5\mathbf{Season} + \\
  &\beta'_1\mathbf{Trend} + \beta'_2\mathbf{LC} + \boldsymbol{\epsilon'}_r
\end{array}
  (\#eq:model2)
\end{equation}

One of the difficulties is to point an exact time to the changes in the vegetation cover in the two regions. In the QLD region, no exact time can be assigned to the land clearing. According to the SLATS reports, the most substantial clearing occurred between 2003 - 2004 . However, the information on the change in type of land cover during the time period is missing. Therefore, four scenarios were initially tested in the analysis. In these scenarios the "after change" period started from: (1) June 2003, (2) January 2004, (3) June 2004 and (4) January 2005. A further complicating factor is the influence of the "millenium drought" over the study period and in particular to change to wet conditions in 2010 - 2011 [@vanDijk2013]. In the NSW/VIC region, severe bush fires were reported in early January 2003. Hence the "tree" cover state was up to December 2002 then it was changed to "removed" state from January 2003. However, there is similar uncertainty about when the tree cover had recovered from the bushfire and how this interacted with climate variability. As a result several scenarios were tested for both locations: 1) analysis up to 7 years post change (assumed to be 2010 for both regions); and 2) analysis of the full period (2015). More detail about this is later in the methods. As a starting date, the regression model was run from 1979 for both regions.

```{r simpleMode,eval=F, echo = F}
## plotting a hypothesed land cover change model
data1 <- data.frame(year=c(1,2,2,3),tc=c(2,2,1,1))
data2 <- data.frame(year=c(1,2,2,3,4),tc=c(2,2,1,1.5,2))

jpeg("../figures/tc_simple.jpg", width=720, height=480)
par(mar=c(5,6,1,1), mfrow=c(1,2))
plot(data1$year,data1$tc,type="l",ylim=c(0.5,2.5),xlab="",ylab="",xaxt="n",yaxt="n")
axis(1,at=c(1,2,3),labels=c("before","LCC","after"),cex.axis=1.3)
axis(2,at=c(1,2),label=c("removed","tree"),cex.axis=1.3,las=2)
par(mar=c(5,6,1,1))
plot(data2$year,data2$tc,type="l",ylim=c(0.5,2.5),xlab="",ylab="",xaxt="n",yaxt="n")
axis(1,at=c(1,2,4),labels=c("before","LCC","after"),cex.axis=1.3)
axis(2,at=c(1,2),label=c("removed","tree"),cex.axis=1.3,las=2)
dev.off()

```

```{r fullRegression,eval=F , echo = F}
## predicting model: land cover is category variable
rain.trend <- function(x,y,var.data) {

  var.data$rain <- as.vector(unlist(y[x]))[1:nrow(var.data)]+0.0001

  #if (!is.na(mean(var.data$rain))) {
#    if (grepl("qld",substitute(y))) {
      #var.data <- na.omit(var.data)
      rain_fit <- gam(rain ~ s(month, k=3, bs="cs")
                  + s(SOI, k=3, bs="cs")
                  + s(IOD, k=3, bs="cs")
                  #+ s(PDO, k=3, bs="cs")
                  + s(NINO3.4, k=3, bs="cs")
                  + s(NINO4, k=3, bs="cs")
                  + cover
                  + trd
                  , data=var.data,family=Gamma(link=log),
                  na.action = na.omit)
    # } else
    #  if (grepl("nsw",substitute(y))) {
    #   var.data <- na.omit(var.data)
    #   rain_fit <- gam(rain ~ s(month, k=3)
    #               + s(SOI, k=3)
    #               + s(IOD, k=3)
    #               + cover
    #               + trd
    #               , data=var.data,family=Gamma(link=log))
   # }
    #
  return(rain_fit)

}

## note: link=log generally has higher r.sq value than link=reverse

## generate variables for the model
trd <- seq(1:444)
month <-rep(1:12,times=37)
mth <- cycle(ts(month,start=c(1979,1),frequency=12))

LC <- c(rep("tree",288),rep("clear",156))
# different QLD scenarios
LC_qld1 <- c(rep("tree",294),rep("clear",150)) ## after-change: from Jun 2003
LC_qld2 <- c(rep("tree",300),rep("clear",144)) ## after- : from Jan 2004
LC_qld3 <- c(rep("tree",306),rep("clear",138)) ## after- : from Jun 2004
LC_qld4 <- c(rep("tree",312),rep("clear",132)) ## after- : from Jan 2005

# NSW
var.data <- data.frame(month, trd, SOI = SOI_30a$SOI,
                       IOD = IOD_30a$IOD,
                       PDO = PDO_100a$PDO[949:1392],
                       NINO3.4 = SST34_30a$SST34,
                       NINO4 = SST4_30a$SST4,
                       cover = LC)
# # 5 years
# var.data5 <- var.data %>%
#   filter(trd <= 348)

# 7 years to 2009
var.data7 <- var.data %>%
  filter(trd <= 372)
# 8 years to 2010
 var.data8 <- var.data %>%
   filter(trd <= 384)

# # 9 years
# var.data9 <- var.data %>%
#   filter(trd <= 396)

# QLD
# test different scenarios
# var.qld1 <- data.frame(month, trd, SOI = SOI_30a$SOI,
#                        IOD = IOD_30a$IOD,
#                        PDO = PDO_100a$PDO[949:1392],
#                        NINO3.4 = SST34_30a$SST34,
#                        NINO4 = SST4_30a$SST4,
#                        cover = LC_qld1)
#
var.qld2 <- data.frame(month, trd, SOI = SOI_30a$SOI,
                       IOD = IOD_30a$IOD,
                       PDO = PDO_100a$PDO[949:1392],
                       NINO3.4 = SST34_30a$SST34,
                       NINO4 = SST4_30a$SST4,
                       cover = LC_qld2)
#
# var.qld3 <- data.frame(month, trd, SOI = SOI_30a$SOI,
#                        IOD = IOD_30a$IOD,
#                        PDO = PDO_100a$PDO[949:1392],
#                        NINO3.4 = SST34_30a$SST34,
#                        NINO4 = SST4_30a$SST4,
#                        cover = LC_qld3)

var.qld4 <- data.frame(month, trd, SOI = SOI_30a$SOI,
                       IOD = IOD_30a$IOD,
                       PDO = PDO_100a$PDO[949:1392],
                       NINO3.4 = SST34_30a$SST34,
                       NINO4 = SST4_30a$SST4,
                       cover = LC_qld4)
# 31 years: 1979 - 2009)
# shorter qld series
var.qld4_7y <- var.qld4 %>%
  filter(trd <= 372)
# shorter series to 2010
var.qld4_8y <- var.qld4 %>%
  filter(trd <= 384)

## run model for all periods
   # coef.rain30.qld1 <- lapply(2:ncol(qld_30),
   #                            function(x) rain.trend(x,qld_30,var.qld1))
coef.rain30.qld2 <- lapply(2:ncol(qld_30),
                            function(x) rain.trend(x,qld_30,var.qld2))
   # coef.rain30.qld3 <- lapply(2:ncol(qld_30),
   #                            function(x) rain.trend(x,qld_30,var.qld3))
coef.rain30.qld4 <- lapply(2:ncol(qld_30),
                             function(x) rain.trend(x,qld_30,var.qld4))
# QUEENSLAND 7 yr series
   coef.rain30.qld4_7y <- lapply(2:ncol(qld_30),
                              function(x) rain.trend(x,qld_30,var.qld4_7y))
# and qld 8 year series
   coef.rain30.qld4_8y <- lapply(2:ncol(qld_30),
                             function(x) rain.trend(x,qld_30,var.qld4_8y))
  # NSW/VIC
  coef.rain30.nswvic <- lapply(2:ncol(nsw_30),
                               function(x) rain.trend(x,nsw_30,var.data))
# # 5 years
# coef.rain30.nswvic5 <- lapply(2:ncol(nsw_30),
#                               function(x) rain.trend(x,nsw_30,var.data5))
# 7 years
 coef.rain30.nswvic7 <- lapply(2:ncol(nsw_30),
                               function(x) rain.trend(x,nsw_30,var.data7))
# 8 years (includes 2010)
coef.rain30.nswvic8 <- lapply(2:ncol(nsw_30),
                              function(x) rain.trend(x,nsw_30,var.data8))
# # 9 years
# coef.rain30.nswvic9 <- lapply(2:ncol(nsw_30),
#                               function(x) rain.trend(x,nsw_30,var.data9))

## extract sign of C (the tree cover effect)
  C30_qld <- unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$p.coeff[2],NA)))
# shorter series QLD (includes 2009)
  C30_qld_7y <- unlist(lapply(1:length(coef.rain30.qld4_7y),function(x) ifelse(summary(coef.rain30.qld4_7y)[x,2]=="gam",summary(coef.rain30.qld4_7y[[x]])$p.coeff[2],NA)))
# shorter series QLD (includes 2010)
  C30_qld_8y <- unlist(lapply(1:length(coef.rain30.qld4_8y),function(x) ifelse(summary(coef.rain30.qld4_8y)[x,2]=="gam",summary(coef.rain30.qld4_8y[[x]])$p.coeff[2],NA)))
# NSW/VIC
C30_nswvic <- unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$p.coeff[2],NA)))
# NSW/VIC 7 year (includes 2009)
 C30_nswvic7 <- unlist(lapply(1:length(coef.rain30.nswvic7),function(x) ifelse(summary(coef.rain30.nswvic7)[x,2]=="gam",summary(coef.rain30.nswvic7[[x]])$p.coeff[2],NA)))
# NSW/VIC 8 year (includes 2010)
 C30_nswvic8 <- unlist(lapply(1:length(coef.rain30.nswvic8),function(x) ifelse(summary(coef.rain30.nswvic8)[x,2]=="gam",summary(coef.rain30.nswvic8[[x]])$p.coeff[2],NA)))

jpeg("../../figures/hist_sign_LC.jpg")
par(mfrow=c(2,2))
hist(C30_nswvic)
hist(C30_nswvic7)
hist(C30_nswvic8)
dev.off()
# also for long term trend in nsw
jpeg("../../figures/hist_trend_nsw.jpg")
par(mfrow=c(2,2))
hist(unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$p.coeff[3],NA))), xlab = "value of trend estimate", main="full data period")
hist(unlist(lapply(1:length(coef.rain30.nswvic7),function(x) ifelse(summary(coef.rain30.nswvic7)[x,2]=="gam",summary(coef.rain30.nswvic7[[x]])$p.coeff[3],NA))), xlab = "value of trend estimate", main="Up to 2009")
hist(unlist(lapply(1:length(coef.rain30.nswvic8),function(x) ifelse(summary(coef.rain30.nswvic8)[x,2]=="gam",summary(coef.rain30.nswvic8[[x]])$p.coeff[3],NA))), xlab = "value of trend estimate", main="Up to 2010")
dev.off()
# shows trds are all positive and small

# C30_nswvic5 <- unlist(lapply(1:length(coef.rain30.nswvic5),function(x) ifelse(summary(coef.rain30.nswvic5)[x,2]=="gam",summary(coef.rain30.nswvic5[[x]])$p.coeff[2],NA)))
# C30_nswvic9 <- unlist(lapply(1:length(coef.rain30.nswvic9),function(x) ifelse(summary(coef.rain30.nswvic9)[x,2]=="gam",summary(coef.rain30.nswvic9[[x]])$p.coeff[2],NA)))

# p-value of C
Cp30_qld4 <- unlist(lapply(1:length(coef.rain30.qld4),
   function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",
                      summary(coef.rain30.qld4[[x]])$p.pv[2],NA)))
# Cp30_qld2 <- unlist(lapply(1:length(coef.rain30.qld2),
#    function(x) ifelse(summary(coef.rain30.qld2)[x,2]=="gam",
#                       summary(coef.rain30.qld2[[x]])$p.pv[2],NA)))
# Qld 7 year series
 Cp30_qld4_7y <- unlist(lapply(1:length(coef.rain30.qld4_7y),
    function(x) ifelse(summary(coef.rain30.qld4_7y)[x,2]=="gam",
                       summary(coef.rain30.qld4_7y[[x]])$p.pv[2],NA)))
# Qld 8 year series
Cp30_qld4_8y <- unlist(lapply(1:length(coef.rain30.qld4_8y),
   function(x) ifelse(summary(coef.rain30.qld4_8y)[x,2]=="gam",
                      summary(coef.rain30.qld4_8y[[x]])$p.pv[2],NA)))
# NSW all data
Cp30_nswvic <- unlist(lapply(1:length(coef.rain30.nswvic),
    function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",
                       summary(coef.rain30.nswvic[[x]])$p.pv[2],
                       NA)))
# NSW 7 year series
 Cp30_nswvic7 <- unlist(lapply(1:length(coef.rain30.nswvic7),
               function(x) ifelse(summary(coef.rain30.nswvic7)[x,2]=="gam",
                             summary(coef.rain30.nswvic7[[x]])$p.pv[2],NA)))
# NSW 8 year series
Cp30_nswvic8 <- unlist(lapply(1:length(coef.rain30.nswvic8),
          function(x) ifelse(summary(coef.rain30.nswvic8)[x,2]=="gam",
                             summary(coef.rain30.nswvic8[[x]])$p.pv[2],NA)))

## p value of seasons and climate indices
p_qld <- do.call(rbind,
                 lapply(1:length(coef.rain30.qld4), function(x) {
                   ifelse(summary(coef.rain30.qld4)[x,2]=="gam",
                      return(summary(coef.rain30.qld4[[x]])$s.pv),
                                       return(rep(NA,6)))}))

p_qld <- as.tibble(p_qld)
colnames(p_qld) = c("month", "SOI", "IOD", "NINO3.4", "NINO4")
p_qld$state <- "QLD"

p_nsw <- do.call(rbind,
                 lapply(1:length(coef.rain30.nswvic), function(x) {
                ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",
                  return(summary(coef.rain30.nswvic[[x]])$s.pv),
                       return(rep(NA,6)))}))
p_nsw <- as.tibble(p_nsw)
colnames(p_nsw) = c("month", "SOI", "IOD", "NINO3.4", "NINO4")
p_nsw$state <- "NSW/VIC"

p_plot_df <- rbind(p_qld,p_nsw)

jpeg("../../figures/pvalues_fullmodel.jpg", height=960, width=960)
p_plot_df %>%
  gather(key="Variable", value="p_value", month:NINO4) %>%
  ggplot(aes(p_value,fill=state)) + geom_histogram(alpha=0.5) +
  facet_wrap(~Variable) +
  scale_fill_manual(values = c("blue","red")) +
  theme(axis.title = element_text(size = rel(2)),
        axis.text = element_text(size=rel(1.5)),
        strip.text = element_text(face="bold", size= rel(2)),
        legend.text = element_text(face="bold", size= rel(1.5)),
        legend.title = element_text(face="bold", size= rel(2)))
dev.off()
rm(p_plot_df)

## extract p value of long term trend
trdp30_qld4 <- unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$p.pv[3],NA)))
trdp30_nswvic <- unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$p.pv[3],NA)))

# show histogram
plot_df <- data_frame(p_trd = c(trdp30_qld4,trdp30_nswvic),
                      state = c(rep("QLD", length(trdp30_qld4)),
                                rep("NSW/VIC", length(trdp30_nswvic))))

jpeg("../../figures/pvalue_trd.jpg", height=480, width=480)
plot_df %>%
  ggplot(aes(p_trd, fill=state)) + geom_histogram() +
  scale_fill_manual(values = c("blue","red")) +
  xlab("pvalue trend of the full gam model")
dev.off()

## extract r.sq of the model
rsq_qld <- unlist(lapply(1:length(coef.rain30.qld4),
          function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",
                    summary(coef.rain30.qld4[[x]])$r.sq,NA)))
rsq_nsw <- unlist(lapply(1:length(coef.rain30.nswvic),
    function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",
                    summary(coef.rain30.nswvic[[x]])$r.sq,NA)))

# show histogram
plot_df <- data_frame(`r.squared` = c(rsq_qld,rsq_nsw),
                      state = c(rep("QLD", length(rsq_qld)),
                                rep("NSW/VIC", length(rsq_nsw))))

jpeg("../../figures/rsq_fullmodel.jpg", height=480, width=480)
plot_df %>%
  ggplot(aes(`r.squared`*100, fill=state)) + geom_histogram() +
  scale_fill_manual(values = c("blue","red")) +
  xlab("r_squared of the full gam model")
dev.off()

plot_df %>%
  group_by(state) %>%
  summarise(mean_rsq = mean(`r.squared`))

## frame of the two regions (100 years data)
qld_tif <- raster("../..\\Data\\BoM\\rainfall\\AWAP\\from1900_FVG\\r19000101_qld.tif")
nsw_tif <- raster("../..\\Data\\BoM\\rainfall\\AWAP\\from1900_FVG\\r19000101_nsw.tif")

  dist_Cp30_qld4 <- raster(qld_tif)
  dist_Cp30_qld4[] <- Cp30_qld4
  dist_Cp30_qld4_8y <- raster(qld_tif)
  dist_Cp30_qld4_8y[] <- Cp30_qld4_8y
  dist_Cp30_qld4_7y <- raster(qld_tif)
  dist_Cp30_qld4_7y[] <- Cp30_qld4_7y

  jpeg("../../figures/Cp_30yrs_qld.jpg", width=1440, height=720)
  par(mfrow=c(1,2))
  par(mar=c(3,3,1,7),xpd=T)
  plot(e111,xlab="",ylab="", cex.lab = 2,
       cex.axis=2)
  image(dist_Cp30_qld4,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
#   plot(aus_reg2,border="grey",add=T)
  plot(mdb_reg2,border="grey",lwd=3,add=T)
  plot(e111,add=T)
  image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
legend(148.7,-27.9,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=1)

  par(mar=c(3,3,1,7),xpd=T)
  plot(e111,xlab="",ylab="", cex.lab = 2,
       cex.axis=2)
  image(dist_Cp30_qld4_7y,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
#   plot(aus_reg2,border="grey",add=T)
  plot(mdb_reg2,border="grey",lwd=3,add=T)
  plot(e111,add=T)

  # this is a totally blue plot
#   par(mar=c(3,3,1,7),xpd=T)
#   plot(e111,xlab="",ylab="", cex.lab = 2,
#        cex.axis=2)
#   image(dist_Cp30_qld4_8y,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
# #   plot(aus_reg2,border="grey",add=T)
#   plot(mdb_reg2,border="grey",lwd=3,add=T)
#   plot(e111,add=T)
#   image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
# legend(148.7,-27.9,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=1)
  dev.off()

    #   dist_Cp30_qld <- crop(reproject(dist_Cp100_qld,CRS=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84"),program="raster"),e1)
  dist_Cp30_nsw <- raster(nsw_tif)
  dist_Cp30_nsw[] <- Cp30_nswvic
  dist_Cp30_nsw_7y <- raster(nsw_tif)
  dist_Cp30_nsw_7y[] <- Cp30_nswvic7
  dist_Cp30_nsw_8y <- raster(nsw_tif)
  dist_Cp30_nsw_8y[] <- Cp30_nswvic8
#   dist_Cp30_nsw <- crop(reproject(dist_Cp30_nsw,CRS=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84"),program="raster"),e2)
#
  jpeg("../../figures/Cp_30yrs_nswvic.jpg", width=1440, height=720)
  par(mfrow=c(1,3))
  par(mar=c(3,3,1,7),xpd=T)
  plot(e222,xlab="",ylab="", cex.lab = 2,
       cex.axis=2) #,border="white")
  image(dist_Cp30_nsw_8y,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
  plot(aus_reg2,border="grey75",lwd=5,add=T)
  plot(mdb_reg2,border="grey75",lwd=5,add=T)
  plot(e222,add=T)
  image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
  #dev.off()

  par(mar=c(3,3,1,7),xpd=T)
  plot(e222,xlab="",ylab="", cex.lab = 2,
       cex.axis=2) #,border="white")
  image(dist_Cp30_nsw_7y,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
  plot(aus_reg2,border="grey75",lwd=5,add=T)
  plot(mdb_reg2,border="grey75",lwd=5,add=T)
  plot(e222,add=T)
  # # image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
  # dev.off()

   # jpeg("../../figures/Cp_30yrs_nswvic_8y.jpg", width=960, height=960)
  par(mar=c(3,3,1,7),xpd=T)
  plot(e222,xlab="",ylab="", cex.lab = 2,
       cex.axis=2) #,border="white")
  image(dist_Cp30_nsw,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
  plot(aus_reg2,border="grey75",lwd=5,add=T)
  plot(mdb_reg2,border="grey75",lwd=5,add=T)
  plot(e222,add=T)
  dev.off()
#

# Now make the final plot for the paper with both sites and 2009 and all data
  jpeg("../../figures/Cp_30yrs.jpg", width=1440, height=1440)
  par(mfrow=c(2,2))
  # NSW first
  par(mar=c(3,3,1,9),xpd=T)
  plot(e222,xlab="",ylab="", cex.lab = 2,
       cex.axis=2) #,border="white")
  image(dist_Cp30_nsw_7y,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
  plot(aus_reg2,border="grey75",lwd=5,add=T)
  plot(mdb_reg2,border="grey75",lwd=5,add=T)
  plot(e222,add=T)
  image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
legend(149,-36.6,c("State","MDB"),col=c("grey","grey"),
       lwd=c(1,3),bty="n",title="Boundary",cex=1.8)
  #dev.off()

  par(mar=c(3,5,1,7),xpd=T)
  plot(e222,xlab="",ylab="", cex.lab = 2,
       cex.axis=2) #,border="white")
  image(dist_Cp30_nsw,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
  plot(aus_reg2,border="grey75",lwd=5,add=T)
  plot(mdb_reg2,border="grey75",lwd=5,add=T)
  plot(e222,add=T)

  # Then qld
  par(mar=c(3,3,1,9),xpd=T)
  plot(e111,xlab="",ylab="", cex.lab = 2,
       cex.axis=2)
  image(dist_Cp30_qld4_7y,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
#   plot(aus_reg2,border="grey",add=T)
  plot(mdb_reg2,border="grey",lwd=3,add=T)
  plot(e111,add=T)
  image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))

  par(mar=c(3,5,1,7),xpd=T)
  plot(e111,xlab="",ylab="", cex.lab = 2,
       cex.axis=2)
  image(dist_Cp30_qld4,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
#   plot(aus_reg2,border="grey",add=T)
  plot(mdb_reg2,border="grey",lwd=3,add=T)
  plot(e111,add=T)

  dev.off()
# plot up to 2010 for the appendix
  jpeg("../../figures/Cp_30yrs_8y_appendix.jpg", width=1440, height=720)
  par(mfrow=c(1,2))
  # NSW first
  par(mar=c(3,3,1,9),xpd=T)
  plot(e222,xlab="",ylab="", cex.lab = 2,
       cex.axis=2) #,border="white")
  image(dist_Cp30_nsw_8y,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
  plot(aus_reg2,border="grey75",lwd=5,add=T)
  plot(mdb_reg2,border="grey75",lwd=5,add=T)
  plot(e222,add=T)
  image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
legend(149,-36.6,c("State","MDB"),col=c("grey","grey"),
       lwd=c(1,3),bty="n",title="Boundary",cex=1.8)
  #dev.off()

  par(mar=c(3,5,1,7),xpd=T)
  plot(e222,xlab="",ylab="", cex.lab = 2,
       cex.axis=2) #,border="white")
  image(dist_Cp30_qld_8y,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
  plot(aus_reg2,border="grey75",lwd=5,add=T)
  plot(mdb_reg2,border="grey75",lwd=5,add=T)
  plot(e222,add=T)
dev.off()

# ## range of predicted values
  # predictmax_qld30 <- unlist(lapply(1:length(coef.rain30.qld4), function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam", max(coef.rain30.qld4[[x]]$fitted.values),NA)))
  # predictmin_qld30 <- unlist(lapply(1:length(coef.rain30.qld4), function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam", min(coef.rain30.qld4[[x]]$fitted.values),NA)))
#
resid_qld <- lapply(1:length(coef.rain30.qld4), function(x) if(summary(coef.rain30.qld4)[x,2]=="gam") {residuals(coef.rain30.qld4[[x]])} else{NA})

# ## calculate mean residual
  test <- do.call(cbind,resid_qld)
  qld.resid.mean <- apply(test,1, mean,na.rm=T)
  residuals <- data_frame(Dates = seq.Date(as.Date("1979-01-01"),
                                           length = length(qld.resid.mean),
                                           by = "month"),
                          QLD = qld.resid.mean)
#
# predictmax_nsw30 <- unlist(lapply(1:length(coef.rain30.nswvic), function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam", max(coef.rain30.nswvic[[x]]$fitted.values),NA)))
# # 70 - 254.7
#   predictmin_nsw30 <- unlist(lapply(1:length(coef.rain30.nswvic), function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam", min(coef.rain30.nswvic[[x]]$fitted.values),NA)))
# # 18.4 - 62.2
#
resid_nswvic <- lapply(1:length(coef.rain30.nswvic),
            function(x) if(summary(coef.rain30.nswvic)[x,2]=="gam")
              {residuals(coef.rain30.nswvic[[x]])} else{NA})

## calculate mean residual
test <- do.call(cbind,resid_nswvic)
nswvic.resid.mean <- apply(test,1, mean,na.rm=T)
residuals$NSWVIC <- nswvic.resid.mean

## plots
jpeg("../../figures/GAMresidualTime.jpeg", height=960, width=960)
residuals %>%
  gather(key="State", value="residual", QLD, NSWVIC) %>%
  ggplot(aes(Dates,residual)) + geom_line(colour="blue") +
  geom_vline(xintercept=as.numeric(as.Date("2003-01-01")), colour="red",
             size=1.5) +
  facet_wrap(~State)  +
  theme(axis.text = element_text(size=rel(1.5)),
        axis.title = element_text(size=rel(2)),
        strip.text = element_text(size=rel(2), face = "bold"))
dev.off()

#
# ## R^2 of the regression
#     rsq_qld30 <- unlist(lapply(1:length(coef.rain30.qld), function(x) ifelse(summary(coef.rain30.qld)[x,2]=="gam", summary(coef.rain30.qld[[x]])$r.sq,NA)))
#     rsq_nswvic30 <- unlist(lapply(1:length(coef.rain30.nswvic), function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam", summary(coef.rain30.nswvic[[x]])$r.sq,NA)))
#
# ## significant of the other two variables
#     v1_qld30 <- unlist(lapply(1:length(coef.rain30.qld), function(x) ifelse(summary(coef.rain30.qld)[x,2]=="gam", summary(coef.rain30.qld[[x]])$s.pv[1],NA)))
#     v1_nswvic30 <- unlist(lapply(1:length(coef.rain30.nswvic), function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam", summary(coef.rain30.nswvic[[x]])$s.pv[1],NA)))
#
## plot gam.check
jpeg("../../figures/gam_check.jpg", width=720, height=960)
  par(mfrow=c(4,2))
  gam.check(coef.rain30.qld4[[1]])
  gam.check(coef.rain30.nswvic[[1]])
  dev.off()

```

## Step trend test

To support the regression analysis, a step trend test was used to detect changes in rainfall as a result of vegetation cover change. This nonparametric statistical test was modified from the Mann-Whitney Rank-Sum test by [@Hirsch1985] and can identify a step change in data which is cross correlated. The gridded rainfall dataset used in this study has a high spatial correlation between neighbouring pixels due to the underlying interpolation method. The advantages of using the Rank-Sum test therefore are: (1) it does not depend on assumptions of the data distribution; (2) it is not restricted to datasets with no missing data; (3) it is robust and not as easily influenced by outliers and negative numbers [@Hirsch1985].

The rainfall residuals from the regression model in Equation \@ref(eq:model2) were used for the step trend test. According to [@Hirsch1985], to detect a step change, using deseasonalised and detrended data is important. Furthermore, as rainfall can only be partially attributed to local sources and conditions, other effects are introduced by large scale dynamics and changes in other climatic factors. The assumption is that the regression model should remove these effects, and additionally deseasonalise and detrend the rainfall data. As a result, the local landuse effects are amplified in the variation of the remaining residuals. The  test, described in the following sub section, subsequently associates trends in the rainfall residuals with tree cover changes.

```{r Residual_Analysis, eval=F, echo=F}
## GAM without land cover variable for step trend test
rain.trend3 <- function(x,y,var.data) {

  var.data$rain <- as.vector(unlist(y[x]))+0.0001

  #if (!is.na(mean(var.data$rain))) {
    # if (grepl("qld",substitute(y))) {
      rain_fit <- gam(rain ~ trd
                      + s(month, k=3, bs="cs")# added month back in
                  + s(SOI, k=3, bs="cs")
                  + s(IOD, k=3, bs="cs")
                  #+ s(PDO, k=3, bs="cs")
                  + s(NINO3.4, k=3, bs="cs")
                  + s(NINO4, k=3, bs="cs")
                    , data=var.data,family=Gamma(link=log),
                  na.action=na.omit)
    # } else
    # if (grepl("nsw",substitute(y))) {
    #   rain_fit <- gam(rain ~ trd
    #               + s(month,k=3)
    #               + s(SOI, k=3)
    #               + s(IOD, k=3)
    #               , data=var.data,family=Gamma(link=log))
    #}

  return(rain_fit)
  #} #else {NA}

}
#
# ## change LC variable to be null
#   var.data <- data.frame(month, trd, SOI = SOI_30a, IOD = IOD_30a)
#
## run model for 37 years
  coef.rain.qld <- lapply(2:ncol(qld_30),
                          function(x) rain.trend3(x,qld_30,var.qld4))
  coef.rain.qld2 <- lapply(2:ncol(qld_30),
                          function(x) rain.trend3(x,qld_30,var.qld2))
  # QLD shorter series
  test <- qld_30[1:372,]
  coef.rain.qld_7y <- lapply(2:ncol(qld_30), function(x) rain.trend3(x,test,var.qld4_7y))
rm(test)
## QLD
  rsq_qld <- unlist(lapply(1:length(coef.rain.qld), function(x) ifelse(summary(coef.rain.qld)[x,2]=="gam", summary(coef.rain.qld[[x]])$r.sq,NA)))
#
#
## extract residual
  residual_qld <- lapply(1:length(coef.rain.qld), function(x) if(summary(coef.rain.qld)[x,2]=="gam") {coef.rain.qld[[x]]$y-fitted(coef.rain.qld[[x]])} else {NA})

  residual_qld2 <- lapply(1:length(coef.rain.qld2), function(x) if(summary(coef.rain.qld2)[x,2]=="gam") {coef.rain.qld2[[x]]$y-fitted(coef.rain.qld2[[x]])} else {NA})

# shorter Qld series
    residual_qld_7y <- lapply(1:length(coef.rain.qld_7y), function(x) if(summary(coef.rain.qld_7y)[x,2]=="gam") {coef.rain.qld_7y[[x]]$y-fitted(coef.rain.qld_7y[[x]])} else {NA})

#
# ## NSW/VIC
#
#NSW/VIC
  coef.rain.nswvic <- lapply(2:ncol(nsw_30), function(x) rain.trend3(x,nsw_30,var.data))
  test <- nsw_30[1:372,]
  coef.rain.nswvic7 <- lapply(2:ncol(nsw_30), function(x) rain.trend3(x,test,var.data7))
  rm(test)
  # coef.rain.nswvic5 <- lapply(2:ncol(nsw_30), function(x) rain.trend3(x,nsw_30[1:348,],30,var.data5))
  # coef.rain.nswvic9 <- lapply(2:ncol(nsw_30), function(x) rain.trend3(x,nsw_30[1:396,],30,var.data9))

  ## explaining power of the partial model
  rsq_nswvic <- unlist(lapply(1:length(coef.rain.nswvic), function(x) ifelse(summary(coef.rain.nswvic)[x,2]=="gam", summary(coef.rain.nswvic[[x]])$r.sq,NA)))
#

#   predictmax_nswvic <- unlist(lapply(1:length(coef.rain.nswvic), function(x) ifelse(summary(coef.rain.nswvic)[x,2]=="gam", max(coef.rain.nswvic[[x]]$fitted.values),NA)))
#   predictmin_nswvic <- unlist(lapply(1:length(coef.rain.nswvic), function(x) ifelse(summary(coef.rain.nswvic)[x,2]=="gam", min(coef.rain.nswvic[[x]]$fitted.values),NA)))

# extract residual
residual_nswvic <- lapply(1:length(coef.rain.nswvic), function(x) if(summary(coef.rain.nswvic)[x,2]=="gam") {coef.rain.nswvic[[x]]$y-fitted(coef.rain.nswvic[[x]])} else {NA})
# 7 year (2003 - 2010)
residual_nswvic7 <- lapply(1:length(coef.rain.nswvic7), function(x) if(summary(coef.rain.nswvic7)[x,2]=="gam") {coef.rain.nswvic7[[x]]$y-fitted(coef.rain.nswvic7[[x]])} else {NA})

# 5 year
# residual_nswvic5 <- lapply(1:length(coef.rain.nswvic5), function(x) if(summary(coef.rain.nswvic5)[x,2]=="gam") {coef.rain.nswvic5[[x]]$y-fitted(coef.rain.nswvic5[[x]])} else {NA})

# 9 year
# residual_nswvic9 <- lapply(1:length(coef.rain.nswvic9), function(x) if(summary(coef.rain.nswvic9)[x,2]=="gam") {coef.rain.nswvic9[[x]]$y-fitted(coef.rain.nswvic9[[x]])} else {NA})

# histogram of partial model rsq
qld_df <-  data.frame(rsq = rsq_qld*100,
                      state = "Qld")
nsw_df <-  data.frame(rsq = rsq_nswvic*100,
                      state = "NSW")

plot_df <- rbind(qld_df, nsw_df)
jpeg("../../figures/pmodel_explHistogram.jpg")
plot_df %>%
  ggplot(aes(rsq)) + geom_histogram(fill="blue") +
  facet_wrap(~state) +
  xlab(expression(paste("% of variability explained (",r^2,")"))) +
  theme(axis.text = element_text(size=rel(1.5)),
        strip.text = element_text(size=rel(2)),
        axis.title = element_text(size=rel(2)))
 dev.off()

# stack the residuals

# create a boxplot with the annual residuals from 1998
resplot_df <- data_frame(Date = rep(ymd(paste(nsw_30[[1]],
                                              "01", sep="-")),
                        length(residual_nswvic)+length(residual_qld)),
                          residuals = c(unlist(residual_nswvic),
                                                  unlist(residual_qld)),
                      State = c(rep("NSW/VIC",
                                    length(unlist(residual_nswvic))),
                                  rep("QLD", length(unlist(residual_qld)))))

jpeg("../../figures/ResidualBoxplot.jpg")
resplot_df %>%
  filter(Date >= ymd("1998-01-01")) %>%
  ggplot(aes(Date,residuals, fill=State)) +
  geom_boxplot(aes(group=year(Date))) +
  facet_wrap(~State, ncol=1)
dev.off()

# residuals before and after change
#par(mfrow=c(1,2))
jpeg("../../figures/ResidualBoxplotchange.jpg")
p1 <- resplot_df %>%
  filter(Date >= ymd("1998-01-01")) %>%
  mutate(Change = ifelse(year(Date) > 2003, "After", "Before")) %>%
  ggplot(aes(Change,residuals, fill= State)) +
  geom_boxplot(aes(group=Change)) + facet_wrap(~State, ncol=1) +
  guides(fill=FALSE) + ggtitle("1998 - 2015")
#dev.off()

#jpeg("../../figures/ResidualBoxplotchange2.jpg")
p2 <- resplot_df %>%
  filter(Date >= ymd("1998-01-01")) %>%
  mutate(Change = ifelse(year(Date) > 2003, "After", "Before")) %>%
  filter(year(Date) < 2010) %>%
  ggplot(aes(Change,residuals, fill = State)) +
  geom_boxplot(aes(group=Change)) + facet_wrap(~State, ncol=1) +
  ylab(label="") + ggtitle("1998 - 2009")
lay <- rbind(c(1,1,2,2,2))
grid.arrange(p1,p2, layout_matrix = lay)
dev.off()

res2 <- resplot_df %>%
  filter(Date >= ymd("1998-01-01")) %>%
  mutate(Change = ifelse(year(Date) > 2003, "After", "Before")) %>%
  filter(year(Date) < 2010) %>%
  filter(State == "QLD")

  summary(lm(residuals~Change-1,data=res2))
  t.test(formula=residuals~Change,data=res2)
# No significant difference if measured to 2010 and slightly less rain (p < 0.05) before if measured to 2015 (~ 2 mm/month)

# 2015
#   data:  residuals by Change
# t = 21.177, df = 513000, p-value < 2.2e-16
# alternative hypothesis: true difference in means is not equal to 0
# 95 percent confidence interval:
#  1.863279 2.243363
# sample estimates:
#  mean in group After mean in group Before
#            0.7206134           -1.3327075

# 2009
#   data:  residuals by Change
# t = 1.6534, df = 414050, p-value = 0.09825
# alternative hypothesis: true difference in means is not equal to 0
# 95 percent confidence interval:
#  -0.03314461  0.39066623
# sample estimates:
#  mean in group After mean in group Before
#            -1.153947            -1.332708

res2 <- resplot_df %>%
  filter(Date >= ymd("1998-01-01")) %>%
  mutate(Change = ifelse(year(Date) > 2003, "After", "Before")) %>%
  filter(year(Date) < 2010) %>%
  filter(State != "QLD")

  summary(lm(residuals~Change-1,data=res2))
  t.test(formula=residuals~Change,data=res2)

  # slightly more rain (p < 0.05) after the change if measured to 2015 (~ 1 mm/month), significantly less rain (p < 0.05) after the change (~ 9 mm/month) if measured to 2009
# 2015
#   data:  residuals by Change
# t = 7.3033, df = 215950, p-value = 2.817e-13
# alternative hypothesis: true difference in means is not equal to 0
# 95 percent confidence interval:
#  0.9041572 1.5674572
# sample estimates:
#  mean in group After mean in group Before
#            -2.919095            -4.154902

# 2009
#   data:  residuals by Change
# t = -55.329, df = 180240, p-value < 2.2e-16
# alternative hypothesis: true difference in means is not equal to 0
# 95 percent confidence interval:
#  -10.144149  -9.450039
# sample estimates:
#  mean in group After mean in group Before
#           -13.951996            -4.154902

```

```{r partialSSTmodel, eval=F, echo=F}
## explaining power of the partial model only SST
rain.trend4 <- function(x,y,var.data) {

  var.data$rain <- as.vector(unlist(y[x]))+0.0001

  #if (!is.na(mean(var.data$rain))) {
#    if (grepl("qld",substitute(y))) {
      rain_fit <- gam(rain ~  s(SOI, k=3, bs="cs")
                  + s(IOD, k=3, bs="cs")
                  #+ s(PDO, k=3, bs="cs")
                  + s(NINO3.4, k=3, bs="cs")
                  + s(NINO4, k=3, bs="cs")
                  , data=var.data,family=Gamma(link=log))
    # } else
    # if (grepl("nsw",substitute(y))) {
    #   rain_fit <- gam(rain ~ s(SOI, k=3)
    #               + s(IOD, k=3)
    #               , data=var.data,family=Gamma(link=log))
    # }
    #
  return(rain_fit)
  #} #else {NA}

}

## explaining power of index
  coef.soi.qld <- lapply(2:ncol(qld_30), function(x) rain.trend4(x,qld_30,var.qld4))
  coef.soi.nswvic <- lapply(2:ncol(nsw_30), function(x) rain.trend4(x,nsw_30,var.data))
  rsq_soi_qld <- unlist(lapply(1:length(coef.soi.qld), function(x) ifelse(summary(coef.soi.qld)[x,2]=="gam", summary(coef.soi.qld[[x]])$r.sq,NA)))
  rsq_soi_nswvic <- unlist(lapply(1:length(coef.soi.nswvic), function(x) ifelse(summary(coef.soi.nswvic)[x,2]=="gam", summary(coef.soi.nswvic[[x]])$r.sq,NA)))

qld_df <-  data.frame(rsq = rsq_soi_qld*100,
                      state = "Qld")
nsw_df <-  data.frame(rsq = rsq_soi_nswvic*100,
                      state = "NSW")

plot_df <- rbind(qld_df, nsw_df)
jpeg("../../figures/soi_explHistogram.jpg", width=960, height=960)
plot_df %>%
  ggplot(aes(rsq)) + geom_histogram(fill="blue") +
  facet_wrap(~state) +
  xlab(expression(paste("% of variability explained (adj",r^2,")"))) +
  theme(axis.text = element_text(size=rel(1.5)),
        strip.text = element_text(size=rel(2)),
        axis.title = element_text(size=rel(2)))
 dev.off()

 plot_df %>%
   group_by(state) %>%
   summarise(mean_rsq = mean(rsq))

```

### Mann-Whitney rank-sum statistic

As indicated, the step trend test is a modified version of the Mann-Whitney rank-sum statistic [@Hirsch1985]. As a nonparametric rank-based test, the Mann-Whitney test does not use the exact values of rainfall but depends on the ranks of the data. For each month, rainfall residuals of each year were ranked in an ascending order. The ranking of January rainfall in a sample pixel k in QLD is illustrated below:

```{r exampleRank, echo=F}
table <- data_frame(Year = c(1998 , 1999 , 2000 , 2001 ,
                             2002 , 2005 , 2006),
                    `Rainfall residuals` =c(-0.3, -60.9, -16.1,
                                                      -71.7, 111.1,
                                             -7.2, -60.5),
                    Rank= c(6, 2, 4,
                                                               1, 7, 5, 3))
colnames(table)[3] <- paste("Rank","$R'_{1k}$")
pander(table, caption="Example of ranking rainfall residuals",
             style="rmarkdown")

```

\noindent Therefore, the smallest or most negative value has rank 1 and the largest value has the maximum rank.

The before and after period in the data formed two groups of samples. The split point of the two periods was based on the timing of the vegetation cover changes. In the QLD region, changes occurred anytime during 2003 and 2004. In contrast to the previous method, the time period covering the land cover change  was excluded, as the nonparametric test allows missing data. @Hirsch1985 also pointed out that the power of the test is higher if the data of the change period is ignored. Hence 2003 and 2004 were excluded from the analysis. As a result, the after-change period was 2005 - 2015 for the Queensland location.

In the case of NSW/VIC, the bushfires broke out in early January 2003. The change was within a relatively short period of the year. Therefore the after change period in this region still started in January 2003. Following @Hirsch1985, the before period was set to five years (1998 - 2002) in both regions. The length of the after period is difficult for the NSW/VIC as the regrowth would at some point have impacted the local effects.

There is a further complication in the data. The year 2009 was the final year of the millenium drought [@vanDijk2013], followed by two very wet years (also visible in Figure \@ref(fig:ts-mean) in 2010 and 2011. As highlighted earlier, this means the the length of after period could influence the analysis by the inclusion of more or fewer years. as a result the period up to (and including) 2009 was compared to the full period up to 2015. Further analysis up to 2010 is included in the supplementary material.

The rank of rainfall in month j year i in pixel k is denoted as $R'_{ijk}$. The sum of ranks of rainfall in month j in pixel k before the known intervention is:
\begin{equation}
  W_{jk} = \sum_{i=1}^{n_1}R'_{ijk}.
  (\#eq:Wj)
\end{equation}
$n_1$ is the number of years before the land cover change. The expected value of $W_{jk}$ is
\begin{equation}
  \mu_w=n_1(n_1+n_2+1)/2
\end{equation}

$n_2$ is the number of years after the change. Hence the expected value of the rank sum before the intervention is the same for all months and all pixels. The sum of ranks for the whole time period is fixed, as $(n_1+n_2)(n_1+n_2+1)/2$. In this study, since there are only two groups (before and after), knowing the rank-sum of one group is the same as knowing the rank-sum of the other group. If the rainfall data is temporally and spatially independent, the variance of $W_{jk}$ is
\begin{equation}
  \sigma^2_w = n_1\cdot n_2(n_1+n_2+1)/m
\end{equation}
where m is the number of months which is 12 in the case of a full year.

### Step trend test

Instead of completing the Mann-Whitney U-test, @Hirsch1985 applied a standard normal Z test to the rank-sum statistics. As highlighted, this modified test accounts for serial and cross correlation in the data. In the case here, the deseasonalised and detrended data shows little autocorrelation in the time series but possesses strong cross correlation between neighbouring pixels, i.e. $R>0.99$.

The sum of $W_{jk}$ for a block of $ns$ pixels over the whole year, $\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk}$, has mean
\begin{equation}
  E(\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk})=12\cdot ns\cdot\mu_W
\end{equation}
and variance
\begin{equation}
  Var(\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk})=\sum_{j=1}^{12}\sum_{k=1}^{ns}\sum_{h=1}^{ns}C(W_{jk},W_{jh}).
\end{equation}
$C(W_{jk},W_{jh})$ is the covariance of the W statistics between pixel k and pixel h in month j. When $k=h$, $C(W_{jk},W_{jh})=\sigma^2_w$. When $k\neq h$,
\begin{equation}
  C(W_{jk},W_{jh})=\sigma^2_w r(R_k,R_h)
\end{equation}
where $r(R_k,R_h)$ is the product moment correlation coefficient of the concurrent ranks in pixel k and h. Here $r$ is calculated on the full time series in each pixel. In the analysis, the test was applied to a square block of four pixels each time. As argued by @Hirsch1985, $ns=4$ is the most optimal solution to balance the cost and the gain in the test power.

The statistic of the step trend test is then defined as
\begin{equation}
  Z'=\frac{\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk}-12\cdot ns\cdot\mu_w}{\sqrt{Var(\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk})}}.
  (\#eq:Z)
\end{equation}
The above statistic is written for a 12 month period. By changing the value 12, it can also be used to test seasonal rainfall change or for other customized periods.

The null hypothesis ($H_0$) in this study is that there was no change in rainfall due to land surface intervention. The results of the step trend test can be interpreted according to the sign of the Z' score (see Table \@ref(tab:Zscore) [Chapter 23, P887 @Hipel1994]). Z' is normally distributed similar to the standard normal statistics Z. Hence it can be compared to a standard normal distribution to determine the p value.

```{r (ref:Zscore), echo=F}
table2 <- data_frame(" "= c("Z' $>$ 0 & rainfall decreases after change",
    "Z' $<$ 0 & rainfall increases after change",
    "Z' = 0 & rainfall does not change"))
pander(table2, caption="The interpretation of Z' score in the step trend test", style="rmarkdown")
```

```{r steptrend_ns4, echo=F, eval=F}
## perform step test on the data using ns = 4
## function to calculate the step trend statistic (Hirsch, 1985)
stepTrend2 <- function(x,y,i, endyear=2015){
#browser()
    n1 <- 2002-1997

    if (i==1){

      z <- 56  # no. of column, related to the dim of the region
      # Qld 57 by 51
      # NSW 38 by 33

    } else {

    z <- 37 }

if (x%%z!=0){ #& x%%z!=(z-1)) {
    y.1 <- y[[x]]
    y.2 <- y[[x+1]]
#     y.3 <- y[[x+2]]
    y.3 <- y[[x+z]]
    y.4 <- y[[x+z+1]]
#     y.6 <- y[[x+z+2]]
#     y.7 <- y[[x+2*z]]
#     y.8 <- y[[x+2*z+1]]
#     y.9 <- y[[x+2*z+2]]

    if(!is.na(y.1[1])&!is.na(y.2[1])&!is.na(y.3[1])&!is.na(y.4[1])) { #|!is.na(y.5[1])|!is.na(y.6[1])|!is.na(y.7[1])|!is.na(y.8[1])|!is.na(y.9[1])) {

    if (i == 1) {
      n2 <- endyear-2004
      v.1 <- ts(y.1[c(229:288,313:length(y.1))],
                start=c(1997,1),frequency=12)
      v.2 <- ts(y.2[c(229:288,313:length(y.1))],
                start=c(1997,1),frequency=12)
      v.3 <- ts(y.3[c(229:288,313:length(y.1))],
                start=c(1997,1),frequency=12)
      v.4 <- ts(y.4[c(229:288,313:length(y.1))],
                start=c(1997,1),frequency=12)
#       v.5 <- ts(y.5[c(229:288,313:348)],start=c(1998,1),frequency=12)
#       v.6 <- ts(y.6[c(229:288,313:348)],start=c(1998,1),frequency=12)
#       v.7 <- ts(y.7[c(229:288,313:348)],start=c(1998,1),frequency=12)
#       v.8 <- ts(y.8[c(229:288,313:348)],start=c(1998,1),frequency=12)
#       v.9 <- ts(y.9[c(229:288,313:348)],start=c(1998,1),frequency=12)
#       v <- (x$y-x$fitted.values)[c(1:288,313:357)]
#       v.s <- ts(v,start=c(1998,1),frequency=12)

    } else {

      n2 <- endyear-2002

      v.1 <- ts(y.1[c(229:length(y.1))],
                start=c(1997,1),frequency=12)
      v.2 <- ts(y.2[c(229:length(y.2))],
                start=c(1997,1),frequency=12)
      v.3 <- ts(y.3[c(229:length(y.3))],
                start=c(1997,1),frequency=12)
      v.4 <- ts(y.4[c(229:length(y.4))],
                start=c(1997,1),frequency=12)
#       v.5 <- ts(y.5[c(229:348)],start=c(1998,1),frequency=12)
#       v.6 <- ts(y.6[c(229:348)],start=c(1998,1),frequency=12)
#       v.7 <- ts(y.7[c(229:348)],start=c(1998,1),frequency=12)
#       v.8 <- ts(y.8[c(229:348)],start=c(1998,1),frequency=12)
#       v.9 <- ts(y.9[c(229:348)],start=c(1998,1),frequency=12)
#       v <- (x$y-x$fitted.values)[c(1:288, 301:357)]
#       v.s <- ts(v,start=c(1998,1),frequency=12)
}

## rank by month
   for (j in 1:12) {
    assign(paste("r1",j,sep=""),rank(subset(v.1,cycle(v.1)==j)))
    assign(paste("r2",j,sep=""),rank(subset(v.2,cycle(v.2)==j)))
    assign(paste("r3",j,sep=""),rank(subset(v.3,cycle(v.3)==j)))
    assign(paste("r4",j,sep=""),rank(subset(v.4,cycle(v.4)==j)))
#     assign(paste("r5",j,sep=""),rank(subset(v.5,cycle(v.5)==j)))
#     assign(paste("r6",j,sep=""),rank(subset(v.6,cycle(v.6)==j)))
#     assign(paste("r7",j,sep=""),rank(subset(v.7,cycle(v.7)==j)))
#     assign(paste("r8",j,sep=""),rank(subset(v.8,cycle(v.8)==j)))
#     assign(paste("r9",j,sep=""),rank(subset(v.9,cycle(v.9)==j)))

    }

## mann-whitney rank sum statistic
    rlist.1 <- list(r11,r12,r13,r14,r15,r16,r17,r18,r19,r110,r111,r112) ## full year
    rlist.2 <- list(r21,r22,r23,r24,r25,r26,r27,r28,r29,r210,r211,r212)
    rlist.3 <- list(r31,r32,r33,r34,r35,r36,r37,r38,r39,r310,r311,r312)
    rlist.4 <- list(r41,r42,r43,r44,r45,r46,r47,r48,r49,r410,r411,r412)
#     rlist.5 <- list(r51,r52,r53,r54,r55,r56,r57,r58,r59,r510,r511,r512)
#     rlist.6 <- list(r61,r62,r63,r64,r65,r66,r67,r68,r69,r610,r611,r612)
#     rlist.7 <- list(r71,r72,r73,r74,r75,r76,r77,r78,r79,r710,r711,r712)
#     rlist.8 <- list(r81,r82,r83,r84,r85,r86,r87,r88,r89,r810,r811,r812)
#     rlist.9 <- list(r91,r92,r93,r94,r95,r96,r97,r98,r99,r910,r911,r912)
    rlist <- list(unlist(rlist.1),unlist(rlist.2),unlist(rlist.3),unlist(rlist.4)) #,unlist(rlist.5),unlist(rlist.6),unlist(rlist.7),unlist(rlist.8),unlist(rlist.9))
## seasonal
#     rlist <- list(unlist(rlist.1[c(9:11)]),unlist(rlist.2[c(9:11)]),unlist(rlist.3[c(9:11)]),unlist(rlist.4[c(9:11)]))
#     rlist <- list(r1,r2,r12) ## summer
#     rlist <- list(r3,r4,r5)  ## autumn
#     rlist <- list(r6,r7,r8)  ## winter
#     rlist <- list(r9,r10,r11) ## spring
    w.s1 <- 0
    w.s2 <- 0
    w.s3 <- 0
    w.s4 <- 0
#     w.s5 <- 0
#     w.s6 <- 0
#     w.s7 <- 0
#     w.s8 <- 0
#     w.s9 <- 0
  for (k in 1:length(rlist.1)) {
#   for (k in c(9:11)) {
    w.s1 <- sum(rlist.1[[k]][1:n1])+w.s1
    w.s2 <- sum(rlist.2[[k]][1:n1])+w.s2
    w.s3 <- sum(rlist.3[[k]][1:n1])+w.s3
    w.s4 <- sum(rlist.4[[k]][1:n1])+w.s4
#     w.s5 <- sum(rlist.5[[k]][1:n1])+w.s5
#     w.s6 <- sum(rlist.6[[k]][1:n1])+w.s6
#     w.s7 <- sum(rlist.7[[k]][1:n1])+w.s7
#     w.s8 <- sum(rlist.8[[k]][1:n1])+w.s8
#     w.s9 <- sum(rlist.9[[k]][1:n1])+w.s9
    }
    w.s <- sum(w.s1,w.s2,w.s3,w.s4) #,w.s5,w.s6,w.s7,w.s8,w.s9)

## other statistic
# mean for full year
    mu <- n1*(n1+n2+1)/2
    mu.s <- 12*4*mu
# ## mean for one season
#     mu <- n1*(n1+n2+1)/2
#     mu.s <- 3*4*mu

## variance (standard variation). The autocorrelation of the residuals can be assumed to be null (checked, cor.test(), true for NSW, mostly true for WA. QLD? there are some situations in WA require further investigation)
# assume only cross correlation
    ss <- n1*n2*(n1+n2+1)/12
    ss.s <- 0
    for (kk in 1:4) {
      for (kkk in 1:4) {
#         for (kkkk in 1:12) {
          ss.s <- 12*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
#         }
      }
    }
## variance for one season
#     ss <- n1*n2*(n1+n2+1)/3
#     ss.s <- 0
#     for (kk in 1:4) {
#       for (kkk in 1:4) {
#           ss.s <- 3*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
#       }
#     }

## step trend statistic
    z <- (w.s - mu.s)/sqrt(ss.s)

    return(z) }

     else {return(NA)} }

     else {return(NA)}

  }

## 30 years
## full year,
## central QLD
  qld.z.new <- unlist(lapply(1:(length(residual_qld)-57),function(x) stepTrend2(x,residual_qld,1)))
  qld.z.new <- c(qld.z.new,rep(NA,57))
# shorter series
    qld.z.new_7y <- unlist(lapply(1:(length(residual_qld_7y)-57),function(x) stepTrend2(x,residual_qld_7y,1, endyear=2009)))
  qld.z.new_7y <- c(qld.z.new_7y,rep(NA,57))

  tr_qld.step.new <- raster(qld_tif)
  tr_qld.step.new[] <- qld.z.new

  tr_qld.step.new7 <- raster(qld_tif)
  tr_qld.step.new7[] <- qld.z.new_7y

## NSW
  nswvic.z.new <- unlist(lapply(1:(length(residual_nswvic)-38),function(x) stepTrend2(x,residual_nswvic,2)))
  nswvic.z.new <- c(nswvic.z.new,rep(NA,38))

  nswvic.z.new7 <- unlist(lapply(1:(length(residual_nswvic7)-38),function(x) stepTrend2(x,residual_nswvic7,2, endyear = 2009)))
  nswvic.z.new7 <- c(nswvic.z.new7,rep(NA,38))

  tr_nswvic.step.new <- raster(nsw_tif)
  tr_nswvic.step.new[] <- nswvic.z.new
  tr_nswvic.step.new7 <- raster(nsw_tif)
  tr_nswvic.step.new7[] <- nswvic.z.new7

# ## seasonal
#   qld.z.sum <- unlist(lapply(1:(length(residual_qld)-56),function(x) stepTrend2(x,residual_qld,1)))
#   qld.z.sum <- c(qld.z.sum,rep(NA,56))
#
#   tr_qld.step.sum <- raster(qld_tif)
#   tr_qld.step.sum[] <- qld.z.sum
#
#   nswvic.z.sum <- unlist(lapply(1:(length(residual_nswvic)-37),function(x) stepTrend2(x,residual_nswvic,2)))
#   nswvic.z.sum <- c(nswvic.z.sum,rep(NA,37))
#
#   tr_nswvic.step.sum <- raster(nsw_tif)
#   tr_nswvic.step.sum[] <- nswvic.z.sum

  plotclt <- rev(brewer.pal(8,"RdYlBu"))

jpeg("../../figures/step_new.jpg", width=1240, height=960, quality=100)
#tiff("../figures/step_new_nswvic_7y.tif", width=960, height=960)
  par(mar=c(3,3,1,9), mfrow=c(2,2), xpd=T)
  plot(e222,xlab="",ylab="", cex.axis=2)#,border="white")
  image(tr_nswvic.step.new7,col=plotclt,breaks=c(-2.3,-1.645,-1.282,-0.842,0,0.842,1.282,1.645,3),legend=F,add=T)
  plot(mdb_reg2,border="grey",add=T,lwd=3)
  plot(aus_reg2,border="grey",add=T)
  plot(e222,add=T)
image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,0,1,2,3,4),lab.breaks=c(0.00,0.05,0.1,0.2,1,0.2,0.1,0.05,0.00),legend.only=T,horizontal=F,col=plotclt,legend.shrink=0.7,legend.width=0.9,legend.args=list(text=c(expression(paste("level of significance (",alpha,")")),"positive Z'","negative Z'"),side=c(2,3,1),line=0.7,cex=1.5,font=2),axis.args=list(cex.axis=1.5))
legend(149,-36.6,c("State","MDB"),col=c("grey","grey"),
       lwd=c(1,3),bty="n",title="Boundary",cex=1.8)

# nsw 7y
  par(mar=c(3,5,1,7))
  plot(e222,xlab="",ylab="", cex.axis=2)#,border="white")
  image(tr_nswvic.step.new,col=plotclt,breaks=c(-2.3,-1.645,-1.282,-0.842,0,0.842,1.282,1.645,3),legend=F,add=T)
  plot(mdb_reg2,border="grey",add=T,lwd=3)
  plot(aus_reg2,border="grey",add=T)
  #plot(e222,add=T)
# image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,0,1,2,3,4),lab.breaks=c(0.00,0.05,0.1,0.2,1,0.2,0.1,0.05,0.00),legend.only=T,horizontal=F,col=plotclt,legend.shrink=0.7,legend.width=0.9,legend.args=list(text=c(expression(paste("level of significance (",alpha,")")),"positive Z'","negative Z'"),side=c(2,3,1),line=0.5,cex=1,font=2),axis.args=list(cex.axis=0.8))

  # Qld
  par(mar=c(3,3,1,9),xpd=T)
  plot(e111,xlab="",ylab="", cex.axis=2)#,border="white")
  image(tr_qld.step.new7,col=plotclt,breaks=c(-3,-1.645,-1.282,-0.842,0,0.842,1.282,1.645,3),legend=F,add=T)
  plot(mdb_reg2,border="grey",add=T,lwd=3)
  plot(e111,add=T)
image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,0,1,2,3,4),lab.breaks=c(0.00,0.05,0.1,0.2,1,0.2,0.1,0.05,0.00),legend.only=T,horizontal=F,col=plotclt,legend.shrink=0.7,legend.width=0.9,legend.args=list(text=c(expression(paste("level of significance (",alpha,")")),"positive Z'","negative Z'"),side=c(2,3,1),line=0.9,cex=1.5,font=2),axis.args=list(cex.axis=1.5))

# qld year
  par(mar=c(3,5,1,7))
  plot(e111,xlab="",ylab="", cex.axis=2)#,border="white")
  image(tr_qld.step.new,col=plotclt,breaks=c(-3,-1.645,-1.282,-0.842,0,0.842,1.282,1.645,3),legend=F,add=T)
  plot(mdb_reg2,border="grey",add=T,lwd=3)
  plot(e111,add=T)
# image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,0,1,2,3,4),lab.breaks=c(0.00,0.05,0.1,0.2,1,0.2,0.1,0.05,0.00),legend.only=T,horizontal=F,col=plotclt,legend.shrink=0.7,legend.width=0.9,legend.args=list(text=c(expression(paste("level of significance (",alpha,")")),"positive Z'","negative Z'"),side=c(2,3,1),line=0.5,cex=1,font=2),axis.args=list(cex.axis=0.8))
  dev.off()

#ns = 4 results
# positive
qld_origp.new <- length(subset(qld.z.new,qld.z.new>=1.645))/
  length(qld.z.new)
# 0.00034
# negative
qld_orign.new <- length(subset(qld.z.new,qld.z.new<=-1.645))/
  length(qld.z.new)
# 0
# shorter series
qld_origp.new_7y <- length(subset(qld.z.new_7y,qld.z.new_7y>=1.645))/
  length(qld.z.new_7y)
# 0.212
qld_orign.new_7y <- length(subset(qld.z.new_7y,qld.z.new_7y<=-1.645))/
  length(qld.z.new_7y)
# 0

nsw_origp.z.new <- length(subset(nswvic.z.new,nswvic.z.new>=1.645))/
  length(nswvic.z.new)
# 0
nsw_orign.z.new <- length(subset(nswvic.z.new,nswvic.z.new<=-1.645))/
  length(nswvic.z.new)
# 0
nsw_origp.z.new7 <- length(subset(nswvic.z.new7,nswvic.z.new7>=1.645))/
  length(nswvic.z.new7)
# 0.03
nsw_orign.z.new7 <- length(subset(nswvic.z.new7,nswvic.z.new7<=-1.645))/
  length(nswvic.z.new7)
# 0

```

```{r bootstrap_ns1, echo=F, eval=F}
## one cell step trend
## perform step test on the data for the bootstrapping test (spatial structure was kept)
## function to calculate the step trend statistic (Hirsch, 1985), ns = 1
stepTrend <- function(x,i, endyear=2015){
#browser()
    n1 <- 2002-1997

  if(!is.na(x[1])) {

    if(i==1) {
      n2 <- endyear-2004

      v <- x[c(229:288,313:(n2*12+313-1))]
#       v <- (x$y-x$fitted.values)[c(1:288,313:357)]
      v.s <- ts(v,start=c(1998,1),frequency=12)
    } else
      if(i==2) {
      n2 <- endyear-2002
      v <- x[c(229:(n2*12+n1*12+229-1))]
#       v <- (x$y-x$fitted.values)[c(1:288, 301:357)]
      v.s <- ts(v,start=c(1998,1),frequency=12)
    } #else
#       {n2 <- 2008-2002
#       v <- (x$y-x$fitted.values)[229:357]
# #       v <- (x$y-x$fitted.values)
#       v.s <- ts(v,start=c(1998,1),frequency=12)}

## rank by month
   for (j in 1:12) {
    assign(paste("r",j,sep=""),rank(subset(v.s,cycle(v.s)==j)))
    }

## mann-whitney rank sum statistic
    rlist <- list(r1,r2,r3,r4,r5,r6,r7,r8,r9,r10,r11,r12) ## full year
#     rlist <- list(r1,r2,r12) ## summer
#     rlist <- list(r3,r4,r5)  ## autumn
#     rlist <- list(r6,r7,r8)  ## winter
#     rlist <- list(r9,r10,r11) ## spring
    w.s <- 0
    for (k in 1:length(rlist)) {
    w.s <- sum(rlist[[k]][1:n1])+w.s
    }

## other statistic
## mean for full year
    mu <- n1*(n1+n2+1)/2
    mu.s <- length(rlist)*mu
## mean for one season
#     mu <- n1*(n1+n2+1)/2
#     mu.s <- 3*mu

## variance (standard variation). The autocorrelation of the residuals can be assumed to be null (checked, cor.test(), true for NSW, mostly true for WA. QLD? there are some situations in WA require further investigation)
## assume only cross correlation
    ss <- n1*n2*(n1+n2+1)/length(rlist)
    ss.s <- length(rlist)*ss
## variance for one season
#     ss <- n1*n2*(n1+n2+1)/3
#     ss.s <- 3*ss

## step trend statistic
    z <- (w.s - mu.s)/sqrt(ss.s)

    return(z) }

     else {return(NA)}

  }

## full year
## central QLD
  Qld.z <- unlist(lapply(residual_qld,function(x) stepTrend(x,1)))
  Qld.z_7y <- unlist(lapply(residual_qld_7y,
                            function(x) stepTrend(x,1, endyear=2010)))
## NSW/VIC
## treatment 1: excluding 2003
  nswvic.z <- unlist(lapply(residual_nswvic,
                            function(x) stepTrend(x,2)))
  nswvic.z_7y <- unlist(lapply(residual_nswvic7,
                            function(x) stepTrend(x,2,
                                                  endyear = 2010)))

# # ns = 1 results note that this is p < 0.1 already
#   Qld_origp <- length(subset(Qld.z,Qld.z>=1.645))
# # 0.002
#   Qld_orign <- length(subset(Qld.z,Qld.z<=-1.645))
# # 0
#   Qld_origp_7y <- length(subset(Qld.z_7y,
#                                 Qld.z_7y>=1.645))
# # 0
#   Qld_orign_7y <- length(subset(Qld.z_7y,
#                                 Qld.z_7y<=-1.645))
# # 0.023
#
#   Nsw_origp <- length(subset(nswvic.z,
#                              nswvic.z>=1.645))
# # 0
#   Nsw_orign <- length(subset(nswvic.z,
#                              nswvic.z<=-1.645))
# # 0
#   Nsw_origp_7y <- length(subset(nswvic.z_7y,
#                                 nswvic.z_7y>=1.645))
# 0
#  Nsw_orign_7y <- length(subset(nswvic.z_7y,
#                                nswvic.z_7y<=-1.645))
# 0

# new step trend test 2: keep spatial structure, but ns = 1
NstepTrend2 <- function(x,y,ii, endyear=2015){

  n1 <- 2002-1997

  if(ii==1) {
      n2 <- endyear-2004
      v <- x[c(229:288,313:length(x))]
    } else
      if(ii==2) {
      n2 <- endyear-2002
      v <- x[c(229:length(x))]
    }

  if(!is.na(unlist(x)[1])) {

  yy<-ts(v,start=c(1998,1),frequency=12)
#browser()
  for(i in 1:12) {
  y.mth <- data.frame(mth=as.vector(window(yy,start=c(1998,i),frequency=1)),o=y[[i]])[order(y[[i]]),]

  ## rank by month
    assign(paste("r",i,sep=""),rank(y.mth$mth) )
  }

## mann-whitney rank sum statistic
    rlist <- list(r1,r2,r3,r4,r5,r6,r7,r8,r9,r10,r11,r12) ## full year

    w.s <- 0
    for (k in 1:length(rlist)) {
    w.s <- sum(rlist[[k]][1:n1])+w.s
    }

## other statistic
## mean for full year
    mu <- n1*(n1+n2+1)/2
    mu.s <- length(rlist)*mu

## variance (standard variation). The autocorrelation of the residuals can be assumed to be null (checked, true for QLD and NSW)
    ss <- n1*n2*(n1+n2+1)/length(rlist)
    ss.s <- length(rlist)*ss

## step trend statistic
    z <- (w.s - mu.s)/sqrt(ss.s)

    return(z)
    }

     else {return(NA)}

  }

## conduct bootstrapping 2
R <- 1000

## QLD
z2pq.values = numeric(R)
z2nq.values = numeric(R)
## sample years without replacement. note that with this method, only the order of years has changed, but months within year do not change
new.order <- list()
# mth.l <- rep(11,12)
# this relates to the total number of years:n1 + n2
#
period1 <- 2002 - 1997
period2 <- 2015 - 2004
mth.l <- rep((period1 + period2),12)

ptm <- proc.time()
for (i in 1:R) {

  for (j in 1:12) {
    new.order[[j]]<-sample(1:mth.l[j])
  }

  temp <- unlist(lapply(residual_qld,
                        function(x,y) NstepTrend2(x,y,1),new.order))
  if(i==1) {write.table(temp,file="bt_zvalues2_qld.txt",row.names=F)
            } else
              {write.table(temp,file="bt_zvalues2_qld.txt",append=T,row.names=F,col.names=F)}
  z2pq.values[i] <- length(subset(temp,temp>=1.645))/length(temp)
  z2nq.values[i] <- length(subset(temp,temp<=-1.645))/length(temp)
}
proc.time()-ptm

write.table(z2pq.values,file="bt_out2p_qld.txt",row.names=F)
write.table(z2nq.values,file="bt_out2n_qld.txt",row.names=F)

## NSW
period1 <- 2002 - 1997
period2 <- 2015 - 2002

mth.l <- rep((period1 + period2),12)

ptm <- proc.time()
for (i in 1:R) {

  for (j in 1:12) {
    new.order[[j]]<-sample(1:mth.l[j])
  }

  temp <- unlist(lapply(residual_nswvic,
                        function(x,y) NstepTrend2(x,y,2),new.order))
  if(i==1) {write.table(temp,file="bt_zvalues2_nswvic.txt",
                        row.names=F)
            } else
            {write.table(temp,file="bt_zvalues2_nswvic.txt",
                           append=T,row.names=F,col.names=F)}
  z2pq.values[i] <- length(subset(temp,temp>=1.645))/length(temp)
  z2nq.values[i] <- length(subset(temp,temp<=-1.645))/length(temp)
}
proc.time()-ptm

write.table(z2pq.values,file="bt_out2p_nswvic.txt",row.names=F)
write.table(z2nq.values,file="bt_out2n_nswvic.txt",row.names=F)

bt1_nswp.values <- unlist(read.table("bt_out2p_nswvic.txt",
                                     header=T,sep=""))*100
bt1_nswn.values <- unlist(read.table("bt_out2n_nswvic.txt",
                                     header=T,sep=""))*100

bt1_qldp.values <- unlist(read.table("bt_out2p_qld.txt",
                                     header=T,sep=""))*100
bt1_qldn.values <- unlist(read.table("bt_out2n_qld.txt",
                                     header=T,sep=""))*100

jpeg("../../figures/bt1_qldp.jpg")
hist(bt1_qldp.values,nclass=100,xlab="% positive Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
points(qld_origp*100,0,pch=16,col="red",cex=1.8)
dev.off()

jpeg("../../figures/bt1_qldn.jpg")
hist(bt1_qldn.values,nclass=100,xlab="% negative Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
points(qld_orign*100,0,pch=16,col="red",cex=1.8)
dev.off()

jpeg("../../figures/bt1_nswp.jpg")
hist(bt1_nswp.values,nclass=100,xlab="% positive Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
points(nsw_origp*100,0,pch=16,col="red",cex=1.8)
dev.off()

jpeg("../../figures/bt1_nswn.jpg")
hist(bt1_nswn.values,nclass=100,xlab="% negative Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
points(nsw_orign*100,0,pch=16,col="red",cex=1.8)
dev.off()

```

```{r bootstrap_ns4, eval=F, echo=F}
# This is the same as chunk 8, but for ns = 4
## multiple cells
# Set up for the shorter period
NstepTrend <- function(x,y,m,i){

    n1 <- 2002-1997

    if (i==1){

      z <- 56  # no. of column, related to the dim of the region
      # Qld 57 by 51
      # NSW 38 by 33

    } else {

    z <- 37 }

if (x%%z!=0){ #& x%%z!=(z-1)) {
    y.1 <- m[[x]]
    y.2 <- m[[x+1]]
    y.3 <- m[[x+z]]
    y.4 <- m[[x+z+1]]

    if(!is.na(y.1[1])&!is.na(y.2[1])&!is.na(y.3[1])&!is.na(y.4[1])) {
    if (i == 1) {
      n2 <- 2009-2004
      v.1 <- ts(y.1[c(229:288,313:length(y.1))],
                start=c(1997,1),frequency=12)
      v.2 <- ts(y.2[c(229:288,313:length(y.2))],
                start=c(1997,1),frequency=12)
      v.3 <- ts(y.3[c(229:288,313:length(y.3))],
                start=c(1997,1),frequency=12)
      v.4 <- ts(y.4[c(229:288,313:length(y.4))],
                start=c(1997,1),frequency=12)

    } else {

      n2 <- 2010-2002

      v.1 <- ts(y.1[c(229:length(y.1))],start=c(1997,1),frequency=12)
      v.2 <- ts(y.2[c(229:length(y.2))],start=c(1997,1),frequency=12)
      v.3 <- ts(y.3[c(229:length(y.3))],start=c(1997,1),frequency=12)
      v.4 <- ts(y.4[c(229:length(y.4))],start=c(1997,1),frequency=12)
}

## rank by month
#browser()
   for (j in 1:12) {
    y1.mth <- data.frame(mth=as.vector(window(v.1,
                                              start=c(1997,i),
                                              frequency=1)),
                         o=y[[j]])[order(y[[j]]),]
    y2.mth <- data.frame(mth=as.vector(window(v.2,
                                              start=c(1997,i),
                                              frequency=1)),
                         o=y[[j]])[order(y[[j]]),]
    y3.mth <- data.frame(mth=as.vector(window(v.3,
                                              start=c(1997,i),
                                              frequency=1)),
                         o=y[[j]])[order(y[[j]]),]
    y4.mth <- data.frame(mth=as.vector(window(v.4,
                                              start=c(1997,i),
                                              frequency=1)),
                         o=y[[j]])[order(y[[j]]),]

    assign(paste("r1",j,sep=""),rank(y1.mth$mth))
    assign(paste("r2",j,sep=""),rank(y2.mth$mth))
    assign(paste("r3",j,sep=""),rank(y3.mth$mth))
    assign(paste("r4",j,sep=""),rank(y4.mth$mth))

    }

## mann-whitney rank sum statistic
    rlist.1 <- list(r11,r12,r13,r14,r15,r16,r17,r18,r19,r110,r111,r112) ## full year
    rlist.2 <- list(r21,r22,r23,r24,r25,r26,r27,r28,r29,r210,r211,r212)
    rlist.3 <- list(r31,r32,r33,r34,r35,r36,r37,r38,r39,r310,r311,r312)
    rlist.4 <- list(r41,r42,r43,r44,r45,r46,r47,r48,r49,r410,r411,r412)
    rlist <- list(unlist(rlist.1),unlist(rlist.2),
                  unlist(rlist.3),unlist(rlist.4))
## seasonal
    w.s1 <- 0
    w.s2 <- 0
    w.s3 <- 0
    w.s4 <- 0
  for (k in 1:length(rlist.1)) {
#   for (k in c(9:11)) {
    w.s1 <- sum(rlist.1[[k]][1:n1])+w.s1
    w.s2 <- sum(rlist.2[[k]][1:n1])+w.s2
    w.s3 <- sum(rlist.3[[k]][1:n1])+w.s3
    w.s4 <- sum(rlist.4[[k]][1:n1])+w.s4
    }
    w.s <- sum(w.s1,w.s2,w.s3,w.s4) #,w.s5,w.s6,w.s7,w.s8,w.s9)

## other statistic
# mean for full year
    mu <- n1*(n1+n2+1)/2
    mu.s <- 12*4*mu
# ## mean for one season
#     mu <- n1*(n1+n2+1)/2
#     mu.s <- 3*4*mu

## variance (standard variation). The autocorrelation of the residuals can be assumed to be null (checked, cor.test(), true for NSW, mostly true for WA. QLD? there are some situations in WA require further investigation)
# assume only cross correlation
    ss <- n1*n2*(n1+n2+1)/12
    ss.s <- 0
    for (kk in 1:4) {
      for (kkk in 1:4) {
#         for (kkkk in 1:12) {
          ss.s <- 12*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
#         }
      }
    }
## variance for one season
#     ss <- n1*n2*(n1+n2+1)/3
#     ss.s <- 0
#     for (kk in 1:4) {
#       for (kkk in 1:4) {
#           ss.s <- 3*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
#       }
#     }

## step trend statistic
    z <- (w.s - mu.s)/sqrt(ss.s)

    return(z) }

     else {return(NA)} }

     else {return(NA)}

  }

R <- 1000
new.order <- list()

zpq.values = numeric(R)
znq.values = numeric(R)
period1 <- 2002 - 1997
period2 <- 2009 - 2004
mth.l <- rep((period1 + period2),12)

ptm <- proc.time()
for (i in 1:R) {

  for (j in 1:12) {
    new.order[[j]]<-sample(1:mth.l[j])
  }

  temp <- unlist(lapply(1:(length(residual_qld_7y)-57),function(x,y) NstepTrend(x,y,residual_qld_7y,1),new.order))
  if(i==1) {write.table(temp,file="bt_zvalues_qld_7y.txt",row.names=F)
            } else
              {write.table(temp,file="bt_zvalues_qld_7y.txt",append=T,row.names=F,col.names=F)}
  zpq.values[i] <- length(subset(temp,temp>=1.645))/length(temp)
  znq.values[i] <- length(subset(temp,temp<=-1.645))/length(temp)
}
proc.time()-ptm

write.table(zpq.values,file="bt_outp_qld_7y.txt",row.names=F)
write.table(znq.values,file="bt_outn_qld_7y.txt",row.names=F)

period1 <- 2002 - 1997
period2 <- 2009 - 2002
mth.l <- rep((period1 + period2),12)

ptm <- proc.time()
for (i in 1:R) {

  for (j in 1:12) {
    new.order[[j]]<-sample(1:mth.l[j])
  }

  temp <- unlist(lapply(1:(length(residual_nswvic7)-38),function(x,y) NstepTrend(x,y,residual_nswvic7,2),new.order))
  if(i==1) {write.table(temp,file="bt_zvalues_nswvic_7y.txt",row.names=F)
            } else
              {write.table(temp,file="bt_zvalues_nswvic_7y.txt",append=T,row.names=F,col.names=F)}
  zpq.values[i] <- length(subset(temp,temp>=1.645))/length(temp)
  znq.values[i] <- length(subset(temp,temp<=-1.645))/length(temp)
}
proc.time()-ptm

write.table(zpq.values,file="bt_outp_nswvic_7y.txt",row.names=F)
write.table(znq.values,file="bt_outn_nswvic_7y.txt",row.names=F)

bt4_nswp.values <- unlist(read.table("bt_outp_nswvic_7y.txt",header=T,sep=""))*100
bt4_nswn.values <- unlist(read.table("bt_outn_nswvic_7y.txt",header=T,sep=""))*100

bt4_qldp.values <- unlist(read.table("bt_outp_qld_7y.txt",header=T,sep=""))*100
bt4_qldn.values <- unlist(read.table("bt_outn_qld_7y.txt",header=T,sep=""))*100

jpeg("../../figures/bt4_hist.jpg")
par(mfrow=c(1,2), mar = c(4,3,2,1)+0.1)
# nsw pos
hist(bt4_nswp.values,nclass=100,xlab="% positive Z'",main="NSW/VIC",cex.lab=1.4,cex.main=1.3) #xlim=c(0,20))
points(nsw_origp.z.new7*100,0,pch=16,col="red",cex=1.8)
# # nsw neg
# hist(bt4_nswn.values,nclass=100,xlab="% negative Z'",main="",cex.lab=1.4,cex.main=1.3, xlim=c(0,20))
# points(nsw_orign.z.new7*100,0,pch=16,col="red",cex=1.8)
#qld pos
hist(bt4_qldp.values,nclass=100,xlab="% positive Z'",main="Qld",cex.lab=1.4,cex.main=1.3) #xlim=c(0,20))
points(qld_origp.new_7y*100,0,pch=16,col="red",cex=1.8)
#qld neg
# hist(bt4_qldn.values,nclass=100,xlab="% negative Z'",main="",cex.lab=1.4,cex.main=1.3, xlim=c(0,20))
# points(qld_orign.new_7y*100,0,pch=16,col="red",cex=1.8)
dev.off()

```

Results
===========

## Tree cover change

The pixels, where the tree cover change based on the MOD44B data was significant ($p \leq 0.05$) in each study region, are shown in Figure \@ref(fig:tctrend) for the time series to 2009 (left panels) and the time series to 2015 (right panels). In both cases it can be seen that the area of negative cover change was greater in the series to 2009 than in the series to 2015. But even the longer series indicates a large change in the NSW/VIC region (bottom row). In the NSW/VIC region, much of the tree loss between 2002 and 2003 was concentrated in the Snowy Mountains close to the border of NSW and VIC. Tree cover loss occurred in large parts of the QLD region between 2002 and 2005, but this tree loss was spatially less concentrated. Most of the clearings appear in the centre of this region in the series to 2009. The tree cover change map (\@ref(fig:tctrend) is consistent with the annual mean EVI trend map (based on DLCD data, not shown here). However in the series to 2015 the tree cover loss seems to have dissappeared (top right Figure \@ref(fig:tctrend)) and suggests an increase in tree cover over the area. The implications of this will be discussed in more detail in the discussion.

```{r tctrend, fig.cap ="The maps show signifcant changes in tree cover identified from the MOD44B data between 2003 and 2009 (top left) and 2003 and 2015 (top right) in the NSW/VIC region and the Qld region (bottom left to 2009 and bottom right to 2015). The amount of change was calculated as the difference in tree cover before and after the specified land cover intervention and it is shown as the percentage of the ground area. Green colour indicates an increase in tree cover, while red colour indicates a decrease in tree cover.", echo=F, out.width="90%"}
include_graphics("figures/tc_figs.jpg")
```

<!-- ```{r tc_trend, fig.cap="The maps show signifcant changes in tree cover identified from the MOD44B data between 2003 and 2009 (top left) and 2003 and 2015 (top right) in the NSW/VIC region and the Qld region (bottom left to 2009 and bottom right to 2015). The amount of change was calculated as the difference in tree cover before and after the specified land cover intervention and it is shown as the percentage of the ground area. Green colour indicates an increase in tree cover, while red colour indicates a decrease in tree cover.", echo=F, out.width="90%"} -->

<!-- include_graphics("figures/tc_figs.jpg") -->
<!-- ``` -->

## Regression Model and significance of Vegetation Cover Changes

Generally the regression model only explains a limited amount of the rainfall variability. The model in Equation \@ref(eq:model2) accounts for around 15\%\footnote{Here the adjusted $R^2$ was reported. Adjusted $R^2$ is the coefficient of determination, a measurement of the amount of variability predicted by the model adjusting for the number of explanatory terms} of the rainfall variation in both regions. The residual analysis shows that the assumptions of the regression model are generally met (Figure \@ref(fig:residuals). The standardised residual plots, however, show some funnelling for the NSW/VIC regions, suggesting non constant variance. The residual patterns are consistent for all pixels within each region.

The model, however, confirms the importance of the climate drivers and the seasonality in Australian rainfall. Even at the grid level, the seasons and several of the climatic indices were significant ($p \leq 0.05$) everywhere in both regions. The explaining power of the model is mostly due to these variables. The climate drivers (at lag zero) accounted, on average, for 6.7\% of the rainfall variability in both the QLD region and the NSW/VIC region (see Figure \@ref(fig:rsq) for the distribution of $R^2$ in these two regions). These figures are within the upper bound of seasonal rainfall predictability by a SST anomaly field reported by [@Westra2010].

```{r residuals, fig.cap ="The residual analysis of a sample pixel in the QLD region (top) and NSW/VIC region (bottom).", echo=F, out.width="90%"}
include_graphics("figures/gam_check.jpg")
```

```{r rsq, fig.cap = "The performance of the regression model if rainfall is only modelled by the climate drivers. It shows the percentage of rainfall variability that can be explained by the climate drivers for the Qld and NSW/VIC region", echo=F, out.width="90%"}
include_graphics("figures/soi_explhistogram.jpg")

```

Statistically significant long term trends were only observed in part of the NSW region (results not shown). However, this result might prove or disprove the existence a long term trend in rainfall. The overall time period is fairly short [@Koutsoyiannis2007] and more pixels in NSW/VIC could indicate a significant step change if the long term trend effect is not removed by the model. As trend free data is an important requirement for the step trend test, the trend term was kept in the regression model to ensure the detection of step change was not due to a possible long term trend.

```{r LCp, fig.cap = "The spatial distribution of significance of land cover step change variable in the GAM model predicting changes in rainfall in the both regions. The top row reflects NSW/VIC, while the bottom relates to the Qld study site. The left hand plots are the data up to 2009, while the right hand plots are for the full series. The p value reported is for the land cover variable in the model.", echo=F, out.width="90%"}
include_graphics("figures/Cp_30yrs.jpg")
```

The land cover variable in the model (Equation \@ref(eq:model2)) aims to identify a step change in the rainfall before and after the observed change in land cover. The  variable was only significant ($p \leq 0.05$) for the rainfall estimates in some areas in NSW/VIC, as shown in Figure \@ref(fig:LCp). However, the number of pixels where the landcover variable was significant was much greater for the series up to 2009, compared to the series to 2015. There was no direct relationship between the areas of bushfires in Figure \@ref(fig:bushfire). No significant step change due to the land cover changes was found in rainfall in the QLD region in the series to 2015, but a small area of change was identified in the period to 2009 (left bottom panel).

Figure \@ref(fig:LCp) for NSW/VIC suggests that a significant step change in rainfall related to land cover change was found in an area larger than where the bushfires has occurred. This might be showing a large scale effect that reaches beyond the vegetation cover change effect.

More generally, the model suggests that the tree cover has a positive impact on rainfall in both regions. The fitted coefficients for the Land cover change variable were consistently positive for the "tree" part of the series. It implies that rainfall was higher when the surface was covered by trees.

## Step Trend Test

```{r steptest30, fig.cap="Spatial distribution of the step trend test Z' statistics in the two study sites. Panels on the top are for the NSW/Vic site, while panels of the bottom ar for Qld. The left panels are for the series to 2009, while the right panels are for the full data series. Warm colours (yellow, orange and red) are for positive Z' values which indicate decreasing rainfall trend due to the land surface intervention. Cold colours (light blue to blue) are for negative Z' values which indicate increasing rainfall trend. The deeper the colour, the more significant the statistic.", echo = F, out.width="90%"}
include_graphics("figures/step_new.jpg")
```

The spatial step trend test Z' scores for the Qld region (top row) and NSWVIC region (bottom row)  are shown in Figure \@ref(fig:steptest30). The figure also contains the period to 2009 (left panels) and the period to 2015 (right panels). This figure provides two types of information: the sign and the significance level. The sign indicates the direction of the step change, as listed in Table \@ref(Zscore). In each region up to 2009, there is a broad area of positive Z' values which implies a decrease in rainfall. There appears to be little relationship between the locations where changes in tree cover are observed (Figure \@ref(fig:tc_trend)) and the patterns in the Z' score. However, there is not necessarily a direct relationship as movement of air masses could mean that actual changes of rainfall are observed close by, but not necessarily exactly at areas with changes of landcover.
There is once again a difference between the left panels and the right panels, indicating that considering a longer time series (to 2015) reduces the pattern and significance of Z' scores. In the QLD region to 2009, 21\% of the pixels obtained a positive Z' score with p $<$ 0.1, and for the period to 2015 no pixels with positive Z' score at p $<$ 0.1 occurred. In the NSW/VIC region to 2009,  only 2.6\% of pixels in the alpine area having a positive Z' score with p $<$ 0.1. In general it is only a small proportion of both study regions.

```{r meandiff, fig.cap ="Boxplots of annual rainfall residuals (estimated based on Equation 2 before and after the land cover intervention during 1979 - 2015 in the study regions. On average, the after period has a significantly lower annual rainfall residual in NSW/VIC, but a signifcantly higher annual rainfall residual in the Qld study area", echo=F, out.width="90%"}

include_graphics("figures/ResidualBoxplotchange.jpg")
```

The rainfall residuals of the two periods (before-change since 1979 and after-change) were also compared (Figure \@ref(fig:meandiff)) using a simple t-test. From the boxplots it is difficult to see that there were significantly different ($p < 0.05$) mean values between the "before" and "after" periods in both regions.  For the Queensland locations, there was no significant difference if measured to 2009 and slightly less rain ($p < 0.05$) before if measured to 2015 ($~ 2 mm/month$). In contrast, for the NSW/VIC locations there was also slightly more rain ($p < 0.05$) after the change if measured to 2015 ($~ 1 mm/month$), but significantly less rain ($p < 0.05$) after the change ($~ 9 mm/month$) if measured to 2009.

```{r dif_mean, echo=F, eval=F}
## calculate mean rainfall difference in two periods up to 2010
## and up to 2015
# exclude 2010 and 2006
# First across all the data (repeat of earlier t-test)

resQld <- resplot_df %>%
  filter(Date >= ymd("1998-01-01")) %>%
  mutate(Change = ifelse(year(Date) > 2003, "After", "Before")) %>%
  filter(year(Date) < 2010 & year(Date) != 2006) %>%
  filter(State == "QLD")

  t.test(formula=residuals~Change,data=resQld)

# no difference to 2015
# data:  residuals by Change
# t = 0.47233, df = 486360, p-value = 0.6367
# alternative hypothesis: true difference in means is not equal to 0
# 95 percent confidence interval:
#  -0.1436579  0.2348813
# sample estimates:
#  mean in group After mean in group Before
#            -1.287096            -1.332708

# Qld to 2009 more rain after  (2 mm/month)
# data:  residuals by Change
# t = 24.139, df = 348950, p-value < 2.2e-16
# alternative hypothesis: true difference in means is not equal to 0
# 95 percent confidence interval:
#  2.574463 3.029469
# sample estimates:
#  mean in group After mean in group Before
#             1.469258            -1.332708

resNSW <- resplot_df %>%
  filter(Date >= ymd("1998-01-01")) %>%
  mutate(Change = ifelse(year(Date) > 2003, "After", "Before")) %>%
  filter(year(Date) < 2010 & year(Date) != 2006) %>%
  filter(State != "QLD")

  t.test(formula=residuals~Change,data=resNSW)
# NSW to 2015 Before less rain (about 2 mm/month)
#   data:  residuals by Change
# t = 12.092, df = 217850, p-value < 2.2e-16
# alternative hypothesis: true difference in means is not equal to 0
# 95 percent confidence interval:
#  1.776336 2.463545
# sample estimates:
#  mean in group After mean in group Before
#            -2.034961            -4.154902

# NSW to 2009 : Less rain After
#  data:  residuals by Change
# t = -28.481, df = 161480, p-value < 2.2e-16
# alternative hypothesis: true difference in means is not equal to 0
# 95 percent confidence interval:
#  -5.698085 -4.964340
# sample estimates:
#  mean in group After mean in group Before
#            -9.486115            -4.154902

# Only need to do this for pos Z score (decrease in rainfall)

# full period
  sel_neg.qld <- residual_qld[which(values(tr_qld.step.new)>=1.645)]
  sel_neg_df <- data_frame(Date = rep(ymd(paste(nsw_30[[1]],
                                              "01", sep="-")),
                        length(sel_neg.qld)),
                          residuals = unlist(sel_neg.qld),
                        pixel = rep(1:length(sel_neg.qld), each = 444))

  data.qld <- sel_neg_df %>%
    group_by(year = year(Date), pixel = pixel) %>%
    summarise(an_residual = sum(residuals)) %>%
  # rearrange
    filter(year > 1997) %>%
    mutate(group = ifelse(year < 2003,"1before","2after"))

  # shorter period to 2010
  sel_neg.qld_7y <- residual_qld_7y[which(values(tr_qld.step.new7)>=1.645)]
  sel_neg_df_7y <- data_frame(Date = rep(ymd(paste(nsw_30[[1]][1:372],
                                              "01", sep="-")),
                        length(sel_neg.qld_7y)),
                          residuals = unlist(sel_neg.qld_7y),
                        pixel = rep(1:length(sel_neg.qld_7y), each = 372))

  data.qld_7y <- sel_neg_df_7y %>%
    group_by(year = year(Date), pixel=pixel) %>%
    summarise(an_residual = sum(residuals)) %>%
    filter(year > 1997) %>%
    mutate(group = ifelse(year < 2003,"1before","2after"))

# NSW/VIC full period no pixels at p < 0.1
  # sel_neg.nswvic <-
  #   residual_nswvic[which(values(tr_nswvic.step.new)>=1.645)]
  # sel_neg_df <- data_frame(Date = rep(ymd(paste(nsw_30[[1]],
  #                                             "01", sep="-")),
  #                       length(sel_neg.nswvic)),
  #                         residuals = unlist(sel_neg.nswvic),
  #                       pixel = rep(1:length(sel_neg.nswvic), each = 444))
  #
  # data.nswvic <- sel_neg_df %>%
  #   group_by(year = year(Date), pixel = pixel) %>%
  #   summarise(an_residual = sum(residuals)) %>%
  # # rearrange
  #   filter(year > 1997) %>%
  #   mutate(group = ifelse(year < 2003,"1before","2after"))

# NSW/VIC shorter period to 2010
  sel_neg.nswvic_7y <-
    residual_nswvic7[which(values(tr_nswvic.step.new7)>=1.645)]
  sel_neg_df_7y <- data_frame(Date = rep(ymd(paste(nsw_30[[1]][1:372],
                                              "01", sep="-")),
                        length(sel_neg.nswvic_7y)),
                          residuals = unlist(sel_neg.nswvic_7y),
                        pixel = rep(1:length(sel_neg.nswvic_7y), each = 372))

  data.nswvic_7y <- sel_neg_df_7y %>%
    group_by(year = year(Date), pixel = pixel) %>%
    summarise(an_residual = sum(residuals)) %>%
  # rearrange
    filter(year > 1997) %>%
    mutate(group = ifelse(year < 2003,"1before","2after"))

# Qld full period
## perform a 2 sample t test
  var.test(unlist(data.qld[data.qld$group=="1before","an_residual"]),
           unlist(data.qld[data.qld$group=="2after" &
                      data.qld$year!=2006 &
                      data.qld$year!=2010 ,3]),alternative="two.sided")
## equal variance p = 0.1802
  t.test(unlist(data.qld[data.qld$group=="1before","an_residual"]),
           unlist(data.qld[data.qld$group=="2after" &
                      data.qld$year!=2006 &
                      data.qld$year!=2010  ,3]),alternative="greater")
## p = 0.1702

# Qld shorter period
## perform a 2 sample t test
  var.test(unlist(data.qld_7y[data.qld_7y$group=="1before","an_residual"]),
           unlist(data.qld_7y[data.qld_7y$group=="2after" &
                      data.qld_7y$year!=2006 &
                      data.qld_7y$year!=2010 ,3]),alternative="two.sided")
## unequal variance p < 0.05
  t.test(unlist(data.qld_7y[data.qld_7y$group=="1before","an_residual"]),
           unlist(data.qld_7y[data.qld_7y$group=="2after" &
                      data.qld_7y$year!=2006 &
                      data.qld_7y$year!=2010,3]),alternative="greater")
## p < 0.05
#   sample estimates:
# mean of x mean of y
#  25.38161  11.14015

# NSW only shorter period
## perform a 2 sample t test
  var.test(unlist(data.nswvic_7y[data.nswvic_7y$group=="1before",
                                 "an_residual"]),
           unlist(data.nswvic_7y[data.nswvic_7y$group=="2after" &
                      data.nswvic_7y$year!=2006 &
                      data.nswvic_7y$year!=2010 ,3]),alternative="two.sided")
## unequal variance p = 0.0543
  t.test(unlist(data.nswvic_7y[data.nswvic_7y$group=="1before",
                               "an_residual"]),
           unlist(data.nswvic_7y[data.nswvic_7y$group=="2after" &
                      data.nswvic_7y$year!=2006 &
                      data.nswvic_7y$year!=2010,3]),alternative="greater")
## p < 0.05
# sample estimates:
# mean of x mean of y
#  49.92258 -51.18393

### REPEAT FOR THE REVERSE ###----------------------------------
  # full period
  sel_neg.qld <- residual_qld[which(values(tr_qld.step.new)<1.645)]
  sel_neg_df <- data_frame(Date = rep(ymd(paste(nsw_30[[1]],
                                              "01", sep="-")),
                        length(sel_neg.qld)),
                          residuals = unlist(sel_neg.qld),
                        pixel = rep(1:length(sel_neg.qld), each = 444))

  data.qld <- sel_neg_df %>%
    group_by(year = year(Date), pixel = pixel) %>%
    summarise(an_residual = sum(residuals)) %>%
  # rearrange
    filter(year > 1997) %>%
    mutate(group = ifelse(year < 2003,"1before","2after"))

  # shorter period to 2010  ## repeat for reverse
  sel_neg.qld_7y <- residual_qld_7y[which(values(tr_qld.step.new7)<1.645)]
  sel_neg_df_7y <- data_frame(Date = rep(ymd(paste(nsw_30[[1]][1:372],
                                              "01", sep="-")),
                        length(sel_neg.qld_7y)),
                          residuals = unlist(sel_neg.qld_7y),
                        pixel = rep(1:length(sel_neg.qld_7y), each = 372))

  data.qld_7y <- sel_neg_df_7y %>%
    group_by(year = year(Date), pixel=pixel) %>%
    summarise(an_residual = sum(residuals)) %>%
    filter(year > 1997) %>%
    mutate(group = ifelse(year < 2003,"1before","2after"))

## REPEAT FOR REVERSE NSW -----------------------
# NSW/VIC full period no pixels at p < 0.1
  # sel_neg.nswvic <-
  #   residual_nswvic[which(values(tr_nswvic.step.new)>=1.645)]
  # sel_neg_df <- data_frame(Date = rep(ymd(paste(nsw_30[[1]],
  #                                             "01", sep="-")),
  #                       length(sel_neg.nswvic)),
  #                         residuals = unlist(sel_neg.nswvic),
  #                       pixel = rep(1:length(sel_neg.nswvic), each = 444))
  #
  # data.nswvic <- sel_neg_df %>%
  #   group_by(year = year(Date), pixel = pixel) %>%
  #   summarise(an_residual = sum(residuals)) %>%
  # # rearrange
  #   filter(year > 1997) %>%
  #   mutate(group = ifelse(year < 2003,"1before","2after"))

# NSW/VIC shorter period to 2010
  sel_neg.nswvic_7y <-
    residual_nswvic7[which(values(tr_nswvic.step.new7)<1.645)]
  sel_neg_df_7y <- data_frame(Date = rep(ymd(paste(nsw_30[[1]][1:372],
                                              "01", sep="-")),
                        length(sel_neg.nswvic_7y)),
                          residuals = unlist(sel_neg.nswvic_7y),
                        pixel = rep(1:length(sel_neg.nswvic_7y), each = 372))

  data.nswvic_7y <- sel_neg_df_7y %>%
    group_by(year = year(Date), pixel = pixel) %>%
    summarise(an_residual = sum(residuals)) %>%
  # rearrange
    filter(year > 1997) %>%
    mutate(group = ifelse(year < 2003,"1before","2after"))

# Qld full period Repeat for reverse
## perform a 2 sample t test
  var.test(unlist(data.qld[data.qld$group=="1before","an_residual"]),
           unlist(data.qld[data.qld$group=="2after" &
                      data.qld$year!=2006 &
                      data.qld$year!=2010 ,3]),alternative="two.sided")
## equal variance p = 0.1802
  t.test(unlist(data.qld[data.qld$group=="1before","an_residual"]),
           unlist(data.qld[data.qld$group=="2after" &
                      data.qld$year!=2006 &
                      data.qld$year!=2010  ,3]),alternative="greater")
## p < 0.05
 #  sample estimates:
 # mean of x  mean of y
 # -8.978994 -18.078116

# Qld shorter period repeat for reverse
## perform a 2 sample t test
  var.test(unlist(data.qld_7y[data.qld_7y$group=="1before","an_residual"]),
           unlist(data.qld_7y[data.qld_7y$group=="2after" &
                      data.qld_7y$year!=2006 &
                      data.qld_7y$year!=2010 ,3]),alternative="two.sided")
## unequal variance p < 0.05
  t.test(unlist(data.qld_7y[data.qld_7y$group=="1before","an_residual"]),
           unlist(data.qld_7y[data.qld_7y$group=="2after" &
                      data.qld_7y$year!=2006 &
                      data.qld_7y$year!=2010,3]),alternative="greater")
## p = 1

# NSW only shorter period
## perform a 2 sample t test
  var.test(unlist(data.nswvic_7y[data.nswvic_7y$group=="1before",
                                 "an_residual"]),
           unlist(data.nswvic_7y[data.nswvic_7y$group=="2after" &
                      data.nswvic_7y$year!=2006 &
                      data.nswvic_7y$year!=2010 ,3]),alternative="two.sided")
## unequal variance p = 0.0543
  t.test(unlist(data.nswvic_7y[data.nswvic_7y$group=="1before",
                               "an_residual"]),
           unlist(data.nswvic_7y[data.nswvic_7y$group=="2after" &
                      data.nswvic_7y$year!=2006 &
                      data.nswvic_7y$year!=2010,3]),alternative="greater")
## p < 0.05
# sample estimates:
#  mean of x  mean of y
#   8.792574 -15.508618

```

A closer look at the rainfall data indicates that rainfall was consistently very low in 2006 for both regions and consistently high for 2010 \@ref(fig:ts-mean). The low rainfall in 2006 was due to a weak El Niño and unrecovered conditions from the previous drought, in contrast, 2010 brough record drought breaking rains across Eastern Australia related to a La Niña [@vanDijk2013]. While the regression model has removed most of the effect of ENSO, the effect of these anomolous years could still be visible in the residuals if the response of the rainfall to the ENSO effect is non-linear, and for example the memory of the past drought is persistent.
This can also explain the significant results in the previous method. To remove the possible outlier effects from 2006 and 2010, the two periods were compared excluding the 2006 and 2010 rainfall. In addition, the group of pixels showing a negative step change in rainfall ($p < 0.10$) and the group of pixels indicating no change were analysed separately for the period to 2009. An unpaired unequal variance two sample t-test was applied to test whether the after period had lower mean rainfall than the before period, excluding the 2006 and 2010 years for the full period, and excluding 2006 for the period to 2009.

For the Queensland locations using the regression model residuals and excluding 2006 and 2010, there was no difference between the period before and after the landuse change for the full data series. However for the period to 2009, there is slightly more rain ($p < 0.05, ~ 2mm/month$) before than after the change. For the NSW/VIC locations for the same analyses, there is slightly more rain ($p < 0.05$) after the change if measured to 2015 ($~ 2 mm/month$), and for the period to 2009 there is still significantly less rain ($p < 0.05, ~ 5 mm/month$).
Focussing on the pixels with a negative step change (positive significant Z' score) and the monthly rainfall totals, only the period to 2009 is of interest, as the full period had too few significant pixels (\@ref(fig:steptest30)). For both regions and excluding 2006, the Qld area had significantly more rain before than after at $p < 0.05$. For the NSW/VIC area, there was similarly significantly more rain before than after the change ($p < 0.05 and 100mm/month$).
For pixels where the Z' score was not significantly positive ($p > 0.1$ and Z' score negative), the Qld area showed no  significant difference in monthly rainfall totals before and after. For the NSW/VIC area there was still significantly less rain after the change ($p < 0.05$) but difference in monthly rainfall totals was smaller than for positive and signifcant Z' scores: 24 mm/month less.

The choice of $ns$ has some impact on the test results [@Hirsch1985]. The cases of $ns = 1$ and $ns = 9$ were also tested. The results for $ns = 1$ showed a lower number of significant pixels(at p $<$ 0.10) compared to the $ns = 4$ test, with only 66 pixels in the Queensland area in the period to 2010 and a slightly higher 6 pixels in the period to 2015. No pixels were significant at p $<$ 0.1 in the NSW/VIC area. The results from the $ns = 9$ test was not different from the $ns = 4$ test. The power of the test does not change much after $ns = 4$, as shown by @Hirsch1985.

As part of the analysis, the "field significance" of the Z' score test was considered to better interpret the step change at regional scales from multiple local tests [@Wilks2006, @Westra2013]. Here, the bootstrapping resampling method from @Westra2013 was used to evaluate the field significance for the period up to 2009 (as the period to 2015 essentially showed no significant positive Z' scores). This means the spatial structure of the pixels was maintained, but the order of the years and months was changed by random resampling. For each resampling the test statistics identifies the percentage of the pixels with significant positive step change for the step trend test. The test statistics on 1000 resampled replicates were used to develop the distribution of these percentage values under the local null hypothesis that there was no step change. The results showed that none of the resampled series indicated any significant step trend for both regions, suggesting that the actual values found are for a unique series of years.

Discussion
=============

```{r summarytable, echo=F, message=F}
library(kableExtra)

table <- read_csv("TestSummaryTable2.csv")
kable(table,caption="Summary table of all tests on the two regions and for the two time periods", format = "latex", booktabs =TRUE) %>%
  kable_styling(latex_options = "scale_down")#%>%
  #kableExtra::landscape()
```

Overall the summary table of the results indicates that the effects of land clearing or bushfire on rainfall are not easily detected, even with a range of different statistical tests. In addition several of the tests appear to indicate contradictory results, in particular in the comparison between the two regions and between the two time periods.

Generally, empirical studies on LCC-precipitation interaction are conducted within an area with known land surface intervention [e.g. @Otterman1990,@Durieux2003,@Negri2004,@Sato2007]. However, these locations are rare and difficult to isolate from real landscape change. Modelling studies are abundant, but are generally not linked to observed data. In this study we therefore tested the effect of land cover change across a broad region, rather than only for locations where changes were known to occur or have occured. The advantage of the suggested approach is that it does not require a long time series of land cover data as this is usually unavailable. Furthermore, it does not assume a specific relationship between vegetation cover change and rainfall but allows the data to show this relationship, by applying the analysis to a broader area outside the boundary of the vegetation cover change. This approach is expected to provide a way to reduce the risk of a false positive paradox, by comparing results between areas with and without vegetation cover change.

While there is some indication that the observed landuse changes (Figure \@ref(fig:tctrend)) cause a decrease in the rainfall, our analysis has not been able to give an unequivocal answer. There are several possible complicating fatcors in the data that could explain the differences between the two regions and the two time periods.

## Rainfall variability

The first and most significant effect on the results remains the variability in the rainfall. The regression model used here is a simple model. We only consider the important effects of a historical trend, seasonality and climate drivers. The purpose of using the regression model is to remove the year to year and month to month variability in rainfall and therefore strengthen the land cover signal in the residuals. However, the model shows that trend, seasonality, ENSO and IOD together explain no more than 30\% of the rainfall variability, and only around 7\% on average is due to the climate drivers (Figure \@ref(fig:rsq) . And while this is consistent with the literature e.g. @Westra2010, a large amount of variation is left in the model residuals. Rainfall is generally considered a stochastic process [e.g. @Fowler2005,@Cowpertwait2009,@Burton2010] and clearly some of the variability is either a different response to a combination of climate factors (as interactions were not tested in the model), or a strongly non-stationary response to the climate drivers. The remaining variability in the residuals increases the difficulty to detect a change in rainfall. This is clearly demonstrated in the difference in land cover variable significance and the Z' score between the series to 2009 and the series to 2015. While 2009 was at the end of the drought, the series to 2015 includes the record breaking year 2010 (Figure \@ref(fig:ts-mean)).

The severe bushfires in 2003 were also triggered by the extreme drought conditions during the millenium drought [@vanDijk2013]. Although the drought on rainfall has been accounted for in part by the model, a further delayed or cumulative impact of drought could be feeding into the local land-atmosphere interaction. As a result, the rainfall feedback to the vegetation cover change could be weak under the dry conditions, and this could have affected the result.

## Vegetation dynamics

The second possible effect is the dynamic nature of the vegetation clearing and recovery. Although land clearing has occurred at a high rate and broad scale in Queensland [@SLATS2017], the clearing does not have a clear start and end point. QLD has a long history of land clearing. According to the series of SLATS reports on land cover changes in QLD released by the Queensland government, land clearing continued in and around the study region between 1988 - 2008. Major broad scale and high rate clearings occurred in 1999 - 2000 and 2002 - 2004 (Figure \@ref(fig:slat)). And even though there was a decrease in land clearing post 2005, it is difficult to define a clear cut change in this region. The more continuous ongoing land clearing could have reduced the significance of a step change.

```{r slat, fig.cap="Woody vegetation clearing rate in QLD for the two major bioregions in the study area. The data were obtained from SLATS (2017), and clearly indicate a sharp decrease in the clearing rate after 2005.", echo=F, out.width="90%"}
include_graphics("figures/slats.jpg")
```

The two approaches used in this study indicate quite different results on the level of change in the NSW/VIC region. The regression model showed that a large area in the NSW/VIC region has experienced a significant land cover effect ($p < 0.05$) on the rainfall after 2003. The effect in the step trend test (Z' scores) is much mroe subdued, even though it was able to detect significant changes ($p < 0.1$) within the area in the period to 2009, but almost no effect in the period to 2015. In contrast, the results are almost reversed in the Qld area. While some of this could be due to the difficulty in removing year on year rainfall variability, some of this might also be due to vegetation regrowth in both locations.

The specific vegetation class in the Queensland area is well-known for rapid regrowth and "thickening" in favourable conditions [@Gowen2016], and this could explain the change in vegetation cover between 2009 and 2015 in the Qld area (Figure \@ref(fig:tctrend)). In particular the favourable rainfall years of 2010 and 2011 would have boosted regrowth, increasing evapotranspiration and therefore decreasing the effect of the rainfall change.

The different causes of vegetation cover change in these two regions could lead to different post-change characteristics. The magnitude of EVI decreasing trends in the QLD region are less than in NSW region, as reported in the DLCD data. This is due to the lower tree density in the QLD region than in the NSW/VIC region before land surface interventions. The significant bushfires (Figure \@ref(fig:fig2bushfire)) would have drastically reduced the vegetation cover and recovery was very slow in some areas within the NSW/VIC area. The persistent drought in the 2000s [@Howden2012,@vanDijk2013] delayed the regrowth of trees. On the other hand, replacing tree cover with pasture and crops in the Qld area might have a relatively subtle impact on the EVI.

## Misalignment rainfall effects and vegetation clearing

Particularly in the NSW/VIC area, there is no direct overlap between the pixels with significant vegetation change (Figure \@ref(fig:tctrend)) and the pixels with significant land cover variables or positive Z'scores (Figure \@ref(fig:LCp) and Figure \@ref(fig:steptest30)). While this seems possibly surprising at first, a plausible explanation could be the general movement of moisture and climate systems over the landscape, resulting in a possible shift of where the moisture is evaporated and where rainfall occurs. Specifically for this reason, this study selected rather large spatial boxes to capture the overall response rather than the pixel by pixel comparison/

## Gridded monthly rainfall data

The rainfall data used in this study is a gridded data set. This data set is robust and consistent over a long time series (from 1900 to current) and has a broad national wide coverage which can provide more information spatially. However, high cross correlation between pixels, due to the interpolation method generating this data set, can also introduce spatial noise. In the step trend test the cross correlation was accounted for. Some other methods are also available which can be used to perform a comparative trial. For example, [@Narisma2007] applied a spatial Gaussian filter on a similar data set and used wavelet analysis to detect step change in rainfall. High quality station data is another option to test whether the observed spatial pattern in the step trend test results was not due to the gridded data itself. Resampling methods, such as bootstrapping and permutation [@Wilks1997,@Kundzewicz2004,@Westra2013], can also be used to further assess the strength of significance of results and incorporate spatial and temporal patterns in the analysis. While the gridded data set is most useful in regions with sparse rain gauge networks, it can actually reduces information where the rain gauge density is high [@Jones2009]. In the NSW/VIC area, the coverage of rainfall stations is more intensive but they are mainly located in the valleys. The interpolated data might not be the best represention of the local rainfall.

<!-- To discuss:   -->

<!-- * Post-fire revegetation and restoration activities in ACT could have reduced the significance of the difference in tree cover before and after the bushfires. It is an indication that deforestation might cause a step change in rainfall but reforestation could take a long time to show effects. Regrowth in Qld and increase in clearing again (Figure \@ref(fig:slat)).    -->
<!-- * Rainfall variation and anomalous years 2006 and 2010.   -->
<!-- * misalingment rainfall change and clearing  -->
<!-- * gridded rainfal    -->
<!-- * general approach   -->

## General approach

Parametric tests are generally more powerful than nonparametric test in detecting a trend, when the data is normally distributed [@Onoz2003,@Kundzewicz2004]. As a non-parametric test, the step trend test has the advantages of distribution free and having no restriction on missing data [@Hirsch1985]. This is particularly useful in rainfall analysis since rainfall data is usually skewed. On the other hand, the disadvantages of non-parametric tests, such as being limited to hypothesis testing and weaker in power, also hold for the step trend test [@Whitley2002].

Overall, the current study provides a clear approach build on several lines of evidence to provide some evidence to reject the null hypothesis (no step change in rainfall occurs as a result of tree cover loss). Limited by the available data, the time frame under study was chosen to include a long lasting drought period [@Holper2011, @vanDijk2013]. The strong impact of this prolonged drought might have suppressed the land-atmosphere interaction and confused the cause and effect relation between rainfall and vegetation cover change. This could be one of the reasons that the LCC effects on the local climate found in other studies [e.g. @Gorgen2006,@McAlpine2007] are not found to be significant here.
Possible future work could focus on a non-drought period, or when a longer series of land cover data is available. Wile the power of the test can be improved with the longer length of the after-intervention period [@Hirsch1985], the dynamic nature of vegetation regrowth in this case study prevents this effect.
A better approach might be to build a global study that investigates multiple locations where drastic landcover changes have taken place, which would also remove some of the climate variability effects due to the larger sample size.

Conclusions
===========

In this study, we present an approach to identify iffound some observationl data based evidence, although not strong, that vegetation cover change has changed local rainfall. The semi-parametric method and non-parametric method did totally agree on detecting a significant step change in rainfall in the hot dry QLD region where land clearing has occurred. On the other hand, the bushfires in the humid, temperate mountain range in the NSW/VIC region has experienced reduced rainfall. But the dry spell also plays an important role in the results.

Drought has had a pronounced impact on the land surface condition during the study period, leading to significant reduction in vegetation and extreme events such as bushfires. The associated lack of rainfall and high temperatures may mask the step change in the vegetation. Hence, the signal of LCC feedback on rainfall is probably weaker under such regional dry conditions, as the impact of LCC on rainfall is mainly through changes in moisture convergence [@Gorgen2006,@Pitman2007].

acknowledgments
===============
CL was supported by an Australian Postgraduate Award for this work.

\newpage
appendix
==============

## Summary of Data

<!--  \begin{sidewaystable} -->
<!--  \caption{Summary of data.} -->
<!--   \label{tab:ch3Data} -->
<!--  \begin{tabular}{lllll} -->
<!--   \hline -->
<!--   \textbf{Data} & \textbf{Source} & \multicolumn{2}{c}{\textbf{Resolution}} & \textbf{Analysis period} \\\cline{3-4} -->
<!--   & & Temporal & Spatial & \\\hline -->
<!--   Percent tree cover & MOD44B & Annual &	250m & 2000-2010\\ -->
<!--   Trend of vegetation cover change  &	DLCD (2009)	& Onetime & 250m	& Trend of Apr 2000 - Apr 2008\\ -->
<!--   Rainfall &	AWAP gridded rainfall data &	Monthly &	0.05\textdegree$\times$0.05\textdegree & Jan 1979- Dec 2008\\ -->
<!--   SOI	& BoM	& Monthly &	N/A &	Jan 1979- Dec 2008\\ -->
<!--   NINO 3, 3.4, 4 &	IRI/LDEO data library	& Monthly	& N/A	& Jan 1979- Dec 2008\\ -->
<!--   PDO	& NOAA & Monthly	& N/A	& Jan 1979- Dec 2008\\ -->
<!--   IOD	& POAMA-2 dataset	& Monthly	& N/A	& Jan 1979- Dec 2008\\ -->
<!--   \hline -->
<!--   \end{tabular} -->
<!-- \end{sidewaystable} -->

\newpage

References {#references .unnumbered}
==========

### 2018-09-04 Manuscripts/Rainfall_Landuse_paper/Rainfall_Landuse_paper.tex
changes for paper

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

  \journal{Journal of Hydrology Regional Studies} % Sets Journal name

\usepackage{lineno} % add
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\bibliographystyle{elsarticle-harv}
\biboptions{sort&compress} % For natbib
\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Detecting influences of changes in land cover on observed rainfall.},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)
% Pandoc header

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}
\begin{frontmatter}

  \title{Detecting influences of changes in land cover on observed rainfall.}
    \author[a]{Chun X. Liang}
   \ead{summer.chunliang@gmail.com}

    \author[a]{Floris F. van Ogtrop}
   \ead{floris.vanogtrop@sydney.edu.au}

    \author[a]{R. Willem Vervoort\corref{c1}}
   \ead{willem.vervoort@sydney.edu.au} 
   \cortext[c1]{Corresponding Author}
      \address[a]{Sydney Institute of Agriculture, The University of Sydney, NSW 2006}

  \begin{abstract}
  Analysis of observational data to identify relationships between
  rainfall and land cover change are difficult due to multiple
  environmental factors that cannot be strictly controlled. In this study
  we present a methodology to investigate the relationship using
  statistical methods on data from best available sources at two sites in
  Australia. Gridded data of rainfall and tree cover were used as
  spatially corresponding local conditions. Large scale effects were
  represented by climatic indicators, such as SOI and IOD. Regression
  analysis and step trend tests were used to assess the effect of abrupt
  land surface intervention. At a Queensland site, significant tree cover
  change between 2002 - 2005 did not result in strong statistically
  significant precipitation changes. On the other hand, results from a
  bushfire affected NSW/VIC region suggests significant changes in the
  rainfall. This indicates the method works better when a abrupt change in
  the data can be clearly identified. The results from the step trend test
  implied a positive relationship between the tree cover and the rainfall
  at 0.1 significance level in both locations in data up to 2009. However,
  high rainfall variability and possible regrowth meant that no signifcant
  changes were observed in alonger time sereies to 2015.
  \end{abstract}

 \end{frontmatter}

\section{Introduction}\label{introduction}

Land use and land cover changes can lead to changes in the local
climate. Empirical and modelling studies have found cloud types and
rainfall are correlated to large scale vegetation cover changes, such as
deforestation in the Amazon and in the Sahel (Chagnon and Bras 2005;
Pinto et al. 2009; Wang et al. 2009; Mei and Wang 2010; Kucharski, Zeng,
and Kalnay 2013; Pitman and Lorenz 2016) and afforestation in south
Israel (Otterman et al. 1990; Ben-Gai et al. 1998). Using airborne
measurement in Western Australia, Junkermann et al. (2009) showed a
significantly higher level of aerosols over an agricultural area
compared to an adjacent natural vegetation. They suggested that a
modification of aerosol concentrations due to deforestation could have
contributed to a reduction of local rainfall, as more, but smaller rain
droplets were observed. Nair et al. (2011) reported from the Bunny Fence
Experiment in Western Australia that local land use change altered the
synoptic west coast trough dynamics and surface roughness, and this
resulted in an observed rainfall decrease. Maximum temperatures were
also found to be sensitive to land cover change in eastern Australia
(McAlpine et al. 2007).

Overall the number of empirical studies analyzing changes to rainfall
due to land cover change from observational data is limited. Most of the
studies mentioned previously were either model simulations, or
comparisons of modelled data with observations. This is because there
are some fundamental experimental difficulties in both space (where does
evaporated water reappear as rainfall?) and in time (how much time does
it take for land cover change effects to appear or disappear?). In
addition, in many areas across the globe rainfall variability is related
to complex set of interactions, of which land use change might only be a
minor component.

Locally, there are two main sources that generate rainfall: moisture
from advective atmospheric transport; and local evapotranspiration
(Eltahir and Bras 1996; Bosilovich and Chern 2006; Dirmeyer, Brubaker,
and DelSole 2009; Gimeno et al. 2010). The local evapotranspiration
component is the component considered to be affected by land use change
(Eltahir and Bras 1996). According to Trenberth (1999), the contribution
of advective moisture partially depends on the availability of external
moisture and atmospheric transport. On the longer time scale, such as
monthly and annually, large scale atmospheric dynamics are affected by
large scale climate drivers. For example, many studies have reported
significant relationships between rainfall in large parts of Australia
and the El Niño-Southern Oscillation (ENSO) (Verdon et al. 2004; Risbey
et al. 2009; Speer, Leslie, and Fierro 2011). In contrast, local ET is
determined by local land surface characteristics, which influence local
scale atmospheric dynamics and hence the amount of rainfall, including
contribution from both main sources. Therefore land surface is suggested
to play an important role in local rainfall.

Although climate drivers demonstrate some capability to predict
Australian rainfall, there is still a large amount of unexplained
variance. Westra and Sharma (2010) pointed out that models based on
global sea surface temperature anomalies can only predict up to 14.7\%
of annual precipitation variance. Some of the remaining variance could
be due to land surface processes as suggested in studies predicting
local rainfall (e.g. Ma et al. 2011; Zeng et al. 2012; Pitman and Lorenz
2016; Saha, Dirmeyer, and Chase 2016). However, they are mostly based on
modelling experiments and little evidence was reported from
observations. However, Pitman et al. (2004) found a good match between
observations and simulated rainfall changes in southwest Western
Australia ,forced by land cover change. Timbal and Arblaster (2006) were
able to reproduce the rainfall decline in south west Australia by
including land cover influence. In addition, local land use change might
not be a primary, but is likely to be a secondary cause of rainfall
change (Nicholls 2006).

Therefore, the aim of this study is to expand the available methodology
to use empirical evidence at regional scales to investigate the cause
and effect relationship between land cover change and local rainfall.
More specifically, we hypothesize that a step change on the land cover
on the surface will cause a step change in the rainfall. To demonstrate
the approach and this effect we applied the methodology to study the
changes in rainfall at two locations in Queensland and NSW/Victoria
where there are possible step changes in land cover change due to land
clearing and bush fires. The methodology is mainly based on statistical
approaches to identify changes in rainfall, which were subsequently
associated with land cover change through spatial comparison.

In this paper, after this section (the introduction), section 2 covers
the case study areas and the observed land use change. Section 3
describes the data used in the study in more detail. Section 4 details
the statistical methods and the underlying assumptions related to the
modelling approach, Section 5 gives the results, which are further
discussed in section 6 and finally section 7 offers the conclusions.

\section{Study regions and tree cover
change}\label{study-regions-and-tree-cover-change}

We use two areas in Australia to demonstrate the approach. In Australia,
significant tree cover change has mainly occurred in the north east of
the continent and on the southeast coast, as well as in the southwest of
Western Australia. According to the National Dynamic Land Cover Dataset
(DLCD) (Lymburner et al. 2010), most of these areas have experienced
decreasing EVI post 2000. As an index for vegetation greenness, the
decreasing values indicate lower biomass over time in the tree cover
regions. The possible EVI reduction might be due to land clearing, bush
fires or drought.

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/map_selreg} \caption{Selected study regions are highlighted by red rectangles in the main map (the red rectangle in the insert indicates the location of the main map). The types of tree cover in 2008 from the DLCD product is shown at the background. In site 1 (the QLD region), the tree cover is mostly sparse. In site 2 (the NSW/VIC region), many areas have open or close forest in which tree cover is denser.}\label{fig:selreg}
\end{figure}

Two regions were selected where significant tree cover change was
reported. The first region is located in south central Queensland to the
north of the Murray Darling Basin (MDB) (site 1 in Figure
\ref{fig:selreg}. High rates of land clearing have been reported in this
region during the early 2000s (Department of Natural Resources and Water
2007). The second study region is located at the border of New South
Wales and Victoria, and includes the Snowy Mountain ranges (site 2 in
Figure \ref{fig:selreg}. Severe bush fires occurred in this area and the
surroundings in early 2003 (see Figure \ref{fig:figure2bushfire}). The
2003 bush fires were the largest and the worst in this area for the last
60 years (The State Government of Victoria 2011). Two thirds of
Kosciuszko national park was heavily burned and regrowth was reported to
be slow due to drought and cold conditions (ABC News 2003) and the type
of species in this region. However, in the longer term, after an early
high transpiration period a recovery of pre-fire evapotranspiration
would be expected (Kuczera 1987). As a result , significant tree cover
loss has happened in both study areas in the last decade, either
permanently or temporarily.

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/bushfire_nswvic} \caption{Location of bushfires occurring in January 2003, in and around the NSW/VIC study region, as shown by the red pixels. The map shows large area in the Kosciuszko national park has been burned. Some locations in the southwest of ACT have also experienced intensive bushfires.}\label{fig:figure2bushfire}
\end{figure}

The two regions have different climate characteristics. The QLD region
is partially grassland and partially subtropical, while the NSW/VIC
region is mainly within the temperate zone, under the Köppen
classification. According to Australian Bureau of Meteorology (BoM), the
NSW/VIC region receives 1000 - 2000 mm rainfall annually, which is more
than double of the rainfall in the QLD region. Evapotranspiration is
similar in both regions. Marine moisture and orographic effects are
likely to be the main contributors to rainfall in the southeast mountain
areas of the NSW/VIC region.

The land use and land cover characteristics in the two regions are also
different. In the Queensland region, the tree cover is sparse over most
of the area. The MODIS satellite tree cover data (discussed in more
detail in section 3) shows that tree cover in this region is generally
below 20\% of total ground area. Grazing is the main activity in this
region, with over 90\% of land used by the grazing industry (ABARES
2010). Our starting assumption is that the main cause of the EVI decline
over large part of the region is due to land clearing. Tree cover has
been cleared at a massive scale over the last decade, especially during
2002 - 2004. The Kosciuszko national park is within the NSW/VIC region.
Here tree cover is denser with open or even closed forest (the tree
cover distribution is bimodal at 10 - 20\% and 60 - 70\%). The dominant
species in the alpine area are Snow Gum and large stand species such as
Alpine Ash and Mountain Gum in the sub-alpine area. These trees can
reach a great height but they take long time to grow. For example,
Alpine Ash would need about 20 years to mature. Although land clearing
is not a major issue in this region, it is vulnerable to fires and
drought.

Therefore two types of land cover changes were studied. The reports from
the Queensland Statewide Land Cover and Trees Study (SLATS) (e.g.
Department of Natural Resources and Mines 2005; Department of Science,
Information Technology and Innovation 2017) were used to investigate the
time and location of land clearing in the QLD region. The MODIS burned
area product, MCD45A1 (Roy, Lewis, and Justice 2002; Roy et al. 2005;
Roy et al. 2008), was used to locate bush fires areas in the NSW/VIC
region, with a grid resolution of 500 m. MCD45A1 provides monthly
burning information on all pixels, which helps to pinpoint an abrupt
event. Due to the nature of the different land cover change, the
post-change vegetation status in the two regions is expected to be
different (see Figure \ref{fig:figure3-tc-simple}).

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/tc_simple} \caption{The expected evolution of the land surface after trees have been removed in (a) the QLD region and (b) the NSW/VIC region.}\label{fig:figure3-tc-simple}
\end{figure}

The hypothesis in this study therefore is that the effect of 2003 - 2004
land clearings in the QLD region and the 2003 bush fires in the NSW/VIC
region cause a step change in the local rainfall. The actual tree cover
change at the pixel level during this time was derived from the 15-year
MODIS data (discussed below). The difference of tree cover before and
after the land disturbance was tested using a Student's t-test. As the
length of the tree cover data is shorter than available rainfall data,
earlier land clearings in the QLD region cannot be identified spatially,
hence they are excluded from the analysis.

\hypertarget{Data}{\subsection{Data}\label{Data}}

Several land surface data sets were used in this study. The main one was
the MOD44B product Global Vegetation Continuous Field data set (version
5). This data set provides estimates of percent tree cover (percentage
of ground surface covered by trees) at a grid resolution of 250 m
(Townshend et al. 2011), which is finer then the earlier mentioned
burned product MCD45A1. The data set is available on an annual time
interval for the study period of 2000 - 2015. The tree cover data was
produced from 16-day Terra MODIS Land Surface Reflectance data and Land
Surface Temperature (Townshend et al. 2011). The National Dynamic Land
Cover Dataset (DLCD) (Lymburner et al. 2010) from the Australian
Collaborative Land Use Mapping Program (ACLUMP) was used to verify the
trend of vegetation cover change calculated from the previous data set.
This data set, developed by Geoscience Australia and Australian Bureau
of Agricultural and Resource Economics and Sciences (ABARES), is the
first nationally consistent and thematically comprehensive land cover
reference for Australia. The DLCD is based on the 16-day Enhanced
Vegetation Index (EVI), again from the MODIS satellite, between April
2000 and April 2015. It also has a grid resolution of 250 m. The data
set provides information on the final land cover types (as in 2015) and
estimated trend of EVI statistics (annual mean, maximum and minimum).

Rainfall data for Australia (Jones, Wang, and Fawcett 2009) was obtained
from the Bureau of Meteorology. The data has been projected onto a
national 0.05\textdegree × 0.05\textdegree grid (approximately 5 km × 5
km). This gridded data set was generated from station observations using
an optimised Barnes successive correction technique. The Barnes
technique combines a weighted averaging process and defined
topographical information to estimate rainfall values between spatial
points (BoM 2009). The resulting data set provides additional
information for data-sparse areas like central Australia but reduces
information in the data-rich areas, such as southeast Australia where
station density is up to 20 per 100 \(\text{km}^2\). The data is
available on a monthly basis from 1900 to current. Here a subset of 36
years (1979 - 2015) was used. The study was conducted on monthly data,
as a land cover change effect on annual rainfall might be negligible but
can often found to be significant in particular months or seasons (e.g.
Otterman et al. 1990; Gaertner et al. 2001; Semazzi and Song 2001;
Oleson et al. 2004; Deo et al. 2009).

Large scale climate drivers are represented by various climatic indices.
The Southern Oscillation Index (SOI) is generally regarded as a good
predictor of Australian rainfall (Risbey et al. 2009; Chowdhury and
Beecham 2010; Westra and Sharma 2010), but its skill is weaker in some
parts of Australia. For example the Southern Annular Mode (SAM) is found
to be more important than ENSO in south Western Australia (Meneghini,
Simmonds, and Smith 2007). The suitability of each index for the regions
of interest was tested in section @ref(sec:reg\_model). The following
climate indices were used as candidate predictors for local rainfall.

\begin{itemize}
\tightlist
\item
  Southern Oscillation Index (SOI). The Troup version of the monthly SOI
  series used in this study was obtained from BoM (available online at
  \url{http://www.bom.gov.au/climate/current/soihtm1.shtml}).\\
\item
  Eastern, East Central and Central Tropical Pacific Sea Surface
  Temperatures (NINO 3, NINO 3.4 and NINO 4). Monthly SST anomalies are
  available from IRI/LDEO data library and the extended NINO data set is
  used (available online at
  \url{http://iridl.ldeo.columbia.edu/SOURCES/.Indices/.nino/.EXTENDED/}).\\
\item
  Pacific Decadal Oscillation (PDO). The Pacific Decadal Oscillation is
  the leading principal component of monthly SST anomaly in the North
  Pacific Ocean.. The monthly PDO series was provided by JISAO (Joint
  Institute for the Study of the Atmosphere and Ocean, University of
  Washington) (available online at
  \url{http://jisao.washington.edu/pdo/PDO.latest}).\\
\item
  Indian Ocean Dipole (IOD). The Indian Ocean dipole is commonly
  measured by the difference between SST anomaly in the western (50 -
  70\textdegree E and 10\textdegree S-10\textdegree N) and eastern (90 -
  110\textdegree E and 0 - 10\textdegree S) equatorial India Ocean (Saji
  et al. 1999). Monthly IOD was obtained from JAMSTEC (the Japan Agency
  for Marine-Earth Science and Technology) (available online at
  \url{http://www.jamstec.go.jp/frcgc/research/d1/iod/DATA/dmi.monthly.txt}).
\end{itemize}

\subsection{Statistical method}\label{statistical-method}

The step changes on rainfall were analysed using two different
statistical methods. A step change is not obvious in the time series
data of the rainfall residuals (Figure \ref{fig:ts-mean}), even though
the data is deseasonalised and detrended. Both methods in this paper
make use of a regression model to remove variability in rainfall due to
climatic factors to strengthen the tree cover change signal. In the
first method, the tree cover change was implemented as a factor variable
in the regression model. In the second method, a rank sum test (step
trend test), was applied to the model residuals after effects of other
major factors were removed. This assumes that after removal of all
climate variation, the vegetation cover change is the only factor
explaining the non-random pattern in the rainfall residuals.

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/Rainfall_resid} \caption{The deseasonalised and detrended rainfall over the 30 years period in (a) the QLD region and (b) the NSW/VIC region. The vertical red lines indicate the year of 2003, in which the studied land cover changes occurred. A change in the time series data is not obvious before and after the land cover changes.}\label{fig:ts-mean}
\end{figure}

\subsubsection{Regression model}\label{reg_model}

As highlighted in the introduction, the Australian climate is influenced
by sea surface temperatures in the tropical Pacific and Indian Oceans,
as well as pressure systems in the Southern Ocean (BoM 2012b). Risbey et
al. (2009) compared five large-scale drivers, including ENSO (measured
by SOI and the Tropical Pacific SSTs), IOD, SAM, MJO (Madden-Julian
oscillation) and blocking, in relation to Australian rainfall
variability. The MJO is a large scale eastward-propagating wave-like
disturbance in equatorial latitudes (Risbey et al. 2009). They
identified SOI as the most important index among all indices tested for
broad parts of Australia (including QLD and NSW/VIC) in almost any
season. In this study, four indices were selected from the main climatic
indicators (see section \protect\hyperlink{Data}{Data}) were used as the
explanatory variables in the model for each study region.

Correlations between rainfall and each climate index were analysed.
Rainfall in each study region was first deseasonalised and detrended
using the seasonal decomposition function \texttt{ds} in the package
\texttt{deseasonalise} in R (R Core Team, 2018). Using detrended data
gives a better indication of the underlying correlation rather than the
correlation between trends in the data (Smith and Timbal 2012). The
cross-correlations between the deseasonalised and detrended rainfall and
the climatic indices were tested using the Pearson's product moment
correlation method, assuming the relationships are linear. Although the
optimal technique for exploring the correlation with each index could be
different as described in Risbey et al. (2009), the Pearson's method was
applied to all indices for consistency. Because the PDO describes the
multi-decadal SST with lower frequency (MacDonald and Case 2005;
Zanchettin et al. 2008; Kamruzzaman, Beecham, and Metcalfe 2011),
instead of 37-year rainfall data, a longer period (115 years, from 1900
to 2015) was used to estimate the correlation with PDO, up to lag 24.
For the other indices, the 37-year data was used.

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/cor_qld} \caption{Cross-correlation of six climate indices and rainfall in QLD study region. For the PDO analysis, 108-year rainfall data (1900 - 2008) are used. Otherwise, 36-year rainfall data are used. The correlation with NINO 3 is not shown as it is very similar to but weaker than for NINO 3.4.}\label{fig:cor-rain-qld}
\end{figure}

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/cor_nswvic} \caption{Cross-correlation of six climate indices and rainfall in NSW/VIC study region.}\label{fig:cor-rain-nsw}
\end{figure}

Based on the correlation between the climatic indices and rainfall (as
shown in Figure \ref{fig:cor-rain-qld} and Figure
\ref{fig:cor-rain-nsw}), it can be concluded that:

\begin{itemize}
  \setlength{\itemsep}{0cm}
  \setlength{\parskip}{0cm}
  \item In QLD, the correlation between rainfall and SOI at zero time lags is the highest across all indices, outweighing the other ENSO indicators. IOD, had a weak influence in QLD.
  \item In NSW/VIC, again the SOI has the highest correlation with rainfall, followed by the IOD. Both occur at the zero time lags. 
  \item in both cases PDO had the weakest correlations, and this factor was not further considered as a predictor.
\end{itemize}

The above findings are consistent with previous studies. Although some
indices are serially correlated with rainfall up to several months, the
lag zero events have the most significant correlation coefficients.
Concurrent climatic index series were generally found most useful in
rainfall prediction (e.g. Risbey et al. 2009; Kamruzzaman, Beecham, and
Metcalfe 2011). The correlations between the climatic indices and
rainfall for each individual season have also been tested, and the
results were similar.

Rainfall in Australia shows strong seasonal patterns (Holper 2011;
Australian Bureau of Statistics 2012). For example, the north part of
the country is summer rainfall dominant with a dry winter, while most of
the southern part has a winter rainfall regime. This character is driven
by the movement of subtropical high pressure systems which dominate the
Australian climate (BoM 2012a). These characteristic summer and winter
rainfall patterns mean that the seasonal component of rainfall has a
periodic pattern which should be included in the model.

Long term trends in the regional rainfall in some parts of Australia are
significant (Hughes 2003; Gallant, Hennessy, and Risbey 2007; Chowdhury
and Beecham 2010). In the northern and eastern parts of the continent,
increasing rainfall is reported over the last century (Hughes 2003). The
presence of such long term trends may be confused with the outcome of a
step change in rainfall. As a result a linear trend term was implemented
in the model to remove any long term effect.

We assumed all the factors are additive smooth components in determining
rainfall following Kamruzzaman, Beecham, and Metcalfe (2011). Generally,
monthly rainfall has a skewed distribution so the normality assumption
of the residuals in a general linear model could be violated. In this
case, the rainfall model is a generalised additive model (GAM) (Hastie
and Tibshirani 1986) with a log link function \texttt{g()} and assuming
the residuals are gamma distributed (see Figure \ref{fig:hist-rain}).
This means all predictors are modelled as smooth functions, in this case
using the shrinkage version of the cubic regression splines (Wood 2011).
\vspace{0.5cm}

\begin{equation}
\begin{array}{lll}
g(E(\mathbf{R}_r)) = &\beta_0 + s_1(\mathbf{SOI}) + s_2(\mathbf{IOD}) + \\                &s_3(\mathbf{Nino3.4}) + s_4(\mathbf{Nino4}) + s_5(\mathbf{Season}) + \\
           &\beta_1\mathbf{Trend} + \boldsymbol{\epsilon}_r
\end{array}
\label{eq:model}
\end{equation}

The bold letters represent the time series vectors. The region is
indicated by \(r\), while \(\beta_u\) (\(_u\)=0, 1) are the fitted
coefficients in the model. \(s_v\) (\(_v\)=1, 2, 3) are the smooth
penalized cubic regression spline functions on the climatic indices and
the season. Apart from dropping the PDO as a predictor, all other
climate indices were included, allowing the model to select the
appropriate predictors.

A possible linear long term trend in the rainfall data is modelled by
\textbf{Trend} = 1,2,3\ldots{}n, where n is the total number of months
in the time series. \textbf{Season} is the seasonal component. The
climatic terms are also modelled with smooth functions. The effect of
large scale drivers on Australian rainfall is more likely to be seasonal
(Murphy and Timbal 2008; Schepen, Wang, and Robertson 2012), and the
smooth spline function is more flexible in reproducing the variability
in impacts of the climatic indices.

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/hist_rainfall} \caption{Distribution of monthly rainfall in (a) QLD and (b) NSW/VIC. Using a Kolmogorov-Smirnov test with shape = 1 and 2.4  for QLD and NSW/VIC respectively, rainfall in both regions can be modelled as a gamma distribution.}\label{fig:hist-rain}
\end{figure}

\subsection{Tree cover change as factor variable}

One of the main difficulties in empirical observation studies on the
effect of land cover change on rainfall is the lack of continuous
monitoring of land surface variables, or even, that no specific variable
can be defined that can clearly represent the land surface process.
Given the lack of a full picture of the land surface process, a factor
variable was used in this study to represent the abrupt land surface
change (see Equation \eqref{eq:model}). The change could be a result of
either land clearing or bush fires as long as it is permanent or takes a
long time to recover. Here we approached the problem with two different
models.

In the first method, the tree cover change was used as a predictor in
the regression model, represented by a factor variable \textbf{LC}. The
significance of the coefficient of \textbf{LC}, denoted as \(\beta'_5\)
in Equation \eqref{eq:model}, can be determined by a ratio test.

\begin{equation}
  \mbox{LC} = \left\{
  \begin{array}{ll}
     \mbox{Trees} \\
     \mbox{Removed}
  \end{array} \right.\\
  \label{eq:dummyC}
\end{equation}

Therefore in both regions, land cover is ``trees'' for the period before
land cover change and ``removed'' for the period after the change. Here
we simply assumed that vegetation cover change has occurred on every
pixel. The remaining term \(\epsilon_r\) is the amount of rainfall that
is attributed to other unspecified factors and random errors. Hence the
regression model becomes \vspace{0.5cm}

\begin{equation}
\begin{array}{lll}
g(E(\mathbf{R}_r)) = &\beta'_0 + s'_1(\mathbf{SOI}) + s'_2(\mathbf{IOD}) + \\
  &s'_3(\mathbf{Nino3.4}) + s'_4(\mathbf{Nino4}) + s'_5\mathbf{Season} + \\
  &\beta'_1\mathbf{Trend} + \beta'_2\mathbf{LC} + \boldsymbol{\epsilon'}_r
\end{array}
  \label{eq:model2}
\end{equation}

One of the difficulties is to point an exact time to the changes in the
vegetation cover in the two regions. In the QLD region, no exact time
can be assigned to the land clearing. According to the SLATS reports,
the most substantial clearing occurred between 2003 - 2004 . However,
the information on the change in type of land cover during the time
period is missing. Therefore, four scenarios were initially tested in
the analysis. In these scenarios the ``after change'' period started
from: (1) June 2003, (2) January 2004, (3) June 2004 and (4) January
2005. A further complicating factor is the influence of the ``millenium
drought'' over the study period and in particular to change to wet
conditions in 2010 - 2011 (Dijk and Viney 2013). In the NSW/VIC region,
severe bush fires were reported in early January 2003. Hence the
``tree'' cover state was up to December 2002 then it was changed to
``removed'' state from January 2003. However, there is similar
uncertainty about when the tree cover had recovered from the bushfire
and how this interacted with climate variability. As a result several
scenarios were tested for both locations: 1) analysis up to 7 years post
change (assumed to be 2010 for both regions); and 2) analysis of the
full period (2015). More detail about this is later in the methods. As a
starting date, the regression model was run from 1979 for both regions.

\subsection{Step trend test}\label{step-trend-test}

To support the regression analysis, a step trend test was used to detect
changes in rainfall as a result of vegetation cover change. This
nonparametric statistical test was modified from the Mann-Whitney
Rank-Sum test by (Hirsch and Gilroy 1985) and can identify a step change
in data which is cross correlated. The gridded rainfall dataset used in
this study has a high spatial correlation between neighbouring pixels
due to the underlying interpolation method. The advantages of using the
Rank-Sum test therefore are: (1) it does not depend on assumptions of
the data distribution; (2) it is not restricted to datasets with no
missing data; (3) it is robust and not as easily influenced by outliers
and negative numbers (Hirsch and Gilroy 1985).

The rainfall residuals from the regression model in Equation
\eqref{eq:model2} were used for the step trend test. According to (Hirsch
and Gilroy 1985), to detect a step change, using deseasonalised and
detrended data is important. Furthermore, as rainfall can only be
partially attributed to local sources and conditions, other effects are
introduced by large scale dynamics and changes in other climatic
factors. The assumption is that the regression model should remove these
effects, and additionally deseasonalise and detrend the rainfall data.
As a result, the local landuse effects are amplified in the variation of
the remaining residuals. The test, described in the following sub
section, subsequently associates trends in the rainfall residuals with
tree cover changes.

\subsubsection{Mann-Whitney rank-sum
statistic}\label{mann-whitney-rank-sum-statistic}

As indicated, the step trend test is a modified version of the
Mann-Whitney rank-sum statistic (Hirsch and Gilroy 1985). As a
nonparametric rank-based test, the Mann-Whitney test does not use the
exact values of rainfall but depends on the ranks of the data. For each
month, rainfall residuals of each year were ranked in an ascending
order. The ranking of January rainfall in a sample pixel k in QLD is
illustrated below:

\begin{longtable}[]{@{}ccc@{}}
\caption{Example of ranking rainfall residuals}\tabularnewline
\toprule
Year & Rainfall residuals & Rank \(R'_{1k}\)\tabularnewline
\midrule
\endfirsthead
\toprule
Year & Rainfall residuals & Rank \(R'_{1k}\)\tabularnewline
\midrule
\endhead
1998 & -0.3 & 6\tabularnewline
1999 & -60.9 & 2\tabularnewline
2000 & -16.1 & 4\tabularnewline
2001 & -71.7 & 1\tabularnewline
2002 & 111.1 & 7\tabularnewline
2005 & -7.2 & 5\tabularnewline
2006 & -60.5 & 3\tabularnewline
\bottomrule
\end{longtable}

\noindent Therefore, the smallest or most negative value has rank 1 and
the largest value has the maximum rank.

The before and after period in the data formed two groups of samples.
The split point of the two periods was based on the timing of the
vegetation cover changes. In the QLD region, changes occurred anytime
during 2003 and 2004. In contrast to the previous method, the time
period covering the land cover change was excluded, as the nonparametric
test allows missing data. Hirsch and Gilroy (1985) also pointed out that
the power of the test is higher if the data of the change period is
ignored. Hence 2003 and 2004 were excluded from the analysis. As a
result, the after-change period was 2005 - 2015 for the Queensland
location.

In the case of NSW/VIC, the bushfires broke out in early January 2003.
The change was within a relatively short period of the year. Therefore
the after change period in this region still started in January 2003.
Following Hirsch and Gilroy (1985), the before period was set to five
years (1998 - 2002) in both regions. The length of the after period is
difficult for the NSW/VIC as the regrowth would at some point have
impacted the local effects.

There is a further complication in the data. The year 2009 was the final
year of the millenium drought (Dijk and Viney 2013), followed by two
very wet years (also visible in Figure \ref{fig:ts-mean} in 2010 and
2011. As highlighted earlier, this means the the length of after period
could influence the analysis by the inclusion of more or fewer years. as
a result the period up to (and including) 2009 was compared to the full
period up to 2015. Further analysis up to 2010 is included in the
supplementary material.

The rank of rainfall in month j year i in pixel k is denoted as
\(R'_{ijk}\). The sum of ranks of rainfall in month j in pixel k before
the known intervention is:

\begin{equation}
  W_{jk} = \sum_{i=1}^{n_1}R'_{ijk}.
  \label{eq:Wj}
\end{equation}

\(n_1\) is the number of years before the land cover change. The
expected value of \(W_{jk}\) is

\begin{equation}
  \mu_w=n_1(n_1+n_2+1)/2
\end{equation}

\(n_2\) is the number of years after the change. Hence the expected
value of the rank sum before the intervention is the same for all months
and all pixels. The sum of ranks for the whole time period is fixed, as
\((n_1+n_2)(n_1+n_2+1)/2\). In this study, since there are only two
groups (before and after), knowing the rank-sum of one group is the same
as knowing the rank-sum of the other group. If the rainfall data is
temporally and spatially independent, the variance of \(W_{jk}\) is

\begin{equation}
  \sigma^2_w = n_1\cdot n_2(n_1+n_2+1)/m
\end{equation}

where m is the number of months which is 12 in the case of a full year.

\subsubsection{Step trend test}\label{step-trend-test-1}

Instead of completing the Mann-Whitney U-test, Hirsch and Gilroy (1985)
applied a standard normal Z test to the rank-sum statistics. As
highlighted, this modified test accounts for serial and cross
correlation in the data. In the case here, the deseasonalised and
detrended data shows little autocorrelation in the time series but
possesses strong cross correlation between neighbouring pixels, i.e.
\(R>0.99\).

The sum of \(W_{jk}\) for a block of \(ns\) pixels over the whole year,
\(\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk}\), has mean

\begin{equation}
  E(\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk})=12\cdot ns\cdot\mu_W
\end{equation}

and variance

\begin{equation}
  Var(\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk})=\sum_{j=1}^{12}\sum_{k=1}^{ns}\sum_{h=1}^{ns}C(W_{jk},W_{jh}).
\end{equation}

\(C(W_{jk},W_{jh})\) is the covariance of the W statistics between pixel
k and pixel h in month j. When \(k=h\), \(C(W_{jk},W_{jh})=\sigma^2_w\).
When \(k\neq h\),

\begin{equation}
  C(W_{jk},W_{jh})=\sigma^2_w r(R_k,R_h)
\end{equation}

where \(r(R_k,R_h)\) is the product moment correlation coefficient of
the concurrent ranks in pixel k and h. Here \(r\) is calculated on the
full time series in each pixel. In the analysis, the test was applied to
a square block of four pixels each time. As argued by Hirsch and Gilroy
(1985), \(ns=4\) is the most optimal solution to balance the cost and
the gain in the test power.

The statistic of the step trend test is then defined as

\begin{equation}
  Z'=\frac{\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk}-12\cdot ns\cdot\mu_w}{\sqrt{Var(\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk})}}.
  \label{eq:Z}
\end{equation}

The above statistic is written for a 12 month period. By changing the
value 12, it can also be used to test seasonal rainfall change or for
other customized periods.

The null hypothesis (\(H_0\)) in this study is that there was no change
in rainfall due to land surface intervention. The results of the step
trend test can be interpreted according to the sign of the Z' score (see
Table \ref{tab:Zscore} (Chapter 23, P887 Hipel and McLeod 1994)). Z' is
normally distributed similar to the standard normal statistics Z. Hence
it can be compared to a standard normal distribution to determine the p
value.

\begin{longtable}[]{@{}c@{}}
\caption{The interpretation of Z' score in the step trend
test}\tabularnewline
\toprule
Z' \(>\) 0 \& rainfall decreases after change\tabularnewline
Z' \(<\) 0 \& rainfall increases after change\tabularnewline
Z' = 0 \& rainfall does not change\tabularnewline
\bottomrule
\end{longtable}

\section{Results}\label{results}

\subsection{Tree cover change}\label{tree-cover-change}

The pixels, where the tree cover change based on the MOD44B data was
significant (\(p \leq 0.05\)) in each study region, are shown in Figure
\ref{fig:tctrend} for the time series to 2009 (left panels) and the time
series to 2015 (right panels). In both cases it can be seen that the
area of negative cover change was greater in the series to 2009 than in
the series to 2015. But even the longer series indicates a large change
in the NSW/VIC region (bottom row). In the NSW/VIC region, much of the
tree loss between 2002 and 2003 was concentrated in the Snowy Mountains
close to the border of NSW and VIC. Tree cover loss occurred in large
parts of the QLD region between 2002 and 2005, but this tree loss was
spatially less concentrated. Most of the clearings appear in the centre
of this region in the series to 2009. The tree cover change map
(\ref{fig:tctrend} is consistent with the annual mean EVI trend map
(based on DLCD data, not shown here). However in the series to 2015 the
tree cover loss seems to have dissappeared (top right Figure
\ref{fig:tctrend}) and suggests an increase in tree cover over the area.
The implications of this will be discussed in more detail in the
discussion.

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/tc_figs} \caption{The maps show signifcant changes in tree cover identified from the MOD44B data between 2003 and 2009 (top left) and 2003 and 2015 (top right) in the NSW/VIC region and the Qld region (bottom left to 2009 and bottom right to 2015). The amount of change was calculated as the difference in tree cover before and after the specified land cover intervention and it is shown as the percentage of the ground area. Green colour indicates an increase in tree cover, while red colour indicates a decrease in tree cover.}\label{fig:tctrend}
\end{figure}

\subsection{Regression Model and significance of Vegetation Cover
Changes}\label{regression-model-and-significance-of-vegetation-cover-changes}

Generally the regression model only explains a limited amount of the
rainfall variability. The model in Equation \eqref{eq:model2} accounts for
around
15\%\footnote{Here the adjusted $R^2$ was reported. Adjusted $R^2$ is the coefficient of determination, a measurement of the amount of variability predicted by the model adjusting for the number of explanatory terms}
of the rainfall variation in both regions. The residual analysis shows
that the assumptions of the regression model are generally met (Figure
\ref{fig:residuals}. The standardised residual plots, however, show some
funnelling for the NSW/VIC regions, suggesting non constant variance.
The residual patterns are consistent for all pixels within each region.

The model, however, confirms the importance of the climate drivers and
the seasonality in Australian rainfall. Even at the grid level, the
seasons and several of the climatic indices were significant
(\(p \leq 0.05\)) everywhere in both regions. The explaining power of
the model is mostly due to these variables. The climate drivers (at lag
zero) accounted, on average, for 6.7\% of the rainfall variability in
both the QLD region and the NSW/VIC region (see Figure \ref{fig:rsq} for
the distribution of \(R^2\) in these two regions). These figures are
within the upper bound of seasonal rainfall predictability by a SST
anomaly field reported by (Westra and Sharma 2010).

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/gam_check} \caption{The residual analysis of a sample pixel in the QLD region (top) and NSW/VIC region (bottom).}\label{fig:residuals}
\end{figure}

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/soi_explhistogram} \caption{The performance of the regression model if rainfall is only modelled by the climate drivers. It shows the percentage of rainfall variability that can be explained by the climate drivers for the Qld and NSW/VIC region}\label{fig:rsq}
\end{figure}

Statistically significant long term trends were only observed in part of
the NSW region (results not shown). However, this result might prove or
disprove the existence a long term trend in rainfall. The overall time
period is fairly short (Koutsoyiannis 2006) and more pixels in NSW/VIC
could indicate a significant step change if the long term trend effect
is not removed by the model. As trend free data is an important
requirement for the step trend test, the trend term was kept in the
regression model to ensure the detection of step change was not due to a
possible long term trend.

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/Cp_30yrs} \caption{The spatial distribution of significance of land cover step change variable in the GAM model predicting changes in rainfall in the both regions. The top row reflects NSW/VIC, while the bottom relates to the Qld study site. The left hand plots are the data up to 2009, while the right hand plots are for the full series. The p value reported is for the land cover variable in the model.}\label{fig:LCp}
\end{figure}

The land cover variable in the model (Equation \eqref{eq:model2}) aims to
identify a step change in the rainfall before and after the observed
change in land cover. The variable was only significant
(\(p \leq 0.05\)) for the rainfall estimates in some areas in NSW/VIC,
as shown in Figure \ref{fig:LCp}. However, the number of pixels where
the landcover variable was significant was much greater for the series
up to 2009, compared to the series to 2015. There was no direct
relationship between the areas of bushfires in Figure
\ref{fig:bushfire}. No significant step change due to the land cover
changes was found in rainfall in the QLD region in the series to 2015,
but a small area of change was identified in the period to 2009 (left
bottom panel).

Figure \ref{fig:LCp} for NSW/VIC suggests that a significant step change
in rainfall related to land cover change was found in an area larger
than where the bushfires has occurred. This might be showing a large
scale effect that reaches beyond the vegetation cover change effect.

More generally, the model suggests that the tree cover has a positive
impact on rainfall in both regions. The fitted coefficients for the Land
cover change variable were consistently positive for the ``tree'' part
of the series. It implies that rainfall was higher when the surface was
covered by trees.

\subsection{Step Trend Test}\label{step-trend-test-2}

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/step_new} \caption{Spatial distribution of the step trend test Z' statistics in the two study sites. Panels on the top are for the NSW/Vic site, while panels of the bottom ar for Qld. The left panels are for the series to 2009, while the right panels are for the full data series. Warm colours (yellow, orange and red) are for positive Z' values which indicate decreasing rainfall trend due to the land surface intervention. Cold colours (light blue to blue) are for negative Z' values which indicate increasing rainfall trend. The deeper the colour, the more significant the statistic.}\label{fig:steptest30}
\end{figure}

The spatial step trend test Z' scores for the Qld region (top row) and
NSWVIC region (bottom row) are shown in Figure \ref{fig:steptest30}. The
figure also contains the period to 2009 (left panels) and the period to
2015 (right panels). This figure provides two types of information: the
sign and the significance level. The sign indicates the direction of the
step change, as listed in Table \ref{Zscore}. In each region up to 2009,
there is a broad area of positive Z' values which implies a decrease in
rainfall. There appears to be little relationship between the locations
where changes in tree cover are observed (Figure @ref(fig:tc\_trend))
and the patterns in the Z' score. However, there is not necessarily a
direct relationship as movement of air masses could mean that actual
changes of rainfall are observed close by, but not necessarily exactly
at areas with changes of landcover. There is once again a difference
between the left panels and the right panels, indicating that
considering a longer time series (to 2015) reduces the pattern and
significance of Z' scores. In the QLD region to 2009, 21\% of the pixels
obtained a positive Z' score with p \(<\) 0.1, and for the period to
2015 no pixels with positive Z' score at p \(<\) 0.1 occurred. In the
NSW/VIC region to 2009, only 2.6\% of pixels in the alpine area having a
positive Z' score with p \(<\) 0.1. In general it is only a small
proportion of both study regions.

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/ResidualBoxplotchange} \caption{Boxplots of annual rainfall residuals (estimated based on Equation 2 before and after the land cover intervention during 1979 - 2015 in the study regions. On average, the after period has a significantly lower annual rainfall residual in NSW/VIC, but a signifcantly higher annual rainfall residual in the Qld study area}\label{fig:meandiff}
\end{figure}

The rainfall residuals of the two periods (before-change since 1979 and
after-change) were also compared (Figure \ref{fig:meandiff}) using a
simple t-test. From the boxplots it is difficult to see that there were
significantly different (\(p < 0.05\)) mean values between the
``before'' and ``after'' periods in both regions. For the Queensland
locations, there was no significant difference if measured to 2009 and
slightly less rain (\(p < 0.05\)) before if measured to 2015
(\(~ 2 mm/month\)). In contrast, for the NSW/VIC locations there was
also slightly more rain (\(p < 0.05\)) after the change if measured to
2015 (\(~ 1 mm/month\)), but significantly less rain (\(p < 0.05\))
after the change (\(~ 9 mm/month\)) if measured to 2009.

A closer look at the rainfall data indicates that rainfall was
consistently very low in 2006 for both regions and consistently high for
2010 \ref{fig:ts-mean}. The low rainfall in 2006 was due to a weak El
Niño and unrecovered conditions from the previous drought, in contrast,
2010 brough record drought breaking rains across Eastern Australia
related to a La Niña (Dijk and Viney 2013). While the regression model
has removed most of the effect of ENSO, the effect of these anomolous
years could still be visible in the residuals if the response of the
rainfall to the ENSO effect is non-linear, and for example the memory of
the past drought is persistent. This can also explain the significant
results in the previous method. To remove the possible outlier effects
from 2006 and 2010, the two periods were compared excluding the 2006 and
2010 rainfall. In addition, the group of pixels showing a negative step
change in rainfall (\(p < 0.10\)) and the group of pixels indicating no
change were analysed separately for the period to 2009. An unpaired
unequal variance two sample t-test was applied to test whether the after
period had lower mean rainfall than the before period, excluding the
2006 and 2010 years for the full period, and excluding 2006 for the
period to 2009.

For the Queensland locations using the regression model residuals and
excluding 2006 and 2010, there was no difference between the period
before and after the landuse change for the full data series. However
for the period to 2009, there is slightly more rain
(\(p < 0.05, ~ 2mm/month\)) before than after the change. For the
NSW/VIC locations for the same analyses, there is slightly more rain
(\(p < 0.05\)) after the change if measured to 2015 (\(~ 2 mm/month\)),
and for the period to 2009 there is still significantly less rain
(\(p < 0.05, ~ 5 mm/month\)). Focussing on the pixels with a negative
step change (positive significant Z' score) and the monthly rainfall
totals, only the period to 2009 is of interest, as the full period had
too few significant pixels (\ref{fig:steptest30}). For both regions and
excluding 2006, the Qld area had significantly more rain before than
after at \(p < 0.05\). For the NSW/VIC area, there was similarly
significantly more rain before than after the change
(\(p < 0.05 and 100mm/month\)). For pixels where the Z' score was not
significantly positive (\(p > 0.1\) and Z' score negative), the Qld area
showed no significant difference in monthly rainfall totals before and
after. For the NSW/VIC area there was still significantly less rain
after the change (\(p < 0.05\)) but difference in monthly rainfall
totals was smaller than for positive and signifcant Z' scores: 24
mm/month less.

The choice of \(ns\) has some impact on the test results (Hirsch and
Gilroy 1985). The cases of \(ns = 1\) and \(ns = 9\) were also tested.
The results for \(ns = 1\) showed a lower number of significant
pixels(at p \(<\) 0.10) compared to the \(ns = 4\) test, with only 66
pixels in the Queensland area in the period to 2010 and a slightly
higher 6 pixels in the period to 2015. No pixels were significant at p
\(<\) 0.1 in the NSW/VIC area. The results from the \(ns = 9\) test was
not different from the \(ns = 4\) test. The power of the test does not
change much after \(ns = 4\), as shown by Hirsch and Gilroy (1985).

As part of the analysis, the ``field significance'' of the Z' score test
was considered to better interpret the step change at regional scales
from multiple local tests (Wilks 2006, Westra, Alexander, and Zwiers
(2013)). Here, the bootstrapping resampling method from Westra,
Alexander, and Zwiers (2013) was used to evaluate the field significance
for the period up to 2009 (as the period to 2015 essentially showed no
significant positive Z' scores). This means the spatial structure of the
pixels was maintained, but the order of the years and months was changed
by random resampling. For each resampling the test statistics identifies
the percentage of the pixels with significant positive step change for
the step trend test. The test statistics on 1000 resampled replicates
were used to develop the distribution of these percentage values under
the local null hypothesis that there was no step change. The results
showed that none of the resampled series indicated any significant step
trend for both regions, suggesting that the actual values found are for
a unique series of years.

\section{Discussion}\label{discussion}

\begin{table}

\caption{\label{tab:summarytable}Summary table of all tests on the two regions and for the two time periods}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lllll}
\toprule
Test & Qld 2009 & Qld 2015 & NSW/VIC 2009 & NSW/VIC 2015\\
\midrule
LC variable & small area on west side & no pixels significant & Large area & Smaller area than 2009\\
t-test regression \textbackslash{}\& model residuals & not significant & less rain before & less rain after & slightly less rain after\\
t-test residuals \textbackslash{}\& excluding 2006 and 2010 & slightly more rain after & no significant difference & significantly less rain after & slightly more rain after\\
Z' score & 21\% of pixels at p < 0.1 & 1 pixel at p < 0.1 & 2.6\% of pixels at p < 0.1 & no pixels at p < 0.1\\
t-test pixels positive Z' score & significantly less rain after & NA & 100 mm/month less rain after (p < 0.05) & NA\\
t-test other pixels & no difference & NA & 24 mm/month less rain after (p < 0.05) & NA\\
\bottomrule
\end{tabular}}
\end{table}

Overall the summary table of the results indicates that the effects of
land clearing or bushfire on rainfall are not easily detected, even with
a range of different statistical tests. In addition several of the tests
appear to indicate contradictory results, in particular in the
comparison between the two regions and between the two time periods.

Generally, empirical studies on LCC-precipitation interaction are
conducted within an area with known land surface intervention (e.g.
Otterman et al. 1990,Durieux, Machado, and Laurent (2003),Negri et al.
(2004),Sato, Kimura, and Hasegawa (2007)). However, these locations are
rare and difficult to isolate from real landscape change. Modelling
studies are abundant, but are generally not linked to observed data. In
this study we therefore tested the effect of land cover change across a
broad region, rather than only for locations where changes were known to
occur or have occured. The advantage of the suggested approach is that
it does not require a long time series of land cover data as this is
usually unavailable. Furthermore, it does not assume a specific
relationship between vegetation cover change and rainfall but allows the
data to show this relationship, by applying the analysis to a broader
area outside the boundary of the vegetation cover change. This approach
is expected to provide a way to reduce the risk of a false positive
paradox, by comparing results between areas with and without vegetation
cover change.

While there is some indication that the observed landuse changes (Figure
\ref{fig:tctrend}) cause a decrease in the rainfall, our analysis has
not been able to give an unequivocal answer. There are several possible
complicating fatcors in the data that could explain the differences
between the two regions and the two time periods.

\subsection{Rainfall variability}\label{rainfall-variability}

The first and most significant effect on the results remains the
variability in the rainfall. The regression model used here is a simple
model. We only consider the important effects of a historical trend,
seasonality and climate drivers. The purpose of using the regression
model is to remove the year to year and month to month variability in
rainfall and therefore strengthen the land cover signal in the
residuals. However, the model shows that trend, seasonality, ENSO and
IOD together explain no more than 30\% of the rainfall variability, and
only around 7\% on average is due to the climate drivers (Figure
\ref{fig:rsq} . And while this is consistent with the literature e.g.
Westra and Sharma (2010), a large amount of variation is left in the
model residuals. Rainfall is generally considered a stochastic process
(e.g. Fowler et al. 2005,Cowpertwait, Salinger, and Mullen (2009),Burton
et al. (2010)) and clearly some of the variability is either a different
response to a combination of climate factors (as interactions were not
tested in the model), or a strongly non-stationary response to the
climate drivers. The remaining variability in the residuals increases
the difficulty to detect a change in rainfall. This is clearly
demonstrated in the difference in land cover variable significance and
the Z' score between the series to 2009 and the series to 2015. While
2009 was at the end of the drought, the series to 2015 includes the
record breaking year 2010 (Figure \ref{fig:ts-mean}).

The severe bushfires in 2003 were also triggered by the extreme drought
conditions during the millenium drought (Dijk and Viney 2013). Although
the drought on rainfall has been accounted for in part by the model, a
further delayed or cumulative impact of drought could be feeding into
the local land-atmosphere interaction. As a result, the rainfall
feedback to the vegetation cover change could be weak under the dry
conditions, and this could have affected the result.

\subsection{Vegetation dynamics}\label{vegetation-dynamics}

The second possible effect is the dynamic nature of the vegetation
clearing and recovery. Although land clearing has occurred at a high
rate and broad scale in Queensland (Department of Science, Information
Technology and Innovation 2017), the clearing does not have a clear
start and end point. QLD has a long history of land clearing. According
to the series of SLATS reports on land cover changes in QLD released by
the Queensland government, land clearing continued in and around the
study region between 1988 - 2008. Major broad scale and high rate
clearings occurred in 1999 - 2000 and 2002 - 2004 (Figure
\ref{fig:slat}). And even though there was a decrease in land clearing
post 2005, it is difficult to define a clear cut change in this region.
The more continuous ongoing land clearing could have reduced the
significance of a step change.

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/slats} \caption{Woody vegetation clearing rate in QLD for the two major bioregions in the study area. The data were obtained from SLATS (2017), and clearly indicate a sharp decrease in the clearing rate after 2005.}\label{fig:slat}
\end{figure}

The two approaches used in this study indicate quite different results
on the level of change in the NSW/VIC region. The regression model
showed that a large area in the NSW/VIC region has experienced a
significant land cover effect (\(p < 0.05\)) on the rainfall after 2003.
The effect in the step trend test (Z' scores) is much mroe subdued, even
though it was able to detect significant changes (\(p < 0.1\)) within
the area in the period to 2009, but almost no effect in the period to
2015. In contrast, the results are almost reversed in the Qld area.
While some of this could be due to the difficulty in removing year on
year rainfall variability, some of this might also be due to vegetation
regrowth in both locations.

The specific vegetation class in the Queensland area is well-known for
rapid regrowth and ``thickening'' in favourable conditions (Gowen and
Bray 2016), and this could explain the change in vegetation cover
between 2009 and 2015 in the Qld area (Figure \ref{fig:tctrend}). In
particular the favourable rainfall years of 2010 and 2011 would have
boosted regrowth, increasing evapotranspiration and therefore decreasing
the effect of the rainfall change.

The different causes of vegetation cover change in these two regions
could lead to different post-change characteristics. The magnitude of
EVI decreasing trends in the QLD region are less than in NSW region, as
reported in the DLCD data. This is due to the lower tree density in the
QLD region than in the NSW/VIC region before land surface interventions.
The significant bushfires (Figure \ref{fig:fig2bushfire}) would have
drastically reduced the vegetation cover and recovery was very slow in
some areas within the NSW/VIC area. The persistent drought in the 2000s
(Howden 2012,Dijk and Viney (2013)) delayed the regrowth of trees. On
the other hand, replacing tree cover with pasture and crops in the Qld
area might have a relatively subtle impact on the EVI.

\subsection{Misalignment rainfall effects and vegetation
clearing}\label{misalignment-rainfall-effects-and-vegetation-clearing}

Particularly in the NSW/VIC area, there is no direct overlap between the
pixels with significant vegetation change (Figure \ref{fig:tctrend}) and
the pixels with significant land cover variables or positive Z'scores
(Figure \ref{fig:LCp} and Figure \ref{fig:steptest30}). While this seems
possibly surprising at first, a plausible explanation could be the
general movement of moisture and climate systems over the landscape,
resulting in a possible shift of where the moisture is evaporated and
where rainfall occurs. Specifically for this reason, this study selected
rather large spatial boxes to capture the overall response rather than
the pixel by pixel comparison/

\subsection{Gridded monthly rainfall
data}\label{gridded-monthly-rainfall-data}

The rainfall data used in this study is a gridded data set. This data
set is robust and consistent over a long time series (from 1900 to
current) and has a broad national wide coverage which can provide more
information spatially. However, high cross correlation between pixels,
due to the interpolation method generating this data set, can also
introduce spatial noise. In the step trend test the cross correlation
was accounted for. Some other methods are also available which can be
used to perform a comparative trial. For example, (Narisma et al. 2007)
applied a spatial Gaussian filter on a similar data set and used wavelet
analysis to detect step change in rainfall. High quality station data is
another option to test whether the observed spatial pattern in the step
trend test results was not due to the gridded data itself. Resampling
methods, such as bootstrapping and permutation (Wilks 1997,Kundzewicz
and Robson (2004),Westra, Alexander, and Zwiers (2013)), can also be
used to further assess the strength of significance of results and
incorporate spatial and temporal patterns in the analysis. While the
gridded data set is most useful in regions with sparse rain gauge
networks, it can actually reduces information where the rain gauge
density is high (Jones, Wang, and Fawcett 2009). In the NSW/VIC area,
the coverage of rainfall stations is more intensive but they are mainly
located in the valleys. The interpolated data might not be the best
represention of the local rainfall.

\subsection{General approach}\label{general-approach}

Parametric tests are generally more powerful than nonparametric test in
detecting a trend, when the data is normally distributed (Onoz and
Bayazit 2003,Kundzewicz and Robson (2004)). As a non-parametric test,
the step trend test has the advantages of distribution free and having
no restriction on missing data (Hirsch and Gilroy 1985). This is
particularly useful in rainfall analysis since rainfall data is usually
skewed. On the other hand, the disadvantages of non-parametric tests,
such as being limited to hypothesis testing and weaker in power, also
hold for the step trend test (Whitley and Ball 2002).

Overall, the current study provides a clear approach build on several
lines of evidence to provide some evidence to reject the null hypothesis
(no step change in rainfall occurs as a result of tree cover loss).
Limited by the available data, the time frame under study was chosen to
include a long lasting drought period (Holper 2011, Dijk and Viney
(2013)). The strong impact of this prolonged drought might have
suppressed the land-atmosphere interaction and confused the cause and
effect relation between rainfall and vegetation cover change. This could
be one of the reasons that the LCC effects on the local climate found in
other studies (e.g. Görgen et al. 2006,McAlpine et al. (2007)) are not
found to be significant here. Possible future work could focus on a
non-drought period, or when a longer series of land cover data is
available. Wile the power of the test can be improved with the longer
length of the after-intervention period (Hirsch and Gilroy 1985), the
dynamic nature of vegetation regrowth in this case study prevents this
effect. A better approach might be to build a global study that
investigates multiple locations where drastic landcover changes have
taken place, which would also remove some of the climate variability
effects due to the larger sample size.

\section{Conclusions}\label{conclusions}

In this study, we present an approach to identify iffound some
observationl data based evidence, although not strong, that vegetation
cover change has changed local rainfall. The semi-parametric method and
non-parametric method did totally agree on detecting a significant step
change in rainfall in the hot dry QLD region where land clearing has
occurred. On the other hand, the bushfires in the humid, temperate
mountain range in the NSW/VIC region has experienced reduced rainfall.
But the dry spell also plays an important role in the results.

Drought has had a pronounced impact on the land surface condition during
the study period, leading to significant reduction in vegetation and
extreme events such as bushfires. The associated lack of rainfall and
high temperatures may mask the step change in the vegetation. Hence, the
signal of LCC feedback on rainfall is probably weaker under such
regional dry conditions, as the impact of LCC on rainfall is mainly
through changes in moisture convergence (Görgen et al. 2006,Pitman and
Hesse (2007)).

\section{acknowledgments}\label{acknowledgments}

CL was supported by an Australian Postgraduate Award for this work.

\newpage

\section{appendix}\label{appendix}

\subsection{Summary of Data}\label{summary-of-data}

\newpage

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\hypertarget{ref-ABARES2010}{}
ABARES. 2010. ``Land Use of Australia, Version 4, 2005/2006.''

\hypertarget{ref-ABC2003}{}
ABC News. 2003. ``Kosciuszko Slow to Recover from Bushfires.''

\hypertarget{ref-ABS2012}{}
Australian Bureau of Statistics. 2012. ``Geography and Climate -
Australia's Climate.''

\hypertarget{ref-Ben-Gai1998}{}
Ben-Gai, T., A. Bitan, A. Manes, P. Alpert, and S. Rubin. 1998.
``Spatial and Temporal Changes in Rainfall Frequency Distribution
Patterns in Israel.'' \emph{Theoretical and Applied Climatology} 61
(3-4): 177--90.
\href{\%3CGo\%20to\%20ISI\%3E://000078025000005}{\textless{}Go to ISI\textgreater{}://000078025000005}.

\hypertarget{ref-BoM2009}{}
BoM. 2009. ``Gridded Monthly Rainfall Metadata.'' Australian Bureau of
Meteorology.
\url{http://www.bom.gov.au/climate/austmaps/metadata-monthly-rainfall.shtml}.

\hypertarget{ref-BoM2012a}{}
---------. 2012a. ``Australia - Climate of Our Continent.''

\hypertarget{ref-BoM2012}{}
---------. 2012b. ``Australian Climate Influence.''

\hypertarget{ref-Bosilovich2006}{}
Bosilovich, Michael G., and Jiun-Dar Chern. 2006. ``Simulation of Water
Sources and Precipitation Recycling for the MacKenzie, Mississippi, and
Amazon River Basins.'' \emph{Journal of Hydrometeorology} 7 (3):
312--29. \url{http://dx.doi.org/10.1175\%2FJHM501.1}.

\hypertarget{ref-Burton2010}{}
Burton, A, HJ Fowler, S Blenkinsop, and CG Kilsby. 2010. ``Downscaling
Transient Climate Change Using a Neyman-Scott Rectangular Pulses
Stochastic Rainfall Model.'' \emph{Journal of Hydrology} 381 (1):
18--32.

\hypertarget{ref-Chagnon2005}{}
Chagnon, F. J. F., and R. L. Bras. 2005. ``Contemporary Climate Change
in the Amazon.'' \emph{Geophysical Research Letters} 32 (13).
\href{\%3CGo\%20to\%20ISI\%3E://WOS:000230640100004\%20http://www.agu.org/journals/gl/gl0513/2005GL022722/2005GL022722.pdf}{\textless{}Go to ISI\textgreater{}://WOS:000230640100004 http://www.agu.org/journals/gl/gl0513/2005GL022722/2005GL022722.pdf}.

\hypertarget{ref-Chowdhury2010}{}
Chowdhury, R. K., and S. Beecham. 2010. ``Australian Rainfall Trends and
Their Relation to the Southern Oscillation Index.'' \emph{Hydrological
Processes} 24 (4). John Wiley \& Sons, Ltd.: 504--14.
\url{http://dx.doi.org/10.1002/hyp.7504}.

\hypertarget{ref-Cowpertwait2009}{}
Cowpertwait, PSP, Jim Salinger, and B Mullen. 2009. ``A Spatial-Temporal
Stochastic Rainfall Model for Auckland City: Scenarios for Current and
Future Climates.'' \emph{Journal of Hydrology (NZ)} 48: 95--109.

\hypertarget{ref-Deo2009}{}
Deo, RC, JI Syktus, CA McAlpine, and KK Wong. 2009. ``The Simulated
Impact of Land Cover Change on Climate Extremes in Eastern Australia.''
In \emph{Proceedings of the 18th World Imacs Congress and Modsim09
International Congress on Modelling and Simulation}, 2035--41.
Modelling; Simulation Society of Australia; New Zealand Inc.;
International Association for Mathematics; Computers in Simulation.

\hypertarget{ref-SLATS2001}{}
Department of Natural Resources and Mines. 2005. ``Land Cover Change in
Queensland 2001-2003, Incorporating 2001-2002 Adn 2002-2003 Change
Periods: A Statewide Landcover and Trees Study (SLATS) Report.''
Brisbane: Department of Natural Resources; Mines.

\hypertarget{ref-SLATS2004}{}
Department of Natural Resources and Water. 2007. ``Land Cover Change in
Queensland 2004-2005: A Statewide Landcover and Trees Study (SLATS)
Report.'' Brisbane: Department of Natural Resources; Water.

\hypertarget{ref-SLATS2017}{}
Department of Science, Information Technology and Innovation. 2017.
``Land Cover Change in Queensland 2015-2016: A Statewide Landcover and
Trees Study (SLATS) Report.'' Brisbane: Department of Science,
Information Technology; Innovation.

\hypertarget{ref-vanDijk2013}{}
Dijk, Beck van, Albert I. J. M., and Neil R. Viney. 2013. ``The
Millennium Drought in Southeast Australia (2001--2009): Natural and
Human Causes and Implications for Water Resources, Ecosystems, Economy,
and Society.'' \emph{Water Resources Research} 49 (2): 1040--57.
doi:\href{https://doi.org/10.1002/wrcr.20123}{10.1002/wrcr.20123}.

\hypertarget{ref-Dirmeyer2009}{}
Dirmeyer, P. A., K. L. Brubaker, and T. DelSole. 2009. ``Import and
Export of Atmospheric Water Vapor Between Nations.'' \emph{Journal of
Hydrology} 365 (1-2): 11--22.
\href{\%3CGo\%20to\%20ISI\%3E://000263575600003}{\textless{}Go to ISI\textgreater{}://000263575600003}.

\hypertarget{ref-Durieux2003}{}
Durieux, Laurent, Luiz Augusto Toledo Machado, and Henri Laurent. 2003.
``The Impact of Deforestation on Cloud Cover over the Amazon Arc of
Deforestation.'' \emph{Remote Sensing of Environment} 86 (1): 132--40.
\url{http://www.sciencedirect.com/science/article/pii/S0034425703000956}.

\hypertarget{ref-Eltahir1996}{}
Eltahir, E. A. B., and R. L. Bras. 1996. ``Precipitation Recycling.''
\emph{Reviews of Geophysics} 34 (3): 367--78.
\href{\%3CGo\%20to\%20ISI\%3E://A1996VD58700003}{\textless{}Go to ISI\textgreater{}://A1996VD58700003}.

\hypertarget{ref-Fowler2005}{}
Fowler, HJ, CG Kilsby, PE O'connell, and A Burton. 2005. ``A
Weather-Type Conditioned Multi-Site Stochastic Rainfall Model for the
Generation of Scenarios of Climatic Variability and Change.''
\emph{Journal of Hydrology} 308 (1): 50--66.

\hypertarget{ref-Gaertner2001}{}
Gaertner, M. A., O. B. Christensen, J. A. Prego, J. Polcher, C.
Gallardo, and M. Castro. 2001. ``The Impact of Deforestation on the
Hydrological Cycle in the Western Mediterranean: An Ensemble Study with
Two Regional Climate Models.'' \emph{Climate Dynamics} 17 (11): 857--73.

\hypertarget{ref-Gallant2007}{}
Gallant, Ailie J. E., Kevin J. Hennessy, and James Risbey. 2007.
``Trends in Rainfall Indices for Six Australian Regions: 1910 - 2005.''
\emph{Australian Meteorological Magazine} 56: 223--39.

\hypertarget{ref-Gimeno2010}{}
Gimeno, Luis, Anita Drumond, Raquel Nieto, Ricardo M. Trigo, and Andreas
Stohl. 2010. ``On the Origin of Continental Precipitation.''
\emph{Geophysical Research Letters} 37 (13). AGU.
\url{http://dx.doi.org/10.1029/2010GL043712}.

\hypertarget{ref-Gowen2016}{}
Gowen, Rebecca, and Steven G. Bray. 2016. ``Bioeconomic Modelling of
Woody Regrowth Carbon Offset Options in Productive Grazing Systems.''
\emph{The Rangeland Journal} 38 (3): 307--17.
doi:\href{https://doi.org/https://doi.org/10.1071/RJ15084}{https://doi.org/10.1071/RJ15084}.

\hypertarget{ref-Gorgen2006}{}
Görgen, K., A. H. Lynch, A. G. Marshall, and J. Beringer. 2006. ``Impact
of Abrupt Land Cover Changes by Savanna Fire on Northern Australian
Climate.'' \emph{Journal of Geophysical Research-Atmospheres} 111 (D19):
D19106. \url{http://dx.doi.org/10.1029/2005JD006860}.

\hypertarget{ref-Hastie1986}{}
Hastie, Trevor, and Robert Tibshirani. 1986. ``Generalized Additive
Models.'' \emph{Statistical Science}, 297--310.

\hypertarget{ref-Hipel1994}{}
Hipel, K.W., and A.I. McLeod. 1994. \emph{Time Series Modelling of Water
Resources and Environmental Systems}. Vol. 45. Elsevier Science Ltd.

\hypertarget{ref-Hirsch1985}{}
Hirsch, Robert M., and Edward J. Gilroy. 1985. ``Detectability of Step
Trends in the Rate of Atmospheric Deposition of Sulfate.'' \emph{JAWRA
Journal of the American Water Resources Association} 21 (5): 773--84.
\url{http://dx.doi.org/10.1111/j.1752-1688.1985.tb00171.x}.

\hypertarget{ref-Holper2011}{}
Holper, Paul N. 2011. \emph{Climate Change, Science Information Paper:
Australian Rainfall-Past, Present and Future}. Canberra: CSIRO.

\hypertarget{ref-Howden2012}{}
Howden, Saffron. 2012. ``It's Official: Australia No Longer in
Drought.'' \emph{Brisbane Times}.

\hypertarget{ref-Hughes2003}{}
Hughes, Lesley. 2003. ``Climate Change and Australia: Trends,
Projections and Impacts.'' \emph{Austral Ecology} 28 (4). Blackwell
Science Pty: 423--43.
\url{http://dx.doi.org/10.1046/j.1442-9993.2003.01300.x}.

\hypertarget{ref-Jones2009}{}
Jones, D.A., W. Wang, and R. Fawcett. 2009. ``High-Quality Spatial
Climate Data-Sets for Australia.'' \emph{Australian Meteorological and
Oceanographic Journal} 58 (4): 233--48.

\hypertarget{ref-Junkermann2009}{}
Junkermann, W., J. Hacker, T. Lyons, and U. Nair. 2009. ``Land Use
Change Suppresses Precipitation.'' \emph{Atmospheric Chemistry and
Physics} 9 (17): 6531--9.
\href{\%3CGo\%20to\%20ISI\%3E://WOS:000269778500018}{\textless{}Go to ISI\textgreater{}://WOS:000269778500018}.

\hypertarget{ref-Kamruzzaman2011}{}
Kamruzzaman, M., S. Beecham, and A. V. Metcalfe. 2011.
``Non-Stationarity in Rainfall and Temperature in the Murray Darling
Basin.'' \emph{Hydrological Processes} 25 (10). John Wiley \& Sons,
Ltd.: 1659--75. \url{http://dx.doi.org/10.1002/hyp.7928}.

\hypertarget{ref-Koutsoyiannis2007}{}
Koutsoyiannis, Demetris. 2006. ``Nonstationarity Versus Scaling in
Hydrology.'' Journal Article. \emph{Journal of Hydrology} 324 (1-4):
239--54.
\href{http://www.sciencedirect.com/science/article/B6V6C-4HP6GJ9-1/2/67dcc2e8f8e75b7abb918617f5a79a37\%20}{http://www.sciencedirect.com/science/article/B6V6C-4HP6GJ9-1/2/67dcc2e8f8e75b7abb918617f5a79a37}.

\hypertarget{ref-kucharski_further_2013}{}
Kucharski, Fred, Ning Zeng, and Eugenia Kalnay. 2013. ``A Further
Assessment of Vegetation Feedback on Decadal Sahel Rainfall
Variability.'' \emph{Climate Dynamics} 40 (5-6): 1453--66.
\url{http://link.springer.com/article/10.1007/s00382-012-1397-x}.

\hypertarget{ref-Kuzcera1987}{}
Kuczera, George. 1987. ``Prediction of Water Yield Reductions Following
a Bushfire in Ash-Mixed Species Eucalypt Forest.'' \emph{Journal of
Hydrology} 94 (3-4): 215--36.
doi:\href{https://doi.org/https://10.1016/0022-1694(87)90054-0}{https://10.1016/0022-1694(87)90054-0}.

\hypertarget{ref-Kundzewicz2004}{}
Kundzewicz, Z.W., and A.J. Robson. 2004. ``Change Detection in
Hydrological Records-a Review of the Methodology/Revue Méthodologique de
La Détection de Changements Dans Les Chroniques Hydrologiques.''
\emph{Hydrological Sciences Journal} 49 (1).

\hypertarget{ref-Lymburner2010}{}
Lymburner, Leo, Peter Tan, Norman Mueller, Richard Thackway, Adam Lewis,
Medhavy Thankappan, Lucy Randall, Anisul Islam, and Udaya Senarath.
2010. ``250 Metre Dynamic Land Cover Dataset.'' Geoscience Australia,
Canberra.

\hypertarget{ref-Ma2011}{}
Ma, H. Y., C. R. Mechoso, Y. Xue, H. Xiao, C. M. Wu, J. L. Li, and F. De
Sales. 2011. ``Impact of Land Surface Processes on the South American
Warm Season Climate.'' \emph{Climate Dynamics} 37 (1-2): 187--203.
\href{\%3CGo\%20to\%20ISI\%3E://WOS:000293403000012}{\textless{}Go to ISI\textgreater{}://WOS:000293403000012}.

\hypertarget{ref-MacDonald2005}{}
MacDonald, Glen M., and Roslyn A. Case. 2005. ``Variations in the
Pacific Decadal Oscillation over the Past Millennium.''
\emph{Geophysical Research Letters} 32 (8): L08703.
\url{http://dx.doi.org/10.1029/2005GL022478}.

\hypertarget{ref-McAlpine2007}{}
McAlpine, C. A., J. Syktus, R. C. Deo, P. J. Lawrence, H. A. McGowan, I.
G. Watterson, and S. R. Phinn. 2007. ``Modeling the Impact of Historical
Land Cover Change on Australia's Regional Climate.'' \emph{Geophysical
Research Letters} 34 (22).
\href{\%3CGo\%20to\%20ISI\%3E://000251345100005}{\textless{}Go to ISI\textgreater{}://000251345100005}.

\hypertarget{ref-Mei2010}{}
Mei, R., and G. L. Wang. 2010. ``Rain Follows Logging in the Amazon?
Results from CAM3-CLM3.'' \emph{Climate Dynamics} 34 (7-8): 983--96.
\href{\%3CGo\%20to\%20ISI\%3E://WOS:000278088400005}{\textless{}Go to ISI\textgreater{}://WOS:000278088400005}.

\hypertarget{ref-Meneghini2007}{}
Meneghini, Belinda, Ian Simmonds, and Ian N. Smith. 2007. ``Association
Between Australian Rainfall and the Southern Annular Mode.''
\emph{International Journal of Climatology} 27 (1): 109--21.
\url{http://dx.doi.org/10.1002/joc.1370}.

\hypertarget{ref-Murphy2008}{}
Murphy, Bradley F., and Bertrand Timbal. 2008. ``A Review of Recent
Climate Variability and Climate Change in Southeastern Australia.''
\emph{International Journal of Climatology} 28 (7): 859--79.
\url{http://dx.doi.org/10.1002/joc.1627}.

\hypertarget{ref-Nair2011}{}
Nair, Udaysankar S., Y. Wu, J. Kala, T. J. Lyons, Sr. Pielke R. A., and
J. M. Hacker. 2011. ``The Role of Land Use Change on the Development and
Evolution of the West Coast Trough, Convective Clouds, and Precipitation
in Southwest Australia.'' \emph{Journal of Geophysical
Research-Atmospheres} 116 (D7). AGU: D07103.
\url{http://dx.doi.org/10.1029/2010JD014950}.

\hypertarget{ref-Narisma2007}{}
Narisma, G.T., J.A. Foley, R. Licker, and N. Ramankutty. 2007. ``Abrupt
Changes in Rainfall During the Twentieth Century.'' \emph{Geophysical
Research Letters} 34 (6): L06710.

\hypertarget{ref-Negri2004}{}
Negri, A. J., R. F. Adler, L. M. Xu, and J. Surratt. 2004. ``The Impact
of Amazonian Deforestation on Dry Season Rainfall.'' \emph{Journal of
Climate} 17 (6): 1306--19.
\href{\%3CGo\%20to\%20ISI\%3E://WOS:000220450100013\%20http://journals.ametsoc.org/doi/pdf/10.1175/1520-0442\%282004\%29017\%3C1306\%3ATIOADO\%3E2.0.CO\%3B2}{\textless{}Go to ISI\textgreater{}://WOS:000220450100013 http://journals.ametsoc.org/doi/pdf/10.1175/1520-0442\%282004\%29017\%3C1306\%3ATIOADO\%3E2.0.CO\%3B2}.

\hypertarget{ref-Nicholls2006}{}
Nicholls, N. 2006. ``Detecting and Attributing Australian Climate
Change: A Review.'' \emph{Australian Meteorological Magazine} 55 (3).
Bureau of Meteorology.: 199--211.

\hypertarget{ref-Oleson2004}{}
Oleson, K. W., G. B. Bonan, S. Levis, and M. Vertenstein. 2004.
``Effects of Land Use Change on North American Climate: Impact of
Surface Datasets and Model Biogeophysics.'' \emph{Climate Dynamics} 23
(2): 117--32.

\hypertarget{ref-Onoz2003}{}
Onoz, B., and M. Bayazit. 2003. ``The Power of Statistical Tests for
Trend Detection.'' \emph{Turkish Journal of Engineering and
Environmental Sciences} 27 (4): 247--51.

\hypertarget{ref-Otterman1990}{}
Otterman, J., A. Manes, S. Rubin, P. Alpert, and D. Starr. 1990. ``An
Increase of Early Rains in Southern Israel Following Land-Use Change?''
\emph{Boundary-Layer Meteorology} 53 (4): 333--51.

\hypertarget{ref-Pinto2009}{}
Pinto, E., Y. Shin, S. A. Cowling, and C. D. Jones. 2009. ``Past,
Present and Future Vegetation-Cloud Feedbacks in the Amazon Basin.''
\emph{Climate Dynamics} 32 (6): 741--51.
\href{\%3CGo\%20to\%20ISI\%3E://WOS:000264118400001}{\textless{}Go to ISI\textgreater{}://WOS:000264118400001}.

\hypertarget{ref-Pitman2007}{}
Pitman, A. J., and P. P. Hesse. 2007. ``The Significance of Large-Scale
Land Cover Change on the Australian Palaeomonsoon.'' \emph{Quaternary
Science Reviews} 26 (1-2): 189--200.
\url{http://www.sciencedirect.com/science/article/B6VBC-4MC71SN-1/2/0e1d4dff6e15a1de79cefbfa612f9b37}.

\hypertarget{ref-pitman_scale_2016}{}
Pitman, A. J., and R. Lorenz. 2016. ``Scale Dependence of the Simulated
Impact of Amazonian Deforestation on Regional Climate.''
\emph{Environmental Research Letters} 11 (9): 094025.
doi:\href{https://doi.org/10.1088/1748-9326/11/9/094025}{10.1088/1748-9326/11/9/094025}.

\hypertarget{ref-Pitman2004}{}
Pitman, A. J., G. T. Narisma, R. A. Pielke, and N. J. Holbrook. 2004.
``Impact of Land Cover Change on the Climate of Southwest Western
Australia.'' \emph{Journal of Geophysical Research-Atmospheres} 109
(D18).
\href{\%3CGo\%20to\%20ISI\%3E://WOS:000224126500001}{\textless{}Go to ISI\textgreater{}://WOS:000224126500001}.

\hypertarget{ref-Rstats2018}{}
R Core Team, 2018. \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria: R Foundation for Statistical Computing.
\url{http://www.R-project.org/}.

\hypertarget{ref-Risbey2009}{}
Risbey, James S., Michael J. Pook, Peter C. McIntosh, Matthew C.
Wheeler, and Harry H. Hendon. 2009. ``On the Remote Drivers of Rainfall
Variability in Australia.'' \emph{Monthly Weather Review} 137 (10):
3233--53. \url{http://dx.doi.org/10.1175/2009MWR2861.1}.

\hypertarget{ref-Roy2008}{}
Roy, David P, Luigi Boschetti, Christopher O Justice, and J Ju. 2008.
``The Collection 5 Modis Burned Area Product---Global Evaluation by
Comparison with the Modis Active Fire Product.'' \emph{Remote Sensing of
Environment} 112 (9). Elsevier: 3690--3707.

\hypertarget{ref-Roy2005}{}
Roy, DP, Y Jin, PE Lewis, and CO Justice. 2005. ``Prototyping a Global
Algorithm for Systematic Fire-Affected Area Mapping Using Modis Time
Series Data.'' \emph{Remote Sensing of Environment} 97 (2). Elsevier:
137--62.

\hypertarget{ref-Roy2002}{}
Roy, DP, PE Lewis, and CO Justice. 2002. ``Burned Area Mapping Using
Multi-Temporal Moderate Spatial Resolution Data---A Bi-Directional
Reflectance Model-Based Expectation Approach.'' \emph{Remote Sensing of
Environment} 83 (1). Elsevier: 263--86.

\hypertarget{ref-saha_investigating_2016}{}
Saha, Subodh K., Paul A. Dirmeyer, and Thomas N. Chase. 2016.
``Investigating the Impact of Land-Use Land-Cover Change on Indian
Summer Monsoon Daily Rainfall and Temperature During 1951-2005 Using a
Regional Climate Model.'' \emph{Hydrology and Earth System Sciences} 20
(5): 1765.
\url{http://search.proquest.com/openview/c4242c9ecce075e3545390b1d00f2751/1?pq-origsite=gscholar\&cbl=105724}.

\hypertarget{ref-Saji1999}{}
Saji, NH, B.N. Goswami, PN Vinayachandran, and T. Yamagata. 1999. ``A
Dipole Mode in the Tropical Indian Ocean.'' \emph{Nature} 401 (6751):
360--63.

\hypertarget{ref-Sato2007}{}
Sato, T., F. Kimura, and A. S. Hasegawa. 2007. ``Vegetation and
Topographic Control of Cloud Activity over Arid/Semiarid Asia.''
\emph{Journal of Geophysical Research-Atmospheres} 112 (D24).
\href{\%3CGo\%20to\%20ISI\%3E://WOS:000252013700002\%20http://www.agu.org/journals/jd/jd0724/2006JD008129/2006JD008129.pdf}{\textless{}Go to ISI\textgreater{}://WOS:000252013700002 http://www.agu.org/journals/jd/jd0724/2006JD008129/2006JD008129.pdf}.

\hypertarget{ref-Schepen2012}{}
Schepen, Andrew, Q. J. Wang, and David Robertson. 2012. ``Evidence for
Using Lagged Climate Indices to Forecast Australian Seasonal Rainfall.''
\emph{Journal of Climate} 25 (4). American Meteorological Society:
1230--46.

\hypertarget{ref-Semazzi2001}{}
Semazzi, F. H. M., and Y. Song. 2001. ``A GCM Study of Climate Change
Induced by Deforestation in Africa.'' \emph{Climate Research} 17 (2):
169--82.

\hypertarget{ref-Smith2012}{}
Smith, Ian N., and Bertrand Timbal. 2012. ``Links Between Tropical
Indices and Southern Australian Rainfall.'' \emph{International Journal
of Climatology} 32 (1): 33--40.

\hypertarget{ref-Speer2011}{}
Speer, Milton, Lance Leslie, and Alexandre Fierro. 2011. ``Australian
East Coast Rainfall Decline Related to Large Scale Climate Drivers.''
\emph{Climate Dynamics} 36 (7): 1419--29.
\url{http://dx.doi.org/10.1007/s00382-009-0726-1}.

\hypertarget{ref-Fire2011}{}
The State Government of Victoria. 2011. ``Bushfire History.''

\hypertarget{ref-Timbal2006}{}
Timbal, B., and J.M. Arblaster. 2006. ``Land Cover Change as an
Additional Forcing to Explain the Rainfall Decline in the South West of
Australia.'' \emph{Geophysical Research Letters} 33 (7): L07717.

\hypertarget{ref-Townshend2011}{}
Townshend, J.R.G., M. Carroll, C. Dimiceli, R. Sohlberg, M. Hansen, and
R. DeFries. 2011. ``Vegetation Continuous Fields MOD44B, 2001 Percent
Tree Cover, Collection 5.'' University of Maryland, College Park,
Maryland, 2001.

\hypertarget{ref-Trenberth1999}{}
Trenberth, K. E. 1999. ``Atmospheric Moisture Recycling: Role of
Advection and Local Evaporation.'' \emph{Journal of Climate} 12 (5):
1368--81.
\href{\%3CGo\%20to\%20ISI\%3E://000080145700002}{\textless{}Go to ISI\textgreater{}://000080145700002}.

\hypertarget{ref-Verdon2004}{}
Verdon, Danielle C., Adam M. Wyatt, Anthony S. Kiem, and Stewart W.
Franks. 2004. ``Multidecadal Variability of Rainfall and Streamflow:
Eastern Australia.'' \emph{Water Resources Research} 40 (10): W10201.
\url{http://dx.doi.org/10.1029/2004WR003234}.

\hypertarget{ref-Wang2009}{}
Wang, Jingfeng, Frederic Chagnon, Earle Williams, Alan Betts, Nilton
Renno, Luiz Machado, Gautam Bisht, Ryan Knox, and Rafael Bras. 2009.
``Impact of Deforestation in the Amazon Basin on Cloud Climatology.''
\emph{Proceedings of the National Academy of Sciences of the United
States of America} 106 (10): 3670.
\url{http://ezproxy.library.usyd.edu.au/login?url=http://search.proquest.com/docview/201417978?accountid=14757}.

\hypertarget{ref-Westra2013}{}
Westra, Seth, Lisa V. Alexander, and Francis W. Zwiers. 2013. ``Global
Increasing Trends in Annual Maximum Daily Precipitation.'' \emph{Journal
of Climate} 26 (11). American Meteorological Society: 3904--18.
\url{http://dx.doi.org/10.1175/JCLI-D-12-00502.1}.

\hypertarget{ref-Westra2010}{}
Westra, Seth, and Ashish Sharma. 2010. ``An Upper Limit to Seasonal
Rainfall Predictability?'' \emph{Journal of Climate} 23 (12). American
Meteorological Society: 3332--51.
\href{http://ezproxy.library.usyd.edu.au/login?url=http://search.proquest.com/docview/814127501?accountid=14757\%20http://DD8GH5YX7K.search.serialssolutions.com/?SS_Source=3\&genre=article\&sid=ProQ:\&atitle=An+Upper+Limit+to+Seasonal+Rainfall+Predictability\%3F\&title=Journal+of+Climate\&issn=0894-8755\&date=2010-06-01\&volume=23\&issue=12\&spage=3332\&SS_docid=814127501\&author=Westra\%2C+Seth\%3BSharma\%2C+Ashish}{http://ezproxy.library.usyd.edu.au/login?url=http://search.proquest.com/docview/814127501?accountid=14757 http://DD8GH5YX7K.search.serialssolutions.com/?SS\_Source=3\&genre=article\&sid=ProQ:\&atitle=An+Upper+Limit+to+Seasonal+Rainfall+Predictability\%3F\&title=Journal+of+Climate\&issn=0894-8755\&date=2010-06-01\&volume=23\&issue=12\&spage=3332\&SS\_docid=814127501\&author=Westra\%2C+Seth\%3BSharma\%2C+Ashish}.

\hypertarget{ref-Whitley2002}{}
Whitley, Elise, and Jonathan Ball. 2002. ``Statistics Review 6:
Nonparametric Methods.'' \emph{Critical Care, London} 6 (6): 509--13.

\hypertarget{ref-Wilks2006}{}
Wilks, D. S. 2006. ``On `Field Significance' and the False Discovery
Rate.'' \emph{Journal of Applied Meteorology and Climatology} 45 (9).
American Meteorological Society: 1181--9.
\href{http://ezproxy.library.usyd.edu.au/login?url=http://search.proquest.com/docview/224582143?accountid=14757\%20http://DD8GH5YX7K.search.serialssolutions.com/?ctx_ver=Z39.88-2004\&ctx_enc=info:ofi/enc:UTF-8\&rfr_id=info:sid/ProQ\%3Amilitary\&rft_val_fmt=info:ofi/fmt:kev:mtx:journal\&rft.genre=article\&rft.jtitle=Journal+of+Applied+Meteorology+and+Climatology\&rft.atitle=On+\%22Field+Significance\%22+and+the+False+Discovery+Rate\&rft.au=Wilks\%2C+D+S\&rft.aulast=Wilks\&rft.aufirst=D\&rft.date=2006-09-01\&rft.volume=45\&rft.issue=9\&rft.spage=1181\&rft.isbn=\&rft.btitle=\&rft.title=Journal+of+Applied+Meteorology+and+Climatology\&rft.issn=15588424}{http://ezproxy.library.usyd.edu.au/login?url=http://search.proquest.com/docview/224582143?accountid=14757 http://DD8GH5YX7K.search.serialssolutions.com/?ctx\_ver=Z39.88-2004\&ctx\_enc=info:ofi/enc:UTF-8\&rfr\_id=info:sid/ProQ\%3Amilitary\&rft\_val\_fmt=info:ofi/fmt:kev:mtx:journal\&rft.genre=article\&rft.jtitle=Journal+of+Applied+Meteorology+and+Climatology\&rft.atitle=On+\%22Field+Significance\%22+and+the+False+Discovery+Rate\&rft.au=Wilks\%2C+D+S\&rft.aulast=Wilks\&rft.aufirst=D\&rft.date=2006-09-01\&rft.volume=45\&rft.issue=9\&rft.spage=1181\&rft.isbn=\&rft.btitle=\&rft.title=Journal+of+Applied+Meteorology+and+Climatology\&rft.issn=15588424}.

\hypertarget{ref-Wilks1997}{}
Wilks, DS. 1997. ``Resampling Hypothesis Tests for Autocorrelated
Fields.'' \emph{Journal of Climate} 10 (1): 65--82.

\hypertarget{ref-Wood2011}{}
Wood, S. N. 2011. ``Fast Stable Restricted Maximum Likelihood and
Marginal Likelihood Estimation of Semiparametric Generalized Linear
Models.'' \emph{Journal of the Royal Statistical Society (B)} 73 (1):
3--36.

\hypertarget{ref-Zanchettin2008}{}
Zanchettin, Davide, Stewart W. Franks, Pietro Traverso, and Mario
Tomasino. 2008. ``On Enso Impacts on European Wintertime Rainfalls and
Their Modulation by the Nao and the Pacific Multi-Decadal Variability
Described Through the Pdo Index.'' \emph{International Journal of
Climatology} 28 (8). John Wiley \& Sons, Ltd.: 995--1006.
\url{http://dx.doi.org/10.1002/joc.1601}.

\hypertarget{ref-Zeng2012}{}
Zeng, X. M., Z. H. Wu, S. Song, S. Y. Xiong, Y. Q. Zheng, Z. G. Zhou,
and H. Q. Liu. 2012. ``Effects of Land Surface Schemes on the Simulation
of a Heavy Rainfall Event by WRF.'' \emph{Chinese Journal of
Geophysics-Chinese Edition} 55 (1): 16--28.
\href{\%3CGo\%20to\%20ISI\%3E://WOS:000300129300002}{\textless{}Go to ISI\textgreater{}://WOS:000300129300002}.

\end{document}

### 2018-09-04 Manuscripts/chap3_paper_WVversion.Rnw
changes for paper

--- 2017-11-08 chap3_paper_WVversion.Rnw
@@
   mod44_tc <- brick("Data\\Satellite\\MOD44\\B\\mod44_tc_aus.tif")
-  tm <- seq(as.Date("2000-03-06"), as.Date("2010-03-06"), "year")
+  tm <- seq(as.Date("2000-03-06"), as.Date("2016-12-06"), "year")
   mod44_tc <- setZ(mod44_tc, tm, "year")

### 2018-09-04 Manuscripts/chap3_paper_new.Rnw
changes for paper

--- 2017-11-08 chap3_paper_new.Rnw
@@
                        NA)))
-
+# p-value of C
 Cp30_qld4 <- unlist(lapply(1:length(coef.rain30.qld4),

### 2018-09-04 Manuscripts/chap3_paper_revision.Rnw
changes for paper

--- 2018-09-04 Manuscripts/chap3_paper_WVversion.Rnw
@@
 \begin{abstract}
-(text here)
+Analysis of observational data to identify relationships between rainfall and land cover change are scarce due to multiple environmental factors that cannot be controlled.
+In this study we investigated the relationship using statistical methods on data from best available sources at two sites in Australia.
+Gridded data of rainfall and tree cover were used as spatially corresponding local conditions.
+Large scale effects were represented by climatic indicators, such as SOI and IOD.
+Regression analysis and step trend tests were used to assess the effect of abrupt land surface intervention.
+At a Queensland site, dramatic tree cover change between 2002 - 2005 did not result in strong statistically significant precipitation changes.
+On the other hand, results from step trend test on 1998 - 2008 rainfall data in a bushfire affected NSW/VIC region has a better match with the tree cover change map.
+This indicates the method works better when a abrupt change in the data can be clearly identified.
+The results from the step trend test implied a positive relationship between the tree cover and the rainfall at 0.1 significance level.
 \end{abstract}
@@
 Land use and land cover changes can lead to changes in the local climate.
-Both empirical and modelling studies have found cloud types and rainfall are correlated to large scale vegetation cover changes, such as deforestation in the Amazon and in the Sahel \citep{Chagnon2005,Pinto2009, Wang2009, Mei2010, kucharski_further_2013,pitman_scale_2016} and afforestation in south Israel \citep{Otterman1990, Ben-Gai1998}.
+Empirical and modelling studies have found cloud types and rainfall are correlated to large scale vegetation cover changes, such as deforestation in the Amazon and in the Sahel \citep{Chagnon2005,Pinto2009, Wang2009, Mei2010, kucharski_further_2013,pitman_scale_2016} and afforestation in south Israel \citep{Otterman1990, Ben-Gai1998}.
 Using airborne measurement in Western Australia, \citet{Junkermann2009} showed a significantly higher level of aerosols over an agricultural area compared to an adjacent natural vegetation.
-They suggested that a modification of aerosol concentrations due to deforestation could have contributed to a reduction of local rainfall, as more but smaller rain droplets were observed.
-\citet{Nair2011} reported from the Bunny Fence Experiment in Western Australia that local land use change altered the synoptic west coast trough dynamics and surface roughness, and this resulted in the observed rainfall decrease.
-Climate sensitivity to land cover change is also found in eastern Australia \citep{McAlpine2007}.
-
-Rainfall over land is generally influenced by multiple factors.
-Locally, there are two main sources: moisture from advective atmospheric transport; and local evapotranspiration \citep{Eltahir1996,Bosilovich2006,Dirmeyer2009,Gimeno2010}.
+They suggested that a modification of aerosol concentrations due to deforestation could have contributed to a reduction of local rainfall, as more, but smaller rain droplets were observed.
+\citet{Nair2011} reported from the Bunny Fence Experiment in Western Australia that local land use change altered the synoptic west coast trough dynamics and surface roughness, and this resulted in an observed rainfall decrease.
+Maximum tempratures were also found to be sesnitive to land cover change in eastern Australia \citep{McAlpine2007}.
+
+Overall the number of empirical studies analyzing changes to rainfall due to land cover change from observational data is limited.
+Most of the studies mentioned previously were either model simulations, or comparisons of modelled data with observations.
+This is because there are some fundamental experimental difficulties in both space (where does evaporated water reappear as rainfall?) and in time (how much time does it take for land cover change effects to appear or disappear?).
+In addition, in many areas across the globe rainfall variability is related to complex set of interactions, of which landuse change might only be a minor component.
+
+Locally, there are two main sources that generate rainfall: moisture from advective atmospheric transport; and local evapotranspiration \citep{Eltahir1996,Bosilovich2006,Dirmeyer2009,Gimeno2010}.
+The local evapotranspiration component is the component considered to be affected by landuse change \citep{Eltahir1996}.
 According to \citet{Trenberth1999}, the contribution of advective moisture partially depends on the availability of external moisture and atmospheric transport.
 On the longer time scale, such as monthly and annually, large scale atmospheric dynamics are affected by large scale climate drivers.
-Many studies have reported significant relationships between rainfall in large parts of Australia and the El Ni\~{n}o-Southern Oscillation (ENSO) \citep{Verdon2004,Risbey2009,Speer2011}.
-ENSO can be used to represent longer term cycles in rainfall data such as drought.
-On the other hand, local ET is determined by local land surface characteristics.
-Local land surface characteristics further influence local scale atmospheric dynamics and hence the amount of rainfall, including contribution from both sources.
-Therefore land surface plays an important role in local rainfall.
-
-Conventionally, large scale climate drivers are used as predictors for seasonal rainfall forecasts.
+For example, many studies have reported significant relationships between rainfall in large parts of Australia and the El Ni\~{n}o-Southern Oscillation (ENSO) \citep{Verdon2004,Risbey2009,Speer2011}.
+In contrast, local ET is determined by local land surface characteristics, which influence local scale atmospheric dynamics and hence the amount of rainfall, including contribution from both main sources.
+Therefore land surface is suggested to play an important role in local rainfall.
+
+%%Conventionally, large scale climate drivers are used as predictors for seasonal rainfall forecasts.
 According to the Australian Bureau of Meteorology (BoM), probablistic seasonal outlooks are developed based on Australian rainfall and temperature as well as sea surface temperature records from the tropical Pacific and Indian Oceans \citep{BoM2012c}.
 %% comment Expand this section with section from methods (section 4.1)?
-Different parts of the Australian continent can be more or less influenced by different climatic drivers \citep{Mla2008}.
+%%Different parts of the Australian continent can be more or less influenced by different climatic drivers \citep{Mla2008}.
 Using both SOI and PDO in a prediction model, \citet{Kamruzzaman2011} reported that PDO was rarely significant for rainfall stations in the MDB and on the southeast coast of Australia, while SOI was at least significant at the 1\% level.
@@
 Although climate drivers demonstrate some capability to predict Australian rainfall, there is still a large amount of unexplained variance.
-\citet{Westra2010} pointed out that models based on global sea surface temperature anomalies can only predict up to 14.7\% of precipitation variance.
-Recent studies suggest that land surface processes are important for predicting local rainfall \citep[e.g.][]{Ma2011,Zeng2012, pitman_scale_2016, saha_investigating_2016}.
+\citet{Westra2010} pointed out that models based on global sea surface temperature anomalies can only predict up to 14.7\% of annual precipitation variance.
+Some of the remaining variance could be due to land surface processes as suggested in studies predicting local rainfall \citep[e.g.][]{Ma2011,Zeng2012, pitman_scale_2016, saha_investigating_2016}.
 However, they are mostly based on modelling experiments and little evidence was reported from observations.
-\citet{Pitman2004} found a good match between observations and simulated rainfall changes in southwest Western Australia ,forced by land cover change.
+However, \citet{Pitman2004} found a good match between observations and simulated rainfall changes in southwest Western Australia ,forced by land cover change.
 \citet{Timbal2006} were able to reproduce the rainfall decline in south west Australia by including land cover influence.
-Local land use change might not be a primary, but is likely to be a secondary cause of rainfall change \citep{Nicholls2006}.
-Therefore, land surface modification has, at least partially, contributed to local rainfall variability.
-
-Overall the number of empirical studies related to changes to rainfall due to land cover change is limited.
-This is because there are some fundamental experimental difficulties in both space (where does evaporated water reappear as rainfall?) and in time (how much time does it take for land cover change effects to appear or disappear?).
-Therefore, the aim of this study is to use empirical evidence at regional scales to investigate the cause and effect relationship between land cover change and local rainfall using empirical evidence.
+INaddititon, local land use change might not be a primary, but is likely to be a secondary cause of rainfall change \citep{Nicholls2006}.
+
+Therefore, the aim of this study is to expand the available methodology to use empirical evidence at regional scales to investigate the cause and effect relationship between land cover change and local rainfall.
 More specifically, we hypothesize that a step change on the land cover on the surface will cause a step change in the rainfall.
-To demonstrate the approach and this effect we studied step changes in rainfall at two locations in Queensland and NSW/Victoria where there are possible step changes in land cover change due to land clearing and bushfires based on tree cover data.
-Statistical approaches were used to identify the step changes in rainfall, which were subsequently associated with land cover change through spatial comparison.
-
-In this paper, after this section (the introduction), section 2 covers the case study areas and the observed land use changed.
+To demonstrate the approach and this effect we applied the methodlogy to study the changes in rainfall at two locations in Queensland and NSW/Victoria where there are possible step changes in land cover change due to land clearing and bushfires.
+The methodology is mainly based on statistical approaches to identify changes in rainfall, which were subsequently associated with land cover change through spatial comparison.
+
+In this paper, after this section (the introduction), section 2 covers the case study areas and the observed land use change.
 Section 3 describes the data used in the study in more detail.
@@
 
+We use two areas in Australia to demonstrate the approach.
 In Australia, significant tree cover change has mainly occurred in the north east of the continent and on the southeast coast, as well as in the southwest of Western Australia.
@@
 
-Two regions were selected where significant tree cover change was present.
+Two regions were selected where significant tree cover change was reported.
 The first region is located in south central Queensland to the north of the Murray Darling Basin (MDB) (site 1 in Figure~\ref{fig:sel_reg}).
 High rates of land clearing have been reported in this region during the early 2000s \citep{SLATS2004}.
-The second study region is located at the border of New South Wales and Victoria, and includes the Snowy Mountains ranges (site 2 in Figure~\ref{fig:sel_reg}).
+The second study region is located at the border of New South Wales and Victoria, and includes the Snowy Mountain ranges (site 2 in Figure~\ref{fig:sel_reg}).
 Severe bushfires occurred in this area and the surroundings in early 2003 (see Figure~\ref{fig:bushfire}).
@@
 Two thirds of Kosciuszko national park was heavily burned and regrowth was reported to be slow due to drought and cold conditions \citep{ABC2003} and the type of species in this region.
-In both study regions, significant tree cover loss has happened within the last decade, either permanently or temporarily.
-It makes the selected areas suitable for this land cover change study.
+However, in the longer term, after an early high transpiration period a recovery of pre-fire evapotranspiration would be expected \citep{Kuzcera1987}.
+As a result , significant tree cover loss has happened in both study areas in the last decade, either permanently or temporarily.
 %\footnote{Bushfire history, \url{http://www.dse.vic.gov.au/fire-and-other-emergencies/major-bushfires-in-victoria}, viewed on 12/12/2011}
@@
 
-<<mapping,eval=F, echo = F, results = hide>>=
-
+<<packages,eval=T, echo = F, results = hide>>=
+
+#library(chron)
+library(fields)
+library(graphics)
+library(maptools)
 library(plotKML)
 library(raster)
+library(reshape)
 library(rgeos)
-library(maptools)
-library(fields)
+library(rgdal)
 library(RColorBrewer)
+#library(zoo)
+library(tidyverse)
+require(lubridate)
+# library(classInt)     # finds class intervals for continuous variables
+library(deseasonalize)
+library(mgcv)
+library(ncdf4)
+library(nnet)
+@
+
+<<mapping,eval=T, echo = F, results = hide>>=
 
@@
    e2 <- extent(c(1346395.141,1534425.527,-1169402.246,-607644.171))
-
+  # 
    e11 <- extent(c(145.425, 149.025, -28.475, -25.175))
@@
 
-  e111 <- extent(c(145.825, 148.625, -28.075, -25.575)) #as(tc_r1,"SpatialPolygons")
-  e222 <- extent(c(147.075, 148.925, -36.675, -35.075)) #as(tc_r2,"SpatialPolygons")
+  e111 <- extent(c(145.825, 148.625, -28.075, -25.575)) 
+  e222 <- extent(c(147.075, 148.925, -36.675, -35.075))
 
 ## the selected regions for mapping
-
-  aus_reg <- readShapePoly("Data\\Geo\\aus_state2011\\simplify_state11_uproj.shp",proj4string=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0"))
+root_dir <- "c:/users/rver4657/owncloud/chun/chun/data"
+aus_reg <- readShapePoly(paste(root_dir,
+                               "Geo\\aus_state2011\\STE11aAust.shp",sep="/"),
+                         proj4string=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0"))
 aus_reg2 <- gSimplify(aus_reg,0.01,topologyPreserve=T)
 
-  mdb_reg <- readShapePoly("Data\\Geo\\MDB\\mdb.shp",proj4string=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0"))
+mdb_reg <- readShapePoly(paste(root_dir,
+                               "Geo\\MDB\\mdb_boundary.shp", sep="/"),
+                               proj4string=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0"))
 mdb_reg2 <- gSimplify(mdb_reg,0.01,topologyPreserve=TRUE)
 
-  remove(list=c("aus_reg","mdb_reg"))
+#remove(list=c("aus_reg","mdb_reg"))
 
@@
 
-fire_0301 <- raster("Data\\Other\\bushfires\\Burnt_Area\\tmpA2003001.burndate.tif")
+fire_0301 <- raster(paste(root_dir,
+                          "Other\\bushfires\\Burnt_Area\\tmpA2003001.burndate.tif",sep="/"))
 fire_0301nsw <- crop(fire_0301,e2)
@@
 
-  setEPS()
-  postscript("bushfire_nswvic.eps")
-#  jpg("bushfire_nswvic.jpg")
+  jpeg("../figures/bushfire_nswvic.jpg", width=960, height=960)
   par(mar=c(3,3,1,6),xpd=T)
@@
 
-## trend of evi (MODIS product) from GA and ABARES
-
-## tree cover data
-  mod44_tc <- brick("Data\\Satellite\\MOD44\\B\\mod44_tc_aus.tif")
-  tm <- seq(as.Date("2000-03-06"), as.Date("2016-12-06"), "year")
-  mod44_tc <- setZ(mod44_tc, tm, "year")
-
-## tree cover in sub-region
-  mod_tc_r11 <- crop(mod44_tc,e1)
-  mod_tc_r111 <- reproject(mod_tc_r11,CRS=CRS("+proj=longlat +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84"))
-  mod_tc_r1 <- crop(mod_tc_r111,e111)
-
-  mod_tc_r22 <- crop(mod44_tc,e2)
-  mod_tc_r222 <- reproject(mod_tc_r22,CRS=CRS("+proj=longlat +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84"))
- mod_tc_r2 <- crop(mod_tc_r222,e222)
-
- remove(list=c("mod44_tc","mod_tc_r11","mod_tc_r111","mod_tc_r22","mod_tc_r222"))
+rootdir <-"C:/users/rver4657/owncloud/chun2/data/satellite/MOD44/B"
+
+# read in the file list from the directory
+x <- dir(path=rootdir,pattern="Percent_Tree_Cover.tif")
+# Identify the different states in the images
+# NSW
+NSW <- x[grep("NSW",x)]
+# Qld
+Qld <- x[grep("Qld",x)]
+
+img <- list()
+# raster and brick
+for(i in 1:length(NSW)) {
+  img[[i]] <- raster(paste(rootdir,NSW[i],sep="/"))
+}
+
+imgQ <- list()
+# raster and brick
+for(i in 1:length(Qld)) {
+  imgQ[[i]] <- raster(paste(rootdir,Qld[i],sep="/"))
+}
+
+NSW_b <- do.call(brick,img)
+Qld_b <- do.call(brick, imgQ)
+
+tm <- seq(as.Date("2000-03-06"), as.Date("2016-03-06"), "year")
+mod_tc_r22 <- setZ(NSW_b, tm, "year")
+
+mod_tc_r11 <- setZ(Qld_b, tm, "year")
 
@@
 fun <- function(x,m) {
-    var_x <- sd(x)
-    mu_1 <- mean(x[1:3])
-    if (m==1) {y <- x[6:11]} else {y <- x[4:11]}   
-    mu_2 <- mean(y)
+  var_x <- sd(x,na.rm=T)
+  mu_1 <- mean(x[1:3],na.rm=T)
+  # m = 1 Queensland, 
+  if (m==1) {y <- x[4:16]} else {
+    # NSW whole series m = 2
+    if (m ==2) {y <- x[6:16]} else {
+      # build in other possibilities for length series
+      y <- x[6:(16 - m)]
+    }  
+  mu_2 <- mean(y,na.rm=T)
   z <- (mu_2-mu_1)/var_x
-    if (!is.na(z) & abs(z)>1.645) {return(y[1]-x[3])} else {return(NA)}
-  }
-
-  mod_tc_r1.z <- calc(mod_tc_r1,function(x) fun(x,1))
-  mod_tc_r2.z <- calc(mod_tc_r2,function(x) fun(x,2))
-
-  remove(list=c("mod_tc_r1","mod_tc_r2"))
-
-## create plot
-  plotclr <- brewer.pal(6,"PiYG")
-#"#C51B7D" "#E9A3C9" "#FDE0EF" "#E6F5D0" "#A1D76A" "#4D9221"
-
-  setEPS()
-  postscript("tc_qld.eps")
+  if (!is.na(z) & abs(z)>1.645) {return(mu_2-mu_1)} else {return(NA)}
+  }
+}
+
+# Qld
+mod_tc_r1.z <- calc(mod_tc_r11,function(x) fun(x,1))
+# NSW
+mod_tc_r2.z <- calc(mod_tc_r22,function(x) fun(x,2))
+# NSW only 5 years after
+mod_tc_r2_5y.z <- calc(mod_tc_r22,function(x) fun(x,7))
+# NSW 7 years after
+mod_tc_r2_7y.z <- calc(mod_tc_r22,function(x) fun(x,5))
+# NSW 9 years after
+mod_tc_r2_9y.z <- calc(mod_tc_r22,function(x) fun(x,3))
+
+require(RColorBrewer)
+plotclr <- brewer.pal(8,"PiYG")
+
+jpeg("../figures/tc_nswvic_5y.jpg", width=960, height=960)
+par(mar=c(3,3,1,7))
+plot(e222)
+image(mod_tc_r2_5y.z,col=plotclr,breaks=c(-70,-20,-10,-5,0,5,10,20,70),legend=F,add=T)
+#plot(e111,add=T)
+plot(aus_reg2,border="grey",lwd=3,add=T)
+image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,-0.2,0.2,1,2,3,4),
+           lab.breaks=c(-70,-20,-10,-5,"no tree","0 or",5,10,20,70),
+           legend.only=T,horizontal=F,
+           col=c("#C51B7D","#DE77AE","#F1B6DA","#FDE0EF","white",
+                 "#E6F5D0","#B8E186","#7FBC41","#4D9221"),
+           legend.shrink=0.7,
+           legend.width=0.7,
+           legend.args=list(text=expression(paste(Delta,
+                            "tree (% ground)",sep="")),
+                            line=0.5,cex=0.9,font=2),
+           axis.args=list(cex.axis=0.8),graphics.reset=F)
+dev.off()
+
+jpeg("../figures/tc_nswvic_7y.jpg", width=960, height=960)
+par(mar=c(3,3,1,7))
+plot(e222)
+image(mod_tc_r2_7y.z,col=plotclr,breaks=c(-70,-20,-10,-5,0,5,10,20,70),legend=F,add=T)
+#plot(e111,add=T)
+plot(aus_reg2,border="grey",lwd=3,add=T)
+image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,-0.2,0.2,1,2,3,4),
+           lab.breaks=c(-70,-20,-10,-5,"no tree","0 or",5,10,20,70),
+           legend.only=T,horizontal=F,
+           col=c("#C51B7D","#DE77AE","#F1B6DA","#FDE0EF","white",
+                 "#E6F5D0","#B8E186","#7FBC41","#4D9221"),
+           legend.shrink=0.7,
+           legend.width=0.7,
+           legend.args=list(text=expression(paste(Delta,
+                            "tree (% ground)",sep="")),
+                            line=0.5,cex=0.9,font=2),
+           axis.args=list(cex.axis=0.8),graphics.reset=F)
+dev.off()
+
+jpeg("../figures/tc_nswvic_9y.jpg", width=960, height=960)
+par(mar=c(3,3,1,7))
+plot(e222)
+image(mod_tc_r2_9y.z,col=plotclr,breaks=c(-70,-20,-10,-5,0,5,10,20,70),legend=F,add=T)
+#plot(e111,add=T)
+plot(aus_reg2,border="grey",lwd=3,add=T)
+image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,-0.2,0.2,1,2,3,4),
+           lab.breaks=c(-70,-20,-10,-5,"no tree","0 or",5,10,20,70),
+           legend.only=T,horizontal=F,
+           col=c("#C51B7D","#DE77AE","#F1B6DA","#FDE0EF","white",
+                 "#E6F5D0","#B8E186","#7FBC41","#4D9221"),
+           legend.shrink=0.7,
+           legend.width=0.7,
+           legend.args=list(text=expression(paste(Delta,
+                            "tree (% ground)",sep="")),
+                            line=0.5,cex=0.9,font=2),
+           axis.args=list(cex.axis=0.8),graphics.reset=F)
+dev.off()
+
+jpeg("../figures/tc_nswvic.jpg", width=960, height=960)
+par(mar=c(3,3,1,7))
+plot(e222)
+image(mod_tc_r2.z,col=plotclr,breaks=c(-70,-20,-10,-5,0,5,10,20,70),legend=F,add=T)
+#plot(e111,add=T)
+plot(aus_reg2,border="grey",lwd=3,add=T)
+image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,-0.2,0.2,1,2,3,4),
+           lab.breaks=c(-70,-20,-10,-5,"no tree","0 or",5,10,20,70),
+           legend.only=T,horizontal=F,
+           col=c("#C51B7D","#DE77AE","#F1B6DA","#FDE0EF","white",
+                 "#E6F5D0","#B8E186","#7FBC41","#4D9221"),
+           legend.shrink=0.7,
+           legend.width=0.7,
+           legend.args=list(text=expression(paste(Delta,
+                            "tree (% ground)",sep="")),
+                            line=0.5,cex=0.9,font=2),
+           axis.args=list(cex.axis=0.8),graphics.reset=F)
+dev.off()
+
+jpeg("../figures/tc_qld.jpg", width=960, height=960)
 par(mar=c(3,3,1,7))
 plot(e111)
-  image(mod_tc_r1.z,col=plotclr,breaks=c(-70,-50,-20,0,20,50,70),legend=F,add=T)
-  plot(e111,add=T)
-  plot(mdb_reg2,border="grey",lwd=3,add=T)
-  image.plot(zlim=c(-3,3),breaks=c(-3,-2,-1,-0.2,0.2,1,2,3),lab.breaks=c(-70,-50,-20,"no tree","0 or",20,50,70),legend.only=T,horizontal=F,col=c("#C51B7D","#E9A3C9","#FDE0EF","white","#E6F5D0","#A1D76A","#4D9221"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text=expression(paste(Delta,"tree (% ground)",sep="")),line=0.5,cex=0.9,font=2),axis.args=list(cex.axis=0.8),graphics.reset=F)
+image(mod_tc_r1.z,col=plotclr,breaks=c(-70,-20,-10,-5,0,5,10,20,70),legend=F,add=T)
+#plot(e111,add=T)
+plot(aus_reg2,border="grey",lwd=3,add=T)
+image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,-0.2,0.2,1,2,3,4),
+           lab.breaks=c(-70,-20,-10,-5,"no tree","0 or",5,10,20,70),
+           legend.only=T,horizontal=F,
+           col=c("#C51B7D","#DE77AE","#F1B6DA","#FDE0EF","white",
+                 "#E6F5D0","#B8E186","#7FBC41","#4D9221"),
+           legend.shrink=0.7,
+           legend.width=0.7,
+           legend.args=list(text=expression(paste(Delta,
+                            "tree (% ground)",sep="")),
+                            line=0.5,cex=0.9,font=2),
+           axis.args=list(cex.axis=0.8),graphics.reset=F)
 dev.off()
 
-  setEPS()
-  postscript("tc_nswvic.eps")
-  par(mar=c(3,3,1,7))
-  plot(e222)
-  image(mod_tc_r2.z,col=plotclr,breaks=c(-70,-50,-20,0,20,50,70),legend=F,add=T)
-  plot(e222,add=T)
-  plot(aus_reg2,border="grey",add=T)
-  plot(mdb_reg2,border="grey",lwd=3,add=T)
-#   image.plot(zlim=c(-3,3),breaks=c(-3,-2,-1,-0.2,0.2,1,2,3),lab.breaks=c(-70,-50,-20,"no tree","0 or",20,50,70),legend.only=T,horizontal=F,col=c("#C51B7D","#E9A3C9","#FDE0EF","white","#E6F5D0","#A1D76A","#4D9221"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text=expression(paste(Delta,"tree (% ground)",sep="")),line=0.5,cex=0.7,font=2),axis.args=list(cex.axis=0.8))
-# legend(149,-36.5,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=0.9)
-  dev.off()
-
-setEPS()
-postscript("map.eps")
+jpeg("../figures/map.jpg", width=960, height=960)
 par(mar=c(1.5,1.5,1.5,1.5))
@@
 
-Specifically, this study focused on the effect of 2003 - 2004 land clearings in the QLD region and the effect of 2003 bushfires in the NSW/VIC region.
-These events are expected to cause a step change in the local rainfall.
-The actual tree cover change at the pixel level during this time was derived from the 11-year MODIS data (discussed below).
+The hypothesis in this study therefore is that the effect of 2003 - 2004 land clearings in the QLD region and the 2003 bushfires in the NSW/VIC region cause a step change in the local rainfall.
+The actual tree cover change at the pixel level during this time was derived from the 15-year MODIS data (discussed below).
 The difference of tree cover before and after the land disturbance was tested using a Student's t-test.
-As the length of the tree cover data is shorter than the length of the rainfall data, earlier land clearings in the QLD region cannot be identified spatially, hence they are excluded from the analysis.
+As the length of the tree cover data is shorter than available rainfall data, earlier land clearings in the QLD region cannot be identified spatially, hence they are excluded from the analysis.
 
@@
 This dataset provides estimates of percent tree cover (percentage of ground surface covered by trees) at a grid resolution of 250 m \citep{Townshend2011}, which is finer then the earlier mentioned burned product MCD45A1.
-The dataset is available on an annual time interval for the period of 2000 - 2010.
+The dataset is available on an annual time interval for the study period of 2000 - 2015.
 The tree cover data was produced from 16-day Terra MODIS Land Surface Reflectance data and Land Surface Temperature \citep{Townshend2011}.
@@
 This dataset, developed by Geoscience Australia and Australian Bureau of Agricultural and Resource Economics and Sciences (ABARES), is the first nationally consistent and thematically comprehensive land cover reference for Australia.
-The DLCD is based on the 16-day Enhanced Vegetation Index (EVI), again from the MODIS satellite, between April 2000 and April 2008.
+The DLCD is based on the 16-day Enhanced Vegetation Index (EVI), again from the MODIS satellite, between April 2000 and April 2015.
 It also has a grid resolution of 250 m.
-The dataset provides information on the final land cover types (as in 2008) and estimated trend of EVI statistics (annual mean, maximum and minimum).
+The dataset provides information on the final land cover types (as in 2015) and estimated trend of EVI statistics (annual mean, maximum and minimum).
 
@@
 
-Rainfall data for Australia \citep{Jones2009} was obtained from BoM.
+Rainfall data for Australia \citep{Jones2009} was obtained from the Bureau of Meteorology.
 The data has been projected onto a national 0.05\textdegree $\times$ 0.05\textdegree grid (approximately 5 km$\times$5 km).
@@
 The data is available on a monthly basis from 1900 to current.
-Here a subset of 30 years (1979 - 2008) was used.
+Here a subset of 36 years (1979 - 2015) was used.
 The study was conducted on monthly data, as a land cover change effect on annual rainfall might be negligible but can often found to be significant in particular months or seasons \citep[e.g.][]{Otterman1990,Gaertner2001,Semazzi2001,Oleson2004,Deo2009}.
@@
 
-# setwd("R:\\AVSproject\\WrittenWork\\PhD_thesis\\chap3")
-# setwd("B:/My Study/PhD/AVSproject_laptop/WrittenWork/PhD_thesis/chap3")
-#setwd("B:/My Study/PhD/AVSproject_PhD/WrittenWork/Paper/Paper 1 - chapter 3")
-
-library(reshape)
-library(chron)
-library(fields)
-library(rgdal)
-library(raster)
-library(maptools)
-library(graphics)
-library(zoo)
-library(RColorBrewer) # creates nice color schemes
-# library(classInt)     # finds class intervals for continuous variables
-
-library(mgcv)
-library(ncdf)
-library(nnet)
-
 ## climatic indicator data
-
-file1 <- "Data\\BoM\\SOI\\SOI.csv"
-SOI <- read.csv(file1,sep=",",skip=6)
-SOI <- melt(SOI,id="Year")
-names(SOI) <- c("Year","Month","SOI")
-SOI <- SOI[order(SOI$Year,SOI$Month),]
-SOI_ts <- ts(SOI$SOI,start=c(1876,1),end=c(2014,4),frequency=12)
-SOI_100a <- window(SOI_ts,start=c(1900,1),end=c(2008,12))
-SOI_30a <- window(SOI_ts,start=c(1979,1),end=c(2008,12))
-## deseasonalised SOI
-SOI_100 <- stl(SOI_100a,"period")$time.series[,3]+stl(SOI_100a,"period")$time.series[,2]
-SOI_30 <- stl(SOI_30a,"period")$time.series[,3]+stl(SOI_30a,"period")$time.series[,2]
-
-file2 <- "Data\\BoM\\SST\\SST34.cdf"
-SST34 <- open.ncdf(file2)
-SST34 <- get.var.ncdf(SST34,SST34$var[[1]])
-SST34_ts <- ts(SST34,start=c(1856,1),end=c(2014,4),frequency=12)
-SST34_30a <- window(SST34_ts,start=c(1979,1),end=c(2008,12))
-## deseasonalised SST34
-SST34_30 <- stl(SST34_30a,"period")$time.series[,3]+stl(SST34_30a,"period")$time.series[,2]
-
-file3 <- "Data\\BoM\\SST\\SST3.cdf"
-SST3 <- open.ncdf(file3)
-SST3 <- get.var.ncdf(SST3,SST3$var[[1]])
-SST3_ts <- ts(SST3,start=c(1856,1),end=c(2014,4),frequency=12)
-SST3_30a <- window(SST3_ts,start=c(1979,1),end=c(2008,12))
-## deseasonalised SST34
-SST3_30 <- stl(SST3_30a,"period")$time.series[,3]+stl(SST3_30a,"period")$time.series[,2]
-
-file4 <- "Data\\BoM\\SST\\SST4.cdf"
-SST4 <- open.ncdf(file4)
-SST4 <- get.var.ncdf(SST4,SST4$var[[1]])
-SST4_ts <- ts(SST4,start=c(1856,1),end=c(2014,5),frequency=12)
-SST4_30a <- window(SST4_ts,start=c(1979,1),end=c(2008,12))
-## deseasonalised SST4
-SST4_30 <- stl(SST4_30a,"period")$time.series[,3]+stl(SST4_30a,"period")$time.series[,2]
-
-file5 <- "Data\\BoM\\PDO\\PDO.latest_1900-.txt"
-pdo_mth <- read.csv(file5,sep="",header=T)
-pdo_mth <- melt.data.frame(pdo_mth,id="YEAR")
-names(pdo_mth) <- c("Year","Month","PDO")
-pdo_mth <- pdo_mth[order(pdo_mth$Year,pdo_mth$Month),]
-PDO_ts <- ts(pdo_mth$PDO,start=c(1900,1),end=c(2014,4),frequency=12)
-PDO_100a <- window(PDO_ts,start=c(1900,1),end=c(2008,12))
-## deseasonalised PDO
-PDO_100 <- stl(PDO_100a,"period")$time.series[,3]+stl(PDO_100a,"period")$time.series[,2]
-
-file7 <- "Data\\BoM\\IOD\\iod_jamstec.txt"
-IOD <- read.csv(file7,sep="",header=T)
-IOD <- IOD[,4]
-IOD_ts <- ts(IOD,start=c(1870,1),frequency=12)
-IOD_30a <- window(IOD_ts,start=c(1979,1),end=c(2008,12))
-## deseasonalised IOD
-IOD_30 <- stl(IOD_30a,"period")$time.series[,3]+stl(IOD_30a,"period")$time.series[,2]
-# end of read data
-
+# 
+# file1 <- "..\\Data\\BoM\\SOI\\SOI.csv"
+# SOI <- read_csv(file1)
+# SOI <- gather(SOI,key="Month", value="SOI",Jan:Dec)
+# SOI <- SOI[order(as.Date(paste(SOI$Year,SOI$Month,
+#                                         "01",sep="-"), "%Y-%b-%d")),]
+# SOI_100a <- SOI %>%
+#   filter(Year > 1899 & Year < 2016)
+# SOI_30a <- SOI %>%
+#   filter(Year > 1978 & Year < 2016)
+# ## deseasonalised SOI
+# SOI_100 <- ds(SOI_100a$SOI, ic = "AIC", type = "monthly",
+#               standardizeQ=F)$z
+# SOI_30 <- ds(SOI_30a$SOI, ic = "AIC", type = "monthly",
+#              standardizeQ=F)$z
+# 
+# file2 <- "..\\Data\\BoM\\SST\\SST34_anomaly.nc"
+# SST34 <- nc_open(file2)
+# SST34_var <- ncvar_get(SST34,SST34$var[[1]])
+# nc_close(SST34)
+# SST34_df <- tibble(Date = seq.Date(as.Date("1856-01-01"),
+#                                    as.Date("2018-04-01"),
+#                                    by="month"),
+#                    SST34 = SST34_var)
+# SST34_100a <- SST34_df %>%
+#   filter(Date >= "1900-01-01" & Date < "2016-01-01")
+# SST34_30a <- SST34_df %>%
+#   filter(Date >= "1979-01-01" & Date < "2016-01-01")
+# ## deseasonalised SST34
+# SST34_30 <- ds(SST34_30a$SST34,ic = "AIC", type="monthly",
+#                standardizeQ=F)$z
+# SST34_100 <- ds(SST34_100a$SST34,ic = "AIC", type="monthly",
+#                standardizeQ=F)$z
+# 
+# file3 <- "..\\Data\\BoM\\SST\\SST3_anomaly.nc"
+# SST3 <- nc_open(file3)
+# SST3_var <- ncvar_get(SST3,SST3$var[[1]])
+# nc_close(SST3)
+# SST3_df <- tibble(Date = seq.Date(as.Date("1856-01-01"),
+#                                    as.Date("2018-04-01"),
+#                                    by="month"),
+#                    SST3 = SST3_var)
+# SST3_100a <- SST3_df %>%
+#   filter(Date >= "1900-01-01" & Date < "2016-01-01")
+# SST3_30a <- SST3_df %>%
+#   filter(Date >= "1979-01-01" & Date < "2016-01-01")
+# ## deseasonalised SST3
+# SST3_30 <- ds(SST3_30a$SST3,ic = "AIC", type="monthly",
+#               standardizeQ=F)$z
+# SST3_100 <- ds(SST3_100a$SST3,ic = "AIC", type="monthly",
+#               standardizeQ=F)$z
+# 
+# file4 <- "..\\Data\\BoM\\SST\\SST4_anomaly.nc"
+# SST4 <- nc_open(file4)
+# SST4_var <- ncvar_get(SST4,SST4$var[[1]])
+# nc_close(SST4)
+# SST4_df <- tibble(Date = seq.Date(as.Date("1856-01-01"),
+#                                    as.Date("2018-04-01"),
+#                                    by="month"),
+#                    SST4 = SST4_var)
+# SST4_100a <- SST4_df %>%
+#   filter(Date >= "1900-01-01" & Date < "2016-01-01")
+# SST4_30a <- SST4_df %>%
+#   filter(Date >= "1979-01-01" & Date < "2016-01-01")
+# ## deseasonalised SST4
+# SST4_30 <- ds(SST4_30a$SST4,ic = "AIC", type="monthly",
+#               standardizeQ=F)$z
+# SST4_100 <- ds(SST4_100a$SST4,ic = "AIC", type="monthly",
+#               standardizeQ=F)$z
+# 
+# file5 <- "..\\Data\\BoM\\PDO\\PDO.latest_1900-.txt"
+# pdo_mth <- read_table(file5)
+# pdo_mth <- gather(pdo_mth,key="Month", value="PDO", JAN:DEC)
+# pdo_mth$YEAR <- substr(pdo_mth$YEAR,1,4)
+# pdo_mth <- pdo_mth[order(as.Date(paste(pdo_mth$YEAR,pdo_mth$Month,
+#                                         "01",sep="-"), "%Y-%b-%d")),]
+# PDO_100a <- pdo_mth %>%
+#   filter(YEAR > 1899 & YEAR < 2016)
+# ## deseasonalised PDO
+# PDO_100 <- ds(PDO_100a$PDO,ic = "AIC", type = "monthly",
+#               standardizeQ=F)$z
+# 
+# file7 <- "..\\Data\\BoM\\IOD\\iod_jamstec.txt"
+# IOD <- read_table2(file7)
+# IOD <- tibble(Date = seq.Date(as.Date("1870-01-16"),
+#                                    as.Date("2018-03-16"),
+#                                    by="month"),
+#               IOD = IOD$`DMI(=West-East)`)
+# IOD_30a <- IOD %>%
+#   filter(Date > "1979-01-01" & Date < "2016-01-01")
+# IOD_100a <- IOD %>%
+#   filter(Date > "1899-01-01" & Date < "2016-01-01")
+# ## deseasonalised IOD
+# IOD_30 <- ds(IOD_30a$IOD, ic = "AIC", type = "monthly",
+#              standardizeQ=F)$z
+# IOD_100 <- ds(IOD_100a$IOD, ic = "AIC", type = "monthly",
+#              standardizeQ=F)$z
+# # end of read data
+# # should write out useful data
+# save.image("../data/data_environment.Rdata")
 @
@@
 Hence the step changes on rainfall were analysed by two different statistical methods to provide a comparison.
-Both methods make use of a regression model to remove variability in rainfall hypothesised to be caused by factors other than vegetation cover change.
+Both methods make use of a regression model to remove variability in rainfall due to climatic factors to strengthen the vegetation cover change signal.
 In the first method, the tree cover change was implemented as a factor variable in the regression model.
@@
   \centering
-  \includegraphics[scale=0.6]{mean_ts.jpg}
+  \includegraphics[scale=0.6]{../figures/mean_ts.jpg}
   \caption{The deseasonalised and detrended rainfall over the 30 years period in (a) the QLD region and (b) the NSW/VIC region.
@@
 
-Rainfall, even a single event, is generally affected by several factors.
+%% NEEDED??Rainfall, even a single event, is generally affected by several factors.
 Many studies suggest the large-scale climate drivers, which are related to the global or regional atmospheric air mass movement, are the dominant control on the overall rainfall \citep{Maynard2003,DeAngelis2010,Holper2011,Smith2012}.
@@
 %% comment: update references?
+%% NEEDED??
 The difference in available radiation and hence ET in different seasons increase the variability in rainfall.
 The land surface effect on rainfall can be confused by changes in these factors.
-Hence a regression model was used to estimate the amount of variability in rainfall that is due to these important factors, and isolate changes resulting from vegetation cover change.
-As highlighted in the introduction, the Australian climate is influenced by sea surface temperature in the tropical Pacific and Indian Oceans, as well as pressure systems in the Southern Ocean \citep{BoM2012}.
+This study uses a flexible regression model to estimate the amount of variability in rainfall that is due to climate factors.
+This allows amplifying the local effect of evapotranspiration on rainfall.
+As highlighted in the introduction, the Australian climate is influenced by sea surface temperatures in the tropical Pacific and Indian Oceans, as well as pressure systems in the Southern Ocean \citep{BoM2012}.
 \citet{Risbey2009} compared five large-scale drivers, including ENSO (measured by SOI and the Tropical Pacific SSTs), IOD, SAM, MJO\footnote{MJO is a large scale eastward-propagating wave-like disturbance in equatorial latitudes \citep{Risbey2009}.} (Madden-Julian oscillation) and blocking, in relation to Australian rainfall variability.
 They identified SOI as the most important index among all indices tested for broad parts of Australia (including QLD and NSW/VIC) in almost any season.
-In this study, up to two important indices from the seven climatic indicators (see section \ref{sec:Data}) were used as the explanatory variables in the model for each study region.
+In this study, four indices were selected from the seven climatic indicators (see section \ref{sec:Data}) were used as the explanatory variables in the model for each study region.
 
 <<rainData,eval=F , echo = F, results = hide>>=
-
-## 100 years awap data & 30 years
-qld_100 <- read.csv("Data\\rainfall\\QLD_month_rf.csv",header=T)
-qld_30 <- subset(qld_100,as.Date(qld_100[,1],"%d/%m/%Y")>="1979-01-01" & as.Date(qld_100[,1],"%d/%m/%Y")<="2008-12-01")
-qld100_mean <- ts(rowMeans(qld_100[,2:2801],na.rm=T,dims=1),start=c(1900,1),frequency=12)
-qld30_mean <- window(qld100_mean,start=c(1979,1),end=c(2008,12))
-
-nsw_100 <- read.csv("Data\\rainfall\\NSW_month_rf.csv",header=T)
-nsw_30 <- subset(nsw_100,as.Date(nsw_100[,1],"%d/%m/%Y")>="1979-01-01" & as.Date(nsw_100[,1],"%d/%m/%Y")<="2008-12-01")
-nsw100_mean <- ts(rowMeans(nsw_100[,2:1185],na.rm=T,dims=1),start=c(1900,1),frequency=12)
-nsw30_mean <- window(nsw100_mean, start=c(1979,1),end=c(2008,12))
+load("../data/data_environment.Rdata")
+
+# read in the BOM gridded rainfall data extracted earlier
+rain <- readRDS("../data/bom/rainfall/Raindata.rds")
+# dim(rain)
+# 43829 columns (days from 19000101) by 4161 locations (rows)
+
+# split by Qld and NSW
+# NSW has first 1254 locations (columns) Qld other 2907
+rainNSW <- rain[1:1254,]
+rainQld <- rain[1255:4161,]
+rm(rain)
+
+# summarise to monthly
+# NSW
+rainNSW_t <- as.tibble(data.frame(Date = 
+                                    seq.Date(as.Date("1900-01-01"),
+                                    as.Date("2015-12-31"),by=1),
+                    rain=t(rainNSW)))
+rm(rainNSW)
+rainNSW_t <- rainNSW_t %>% 
+  mutate(Month = format(Date,"%Y-%m")) 
+rain_monthNSW <- rainNSW_t %>%
+  group_by(Month) %>%
+  summarise_at(vars(starts_with("rain")), funs(sum = sum(.)))
+
+# Qld
+rainQld_t <- as.tibble(data.frame(Date = 
+                                    seq.Date(as.Date("1900-01-01"),
+                                    as.Date("2015-12-31"),by=1),
+                    rain=t(rainQld)))
+rm(rainQld)
+rainQld_t <- rainQld_t %>% 
+  mutate(Month = format(Date,"%Y-%m")) 
+rain_monthQld <- rainQld_t %>%
+  group_by(Month) %>%
+  summarise_at(vars(starts_with("rain")), funs(sum = sum(.)))
+rm(rainNSW_t, rainQld_t)
+
+## 100 years data & 30 years
+# Qld
+qld_100 <- rain_monthQld
+qld_30 <- qld_100 %>%
+ filter(Month >= "1979-01")
+
+qld100_mean <- qld_100 %>%
+  mutate(Means = rowMeans(.[,-1])) %>%
+  dplyr::select(Month,Means)
+
+qld30_mean <- qld100_mean %>%
+ filter(Month >= "1979-01")
+
+# NSW
+nsw_100 <- rain_monthNSW
+nsw_30 <- nsw_100 %>%
+ filter(Month >= "1979-01")
+
+nsw100_mean <- nsw_100 %>%
+  mutate(Means = rowMeans(.[,-1])) %>%
+  dplyr::select(Month,Means)
+
+nsw30_mean <- nsw100_mean %>%
+ filter(Month >= "1979-01")
 
 require(MASS)
-fitdistr(qld30_mean,"gamma")  ## assume shape=1
-fitdistr(nsw30_mean,"gamma")  ## assume shape =8
-ks.test(qld30_mean,1,0.03)
-ks.test(nsw30_mean,2.4,0.03)
-
-setEPS()
-postscript("hist_qld.eps")
- hist(qld30_mean,main="",xlab=expression(paste("rainfall (mm mont",h^-1,")",sep="")))
+fitdistr(qld30_mean$Means,"gamma")  ## assume shape=0.9
+fitdistr(nsw30_mean$Means,"gamma")  ## assume shape =2.19
+ks.test(qld30_mean$Means,"pgamma",0.90,0.03)
+ks.test(nsw30_mean$Means,"pgamma",2,0.03)
+
+jpeg("../figures/hist_rainfall.jpg", width=960, height=480)
+par(mfrow=c(1,2))
+hist(qld30_mean$Means,main="",xlab=expression(paste("rainfall (mm mont",h^-1,")",sep="")))
+ hist(nsw30_mean$Means,main="",xlab=expression(paste("rainfall (mm mont",h^-1,")",sep="")))
 dev.off()
 
-setEPS()
-postscript("hist_nswvic.eps")
- hist(nsw30_mean,main="",xlab=expression(paste("rainfall (mm mont",h^-1,")",sep="")))
-dev.off()
-
-## autocorrelation in rainfall
-setEPS()
-postscript("autocorrelation.eps")
-par(mfrow=c(2,2))
-  acf(qld30_mean,lag.max=12,main="QLD")
-  acf(nsw30_mean,lag.max=12,main="NSW & VIC")
-
-  pacf(qld30_mean,lag.max=12,main="")
-  pacf(nsw30_mean,lag.max=12,main="")  
+# autocorrelation in rainfall
+acf_df <- data_frame(lag = 1:12, 
+                     acf_Qld = acf(qld30_mean$Means,
+                                   lag.max=12,main="QLD")$acf[2:13,1,1],
+                     acf_NSW = acf(nsw30_mean$Means,
+                                   lag.max=12,main="NSW & VIC")$acf[2:13,1,1],
+                     pacf_Qld = pacf(qld30_mean$Means,
+                                     lag.max=12,main="")$acf[,1,1],
+                     pacf_NSW = pacf(nsw30_mean$Means,
+                                     lag.max=12,main="")$acf[,1,1])
+
+jpeg("../figures/autocorrelation.jpg")
+acf_df %>%
+  gather(key="acf", value="cor", acf_Qld:pacf_NSW) %>%
+  ggplot(aes(lag, cor)) + geom_bar(stat="identity", fill = "blue") +
+  facet_wrap(~acf) + xlab("Lag") + ylab("correlation")
 dev.off()
@@
 ## use stl function in R
-  qld30_deseason <- stl(qld30_mean,s.window=length(qld30_mean),t.window=length(qld30_mean))$time.series[,3]
-  nsw30_deseason <- stl(nsw30_mean,s.window=length(nsw30_mean),t.window=length(nsw30_mean))$time.series[,3]
-  qld100_deseason <- stl(qld100_mean,s.window=length(qld100_mean),t.window=length(qld100_mean))$time.series[,3]
-  nsw100_deseason <- stl(nsw100_mean,s.window=length(nsw100_mean),t.window=length(nsw100_mean))$time.series[,3]
-# 
-# setEPS()
-# postscript("qld_mean_ts.eps")
-# par(cex.lab=1.3,cex.axis=1.3)
-# plot(qld30_deseason,xlab="Year",ylab="monthly rainfall")
-# abline(v=2003,col="red",lwd=2)
-# text(2004.6,150,"2003",cex=1.5,col="red")
-# dev.off()
-# 
-# setEPS()
-# postscript("nswvic_mean_ts.eps")
-# par(cex.lab=1.3,cex.axis=1.3)
-# plot(nsw30_deseason,xlab="Year",ylab="monthly rainfall")
-# abline(v=2003,col="red",lwd=2)
-# text(2004.6,120,"2003",cex=1.5,col="red")
-# dev.off()
-#   
-# setEPS()
-# postscript("cor_qld.eps")
-# par(mfrow=c(2,3),mar=c(5,4,1,1))
-# ccf(as.vector(SOI_30),as.vector(qld30_deseason),lag.max=12,xlim=c(-12,0),xlab="SOI  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# ccf(as.vector(SST34_30),as.vector(qld30_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO3.4  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# ccf(as.vector(SST4_30),as.vector(qld30_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO4  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# ccf(as.vector(PDO_100),as.vector(qld100_deseason),lag.max=24,xlim=c(-24,0),xlab="PDO Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# ccf(as.vector(PDO_100*SOI_100),as.vector(qld100_deseason),lag.max=12,xlim=c(-12,0),xlab="PDO*SOI  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# ccf(as.vector(IOD_30),as.vector(qld30_deseason),lag.max=12,xlim=c(-12,0),xlab="IOD  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# dev.off()
-# 
-# # ccf(subset(SOI_30,cycle(SOI_30)==10|cycle(SOI_30)==11|cycle(SOI_30)==12),subset(qld_deseason,cycle(qld_deseason)==10|cycle(qld_deseason)==11|cycle(qld_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="SOI  Lag",ylab="CCF") ## SOI
-# # ccf(subset(SST34_30,cycle(SST34_30)==10|cycle(SST34_30)==11|cycle(SST34_30)==12),subset(qld_deseason,cycle(qld_deseason)==10|cycle(qld_deseason)==11|cycle(qld_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="SST34  Lag",ylab="CCF") ## SST34
-# # ccf(subset(SST3_30,cycle(SST3_30)==10|cycle(SST3_30)==11|cycle(SST3_30)==12),subset(qld_deseason,cycle(qld_deseason)==10|cycle(qld_deseason)==11|cycle(qld_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="SST3  Lag",ylab="CCF") ## SST3
-# # ccf(subset(PDO_30,cycle(PDO_30)==10|cycle(PDO_30)==11|cycle(PDO_30)==12),subset(qld_deseason,cycle(qld_deseason)==10|cycle(qld_deseason)==11|cycle(qld_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="PDO  Lag",ylab="CCF") ## PDO
-# # ccf(subset(PDO_30,cycle(PDO_30)==10|cycle(PDO_30)==11|cycle(PDO_30)==12)* subset(SOI_30,cycle(SOI_30)==10|cycle(SOI_30)==11|cycle(SOI_30)==12),subset(qld_deseason,cycle(qld_deseason)==10|cycle(qld_deseason)==11|cycle(qld_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="PDO  Lag",ylab="CCF") ## PDO
-# # ccf(subset(IOD_30,cycle(IOD_30)==10|cycle(IOD_30)==11|cycle(IOD_30)==12),subset(qld_deseason,cycle(qld_deseason)==10|cycle(qld_deseason)==11|cycle(qld_deseason)==12),lag.max=3,xlim=c(-1,0),xlab="IOD  Lag",ylab="CCF") ## IOD
-# # 
-# # 
-# # 
+
+## revision June 2018: change this to only deseasonalise using the 
+# deseasonalise package
+qld30_deseason <- ds(qld30_mean$Means,ic="AIC", type="monthly",
+                     standardizeQ=F)$z
+nsw30_deseason <- ds(nsw30_mean$Means,ic="AIC", type="monthly",
+                     standardizeQ=F)$z
+qld100_deseason <- ds(qld100_mean$Means,ic="AIC", type="monthly",
+                      standardizeQ=F)$z
+nsw100_deseason <- ds(qld100_mean$Means,ic="AIC", type="monthly",
+                      standardizeQ=F)$z
+
+# test whether any linear trends are significant
+df_list <- list(qld30 = data.frame(z=qld30_deseason,
+                           trend=1:length(qld30_deseason)),
+                nsw30 = data.frame(z=nsw30_deseason,
+                           trend=1:length(nsw30_deseason)),
+                qld100 = data.frame(z=qld100_deseason,
+                 trend=1:length(qld100_deseason)),
+                nsw100 = data.frame(z=nsw100_deseason,
+                 trend=1:length(nsw100_deseason)))
+result <- lapply(df_list,
+                 function(x) summary(lm(z~trend,
+                                        data=x))$coefficients)
+# none have a trend
+
+# 
+plot_df <- tibble(Dates = rep(ymd(paste(qld30_mean$Month,
+                                        "01",sep="_")),2),
+                  Resid = c(qld30_deseason,nsw30_deseason),
+                  State = c(rep("Qld",length(qld30_deseason)),
+                            rep("NSW",length(nsw30_deseason))))
+
+jpeg("../figures/Rainfall_resid.jpg", width=960, height=960)
+plot_df %>%
+  ggplot(aes(Dates,Resid)) + geom_line() + 
+  geom_vline(xintercept = as.Date("2003-01-01"), 
+             col="red", size=1.5) +
+  geom_text(aes(x=as.Date("2004-12-01"),y=150),
+            label="2003", col="red", size=7) +
+  facet_wrap(~State, ncol = 1) + xlab("Date") +
+  ylab("Deseasonalised rainfall for the NSW and Qld locations") +
+  theme(axis.text=element_text(size=rel(2)),
+        axis.title=element_text(size=rel(2.5)),
+        strip.text =element_text(size=rel(2), face="bold"))
+dev.off()
+
+# Figure 5
+plot_df_qld <- tibble(Lag = ccf(as.vector(SOI_30),
+                            as.vector(qld30_deseason),lag.max=12, 
+                            plot=F)$lag[,,1],
+                  SOI = ccf(as.vector(SOI_30),
+                            as.vector(qld30_deseason),lag.max=12, 
+                            plot=F)$acf[,,1],
+                  NINO3.4 = ccf(as.vector(SST34_30),
+                                as.vector(qld30_deseason),lag.max=12,
+                                plot=F)$acf[,,1],
+                  NINO4 = ccf(as.vector(SST4_30),
+                              as.vector(qld30_deseason),lag.max=12,
+                              plot=F)$acf[,,1],
+                  PDO = ccf(as.vector(PDO_100),
+                            as.vector(qld100_deseason),lag.max=12,
+                            plot=F)$acf[,,1],
+                  IOD = ccf(as.vector(IOD_30),
+                            as.vector(qld30_deseason),lag.max=12, 
+                            plot=F)$acf[,,1])
+
+jpeg("../figures/cor_qld.jpg", width=960, height=960)
+plot_df_qld %>%
+  gather(key="SST", value="correlation", SOI:IOD) %>%
+  ggplot(aes(Lag,correlation)) + 
+  geom_bar(fill="blue", stat="identity") +
+  facet_wrap(~SST) + xlab("Lag in months") + ylab("correlation") +
+  xlim(c(-12.5,0.5)) +
+  theme(axis.text=element_text(size=rel(2)),
+        axis.title=element_text(size=rel(2.5)),
+        strip.text =element_text(size=rel(2), face="bold"))
+dev.off()
+
+# to add and calculate ci: qnorm((1 + 0.95)/2)/sqrt(x$n.used)
+
+plot_df <- tibble(Lag = ccf(as.vector(SOI_30),
+                            as.vector(nsw30_deseason),lag.max=12, 
+                            plot=F)$lag[,,1],
+                  SOI = ccf(as.vector(SOI_30),
+                            as.vector(nsw30_deseason),lag.max=12, 
+                            plot=F)$acf[,,1],
+                  NINO3.4 = ccf(as.vector(SST34_30),
+                                as.vector(nsw30_deseason),lag.max=12,
+                                plot=F)$acf[,,1],
+                  NINO4 = ccf(as.vector(SST4_30),
+                              as.vector(nsw30_deseason),lag.max=12,
+                              plot=F)$acf[,,1],
+                  PDO = ccf(as.vector(PDO_100),
+                            as.vector(nsw100_deseason),lag.max=12,
+                            plot=F)$acf[,,1],
+                  IOD = ccf(as.vector(IOD_30),
+                            as.vector(nsw30_deseason),lag.max=12, 
+                            plot=F)$acf[,,1])
+
+jpeg("../figures/cor_nswvic.jpg", width=960, height=960)
+plot_df %>%
+  gather(key="SST", value="correlation", SOI:IOD) %>%
+  ggplot(aes(Lag,correlation)) + 
+  geom_bar(fill="blue", stat="identity") +
+  facet_wrap(~SST) + xlab("Lag in months") + ylab("correlation") +
+  xlim(c(-12.5,0.5)) +
+  theme(axis.text=element_text(size=rel(2)),
+        axis.title=element_text(size=rel(2.5)),
+        strip.text =element_text(size=rel(2), face="bold"))
+dev.off()
+
 # # ## 1,2,3: SOI
@@
 # #
-# 
-# setEPS()
-# postscript("cor_nswvic.eps")
-# par(mfrow=c(2,3),mar=c(5,4,1,1))
-# ccf(as.vector(SOI_30),as.vector(nsw30_deseason),lag.max=12,xlim=c(-12,0),xlab="SOI  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# ccf(as.vector(SST34_30),as.vector(nsw30_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO3.4  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# ccf(as.vector(SST4_30),as.vector(nsw30_deseason),lag.max=12,xlim=c(-12,0),xlab="NINO4  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# ccf(as.vector(PDO_100),as.vector(nsw100_deseason),lag.max=24,xlim=c(-24,0),xlab="PDO Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# ccf(as.vector(PDO_100*SOI_100),as.vector(nsw100_deseason),lag.max=12,xlim=c(-12,0),xlab="PDO*SOI  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# ccf(as.vector(IOD_30),as.vector(nsw30_deseason),lag.max=12,xlim=c(-12,0),xlab="IOD  Lag",ylab="CCF",cex.axis=1.2,cex.lab=1.3)
-# dev.off()
-# 
-# ccf(subset(SOI_30,cycle(SOI_30)==7|cycle(SOI_30)==8|cycle(SOI_30)==9),subset(nswvic_deseason,cycle(nswvic_deseason)==7|cycle(nswvic_deseason)==8|cycle(nswvic_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="SOI  Lag",ylab="CCF") ## SOI
-# ccf(subset(SST34_30,cycle(SST34_30)==7|cycle(SST34_30)==8|cycle(SST34_30)==9),subset(nswvic_deseason,cycle(nswvic_deseason)==7|cycle(nswvic_deseason)==8|cycle(nswvic_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="SST34  Lag",ylab="CCF") ## SST34
-# ccf(subset(SST3_30,cycle(SST3_30)==7|cycle(SST3_30)==8|cycle(SST3_30)==9),subset(nswvic_deseason,cycle(nswvic_deseason)==7|cycle(nswvic_deseason)==8|cycle(nswvic_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="SST3  Lag",ylab="CCF") ## SST3
-# ccf(subset(PDO_30,cycle(PDO_30)==7|cycle(PDO_30)==8|cycle(PDO_30)==9),subset(nswvic_deseason,cycle(nswvic_deseason)==7|cycle(nswvic_deseason)==8|cycle(nswvic_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="PDO  Lag",ylab="CCF") ## PDO
-# ccf(subset(PDO_30,cycle(PDO_30)==7|cycle(PDO_30)==8|cycle(PDO_30)==9)* subset(SOI_30,cycle(SOI_30)==7|cycle(SOI_30)==8|cycle(SOI_30)==9),subset(nswvic_deseason,cycle(nswvic_deseason)==7|cycle(nswvic_deseason)==8|cycle(nswvic_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="PDO  Lag",ylab="CCF") ## PDO
-# ccf(subset(IOD_30,cycle(IOD_30)==7|cycle(IOD_30)==8|cycle(IOD_30)==9),subset(nswvic_deseason,cycle(nswvic_deseason)==7|cycle(nswvic_deseason)==8|cycle(nswvic_deseason)==9),lag.max=3,xlim=c(-1,0),xlab="IOD  Lag",ylab="CCF") ## IOD
-# 
-# 
-# 
-# ## 1,2,3: NONE
-# #4,5,6: SOI
-# # 7, 8, 9: SOI, IOD
-# ## 10, 11, 12: SOI, SST34, SST3, IOD
 
@@
 As a preliminary step, the correlations between rainfall and each climatic index were analysed.
-Rainfall in each study region was first deseasonalised and detrended using the seasonal decomposition function ``stl" in R \citep{Rstats2011}.
+Rainfall in each study region was first deseasonalised and detrended using the seasonal decomposition function `stl` in R \citep{Rstats2018}.
 The reasons for using detrended data was discussed in \citet{Smith2012}.
@@
 \centering
-\includegraphics[bb=1 1 500 350, scale=0.6]{cor_qld.jpg}
- \caption{Cross-correlation of the six indices and rainfall in QLD study region.
-Where the correlations with PDO are sought, 108-year rainfall data (1900 - 2008) are used.
-Otherwise, 30-year rainfall data are used.
-The correlation with NINO 3 is not shown as it is very similar to but weaker than the case of NINO 3.4.
-The blue dashed lines indicate the 95\% confidence interval.}
+\includegraphics[bb=1 1 500 350, scale=0.6]{../figures/cor_qld.tif}
+ \caption{Cross-correlation of six climate indices and rainfall in QLD study region.
+For the PDO analysis, 108-year rainfall data (1900 - 2008) are used.
+Otherwise, 36-year rainfall data are used.
+The correlation with NINO 3 is not shown as it is very similar to but weaker than for NINO 3.4.}
   \label{fig:cor_rain_qld}
@@
 \centering
-\includegraphics[bb=1 1 500 550, scale=0.6]{cor_nswvic.jpg}
- \caption{Cross-correlation of the six indices and rainfall in NSW/VIC study region.}
+\includegraphics[bb=1 1 500 550, scale=0.6]{../figures/cor_nswvic.tif}
+ \caption{Cross-correlation of six climate indices and rainfall in NSW/VIC study region.}
   \label{fig:cor_rain_nsw}
@@
 
-Based on the correlation between the climatic indices and rainfall (as shown in Figure~\ref{fig:cor_rain_qld} and Figure~\ref{fig:cor_rain_nsw}), the following results were found:
+Based on the correlation between the climatic indices and rainfall (as shown in Figure~\ref{fig:cor_rain_qld} and Figure~\ref{fig:cor_rain_nsw}), it can be concluded that:
 \begin{itemize}
@@
   \setlength{\parskip}{0cm}
-  \item In QLD, the correlation between rainfall and SOI at zero time lags is the highest across all indices, outweighing the other ENSO indicators. 
+  \item In QLD, the correlation between rainfall and SOI at zero time lags is the highest across all indices, outweighing the other ENSO indicators. IOD, had a weak influence in QLD.
   \item In NSW/VIC, again the SOI has the highest correlation with rainfall, followed by the IOD. Both occur at the zero time lags. 
+  \item in both cases PDO had the weakest correlations, and this factor was not considered further as a predictor.
 \end{itemize}
@@
 Concurrent climatic index series were generally found most useful in rainfall prediction \citep[e.g.][]{Risbey2009,Kamruzzaman2011}.
-The correlations between the climatic indices and rainfall for each individual season have also been tested.
-The results were similar.
-So there was no change in the results as we were interested at in using no more than two indices in the model.
+The correlations between the climatic indices and rainfall for each individual season have also been tested, and the results were similar.
 
@@
 This character is given by the movement of subtropical high pressure system which dominates Australian climate \citep{BoM2012a}.
-The seasonal component of rainfall has a periodic pattern so it is better modelled as a smooth term.
-A spline function was applied on the months to define a smooth seasonal pattern \citep{Wood2011}.
-%The spline function ``s" in R was used .
+The seasonal component of rainfall has a periodic pattern which means this should be included in the model.
+%The spline function ``s" in R was used.
 
@@
 
-We assumed all the factors are additive components in determining rainfall as in \citet{Kamruzzaman2011}.
+We assumed all the factors are additive smooth components in determining rainfall as in \citet{Kamruzzaman2011}.
 Generally, monthly rainfall has a skewed distribution so the normality assumption of the residuals in a general linear model could be violated.
 In this case, the rainfall model is expressed as a generalised additive model (GAM) \citep{Hastie1986} with a log link function g(), assuming the residuals are gamma distributed (see Figure~\ref{fig:hist_rain}).
+This means all predictors are modelled as smooth functions, in this case using penalized regression splines \citep{Woood2011}.
 \vspace{0.5cm}
@@
 \begin{equation}
-g(E(\mathbf{R}_r)) = \beta_0 + s_1(D_{1,r}\mathbf{SOI}) + s_2(D_{2,r}\mathbf{IOD}) + s_3(\mathbf{Season}) + \beta_1\mathbf{Trend} + \boldsymbol{\epsilon}_r
+g(E(\mathbf{R}_r)) = \beta_0 + s_1(\mathbf{SOI}) + s_2(\mathbf{IOD}) + s_3(\mathbf{Nino3.4}) + s_4(\mathbf{Nino4}) + s_5(\mathbf{Season}) + \beta_1\mathbf{Trend} + \boldsymbol{\epsilon}_r
   \label{eqn:model}
@@
 The bold letters represent the time series vectors.
-The subscript $r$ denotes the region, $r=$qld or nswvic.
-$\beta_u$ ($_u$=0, 1) are the fitted coefficients in the model.
-$s_v$ ($_v$=1, 2, 3) are the smooth functions on the climatic indices and the season.
-$D_{1,r}$ and $D_{2,r}$ switch on/off the corresponding climatic index in the model as discussed previously.
-\begin{equation}
-  D_{1,r} = 1 
-\end{equation}
-as SOI was used in both regions as rainfall predictor.
-\begin{equation}
-  D_{2,r} = \left\{ \begin{array}{ll}
-   1 &\mbox{for NSW/VIC } \\
-   0 &\mbox{for QLD}
-  \end{array} \right.
-  \label{eqn:dummyD}
-\end{equation}
-The linear long term trend in the rainfall data is modelled by \textbf{Trend}=1,2,3...n, where n is the total number of months in the time series.
-\textbf{Season} is the seasonal component which is represented by applying a smooth spline function on the month values (1-12.
-The monthly rainfalls in both regions appear to have a lag 1 autocorrelation, being 0.26 for QLD and 0.33 for NSW/VIC.
+The region is indicated by $r$, while $\beta_u$ ($_u$=0, 1) are the fitted coefficients in the model.
+$s_v$ ($_v$=1, 2, 3) are the smooth penalized regression spline functions on the climatic indices and the season.
+Apart from dropping the PDO as a predictor, all other climate indices were included, allowing the model to select the appropriate predictors.
+
+A possible linear long term trend in the rainfall data is modelled by \textbf{Trend}=1,2,3...n, where n is the total number of months in the time series. \textbf{Season} is the seasonal component.  
 %% comment: Do we have to explain why we don't include lag 1 autocorrelation in the model, or use mixed modelling?
-The SOI and IOD terms are also modelled with spline functions.
-As the effect of large scale drivers on Australian rainfall is more likely to be seasonal \citep{Murphy2008,Schepen2012}, the spline function can closely reproduce the high and low impacts of the climatic indices.
+The climatic terms are also modelled with smooth functions.
+The effect of large scale drivers on Australian rainfall is more likely to be seasonal \citep{Murphy2008,Schepen2012}, and the regression spline function is more flexible in reproducing the variability in impacts of the climatic indices.
 
@@
   \caption{Distribution of monthly rainfall in (a) QLD and (b) NSW/VIC.
-By using a Kolmogorov-Smirnov test with shape = 1 and 2.4 respectively, rainfalls in both regions are shown to have a gamma distribution.}
+Using a Kolmogorov-Smirnov test with shape = 1 and 2.4  for QLD and NSW/VIC respectively, rainfall in both regions can be modelled as a gamma distribution.}
   \label{fig:hist_rain}
@@
 \end{equation}
-Therefore in both regions, land cover is ``trees" for the period before land cover change and ``removed" for the period after the change.
+Therefore in both regions, land cover is "trees" for the period before land cover change and "removed" for the period after the change.
 Here we simply assumed that vegetation cover change has occurred on every pixel.
@@
 \begin{equation}
-g(E(\mathbf{R}_r)) = \beta'_0 + s'_1(D_{1,r}\mathbf{SOI}) + s'_2(D_{2,r}\mathbf{IOD}) + s'_3\mathbf{Season} + \beta'_1\mathbf{Trend} + \beta'_2\mathbf{LC} + \boldsymbol{\epsilon'}_r
+g(E(\mathbf{R}_r)) = \beta'_0 + s'_1(\mathbf{SOI}) + s'_2(\mathbf{IOD}) + s'_3(\mathbf{Nino3.4}) + s'_4(\mathbf{Nino4}) + s'_5\mathbf{Season} + \beta'_1\mathbf{Trend} + \beta'_2\mathbf{LC} + \boldsymbol{\epsilon'}_r
   \label{eqn:model2}
@@
 
-%The valid time lengths are different in the two regions.
-In the QLD region, the exact timing of tree clearing is unknown.
-The information given in SLATS showed that clearings had occurred within 2003 - 2004.
-Hence in the analysis, the time series from 2003 to 2004 were discarded as the \textbf{LC} value for these two years cannot be determined.
-In the NSW/VIC region, severe bushfires were reported in January 2003.
-Storms and flooding were also reported immediately after the fires up to March in the Alpine Shire, Victoria, which could be an aftermath effect of the bushfires \citep{Tryhorn2008}.
-In this case, regression was performed on the time series excluding January - March 2003 to reduce the noise of storm events.
-The regression model was run starting from 1979 in both regions, by assuming the land covers before 2000 were ``trees".
-
 Vegetation cover changes occurred at different times in the two regions.
-In the QLD region, there no exact time can be assigned to the land clearing.
-Clearing has occurred between 2003 - 2004 according to SLATS reports.
-The information on the change in type of land cover during this time period is missing.
+In the QLD region, no exact time can be assigned to the land clearing.
+According to the SLATS reports, the most substantial clearing occurred between 2003 - 2004 .
+However, the information on the change in type of land cover during the time period is missing.
 Therefore, four scenarios were tested in the analysis.
-In these scenarios the ``after change" period started from: (1) June 2003, (2) January 2004, (3) June 2004 and (4) January 2005.
+In these scenarios the "after change" period started from: (1) June 2003, (2) January 2004, (3) June 2004 and (4) January 2005.
 In the NSW/VIC region, severe bushfires were reported in early January 2003.
-Hence the ``tree" cover state was up to December 2002 then it was changed to ``removed" state from January 2003.
-The regression model was run from 1979 for both regions.
-% Storms and flooding were also reported immediately after the fires up to March in the Alpine Shire, Victoria, which could be an aftermath effect of the bushfires \citep{Tryhorn2008}.
-In this case, two scenarios were tested, setting January 2003 and April 2003 as the beginning of the after-change period.
-
-% comment: missing values and gaps in data are complicating issues and require careful consideration \citep{Kundzewicz2004}
-%comment: Hence it was assumed that clearing was at the regional scale. The land surface cover in each individual pixel is homogeneous.
+Hence the "tree" cover state was up to December 2002 then it was changed to "removed" state from January 2003.
+However, as there is similar uncertainty about when the tree cover had recovered from the bushfire, several scenarios were again tested: 1) analysis up to 5 years post bushfire; 2) analysis up to 7 years post bush fire; and 3) analysis up to 9 years post bushfire.
+As a starting date, the regression model was run from 1979 for both regions.
 
@@
 
-setEPS()
-postscript("tc_simple_qld.eps")
-par(mar=c(5,6,1,1))
+jpeg("../figures/tc_simple.jpg", width=720, height=480)
+par(mar=c(5,6,1,1), mfrow=c(1,2))
 plot(data1$year,data1$tc,type="l",ylim=c(0.5,2.5),xlab="",ylab="",xaxt="n",yaxt="n")
@@
 axis(2,at=c(1,2),label=c("removed","tree"),cex.axis=1.3,las=2)
-dev.off()
-
-setEPS()
-postscript("tc_simple_nswvic.eps")
 par(mar=c(5,6,1,1))
@@
 ## predicting model: land cover is category variable
-rain.trend <- function(x,y,m,var.data) {
-
-  var.data$rain <- as.vector(unlist(y[x]))+0.0001
-
-  if (!is.na(mean(var.data$rain))) {
-    if (grepl("qld",substitute(y))) {
-      var.data <- na.omit(var.data)    
-      rain_fit <- gam(rain ~ s(month, k=3)
-                  + s(SOI, k=3)
+rain.trend <- function(x,y,var.data) {
+
+  var.data$rain <- as.vector(unlist(y[x]))[1:nrow(var.data)]+0.0001
+
+  #if (!is.na(mean(var.data$rain))) {
+#    if (grepl("qld",substitute(y))) {
+      #var.data <- na.omit(var.data)    
+      rain_fit <- gam(rain ~ s(month, k=3, bs="cs")
+                  + s(SOI, k=3, bs="cs")
+                  + s(IOD, k=3, bs="cs")
+                  #+ s(PDO, k=3, bs="cs")
+                  + s(NINO3.4, k=3, bs="cs")
+                  + s(NINO4, k=3, bs="cs")
                   + cover  
                   + trd
-                  , data=var.data,family=Gamma(link=log))
-    } else
-     if (grepl("nsw",substitute(y))) {  
-      var.data <- na.omit(var.data)    
-      rain_fit <- gam(rain ~ s(month, k=3)
-                  + s(SOI, k=3)
-                  + s(IOD, k=3)
-                  + cover  
-                  + trd
-                  , data=var.data,family=Gamma(link=log))
-    }
-
+                  , data=var.data,family=Gamma(link=log),
+                  na.action = na.omit)
+    # } else
+    #  if (grepl("nsw",substitute(y))) {  
+    #   var.data <- na.omit(var.data)    
+    #   rain_fit <- gam(rain ~ s(month, k=3)
+    #               + s(SOI, k=3)
+    #               + s(IOD, k=3)
+    #               + cover  
+    #               + trd
+    #               , data=var.data,family=Gamma(link=log))
+   # }        
+    #               
   return(rain_fit)
-  } else {NA}
 
@@
 ## generate variables for the model
-trd <- seq(1:360)
-month <-rep(1:12,times=30)
+trd <- seq(1:444)
+month <-rep(1:12,times=37)
 mth <- cycle(ts(month,start=c(1979,1),frequency=12))
 
-LC <- c(rep("tree",288),rep("clear",72))
-# LC_qld1 <- c(rep("tree",294),rep("clear",66)) ## after-change: from Jun 2003
-# LC_qld2 <- c(rep("tree",300),rep("clear",60)) ## after- : from Jan 2004
-# LC_qld3 <- c(rep("tree",306),rep("clear",54)) ## after- : from Jun 2004
-LC_qld4 <- c(rep("tree",312),rep("clear",48)) ## after- : from Jan 2005
-
-var.data <- data.frame(month, trd, SOI = SOI_30a, IOD = IOD_30a, cover = LC)
-# var.qld1 <- data.frame(month, trd, SOI = SOI_30a, IOD = IOD_30a, cover = LC_qld1)
-# var.qld2 <- data.frame(month, trd, SOI = SOI_30a, IOD = IOD_30a, cover = LC_qld2)
-# var.qld3 <- data.frame(month, trd, SOI = SOI_30a, IOD = IOD_30a, cover = LC_qld3)
-var.qld4 <- data.frame(month, trd, SOI = SOI_30a, IOD = IOD_30a, cover = LC_qld4)
+LC <- c(rep("tree",288),rep("clear",156))
+# different QLD scenarios
+LC_qld1 <- c(rep("tree",294),rep("clear",150)) ## after-change: from Jun 2003
+LC_qld2 <- c(rep("tree",300),rep("clear",144)) ## after- : from Jan 2004
+LC_qld3 <- c(rep("tree",306),rep("clear",138)) ## after- : from Jun 2004
+LC_qld4 <- c(rep("tree",312),rep("clear",132)) ## after- : from Jan 2005
+
+# NSW
+var.data <- data.frame(month, trd, SOI = SOI_30a$SOI,
+                       IOD = IOD_30a$IOD,
+                       PDO = PDO_100a$PDO[949:1392],
+                       NINO3.4 = SST34_30a$SST34,
+                       NINO4 = SST4_30a$SST4,
+                       cover = LC)
+# # 5 years
+# var.data5 <- var.data %>%
+#   filter(trd <= 348)
+
+# 7 years
+var.data7 <- var.data %>%
+  filter(trd <= 372)
+var.data8 <- var.data %>%
+  filter(trd <= 384)
+
+# # 9 years
+# var.data9 <- var.data %>%
+#   filter(trd <= 396)
+
+# QLD
+# test different scenarios
+# var.qld1 <- data.frame(month, trd, SOI = SOI_30a$SOI, 
+#                        IOD = IOD_30a$IOD, 
+#                        PDO = PDO_100a$PDO[949:1392],
+#                        NINO3.4 = SST34_30a$SST34,
+#                        NINO4 = SST4_30a$SST4,
+#                        cover = LC_qld1)
+# 
+var.qld2 <- data.frame(month, trd, SOI = SOI_30a$SOI,
+                       IOD = IOD_30a$IOD,
+                       PDO = PDO_100a$PDO[949:1392],
+                       NINO3.4 = SST34_30a$SST34,
+                       NINO4 = SST4_30a$SST4,
+                       cover = LC_qld2)
+# 
+# var.qld3 <- data.frame(month, trd, SOI = SOI_30a$SOI, 
+#                        IOD = IOD_30a$IOD, 
+#                        PDO = PDO_100a$PDO[949:1392],
+#                        NINO3.4 = SST34_30a$SST34,
+#                        NINO4 = SST4_30a$SST4,
+#                        cover = LC_qld3)
+
+var.qld4 <- data.frame(month, trd, SOI = SOI_30a$SOI, 
+                       IOD = IOD_30a$IOD, 
+                       PDO = PDO_100a$PDO[949:1392],
+                       NINO3.4 = SST34_30a$SST34,
+                       NINO4 = SST4_30a$SST4,
+                       cover = LC_qld4)
+# shorter qld series
+var.qld4_7y <- var.qld4 %>%
+  filter(trd <= 372)
+var.qld4_8y <- var.qld4 %>%
+  filter(trd <= 384)
+
+var.qld2_7y <- var.qld4 %>%
+  filter(trd <= 372)
+var.qld2_8y <- var.qld4 %>%
+  filter(trd <= 384)
 
 ## run model for 30 years
-#   coef.rain30.qld1 <- lapply(2:ncol(qld_30), function(x) rain.trend(x,qld_30,30,var.qld1))
-#   coef.rain30.qld2 <- lapply(2:ncol(qld_30), function(x) rain.trend(x,qld_30,30,var.qld2))
-#   coef.rain30.qld3 <- lapply(2:ncol(qld_30), function(x) rain.trend(x,qld_30,30,var.qld3))
-  coef.rain30.qld4 <- lapply(2:ncol(qld_30), function(x) rain.trend(x,qld_30,30,var.qld4))
-  coef.rain30.nswvic <- lapply(2:ncol(nsw_30), function(x) rain.trend(x,nsw_30,30,var.data))
-
-## extract sign of C
+   # coef.rain30.qld1 <- lapply(2:ncol(qld_30), 
+   #                            function(x) rain.trend(x,qld_30,var.qld1))
+coef.rain30.qld2 <- lapply(2:ncol(qld_30), 
+                            function(x) rain.trend(x,qld_30,var.qld2))
+   # coef.rain30.qld3 <- lapply(2:ncol(qld_30), 
+   #                            function(x) rain.trend(x,qld_30,var.qld3))
+coef.rain30.qld4 <- lapply(2:ncol(qld_30), 
+                             function(x) rain.trend(x,qld_30,var.qld4))
+# QUEENSLAND 7 yr series
+  coef.rain30.qld4_7y <- lapply(2:ncol(qld_30), 
+                             function(x) rain.trend(x,qld_30,var.qld4_7y))
+  coef.rain30.qld4_8y <- lapply(2:ncol(qld_30), 
+                             function(x) rain.trend(x,qld_30,var.qld4_8y))
+  # NSW/VIC
+  coef.rain30.nswvic <- lapply(2:ncol(nsw_30), 
+                               function(x) rain.trend(x,nsw_30,var.data))
+# # 5 years
+# coef.rain30.nswvic5 <- lapply(2:ncol(nsw_30), 
+#                               function(x) rain.trend(x,nsw_30,var.data5))
+# 7 years
+coef.rain30.nswvic7 <- lapply(2:ncol(nsw_30), 
+                              function(x) rain.trend(x,nsw_30,var.data7))
+# 8 years (includes 2010)
+coef.rain30.nswvic8 <- lapply(2:ncol(nsw_30), 
+                              function(x) rain.trend(x,nsw_30,var.data8))
+# # 9 years
+# coef.rain30.nswvic9 <- lapply(2:ncol(nsw_30), 
+#                               function(x) rain.trend(x,nsw_30,var.data9))
+
+## extract sign of C (the tree cover effect)
   C30_qld <- unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$p.coeff[2],NA)))
+# shorter series QLD
+  C30_qld_7y <- unlist(lapply(1:length(coef.rain30.qld4_7y),function(x) ifelse(summary(coef.rain30.qld4_7y)[x,2]=="gam",summary(coef.rain30.qld4_7y[[x]])$p.coeff[2],NA)))
+# NSW/VIC
 C30_nswvic <- unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$p.coeff[2],NA)))
-
-## extract p value of C
-#   Cp30_qld1 <- unlist(lapply(1:length(coef.rain30.qld1),function(x) ifelse(summary(coef.rain30.qld1)[x,2]=="gam",summary(coef.rain30.qld1[[x]])$p.pv[2],NA)))
-# Cp30_qld2 <- unlist(lapply(1:length(coef.rain30.qld2),function(x) ifelse(summary(coef.rain30.qld2)[x,2]=="gam",summary(coef.rain30.qld2[[x]])$p.pv[2],NA)))
-# Cp30_qld3 <- unlist(lapply(1:length(coef.rain30.qld3),function(x) ifelse(summary(coef.rain30.qld3)[x,2]=="gam",summary(coef.rain30.qld3[[x]])$p.pv[2],NA)))
-Cp30_qld4 <- unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$p.pv[2],NA)))
-Cp30_nswvic <- unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$p.pv[2],NA)))
+# NSW/VIC 7 year
+ C30_nswvic7 <- unlist(lapply(1:length(coef.rain30.nswvic7),function(x) ifelse(summary(coef.rain30.nswvic7)[x,2]=="gam",summary(coef.rain30.nswvic7[[x]])$p.coeff[2],NA)))
+
+# also for long term trend in nsw
+trd30_nswvic <- unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$p.coeff[3],NA)))
+hist(trd30_nswvic)
+# shows trds are all positive and small
+
+# C30_nswvic5 <- unlist(lapply(1:length(coef.rain30.nswvic5),function(x) ifelse(summary(coef.rain30.nswvic5)[x,2]=="gam",summary(coef.rain30.nswvic5[[x]])$p.coeff[2],NA)))
+# C30_nswvic9 <- unlist(lapply(1:length(coef.rain30.nswvic9),function(x) ifelse(summary(coef.rain30.nswvic9)[x,2]=="gam",summary(coef.rain30.nswvic9[[x]])$p.coeff[2],NA)))
+
+# p-value of C
+Cp30_qld4 <- unlist(lapply(1:length(coef.rain30.qld4),
+   function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",
+                      summary(coef.rain30.qld4[[x]])$p.pv[2],NA)))
+Cp30_qld2 <- unlist(lapply(1:length(coef.rain30.qld2),
+   function(x) ifelse(summary(coef.rain30.qld2)[x,2]=="gam",
+                      summary(coef.rain30.qld2[[x]])$p.pv[2],NA)))
+# Qld 7 year series
+Cp30_qld4_7y <- unlist(lapply(1:length(coef.rain30.qld4_7y),
+   function(x) ifelse(summary(coef.rain30.qld4_7y)[x,2]=="gam",
+                      summary(coef.rain30.qld4_7y[[x]])$p.pv[2],NA)))
+Cp30_qld4_8y <- unlist(lapply(1:length(coef.rain30.qld4_8y),
+   function(x) ifelse(summary(coef.rain30.qld4_8y)[x,2]=="gam",
+                      summary(coef.rain30.qld4_8y[[x]])$p.pv[2],NA)))
+
+Cp30_nswvic <- unlist(lapply(1:length(coef.rain30.nswvic),
+    function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",
+                       summary(coef.rain30.nswvic[[x]])$p.pv[2],
+                       NA)))
+# Cp30_nswvic5 <- unlist(lapply(1:length(coef.rain30.nswvic5),function(x) ifelse(summary(coef.rain30.nswvic5)[x,2]=="gam",summary(coef.rain30.nswvic5[[x]])$p.pv[2],NA)))
+Cp30_nswvic7 <- unlist(lapply(1:length(coef.rain30.nswvic7),function(x) ifelse(summary(coef.rain30.nswvic7)[x,2]=="gam",summary(coef.rain30.nswvic7[[x]])$p.pv[2],NA)))
+Cp30_nswvic8 <- unlist(lapply(1:length(coef.rain30.nswvic8),function(x) ifelse(summary(coef.rain30.nswvic8)[x,2]=="gam",summary(coef.rain30.nswvic8[[x]])$p.pv[2],NA)))
+# Cp30_nswvic9 <- unlist(lapply(1:length(coef.rain30.nswvic9),function(x) ifelse(summary(coef.rain30.nswvic9)[x,2]=="gam",summary(coef.rain30.nswvic9[[x]])$p.pv[2],NA)))
 
 ## p value of seasons and climate indices
-p_qld <- unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$s.pv[1],NA)))
-p_nsw <- unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$s.pv[1],NA)))
+p_qld <- do.call(rbind,
+                 lapply(1:length(coef.rain30.qld4), function(x) {
+                   ifelse(summary(coef.rain30.qld4)[x,2]=="gam",
+                      return(summary(coef.rain30.qld4[[x]])$s.pv),
+                                       return(rep(NA,6)))}))
+
+p_qld <- as.tibble(p_qld)
+colnames(p_qld) = c("month", "SOI", "IOD", "NINO3.4", "NINO4")
+p_qld$state <- "QLD"
+
+p_nsw <- do.call(rbind,
+                 lapply(1:length(coef.rain30.nswvic), function(x) {
+                ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",
+                  return(summary(coef.rain30.nswvic[[x]])$s.pv),
+                       return(rep(NA,6)))}))
+p_nsw <- as.tibble(p_nsw)
+colnames(p_nsw) = c("month", "SOI", "IOD", "NINO3.4", "NINO4")
+p_nsw$state <- "NSW/VIC"
+
+p_plot_df <- rbind(p_qld,p_nsw)
+
+jpeg("../figures/pvalues_fullmodel.jpg", height=960, width=960)
+p_plot_df %>%
+  gather(key="Variable", value="p_value", month:NINO4) %>%
+  ggplot(aes(p_value,fill=state)) + geom_histogram(alpha=0.5) +
+  facet_wrap(~Variable) +
+  scale_fill_manual(values = c("blue","red")) +
+  theme(axis.title = element_text(size = rel(2)),
+        axis.text = element_text(size=rel(1.5)),
+        strip.text = element_text(face="bold", size= rel(2)),
+        legend.text = element_text(face="bold", size= rel(1.5)),
+        legend.title = element_text(face="bold", size= rel(2)))
+dev.off()
+rm(p_plot_df)
+
+## extract p value of long term trend
+trdp30_qld4 <- unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$p.pv[3],NA)))
+trdp30_nswvic <- unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$p.pv[3],NA)))
+
+# show histogram
+plot_df <- data_frame(p_trd = c(trdp30_qld4,trdp30_nswvic),
+                      state = c(rep("QLD", length(trdp30_qld4)),
+                                rep("NSW/VIC", length(trdp30_nswvic))))
+
+jpeg("../figures/pvalue_trd.jpg", height=480, width=480)
+plot_df %>%
+  ggplot(aes(p_trd, fill=state)) + geom_histogram() +
+  scale_fill_manual(values = c("blue","red")) +
+  xlab("pvalue trend of the full gam model")
+dev.off()
 
 ## extract r.sq of the model
-rsq_qld <- unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$r.sq,NA)))
-rsq_nsw <- unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$r.sq,NA)))
+rsq_qld <- unlist(lapply(1:length(coef.rain30.qld4),
+          function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",
+                    summary(coef.rain30.qld4[[x]])$r.sq,NA)))
+rsq_nsw <- unlist(lapply(1:length(coef.rain30.nswvic),
+    function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",
+                    summary(coef.rain30.nswvic[[x]])$r.sq,NA)))
+
+# show histogram
+plot_df <- data_frame(`r.squared` = c(rsq_qld,rsq_nsw),
+                      state = c(rep("QLD", length(rsq_qld)),
+                                rep("NSW/VIC", length(rsq_nsw))))
+
+jpeg("../figures/rsq_fullmodel.jpg", height=480, width=480)
+plot_df %>%
+  ggplot(aes(`r.squared`, fill=state)) + geom_histogram() +
+  scale_fill_manual(values = c("blue","red")) +
+  xlab("r_squared of the full gam model")
+dev.off()
 
 ## frame of the two regions (100 years data)
-qld_tif <- raster("Data\\BoM\\rainfall\\AWAP\\from1900_FVG\\r19000101_qld.tif")
-nsw_tif <- raster("Data\\BoM\\rainfall\\AWAP\\from1900_FVG\\r19000101_nsw.tif")
+qld_tif <- raster("..\\Data\\BoM\\rainfall\\AWAP\\from1900_FVG\\r19000101_qld.tif")
+nsw_tif <- raster("..\\Data\\BoM\\rainfall\\AWAP\\from1900_FVG\\r19000101_nsw.tif")
 
@@
   dist_Cp30_qld4[] <- Cp30_qld4
-#   dist_Cp30_qld <- crop(reproject(dist_Cp100_qld,CRS=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84"),program="raster"),e1)
-  dist_Cp30_nsw <- raster(nsw_tif)
-  dist_Cp30_nsw[] <- Cp30_nswvic
-#   dist_Cp30_nsw <- crop(reproject(dist_Cp30_nsw,CRS=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84"),program="raster"),e2)
-# 
-#   dist_C30_qld <- raster(reg_qld)
-#   dist_C30_qld[] <- C30_qld
-#   dist_C30_nswvic <- raster(reg_nswvic)
-#   dist_C30_nswvic[] <- C30_nswvic
-# 
-# ## range of predicted values
-  predictmax_qld30 <- unlist(lapply(1:length(coef.rain30.qld4), function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam", max(coef.rain30.qld4[[x]]$fitted.values),NA)))
-  predictmin_qld30 <- unlist(lapply(1:length(coef.rain30.qld4), function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam", min(coef.rain30.qld4[[x]]$fitted.values),NA)))
-# 
-mean_qld <- lapply(1:length(coef.rain30.qld4), function(x) if(summary(coef.rain30.qld4)[x,2]=="gam") {fitted(coef.rain30.qld4[[x]])} else{NA})
-
-## calculate mean residual
-  cc <- 1
-  qld.sum <- mean_qld[[1]]
-  for (i in 2:length(mean_qld)) {
-    if(!is.na(mean_qld[[i]])) {
-      qld.sum <- qld.sum + mean_qld[[i]]
-      cc <- cc+1
-    }
-  }
-  qld.mean <- qld.sum/cc
-  qld.mean <- ts(qld.mean,start=c(1979,1),frequency=12)
-
-predictmax_nsw30 <- unlist(lapply(1:length(coef.rain30.nswvic), function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam", max(coef.rain30.nswvic[[x]]$fitted.values),NA)))
-# 70 - 254.7
-  predictmin_nsw30 <- unlist(lapply(1:length(coef.rain30.nswvic), function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam", min(coef.rain30.nswvic[[x]]$fitted.values),NA)))
-# 18.4 - 62.2
-
-mean_nswvic <- lapply(1:length(coef.rain30.nswvic), function(x) if(summary(coef.rain30.nswvic)[x,2]=="gam") {fitted(coef.rain30.nswvic[[x]])} else{NA})
-
-## calculate mean residual
-  cc <- 1
-  nswvic.sum <- mean_nswvic[[1]]
-  for (i in 2:length(mean_nswvic)) {
-    if(!is.na(mean_nswvic[[i]])) {
-      nswvic.sum <- nswvic.sum + mean_nswvic[[i]]
-      cc <- cc+1
-    }
-  }
-  nswvic.mean <- nswvic.sum/cc
-  nswvic.mean <- ts(nswvic.mean,start=c(1979,1),frequency=12)
-
-## plots
-  setEPS()
-  postscript("Cp_30yrs_qld.eps")
-  par(mar=c(3,3,1,6),xpd=T)
-  plot(e111,xlab="",ylab="")
+  dist_Cp30_qld4_7y <- raster(qld_tif)
+  dist_Cp30_qld4_7y[] <- Cp30_qld4_7y
+  dist_Cp30_qld4_8y <- raster(qld_tif)
+  dist_Cp30_qld4_8y[] <- Cp30_qld4_8y
+  dist_Cp30_qld2 <- raster(qld_tif)
+  dist_Cp30_qld2[] <- Cp30_qld2
+
+  jpeg("../figures/Cp_30yrs_qld.jpg", height=960, width=960)
+  par(mar=c(3,3,1,7),xpd=T)
+  plot(e111,xlab="",ylab="", cex.lab = 2, 
+       cex.axis=2)
   image(dist_Cp30_qld4,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
@@
   plot(e111,add=T)
-image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.7,legend.width=0.9,legend.args=list(text="p value",line=0.5,cex=1.2,font=1),axis.args=list(cex.axis=0.8))
+  image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
 legend(148.7,-27.9,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=1)
@@
 
-  setEPS()
-  postscript("Cp_30yrs_nswvic.eps")
-  par(mar=c(3,3,1,6),xpd=T)  
-  plot(e222,xlab="",ylab="") #,border="white")
+  jpeg("../figures/Cp_30yrs_qld_7y.jpg", height=960, width=960)
+  par(mar=c(3,3,1,7),xpd=T)
+  plot(e111,xlab="",ylab="", cex.lab = 2, 
+       cex.axis=2)
+  image(dist_Cp30_qld4_7y,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
+#   plot(aus_reg2,border="grey",add=T)
+  plot(mdb_reg2,border="grey",lwd=3,add=T)
+  plot(e111,add=T)
+  image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
+legend(148.7,-27.9,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=1)
+  dev.off()
+
+  jpeg("../figures/Cp_30yrs_qld_8y.jpg", height=960, width=960)
+  par(mar=c(3,3,1,7),xpd=T)
+  plot(e111,xlab="",ylab="", cex.lab = 2, 
+       cex.axis=2)
+  image(dist_Cp30_qld4_8y,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
+#   plot(aus_reg2,border="grey",add=T)
+  plot(mdb_reg2,border="grey",lwd=3,add=T)
+  plot(e111,add=T)
+  image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
+legend(148.7,-27.9,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=1)
+  dev.off()
+
+    #   dist_Cp30_qld <- crop(reproject(dist_Cp100_qld,CRS=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84"),program="raster"),e1)
+  dist_Cp30_nsw <- raster(nsw_tif)
+  dist_Cp30_nsw[] <- Cp30_nswvic
+  dist_Cp30_nsw_7y <- raster(nsw_tif)
+  dist_Cp30_nsw_7y[] <- Cp30_nswvic7
+  dist_Cp30_nsw_8y <- raster(nsw_tif)
+  dist_Cp30_nsw_8y[] <- Cp30_nswvic8
+#   dist_Cp30_nsw <- crop(reproject(dist_Cp30_nsw,CRS=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84"),program="raster"),e2)
+# 
+  jpeg("../figures/Cp_30yrs_nswvic.jpg", width=960, height=960)
+  par(mar=c(3,3,1,7),xpd=T)  
+  plot(e222,xlab="",ylab="", cex.lab = 2, 
+       cex.axis=2) #,border="white")
   image(dist_Cp30_nsw,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
-  plot(aus_reg2,border="grey",add=T)
-  plot(mdb_reg2,border="grey",lwd=3,add=T)
+  plot(aus_reg2,border="grey75",lwd=5,add=T)
+  plot(mdb_reg2,border="grey75",lwd=5,add=T)
   plot(e222,add=T)
-  #image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.8,legend.width=0.7,legend.args=list(text="p value",line=0.5,cex=1,font=2),axis.args=list(cex.axis=0.8))
+  # image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
   dev.off()
+
+    jpeg("../figures/Cp_30yrs_nswvic_7y.jpg", width=960, height=960)
+  par(mar=c(3,3,1,7),xpd=T)  
+  plot(e222,xlab="",ylab="", cex.lab = 2, 
+       cex.axis=2) #,border="white")
+  image(dist_Cp30_nsw_7y,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
+  plot(aus_reg2,border="grey75",lwd=5,add=T)
+  plot(mdb_reg2,border="grey75",lwd=5,add=T)
+  plot(e222,add=T)
+  # image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
+  dev.off()
+
+    jpeg("../figures/Cp_30yrs_nswvic_8y.jpg", width=960, height=960)
+  par(mar=c(3,3,1,7),xpd=T)  
+  plot(e222,xlab="",ylab="", cex.lab = 2, 
+       cex.axis=2) #,border="white")
+  image(dist_Cp30_nsw_8y,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
+  plot(aus_reg2,border="grey75",lwd=5,add=T)
+  plot(mdb_reg2,border="grey75",lwd=5,add=T)
+  plot(e222,add=T)
+  # image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
+  dev.off()
+# 
+# ## range of predicted values
+  # predictmax_qld30 <- unlist(lapply(1:length(coef.rain30.qld4), function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam", max(coef.rain30.qld4[[x]]$fitted.values),NA)))
+  # predictmin_qld30 <- unlist(lapply(1:length(coef.rain30.qld4), function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam", min(coef.rain30.qld4[[x]]$fitted.values),NA)))
+# 
+resid_qld <- lapply(1:length(coef.rain30.qld4), function(x) if(summary(coef.rain30.qld4)[x,2]=="gam") {residuals(coef.rain30.qld4[[x]])} else{NA})
+
+# ## calculate mean residual
+  test <- do.call(cbind,resid_qld)
+  qld.resid.mean <- apply(test,1, mean,na.rm=T)
+  residuals <- data_frame(Dates = seq.Date(as.Date("1979-01-01"),
+                                           length = length(qld.resid.mean),
+                                           by = "month"),
+                          QLD = qld.resid.mean)
+# 
+# predictmax_nsw30 <- unlist(lapply(1:length(coef.rain30.nswvic), function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam", max(coef.rain30.nswvic[[x]]$fitted.values),NA)))
+# # 70 - 254.7
+#   predictmin_nsw30 <- unlist(lapply(1:length(coef.rain30.nswvic), function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam", min(coef.rain30.nswvic[[x]]$fitted.values),NA)))
+# # 18.4 - 62.2
+# 
+resid_nswvic <- lapply(1:length(coef.rain30.nswvic),
+            function(x) if(summary(coef.rain30.nswvic)[x,2]=="gam")
+              {residuals(coef.rain30.nswvic[[x]])} else{NA})
+
+## calculate mean residual
+test <- do.call(cbind,resid_nswvic)
+nswvic.resid.mean <- apply(test,1, mean,na.rm=T)
+residuals$NSWVIC <- nswvic.resid.mean
+
+## plots
+jpeg("../figures/GAMresidualTime.jpeg", height=960, width=960)
+residuals %>%
+  gather(key="State", value="residual", QLD, NSWVIC) %>%
+  ggplot(aes(Dates,residual)) + geom_line(colour="blue") +
+  geom_vline(xintercept=as.numeric(as.Date("2003-01-01")), colour="red",
+             size=1.5) +
+  facet_wrap(~State)  +
+  theme(axis.text = element_text(size=rel(1.5)),
+        axis.title = element_text(size=rel(2)),
+        strip.text = element_text(size=rel(2), face = "bold"))
+dev.off()
+
 # 
@@
 ## plot gam.check
-  setEPS()
-  postscript("qld_check.eps")
+jpeg("../figures/gam_check.jpg", width=720, height=960)
+  par(mfrow=c(4,2))
   gam.check(coef.rain30.qld4[[1]])
+  gam.check(coef.rain30.nswvic[[1]])
   dev.off()
 
-  setEPS()
-  postscript("nswvic_check.eps")
-  gam.check(coef.rain30.nswvic[[which(values(dist_Cp30_nsw)<0.05)[6]]])
-  dev.off()
+  rm(coef.rain30.nswvic5,coef.rain30.nswvic7,coef.rain30.nswvic9,coef.rain30.nswvic11, coef.rain30.qld1, coef.rain30.qld2, coef.rain30.qld3, Cp30_nswvic8,Cp30_nswvic7,Cp30_qld4_7y,Cp30_qld4_8y, Cp30_qld4_7y,Cp30_qld4_8y, resid_nswvic,resid_qld)
 
@@
 
-A step trend test was used to detect changes in rainfall after the vegetation cover change.
-This nonparametric statistical test was modified from the Mann-Whitney Rank-Sum test by \citet{Hirsch1985}.
-The test was developed to look for a step change in data which is cross correlated.
-The gridded rainfall dataset was built by an interpolation method and as a result there is high spatial correlation between neighbouring pixels.
-Hence the step trend test is suitable for the analysis of such a dataset.
-The advantages of this test are: (1) it does not depend on assumptions of the data distribution; (2) it is not restricted to datasets with no missing data; (3) it is robust and not as easily influenced by outliers and negative numbers \citep{Hirsch1985}.
-
-\subsubsection{Rainfall residuals} %% do we need this heading?
-
-The rainfall residuals from the regression model in Equation~\ref{eqn:model} were used for this test.
-According to \citet{Hirsch1985}, using deseasonalised and detrended data is important in a test to detect step change.
-Furthermore, since rainfall is only partially attributed to local sources and conditions, noise can be introduced by large scale dynamics and changes in other climatic factors.
-The assumption is that the regression model should remove this noise.
-Hence the model residuals $\epsilon'_r$ excludes the effects due to climate drivers, seasonality and the long term trend.
-As a result local effects are assumed to explain at least part of the variation in the residuals.
-The  test, described in the following sub section, associates changes in rainfall trends with the tree cover changes.
-%The step change of tree cover were expected to be identified from the rainfall residuals. 
-%However deseasonalisation is unnecessary as the test was set up to count for the difference between months.
-Here the season term was not used in the model to reduce the risk of overfitting \citep{Wood2011}.
+To support the regression analysis, a step trend test to detect changes in rainfall as a result of vegetation cover change was used.
+This nonparametric statistical test was modified from the Mann-Whitney Rank-Sum test by \citet{Hirsch1985} and can identify a step change in data which is cross correlated.
+The gridded rainfall dataset has high spatial correlation between neighbouring pixels due to the underlying interpolation method.
+The advantages of using the Rank-Sum test are: (1) it does not depend on assumptions of the data distribution; (2) it is not restricted to datasets with no missing data; (3) it is robust and not as easily influenced by outliers and negative numbers \citep{Hirsch1985}.
+
+The rainfall residuals from the regression model in Equation~\ref{eqn:model} were used for the step trend test.
+According to \citet{Hirsch1985}, to detect a step change, using deseasonalised and detrended data is important.
+Furthermore, as rainfall can only be partially attributed to local sources and conditions, other effects are introduced by large scale dynamics and changes in other climatic factors.
+The assumption is that the regression model should remove these effects, and additionally deseasonalise and detrend the rainfall data.
+As a result, the local landuse effects are amplified in the variation of the remaining residuals.
+The  test, described in the following sub section, subsequently associates trends in the rainfall residuals with tree cover changes.
 
@@
 ## without land cover variable for step rend test
-rain.trend3 <- function(x,y,m,var.data) {
+rain.trend3 <- function(x,y,var.data) {
 
@@
 
-  if (!is.na(mean(var.data$rain))) {
-    if (grepl("qld",substitute(y))) {
-      rain_fit <- gam(rain ~ trd  # 
-                  + s(month,k=3)
-                  + s(SOI, k=3)
-                  , data=var.data,family=Gamma(link=log))
-    } else
-    if (grepl("nsw",substitute(y))) {
+  #if (!is.na(mean(var.data$rain))) {
+    # if (grepl("qld",substitute(y))) {
       rain_fit <- gam(rain ~ trd
-                  + s(month,k=3)
-                  + s(SOI, k=3)
-                  + s(IOD, k=3) 
-                  , data=var.data,family=Gamma(link=log))
-    }
+                      + s(month, k=3, bs="cs")# added month back in
+                  + s(SOI, k=3, bs="cs")
+                  + s(IOD, k=3, bs="cs")
+                  #+ s(PDO, k=3, bs="cs")
+                  + s(NINO3.4, k=3, bs="cs")
+                  + s(NINO4, k=3, bs="cs")
+                    , data=var.data,family=Gamma(link=log),
+                  na.action=na.omit)
+    # } else
+    # if (grepl("nsw",substitute(y))) {
+    #   rain_fit <- gam(rain ~ trd
+    #               + s(month,k=3)
+    #               + s(SOI, k=3)
+    #               + s(IOD, k=3) 
+    #               , data=var.data,family=Gamma(link=log))
+    #}
 
   return(rain_fit)
-  } else {NA}
+  #} #else {NA}
 
@@
 # 
-## run model for 30 years
-  coef.rain.qld <- lapply(2:ncol(qld_30), function(x) rain.trend3(x,qld_30,30,var.data))
-  coef.rain.nswvic <- lapply(2:ncol(nsw_30), function(x) rain.trend3(x,nsw_30,30,var.data))
-
-## explaining power of the partial model
+## run model for 37 years
+  coef.rain.qld <- lapply(2:ncol(qld_30), 
+                          function(x) rain.trend3(x,qld_30,var.qld4))
+  coef.rain.qld2 <- lapply(2:ncol(qld_30), 
+                          function(x) rain.trend3(x,qld_30,var.qld2))
+  # QLD shorter series
+  test <- qld_30[1:372,]
+  coef.rain.qld_7y <- lapply(2:ncol(qld_30), function(x) rain.trend3(x,test,var.qld4_7y))
+rm(test)  
 ## QLD
@@
 # 
-#   predictmax_qld <- unlist(lapply(1:length(coef.rain.qld), function(x) ifelse(summary(coef.rain.qld)[x,2]=="gam", max(coef.rain.qld[[x]]$fitted.values),NA)))
-#   predictmin_qld <- unlist(lapply(1:length(coef.rain.qld), function(x) ifelse(summary(coef.rain.qld)[x,2]=="gam", min(coef.rain.qld[[x]]$fitted.values),NA)))
 # 
@@
 
-# ## calculate mean residual
-#   cc <- 1
-#   qld.rsum <- residual_qld[[1]]
-#   for (i in 2:length(residual_qld)) {
-#     if(!is.na(residual_qld[[i]])) {
-#       qld.rsum <- qld.rsum + residual_qld[[i]]
-#       cc <- cc+1
-#     }
-#   }
-#   qld.rm <- qld.rsum/cc
-#   qld.rm <- ts(qld.rm,start=c(1979,1),frequency=12)
+  residual_qld2 <- lapply(1:length(coef.rain.qld2), function(x) if(summary(coef.rain.qld2)[x,2]=="gam") {coef.rain.qld2[[x]]$y-fitted(coef.rain.qld2[[x]])} else {NA})
+
+# shorter Qld series
+    residual_qld_7y <- lapply(1:length(coef.rain.qld_7y), function(x) if(summary(coef.rain.qld_7y)[x,2]=="gam") {coef.rain.qld_7y[[x]]$y-fitted(coef.rain.qld_7y[[x]])} else {NA})
+
 # 
@@
 #
+#NSW/VIC
+  coef.rain.nswvic <- lapply(2:ncol(nsw_30), function(x) rain.trend3(x,nsw_30,var.data))
+  test <- nsw_30[1:372,]
+  coef.rain.nswvic7 <- lapply(2:ncol(nsw_30), function(x) rain.trend3(x,test,var.data7))
+  rm(test)
+  # coef.rain.nswvic5 <- lapply(2:ncol(nsw_30), function(x) rain.trend3(x,nsw_30[1:348,],30,var.data5))
+  # coef.rain.nswvic9 <- lapply(2:ncol(nsw_30), function(x) rain.trend3(x,nsw_30[1:396,],30,var.data9))
+
+  ## explaining power of the partial model
   rsq_nswvic <- unlist(lapply(1:length(coef.rain.nswvic), function(x) ifelse(summary(coef.rain.nswvic)[x,2]=="gam", summary(coef.rain.nswvic[[x]])$r.sq,NA)))
 #
+
 #   predictmax_nswvic <- unlist(lapply(1:length(coef.rain.nswvic), function(x) ifelse(summary(coef.rain.nswvic)[x,2]=="gam", max(coef.rain.nswvic[[x]]$fitted.values),NA)))
@@
 residual_nswvic <- lapply(1:length(coef.rain.nswvic), function(x) if(summary(coef.rain.nswvic)[x,2]=="gam") {coef.rain.nswvic[[x]]$y-fitted(coef.rain.nswvic[[x]])} else {NA})
+# 7 year (2003 - 2010)
+residual_nswvic7 <- lapply(1:length(coef.rain.nswvic7), function(x) if(summary(coef.rain.nswvic7)[x,2]=="gam") {coef.rain.nswvic7[[x]]$y-fitted(coef.rain.nswvic7[[x]])} else {NA})
+
+# 5 year
+# residual_nswvic5 <- lapply(1:length(coef.rain.nswvic5), function(x) if(summary(coef.rain.nswvic5)[x,2]=="gam") {coef.rain.nswvic5[[x]]$y-fitted(coef.rain.nswvic5[[x]])} else {NA})
+
+# 9 year
+# residual_nswvic9 <- lapply(1:length(coef.rain.nswvic9), function(x) if(summary(coef.rain.nswvic9)[x,2]=="gam") {coef.rain.nswvic9[[x]]$y-fitted(coef.rain.nswvic9[[x]])} else {NA})
+
+# histogram of partial model rsq
+qld_df <-  data.frame(rsq = rsq_qld*100,
+                      state = "Qld")
+nsw_df <-  data.frame(rsq = rsq_nswvic*100,
+                      state = "NSW")
+
+plot_df <- rbind(qld_df, nsw_df)
+jpeg("../figures/pmodel_explHistogram.jpg")
+plot_df %>%
+  ggplot(aes(rsq)) + geom_histogram(fill="blue") + 
+  facet_wrap(~state) +
+  xlab(expression(paste("% of variability explained (",r^2,")"))) +
+  theme(axis.text = element_text(size=rel(1.5)),
+        strip.text = element_text(size=rel(2)),
+        axis.title = element_text(size=rel(2)))
+ dev.off()
 
@@
 ## explaining power of the partial model
-rain.trend4 <- function(x,y,m,var.data) {
+rain.trend4 <- function(x,y,var.data) {
 
@@
 
-  if (!is.na(mean(var.data$rain))) {
-    if (grepl("qld",substitute(y))) {
-      rain_fit <- gam(rain ~ s(SOI, k=3)
+  #if (!is.na(mean(var.data$rain))) {
+#    if (grepl("qld",substitute(y))) {
+      rain_fit <- gam(rain ~  s(SOI, k=3, bs="cs")
+                  + s(IOD, k=3, bs="cs")
+                  #+ s(PDO, k=3, bs="cs")
+                  + s(NINO3.4, k=3, bs="cs")
+                  + s(NINO4, k=3, bs="cs")
                   , data=var.data,family=Gamma(link=log))
-    } else
-    if (grepl("nsw",substitute(y))) {
-      rain_fit <- gam(rain ~ s(SOI, k=3)
-                  + s(IOD, k=3) 
-                  , data=var.data,family=Gamma(link=log))
-    }
-
+    # } else
+    # if (grepl("nsw",substitute(y))) {
+    #   rain_fit <- gam(rain ~ s(SOI, k=3)
+    #               + s(IOD, k=3) 
+    #               , data=var.data,family=Gamma(link=log))
+    # }        
+    #               
   return(rain_fit)
-  } else {NA}
+  #} #else {NA}
 
@@
 ## explaining power of index
-  coef.soi.qld <- lapply(2:ncol(qld_30), function(x) rain.trend4(x,qld_30,30,var.data))
-  coef.soi.nswvic <- lapply(2:ncol(nsw_30), function(x) rain.trend4(x,nsw_30,30,var.data))
+  coef.soi.qld <- lapply(2:ncol(qld_30), function(x) rain.trend4(x,qld_30,var.qld4))
+  coef.soi.nswvic <- lapply(2:ncol(nsw_30), function(x) rain.trend4(x,nsw_30,var.data))
   rsq_soi_qld <- unlist(lapply(1:length(coef.soi.qld), function(x) ifelse(summary(coef.soi.qld)[x,2]=="gam", summary(coef.soi.qld[[x]])$r.sq,NA)))
@@
 
-setEPS()
-postscript("soi_qld.eps")
- hist(rsq_soi_qld*100,main="",xlab=expression(paste("% of variability explained (",R^2,")")),cex.lab=1.2)
+qld_df <-  data.frame(rsq = rsq_soi_qld*100,
+                      state = "Qld")
+nsw_df <-  data.frame(rsq = rsq_soi_nswvic*100,
+                      state = "NSW")
+
+plot_df <- rbind(qld_df, nsw_df)
+jpeg("../figures/soi_explHistogram.jpg", width=960, height=960)
+plot_df %>%
+  ggplot(aes(rsq)) + geom_histogram(fill="blue") + 
+  facet_wrap(~state) +
+  xlab(expression(paste("% of variability explained (",r^2,")"))) +
+  theme(axis.text = element_text(size=rel(1.5)),
+        strip.text = element_text(size=rel(2)),
+        axis.title = element_text(size=rel(2)))
  dev.off()
 
- setEPS()
-postscript("soi_nswvic.eps")
- hist(rsq_soi_nswvic*100,main="",xlab=expression(paste("% of variability explained (",R^2,")")),cex.lab=1.2)
- dev.off()
-
 @
@@
 
-The step trend test is a modified version of the Mann-Whitney rank-sum statistic \citep{Hirsch1985}.
+As indicated, the step trend test is a modified version of the Mann-Whitney rank-sum statistic \citep{Hirsch1985}.
 As a nonparametric rank-based test, the Mann-Whitney test does not use the exact values of rainfall but depends on the ranks of the data.
@@
 
-The before period and the after period formed two groups of samples.
+The before and after period in the data formed two groups of samples.
 The split point of the two periods was based on the timing of the vegetation cover changes.
 In the QLD region, changes occurred anytime during 2003 and 2004.
-In contrast to the previous method, the time period of where the land cover chang occurred was excluded, as the nonparametric test allows missing data.
+In contrast to the previous method, the time period covering the land cover change  was excluded, as the nonparametric test allows missing data.
 \citet{Hirsch1985} also pointed out that the power of the test is higher if the data of the change period is ignored.
 Hence 2003 and 2004 were excluded from the analysis.
-As a result, the after-change period was 2005 - 2007.
+As a result, the after-change period was 2005 - 2015 for the Queensland location.
 
@@
 The change was within a relatively short period of the year.
-Therefore the after-change period in this region still started in January 2003.
+Therefore the after change period in this region still started in January 2003.
 Following \citet{Hirsch1985}, the before period was set to five years (1998 - 2002) in both regions.
-%The test does not require a long time series before and after the intervention as shown in \citet{Hirsch1985}.
+The length of the after period is difficult for the NSW/VIC as the regrowth would at some point have impacted the local effects.
+Therefore we tested a short period 2003 - 2010 and a longer period 2003 - 2015.
 
@@
 Instead of completing the Mann-Whitney U-test, \citet{Hirsch1985} applied the rank-sum statistics in a standard normal Z test.
-The modified test can be used to detect step change and it accounts for serial and cross correlation in the data.
+As highlighted, this modified test accounts for serial and cross correlation in the data.
 In the case here, the deseasonalised and detrended data shows little autocorrelation in the time series but possesses strong cross correlation between neighbouring pixels, i.e.
 $R>0.99$.
-Hence the covariance between pixels would need to be considered in the test.
-
-% Assuming the distribution of $W_{jk}$ is normal and independent,
 
@@
 Here $r$ is calculated on the full time series in each pixel.
-In this analysis, the test was applied to a square block of four pixels each time.
+In the analysis, the test was applied to a square block of four pixels each time.
 As argued by \citet{Hirsch1985}, $ns=4$ is the most optimal solution to balance the cost and the gain in the test power.
-% \begin{equation}
-%   \sigma^2_{w12} = 12\cdot\sigma^2_W = n_1\cdot n_2(n_1+n_2+1).
-% \end{equation}
 
@@
 The above statistic is written for a 12 month period. By changing the value 12, it can also be used to test seasonal rainfall change or for other customized periods.
-
-%The Z' test is based on the assumption that the sum of $W_j$ is normally distributed.
-Small numbers of $n_1$ and $n_2$ might weaken this assumption.
-On the other hand, because there is a much longer period before the intervention in the rainfall data, the possible changes at some other time could increase the noise in the analysis.
-Given the rainfall time series data was only available up to 2008 and the specified land intervention was closer to the end of this time series, a subset of the rainfall time series was selected from 1998 to 2008.
-This 11-year rainfall data analysis was compared to the time series analysis for the full period.
 
@@
 
-## perform step test on the data
+## perform step test on the data using ns = 4
 ## function to calculate the step trend statistic (Hirsch, 1985)
-stepTrend2 <- function(x,y,i){
-
+stepTrend2 <- function(x,y,i, endyear=2015){  
+#browser()
     n1 <- 2002-1997
@@
 
-      z <- 56  # no. of column
+      z <- 56  # no. of column, related to the dim of the region
+      # Qld 57 by 51
+      # NSW 38 by 33
 
@@
     if (i == 1) {
-      n2 <- 2008-2004    
-      v.1 <- ts(y.1[c(229:288,313:360)],start=c(1998,1),frequency=12)
-      v.2 <- ts(y.2[c(229:288,313:360)],start=c(1998,1),frequency=12)
-      v.3 <- ts(y.3[c(229:288,313:360)],start=c(1998,1),frequency=12)
-      v.4 <- ts(y.4[c(229:288,313:360)],start=c(1998,1),frequency=12)
+      n2 <- endyear-2004    
+      v.1 <- ts(y.1[c(229:288,313:length(y.1))],
+                start=c(1997,1),frequency=12)
+      v.2 <- ts(y.2[c(229:288,313:length(y.1))],
+                start=c(1997,1),frequency=12)
+      v.3 <- ts(y.3[c(229:288,313:length(y.1))],
+                start=c(1997,1),frequency=12)
+      v.4 <- ts(y.4[c(229:288,313:length(y.1))],
+                start=c(1997,1),frequency=12)
 #       v.5 <- ts(y.5[c(229:288,313:348)],start=c(1998,1),frequency=12)
@@
 
-      n2 <- 2008-2002
-
-      v.1 <- ts(y.1[c(229:360)],start=c(1998,1),frequency=12)
-      v.2 <- ts(y.2[c(229:360)],start=c(1998,1),frequency=12)
-      v.3 <- ts(y.3[c(229:360)],start=c(1998,1),frequency=12)
-      v.4 <- ts(y.4[c(229:360)],start=c(1998,1),frequency=12)
+      n2 <- endyear-2002
+
+      v.1 <- ts(y.1[c(229:length(y.1))],
+                start=c(1997,1),frequency=12)
+      v.2 <- ts(y.2[c(229:length(y.2))],
+                start=c(1997,1),frequency=12)
+      v.3 <- ts(y.3[c(229:length(y.3))],
+                start=c(1997,1),frequency=12)
+      v.4 <- ts(y.4[c(229:length(y.4))],
+                start=c(1997,1),frequency=12)
 #       v.5 <- ts(y.5[c(229:348)],start=c(1998,1),frequency=12)
@@
 ## central QLD
-  qld.z.new <- unlist(lapply(1:(length(residual_qld)-56),function(x) stepTrend2(x,residual_qld,1)))
-  qld.z.new <- c(qld.z.new,rep(NA,56))
+  qld.z.new <- unlist(lapply(1:(length(residual_qld)-57),function(x) stepTrend2(x,residual_qld,1)))
+  qld.z.new <- c(qld.z.new,rep(NA,57))
+# shorter series
+    qld.z.new_7y <- unlist(lapply(1:(length(residual_qld_7y)-57),function(x) stepTrend2(x,residual_qld_7y,1, endyear=2009)))
+  qld.z.new_7y <- c(qld.z.new_7y,rep(NA,57))
 
@@
   tr_qld.step.new[] <- qld.z.new
+
+  tr_qld.step.new <- raster(qld_tif)
+  tr_qld.step.new[] <- qld.z.new_7y
+
 ## NSW  
-  nswvic.z.new <- unlist(lapply(1:(length(residual_nswvic)-37),function(x) stepTrend2(x,residual_nswvic,2)))
-  nswvic.z.new <- c(nswvic.z.new,rep(NA,37))
+  nswvic.z.new <- unlist(lapply(1:(length(residual_nswvic)-38),function(x) stepTrend2(x,residual_nswvic,2)))
+  nswvic.z.new <- c(nswvic.z.new,rep(NA,38))
+
+  nswvic.z.new7 <- unlist(lapply(1:(length(residual_nswvic7)-38),function(x) stepTrend2(x,residual_nswvic7,2, endyear = 2009)))
+  nswvic.z.new7 <- c(nswvic.z.new7,rep(NA,38))
 
   tr_nswvic.step.new <- raster(nsw_tif)
-  tr_nswvic.step.new[] <- nswvic.z.new
+  tr_nswvic.step.new[] <- nswvic.z.new7
 
@@
 
-setEPS()
-postscript("step_new_qld.eps")
+tiff("../figures/step_new_qld_7y.tif", width=960, height=960)
   par(mar=c(3,3,1,7),xpd=T)
@@
 
-setEPS()
-  postscript("step_new_nswvic.eps")
+tiff("../figures/step_new_nswvic_7y.tif", width=960, height=960) 
 par(mar=c(3,3,1,7),xpd=T)
@@
 
-qld_per_origp <- length(subset(qld.z.new,qld.z.new>=1.645))/length(qld.z.new)
-# 0.0125
-qld_per_orign <- length(subset(qld.z.new,qld.z.new<=-1.645))/length(qld.z.new)
-# 0
-# 
-nsw_per_origp <- length(subset(nswvic.z.new,nswvic.z.new>=1.645))/length(nswvic.z.new)
-# 0.015
-nsw_per_orign <- length(subset(nswvic.z.new,nswvic.z.new<=-1.645))/length(nswvic.z.new)
+# qld_per_origp <- length(subset(qld.z.new,qld.z.new>=1.645))/length(qld.z.new)
+# # 0.0125
+# qld_per_orign <- length(subset(qld.z.new,qld.z.new<=-1.645))/length(qld.z.new)
+# # 0
+# # 
+# nsw_per_origp <- length(subset(nswvic.z.new,nswvic.z.new>=1.645))/length(nswvic.z.new)
+# # 0.015
+# nsw_per_orign <- length(subset(nswvic.z.new,nswvic.z.new<=-1.645))/length(nswvic.z.new)
 # 0
@@
 ## perform step test on the data for the bootstrapping test (spatial structure was kept)
-## function to calculate the step trend statistic (Hirsch, 1985)
-stepTrend <- function(x,i){
-
+## function to calculate the step trend statistic (Hirsch, 1985), ns = 1 
+stepTrend <- function(x,i, endyear=2015){  
+#browser()
     n1 <- 2002-1997
@@
     if(i==1) {
-      n2 <- 2008-2004    
-      v <- x[c(229:288,313:360)]
+      n2 <- endyear-2004
+
+      v <- x[c(229:288,313:(n2*12+313-1))]
 #       v <- (x$y-x$fitted.values)[c(1:288,313:357)]
@@
       if(i==2) {
-      n2 <- 2008-2002
-      v <- x[c(229:360)]
+      n2 <- endyear-2002
+      v <- x[c(229:(n2*12+n1*12+229-1))]
 #       v <- (x$y-x$fitted.values)[c(1:288, 301:357)]
@@
 ## central QLD
-  qld.z <- unlist(lapply(residual_qld,function(x) stepTrend(x,1)))
+  Qld.z <- unlist(lapply(residual_qld,function(x) stepTrend(x,1)))
+  Qld.z_7y <- unlist(lapply(residual_qld_7y,
+                            function(x) stepTrend(x,1, endyear=2010)))
 ## NSW/VIC
 ## treatment 1: excluding 2003
-  nswvic.z <- unlist(lapply(residual_nswvic,function(x) stepTrend(x,2)))
-
-  qld_origp <- length(subset(qld.z,qld.z>=1.645))/length(qld.z)
-# 0.01
-  qld_orign <- length(subset(qld.z,qld.z<=-1.645))/length(qld.z)
+  nswvic.z <- unlist(lapply(residual_nswvic,
+                            function(x) stepTrend(x,2)))
+  nswvic.z_7y <- unlist(lapply(residual_nswvic7,
+                            function(x) stepTrend(x,2,
+                                                  endyear = 2010)))
+
+# ns = 1 results
+  Qld_origp <- length(subset(Qld.z,Qld.z>=1.645))/length(Qld.z)
+# 0.002
+  Qld_orign <- length(subset(Qld.z,Qld.z<=-1.645))/length(Qld.z)
 # 0
-
-  nsw_origp <- length(subset(nswvic.z,nswvic.z>=1.645))/length(nswvic.z)
-# 0.022
+  Qld_origp_7y <- length(subset(Qld.z_7y,Qld.z_7y>=1.645))/length(Qld.z_7y)
+# 0
+  Qld_orign_7y <- length(subset(Qld.z_7y,Qld.z_7y<=-1.645))/length(Qld.z)
+# 0.023
+
+  nsw_origp <- length(subset(nswvic.z,nswvic.z>=1.645))/length(nswv
+# 0
   nsw_orign <- length(subset(nswvic.z,nswvic.z<=-1.645))/length(nswvic.z)
-
-# new step trend test 2: keep spatial structure
-NstepTrend2 <- function(x,y,ii){
+# 0
+  nsw_origp_7y <- length(subset(nswvic.z_7y,nswvic.z_7y>=1.645))/length(nswvic.z_7y)
+# 0
+  nsw_orign_7y <- length(subset(nswvic.z_7y,nswvic.z_7y<=-1.645))/length(nswvic.z_7y)
+# 0
+
+#ns = 4 results
+# positive
+  qld_origp.new <- length(subset(qld.z.new,qld.z.new>=1.645))/length(qld.z.new)
+# 0.00034
+# negative
+  qld_orign.new <- length(subset(qld.z.new,qld.z.new<=-1.645))/length(qld.z.new)
+# 0
+# shorter series
+  qld_origp.new_7y <- length(subset(qld.z.new_7y,qld.z.new_7y>=1.645))/length(qld.z.new_7y)
+# 0
+  qld_orign.new_7y <- length(subset(qld.z.new_7y,qld.z.new_7y<=-1.645))/length(qld.z.new_7y)
+# 0.02
+
+  nsw_origp.z.new <- length(subset(nswvic.z.new,nswvic.z.new>=1.645))/length(nswvic.z.new)
+# 0
+  nsw_orign.z.new <- length(subset(nswvic.z.new,nswvic.z.new<=-1.645))/length(nswvic.z.new)
+# 0
+  nsw_origp.z.new7 <- length(subset(nswvic.z.new7,nswvic.z.new7>=1.645))/length(nswvic.z.new7)
+# 0
+  nsw_orign.z.new7 <- length(subset(nswvic.z.new7,nswvic.z.new7<=-1.645))/length(nswvic.z.new7)
+# 0
+
+# new step trend test 2: keep spatial structure, but ns = 1
+NstepTrend2 <- function(x,y,ii, endyear=2015){
 
@@
   if(ii==1) {
-      n2 <- 2008-2004    
-      v <- x[c(229:288,313:360)]    
+      n2 <- endyear-2004    
+      v <- x[c(229:288,313:length(x))]    
     } else
       if(ii==2) {
-      n2 <- 2008-2002
-      v <- x[c(229:360)]
+      n2 <- endyear-2002
+      v <- x[c(229:length(x))]
     }
@@
   yy<-ts(v,start=c(1998,1),frequency=12)
-
+#browser()  
   for(i in 1:12) { 
@@
 R <- 1000
+
+## QLD
 z2pq.values = numeric(R)
@@
 # mth.l <- rep(11,12)
-mth.l <- rep(9,12)
+# this relates to the total number of years:n1 + n2
+# 
+period1 <- 2002 - 1997
+period2 <- 2015 - 2004
+mth.l <- rep((period1 + period2),12)
 
@@
 
-  temp <- unlist(lapply(residual_qld,function(x,y) NstepTrend2(x,y,1),new.order))
+  temp <- unlist(lapply(residual_qld,
+                        function(x,y) NstepTrend2(x,y,1),new.order))
   if(i==1) {write.table(temp,file="bt_zvalues2_qld.txt",row.names=F)
@@
 
+## NSW
+period1 <- 2002 - 1997
+period2 <- 2015 - 2002
+
+mth.l <- rep((period1 + period2),12)
+
+ptm <- proc.time()
+for (i in 1:R) {
+
+  for (j in 1:12) {
+    new.order[[j]]<-sample(1:mth.l[j])
+  }
+
+  temp <- unlist(lapply(residual_nswvic,
+                        function(x,y) NstepTrend2(x,y,2),new.order))
+  if(i==1) {write.table(temp,file="bt_zvalues2_nswvic.txt",
+                        row.names=F)
+            } else
+            {write.table(temp,file="bt_zvalues2_nswvic.txt",
+                           append=T,row.names=F,col.names=F)}
+  z2pq.values[i] <- length(subset(temp,temp>=1.645))/length(temp)
+  z2nq.values[i] <- length(subset(temp,temp<=-1.645))/length(temp)
+}
+proc.time()-ptm
+
+write.table(z2pq.values,file="bt_out2p_nswvic.txt",row.names=F)
+write.table(z2nq.values,file="bt_out2n_nswvic.txt",row.names=F)
+
 # zp.values <- unlist(read.table("B:/My Study\\AVSproject\\WrittenWork\\PhD_thesis\\chap3\\bt_outp.txt",header=T,sep=""))*100
@@
 # 
-bt1_nswp.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_out2p_nsw.txt",header=T,sep=""))*100
-bt1_nswn.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_out2n_nsw.txt",header=T,sep=""))*100
-
-bt1_qldp.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_out2p_qld.txt",header=T,sep=""))*100
-bt1_qldn.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_out2n_qld.txt",header=T,sep=""))*100
+bt1_nswp.values <- unlist(read.table("bt_out2p_nswvic.txt",
+                                     header=T,sep=""))*100
+bt1_nswn.values <- unlist(read.table("bt_out2n_nswvic.txt",
+                                     header=T,sep=""))*100
+
+bt1_qldp.values <- unlist(read.table("bt_out2p_qld.txt",
+                                     header=T,sep=""))*100
+bt1_qldn.values <- unlist(read.table("bt_out2n_qld.txt",
+                                     header=T,sep=""))*100
+#
+# setEPS()
+# postscript("boot1.eps")
+# hist(zp.values,xlim=c(0,9),xlab="% positive z",main="no spatial dependency")
+# points(per_origp*100,0,pch=16)
+# dev.off()
+#
+# setEPS()
+# postscript("boot2.eps")
+# hist(zn.values,xlim=c(0,25),xlab="% negative z",main="no spatial dependency")
+# points(per_orign*100,0,pch=16)
+# dev.off()
 # #
-# # setEPS()
-# # postscript("boot1.eps")
-# # hist(zp.values,xlim=c(0,9),xlab="% positive z",main="no spatial dependency")
-# # points(per_origp*100,0,pch=16)
-# # dev.off()
-# # 
-# # setEPS()
-# # postscript("boot2.eps")
-# # hist(zn.values,xlim=c(0,25),xlab="% negative z",main="no spatial dependency")
-# # points(per_orign*100,0,pch=16)
-# # dev.off()
-# # # 
-# setEPS()
-# postscript("bt1_qldp.eps")
-# hist(bt1_qldp.values,nclass=100,xlab="% positive Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
-# points(qld_origp*100,0,pch=16,col="red",cex=1.8)
-# dev.off()
-# 
-# setEPS()
-# postscript("bt1_qldn.eps")
-# hist(bt1_qldn.values,nclass=100,xlab="% negative Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
-# points(qld_orign*100,0,pch=16,col="red",cex=1.8)
-# dev.off()
-# 
-# setEPS()
-# postscript("bt1_nswp.eps")
-# hist(bt1_nswp.values,nclass=100,xlab="% positive Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
-# points(nsw_origp*100,0,pch=16,col="red",cex=1.8)
-# dev.off()
-# 
-# setEPS()
-# postscript("bt1_nswn.eps")
-# hist(bt1_nswn.values,nclass=100,xlab="% negative Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
-# points(nsw_orign*100,0,pch=16,col="red",cex=1.8)
-# dev.off()
-# 
-# ## multiple cells
-# NstepTrend <- function(x,y,m,i){  
-# 
-#     n1 <- 2002-1997
-#     
-#     if (i==1){
-#       
-#       z <- 56  # no. of column
-#       
-#     } else {
-#     
-#     z <- 37 }
-#     
-# if (x%%z!=0){ #& x%%z!=(z-1)) {
-#     y.1 <- m[[x]]
-#     y.2 <- m[[x+1]]
-# #     y.3 <- y[[x+2]]
-#     y.3 <- m[[x+z]]
-#     y.4 <- m[[x+z+1]]
-# #     y.6 <- y[[x+z+2]]
-# #     y.7 <- y[[x+2*z]]
-# #     y.8 <- y[[x+2*z+1]]
-# #     y.9 <- y[[x+2*z+2]]
-#     
-#     if(!is.na(y.1[1])&!is.na(y.2[1])&!is.na(y.3[1])&!is.na(y.4[1])) { #|!is.na(y.5[1])|!is.na(y.6[1])|!is.na(y.7[1])|!is.na(y.8[1])|!is.na(y.9[1])) {
-#         
-#     if (i == 1) {
-#       n2 <- 2008-2004    
-#       v.1 <- ts(y.1[c(229:288,313:360)],start=c(1998,1),frequency=12)
-#       v.2 <- ts(y.2[c(229:288,313:360)],start=c(1998,1),frequency=12)
-#       v.3 <- ts(y.3[c(229:288,313:360)],start=c(1998,1),frequency=12)
-#       v.4 <- ts(y.4[c(229:288,313:360)],start=c(1998,1),frequency=12)
-# #       v.5 <- ts(y.5[c(229:288,313:348)],start=c(1998,1),frequency=12)
-# #       v.6 <- ts(y.6[c(229:288,313:348)],start=c(1998,1),frequency=12)
-# #       v.7 <- ts(y.7[c(229:288,313:348)],start=c(1998,1),frequency=12)
-# #       v.8 <- ts(y.8[c(229:288,313:348)],start=c(1998,1),frequency=12)
-# #       v.9 <- ts(y.9[c(229:288,313:348)],start=c(1998,1),frequency=12)
-# #       v <- (x$y-x$fitted.values)[c(1:288,313:357)]
-# #       v.s <- ts(v,start=c(1998,1),frequency=12)
-#       
-#     } else {
-#     
-#       n2 <- 2008-2002
-#   
-#       v.1 <- ts(y.1[c(229:360)],start=c(1998,1),frequency=12)
-#       v.2 <- ts(y.2[c(229:360)],start=c(1998,1),frequency=12)
-#       v.3 <- ts(y.3[c(229:360)],start=c(1998,1),frequency=12)
-#       v.4 <- ts(y.4[c(229:360)],start=c(1998,1),frequency=12)
-# #       v.5 <- ts(y.5[c(229:348)],start=c(1998,1),frequency=12)
-# #       v.6 <- ts(y.6[c(229:348)],start=c(1998,1),frequency=12)
-# #       v.7 <- ts(y.7[c(229:348)],start=c(1998,1),frequency=12)
-# #       v.8 <- ts(y.8[c(229:348)],start=c(1998,1),frequency=12)
-# #       v.9 <- ts(y.9[c(229:348)],start=c(1998,1),frequency=12)      
-# #       v <- (x$y-x$fitted.values)[c(1:288, 301:357)]
-# #       v.s <- ts(v,start=c(1998,1),frequency=12)    
-# }
-#   
-# ## rank by month
-#    for (j in 1:12) {  
-#     y1.mth <- data.frame(mth=as.vector(window(v.1,start=c(1998,i),frequency=1)),o=y[[j]])[order(y[[j]]),]
-#     y2.mth <- data.frame(mth=as.vector(window(v.2,start=c(1998,i),frequency=1)),o=y[[j]])[order(y[[j]]),]
-#     y3.mth <- data.frame(mth=as.vector(window(v.3,start=c(1998,i),frequency=1)),o=y[[j]])[order(y[[j]]),]
-#     y4.mth <- data.frame(mth=as.vector(window(v.4,start=c(1998,i),frequency=1)),o=y[[j]])[order(y[[j]]),]
-#     
-#     assign(paste("r1",j,sep=""),rank(y1.mth$mth))
-#     assign(paste("r2",j,sep=""),rank(y2.mth$mth))
-#     assign(paste("r3",j,sep=""),rank(y3.mth$mth))
-#     assign(paste("r4",j,sep=""),rank(y4.mth$mth)) 
-# #     assign(paste("r5",j,sep=""),rank(subset(v.5,cycle(v.5)==j)))
-# #     assign(paste("r6",j,sep=""),rank(subset(v.6,cycle(v.6)==j)))
-# #     assign(paste("r7",j,sep=""),rank(subset(v.7,cycle(v.7)==j)))
-# #     assign(paste("r8",j,sep=""),rank(subset(v.8,cycle(v.8)==j)))
-# #     assign(paste("r9",j,sep=""),rank(subset(v.9,cycle(v.9)==j)))
-#     
-#     }
-#   
-# ## mann-whitney rank sum statistic
-#     rlist.1 <- list(r11,r12,r13,r14,r15,r16,r17,r18,r19,r110,r111,r112) ## full year
-#     rlist.2 <- list(r21,r22,r23,r24,r25,r26,r27,r28,r29,r210,r211,r212)
-#     rlist.3 <- list(r31,r32,r33,r34,r35,r36,r37,r38,r39,r310,r311,r312)
-#     rlist.4 <- list(r41,r42,r43,r44,r45,r46,r47,r48,r49,r410,r411,r412)
-# #     rlist.5 <- list(r51,r52,r53,r54,r55,r56,r57,r58,r59,r510,r511,r512)
-# #     rlist.6 <- list(r61,r62,r63,r64,r65,r66,r67,r68,r69,r610,r611,r612)    
-# #     rlist.7 <- list(r71,r72,r73,r74,r75,r76,r77,r78,r79,r710,r711,r712)    
-# #     rlist.8 <- list(r81,r82,r83,r84,r85,r86,r87,r88,r89,r810,r811,r812)
-# #     rlist.9 <- list(r91,r92,r93,r94,r95,r96,r97,r98,r99,r910,r911,r912)    
-#     rlist <- list(unlist(rlist.1),unlist(rlist.2),unlist(rlist.3),unlist(rlist.4)) #,unlist(rlist.5),unlist(rlist.6),unlist(rlist.7),unlist(rlist.8),unlist(rlist.9))
-# ## seasonal
-# #     rlist <- list(unlist(rlist.1[c(9:11)]),unlist(rlist.2[c(9:11)]),unlist(rlist.3[c(9:11)]),unlist(rlist.4[c(9:11)]))
-# #     rlist <- list(r1,r2,r12) ## summer
-# #     rlist <- list(r3,r4,r5)  ## autumn
-# #     rlist <- list(r6,r7,r8)  ## winter
-# #     rlist <- list(r9,r10,r11) ## spring
-#     w.s1 <- 0
-#     w.s2 <- 0
-#     w.s3 <- 0
-#     w.s4 <- 0
-# #     w.s5 <- 0
-# #     w.s6 <- 0
-# #     w.s7 <- 0
-# #     w.s8 <- 0
-# #     w.s9 <- 0
-#   for (k in 1:length(rlist.1)) {
-# #   for (k in c(9:11)) {
-#     w.s1 <- sum(rlist.1[[k]][1:n1])+w.s1
-#     w.s2 <- sum(rlist.2[[k]][1:n1])+w.s2
-#     w.s3 <- sum(rlist.3[[k]][1:n1])+w.s3
-#     w.s4 <- sum(rlist.4[[k]][1:n1])+w.s4
-# #     w.s5 <- sum(rlist.5[[k]][1:n1])+w.s5
-# #     w.s6 <- sum(rlist.6[[k]][1:n1])+w.s6
-# #     w.s7 <- sum(rlist.7[[k]][1:n1])+w.s7
-# #     w.s8 <- sum(rlist.8[[k]][1:n1])+w.s8
-# #     w.s9 <- sum(rlist.9[[k]][1:n1])+w.s9
-#     }
-#     w.s <- sum(w.s1,w.s2,w.s3,w.s4) #,w.s5,w.s6,w.s7,w.s8,w.s9)
-#   
-# ## other statistic
-# # mean for full year
+
+tiff("../figures/bt1_qldp.tiff")
+hist(bt1_qldp.values,nclass=100,xlab="% positive Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
+points(qld_origp*100,0,pch=16,col="red",cex=1.8)
+dev.off()
+
+tiff("../figures/bt1_qldn.tiff")
+hist(bt1_qldn.values,nclass=100,xlab="% negative Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
+points(qld_orign*100,0,pch=16,col="red",cex=1.8)
+dev.off()
+
+tiff("../figures/bt1_nswp.tiff")
+hist(bt1_nswp.values,nclass=100,xlab="% positive Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
+points(nsw_origp*100,0,pch=16,col="red",cex=1.8)
+dev.off()
+
+tiff("../figures/bt1_nswn.tiff")
+hist(bt1_nswn.values,nclass=100,xlab="% negative Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
+points(nsw_orign*100,0,pch=16,col="red",cex=1.8)
+dev.off()
+@
+
+<<chunk8b,eval=F , echo = F, results = hide>>=
+# This is the same as chunk 8, but for ns = 4
+## multiple cells
+NstepTrend <- function(x,y,m,i){
+
+    n1 <- 2002-1997
+
+    if (i==1){
+
+      z <- 56  # no. of column, related to the dim of the region
+      # Qld 57 by 51
+      # NSW 38 by 33
+
+    } else {
+
+    z <- 37 }
+
+if (x%%z!=0){ #& x%%z!=(z-1)) {
+    y.1 <- m[[x]]
+    y.2 <- m[[x+1]]
+    y.3 <- m[[x+z]]
+    y.4 <- m[[x+z+1]]
+
+    if(!is.na(y.1[1])&!is.na(y.2[1])&!is.na(y.3[1])&!is.na(y.4[1])) { 
+    if (i == 1) {
+      n2 <- 2015-2004
+      v.1 <- ts(y.1[c(229:288,313:length(y.1))],
+                start=c(1998,1),frequency=12)
+      v.2 <- ts(y.2[c(229:288,313:length(y.2))],
+                start=c(1998,1),frequency=12)
+      v.3 <- ts(y.3[c(229:288,313:length(y.3))],
+                start=c(1998,1),frequency=12)
+      v.4 <- ts(y.4[c(229:288,313:length(y.4))],
+                start=c(1998,1),frequency=12)
+
+    } else {
+
+      n2 <- 2015-2002
+
+      v.1 <- ts(y.1[c(229:length(y.1))],start=c(1998,1),frequency=12)
+      v.2 <- ts(y.2[c(229:length(y.2))],start=c(1998,1),frequency=12)
+      v.3 <- ts(y.3[c(229:length(y.3))],start=c(1998,1),frequency=12)
+      v.4 <- ts(y.4[c(229:length(y.4))],start=c(1998,1),frequency=12)
+}
+
+## rank by month
+   for (j in 1:12) {
+    y1.mth <- data.frame(mth=as.vector(window(v.1,
+                                              start=c(1998,i),
+                                              frequency=1)),
+                         o=y[[j]])[order(y[[j]]),]
+    y2.mth <- data.frame(mth=as.vector(window(v.2,
+                                              start=c(1998,i),
+                                              frequency=1)),
+                         o=y[[j]])[order(y[[j]]),]
+    y3.mth <- data.frame(mth=as.vector(window(v.3,
+                                              start=c(1998,i),
+                                              frequency=1)),
+                         o=y[[j]])[order(y[[j]]),]
+    y4.mth <- data.frame(mth=as.vector(window(v.4,
+                                              start=c(1998,i),
+                                              frequency=1)),
+                         o=y[[j]])[order(y[[j]]),]
+
+    assign(paste("r1",j,sep=""),rank(y1.mth$mth))
+    assign(paste("r2",j,sep=""),rank(y2.mth$mth))
+    assign(paste("r3",j,sep=""),rank(y3.mth$mth))
+    assign(paste("r4",j,sep=""),rank(y4.mth$mth))
+
+    }
+
+## mann-whitney rank sum statistic
+    rlist.1 <- list(r11,r12,r13,r14,r15,r16,r17,r18,r19,r110,r111,r112) ## full year
+    rlist.2 <- list(r21,r22,r23,r24,r25,r26,r27,r28,r29,r210,r211,r212)
+    rlist.3 <- list(r31,r32,r33,r34,r35,r36,r37,r38,r39,r310,r311,r312)
+    rlist.4 <- list(r41,r42,r43,r44,r45,r46,r47,r48,r49,r410,r411,r412)
+    rlist <- list(unlist(rlist.1),unlist(rlist.2),
+                  unlist(rlist.3),unlist(rlist.4)) 
+## seasonal
+    w.s1 <- 0
+    w.s2 <- 0
+    w.s3 <- 0
+    w.s4 <- 0
+  for (k in 1:length(rlist.1)) {
+#   for (k in c(9:11)) {
+    w.s1 <- sum(rlist.1[[k]][1:n1])+w.s1
+    w.s2 <- sum(rlist.2[[k]][1:n1])+w.s2
+    w.s3 <- sum(rlist.3[[k]][1:n1])+w.s3
+    w.s4 <- sum(rlist.4[[k]][1:n1])+w.s4
+    }
+    w.s <- sum(w.s1,w.s2,w.s3,w.s4) #,w.s5,w.s6,w.s7,w.s8,w.s9)
+
+## other statistic
+# mean for full year
+    mu <- n1*(n1+n2+1)/2
+    mu.s <- 12*4*mu
+# ## mean for one season
 #     mu <- n1*(n1+n2+1)/2
-#     mu.s <- 12*4*mu
-# # ## mean for one season
-# #     mu <- n1*(n1+n2+1)/2
-# #     mu.s <- 3*4*mu
-#   
-# ## variance (standard variation).
+#     mu.s <- 3*4*mu
+
+## variance (standard variation).
 The autocorrelation of the residuals can be assumed to be null (checked, cor.test(), true for NSW, mostly true for WA.
@@
 there are some situations in WA require further investigation)
-# # assume only cross correlation
-#     ss <- n1*n2*(n1+n2+1)/12
+# assume only cross correlation
+    ss <- n1*n2*(n1+n2+1)/12
+    ss.s <- 0
+    for (kk in 1:4) {
+      for (kkk in 1:4) {
+#         for (kkkk in 1:12) {
+          ss.s <- 12*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
+#         }
+      }
+    }
+## variance for one season
+#     ss <- n1*n2*(n1+n2+1)/3
 #     ss.s <- 0
@@
 #       for (kkk in 1:4) {
-# #         for (kkkk in 1:12) {
-#           ss.s <- 12*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
-# #         }
+#           ss.s <- 3*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
 #       }
 #     }
-# ## variance for one season   
-# #     ss <- n1*n2*(n1+n2+1)/3
-# #     ss.s <- 0
-# #     for (kk in 1:4) {
-# #       for (kkk in 1:4) {
-# #           ss.s <- 3*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
-# #       }
-# #     }
-#   
-# ## step trend statistic
-#     z <- (w.s - mu.s)/sqrt(ss.s)
-#   
-#     return(z) }
-#     
-#      else {return(NA)} }
-#      
-#      else {return(NA)}
-#   
-#   }
-# 
-# 
-# zpq.values = numeric(R)
-# znq.values = numeric(R)
-# mth.lnsw <- rep(11,12)
-# mth.lqld <- rep(9,12)
-# 
-# ptm <- proc.time()
-# for (i in 1:R) {
-#   
-#   for (j in 1:12) {
-#     new.order[[j]]<-sample(1:mth.lqld[j])
-#   }
-#   
-#   temp <- unlist(lapply(1:(length(residual_qld)-56),function(x,y) NstepTrend(x,y,residual_qld,1),new.order))
-#   if(i==1) {write.table(temp,file="bt_zvalues_qld.txt",row.names=F)
-#             } else
-#               {write.table(temp,file="bt_zvalues_qld.txt",append=T,row.names=F,col.names=F)}
-#   zpq.values[i] <- length(subset(temp,temp>=1.645))/length(temp)
-#   znq.values[i] <- length(subset(temp,temp<=-1.645))/length(temp)
-# }
-# proc.time()-ptm
-# 
-# write.table(zpq.values,file="bt_outp_qld.txt",row.names=F)
-# write.table(znq.values,file="bt_outn_qld.txt",row.names=F)
-# 
-bt4_nswp.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_outp_nsw.txt",header=T,sep=""))*100
-bt4_nswn.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_outn_nsw.txt",header=T,sep=""))*100
-
-bt4_qldp.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_outp_qld.txt",header=T,sep=""))*100
-bt4_qldn.values <- unlist(read.table("B:/My Study/PhD\\AVSproject_laptop\\WrittenWork\\PhD_thesis\\chap3\\bt_outn_qld.txt",header=T,sep=""))*100
-
-setEPS()
-postscript("bt4_nswp.eps")
-hist(bt4_nswp.values,nclass=100,xlab="% positive Z'",main="4-pixel step trend",cex.lab=1.4,cex.main=1.3)
-points(nsw_origp*100,0,pch=16,col="red",cex=1.8)
-dev.off()
-
-setEPS()
-postscript("bt4_nswn.eps")
-hist(bt1_nswn.values,nclass=100,xlab="% negative Z'",main="4-pixel step trend",cex.lab=1.4,cex.main=1.3)
-points(nsw_orign*100,0,pch=16,col="red",cex=1.8)
-dev.off()
-
-setEPS()
-postscript("bt4_qldp.eps")
-hist(bt4_qldp.values,nclass=100,xlab="% positive Z'",main="4-pixel step trend",cex.lab=1.4,cex.main=1.3)
-points(qld_origp*100,0,pch=16,col="red",cex=1.8)
-dev.off()
-
-setEPS()
-postscript("bt4_qldn.eps")
-hist(bt4_qldn.values,nclass=100,xlab="% negative Z'",main="4-pixel step trend",cex.lab=1.4,cex.main=1.3)
-points(qld_orign*100,0,pch=16,col="red",cex=1.8)
+
+## step trend statistic
+    z <- (w.s - mu.s)/sqrt(ss.s)
+
+    return(z) }
+
+     else {return(NA)} }
+
+     else {return(NA)}
+
+  }
+
+zpq.values = numeric(R)
+znq.values = numeric(R)
+mth.lnsw <- rep(18,12)
+mth.lqld <- rep(16,12)
+
+ptm <- proc.time()
+for (i in 1:R) {
+
+  for (j in 1:12) {
+    new.order[[j]]<-sample(1:mth.lqld[j])
+  }
+
+  temp <- unlist(lapply(1:(length(residual_qld)-57),function(x,y) NstepTrend(x,y,residual_qld,1),new.order))
+  if(i==1) {write.table(temp,file="bt_zvalues_qld.txt",row.names=F)
+            } else
+              {write.table(temp,file="bt_zvalues_qld.txt",append=T,row.names=F,col.names=F)}
+  zpq.values[i] <- length(subset(temp,temp>=1.645))/length(temp)
+  znq.values[i] <- length(subset(temp,temp<=-1.645))/length(temp)
+}
+proc.time()-ptm
+
+write.table(zpq.values,file="bt_outp_qld.txt",row.names=F)
+write.table(znq.values,file="bt_outn_qld.txt",row.names=F)
+
+ptm <- proc.time()
+for (i in 1:R) {
+
+  for (j in 1:12) {
+    new.order[[j]]<-sample(1:mth.lnsw[j])
+  }
+
+  temp <- unlist(lapply(1:(length(residual_nswvic)-38),function(x,y) NstepTrend(x,y,residual_nswvic,2),new.order))
+  if(i==1) {write.table(temp,file="bt_zvalues_nswvic.txt",row.names=F)
+            } else
+              {write.table(temp,file="bt_zvalues_nswvic.txt",append=T,row.names=F,col.names=F)}
+  zpq.values[i] <- length(subset(temp,temp>=1.645))/length(temp)
+  znq.values[i] <- length(subset(temp,temp<=-1.645))/length(temp)
+}
+proc.time()-ptm
+
+write.table(zpq.values,file="bt_outp_nswvic.txt",row.names=F)
+write.table(znq.values,file="bt_outn_nswvic.txt",row.names=F)
+
+bt4_nswp.values <- unlist(read.table("bt_outp_nswvic.txt",header=T,sep=""))*100
+bt4_nswn.values <- unlist(read.table("bt_outn_nswvic.txt",header=T,sep=""))*100
+
+bt4_qldp.values <- unlist(read.table("bt_outp_qld.txt",header=T,sep=""))*100
+bt4_qldn.values <- unlist(read.table("bt_outn_qld.txt",header=T,sep=""))*100
+
+jpeg("../figures/bt4_hist.jpg")
+par(mfrow=c(2,2), mar = c(4,3,2,1)+0.1)
+# nsw pos
+hist(bt4_nswp.values,nclass=100,xlab="% positive Z'",main="",cex.lab=1.4,cex.main=1.3, xlim=c(0,20))
+points(nsw_origp.z.new*100,0,pch=16,col="red",cex=1.8)
+# nsw neg
+hist(bt1_nswn.values,nclass=100,xlab="% negative Z'",main="",cex.lab=1.4,cex.main=1.3, xlim=c(0,20))
+points(nsw_orign.z.new*100,0,pch=16,col="red",cex=1.8)
+#qld pos
+hist(bt4_qldp.values,nclass=100,xlab="% positive Z'",main="",cex.lab=1.4,cex.main=1.3, xlim=c(0,20))
+points(qld_origp.new*100,0,pch=16,col="red",cex=1.8)
+#qld neg
+hist(bt4_qldn.values,nclass=100,xlab="% negative Z'",main="",cex.lab=1.4,cex.main=1.3, xlim=c(0,20))
+points(qld_orign.new*100,0,pch=16,col="red",cex=1.8)
 dev.off()
@@
 <<chunk9,eval=F , echo = F, results = hide>>=
+# THIS IS THE APPLIED BOOTSTRAP ANALYSIS 20180724
+# This is the same as Chunk 8
 
@@
 write.table(z3nn.values,file="bt_out3n_nsw.txt",row.names=F)
-# 
-# setEPS()
-# postscript("bt3_qldp.eps")
-# hist(z3pq.values,nclass=100,xlab="% positive Z'",main="QLD",cex.lab=1.4,cex.main=1.3)
-# points(qld_origp,0,pch=16,col="red",cex=1.8)
-# dev.off()
-# 
-# setEPS()
-# postscript("bt3_qldn.eps")
-# hist(z3nq.values,nclass=100,xlab="% negative Z'",main="QLD",cex.lab=1.4,cex.main=1.3)
-# points(qld_orign,0,pch=16,col="red",cex=1.8)
-# dev.off()
-# 
-# setEPS()
-# postscript("bt3_nswp.eps")
-# hist(z3pn.values,nclass=100,xlab="% positive Z'",main="NSW",cex.lab=1.4,cex.main=1.3)
-# points(nsw_origp,0,pch=16,col="red",cex=1.8)
-# dev.off()
-# 
-# setEPS()
-# postscript("bt3_nswn.eps")
-# hist(z3nn.values,nclass=100,xlab="% negative Z'",main="NSW",cex.lab=1.4,cex.main=1.3)
-# points(nsw_orign,0,pch=16,col="red",cex=1.8)
-# dev.off()
+
+setEPS()
+postscript("bt3_qldp.eps")
+hist(z3pq.values,nclass=100,xlab="% positive Z'",main="QLD",cex.lab=1.4,cex.main=1.3)
+points(qld_origp,0,pch=16,col="red",cex=1.8)
+dev.off()
+
+setEPS()
+postscript("bt3_qldn.eps")
+hist(z3nq.values,nclass=100,xlab="% negative Z'",main="QLD",cex.lab=1.4,cex.main=1.3)
+points(qld_orign,0,pch=16,col="red",cex=1.8)
+dev.off()
+
+setEPS()
+postscript("bt3_nswp.eps")
+hist(z3pn.values,nclass=100,xlab="% positive Z'",main="NSW",cex.lab=1.4,cex.main=1.3)
+points(nsw_origp,0,pch=16,col="red",cex=1.8)
+dev.off()
+
+setEPS()
+postscript("bt3_nswn.eps")
+hist(z3nn.values,nclass=100,xlab="% negative Z'",main="NSW",cex.lab=1.4,cex.main=1.3)
+points(nsw_orign,0,pch=16,col="red",cex=1.8)
+dev.off()
 
@@
 
-require(SpatialVx)
-
-## reconstruct the data into format required in LocSig
-qld_res_mat <- matrix(unlist(residual_qld),nrow=360)
-nsw_res_mat <- matrix(unlist(residual_nswvic),nrow=360)
-
-## the step trend function for LocSig
-NstepTrend4 <- function(x,ii){
-
-  n1 <- sample(c(4:8),1)
-  zz <- vector()
-
-  for(m in 1:ncol(x)) {
-    if(!is.na(x[1,m])) {
-
-  v <- x[c(229:360),m]
-  ttrd <- c(1:132)
-  v <- residuals(lm(v~ttrd))
-
-  if(ii==1) {    
-      yy <- ts(v[c(1:60,85:132)],start=c(1998,1),frequency=12)    
-      n2 <- 9-n1
-    } else
-      if(ii==2) {
-      yy <- ts(v,start=c(1998,1),frequency=12)
-      n2 <- 11-n1
-    }
-
-  for(i in 1:12) { 
-  y.mth <- as.vector(window(yy,start=c(1998,i),frequency=1))
-
-  ## rank by month   
-    assign(paste("r",i,sep=""),rank(y.mth) )
-  }
-
-## mann-whitney rank sum statistic
-    rlist <- list(r1,r2,r3,r4,r5,r6,r7,r8,r9,r10,r11,r12) ## full year
-
-    w.s <- 0
-    for (k in 1:length(rlist)) {
-    w.s <- sum(rlist[[k]][1:n1])+w.s
-    }
-
-## other statistic
-## mean for full year
-    mu <- n1*(n1+n2+1)/2
-    mu.s <- length(rlist)*mu
-
-## variance (standard variation). The autocorrelation of the residuals can be assumed to be null (checked, true for QLD and NSW)
-    ss <- n1*n2*(n1+n2+1)/length(rlist)
-    ss.s <- length(rlist)*ss
-
-## step trend statistic
-    z <- (w.s - mu.s)/sqrt(ss.s)
-    zz[m] <- z
-    }
-
-     else {zz[m]<-NA}
-  }
-  return(zz)
-
-  }
+# require(SpatialVx)
+# 
+# ## reconstruct the data into format required in LocSig
+# qld_res_mat <- matrix(unlist(residual_qld),nrow=360)
+# nsw_res_mat <- matrix(unlist(residual_nswvic),nrow=360)
+# 
+# ## the step trend function for LocSig
+# NstepTrend4 <- function(x,ii){
+# 
+#   n1 <- sample(c(4:8),1)
+#   zz <- vector()
+# 
+#   for(m in 1:ncol(x)) {
+#     if(!is.na(x[1,m])) {
+# 
+#   v <- x[c(229:360),m]
+#   ttrd <- c(1:132)
+#   v <- residuals(lm(v~ttrd))
+# 
+#   if(ii==1) {
+#       yy <- ts(v[c(1:60,85:132)],start=c(1998,1),frequency=12)
+#       n2 <- 9-n1
+#     } else
+#       if(ii==2) {
+#       yy <- ts(v,start=c(1998,1),frequency=12)
+#       n2 <- 11-n1
+#     }
+# 
+#   for(i in 1:12) {
+#   y.mth <- as.vector(window(yy,start=c(1998,i),frequency=1))
+# 
+#   ## rank by month
+#     assign(paste("r",i,sep=""),rank(y.mth) )
+#   }
+# 
+# 
+# ## mann-whitney rank sum statistic
+#     rlist <- list(r1,r2,r3,r4,r5,r6,r7,r8,r9,r10,r11,r12) ## full year
+# 
+#     w.s <- 0
+#     for (k in 1:length(rlist)) {
+#     w.s <- sum(rlist[[k]][1:n1])+w.s
+#     }
+# 
+# ## other statistic
+# ## mean for full year
+#     mu <- n1*(n1+n2+1)/2
+#     mu.s <- length(rlist)*mu
+# 
+# 
+# ## variance (standard variation). The autocorrelation of the residuals can be assumed to be null (checked, true for QLD and NSW)
+#     ss <- n1*n2*(n1+n2+1)/length(rlist)
+#     ss.s <- length(rlist)*ss
+# 
+# ## step trend statistic
+#     z <- (w.s - mu.s)/sqrt(ss.s)
+#     zz[m] <- z
+#     }
+# 
+#      else {zz[m]<-NA}
+#   }
+#   return(zz)
+#
+#  }
 @
@@
 The model in Equation~\ref{eqn:model2} accounts for around 13\%\footnote{Here the adjusted $R^2$ was reported.
-Adjusted $R^2$ is the coefficient of determination, a measurement of the amount of variability predicted by the model adjusting for the number of explanatory terms} of the rainfall variations in the QLD region and 19\% in the NSW/VIC region on average.
+Adjusted $R^2$ is the coefficient of determination, a measurement of the amount of variability predicted by the model adjusting for the number of explanatory terms} of the rainfall variation in the QLD region and 19\% in the NSW/VIC region.
 The residual analysis shows that the assumptions of the regression model are generally met.
@@
 
-# ## calculate mean rainfall difference in two period for 11-year
-# 
-#   sel_neg.qld <- residual_qld[which(values(tr_qld.step.new)>=1.645)]
-# #   sel_pos.qld <- coef.rain.qld[which(values(tr_qld.step)<(-1.645))]
-#   
-#   sel_neg.nswvic <- residual_nswvic[which(values(tr_nswvic.step.new)>=1.645)]
-# 
-# ## compare mean of before and after
-# ## boxplot
-#   data.qld <- data.frame(year=seq(1998,2008,1))
-#   for (i in 1:length(sel_neg.qld)) {
-#     data.qld[,i+1] <- as.vector(rollapply(as.zoo(sel_neg.qld[[i]])[229:360],width=12,sum,by=12))
-#   }
-#   data2.qld <- melt(data.qld,id="year")
-#   data2.qld$group[data2.qld$year<2003] <- "1before"
-#   data2.qld$group[data2.qld$year>2004] <- "2after"
-# 
-# 
-#   data.nswvic <- data.frame(year=seq(1998,2008,1))
-#   for (i in 1:length(sel_neg.nswvic)) {
-#     data.nswvic[,i+1] <- as.vector(rollapply(as.zoo(sel_neg.nswvic[[i]])[229:360],width=12,sum,by=12))
+## calculate mean rainfall difference in two period for 11-year
+
+  sel_neg.qld <- residual_qld[which(values(tr_qld.step.new)>=1.4395)]
+#   sel_pos.qld <- coef.rain.qld[which(values(tr_qld.step)<(-1.645))]
+
+  sel_neg.nswvic <- residual_nswvic[which(values(tr_nswvic.step.new)>=1.4395)]
+
+## compare mean of before and after
+## boxplot
+
+# Qld  
+  years <- rep(seq(1979,2015,1), each=12)
+  data.qld <- data.frame(year=seq(1979,2015,1))
+# aggregate by year    
+  sel_neg.qld_an <-lapply(sel_neg.qld,
+                           function(x) aggregate(x,list(Year=years),
+                                                 sum))
+
+# add to year column                               
+  for (i in 1:length(sel_neg.qld_an)) {
+    data.qld[,i+1] <- sel_neg.qld_an[[i]]$x
+  }
+# rearrange  
+data2.qld <- data.qld %>%
+  filter(year > 1997) %>%
+  gather(key="point", value="Value", V2:V34) %>%
+  mutate(group = ifelse(year < 2003,"1before","2after"))
+
+# NSW
+  data.nswvic <- data.frame(year=seq(1979,2015,1))
+# aggregate by year  
+  sel_neg.nswvic_an <-lapply(sel_neg.nswvic,
+                           function(x) aggregate(x,list(Year=years),
+                                                 sum))
+
+# add to year column                               
+  for (i in 1:length(sel_neg.nswvic_an)) {
+    data.nswvic[,i+1] <- sel_neg.nswvic_an[[i]]$x
+  }
+
+# select and rearrange
+data2.nswvic <- data.nswvic %>%
+  filter(year > 1997) %>%
+  gather(key="point", value="Value", V2:V34) %>%
+  mutate(group = ifelse(year < 2003,"1before","2after"))
+
+#   nsw_sig <- nsw_30[which(Cp30_nswvic<=0.05)]
+#   data.nswvic <- data.frame(year=seq(1979,2008,1))
+#   for (i in 1:length(nsw_sig)) {
+#     data.nswvic[,i+1] <- as.vector(rollapply(as.zoo(nsw_sig[i]),width=12,sum,by=12))
 #   }
@@
 #   data2.nswvic$group[data2.nswvic$year>2002] <- "2after"
-# 
-# #   nsw_sig <- nsw_30[which(Cp30_nswvic<=0.05)]
-# #   data.nswvic <- data.frame(year=seq(1979,2008,1))
-# #   for (i in 1:length(nsw_sig)) {
-# #     data.nswvic[,i+1] <- as.vector(rollapply(as.zoo(nsw_sig[i]),width=12,sum,by=12))
-# #   }
-# #   data2.nswvic <- melt(data.nswvic,id="year")
-# #   data2.nswvic$group[data2.nswvic$year<2003] <- "1before"
-# #   data2.nswvic$group[data2.nswvic$year>2002] <- "2after"
-# 
-# 
-# ## perform a 2 sample t test
-#   var.test(data2.qld[data2.qld$year<2003,3],data2.qld[data2.qld$year>2004 & data2.qld$year!=2006,3],alternative="two.sided")
-# ## unequal variance
-#   t.test(data2.qld[data2.qld$year<2003,3],data2.qld[data2.qld$year>2004 & data2.qld$year!=2006,3],alternative="greater")
-# ## p = 2.2e-16
-# 
-#   var.test(data2.nswvic[data2.nswvic$year<2003,3],data2.nswvic[data2.nswvic$year>2002& data2.nswvic$year!=2006,3],alternative="two.sided")
-# ## unequal variance
-#   t.test(data2.nswvic[data2.nswvic$year<2003,3],data2.nswvic[data2.nswvic$year>2002& data2.nswvic$year!=2006,3],alternative="greater")
-# ## p < 1.4e-13
-# 
-# 
-#   setEPS()
-#   postscript("compare_mean.eps")
-#   par(mfrow=c(1,2))
-#   boxplot(value~group,data=data2.qld,names=c("before","after"),boxwex=0.4,main="QLD site",ylab="rainfall (mm)")
-# 
-#   boxplot(value~group,data=data2.nswvic,names=c("before","after"),boxwex=0.4,main="NSW/VIC site")
-#   dev.off()
+
+## perform a 2 sample t test
+  var.test(data2.qld[data2.qld$year<2003,3],data2.qld[data2.qld$year>2004 & data2.qld$year!=2006,3],alternative="two.sided")
+## unequal variance
+  t.test(data2.qld[data2.qld$year<2003,3],data2.qld[data2.qld$year>2004 & data2.qld$year!=2006,3],alternative="greater")
+## p = 2.2e-16
+
+  var.test(data2.nswvic[data2.nswvic$year<2003,3],data2.nswvic[data2.nswvic$year>2002& data2.nswvic$year!=2006,3],alternative="two.sided")
+## unequal variance
+  t.test(data2.nswvic[data2.nswvic$year<2003,3],data2.nswvic[data2.nswvic$year>2002& data2.nswvic$year!=2006,3],alternative="greater")
+## p < 1.4e-13
+
+  setEPS()
+  postscript("compare_mean.eps")
+  par(mfrow=c(1,2))
+  boxplot(value~group,data=data2.qld,names=c("before","after"),boxwex=0.4,main="QLD site",ylab="rainfall (mm)")
+
+  boxplot(value~group,data=data2.nswvic,names=c("before","after"),boxwex=0.4,main="NSW/VIC site")
+  dev.off()
 

### 2018-09-04 Scripts/Extractrainfalldata.Rmd
changes for paper

---
title: "Extracting Gridded Rainfall Data"
author: "Willem Vervoort, Chun Liang, Floris van Ogtrop"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/rver4657/owncloud/Chun/chun")
require(tidyverse)
```

# Introduction
This document is related to the paper: Empirical study of land surface effects on local rainfall (Liang et al.) and describes the script to extract the gridded rainfall data from the Australian gridded rainfall data set from BOM. We have a paid licence for this data in the group and have the data stored on a shared server. The main contact for the data in the group is [A/Prof Thomas Bishop](mailto:thomas.bishop@sydney.edu.au).
The formal metadata for the Gridded High Resolution Daily Rainfall is [at the BOM website](http://www.bom.gov.au/climate/austmaps/metadata-daily-rainfall.shtml).

# Defining the extent of the area and generating the points  
The data is produced on a 0.05^o × 0.05^o grid. The study works with two areas, one in Queensland and one in NSW. The extent of these areas are defined below and the points generated to extract the rainfall data.

```{r generatePoints}
Lats_Qld <- seq(-28.075, -25.575, by = 0.05)
Longs_Qld <- seq(145.825, 148.625, by = 0.05)

Lats_NSW <- seq(-36.675, -35.075, by = 0.05)
Longs_NSW <- seq(147.075, 148.925, by = 0.05)

# use the vectors to generate a set of locations
Loc_NSW <- tibble(ID = numeric(length=length(Lats_NSW)*length(Longs_NSW)),
                  longs = numeric(length=length(Lats_NSW)*length(Longs_NSW)),
                  lats = numeric(length=length(Lats_NSW)*length(Longs_NSW)))
Loc_Qld <- tibble(ID = numeric(length=length(Lats_Qld)*length(Longs_Qld)),
                  longs = numeric(length=length(Lats_Qld)*length(Longs_Qld)),
                  lats = numeric(length=length(Lats_Qld)*length(Longs_Qld)))

for (i in 1:length(Lats_NSW)) {
  for (j in 1:length(Longs_NSW)) {
    Loc_NSW[(i-1)*length(Longs_NSW)+j,] <- c((i-1)*length(Longs_NSW)+j,
                                             Longs_NSW[j],Lats_NSW[i])
  }
}

for (i in 1:length(Lats_Qld)) {
  for (j in 1:length(Longs_Qld)) {
    Loc_Qld[(i-1)*length(Longs_Qld)+j,] <- c((i-1)*length(Longs_Qld)+j,
                                             Longs_Qld[j],Lats_Qld[i])
  }
}

# how many sites in each?
nrow(Loc_NSW)

nrow(Loc_Qld)
```

We can check whether this went OK by plotting the resulting lat and long values.

```{r checkplot}
Loc_NSW %>%
  ggplot(aes(longs,lats, colour=ID)) + geom_point() + ggtitle("NSW")

Loc_Qld %>%
  ggplot(aes(longs,lats, colour=ID)) + geom_point() + ggtitle("Qld")
```

Now combine the two frames to make one final dataset for the extraction of the rainfall data

```{r combine}
Loc <- rbind(Loc_NSW,Loc_Qld)
```

# Extract rainfall data using Parallel

Extracting the rainfall takes time, so the easiest way to do this is using parallel computing, so we can use multiple cores on the PC.

```{r parallel}
# we need some packages

library(zoo)
library(raster)
library(rgdal)
library(doParallel)

# directory where the data is stored
datadir <- "R:/grp-hgis/public/BOM_GriddedRainfallData"

# read in the list of decades
decades <- dir(paste(datadir,"Daily-rainfall",sep="/"), pattern="rainfall_")

# store to put data in
Store <- list()

# make the parallel cluster
cl <- makeCluster(4)
registerDoParallel(cl)
# run the foreach loop and export packages to slaves
Store<- foreach(i=1:(length(decades)-1), 
                 .packages=c("raster","rgdal")) %dopar% {    
  # read in the list of years
  Years <- dir(paste(datadir,"Daily-rainfall",decades[i],sep="/"))
  yearlist <- list()

  for (j in 1:length(Years)) {
    # read the grids
    files= list.files(paste(datadir,"Daily-rainfall",decades[i],Years[j],sep="/"),
                      pattern=".txt", full.names=TRUE)
    # stack the grids
    s <- stack(files)
    # extract the stations
    df <- extract(s, 
                  SpatialPoints(as.data.frame(Loc[,2:3]),
  proj4string=CRS("+proj=longlat +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +no_defs")),
              df=TRUE, method='simple')
    # store in the list
    yearlist[[j]] <- df
  }
  yearlist
}
stopCluster(cl)
```

Once this is done, we need to do some combining and massaging to convert the list to a data frame that can be used in analyses.

```{r wrangling}
output1 <- list()
for (i in 1:(length(decades)-1)) {
  output1[[i]] <- do.call(cbind,Store[[i]])
}

#remove ID column
for (i in 1:(length(decades)-1)) {
  remove <- grep("ID",names(output1[[i]]))
  output1[[i]] <- output1[[i]][,-as.numeric(remove)]
}

output2 <- do.call(cbind,output1)
# rename and save
Raindata <- output2
# save as an Rdata file
saveRDS(Raindata,file="Data/bom/rainfall/Raindata.RDS")

```

### 2018-09-04 DATA/other/bushfires/read_code.r

@@ -1,0 +1,3 @@
+version https://git-lfs.github.com/spec/v1
+oid sha256:dd27652b8cc39e546805696a24053ed91a00527e0c6df292f4908f61da0ef33e
+size 3783

### 2018-09-04 Scripts/MODIS_download1.R

@@ -1,0 +1,123 @@
+require(RCurl)
+require(httr)
+require(tidyverse)
+# MOD44B Chun's project (annual data)
+# FTP server
+FTP <- "https://e4ftl01.cr.usgs.gov/MOLT/MOD44B.006/"
+# read in list of folders
+items <- try(strsplit(getURL(FTP,ssl.verifypeer = FALSE,dirlistonly = TRUE), "\r*\n")[[1]],silent=TRUE)
+items2 <- items[grep("folder.gif",items)]
+
+# split out folder names
+dirs2 <- items2 %>%
+  str_split("\"", simplify=T)
+dirs3 <- dirs2[,6] %>%
+  str_sub(1,10)
+
+
+start.date <- "2000.01.01"
+end.date <- "2016.06.30"
+
+##dates
+dates=dirs3[dirs3 <= end.date & dirs3 >= start.date]
+
+## locations (h and v bands for satellite)
+# Queensland
+#h=c(30,31) 
+#v=11
+# NSW
+h = c(29,30)
+v = 12
+
+## Get image filenames 
+getlist2 <- list()
+for(j in 1:length(h)){
+  for(i in 1:length(dates)){
+  	getlist <- try(strsplit(getURL(paste(FTP,dates[i], "/", sep="")), "\r*\n")[[1]],silent=TRUE)
+  	getlist1 <- grep(paste("h",h[j],"v",v,sep=""), getlist)[2]
+  	getlist2[[(j-1)*length(dates)+i]] <- getlist[getlist1]
+  }
+}
+# split out the image names from the paths
+image1 <- lapply(getlist2, function(x) str_split(x, "\"", simplify = TRUE))
+image2 <- na.omit(unlist(lapply(image1, function(x) as.vector(unlist(x))[6])))
+
+# definition of FTP server and working dir
+WDIR <- "C:\\Users\\rver4657\\owncloud\\Chun2\\Data\\Satellite\\MOD44"
+FTP1 <- "https://e4ftl01.cr.usgs.gov/MOLT/MOD44B.006/"
+
+dates2 <- c(dates,dates)
+
+## Download images
+for(i in 1:length(image2)){
+	GET(paste(FTP1, dates2[i], image2[i], sep = "/") 
+	, write_disk(paste(WDIR, image2[i], sep = "/"), overwrite=TRUE) 
+	, authenticate("willemv", "WsOL2010"))
+}
+
+
+
+## --------------------------------------------------------------------
[…]

### 2018-09-04 Scripts/Modis_MRT_merge.R

@@ -1,0 +1,122 @@
+## MRT operations from R
+# This file contains two different MRT functions 
+# to resample/reproject and to mosaic
+# Niranjan, Floris and Willem
+# a specific example application is given
+
+## --------------------------------------------------------------------
+## reproject images if you want to do this
+# custom function to write parameter file for MRT
+# The easiest way to get the inputs right for MRT is to check the manual
+# and run the GUI on a test example
+MRTproj <- function(fname="tmp.prm",hdfName,outname,MRTpath,UL="",LR="",resample_type="NEAREST_NEIGHBOR",proj_type="UTM",
+                    bands_subset="",proj_params="0 0 0 0 0 0 0 0 0 0 0 0",datum="WGS84",utm_zone=NA,pixel_size=1000){
+  
+  filename = file(fname, open="wt")
+  write(paste("INPUT_FILENAME = ", getwd(), "/",hdfName, sep=""), filename) 
+  # 	browser()
+  if (bands_subset != "") {
+    write(paste("SPECTRAL_SUBSET = ( ",bands_subset," )",sep=""),filename,append=TRUE)
+  }
+  if (UL[1] != "" & LR[1] != "") {
+    write("SPATIAL_SUBSET_TYPE = INPUT_LAT_LONG", filename, append=TRUE)
+    write(paste("SPATIAL_SUBSET_UL_CORNER = ( ", as.character(UL[1])," ",as.character(UL[2])," )",sep=""), filename, append=TRUE)
+    write(paste("SPATIAL_SUBSET_LR_CORNER = ( ", as.character(LR[1])," ",as.character(LR[2])," )",sep=""), filename, append=TRUE)
+  }
+  write(paste("OUTPUT_FILENAME = ", outname, sep=""), filename, append=TRUE)
+  write(paste("RESAMPLING_TYPE = ",resample_type,sep=""), filename, append=TRUE)
+  write(paste("OUTPUT_PROJECTION_TYPE = ",proj_type,sep=""), filename, append=TRUE)
+  write(paste("OUTPUT_PROJECTION_PARAMETERS = ( ",proj_params," )",sep=""), filename, append=TRUE, ncolumns=5)
+  write(paste("DATUM = ",datum,sep=""), filename, append=TRUE)
+  if (proj_type == "UTM") write(paste("UTM_ZONE = ",utm_zone,sep=""), filename, append=TRUE)
[…]

### 2018-09-04 Scripts/Test_MOD44B_analysis.R

@@ -1,0 +1,101 @@
+# create raster brick for both NSW and Qld sites
+require(raster)
+#library(plotKML)
+library(rgeos) #gSimplify
+library(maptools) # readShapePoly
+library(fields) # image.plot
+
+
+setwd("C:/users/rver4657/owncloud/chun2/data/satellite/MOD44/B")
+
+# read in the file list from the directory
+x <- dir(pattern="Percent_Tree_Cover.tif")
+# Identify the different states in the images
+# NSW
+NSW <- x[grep("NSW",x)]
+# Qld
+Qld <- x[grep("Qld",x)]
+
+img <- list()
+# raster and brick
+for(i in 1:length(NSW)) {
+  img[[i]] <- raster(NSW[i])
+}
+
+imgQ <- list()
+# raster and brick
+for(i in 1:length(Qld)) {
+  imgQ[[i]] <- raster(Qld[i])
+}
+
+NSW_b <- do.call(brick,img)
+Qld_b <- do.call(brick, imgQ)
+
+tm <- seq(as.Date("2000-03-06"), as.Date("2016-03-06"), "year")
+mod_tc_r22 <- setZ(NSW_b, tm, "year")
+
+mod_tc_r11 <- setZ(Qld_b, tm, "year")
+
+## apply t-test to tree cover data
+fun <- function(x,m) {
+  var_x <- sd(x,na.rm=T)
+  mu_1 <- mean(x[1:3],na.rm=T)
+  if (m==1) {y <- x[4:17]} else {y <- x[6:17]}   
+  mu_2 <- mean(y,na.rm=T)
+  z <- (mu_2-mu_1)/var_x
+  if (!is.na(z) & abs(z)>1.645) {return(mu_2-mu_1)} else {return(NA)}
+}
+
+mod_tc_r1.z <- calc(mod_tc_r11,function(x) fun(x,1))
+mod_tc_r2.z <- calc(mod_tc_r22,function(x) fun(x,2))
+
+root_dir <- "c:/users/rver4657/owncloud/chun/chun/data"
+aus_reg <- readShapePoly(paste(root_dir,
+                               "Geo\\aus_state2011\\STE11aAust.shp",sep="/"),
+                         proj4string=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0"))
+aus_reg2 <- gSimplify(aus_reg,0.01,topologyPreserve=T)
+
[…]

### 2018-09-05 DATA/other/bushfires/read_code.r
updated lfs again

@@ -1,3 +1,0 @@
-version https://git-lfs.github.com/spec/v1
-oid sha256:dd27652b8cc39e546805696a24053ed91a00527e0c6df292f4908f61da0ef33e
-size 3783

### 2019-02-28 Manuscripts/PeerJRainfall_Landuse/Rainfall_Landuse.Rmd
update for PeerJ paper

--- 2018-09-04 Manuscripts/Rainfall_Landuse_paper/Rainfall_Landuse_paper.Rmd
@@
 ---
-title: Detecting influences of changes in land cover on observed rainfall.
+title: Detecting the impact of land cover change on observed rainfall.
+preprint: false
 author:
   - name: Chun X. Liang
-    email: summer.chunliang@gmail.com 
-    affiliation: a
+    affiliation: 1
   - name: Floris F. van Ogtrop
-    email: floris.vanogtrop@sydney.edu.au 
-    affiliation: a
+    affiliation: 1
   - name: R. Willem Vervoort
+    affiliation: 1
+    corresponding: true
     email: willem.vervoort@sydney.edu.au
-    affiliation: a
-    footnote: Corresponding Author
-address:
-  - code: a
+affiliation:
+  - code: 1
     address: Sydney Institute of Agriculture, The University of Sydney, NSW 2006
-abstract: |
+abstract: >
  Analysis of observational data to identify relationships between rainfall and land cover change are difficult due to multiple environmental factors that cannot be strictly controlled.
@@
 However, high rainfall variability and possible regrowth meant that no signifcant changes were observed in alonger time sereies to 2015.
-
-journal: "Journal of Hydrology Regional Studies"
+header-includes: >
+  \usepackage{lipsum}
+  \usepackage{textcomp}
+bibliography: reflist_chap3.bib
 date: "`r Sys.Date()`"
-bibliography: reflist_chap3.bib
 output: 
- bookdown::pdf_book:
-    base_format: rticles::elsevier_article
+  rticles::peerj_article:
+    base_format: bookdown::pdf_document2 # bookdown::pdf_document2 # for using \@ref()
 ---
@@
 library(maptools)
-#library(plotKML)
+library(plotKML)
 library(raster)
@@
 knitr::opts_knit$set(root.dir = 
-                       "C:/Users/rver4657/ownCloud/chun/chun/manuscripts/Rainfall_Landuse_paper")
-```
-
-Introduction
-==========
+                       "E:/cloudstor/chun/chun/manuscripts/PeerJRainfall_Landuse")
+```
+
+# Introduction {-}
 
 Land use and land cover changes can lead to changes in the local climate.
-Empirical and modelling studies have found cloud types and rainfall are correlated to large scale vegetation cover changes, such as deforestation in the Amazon and in the Sahel [@Chagnon2005;@Pinto2009; @Wang2009; @Mei2010; @kucharski_further_2013; @pitman_scale_2016] and afforestation in south Israel [@Otterman1990; @Ben-Gai1998].
+Empirical and modelling studies have found that cloud types and rainfall are correlated to large scale vegetation cover changes, such as deforestation in the Amazon and in the Sahel [@Chagnon2005;@Pinto2009; @Wang2009; @Mei2010; @kucharski_further_2013; @pitman_scale_2016] and afforestation in south Israel [@Otterman1990; @Ben-Gai1998].
 Using airborne measurement in Western Australia, @Junkermann2009 showed a significantly higher level of aerosols over an agricultural area compared to an adjacent natural vegetation.
@@
 This is because there are some fundamental experimental difficulties in both space (where does evaporated water reappear as rainfall?) and in time (how much time does it take for land cover change effects to appear or disappear?).
-In addition, in many areas across the globe rainfall variability is related to complex set of interactions, of which land use change might only be a minor component.
+In addition, in many areas across the globe, rainfall variability is related to a complex set of interactions, of which land use change might only be a minor component.
 
@@
 In contrast, local ET is determined by local land surface characteristics, which influence local scale atmospheric dynamics and hence the amount of rainfall, including contribution from both main sources.
-Therefore land surface is suggested to play an important role in local rainfall.
 
@@
 @Ma2011; @Zeng2012; @pitman_scale_2016; @saha_investigating_2016].
-However, they are mostly based on modelling experiments and little evidence was reported from observations.
-However, @Pitman2004 found a good match between observations and simulated rainfall changes in southwest Western Australia ,forced by land cover change.
+However, most are based on modelling experiments and few empirical observational studies have been reported.
+However, @Pitman2004 found a good match between observations and simulated rainfall changes in southwest Western Australia, forced by land cover change.
 @Timbal2006 were able to reproduce the rainfall decline in south west Australia by including land cover influence.
@@
 
-Therefore, the aim of this study is to expand the available methodology to use empirical evidence at regional scales to investigate the cause and effect relationship between land cover change and local rainfall.
+Therefore, the aim of this study is to use a statistical approach on rainfall data at regional scales to investigate the cause and effect relationship between land cover change and local rainfall, which is demonstrated in many modelling studies.
 More specifically, we hypothesize that a step change on the land cover on the surface will cause a step change in the rainfall.
-To demonstrate the approach and this effect we applied the methodology to study the changes in rainfall at two locations in Queensland and NSW/Victoria where there are possible step changes in land cover change due to land clearing and bush fires.
-The methodology is mainly based on statistical approaches to identify changes in rainfall, which were subsequently associated with land cover change through spatial comparison.
+To demonstrate this we study changes in observed rainfall over time at a Queensland and NSW/Victoria location where there are possible step changes in land cover change due to land clearing and bush fires.
+The methodology uses statistical approaches to identify changes in rainfall, which are subsequently associated with land cover change through spatial comparison.
 
@@
 
-We use two areas in Australia to demonstrate the approach.
-In Australia, significant tree cover change has mainly occurred in the north east of the continent and on the southeast coast, as well as in the southwest of Western Australia.
-According to the National Dynamic Land Cover Dataset (DLCD) [@Lymburner2010], most of these areas have experienced decreasing EVI post 2000.
+In Australia, significant tree cover change has mainly occurred in the north east  and south east of the continent,  as well as in the southwest of Western Australia.
+According to the National Dynamic Land Cover Dataset (DLCD) [@Lymburner2010], most of these areas have experienced decreases in the Enhanced Vegetation Index (EVI) post 2000 aws derived from satellite data.
 As an index for vegetation greenness, the decreasing values indicate lower biomass over time in the tree cover regions.
@@
 In site 1 (the QLD region), the tree cover is mostly sparse.
-In site 2 (the NSW/VIC region), many areas have open or close forest in which tree cover is denser.", out.width="90%", echo=F}
+In site 2 (the NSW/VIC region), many areas have open or close forest where the tree cover is denser.", out.width="90%", echo=F}
 
@@
 
-Two regions were selected where significant tree cover change was reported.
-The first region is located in south central Queensland to the north of the Murray Darling Basin (MDB) (site 1 in Figure \@ref(fig:selreg).
+Two regions were selected where significant tree cover change since 2000 was reported.
+The first region is located in south central Queensland partly covering the north of the Murray Darling Basin (MDB) (site 1 in Figure \@ref(fig:selreg)).
 High rates of land clearing have been reported in this region during the early 2000s [@SLATS2004].
-The second study region is located at the border of New South Wales and Victoria, and includes the Snowy Mountain ranges (site 2 in Figure \@ref(fig:selreg).
-Severe bush fires occurred in this area and the surroundings in early 2003 (see Figure \@ref(fig:figure2bushfire)).
-The 2003 bush fires were the largest and the worst in this area for the last 60 years [@Fire2011].
-Two thirds of Kosciuszko national park was heavily burned and regrowth was reported to be slow due to drought and cold conditions [@ABC2003] and the type of species in this region.
+The second study region is located at the border of New South Wales and Victoria, and includes the Snowy Mountain ranges (site 2 in Figure \@ref(fig:selreg)).
+Severe bush fires occurred in this area  in early 2003 (see Figure \@ref(fig:figure2bushfire)).
+The 2003 bush fires were the largest and the worst in the last 60 years [@Fire2011].
+Two thirds of Kosciuszko national park was heavily burned and regrowth was reported to be slow due to drought and cold conditions [@ABC2003], and the type of species in this region.
 However, in the longer term, after an early high transpiration period a recovery of pre-fire evapotranspiration would be expected [@Kuzcera1987].
-As a result , significant tree cover loss has happened in both study areas in the last decade, either permanently or temporarily.
+For the purpose of this study, significant tree cover loss has happened in both study areas in the last decade, either permanently or temporarily.
 
 ```{r figure2bushfire, fig.cap = "Location of bushfires occurring in January 2003, in and around the NSW/VIC study region, as shown by the red pixels.
-The map shows large area in the Kosciuszko national park has been burned.
-Some locations in the southwest of ACT have also experienced intensive bushfires.", out.width="90%", echo=F}
+The map shows the large area in the Kosciuszko national park that has been burned.
+Some locations in the southwest of teh Australian Capital Territory (ACT) have also experienced intensive bushfires.", out.width="90%", echo=F}
 include_graphics("figures/bushfire_nswvic.jpg")
@@
 
-The two regions have different climate characteristics.
-The QLD region is partially grassland and partially subtropical, while the NSW/VIC region is mainly within the temperate zone, under the Köppen classification.
-According to Australian Bureau of Meteorology (BoM), the NSW/VIC region receives 1000 - 2000 mm rainfall annually, which is more than double of the rainfall in the QLD region.
+The two regions have different characteristics.
+The QLD region is partially grassland and subtropical, while the NSW/VIC region is mainly within the temperate zone, under the Köppen classification.
+According to Australian Bureau of Meteorology (BoM), the NSW/VIC region receives 1000 - 2000 mm rainfall annually, which is more than double the annual rainfall in the QLD region.
 Evapotranspiration is similar in both regions.
@@
 Tree cover has been cleared at a massive scale over the last decade, especially during 2002 - 2004.
+The reports from the Queensland Statewide Land Cover and Trees Study (SLATS) [e.g.
+@SLATS2001; @SLATS2017] were used to investigate the time and location of the land clearing in the QLD region.
+
 The Kosciuszko national park is within the NSW/VIC region.
@@
 Although land clearing is not a major issue in this region, it is vulnerable to fires and drought.
-
-Therefore two types of land cover changes were studied.
-The reports from the Queensland Statewide Land Cover and Trees Study (SLATS) [e.g.
-@SLATS2001; @SLATS2017] were used to investigate the time and location of land clearing in the QLD region.
 The MODIS burned area product, MCD45A1 [@Roy2002; @Roy2005; @Roy2008], was used to locate bush fires areas in the NSW/VIC region, with a grid resolution of 500 m.
 MCD45A1 provides monthly burning information on all pixels, which helps to pinpoint an abrupt event.
-Due to the nature of the different land cover change, the post-change vegetation status in the two regions is expected to be different (see Figure \@ref(fig:figure3-tc-simple)).
+
+Due to the difference in nature of the land cover change in the two regions, the post-change vegetation status is hypothesized to be different as well (see Figure \@ref(fig:figure3-tc-simple)).
 
@@
 ## the selected regions for mapping
-root_dir <- "c:/users/rver4657/owncloud/chun/chun/data"
+root_dir <- "e:/cloudstor/chun/chun/data"
 aus_reg <- readShapePoly(paste(root_dir,
@@
 fire_0301 <- raster(paste(root_dir,
-          "Other\\bushfires\\Burnt_Area\\tmpA2003001.burndate.jpg",
+          "Other\\bushfires\\Burnt_Area\\tmpA2003001.burndate.tif",
                           sep="/"))
 fire_0301nsw <- crop(fire_0301,e2)
-fire_loc <- crop(reproject(fire_0301nsw,
-                           CRS=CRS("+proj=longlat +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84")),
-                 e222)
+fire_0301nsw_t <- projectRaster(fire_0301nsw,
+                           crs = "+proj=longlat +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84")
+fire_loc <- crop(fire_0301nsw_t,e222)
 remove(list=c("fire_0301","fire_0301nsw"))
@@
 
-rootdir <-"C:/users/rver4657/owncloud/chun2/data/satellite/MOD44/B"
+rootdir <-"E:/cloudstor/chun2/data/satellite/MOD44/B"
 
@@
 # Qld 2011
-mod_tc_r1_7y.z <- calc(mod_tc_r11,function(x) fun(x,5))
+#mod_tc_r1_7y.z <- calc(mod_tc_r11,function(x) fun(x,5))
 # NSW
@@
 # NSW 7 years after (2011)
-mod_tc_r2_7y.z <- calc(mod_tc_r22,function(x) fun(x,5))
+#mod_tc_r2_7y.z <- calc(mod_tc_r22,function(x) fun(x,5))
 # # NSW 9 years after
@@
 
-jpeg("../../figures/tc_figs.jpg", width=1280, height=960, quality=100)
-par(mar=c(3,3,2,11),mfrow=c(2,2))
-# Qld 7 year 2011
-plot(e111, cex.axis=1.5, font.axis=2, ylab="")
-image(mod_tc_r1_7y.z,col=plotclr,breaks=c(-70,-20,-10,-5,0,5,10,20,70),
+jpeg("../../figures/tc_figs.jpg", width=1280, height=640, quality=100)
+par(mfrow=c(1,2), mar=c(5,3,5,5))
+
+#par(mar=c(3,9,2,5))
+plot(e111, cex.axis=1.5, font.axis=2, ylab="",xlab="")
+image(mod_tc_r1.z,col=plotclr,breaks=c(-70,-20,-10,-5,0,5,10,20,70),
       legend=F,add=T)
 #plot(e111,add=T)
-plot(aus_reg2,border="grey",add=T)
-plot(mdb_reg2,border="grey",lwd=3,add=T)
-legend(146,-27.5,c("State","MDB"),col=c("grey","grey"),
-       lwd=c(1,3),bty="n",title="Boundary",cex=1.8)
+plot(aus_reg2,border="grey",add=T,lwd=3,ylab="")
+plot(mdb_reg2,border="grey",lwd=5,add=T,ylab="")
+# FULL PERIOD NSW
+par(mar=c(5,0,5,11))
+plot(e222, cex.axis=1.5, font.axis=2, ylab="",xlab="")
+image(mod_tc_r2.z,col=plotclr,breaks=c(-70,-20,-10,-5,0,5,10,20,70),
+      legend=F,add=T)
+#plot(e111,add=T)
+plot(aus_reg2,border="grey",add=T, lwd=3,ylab="")
+plot(mdb_reg2,border="grey",lwd=5,add=T,ylab="")
 image.plot(zlim=c(-4,4), breaks=c(-4,-3,-2,-1,-0.2,0.2,1,2,3,4),
@@
            legend.shrink=0.8,
-           legend.width=1.5,
+           legend.width=3,
+           legend.mar=5.1,
            legend.args=list(text=expression(paste(Delta,
                             "tree (% ground)",sep="")),
-                            line=1,cex=1.3,font=2),
+                            line=1,cex=1.5,font=2),
            axis.args=list(cex.axis=1.5),graphics.reset=F)
-
-par(mar=c(3,9,2,5))
-plot(e111, cex.axis=1.5, font.axis=2, ylab="")
-image(mod_tc_r1.z,col=plotclr,breaks=c(-70,-20,-10,-5,0,5,10,20,70),
-      legend=F,add=T)
-#plot(e111,add=T)
-plot(aus_reg2,border="grey",add=T)
-plot(mdb_reg2,border="grey",lwd=3,add=T)
-# image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,-0.2,0.2,1,2,3,4),
-#            lab.breaks=c(-70,-20,-10,-5,"no tree","0 or",5,10,20,70),
-#            legend.only=T,horizontal=F,
-#            col=c("#C51B7D","#DE77AE","#F1B6DA","#FDE0EF","white",
-#                  "#E6F5D0","#B8E186","#7FBC41","#4D9221"),
-#            legend.shrink=0.7,
-#            legend.width=0.7,
-#            legend.args=list(text=expression(paste(Delta,
-#                             "tree (% ground)",sep="")),
-#                             line=0.5,cex=0.9,font=2),
-#            axis.args=list(cex.axis=0.8),graphics.reset=F)
-# 2011 NSW
-par(mar=c(3,3,1,11))
-plot(e222, cex.axis=1.5, font.axis=2, ylab="")
-image(mod_tc_r2_7y.z,col=plotclr,breaks=c(-70,-20,-10,-5,0,5,10,20,70),
-      legend=F,add=T)
-#plot(e111,add=T)
-plot(aus_reg2,border="grey",add=T)
-plot(mdb_reg2,border="grey",lwd=3,add=T)
-image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,-0.2,0.2,1,2,3,4),
-           lab.breaks=c(-70,-20,-10,-5,"no tree","0 or",5,10,20,70),
-           legend.only=T,horizontal=F,
-           col=c("#C51B7D","#DE77AE","#F1B6DA","#FDE0EF","white",
-                 "#E6F5D0","#B8E186","#7FBC41","#4D9221"),
-           legend.shrink=0.8,
-           legend.width=1.5,
-           legend.args=list(text=expression(paste(Delta,
-                            "tree (% ground)",sep="")),
-                            line=1,cex=1.3,font=2),
-           axis.args=list(cex.axis=1.5),graphics.reset=F)
-
-# FULL PERIOD NSW
-par(mar=c(3,9,1,5))
-plot(e222, cex.axis=1.5, font.axis=2, ylab="")
-image(mod_tc_r2.z,col=plotclr,breaks=c(-70,-20,-10,-5,0,5,10,20,70),
-      legend=F,add=T)
-#plot(e111,add=T)
-plot(aus_reg2,border="grey",add=T)
-plot(mdb_reg2,border="grey",lwd=3,add=T)
-# image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,-0.2,0.2,1,2,3,4),
-#            lab.breaks=c(-70,-20,-10,-5,"no tree","0 or",5,10,20,70),
-#            legend.only=T,horizontal=F,
-#            col=c("#C51B7D","#DE77AE","#F1B6DA","#FDE0EF","white",
-#                  "#E6F5D0","#B8E186","#7FBC41","#4D9221"),
-#            legend.shrink=0.7,
-#            legend.width=0.7,
-#            legend.args=list(text=expression(paste(Delta,
-#                             "tree (% ground)",sep="")),
-#                             line=0.5,cex=0.9,font=2),
-#            axis.args=list(cex.axis=0.8),graphics.reset=F)
-dev.off()
-
-jpeg("../../figures/map.jpeg", width=960, height=960)
-par(mar=c(1.5,1.5,1.5,1.5))
-plot(mdb_reg2,border="grey",lwd=3)
-plot(aus_reg2,border="grey",add=T)
-legend(150.2,-36,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=1.5)
-#invisible(text(getSpPPolygonsLabptSlots(aus_reg2), labels=as.character(aus_reg2$NAME), cex=0.7,adj=0.5))
-box(lty=1)
-plot(e111,col="red",add=T)
-plot(e222,col="red",add=T)
-text(147.2,-26.8,"QLD",cex=1.5)
-text(148,-34.7,"NSW/VIC",cex=1.5)
-dev.off()
-
-```
-
-The hypothesis in this study therefore is that the effect of 2003 - 2004 land clearings in the QLD region and the 2003 bush fires in the NSW/VIC region cause a step change in the local rainfall.
-The actual tree cover change at the pixel level during this time was derived from the 15-year MODIS data (discussed below).
+legend(147.1,-36.2,c("State","MDB"),col=c("grey","grey"),lwd=c(2,5),bty="n",title="Boundary",cex=1.5)
+dev.off()
+
+```
+
+The overall hypothesis is that the effect of 2003 - 2004 land clearings in the QLD region and the 2003 bush fires in the NSW/VIC region cause a step change in the local rainfall.
+The actual tree cover change during this time at the pixel level was derived from the 15-year MODIS data (discussed below).
 The difference of tree cover before and after the land disturbance was tested using a Student's t-test.
-As the length of the tree cover data is shorter than available rainfall data, earlier land clearings in the QLD region cannot be identified spatially, hence they are excluded from the analysis.
-
-## Data {#Data}
-
+As the length of the tree cover data is shorter than available rainfall data, earlier land clearing in the QLD region cannot be identified spatially, hence they are excluded from the analysis.
+
+Data {#Data}
+==============
 Several land surface data sets were used in this study.
@@
 
-Rainfall data for Australia [@Jones2009] was obtained from the Bureau of Meteorology.
-The data has been projected onto a national 0.05\textdegree × 0.05\textdegree grid (approximately 5 km × 5 km).
-This gridded data set was generated from station observations using an optimised Barnes successive correction technique.
-The Barnes technique combines a weighted averaging process and defined topographical information to estimate rainfall values between spatial points [@BoM2009].
-The resulting data set provides additional information for data-sparse areas like central Australia but reduces information in the data-rich areas, such as southeast Australia where station density is up to 20 per 100 $\text{km}^2$.
-The data is available on a monthly basis from 1900 to current.
+The SILO Rainfall product data for Australia was used [@SILO2001] (available online at \url{https://silo.longpaddock.qld.gov.au/})) .
+The data has been projected onto a national 0.05\textdegree by 0.05\textdegree grid (approximately 5 km by 5 km).
+This gridded data set was generated from station observations using spline interpolation and kriging [@SILO2001].
+The data has been compared to other gridded products and observed data and is generally of high quality [@Beesley2009; @Tozer2012].
+The data is available on a daily and monthly basis from 1889 to current.
 Here a subset of 36 years (1979 - 2015) was used.
-The study was conducted on monthly data, as a land cover change effect on annual rainfall might be negligible but can often found to be significant in particular months or seasons [e.g.
+The study was conducted on monthly data, as a land cover change effect on annual rainfall might be negligible, but can often  be significant in particular months or seasons [e.g.
 @Otterman1990; @Gaertner2001; @Semazzi2001; @Oleson2004; @Deo2009].
@@
 For example the Southern Annular Mode (SAM) is found to be more important than ENSO in south Western Australia [@Meneghini2007].
-The suitability of each index for the regions of interest was tested in section \@ref(sec:reg_model).
+The testing of the suitability of each index for the regions of interest is described in a later section.
 The following climate indices were used as candidate predictors for local rainfall.
@@
 
-## Statistical method
-
-The step changes on rainfall were analysed using two different statistical methods.
-A step change is not obvious in the time series data of the rainfall residuals (Figure \@ref(fig:ts-mean)), even though the data is deseasonalised and detrended.
-Both methods in this paper make use of a regression model to remove variability in rainfall due to climatic factors to strengthen the tree cover change signal.
-In the first method, the tree cover change was implemented as a factor variable in the regression model.
-In the second method, a rank sum test (step trend test), was applied to the model residuals after effects of other major factors were removed.
-This assumes that after removal of all climate variation, the vegetation cover change is the only factor explaining the non-random pattern in the rainfall residuals.
+Statistical method
+===================
+
+As an initial analysis a simple boxplot and t-test is used to analyse whether there is a significant change in tree cover in time, and before and after the suspected change in the regions.
+
+To assess the actual causal relationship between the tree cover and the rainfall a flexible regression model is applied (discussed in detail below).
+A step change is not directly obvious in the time series  of the rainfall anomalies (Figure \@ref(fig:ts-mean)) for both regions, even though the data is deseasonalised and detrended.
+In this study we apply different statistical methods to analyse the effect of tree cover change on rainfall.
+Both methods  make use of a regression model to remove year-on-year variability in rainfall to strengthen the tree cover change signal.
+
+In the first method, the tree cover change is implemented as a factor variable in the regression model, and the significance of this variable is tested.
+In the second method, a rank sum test (step trend test), is applied to the regression model residuals after effects of other major factors were removed.
+This assumes that after removal of all climate, long term linear trend and seasonal variation, the vegetation cover change is the only factor explaining the non-random pattern in the rainfall residuals.
 
@@
 As highlighted in the introduction, the Australian climate is influenced by sea surface temperatures in the tropical Pacific and Indian Oceans, as well as pressure systems in the Southern Ocean [@BoM2012].
-@Risbey2009 compared five large-scale drivers, including ENSO (measured by SOI and the Tropical Pacific SSTs), IOD, SAM, MJO (Madden-Julian oscillation) and blocking, in relation to Australian rainfall variability.
-The MJO is a large scale eastward-propagating wave-like disturbance in equatorial latitudes [@Risbey2009].
-They identified SOI as the most important index among all indices tested for broad parts of Australia (including QLD and NSW/VIC) in almost any season.
-In this study, four indices were selected from the main climatic indicators (see section [Data]) were used as the explanatory variables in the model for each study region.
+@Risbey2009 compared five large-scale drivers, including ENSO (measured by SOI and the Tropical Pacific Sea surface temperatures (SSTs)), IOD, SAM, MJO (Madden-Julian oscillation) and blocking, in relation to Australian rainfall variability.
+The MJO is a large scale eastward-propagating wave-like disturbance located around equatorial latitudes [@Risbey2009].
+They identified SOI as the most important index among all climate indices tested for broad parts of Australia (including QLD and NSW/VIC) in almost any season.
+In this study, four climate indices were selected from the main climatic indicators (see section [Data]) and used as the explanatory variables in the model for each study region.
+A further complicating factor is the influence of the "millenium drought" over the study period and in particular the change to wet conditions in 2010 - 2011 [@vanDijk2013].
+Therefore, the spatially averaged monthly rainfall in the Murray Darling Basin (MDB, downloaded from [the Bureau of Meteorology](http://www.bom.gov.au/web01/ncc/www/cli_chg/timeseries/rain/allmonths/mdb/latest.txt)) was used to explain the year-on-year variation in the rainfall in the regions.
+As climate alone are unlikely able to fully explain the temporal variation in rainfall in the regions, including the drought [@vanDijk2013; @Westra2010], this variable was included.
+Since both regions at least partly overlap with the MDB, the average rainfall across was assumed to be a useful explaining variable.
 
@@
 
-# read in the BOM gridded rainfall data extracted earlier
-rain <- readRDS("../../data/bom/rainfall/Raindata.rds")
-# dim(rain)
-# 43829 columns (days from 19000101) by 4161 locations (rows)
-
-# split by Qld and NSW
-# NSW has first 1254 locations (columns) Qld other 2907
-rainNSW <- rain[1:1254,]
-rainQld <- rain[1255:4161,]
-rm(rain)
-
-# summarise to monthly
+# SILO data
+
+## ----extract_Silo-----------------------------------------------------
+years <- seq(1979,2016,by=1)
+
+SILOdata <- list()
+
+# run a loop
+for (i in 1:length(years)) {
+   # url <- paste("https://s3-ap-southeast-2.amazonaws.com/silo-open-data/annual/monthly_rain/",
+   #              years[i],
+   #              ".monthly_rain.nc",sep="")
+   # curl_download(url,
+   #               paste("data/monthly_rainfall_",years[i],".nc",sep=""))
+  SILOdata[[i]] <- brick(paste("../../data/monthly_rainfall_",years[i],".nc",
+                          sep=""),varname="monthly_rain")
+}
+
+SILOdata__b <- do.call(stack, SILOdata)
+
+#crop to extents
+# e111 <- extent(c(145.825, 148.625, -28.075, -25.575)) 
+# e222 <- extent(c(147.075, 148.925, -36.675, -35.075))
+
 #  NSW
-rainNSW_t <- as.tibble(data.frame(Date = 
-                                    seq.Date(as.Date("1900-01-01"),
-                                    as.Date("2015-12-31"),by=1),
-                    rain=t(rainNSW)))
-rm(rainNSW)
-rainNSW_t <- rainNSW_t %>% 
-  mutate(Month = format(Date,"%Y-%m")) 
-rain_monthNSW <- rainNSW_t %>%
-  group_by(Month) %>%
-  summarise_at(vars(starts_with("rain")), funs(sum = sum(.)))
+SILOdata.NSW <- crop(SILOdata__b, extent(e222)) 
+#QLD
+SILOdata.QLD <- crop(SILOdata__b, extent(e111))
+# # MDB
+# mdb_reg <- readShapePoly("../../data/Geo\\MDB\\mdb_boundary.shp",
+#                                proj4string=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0"))
+SILOdata.MDB <- crop(SILOdata__b, extent(mdb_reg))
+# 
+# aus_reg <- readShapePoly("../../data/Geo\\aus_state2011\\STE11aAust.shp",
+#                          proj4string=CRS("+proj=lcc +lat_1=-18 +lat_2=-36")) 
+#
+
+# remove orig
+rm(SILOdata)
+rm(SILOdata__b)
+# test -----------------------------
+# require(rasterVis)
+# gplot(SILOdata.NSW[[1]])  + geom_tile(aes(fill = value)) + coord_equal() +
+#   facet_wrap(~variable) + scale_fill_gradient(low="darksalmon", high="blue",
+#                                         guide="colorbar",na.value="white") +
+#   xlab("Longitude") + ylab("Latitude")
+# ------------------------
+
+# rewrite as a monthly dataframe
+# NSW
+rainNSW_t <- as.tibble(data.frame(Date =  seq.Date(as.Date("1979-01-01"),
+                                   as.Date("2016-12-31"),by="month"),
+                                  rain=t(SILOdata.NSW@data@values)))
+
+# calculate mean by row
+nsw_mean <- rainNSW_t %>%
+  mutate(Means = rowMeans(.[,-1], na.rm = T)) %>%
+  dplyr::select(Date,Means)
+
+rm(SILOdata.NSW)
 
 # Qld
-rainQld_t <- as.tibble(data.frame(Date = 
-                                    seq.Date(as.Date("1900-01-01"),
-                                    as.Date("2015-12-31"),by=1),
-                    rain=t(rainQld)))
-rm(rainQld)
-rainQld_t <- rainQld_t %>% 
-  mutate(Month = format(Date,"%Y-%m")) 
-rain_monthQld <- rainQld_t %>%
-  group_by(Month) %>%
-  summarise_at(vars(starts_with("rain")), funs(sum = sum(.)))
-rm(rainNSW_t, rainQld_t)
-
-## 100 years data & 30 years
-# Qld
-qld_100 <- rain_monthQld
-qld_30 <- qld_100 %>%
- filter(Month >= "1979-01")
-
-qld100_mean <- qld_100 %>%
-  mutate(Means = rowMeans(.[,-1])) %>%
-  dplyr::select(Month,Means)
-
-qld30_mean <- qld100_mean %>%
- filter(Month >= "1979-01")
-
-# NSW
-nsw_100 <- rain_monthNSW
-nsw_30 <- nsw_100 %>%
- filter(Month >= "1979-01")
-
-nsw100_mean <- nsw_100 %>%
-  mutate(Means = rowMeans(.[,-1])) %>%
-  dplyr::select(Month,Means)
-
-nsw30_mean <- nsw100_mean %>%
- filter(Month >= "1979-01")
-
-require(MASS)
-fitdistr(qld30_mean$Means,"gamma")  ## assume shape=0.9
-fitdistr(nsw30_mean$Means,"gamma")  ## assume shape =2.19
-ks.test(qld30_mean$Means,"pgamma",0.90,0.03)
-ks.test(nsw30_mean$Means,"pgamma",2,0.03)
-
-jpeg("../../figures/hist_rainfall.jpg", width=960, height=480)
-par(mfrow=c(1,2))
-hist(qld30_mean$Means,main="",xlab=expression(paste("rainfall (mm mont",h^-1,")",sep="")))
- hist(nsw30_mean$Means,main="",xlab=expression(paste("rainfall (mm mont",h^-1,")",sep="")))
+rainQld_t <- as.tibble(data.frame(Date =  seq.Date(as.Date("1979-01-01"),
+                                   as.Date("2016-12-31"),by="month"),
+                                  rain=t(SILOdata.QLD@data@values)))
+
+# calculate mean by row
+qld_mean <- rainQld_t %>%
+  mutate(Means = rowMeans(.[,-1], na.rm = T)) %>%
+  dplyr::select(Date,Means)
+
+rm(SILOdata.QLD)
+
+# MDB
+rainMDB_t <- as.tibble(data.frame(Date =  seq.Date(as.Date("1979-01-01"),
+                                   as.Date("2016-12-31"),by="month"),
+                                  rain=t(SILOdata.MDB@data@values)))
+
+# calculate mean by row
+MDB_mean <- rainMDB_t %>%
+  mutate(MDBrain = rowMeans(.[,-1], na.rm = T)) %>%
+  dplyr::select(Date,MDBrain) %>%
+  filter(Date < "2016-01-01")
+
+rm(SILOdata.MDB)
+
+# rainfall histogram
+
+rain <- tibble(Rainfall = c(nsw_mean$Means,qld_mean$Means),
+               State = c(rep("NSW/VIC",length(nsw_mean$Means)),
+                         rep("QLD",length(qld_mean$Means))))
+
+jpeg("../../figures/hist_rainfall.jpg")
+rain %>%
+  ggplot(aes(Rainfall)) + geom_histogram(fill="blue") +
+  facet_wrap(~State) + xlab("Monthly Rainfall (mm)") + 
+  ylab("Frequency") +
+  theme(axis.text=element_text(size=rel(2)),
+        axis.title=element_text(size=rel(2.5)),
+        strip.text =element_text(size=rel(2), face="bold"))
 dev.off()
@@
 acf_df <- data_frame(lag = 1:12, 
-                     acf_Qld = acf(qld30_mean$Means,
+                     acf_Qld = acf(qld_mean$Means,
                       lag.max=12,main="QLD")$acf[2:13,1,1],
-                     acf_NSW = acf(nsw30_mean$Means,
+                     acf_NSW = acf(nsw_mean$Means,
                   lag.max=12,main="NSW & VIC")$acf[2:13,1,1],
-                     pacf_Qld = pacf(qld30_mean$Means,
+                     pacf_Qld = pacf(qld_mean$Means,
                              lag.max=12,main="")$acf[,1,1],
-                     pacf_NSW = pacf(nsw30_mean$Means,
+                     pacf_NSW = pacf(nsw_mean$Means,
                            lag.max=12,main="")$acf[,1,1])
@@
 # deseasonalise package
-qld30_deseason <- ds(qld30_mean$Means,ic="AIC", type="monthly",
+qld_deseason <- ds(qld_mean$Means,ic="AIC", type="monthly",
                      standardizeQ=F)$z
-nsw30_deseason <- ds(nsw30_mean$Means,ic="AIC", type="monthly",
+nsw_deseason <- ds(nsw_mean$Means,ic="AIC", type="monthly",
                      standardizeQ=F)$z
-qld100_deseason <- ds(qld100_mean$Means,ic="AIC", type="monthly",
+qld100_deseason <- ds(qld_mean$Means,ic="AIC", type="monthly",
                       standardizeQ=F)$z
-nsw100_deseason <- ds(qld100_mean$Means,ic="AIC", type="monthly",
+nsw100_deseason <- ds(qld_mean$Means,ic="AIC", type="monthly",
                       standardizeQ=F)$z
@@
 # test whether any linear trends are significant
-df_list <- list(qld30 = data.frame(z=qld30_deseason,
-                           trend=1:length(qld30_deseason)),
-                nsw30 = data.frame(z=nsw30_deseason,
-                           trend=1:length(nsw30_deseason)),
+df_list <- list(qld = data.frame(z=qld_deseason,
+                           trend=1:length(qld_deseason)),
+                nsw = data.frame(z=nsw_deseason,
+                           trend=1:length(nsw_deseason)),
                 qld100 = data.frame(z=qld100_deseason,
-                 trend=1:length(qld100_deseason)),
+                 trend=1:length(qld_deseason)),
                 nsw100 = data.frame(z=nsw100_deseason,
@@
 # 
-plot_df <- tibble(Dates = rep(ymd(paste(qld30_mean$Month,
-                                        "01",sep="_")),2),
-                  Resid = c(qld30_deseason,nsw30_deseason),
-                  State = c(rep("Qld",length(qld30_deseason)),
-                            rep("NSW",length(nsw30_deseason))))
+plot_df <- tibble(Dates = rep(qld_mean$Date,2),
+                  Resid = c(qld_deseason,nsw_deseason),
+                  State = c(rep("Qld",length(qld_deseason)),
+                            rep("NSW",length(nsw_deseason))))
 
@@
 plot_df_qld <- tibble(Lag = ccf(as.vector(SOI_30),
-                            as.vector(qld30_deseason),lag.max=12, 
+                            as.vector(qld_deseason),lag.max=12, 
                             plot=F)$lag[,,1],
                   SOI = ccf(as.vector(SOI_30),
-                            as.vector(qld30_deseason),lag.max=12, 
+                            as.vector(qld_deseason),lag.max=12, 
                             plot=F)$acf[,,1],
                   NINO3.4 = ccf(as.vector(SST34_30),
-                                as.vector(qld30_deseason),lag.max=12,
+                                as.vector(qld_deseason),lag.max=12,
                                 plot=F)$acf[,,1],
                   NINO4 = ccf(as.vector(SST4_30),
-                              as.vector(qld30_deseason),lag.max=12,
+                              as.vector(qld_deseason),lag.max=12,
                               plot=F)$acf[,,1],
                   PDO = ccf(as.vector(PDO_100),
-                            as.vector(qld100_deseason),lag.max=12,
+                            as.vector(qld_deseason),lag.max=12,
                             plot=F)$acf[,,1],
                   IOD = ccf(as.vector(IOD_30),
-                            as.vector(qld30_deseason),lag.max=12, 
+                            as.vector(qld_deseason),lag.max=12, 
                             plot=F)$acf[,,1])
@@
 plot_df <- tibble(Lag = ccf(as.vector(SOI_30),
-                            as.vector(nsw30_deseason),lag.max=12, 
+                            as.vector(nsw_deseason),lag.max=12, 
                             plot=F)$lag[,,1],
                   SOI = ccf(as.vector(SOI_30),
-                            as.vector(nsw30_deseason),lag.max=12, 
+                            as.vector(nsw_deseason),lag.max=12, 
                             plot=F)$acf[,,1],
                   NINO3.4 = ccf(as.vector(SST34_30),
-                                as.vector(nsw30_deseason),lag.max=12,
+                                as.vector(nsw_deseason),lag.max=12,
                                 plot=F)$acf[,,1],
                   NINO4 = ccf(as.vector(SST4_30),
-                              as.vector(nsw30_deseason),lag.max=12,
+                              as.vector(nsw_deseason),lag.max=12,
                               plot=F)$acf[,,1],
@@
                   IOD = ccf(as.vector(IOD_30),
-                            as.vector(nsw30_deseason),lag.max=12, 
+                            as.vector(nsw_deseason),lag.max=12, 
                             plot=F)$acf[,,1])
@@
 Rainfall in each study region was first deseasonalised and detrended using the seasonal decomposition function `ds` in the package `deseasonalise` in R [@Rstats2018].
-Using detrended data gives a better indication of the underlying correlation rather than the correlation between trends in the data [@Smith2012].
+Using detrended data gives a better indication of the underlying correlation compared to the correlation between trends in the data [@Smith2012].
 The cross-correlations between the deseasonalised and detrended rainfall and the climatic indices were tested using the Pearson's product moment correlation method, assuming the relationships are linear.
@@
 
-Based on the correlation between the climatic indices and rainfall (as shown in Figure \@ref(fig:cor-rain-qld) and Figure \@ref(fig:cor-rain-nsw)), it can be concluded that:
+Based on the correlation between the climatic indices and rainfall in the regions (as shown in Figure \@ref(fig:cor-rain-qld) and Figure \@ref(fig:cor-rain-nsw)), it can be concluded that:
 \begin{itemize}
@@
   \setlength{\parskip}{0cm}
-  \item In QLD, the correlation between rainfall and SOI at zero time lags is the highest across all indices, outweighing the other ENSO indicators. IOD, had a weak influence in QLD.
-  \item In NSW/VIC, again the SOI has the highest correlation with rainfall, followed by the IOD. Both occur at the zero time lags. 
+  \item In QLD, the correlation between rainfall and SOI at zero time lags is the strongest across all indices, outweighing the other ENSO indicators. IOD and PDO have a weak influence in QLD.
+  \item In NSW/VIC, again the SOI has the strongest correlation with rainfall, followed by the IOD. For both, the strongest correlations occur at the zero time lags. 
   \item in both cases PDO had the weakest correlations, and this factor was not further considered as a predictor.
+  \item in general, for the better correlated indices, strongest correlations occured at zero time lags.
 \end{itemize}
 The above findings are consistent with previous studies.
-Although some indices are serially correlated with rainfall up to several months, the lag zero events have the most significant correlation coefficients.
-Concurrent climatic index series were generally found most useful in rainfall prediction [e.g.
+Although some indices are serially correlated with rainfall up to several months, the lag zero events have the greatest correlation coefficients.
+Using multiple climatic index series was generally found most useful in rainfall prediction [e.g.
 @Risbey2009; @Kamruzzaman2011].
-The correlations between the climatic indices and rainfall for each individual season have also been tested, and the results were similar.
 
@@
 We assumed all the factors are additive smooth components in determining rainfall following @Kamruzzaman2011.
-Generally, monthly rainfall has a skewed distribution so the normality assumption of the residuals in a general linear model could be violated.
 In this case, the rainfall model is a generalised additive model (GAM) [@Hastie1986] with a log link function `g()` and assuming the residuals are gamma distributed (see Figure \@ref(fig:hist-rain)).
 This means all predictors are modelled as smooth functions, in this case using the shrinkage version of the cubic regression splines [@Wood2011].
+All splines were limited to 3 knots in flexibility [@Wood2011], to reduce the risk of overfitting.
 \vspace{0.5cm}
@@
 \begin{array}{lll}
-g(E(\mathbf{R}_r)) = &\beta_0 + s_1(\mathbf{SOI}) + s_2(\mathbf{IOD}) + \\                &s_3(\mathbf{Nino3.4}) + s_4(\mathbf{Nino4}) + s_5(\mathbf{Season}) + \\
+g(E(\mathbf{R}_r)) = &\beta_0 + s_1(\mathbf{MDB_{monthlyRain}} + \\ &s_2(\mathbf{SOI}) + s_3(\mathbf{IOD}) + \\                &s_4(\mathbf{Nino3.4}) + s_5(\mathbf{Nino4}) + s_6(\mathbf{Season}) + \\
            &\beta_1\mathbf{Trend} + \boldsymbol{\epsilon}_r
@@
 The region is indicated by $r$, while $\beta_u$ ($_u$=0, 1) are the fitted coefficients in the model.
-$s_v$ ($_v$=1, 2, 3) are the smooth penalized cubic regression spline functions on the climatic indices and the season.
-Apart from dropping the PDO as a predictor, all other climate indices were included, allowing the model to select the appropriate predictors.
+$s_v$ ($_v$=1, 2, 3,...) are the smooth penalized cubic regression spline functions on the climatic indices and the season.
+MDB_{monthlyRain} is the spatially averaged monthly rainfall across the Murray Darling Basin.
+Apart from dropping PDO as a predictor, all other climate indices were included, allowing the model to select the appropriate predictors.
 
@@
 
-One of the main difficulties in empirical observation studies on the effect of land cover change on rainfall is the lack of continuous monitoring of land surface variables, or even, that no specific variable can be defined that can clearly represent the land surface process.
-Given the lack of a full picture of the land surface process, a factor variable was used in this study to represent the abrupt land surface change (see Equation \@ref(eq:model)).
+One of the main difficulties in empirical observation studies related to the effect of land cover change on rainfall is the lack of continuous monitoring of land surface variables.
+Moreover, no specific variable can possibly be defined that can clearly represent the land surface process.
+Given the lack of a full picture of the land surface process, a factor variable was used in the regression model to represent the abrupt land surface change (see Equation \@ref(eq:model)).
 The change could be a result of either land clearing or bush fires as long as it is permanent or takes a long time to recover.
-Here we approached the problem with two different models.
+As indicated we approached the problem in two different models ways.
 
@@
 \begin{array}{lll}
-g(E(\mathbf{R}_r)) = &\beta'_0 + s'_1(\mathbf{SOI}) + s'_2(\mathbf{IOD}) + \\
-  &s'_3(\mathbf{Nino3.4}) + s'_4(\mathbf{Nino4}) + s'_5\mathbf{Season} + \\
+g(E(\mathbf{R}_r)) = &\beta'_0 + s'_1(\mathbf{MDB_{monthlyRain}} + \\
+  &s'_2(\mathbf{SOI}) + s'_3(\mathbf{IOD}) + \\
+  &s'_4(\mathbf{Nino3.4}) + s'_5(\mathbf{Nino4}) + s'_6\mathbf{Season} + \\
   &\beta'_1\mathbf{Trend} + \beta'_2\mathbf{LC} + \boldsymbol{\epsilon'}_r
@@
 However, the information on the change in type of land cover during the time period is missing.
-Therefore, four scenarios were initially tested in the analysis.
-In these scenarios the "after change" period started from: (1) June 2003, (2) January 2004, (3) June 2004 and (4) January 2005.
-A further complicating factor is the influence of the "millenium drought" over the study period and in particular to change to wet conditions in 2010 - 2011 [@vanDijk2013].
+Therefore, four scenarios were initially tested in the analysis (**see supplementary material**).
 In the NSW/VIC region, severe bush fires were reported in early January 2003.
 Hence the "tree" cover state was up to December 2002 then it was changed to "removed" state from January 2003.
-However, there is similar uncertainty about when the tree cover had recovered from the bushfire and how this interacted with climate variability.
-As a result several scenarios were tested for both locations: 1) analysis up to 7 years post change (assumed to be 2010 for both regions); and 2) analysis of the full period (2015).
-More detail about this is later in the methods.
 As a starting date, the regression model was run from 1979 for both regions.
@@
 rain.trend <- function(x,y,var.data) {
-
+  # add a very small number to take out 0 values
   var.data$rain <- as.vector(unlist(y[x]))[1:nrow(var.data)]+0.0001
 
-  #if (!is.na(mean(var.data$rain))) {
-#    if (grepl("qld",substitute(y))) {
-      #var.data <- na.omit(var.data)
-      rain_fit <- gam(rain ~ s(month, k=3, bs="cs")
+      rain_fit <- gam(rain ~ MDB
                   + s(SOI, k=3, bs="cs")
@@
                   + trd
-                  , data=var.data,family=Gamma(link=log),
+                  , data=var.data,#family=Gamma(link=log),
                   na.action = na.omit)
-    # } else
-    #  if (grepl("nsw",substitute(y))) {
-    #   var.data <- na.omit(var.data)
-    #   rain_fit <- gam(rain ~ s(month, k=3)
-    #               + s(SOI, k=3)
-    #               + s(IOD, k=3)
-    #               + cover
-    #               + trd
-    #               , data=var.data,family=Gamma(link=log))
-   # }
     #
@@
 
-## note: link=log generally has higher r.sq value than link=reverse
-
 ## generate variables for the model
 trd <- seq(1:444)
-month <-rep(1:12,times=37)
-mth <- cycle(ts(month,start=c(1979,1),frequency=12))
 
@@
 # NSW
-var.data <- data.frame(month, trd, SOI = SOI_30a$SOI,
+var.data <- data.frame(trd, MDB = MDB_mean$MDBrain, 
+                       SOI = SOI_30a$SOI,
                        IOD = IOD_30a$IOD,
-                       PDO = PDO_100a$PDO[949:1392],
+                       #PDO = PDO_100a$PDO,
                        NINO3.4 = SST34_30a$SST34,
@@
                        cover = LC)
-# # 5 years
-# var.data5 <- var.data %>%
-#   filter(trd <= 348)
-
-# 7 years to 2009
-var.data7 <- var.data %>%
-  filter(trd <= 372)
-# 8 years to 2010
- var.data8 <- var.data %>%
-   filter(trd <= 384)
-
-# # 9 years
-# var.data9 <- var.data %>%
-#   filter(trd <= 396)
 
@@
 # test different scenarios
-# var.qld1 <- data.frame(month, trd, SOI = SOI_30a$SOI,
-#                        IOD = IOD_30a$IOD,
-#                        PDO = PDO_100a$PDO[949:1392],
-#                        NINO3.4 = SST34_30a$SST34,
-#                        NINO4 = SST4_30a$SST4,
-#                        cover = LC_qld1)
 #
-var.qld2 <- data.frame(month, trd, SOI = SOI_30a$SOI,
+var.qld2 <- data.frame(trd, MDB = MDB_mean$MDBrain, 
+                       SOI = SOI_30a$SOI,
                        IOD = IOD_30a$IOD,
-                       PDO = PDO_100a$PDO[949:1392],
+                       #PDO = PDO_100a$PDO,
                        NINO3.4 = SST34_30a$SST34,
@@
                        cover = LC_qld2)
-#
-# var.qld3 <- data.frame(month, trd, SOI = SOI_30a$SOI,
-#                        IOD = IOD_30a$IOD,
-#                        PDO = PDO_100a$PDO[949:1392],
-#                        NINO3.4 = SST34_30a$SST34,
-#                        NINO4 = SST4_30a$SST4,
-#                        cover = LC_qld3)
-
-var.qld4 <- data.frame(month, trd, SOI = SOI_30a$SOI,
+
+var.qld4 <- data.frame(trd, MDB = MDB_mean$MDBrain,
+                       SOI = SOI_30a$SOI,
                        IOD = IOD_30a$IOD,
-                       PDO = PDO_100a$PDO[949:1392],
+                       #PDO = PDO_100a$PDO,
                        NINO3.4 = SST34_30a$SST34,
@@
                        cover = LC_qld4)
-# 31 years: 1979 - 2009)
-# shorter qld series
-var.qld4_7y <- var.qld4 %>%
-  filter(trd <= 372)
-# shorter series to 2010
-var.qld4_8y <- var.qld4 %>%
-  filter(trd <= 384)
 
 ## run model for all periods
-   # coef.rain30.qld1 <- lapply(2:ncol(qld_30),
-   #                            function(x) rain.trend(x,qld_30,var.qld1))
-coef.rain30.qld2 <- lapply(2:ncol(qld_30),
-                            function(x) rain.trend(x,qld_30,var.qld2))
-   # coef.rain30.qld3 <- lapply(2:ncol(qld_30),
-   #                            function(x) rain.trend(x,qld_30,var.qld3))
-coef.rain30.qld4 <- lapply(2:ncol(qld_30),
-                             function(x) rain.trend(x,qld_30,var.qld4))
-# QUEENSLAND 7 yr series
-   coef.rain30.qld4_7y <- lapply(2:ncol(qld_30),
-                              function(x) rain.trend(x,qld_30,var.qld4_7y))
-# and qld 8 year series
-   coef.rain30.qld4_8y <- lapply(2:ncol(qld_30),
-                             function(x) rain.trend(x,qld_30,var.qld4_8y))
+## QLD
+coef.rain30.qld2 <- lapply(2:ncol(rainQld_t),
+                    function(x) rain.trend(x,rainQld_t,var.qld2))
+coef.rain30.qld4 <- lapply(2:ncol(rainQld_t),
+                       function(x) rain.trend(x,rainQld_t,var.qld4))
 # NSW/VIC
-  coef.rain30.nswvic <- lapply(2:ncol(nsw_30),
-                               function(x) rain.trend(x,nsw_30,var.data))
-# # 5 years
-# coef.rain30.nswvic5 <- lapply(2:ncol(nsw_30),
-#                               function(x) rain.trend(x,nsw_30,var.data5))
-# 7 years
- coef.rain30.nswvic7 <- lapply(2:ncol(nsw_30),
-                               function(x) rain.trend(x,nsw_30,var.data7))
-# 8 years (includes 2010)
-coef.rain30.nswvic8 <- lapply(2:ncol(nsw_30),
-                              function(x) rain.trend(x,nsw_30,var.data8))
-# # 9 years
-# coef.rain30.nswvic9 <- lapply(2:ncol(nsw_30),
-#                               function(x) rain.trend(x,nsw_30,var.data9))
-
+coef.rain30.nswvic <- lapply(2:ncol(rainNSW_t),
+                   function(x) rain.trend(x,rainNSW_t,var.data))
+```
+
+```{r ExtractMDB, eval=F, echo=F}
+## extract sign of MDB, which is assumed to have a linear relationship with
+## the rainfall in the area
+# Qld
+MDB_qld <- unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$p.coeff[2],NA)))
+# NSW/VIC
+MDB_nswvic <- unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$p.coeff[2],NA)))
+
+par(mfrow=c(2,1))
+hist(MDB_nswvic, xlab = "value of MDB rainfall estimate", main="NSW full data period")
+hist(MDB_qld, xlab = "value of MDB rainfall estimate", main="Qld full data period")
+
+```
+
+```{r ExtractC, eval=F, echo=F}
 ## extract sign of C (the tree cover effect)
-  C30_qld <- unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$p.coeff[2],NA)))
-# shorter series QLD (includes 2009)
-  C30_qld_7y <- unlist(lapply(1:length(coef.rain30.qld4_7y),function(x) ifelse(summary(coef.rain30.qld4_7y)[x,2]=="gam",summary(coef.rain30.qld4_7y[[x]])$p.coeff[2],NA)))
-# shorter series QLD (includes 2010)
-  C30_qld_8y <- unlist(lapply(1:length(coef.rain30.qld4_8y),function(x) ifelse(summary(coef.rain30.qld4_8y)[x,2]=="gam",summary(coef.rain30.qld4_8y[[x]])$p.coeff[2],NA)))
+C30_qld <- unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$p.coeff[3],NA)))
 # NSW/VIC
-C30_nswvic <- unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$p.coeff[2],NA)))
+C30_nswvic <- unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$p.coeff[3],NA)))
 # NSW/VIC 7 year (includes 2009)
- C30_nswvic7 <- unlist(lapply(1:length(coef.rain30.nswvic7),function(x) ifelse(summary(coef.rain30.nswvic7)[x,2]=="gam",summary(coef.rain30.nswvic7[[x]])$p.coeff[2],NA)))
-# NSW/VIC 8 year (includes 2010)
- C30_nswvic8 <- unlist(lapply(1:length(coef.rain30.nswvic8),function(x) ifelse(summary(coef.rain30.nswvic8)[x,2]=="gam",summary(coef.rain30.nswvic8[[x]])$p.coeff[2],NA)))
+#  C30_nswvic7 <- unlist(lapply(1:length(coef.rain30.nswvic7),function(x) ifelse(summary(coef.rain30.nswvic7)[x,2]=="gam",summary(coef.rain30.nswvic7[[x]])$p.coeff[2],NA)))
+# # NSW/VIC 8 year (includes 2010)
+#  C30_nswvic8 <- unlist(lapply(1:length(coef.rain30.nswvic8),function(x) ifelse(summary(coef.rain30.nswvic8)[x,2]=="gam",summary(coef.rain30.nswvic8[[x]])$p.coeff[2],NA)))
 
 jpeg("../../figures/hist_sign_LC.jpg")
-par(mfrow=c(2,2))
-hist(C30_nswvic)
-hist(C30_nswvic7)
-hist(C30_nswvic8)
-dev.off()
-# also for long term trend in nsw
-jpeg("../../figures/hist_trend_nsw.jpg")
-par(mfrow=c(2,2))
-hist(unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$p.coeff[3],NA))), xlab = "value of trend estimate", main="full data period")
-hist(unlist(lapply(1:length(coef.rain30.nswvic7),function(x) ifelse(summary(coef.rain30.nswvic7)[x,2]=="gam",summary(coef.rain30.nswvic7[[x]])$p.coeff[3],NA))), xlab = "value of trend estimate", main="Up to 2009")
-hist(unlist(lapply(1:length(coef.rain30.nswvic8),function(x) ifelse(summary(coef.rain30.nswvic8)[x,2]=="gam",summary(coef.rain30.nswvic8[[x]])$p.coeff[3],NA))), xlab = "value of trend estimate", main="Up to 2010")
-dev.off()
-# shows trds are all positive and small
-
-# C30_nswvic5 <- unlist(lapply(1:length(coef.rain30.nswvic5),function(x) ifelse(summary(coef.rain30.nswvic5)[x,2]=="gam",summary(coef.rain30.nswvic5[[x]])$p.coeff[2],NA)))
-# C30_nswvic9 <- unlist(lapply(1:length(coef.rain30.nswvic9),function(x) ifelse(summary(coef.rain30.nswvic9)[x,2]=="gam",summary(coef.rain30.nswvic9[[x]])$p.coeff[2],NA)))
+par(mfrow=c(2,1))
+hist(unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$p.coeff[3],NA))), xlab = "value of tree cover estimate", main="NSW full data period")
+hist(unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$p.coeff[3],NA))), xlab = "value of tree cover estimate", main="Qld full data period")
+dev.off()
 
@@
    function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",
-                      summary(coef.rain30.qld4[[x]])$p.pv[2],NA)))
-# Cp30_qld2 <- unlist(lapply(1:length(coef.rain30.qld2),
-#    function(x) ifelse(summary(coef.rain30.qld2)[x,2]=="gam",
-#                       summary(coef.rain30.qld2[[x]])$p.pv[2],NA)))
-# Qld 7 year series
- Cp30_qld4_7y <- unlist(lapply(1:length(coef.rain30.qld4_7y),
-    function(x) ifelse(summary(coef.rain30.qld4_7y)[x,2]=="gam",
-                       summary(coef.rain30.qld4_7y[[x]])$p.pv[2],NA)))
-# Qld 8 year series
-Cp30_qld4_8y <- unlist(lapply(1:length(coef.rain30.qld4_8y),
-   function(x) ifelse(summary(coef.rain30.qld4_8y)[x,2]=="gam",
-                      summary(coef.rain30.qld4_8y[[x]])$p.pv[2],NA)))
+                      summary(coef.rain30.qld4[[x]])$p.pv[3],NA)))
 # NSW all data
@@
     function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",
-                       summary(coef.rain30.nswvic[[x]])$p.pv[2],
+                       summary(coef.rain30.nswvic[[x]])$p.pv[3],
                        NA)))
-# NSW 7 year series
- Cp30_nswvic7 <- unlist(lapply(1:length(coef.rain30.nswvic7),
-               function(x) ifelse(summary(coef.rain30.nswvic7)[x,2]=="gam",
-                             summary(coef.rain30.nswvic7[[x]])$p.pv[2],NA)))
-# NSW 8 year series
-Cp30_nswvic8 <- unlist(lapply(1:length(coef.rain30.nswvic8),
-          function(x) ifelse(summary(coef.rain30.nswvic8)[x,2]=="gam",
-                             summary(coef.rain30.nswvic8[[x]])$p.pv[2],NA)))
+
+## frame of the two regions (100 years data)
+qld_tif <- raster("../..\\Data\\BoM\\rainfall\\AWAP\\from1900_FVG\\r19000101_qld.tif")
+nsw_tif <- raster("../..\\Data\\BoM\\rainfall\\AWAP\\from1900_FVG\\r19000101_nsw.tif")
+
+  dist_Cp30_qld4 <- raster(qld_tif)
+  dist_Cp30_qld4[] <- Cp30_qld4
+
+  dist_Cp30_nsw <- raster(nsw_tif)
+  dist_Cp30_nsw[] <- Cp30_nswvic
+
+# Now make the final plot for the paper with both sites and 2009 and all data
+  jpeg("../../figures/Cp_30yrs.jpg", width=1440, height=720)
+  par(mfrow=c(1,2))
+  par(mar=c(5,5,3,7))
+  # QLD first
+  plot(e111,xlab="",ylab="", cex.lab = 2,
+       cex.axis=2)
+  image(dist_Cp30_qld4,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
+#   plot(aus_reg2,border="grey",add=T)
+  plot(mdb_reg2,border="grey",lwd=5,add=T)
+  plot(e111,add=T)
+legend(148,-27.5,c("State","MDB"),col=c("grey","grey"),lwd=c(2,5),bty="n",title="Boundary",cex=1.5)
+
+  # Then NSW
+  par(mar=c(5,0,3,9),xpd=T)
+  plot(e222,xlab="",ylab="", cex.lab = 2,
+       cex.axis=2) #,border="white")
+  image(dist_Cp30_nsw,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
+  plot(aus_reg2,border="grey75",lwd=2,add=T)
+  plot(mdb_reg2,border="grey75",lwd=5,add=T)
+  plot(e222,add=T)
+  image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=1.2,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
+
+  dev.off()
+
+```
+
+```{r pvalues_allvars, echo=F, eval=F}
+
+# p-value of the linear variables
+pvalues_qld4 <- lapply(1:length(coef.rain30.qld4),
+   function(x) summary(coef.rain30.qld4[[x]])$p.pv[2:4])
+
+plinear_qld <- as.tibble(do.call(rbind,pvalues_qld4))
+colnames(plinear_qld) = c("MDB", "LC", "trd")
+plinear_qld$state <- "QLD"
+
+# NSW all data
+pvalues_nswvic <- lapply(1:length(coef.rain30.nswvic),
+    function(x) summary(coef.rain30.nswvic[[x]])$p.pv[2:4])
+
+plinear_nsw <- as.tibble(do.call(rbind,pvalues_nswvic))
+colnames(plinear_nsw) = c("MDB", "LC", "trd")
+plinear_nsw$state <- "NSW/VIC"
+
+# plotting
+p_plot_df <- rbind(plinear_qld,plinear_nsw)
+p_plot_df %>%
+  gather(key="Variable", value="p_value", MDB:trd) %>%
+  ggplot(aes(p_value,fill=state)) + geom_histogram(alpha=0.5) +
+  facet_wrap(~Variable) +
+  scale_fill_manual(values = c("blue","red")) +
+  theme(axis.title = element_text(size = rel(2)),
+        axis.text = element_text(size=rel(1)),
+        strip.text = element_text(face="bold", size= rel(2)),
+        legend.text = element_text(face="bold", size= rel(1)),
+        legend.title = element_text(face="bold", size= rel(2)))
+# produce also as jpeg
+jpeg("../../figures/pvalues_linear_fullmodel.jpg", height=960, width=960)
+p_plot_df %>%
+  gather(key="Variable", value="p_value", MDB:trd) %>%
+  ggplot(aes(p_value,fill=state)) + geom_histogram(alpha=0.5) +
+  facet_wrap(~Variable) +
+  scale_fill_manual(values = c("blue","red")) +
+  theme(axis.title = element_text(size = rel(2)),
+        axis.text = element_text(size=rel(1)),
+        strip.text = element_text(face="bold", size= rel(2)),
+        legend.text = element_text(face="bold", size= rel(1)),
+        legend.title = element_text(face="bold", size= rel(2)))
+dev.off()
+
+rm(p_plot_df)
 
@@
 p_qld <- as.tibble(p_qld)
-colnames(p_qld) = c("month", "SOI", "IOD", "NINO3.4", "NINO4")
+colnames(p_qld) = c("SOI", "IOD", "NINO3.4", "NINO4")
 p_qld$state <- "QLD"
@@
 p_nsw <- as.tibble(p_nsw)
-colnames(p_nsw) = c("month", "SOI", "IOD", "NINO3.4", "NINO4")
+colnames(p_nsw) = c("SOI", "IOD", "NINO3.4", "NINO4")
 p_nsw$state <- "NSW/VIC"
@@
 
-jpeg("../../figures/pvalues_fullmodel.jpg", height=960, width=960)
 p_plot_df %>%
-  gather(key="Variable", value="p_value", month:NINO4) %>%
+  gather(key="Variable", value="p_value", SOI:NINO4) %>%
   ggplot(aes(p_value,fill=state)) + geom_histogram(alpha=0.5) +
@@
         legend.title = element_text(face="bold", size= rel(2)))
+# produce also as jpg
+jpeg("../../figures/pvalues_fullmodel.jpg", height=960, width=960)
+p_plot_df %>%
+  gather(key="Variable", value="p_value", SOI:NINO4) %>%
+  ggplot(aes(p_value,fill=state)) + geom_histogram(alpha=0.5) +
+  facet_wrap(~Variable) +
+  scale_fill_manual(values = c("blue","red")) +
+  theme(axis.title = element_text(size = rel(2)),
+        axis.text = element_text(size=rel(1.5)),
+        strip.text = element_text(face="bold", size= rel(2)),
+        legend.text = element_text(face="bold", size= rel(1.5)),
+        legend.title = element_text(face="bold", size= rel(2)))
 dev.off()
@@
 
-## extract p value of long term trend
-trdp30_qld4 <- unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$p.pv[3],NA)))
-trdp30_nswvic <- unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$p.pv[3],NA)))
-
-# show histogram
-plot_df <- data_frame(p_trd = c(trdp30_qld4,trdp30_nswvic),
-                      state = c(rep("QLD", length(trdp30_qld4)),
-                                rep("NSW/VIC", length(trdp30_nswvic))))
-
-jpeg("../../figures/pvalue_trd.jpg", height=480, width=480)
-plot_df %>%
-  ggplot(aes(p_trd, fill=state)) + geom_histogram() +
-  scale_fill_manual(values = c("blue","red")) +
-  xlab("pvalue trend of the full gam model")
-dev.off()
-
+```
+
+```{r value_rsqModel, eval=F, echo=F}
 ## extract r.sq of the model
@@
   scale_fill_manual(values = c("blue","red")) +
-  xlab("r_squared of the full gam model")
+  xlab(expression(paste("% of variability explained (",r^2,")")))
 dev.off()
@@
 
-## frame of the two regions (100 years data)
-qld_tif <- raster("../..\\Data\\BoM\\rainfall\\AWAP\\from1900_FVG\\r19000101_qld.tif")
-nsw_tif <- raster("../..\\Data\\BoM\\rainfall\\AWAP\\from1900_FVG\\r19000101_nsw.tif")
-
-  dist_Cp30_qld4 <- raster(qld_tif)
-  dist_Cp30_qld4[] <- Cp30_qld4
-  dist_Cp30_qld4_8y <- raster(qld_tif)
-  dist_Cp30_qld4_8y[] <- Cp30_qld4_8y
-  dist_Cp30_qld4_7y <- raster(qld_tif)
-  dist_Cp30_qld4_7y[] <- Cp30_qld4_7y
-
-  jpeg("../../figures/Cp_30yrs_qld.jpg", width=1440, height=720)
-  par(mfrow=c(1,2))
-  par(mar=c(3,3,1,7),xpd=T)
-  plot(e111,xlab="",ylab="", cex.lab = 2,
-       cex.axis=2)
-  image(dist_Cp30_qld4,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
-#   plot(aus_reg2,border="grey",add=T)
-  plot(mdb_reg2,border="grey",lwd=3,add=T)
-  plot(e111,add=T)
-  image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
-legend(148.7,-27.9,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=1)
-
-  par(mar=c(3,3,1,7),xpd=T)
-  plot(e111,xlab="",ylab="", cex.lab = 2,
-       cex.axis=2)
-  image(dist_Cp30_qld4_7y,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
-#   plot(aus_reg2,border="grey",add=T)
-  plot(mdb_reg2,border="grey",lwd=3,add=T)
-  plot(e111,add=T)
-
-  # this is a totally blue plot
-#   par(mar=c(3,3,1,7),xpd=T)
-#   plot(e111,xlab="",ylab="", cex.lab = 2,
-#        cex.axis=2)
-#   image(dist_Cp30_qld4_8y,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
-# #   plot(aus_reg2,border="grey",add=T)
-#   plot(mdb_reg2,border="grey",lwd=3,add=T)
-#   plot(e111,add=T)
-#   image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
-# legend(148.7,-27.9,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=1)
-  dev.off()
-
-    #   dist_Cp30_qld <- crop(reproject(dist_Cp100_qld,CRS=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84"),program="raster"),e1)
-  dist_Cp30_nsw <- raster(nsw_tif)
-  dist_Cp30_nsw[] <- Cp30_nswvic
-  dist_Cp30_nsw_7y <- raster(nsw_tif)
-  dist_Cp30_nsw_7y[] <- Cp30_nswvic7
-  dist_Cp30_nsw_8y <- raster(nsw_tif)
-  dist_Cp30_nsw_8y[] <- Cp30_nswvic8
-#   dist_Cp30_nsw <- crop(reproject(dist_Cp30_nsw,CRS=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84"),program="raster"),e2)
-#
-  jpeg("../../figures/Cp_30yrs_nswvic.jpg", width=1440, height=720)
-  par(mfrow=c(1,3))
-  par(mar=c(3,3,1,7),xpd=T)
-  plot(e222,xlab="",ylab="", cex.lab = 2,
-       cex.axis=2) #,border="white")
-  image(dist_Cp30_nsw_8y,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
-  plot(aus_reg2,border="grey75",lwd=5,add=T)
-  plot(mdb_reg2,border="grey75",lwd=5,add=T)
-  plot(e222,add=T)
-  image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
-  #dev.off()
-
-  par(mar=c(3,3,1,7),xpd=T)
-  plot(e222,xlab="",ylab="", cex.lab = 2,
-       cex.axis=2) #,border="white")
-  image(dist_Cp30_nsw_7y,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
-  plot(aus_reg2,border="grey75",lwd=5,add=T)
-  plot(mdb_reg2,border="grey75",lwd=5,add=T)
-  plot(e222,add=T)
-  # # image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
-  # dev.off()
-
-   # jpeg("../../figures/Cp_30yrs_nswvic_8y.jpg", width=960, height=960)
-  par(mar=c(3,3,1,7),xpd=T)
-  plot(e222,xlab="",ylab="", cex.lab = 2,
-       cex.axis=2) #,border="white")
-  image(dist_Cp30_nsw,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
-  plot(aus_reg2,border="grey75",lwd=5,add=T)
-  plot(mdb_reg2,border="grey75",lwd=5,add=T)
-  plot(e222,add=T)
-  dev.off()
-#
-
-# Now make the final plot for the paper with both sites and 2009 and all data
-  jpeg("../../figures/Cp_30yrs.jpg", width=1440, height=1440)
-  par(mfrow=c(2,2))
-  # NSW first
-  par(mar=c(3,3,1,9),xpd=T)
-  plot(e222,xlab="",ylab="", cex.lab = 2,
-       cex.axis=2) #,border="white")
-  image(dist_Cp30_nsw_7y,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
-  plot(aus_reg2,border="grey75",lwd=5,add=T)
-  plot(mdb_reg2,border="grey75",lwd=5,add=T)
-  plot(e222,add=T)
-  image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
-legend(149,-36.6,c("State","MDB"),col=c("grey","grey"),
-       lwd=c(1,3),bty="n",title="Boundary",cex=1.8)
-  #dev.off()
-
-  par(mar=c(3,5,1,7),xpd=T)
-  plot(e222,xlab="",ylab="", cex.lab = 2,
-       cex.axis=2) #,border="white")
-  image(dist_Cp30_nsw,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
-  plot(aus_reg2,border="grey75",lwd=5,add=T)
-  plot(mdb_reg2,border="grey75",lwd=5,add=T)
-  plot(e222,add=T)
-
-  # Then qld
-  par(mar=c(3,3,1,9),xpd=T)
-  plot(e111,xlab="",ylab="", cex.lab = 2,
-       cex.axis=2)
-  image(dist_Cp30_qld4_7y,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
-#   plot(aus_reg2,border="grey",add=T)
-  plot(mdb_reg2,border="grey",lwd=3,add=T)
-  plot(e111,add=T)
-  image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
-
-  par(mar=c(3,5,1,7),xpd=T)
-  plot(e111,xlab="",ylab="", cex.lab = 2,
-       cex.axis=2)
-  image(dist_Cp30_qld4,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
-#   plot(aus_reg2,border="grey",add=T)
-  plot(mdb_reg2,border="grey",lwd=3,add=T)
-  plot(e111,add=T)
-
-  dev.off()
-# plot up to 2010 for the appendix
-  jpeg("../../figures/Cp_30yrs_8y_appendix.jpg", width=1440, height=720)
-  par(mfrow=c(1,2))
-  # NSW first
-  par(mar=c(3,3,1,9),xpd=T)
-  plot(e222,xlab="",ylab="", cex.lab = 2,
-       cex.axis=2) #,border="white")
-  image(dist_Cp30_nsw_8y,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
-  plot(aus_reg2,border="grey75",lwd=5,add=T)
-  plot(mdb_reg2,border="grey75",lwd=5,add=T)
-  plot(e222,add=T)
-  image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
-legend(149,-36.6,c("State","MDB"),col=c("grey","grey"),
-       lwd=c(1,3),bty="n",title="Boundary",cex=1.8)
-  #dev.off()
-
-  par(mar=c(3,5,1,7),xpd=T)
-  plot(e222,xlab="",ylab="", cex.lab = 2,
-       cex.axis=2) #,border="white")
-  image(dist_Cp30_qld_8y,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
-  plot(aus_reg2,border="grey75",lwd=5,add=T)
-  plot(mdb_reg2,border="grey75",lwd=5,add=T)
-  plot(e222,add=T)
-dev.off()
-
-# ## range of predicted values
-  # predictmax_qld30 <- unlist(lapply(1:length(coef.rain30.qld4), function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam", max(coef.rain30.qld4[[x]]$fitted.values),NA)))
-  # predictmin_qld30 <- unlist(lapply(1:length(coef.rain30.qld4), function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam", min(coef.rain30.qld4[[x]]$fitted.values),NA)))
+```
+
+```{r GAMresiduals, echo=F, eval=F}
 #
@@
                           QLD = qld.resid.mean)
-#
-# predictmax_nsw30 <- unlist(lapply(1:length(coef.rain30.nswvic), function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam", max(coef.rain30.nswvic[[x]]$fitted.values),NA)))
-# # 70 - 254.7
-#   predictmin_nsw30 <- unlist(lapply(1:length(coef.rain30.nswvic), function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam", min(coef.rain30.nswvic[[x]]$fitted.values),NA)))
-# # 18.4 - 62.2
 #
@@
 #
-# ## R^2 of the regression
-#     rsq_qld30 <- unlist(lapply(1:length(coef.rain30.qld), function(x) ifelse(summary(coef.rain30.qld)[x,2]=="gam", summary(coef.rain30.qld[[x]])$r.sq,NA)))
-#     rsq_nswvic30 <- unlist(lapply(1:length(coef.rain30.nswvic), function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam", summary(coef.rain30.nswvic[[x]])$r.sq,NA)))
-#
-# ## significant of the other two variables
-#     v1_qld30 <- unlist(lapply(1:length(coef.rain30.qld), function(x) ifelse(summary(coef.rain30.qld)[x,2]=="gam", summary(coef.rain30.qld[[x]])$s.pv[1],NA)))
-#     v1_nswvic30 <- unlist(lapply(1:length(coef.rain30.nswvic), function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam", summary(coef.rain30.nswvic[[x]])$s.pv[1],NA)))
 #
@@
 
-To support the regression analysis, a step trend test was used to detect changes in rainfall as a result of vegetation cover change.
-This nonparametric statistical test was modified from the Mann-Whitney Rank-Sum test by [@Hirsch1985] and can identify a step change in data which is cross correlated.
+To support the regression analysis, a Mann-Whitney Rank-Sum step trend test was used to detect changes in rainfall as a result of vegetation cover change.
+This specific nonparametric statistical test was modified from the Mann-Whitney U test by @Hirsch1985 and can identify a step change in data which is cross-correlated.
 The gridded rainfall dataset used in this study has a high spatial correlation between neighbouring pixels due to the underlying interpolation method.
-The advantages of using the Rank-Sum test therefore are: (1) it does not depend on assumptions of the data distribution; (2) it is not restricted to datasets with no missing data; (3) it is robust and not as easily influenced by outliers and negative numbers [@Hirsch1985].
+The advantages of using the Rank-Sum test are: (1) it does not depend on assumptions of the data distribution; (2) it is not restricted to datasets with no missing data; (3) it is robust and not as easily influenced by outliers and negative numbers [@Hirsch1985].
 
 The rainfall residuals from the regression model in Equation \@ref(eq:model2) were used for the step trend test.
-According to [@Hirsch1985], to detect a step change, using deseasonalised and detrended data is important.
-Furthermore, as rainfall can only be partially attributed to local sources and conditions, other effects are introduced by large scale dynamics and changes in other climatic factors.
+To detect a step change, using deseasonalised and detrended data is important [@Hirsch1985].
+Furthermore, as rainfall can only be partially attributed to local sources and conditions, other effects are introduced by large scale dynamics and changes in climatic factors.
 The assumption is that the regression model should remove these effects, and additionally deseasonalise and detrend the rainfall data.
-As a result, the local landuse effects are amplified in the variation of the remaining residuals.
+As a result, the local landuse effects are amplified in the remaining residuals.
 The  test, described in the following sub section, subsequently associates trends in the rainfall residuals with tree cover changes.
@@
 ```{r Residual_Analysis, eval=F, echo=F}
-## GAM without land cover variable for step trend test
+## predicting model: land cover is category variable
 rain.trend3 <- function(x,y,var.data) {
-
-  var.data$rain <- as.vector(unlist(y[x]))+0.0001
-
-  #if (!is.na(mean(var.data$rain))) {
-    # if (grepl("qld",substitute(y))) {
-      rain_fit <- gam(rain ~ trd
-                      + s(month, k=3, bs="cs")# added month back in
+  # add a very small number to take out 0 values
+  var.data$rain <- as.vector(unlist(y[x]))[1:nrow(var.data)]+0.0001
+
+      rain_fit <- gam(rain ~ MDB
                   + s(SOI, k=3, bs="cs")
@@
                   + s(NINO4, k=3, bs="cs")
-                    , data=var.data,family=Gamma(link=log),
-                  na.action=na.omit)
-    # } else
-    # if (grepl("nsw",substitute(y))) {
-    #   rain_fit <- gam(rain ~ trd
-    #               + s(month,k=3)
-    #               + s(SOI, k=3)
-    #               + s(IOD, k=3)
-    #               , data=var.data,family=Gamma(link=log))
-    #}
-
+                  + trd
+                  , data=var.data,#family=Gamma(link=log),
+                  na.action = na.omit)
+    #
   return(rain_fit)
-  #} #else {NA}
 
 }
-#
-# ## change LC variable to be null
-#   var.data <- data.frame(month, trd, SOI = SOI_30a, IOD = IOD_30a)
-#
+
 ## run model for 37 years
-  coef.rain.qld <- lapply(2:ncol(qld_30),
-                          function(x) rain.trend3(x,qld_30,var.qld4))
-  coef.rain.qld2 <- lapply(2:ncol(qld_30),
-                          function(x) rain.trend3(x,qld_30,var.qld2))
-  # QLD shorter series
-  test <- qld_30[1:372,]
-  coef.rain.qld_7y <- lapply(2:ncol(qld_30), function(x) rain.trend3(x,test,var.qld4_7y))
-rm(test)
+mod_rain_qld4 <- lapply(2:ncol(rainQld_t),
+                       function(x) rain.trend3(x,rainQld_t,var.qld4))
+mod_rain_qld2 <- lapply(2:ncol(rainQld_t),
+                  function(x) rain.trend3(x,rainQld_t,var.qld2))
 ## QLD
-  rsq_qld <- unlist(lapply(1:length(coef.rain.qld), function(x) ifelse(summary(coef.rain.qld)[x,2]=="gam", summary(coef.rain.qld[[x]])$r.sq,NA)))
+rsq_qld <- unlist(lapply(1:length(mod_rain_qld4), 
+                    function(x) summary(mod_rain_qld4[[x]])$r.sq))
 #
@@
 ## extract residual
-  residual_qld <- lapply(1:length(coef.rain.qld), function(x) if(summary(coef.rain.qld)[x,2]=="gam") {coef.rain.qld[[x]]$y-fitted(coef.rain.qld[[x]])} else {NA})
-
-  residual_qld2 <- lapply(1:length(coef.rain.qld2), function(x) if(summary(coef.rain.qld2)[x,2]=="gam") {coef.rain.qld2[[x]]$y-fitted(coef.rain.qld2[[x]])} else {NA})
-
-# shorter Qld series
-    residual_qld_7y <- lapply(1:length(coef.rain.qld_7y), function(x) if(summary(coef.rain.qld_7y)[x,2]=="gam") {coef.rain.qld_7y[[x]]$y-fitted(coef.rain.qld_7y[[x]])} else {NA})
+residual_qld <- lapply(1:length(mod_rain_qld4), 
+      function(x) {mod_rain_qld4[[x]]$y-fitted(mod_rain_qld4[[x]])})
+
+residual_qld2 <- lapply(1:length(mod_rain_qld2), 
+    function(x) {mod_rain_qld2[[x]]$y-fitted(mod_rain_qld2[[x]])})
 
@@
 #NSW/VIC
-  coef.rain.nswvic <- lapply(2:ncol(nsw_30), function(x) rain.trend3(x,nsw_30,var.data))
-  test <- nsw_30[1:372,]
-  coef.rain.nswvic7 <- lapply(2:ncol(nsw_30), function(x) rain.trend3(x,test,var.data7))
-  rm(test)
-  # coef.rain.nswvic5 <- lapply(2:ncol(nsw_30), function(x) rain.trend3(x,nsw_30[1:348,],30,var.data5))
-  # coef.rain.nswvic9 <- lapply(2:ncol(nsw_30), function(x) rain.trend3(x,nsw_30[1:396,],30,var.data9))
+mod_rain_nswvic <- lapply(2:ncol(rainNSW_t), 
+                function(x) rain.trend3(x,rainNSW_t,var.data))
 
 ## explaining power of the partial model
-  rsq_nswvic <- unlist(lapply(1:length(coef.rain.nswvic), function(x) ifelse(summary(coef.rain.nswvic)[x,2]=="gam", summary(coef.rain.nswvic[[x]])$r.sq,NA)))
+rsq_nswvic <- unlist(lapply(1:length(mod_rain_nswvic), 
+              function(x) summary(mod_rain_nswvic[[x]])$r.sq))
 #
 
-#   predictmax_nswvic <- unlist(lapply(1:length(coef.rain.nswvic), function(x) ifelse(summary(coef.rain.nswvic)[x,2]=="gam", max(coef.rain.nswvic[[x]]$fitted.values),NA)))
-#   predictmin_nswvic <- unlist(lapply(1:length(coef.rain.nswvic), function(x) ifelse(summary(coef.rain.nswvic)[x,2]=="gam", min(coef.rain.nswvic[[x]]$fitted.values),NA)))
-
 # extract residual
-residual_nswvic <- lapply(1:length(coef.rain.nswvic), function(x) if(summary(coef.rain.nswvic)[x,2]=="gam") {coef.rain.nswvic[[x]]$y-fitted(coef.rain.nswvic[[x]])} else {NA})
-# 7 year (2003 - 2010)
-residual_nswvic7 <- lapply(1:length(coef.rain.nswvic7), function(x) if(summary(coef.rain.nswvic7)[x,2]=="gam") {coef.rain.nswvic7[[x]]$y-fitted(coef.rain.nswvic7[[x]])} else {NA})
-
-# 5 year
-# residual_nswvic5 <- lapply(1:length(coef.rain.nswvic5), function(x) if(summary(coef.rain.nswvic5)[x,2]=="gam") {coef.rain.nswvic5[[x]]$y-fitted(coef.rain.nswvic5[[x]])} else {NA})
-
-# 9 year
-# residual_nswvic9 <- lapply(1:length(coef.rain.nswvic9), function(x) if(summary(coef.rain.nswvic9)[x,2]=="gam") {coef.rain.nswvic9[[x]]$y-fitted(coef.rain.nswvic9[[x]])} else {NA})
+residual_nswvic <- lapply(1:length(mod_rain_nswvic),
+  function(x) {mod_rain_nswvic[[x]]$y-fitted(mod_rain_nswvic[[x]])})
 
@@
 
+# # create a boxplot with the annual residuals from 1998
+# resplot_df <- data_frame(Date = rep(ymd(paste(nsw_30[[1]],
+#                                               "01", sep="-")),
+#                         length(residual_nswvic)+length(residual_qld)),
+#                           residuals = c(unlist(residual_nswvic),
+#                                                   unlist(residual_qld)),
+#                       State = c(rep("NSW/VIC",
+#                                     length(unlist(residual_nswvic))),
+#                                   rep("QLD", length(unlist(residual_qld)))))
+# 
+# jpeg("../../figures/ResidualBoxplot.jpg")
+# resplot_df %>%
+#   filter(Date >= ymd("1998-01-01")) %>%
+#   ggplot(aes(Date,residuals, fill=State)) +
+#   geom_boxplot(aes(group=year(Date))) +
+#   facet_wrap(~State, ncol=1)
+# dev.off()
+
+# residuals before and after change
 # create a boxplot with the annual residuals from 1998
-resplot_df <- data_frame(Date = rep(ymd(paste(nsw_30[[1]],
-                                              "01", sep="-")),
-                        length(residual_nswvic)+length(residual_qld)),
-                          residuals = c(unlist(residual_nswvic),
-                                                  unlist(residual_qld)),
+resplot_df <- data_frame(Date = rep(ymd(rainNSW_t[[1]][1:444]),
+          length(residual_nswvic) + length(residual_qld)),
+          residuals = c(unlist(residual_nswvic),                                               unlist(residual_qld)),
           State = c(rep("NSW/VIC",
@@
                         rep("QLD", length(unlist(residual_qld)))))
+# resplot_df %>%
+#   filter(Date >= ymd("1998-01-01")) %>%
+#   ggplot(aes(Date,residuals, fill=State)) +
+#   geom_boxplot(aes(group=year(Date))) +
+#   facet_wrap(~State, ncol=1)
 
@@
 #par(mfrow=c(1,2))
-jpeg("../../figures/ResidualBoxplotchange.jpg")
-p1 <- resplot_df %>%
+jpeg("figures/ResidualBoxplotchange.jpg")
+resplot_df %>%
   filter(Date >= ymd("1998-01-01")) %>%
@@
   geom_boxplot(aes(group=Change)) + facet_wrap(~State, ncol=1) +
-  guides(fill=FALSE) + ggtitle("1998 - 2015")
-#dev.off()
-
-#jpeg("../../figures/ResidualBoxplotchange2.jpg")
-p2 <- resplot_df %>%
-  filter(Date >= ymd("1998-01-01")) %>%
-  mutate(Change = ifelse(year(Date) > 2003, "After", "Before")) %>%
-  filter(year(Date) < 2010) %>%
-  ggplot(aes(Change,residuals, fill = State)) +
-  geom_boxplot(aes(group=Change)) + facet_wrap(~State, ncol=1) +
-  ylab(label="") + ggtitle("1998 - 2009")
-lay <- rbind(c(1,1,2,2,2))
-grid.arrange(p1,p2, layout_matrix = lay)
-dev.off()
-
+  guides(fill=FALSE) + ggtitle("whole period")
+dev.off()
+
+# QLD
 res2 <- resplot_df %>%
@@
   mutate(Change = ifelse(year(Date) > 2003, "After", "Before")) %>%
-  filter(year(Date) < 2010) %>%
+  #filter(year(Date) < 2010) %>%
   filter(State == "QLD")
@@
   t.test(formula=residuals~Change,data=res2)
-# No significant difference if measured to 2010 and slightly less rain (p < 0.05) before if measured to 2015 (~ 2 mm/month)
-
-# 2015
-#   data:  residuals by Change
-# t = 21.177, df = 513000, p-value < 2.2e-16
-# alternative hypothesis: true difference in means is not equal to 0
-# 95 percent confidence interval:
-#  1.863279 2.243363
-# sample estimates:
-#  mean in group After mean in group Before
-#            0.7206134           -1.3327075
-
-# 2009
-#   data:  residuals by Change
-# t = 1.6534, df = 414050, p-value = 0.09825
-# alternative hypothesis: true difference in means is not equal to 0
-# 95 percent confidence interval:
-#  -0.03314461  0.39066623
-# sample estimates:
-#  mean in group After mean in group Before
-#            -1.153947            -1.332708
-
-res2 <- resplot_df %>%
+#significant difference  and slightly less mean monthly residual rain (p < 0.05) after (~ 0.45 mm/month)
+
+# NSW/VIC
+res3 <- resplot_df %>%
   filter(Date >= ymd("1998-01-01")) %>%
   mutate(Change = ifelse(year(Date) > 2003, "After", "Before")) %>%
-  filter(year(Date) < 2010) %>%
-  filter(State != "QLD")
-
-  summary(lm(residuals~Change-1,data=res2))
-  t.test(formula=residuals~Change,data=res2)
-
-  # slightly more rain (p < 0.05) after the change if measured to 2015 (~ 1 mm/month), significantly less rain (p < 0.05) after the change (~ 9 mm/month) if measured to 2009
-# 2015
-#   data:  residuals by Change
-# t = 7.3033, df = 215950, p-value = 2.817e-13
-# alternative hypothesis: true difference in means is not equal to 0
-# 95 percent confidence interval:
-#  0.9041572 1.5674572
-# sample estimates:
-#  mean in group After mean in group Before
-#            -2.919095            -4.154902
-
-# 2009
-#   data:  residuals by Change
-# t = -55.329, df = 180240, p-value < 2.2e-16
-# alternative hypothesis: true difference in means is not equal to 0
-# 95 percent confidence interval:
-#  -10.144149  -9.450039
-# sample estimates:
-#  mean in group After mean in group Before
-#           -13.951996            -4.154902
+  #filter(year(Date) < 2010) %>%
+  filter(State == "NSW/VIC")
+
+  summary(lm(residuals~Change-1,data=res3))
+  t.test(formula=residuals~Change,data=res3)
+#significant difference  and less mean monthly residual rain (p < 0.05) after (~ 1.5 mm/month)
 
@@
 
-  var.data$rain <- as.vector(unlist(y[x]))+0.0001
-
-  #if (!is.na(mean(var.data$rain))) {
-#    if (grepl("qld",substitute(y))) {
-      rain_fit <- gam(rain ~  s(SOI, k=3, bs="cs")
+  var.data$rain <- as.vector(unlist(y[x]))[1:nrow(var.data)]+0.0001
+
+      rain_fit <- gam(rain ~
+                  + s(SOI, k=3, bs="cs")
                   + s(IOD, k=3, bs="cs")
@@
                   + s(NINO4, k=3, bs="cs")
-                  , data=var.data,family=Gamma(link=log))
-    # } else
-    # if (grepl("nsw",substitute(y))) {
-    #   rain_fit <- gam(rain ~ s(SOI, k=3)
-    #               + s(IOD, k=3)
-    #               , data=var.data,family=Gamma(link=log))
-    # }
+                  #+ trd
+                  , data=var.data,#family=Gamma(link=log),
+                  na.action = na.omit)
     #
@@
 ## explaining power of index
-  coef.soi.qld <- lapply(2:ncol(qld_30), function(x) rain.trend4(x,qld_30,var.qld4))
-  coef.soi.nswvic <- lapply(2:ncol(nsw_30), function(x) rain.trend4(x,nsw_30,var.data))
-  rsq_soi_qld <- unlist(lapply(1:length(coef.soi.qld), function(x) ifelse(summary(coef.soi.qld)[x,2]=="gam", summary(coef.soi.qld[[x]])$r.sq,NA)))
-  rsq_soi_nswvic <- unlist(lapply(1:length(coef.soi.nswvic), function(x) ifelse(summary(coef.soi.nswvic)[x,2]=="gam", summary(coef.soi.nswvic[[x]])$r.sq,NA)))
+  coef.soi.qld <- lapply(2:ncol(rainQld_t), 
+                         function(x) rain.trend4(x,rainQld_t,var.qld4))
+  coef.soi.nswvic <- lapply(2:ncol(rainNSW_t), 
+                            function(x) rain.trend4(x,rainNSW_t,var.data))
+  rsq_soi_qld <- unlist(lapply(1:length(coef.soi.qld), 
+                         function(x) ifelse(summary(coef.soi.qld)[x,2]=="gam",
+                                            summary(coef.soi.qld[[x]])$r.sq,NA)))
+  rsq_soi_nswvic <- unlist(lapply(1:length(coef.soi.nswvic), 
+                            function(x) ifelse(summary(coef.soi.nswvic)[x,2]=="gam",
+                                         summary(coef.soi.nswvic[[x]])$r.sq,NA)))
 
@@
 
-As indicated, the step trend test is a modified version of the Mann-Whitney rank-sum statistic [@Hirsch1985].
+As indicated, the step trend test is a modified version of the Mann-Whitney U statistic [@Hirsch1985].
 As a nonparametric rank-based test, the Mann-Whitney test does not use the exact values of rainfall but depends on the ranks of the data.
@@
 Following @Hirsch1985, the before period was set to five years (1998 - 2002) in both regions.
-The length of the after period is difficult for the NSW/VIC as the regrowth would at some point have impacted the local effects.
-
-There is a further complication in the data.
-The year 2009 was the final year of the millenium drought [@vanDijk2013], followed by two very wet years (also visible in Figure \@ref(fig:ts-mean) in 2010 and 2011.
-As highlighted earlier, this means the the length of after period could influence the analysis by the inclusion of more or fewer years.
-as a result the period up to (and including) 2009 was compared to the full period up to 2015.
-Further analysis up to 2010 is included in the supplementary material.
+The length of the after period is also difficult for the NSW/VIC region, as the regrowth would at some point have impacted the local effects.
 
@@
 
-The null hypothesis ($H_0$) in this study is that there was no change in rainfall due to land surface intervention.
-The results of the step trend test can be interpreted according to the sign of the Z' score (see Table \@ref(tab:Zscore) [Chapter 23, P887 @Hipel1994]).
-Z' is normally distributed similar to the standard normal statistics Z.
-Hence it can be compared to a standard normal distribution to determine the p value.
-
 ```{r (ref:Zscore), echo=F}
@@
 ```
+
+The null hypothesis ($H_0$) in this study is that there was no change in rainfall due to land surface intervention.
+The results of the step trend test can be interpreted according to the sign of the Z' score (see Table \@ref(tab:Zscore), Chapter 23, P887 [@Hipel1994]).
+Z' is normally distributed similar to the standard normal statistics Z.
+Hence it can be compared to a standard normal distribution to determine the p value.
 
@@
     y.2 <- y[[x+1]]
-#     y.3 <- y[[x+2]]
     y.3 <- y[[x+z]]
     y.4 <- y[[x+z+1]]
-#     y.6 <- y[[x+z+2]]
-#     y.7 <- y[[x+2*z]]
-#     y.8 <- y[[x+2*z+1]]
-#     y.9 <- y[[x+2*z+2]]
-
-    if(!is.na(y.1[1])&!is.na(y.2[1])&!is.na(y.3[1])&!is.na(y.4[1])) { #|!is.na(y.5[1])|!is.na(y.6[1])|!is.na(y.7[1])|!is.na(y.8[1])|!is.na(y.9[1])) {
+
+    if(!is.na(y.1[1])&!is.na(y.2[1])&!is.na(y.3[1])&!is.na(y.4[1])) {
 
@@
                 start=c(1997,1),frequency=12)
-#       v.5 <- ts(y.5[c(229:288,313:348)],start=c(1998,1),frequency=12)
-#       v.6 <- ts(y.6[c(229:288,313:348)],start=c(1998,1),frequency=12)
-#       v.7 <- ts(y.7[c(229:288,313:348)],start=c(1998,1),frequency=12)
-#       v.8 <- ts(y.8[c(229:288,313:348)],start=c(1998,1),frequency=12)
-#       v.9 <- ts(y.9[c(229:288,313:348)],start=c(1998,1),frequency=12)
-#       v <- (x$y-x$fitted.values)[c(1:288,313:357)]
-#       v.s <- ts(v,start=c(1998,1),frequency=12)
 
@@
                 start=c(1997,1),frequency=12)
-#       v.5 <- ts(y.5[c(229:348)],start=c(1998,1),frequency=12)
-#       v.6 <- ts(y.6[c(229:348)],start=c(1998,1),frequency=12)
-#       v.7 <- ts(y.7[c(229:348)],start=c(1998,1),frequency=12)
-#       v.8 <- ts(y.8[c(229:348)],start=c(1998,1),frequency=12)
-#       v.9 <- ts(y.9[c(229:348)],start=c(1998,1),frequency=12)
-#       v <- (x$y-x$fitted.values)[c(1:288, 301:357)]
-#       v.s <- ts(v,start=c(1998,1),frequency=12)
 }
@@
     assign(paste("r4",j,sep=""),rank(subset(v.4,cycle(v.4)==j)))
-#     assign(paste("r5",j,sep=""),rank(subset(v.5,cycle(v.5)==j)))
-#     assign(paste("r6",j,sep=""),rank(subset(v.6,cycle(v.6)==j)))
-#     assign(paste("r7",j,sep=""),rank(subset(v.7,cycle(v.7)==j)))
-#     assign(paste("r8",j,sep=""),rank(subset(v.8,cycle(v.8)==j)))
-#     assign(paste("r9",j,sep=""),rank(subset(v.9,cycle(v.9)==j)))
 
@@
     rlist.4 <- list(r41,r42,r43,r44,r45,r46,r47,r48,r49,r410,r411,r412)
-#     rlist.5 <- list(r51,r52,r53,r54,r55,r56,r57,r58,r59,r510,r511,r512)
-#     rlist.6 <- list(r61,r62,r63,r64,r65,r66,r67,r68,r69,r610,r611,r612)
-#     rlist.7 <- list(r71,r72,r73,r74,r75,r76,r77,r78,r79,r710,r711,r712)
-#     rlist.8 <- list(r81,r82,r83,r84,r85,r86,r87,r88,r89,r810,r811,r812)
-#     rlist.9 <- list(r91,r92,r93,r94,r95,r96,r97,r98,r99,r910,r911,r912)
     rlist <- list(unlist(rlist.1),unlist(rlist.2),unlist(rlist.3),unlist(rlist.4)) #,unlist(rlist.5),unlist(rlist.6),unlist(rlist.7),unlist(rlist.8),unlist(rlist.9))
-## seasonal
-#     rlist <- list(unlist(rlist.1[c(9:11)]),unlist(rlist.2[c(9:11)]),unlist(rlist.3[c(9:11)]),unlist(rlist.4[c(9:11)]))
-#     rlist <- list(r1,r2,r12) ## summer
-#     rlist <- list(r3,r4,r5)  ## autumn
-#     rlist <- list(r6,r7,r8)  ## winter
-#     rlist <- list(r9,r10,r11) ## spring
-    w.s1 <- 0
-    w.s2 <- 0
-    w.s3 <- 0
-    w.s4 <- 0
-#     w.s5 <- 0
-#     w.s6 <- 0
-#     w.s7 <- 0
-#     w.s8 <- 0
-#     w.s9 <- 0
-  for (k in 1:length(rlist.1)) {
-#   for (k in c(9:11)) {
-    w.s1 <- sum(rlist.1[[k]][1:n1])+w.s1
-    w.s2 <- sum(rlist.2[[k]][1:n1])+w.s2
-    w.s3 <- sum(rlist.3[[k]][1:n1])+w.s3
-    w.s4 <- sum(rlist.4[[k]][1:n1])+w.s4
-#     w.s5 <- sum(rlist.5[[k]][1:n1])+w.s5
-#     w.s6 <- sum(rlist.6[[k]][1:n1])+w.s6
-#     w.s7 <- sum(rlist.7[[k]][1:n1])+w.s7
-#     w.s8 <- sum(rlist.8[[k]][1:n1])+w.s8
-#     w.s9 <- sum(rlist.9[[k]][1:n1])+w.s9
-    }
-    w.s <- sum(w.s1,w.s2,w.s3,w.s4) #,w.s5,w.s6,w.s7,w.s8,w.s9)
-
-## other statistic
-# mean for full year
-    mu <- n1*(n1+n2+1)/2
-    mu.s <- 12*4*mu
-# ## mean for one season
-#     mu <- n1*(n1+n2+1)/2
-#     mu.s <- 3*4*mu
-
-## variance (standard variation).
-The autocorrelation of the residuals can be assumed to be null (checked, cor.test(), true for NSW, mostly true for WA.
-QLD?
-there are some situations in WA require further investigation)
-# assume only cross correlation
-    ss <- n1*n2*(n1+n2+1)/12
-    ss.s <- 0
-    for (kk in 1:4) {
-      for (kkk in 1:4) {
-#         for (kkkk in 1:12) {
-          ss.s <- 12*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
-#         }
-      }
-    }
-## variance for one season
-#     ss <- n1*n2*(n1+n2+1)/3
-#     ss.s <- 0
-#     for (kk in 1:4) {
-#       for (kkk in 1:4) {
-#           ss.s <- 3*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
-#       }
-#     }
-
-## step trend statistic
-    z <- (w.s - mu.s)/sqrt(ss.s)
-
-    return(z) }
-
-     else {return(NA)} }
-
-     else {return(NA)}
-
-  }
-
-## 30 years
-## full year,
-## central QLD
-  qld.z.new <- unlist(lapply(1:(length(residual_qld)-57),function(x) stepTrend2(x,residual_qld,1)))
-  qld.z.new <- c(qld.z.new,rep(NA,57))
-# shorter series
-    qld.z.new_7y <- unlist(lapply(1:(length(residual_qld_7y)-57),function(x) stepTrend2(x,residual_qld_7y,1, endyear=2009)))
-  qld.z.new_7y <- c(qld.z.new_7y,rep(NA,57))
-
-  tr_qld.step.new <- raster(qld_tif)
-  tr_qld.step.new[] <- qld.z.new
-
-  tr_qld.step.new7 <- raster(qld_tif)
-  tr_qld.step.new7[] <- qld.z.new_7y
-
-## NSW
-  nswvic.z.new <- unlist(lapply(1:(length(residual_nswvic)-38),function(x) stepTrend2(x,residual_nswvic,2)))
-  nswvic.z.new <- c(nswvic.z.new,rep(NA,38))
-
-  nswvic.z.new7 <- unlist(lapply(1:(length(residual_nswvic7)-38),function(x) stepTrend2(x,residual_nswvic7,2, endyear = 2009)))
-  nswvic.z.new7 <- c(nswvic.z.new7,rep(NA,38))
-
-  tr_nswvic.step.new <- raster(nsw_tif)
-  tr_nswvic.step.new[] <- nswvic.z.new
-  tr_nswvic.step.new7 <- raster(nsw_tif)
-  tr_nswvic.step.new7[] <- nswvic.z.new7
-
-# ## seasonal
-#   qld.z.sum <- unlist(lapply(1:(length(residual_qld)-56),function(x) stepTrend2(x,residual_qld,1)))
-#   qld.z.sum <- c(qld.z.sum,rep(NA,56))
-#
-#   tr_qld.step.sum <- raster(qld_tif)
-#   tr_qld.step.sum[] <- qld.z.sum
-#
-#   nswvic.z.sum <- unlist(lapply(1:(length(residual_nswvic)-37),function(x) stepTrend2(x,residual_nswvic,2)))
-#   nswvic.z.sum <- c(nswvic.z.sum,rep(NA,37))
-#
-#   tr_nswvic.step.sum <- raster(nsw_tif)
-#   tr_nswvic.step.sum[] <- nswvic.z.sum
-
-  plotclt <- rev(brewer.pal(8,"RdYlBu"))
-
-jpeg("../../figures/step_new.jpg", width=1240, height=960, quality=100)
-#tiff("../figures/step_new_nswvic_7y.tif", width=960, height=960)
-  par(mar=c(3,3,1,9), mfrow=c(2,2), xpd=T)
-  plot(e222,xlab="",ylab="", cex.axis=2)#,border="white")
-  image(tr_nswvic.step.new7,col=plotclt,breaks=c(-2.3,-1.645,-1.282,-0.842,0,0.842,1.282,1.645,3),legend=F,add=T)
-  plot(mdb_reg2,border="grey",add=T,lwd=3)
-  plot(aus_reg2,border="grey",add=T)
-  plot(e222,add=T)
-image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,0,1,2,3,4),lab.breaks=c(0.00,0.05,0.1,0.2,1,0.2,0.1,0.05,0.00),legend.only=T,horizontal=F,col=plotclt,legend.shrink=0.7,legend.width=0.9,legend.args=list(text=c(expression(paste("level of significance (",alpha,")")),"positive Z'","negative Z'"),side=c(2,3,1),line=0.7,cex=1.5,font=2),axis.args=list(cex.axis=1.5))
-legend(149,-36.6,c("State","MDB"),col=c("grey","grey"),
-       lwd=c(1,3),bty="n",title="Boundary",cex=1.8)
-
-# nsw 7y
-  par(mar=c(3,5,1,7))
-  plot(e222,xlab="",ylab="", cex.axis=2)#,border="white")
-  image(tr_nswvic.step.new,col=plotclt,breaks=c(-2.3,-1.645,-1.282,-0.842,0,0.842,1.282,1.645,3),legend=F,add=T)
-  plot(mdb_reg2,border="grey",add=T,lwd=3)
-  plot(aus_reg2,border="grey",add=T)
-  #plot(e222,add=T)
-# image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,0,1,2,3,4),lab.breaks=c(0.00,0.05,0.1,0.2,1,0.2,0.1,0.05,0.00),legend.only=T,horizontal=F,col=plotclt,legend.shrink=0.7,legend.width=0.9,legend.args=list(text=c(expression(paste("level of significance (",alpha,")")),"positive Z'","negative Z'"),side=c(2,3,1),line=0.5,cex=1,font=2),axis.args=list(cex.axis=0.8))
-
-  # Qld
-  par(mar=c(3,3,1,9),xpd=T)
-  plot(e111,xlab="",ylab="", cex.axis=2)#,border="white")
-  image(tr_qld.step.new7,col=plotclt,breaks=c(-3,-1.645,-1.282,-0.842,0,0.842,1.282,1.645,3),legend=F,add=T)
-  plot(mdb_reg2,border="grey",add=T,lwd=3)
-  plot(e111,add=T)
-image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,0,1,2,3,4),lab.breaks=c(0.00,0.05,0.1,0.2,1,0.2,0.1,0.05,0.00),legend.only=T,horizontal=F,col=plotclt,legend.shrink=0.7,legend.width=0.9,legend.args=list(text=c(expression(paste("level of significance (",alpha,")")),"positive Z'","negative Z'"),side=c(2,3,1),line=0.9,cex=1.5,font=2),axis.args=list(cex.axis=1.5))
-
-# qld year
-  par(mar=c(3,5,1,7))
-  plot(e111,xlab="",ylab="", cex.axis=2)#,border="white")
-  image(tr_qld.step.new,col=plotclt,breaks=c(-3,-1.645,-1.282,-0.842,0,0.842,1.282,1.645,3),legend=F,add=T)
-  plot(mdb_reg2,border="grey",add=T,lwd=3)
-  plot(e111,add=T)
-# image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,0,1,2,3,4),lab.breaks=c(0.00,0.05,0.1,0.2,1,0.2,0.1,0.05,0.00),legend.only=T,horizontal=F,col=plotclt,legend.shrink=0.7,legend.width=0.9,legend.args=list(text=c(expression(paste("level of significance (",alpha,")")),"positive Z'","negative Z'"),side=c(2,3,1),line=0.5,cex=1,font=2),axis.args=list(cex.axis=0.8))
-  dev.off()
-
-#ns = 4 results
-# positive
-qld_origp.new <- length(subset(qld.z.new,qld.z.new>=1.645))/
-  length(qld.z.new)
-# 0.00034
-# negative
-qld_orign.new <- length(subset(qld.z.new,qld.z.new<=-1.645))/
-  length(qld.z.new)
-# 0
-# shorter series
-qld_origp.new_7y <- length(subset(qld.z.new_7y,qld.z.new_7y>=1.645))/
-  length(qld.z.new_7y)
-# 0.212
-qld_orign.new_7y <- length(subset(qld.z.new_7y,qld.z.new_7y<=-1.645))/
-  length(qld.z.new_7y)
-# 0
-
-nsw_origp.z.new <- length(subset(nswvic.z.new,nswvic.z.new>=1.645))/
-  length(nswvic.z.new)
-# 0
-nsw_orign.z.new <- length(subset(nswvic.z.new,nswvic.z.new<=-1.645))/
-  length(nswvic.z.new)
-# 0
-nsw_origp.z.new7 <- length(subset(nswvic.z.new7,nswvic.z.new7>=1.645))/
-  length(nswvic.z.new7)
-# 0.03
-nsw_orign.z.new7 <- length(subset(nswvic.z.new7,nswvic.z.new7<=-1.645))/
-  length(nswvic.z.new7)
-# 0
-
-```
-
-```{r bootstrap_ns1, echo=F, eval=F}
-## one cell step trend
-## perform step test on the data for the bootstrapping test (spatial structure was kept)
-## function to calculate the step trend statistic (Hirsch, 1985), ns = 1
-stepTrend <- function(x,i, endyear=2015){
-#browser()
-    n1 <- 2002-1997
-
-  if(!is.na(x[1])) {
-
-    if(i==1) {
-      n2 <- endyear-2004
-
-      v <- x[c(229:288,313:(n2*12+313-1))]
-#       v <- (x$y-x$fitted.values)[c(1:288,313:357)]
-      v.s <- ts(v,start=c(1998,1),frequency=12)
-    } else
-      if(i==2) {
-      n2 <- endyear-2002
-      v <- x[c(229:(n2*12+n1*12+229-1))]
-#       v <- (x$y-x$fitted.values)[c(1:288, 301:357)]
-      v.s <- ts(v,start=c(1998,1),frequency=12)
-    } #else
-#       {n2 <- 2008-2002
-#       v <- (x$y-x$fitted.values)[229:357]
-# #       v <- (x$y-x$fitted.values)
-#       v.s <- ts(v,start=c(1998,1),frequency=12)}
-
-## rank by month
-   for (j in 1:12) {
-    assign(paste("r",j,sep=""),rank(subset(v.s,cycle(v.s)==j)))
-    }
-
-## mann-whitney rank sum statistic
-    rlist <- list(r1,r2,r3,r4,r5,r6,r7,r8,r9,r10,r11,r12) ## full year
-#     rlist <- list(r1,r2,r12) ## summer
-#     rlist <- list(r3,r4,r5)  ## autumn
-#     rlist <- list(r6,r7,r8)  ## winter
-#     rlist <- list(r9,r10,r11) ## spring
-    w.s <- 0
-    for (k in 1:length(rlist)) {
-    w.s <- sum(rlist[[k]][1:n1])+w.s
-    }
-
-## other statistic
-## mean for full year
-    mu <- n1*(n1+n2+1)/2
-    mu.s <- length(rlist)*mu
-## mean for one season
-#     mu <- n1*(n1+n2+1)/2
-#     mu.s <- 3*mu
-
-## variance (standard variation).
-The autocorrelation of the residuals can be assumed to be null (checked, cor.test(), true for NSW, mostly true for WA.
-QLD?
-there are some situations in WA require further investigation)
-## assume only cross correlation
-    ss <- n1*n2*(n1+n2+1)/length(rlist)
-    ss.s <- length(rlist)*ss
-## variance for one season
-#     ss <- n1*n2*(n1+n2+1)/3
-#     ss.s <- 3*ss
-
-## step trend statistic
-    z <- (w.s - mu.s)/sqrt(ss.s)
-
-    return(z) }
-
-     else {return(NA)}
-
-  }
-
-## full year
-## central QLD
-  Qld.z <- unlist(lapply(residual_qld,function(x) stepTrend(x,1)))
-  Qld.z_7y <- unlist(lapply(residual_qld_7y,
-                            function(x) stepTrend(x,1, endyear=2010)))
-## NSW/VIC
-## treatment 1: excluding 2003
-  nswvic.z <- unlist(lapply(residual_nswvic,
-                            function(x) stepTrend(x,2)))
-  nswvic.z_7y <- unlist(lapply(residual_nswvic7,
-                            function(x) stepTrend(x,2,
-                                                  endyear = 2010)))
-
-# # ns = 1 results note that this is p < 0.1 already
-#   Qld_origp <- length(subset(Qld.z,Qld.z>=1.645))
-# # 0.002
-#   Qld_orign <- length(subset(Qld.z,Qld.z<=-1.645))
-# # 0
-#   Qld_origp_7y <- length(subset(Qld.z_7y,
-#                                 Qld.z_7y>=1.645))
-# # 0
-#   Qld_orign_7y <- length(subset(Qld.z_7y,
-#                                 Qld.z_7y<=-1.645))
-# # 0.023
-#
-#   Nsw_origp <- length(subset(nswvic.z,
-#                              nswvic.z>=1.645))
-# # 0
-#   Nsw_orign <- length(subset(nswvic.z,
-#                              nswvic.z<=-1.645))
-# # 0
-#   Nsw_origp_7y <- length(subset(nswvic.z_7y,
-#                                 nswvic.z_7y>=1.645))
-# 0
-#  Nsw_orign_7y <- length(subset(nswvic.z_7y,
-#                                nswvic.z_7y<=-1.645))
-# 0
-
-# new step trend test 2: keep spatial structure, but ns = 1
-NstepTrend2 <- function(x,y,ii, endyear=2015){
-
-  n1 <- 2002-1997
-
-  if(ii==1) {
-      n2 <- endyear-2004
-      v <- x[c(229:288,313:length(x))]
-    } else
-      if(ii==2) {
-      n2 <- endyear-2002
-      v <- x[c(229:length(x))]
-    }
-
-  if(!is.na(unlist(x)[1])) {
-
-  yy<-ts(v,start=c(1998,1),frequency=12)
-#browser()
-  for(i in 1:12) {
-  y.mth <- data.frame(mth=as.vector(window(yy,start=c(1998,i),frequency=1)),o=y[[i]])[order(y[[i]]),]
-
-  ## rank by month
-    assign(paste("r",i,sep=""),rank(y.mth$mth) )
-  }
-
-## mann-whitney rank sum statistic
-    rlist <- list(r1,r2,r3,r4,r5,r6,r7,r8,r9,r10,r11,r12) ## full year
-
-    w.s <- 0
-    for (k in 1:length(rlist)) {
-    w.s <- sum(rlist[[k]][1:n1])+w.s
-    }
-
-## other statistic
-## mean for full year
-    mu <- n1*(n1+n2+1)/2
-    mu.s <- length(rlist)*mu
-
-## variance (standard variation). The autocorrelation of the residuals can be assumed to be null (checked, true for QLD and NSW)
-    ss <- n1*n2*(n1+n2+1)/length(rlist)
-    ss.s <- length(rlist)*ss
-
-## step trend statistic
-    z <- (w.s - mu.s)/sqrt(ss.s)
-
-    return(z)
-    }
-
-     else {return(NA)}
-
-  }
-
-## conduct bootstrapping 2
-R <- 1000
-
-## QLD
-z2pq.values = numeric(R)
-z2nq.values = numeric(R)
-## sample years without replacement. note that with this method, only the order of years has changed, but months within year do not change
-new.order <- list()
-# mth.l <- rep(11,12)
-# this relates to the total number of years:n1 + n2
-#
-period1 <- 2002 - 1997
-period2 <- 2015 - 2004
-mth.l <- rep((period1 + period2),12)
-
-ptm <- proc.time()
-for (i in 1:R) {
-
-  for (j in 1:12) {
-    new.order[[j]]<-sample(1:mth.l[j])
-  }
-
-  temp <- unlist(lapply(residual_qld,
-                        function(x,y) NstepTrend2(x,y,1),new.order))
-  if(i==1) {write.table(temp,file="bt_zvalues2_qld.txt",row.names=F)
-            } else
-              {write.table(temp,file="bt_zvalues2_qld.txt",append=T,row.names=F,col.names=F)}
-  z2pq.values[i] <- length(subset(temp,temp>=1.645))/length(temp)
-  z2nq.values[i] <- length(subset(temp,temp<=-1.645))/length(temp)
-}
-proc.time()-ptm
-
-write.table(z2pq.values,file="bt_out2p_qld.txt",row.names=F)
-write.table(z2nq.values,file="bt_out2n_qld.txt",row.names=F)
-
-## NSW
-period1 <- 2002 - 1997
-period2 <- 2015 - 2002
-
-mth.l <- rep((period1 + period2),12)
-
-ptm <- proc.time()
-for (i in 1:R) {
-
-  for (j in 1:12) {
-    new.order[[j]]<-sample(1:mth.l[j])
-  }
-
-  temp <- unlist(lapply(residual_nswvic,
-                        function(x,y) NstepTrend2(x,y,2),new.order))
-  if(i==1) {write.table(temp,file="bt_zvalues2_nswvic.txt",
-                        row.names=F)
-            } else
-            {write.table(temp,file="bt_zvalues2_nswvic.txt",
-                           append=T,row.names=F,col.names=F)}
-  z2pq.values[i] <- length(subset(temp,temp>=1.645))/length(temp)
-  z2nq.values[i] <- length(subset(temp,temp<=-1.645))/length(temp)
-}
-proc.time()-ptm
-
-write.table(z2pq.values,file="bt_out2p_nswvic.txt",row.names=F)
-write.table(z2nq.values,file="bt_out2n_nswvic.txt",row.names=F)
-
-bt1_nswp.values <- unlist(read.table("bt_out2p_nswvic.txt",
-                                     header=T,sep=""))*100
-bt1_nswn.values <- unlist(read.table("bt_out2n_nswvic.txt",
-                                     header=T,sep=""))*100
-
-bt1_qldp.values <- unlist(read.table("bt_out2p_qld.txt",
-                                     header=T,sep=""))*100
-bt1_qldn.values <- unlist(read.table("bt_out2n_qld.txt",
-                                     header=T,sep=""))*100
-
-jpeg("../../figures/bt1_qldp.jpg")
-hist(bt1_qldp.values,nclass=100,xlab="% positive Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
-points(qld_origp*100,0,pch=16,col="red",cex=1.8)
-dev.off()
-
-jpeg("../../figures/bt1_qldn.jpg")
-hist(bt1_qldn.values,nclass=100,xlab="% negative Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
-points(qld_orign*100,0,pch=16,col="red",cex=1.8)
-dev.off()
-
-jpeg("../../figures/bt1_nswp.jpg")
-hist(bt1_nswp.values,nclass=100,xlab="% positive Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
-points(nsw_origp*100,0,pch=16,col="red",cex=1.8)
-dev.off()
-
-jpeg("../../figures/bt1_nswn.jpg")
-hist(bt1_nswn.values,nclass=100,xlab="% negative Z'",main="1-pixel step trend",cex.lab=1.4,cex.main=1.3)
-points(nsw_orign*100,0,pch=16,col="red",cex=1.8)
-dev.off()
-
-```
-
-```{r bootstrap_ns4, eval=F, echo=F}
-# This is the same as chunk 8, but for ns = 4
-## multiple cells
-# Set up for the shorter period
-NstepTrend <- function(x,y,m,i){
-
-    n1 <- 2002-1997
-
-    if (i==1){
-
-      z <- 56  # no. of column, related to the dim of the region
-      # Qld 57 by 51
-      # NSW 38 by 33
-
-    } else {
-
-    z <- 37 }
-
-if (x%%z!=0){ #& x%%z!=(z-1)) {
-    y.1 <- m[[x]]
-    y.2 <- m[[x+1]]
-    y.3 <- m[[x+z]]
-    y.4 <- m[[x+z+1]]
-
-    if(!is.na(y.1[1])&!is.na(y.2[1])&!is.na(y.3[1])&!is.na(y.4[1])) {
-    if (i == 1) {
-      n2 <- 2009-2004
-      v.1 <- ts(y.1[c(229:288,313:length(y.1))],
-                start=c(1997,1),frequency=12)
-      v.2 <- ts(y.2[c(229:288,313:length(y.2))],
-                start=c(1997,1),frequency=12)
-      v.3 <- ts(y.3[c(229:288,313:length(y.3))],
-                start=c(1997,1),frequency=12)
-      v.4 <- ts(y.4[c(229:288,313:length(y.4))],
-                start=c(1997,1),frequency=12)
-
-    } else {
-
-      n2 <- 2010-2002
-
-      v.1 <- ts(y.1[c(229:length(y.1))],start=c(1997,1),frequency=12)
-      v.2 <- ts(y.2[c(229:length(y.2))],start=c(1997,1),frequency=12)
-      v.3 <- ts(y.3[c(229:length(y.3))],start=c(1997,1),frequency=12)
-      v.4 <- ts(y.4[c(229:length(y.4))],start=c(1997,1),frequency=12)
-}
-
-## rank by month
-#browser()
-   for (j in 1:12) {
-    y1.mth <- data.frame(mth=as.vector(window(v.1,
-                                              start=c(1997,i),
-                                              frequency=1)),
-                         o=y[[j]])[order(y[[j]]),]
-    y2.mth <- data.frame(mth=as.vector(window(v.2,
-                                              start=c(1997,i),
-                                              frequency=1)),
-                         o=y[[j]])[order(y[[j]]),]
-    y3.mth <- data.frame(mth=as.vector(window(v.3,
-                                              start=c(1997,i),
-                                              frequency=1)),
-                         o=y[[j]])[order(y[[j]]),]
-    y4.mth <- data.frame(mth=as.vector(window(v.4,
-                                              start=c(1997,i),
-                                              frequency=1)),
-                         o=y[[j]])[order(y[[j]]),]
-
-    assign(paste("r1",j,sep=""),rank(y1.mth$mth))
-    assign(paste("r2",j,sep=""),rank(y2.mth$mth))
-    assign(paste("r3",j,sep=""),rank(y3.mth$mth))
-    assign(paste("r4",j,sep=""),rank(y4.mth$mth))
-
-    }
-
-## mann-whitney rank sum statistic
-    rlist.1 <- list(r11,r12,r13,r14,r15,r16,r17,r18,r19,r110,r111,r112) ## full year
-    rlist.2 <- list(r21,r22,r23,r24,r25,r26,r27,r28,r29,r210,r211,r212)
-    rlist.3 <- list(r31,r32,r33,r34,r35,r36,r37,r38,r39,r310,r311,r312)
-    rlist.4 <- list(r41,r42,r43,r44,r45,r46,r47,r48,r49,r410,r411,r412)
-    rlist <- list(unlist(rlist.1),unlist(rlist.2),
-                  unlist(rlist.3),unlist(rlist.4))
 ## seasonal
@@
     mu.s <- 12*4*mu
-# ## mean for one season
-#     mu <- n1*(n1+n2+1)/2
-#     mu.s <- 3*4*mu
-
-## variance (standard variation).
-The autocorrelation of the residuals can be assumed to be null (checked, cor.test(), true for NSW, mostly true for WA.
-QLD?
-there are some situations in WA require further investigation)
 # assume only cross correlation
@@
     }
-## variance for one season
-#     ss <- n1*n2*(n1+n2+1)/3
-#     ss.s <- 0
-#     for (kk in 1:4) {
-#       for (kkk in 1:4) {
-#           ss.s <- 3*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
-#       }
-#     }
 
@@
 
-R <- 1000
-new.order <- list()
-
-zpq.values = numeric(R)
-znq.values = numeric(R)
-period1 <- 2002 - 1997
-period2 <- 2009 - 2004
-mth.l <- rep((period1 + period2),12)
-
-ptm <- proc.time()
-for (i in 1:R) {
-
-  for (j in 1:12) {
-    new.order[[j]]<-sample(1:mth.l[j])
-  }
-
-  temp <- unlist(lapply(1:(length(residual_qld_7y)-57),function(x,y) NstepTrend(x,y,residual_qld_7y,1),new.order))
-  if(i==1) {write.table(temp,file="bt_zvalues_qld_7y.txt",row.names=F)
-            } else
-              {write.table(temp,file="bt_zvalues_qld_7y.txt",append=T,row.names=F,col.names=F)}
-  zpq.values[i] <- length(subset(temp,temp>=1.645))/length(temp)
-  znq.values[i] <- length(subset(temp,temp<=-1.645))/length(temp)
-}
-proc.time()-ptm
-
-write.table(zpq.values,file="bt_outp_qld_7y.txt",row.names=F)
-write.table(znq.values,file="bt_outn_qld_7y.txt",row.names=F)
-
-period1 <- 2002 - 1997
-period2 <- 2009 - 2002
-mth.l <- rep((period1 + period2),12)
-
-ptm <- proc.time()
-for (i in 1:R) {
-
-  for (j in 1:12) {
-    new.order[[j]]<-sample(1:mth.l[j])
-  }
-
-  temp <- unlist(lapply(1:(length(residual_nswvic7)-38),function(x,y) NstepTrend(x,y,residual_nswvic7,2),new.order))
-  if(i==1) {write.table(temp,file="bt_zvalues_nswvic_7y.txt",row.names=F)
-            } else
-              {write.table(temp,file="bt_zvalues_nswvic_7y.txt",append=T,row.names=F,col.names=F)}
-  zpq.values[i] <- length(subset(temp,temp>=1.645))/length(temp)
-  znq.values[i] <- length(subset(temp,temp<=-1.645))/length(temp)
-}
-proc.time()-ptm
-
-write.table(zpq.values,file="bt_outp_nswvic_7y.txt",row.names=F)
-write.table(znq.values,file="bt_outn_nswvic_7y.txt",row.names=F)
-
-bt4_nswp.values <- unlist(read.table("bt_outp_nswvic_7y.txt",header=T,sep=""))*100
-bt4_nswn.values <- unlist(read.table("bt_outn_nswvic_7y.txt",header=T,sep=""))*100
-
-bt4_qldp.values <- unlist(read.table("bt_outp_qld_7y.txt",header=T,sep=""))*100
-bt4_qldn.values <- unlist(read.table("bt_outn_qld_7y.txt",header=T,sep=""))*100
-
-jpeg("../../figures/bt4_hist.jpg")
-par(mfrow=c(1,2), mar = c(4,3,2,1)+0.1)
-# nsw pos
-hist(bt4_nswp.values,nclass=100,xlab="% positive Z'",main="NSW/VIC",cex.lab=1.4,cex.main=1.3) #xlim=c(0,20))
-points(nsw_origp.z.new7*100,0,pch=16,col="red",cex=1.8)
-# # nsw neg
-# hist(bt4_nswn.values,nclass=100,xlab="% negative Z'",main="",cex.lab=1.4,cex.main=1.3, xlim=c(0,20))
-# points(nsw_orign.z.new7*100,0,pch=16,col="red",cex=1.8)
-#qld pos
-hist(bt4_qldp.values,nclass=100,xlab="% positive Z'",main="Qld",cex.lab=1.4,cex.main=1.3) #xlim=c(0,20))
-points(qld_origp.new_7y*100,0,pch=16,col="red",cex=1.8)
-#qld neg
-# hist(bt4_qldn.values,nclass=100,xlab="% negative Z'",main="",cex.lab=1.4,cex.main=1.3, xlim=c(0,20))
-# points(qld_orign.new_7y*100,0,pch=16,col="red",cex=1.8)
+## 30 years
+## full year,
+## central QLD
+  qld.z.new <- unlist(lapply(1:(length(residual_qld)-57),function(x) stepTrend2(x,residual_qld,1)))
+  qld.z.new <- c(qld.z.new,rep(NA,57))
+  tr_qld.step.new <- raster(qld_tif)
+  tr_qld.step.new[] <- qld.z.new
+
+## NSW
+  nswvic.z.new <- unlist(lapply(1:(length(residual_nswvic)-38),function(x) stepTrend2(x,residual_nswvic,2)))
+  nswvic.z.new <- c(nswvic.z.new,rep(NA,38))
+
+  tr_nswvic.step.new <- raster(nsw_tif)
+  tr_nswvic.step.new[] <- nswvic.z.new
+plotclt <- rev(brewer.pal(8,"RdYlBu"))
+
+jpeg("figures/step_new.jpg", width=1440, height=720, quality=100)
+  par(mar=c(5,3,3,7), mfrow=c(1,2), xpd=T)
+  plot(e111,xlab="",ylab="", cex.axis=2)#,border="white")
+  image(tr_qld.step.new,col=plotclt,
+        breaks=c(-3,-1.645,-1.282,-0.842,0,0.842,1.282,1.645,3),legend=F,add=T)
+  plot(mdb_reg2,border="grey",add=T,lwd=5)
+  plot(e111,add=T)
+# image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,0,1,2,3,4),lab.breaks=c(0.00,0.05,0.1,0.2,1,0.2,0.1,0.05,0.00),legend.only=T,horizontal=F,col=plotclt,legend.shrink=0.7,legend.width=0.9,legend.args=list(text=c(expression(paste("level of significance (",alpha,")")),"positive Z'","negative Z'"),side=c(2,3,1),line=0.7,cex=1.5,font=2),axis.args=list(cex.axis=1.5))
+legend(147.8,-27.5,c("State","MDB"),col=c("grey","grey"),
+       lwd=c(2,5),bty="n",title="Boundary",cex=1.8)
+
+  # NSW
+  par(mar=c(5,0,3,10))
+  plot(e222,xlab="",ylab="", cex.axis=2)#,border="white")
+  image(tr_nswvic.step.new,col=plotclt,breaks=c(-2.3,-1.645,-1.282,-0.842,0,0.842,1.282,1.645,3),legend=F,add=T)
+  plot(mdb_reg2,border="grey",add=T,lwd=5)
+  plot(aus_reg2,border="grey",add=T, lwd=2)
+  plot(e222,add=T)
+  par(mar=c(3,3,1,9),xpd=T)
+image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,0,1,2,3,4),
+           lab.breaks=c(0.00,0.05,0.1,0.2,1,0.2,0.1,0.05,0.00),
+           legend.only=T,horizontal=F,col=plotclt,
+           legend.shrink=0.7,legend.width=1.5,
+           legend.args=list(text=c(expression(paste("
+                    level of significance (",alpha,")")),"positive Z'",
+                           "negative Z'"),
+                            side=c(2,3,1),line=0.9,cex=1.5,font=2),
+           axis.args=list(cex.axis=1.5))
+
+dev.off()
+
+#ns = 4 results
+# positive
+qld_origp.new <- length(subset(qld.z.new,qld.z.new>=1.645))/
+  length(qld.z.new)
+# 0.00034
+# negative
+qld_orign.new <- length(subset(qld.z.new,qld.z.new<=-1.645))/
+  length(qld.z.new)
+# 0
+
+nsw_origp.z.new <- length(subset(nswvic.z.new,nswvic.z.new>=1.645))/
+  length(nswvic.z.new)
+# 0
+nsw_orign.z.new <- length(subset(nswvic.z.new,nswvic.z.new<=-1.0))/
+  length(nswvic.z.new)
+# 0
+
+```
+
+```{r bootstrap_ns1, echo=F, eval=F}
+## 4 cell step trend
+# this was run on the HPC
+NSW_zvalues <- readRDS("../../processed data/NSW_zvalues.RDS")
+QLD_zvalues <- readRDS("../../processed data/QLD_zvalues.RDS")
+
+bt4_nswp.values <- NSW_zvalues$zpq.values*100
+bt4_nswn.values <- NSW_zvalues$znq.values*100 
+bt4_qldp.values <- QLD_zvalues$zpq.values*100
+bt4_qldn.values <- QLD_zvalues$znq.values*100
+
+real_df <- data.frame(zvalue = c(nsw_origp.z.new, nsw_orign.z.new,
+                                 qld_origp.new, qld_orign.new)*100, 
+                      sign = rep(c("postive","negative"),2),
+                      State=c("NSW/VIC","NSW/VIC","QLD","QLD"),
+                      type=rep("real",4))
+
+MC_df <- data.frame(zvalue = c(bt4_nswp.values,bt4_nswn.values,
+                                 bt4_qldp.values, bt4_qldp.values),
+                    sign =c(rep("postive",length(bt4_nswp.values)),
+                            rep("negative",length(bt4_nswn.values)),
+                            rep("postive",length(bt4_qldp.values)),
+                            rep("negative",length(bt4_qldn.values))),
+                    State = c(rep("NSW/VIC",length(bt4_nswn.values)*2),
+                              rep("QLD",length(bt4_qldn.values)*2)),
+                    type = rep("Monte Carlo",
+                               length(bt4_nswn.values)*2 +length(bt4_qldn.values)*2))
+
+jpeg("figures/bt4_hist.jpg", width=740,height=740, quality = 100)
+# A histogram of taus
+hp <- ggplot(MC_df, aes(x=zvalue)) + geom_histogram(binwidth=2,fill="blue")
+# Histogram of p_values by state
+# With panels that have the same scaling, but different range 
+# (and therefore different physical sizes)
+#hp <- hp + facet_wrap(~ State*sign,ncol=2,scales="free")
+# # add a red point for the real slope from the data
+hp <- hp + geom_point(data=real_df,aes(x=zvalue, y=0),colour="red",
+                 shape=16,size=8) + 
+  facet_wrap(~ State*sign,ncol=2, scales="free") #+
+hp <- hp + scale_colour_grey(start = 0, end = 0.6)
+hp <- hp + theme(axis.title = element_text(face ="bold", size=rel(2)),
+                 strip.text = element_text(face ="bold", size=rel(2)),
+                 axis.text = element_text(face ="bold", size=rel(1.5)))
+hp
 dev.off()
@@
 
-The pixels, where the tree cover change based on the MOD44B data was significant ($p \leq 0.05$) in each study region, are shown in Figure \@ref(fig:tctrend) for the time series to 2009 (left panels) and the time series to 2015 (right panels).
-In both cases it can be seen that the area of negative cover change was greater in the series to 2009 than in the series to 2015.
-But even the longer series indicates a large change in the NSW/VIC region (bottom row).
-In the NSW/VIC region, much of the tree loss between 2002 and 2003 was concentrated in the Snowy Mountains close to the border of NSW and VIC.
+The pixels, where the tree cover change based on the MOD44B data was significant ($p \leq 0.05$) in each study region, are shown in Figure \@ref(fig:tctrend) for the NSW/VIC region  (left panel) and the QLD region (right panel).
+In the QLD region there actually has generally been an increase in tree cover after the clearing of native vegetation stopped.
+In the NSW/VIC region, much of the tree loss between 2002 and 2003 was concentrated in the Snowy Mountains close to the border of NSW and VIC, which is also evident in the figure.
 Tree cover loss occurred in large parts of the QLD region between 2002 and 2005, but this tree loss was spatially less concentrated.
-Most of the clearings appear in the centre of this region in the series to 2009.
-The tree cover change map (\@ref(fig:tctrend) is consistent with the annual mean EVI trend map (based on DLCD data, not shown here).
-However in the series to 2015 the tree cover loss seems to have dissappeared (top right Figure \@ref(fig:tctrend)) and suggests an increase in tree cover over the area.
-The implications of this will be discussed in more detail in the discussion.
-
-```{r tctrend, fig.cap ="The maps show signifcant changes in tree cover identified from the MOD44B data between 2003 and 2009 (top left) and 2003 and 2015 (top right) in the NSW/VIC region and the Qld region (bottom left to 2009 and bottom right to 2015).
+```{r tctrend, fig.cap ="The maps show significant changes in tree cover identified from the MOD44B data between 2003 and 2015 in the NSW/VIC region (right) and the Qld region (left).
 The amount of change was calculated as the difference in tree cover before and after the specified land cover intervention and it is shown as the percentage of the ground area.
@@
 
-<!-- ```{r tc_trend, fig.cap="The maps show signifcant changes in tree cover identified from the MOD44B data between 2003 and 2009 (top left) and 2003 and 2015 (top right) in the NSW/VIC region and the Qld region (bottom left to 2009 and bottom right to 2015).
-The amount of change was calculated as the difference in tree cover before and after the specified land cover intervention and it is shown as the percentage of the ground area.
-Green colour indicates an increase in tree cover, while red colour indicates a decrease in tree cover.", echo=F, out.width="90%"} -->
-
-<!-- include_graphics("figures/tc_figs.jpg") -->
-<!-- ``` -->
-
 ## Regression Model and significance of Vegetation Cover Changes
 
-Generally the regression model only explains a limited amount of the rainfall variability.
-The model in Equation \@ref(eq:model2) accounts for around 15\%\footnote{Here the adjusted $R^2$ was reported.
-Adjusted $R^2$ is the coefficient of determination, a measurement of the amount of variability predicted by the model adjusting for the number of explanatory terms} of the rainfall variation in both regions.
+Generally the regression model only explains a limited amount of the rainfall variability (Figure \@reg(fig:rsq_model).
+The model in Equation \@ref(eq:model2) accounts for an average around 33\% of the variation in the NSW/VIC region and about 50% of the variation in the QLD region as indicated by the adjusted $r^2$.
+The adjusted $r^2$ is the coefficient of determination, a measurement of the amount of variability predicted by the model adjusting for the number of explanatory terms.
 The residual analysis shows that the assumptions of the regression model are generally met (Figure \@ref(fig:residuals).
@@
 
-The model, however, confirms the importance of the climate drivers and the seasonality in Australian rainfall.
+```{r rsq_model, fig.cap = "The distribution of the variance explained (adjusted $r^2$ ) by the full regression model", echo=F, out.width="90%"}
+include_graphics("figures/rsq_fullmodel.jpg")
+
+```
+
+In terms of predictors of the local rainfall in the model, logically, the average rainfall across the larger Murray Darling Basin is highly significant.
+This confirms that this variable is a good reflection of the year on year variability in the rainfall.
+The model also confirms the importance of the climate drivers and the seasonality in Australian rainfall, as many of these variables were significant.
 Even at the grid level, the seasons and several of the climatic indices were significant ($p \leq 0.05$) everywhere in both regions.
-The explaining power of the model is mostly due to these variables.
-The climate drivers (at lag zero) accounted, on average, for 6.7\% of the rainfall variability in both the QLD region and the NSW/VIC region (see Figure \@ref(fig:rsq) for the distribution of $R^2$ in these two regions).
-These figures are within the upper bound of seasonal rainfall predictability by a SST anomaly field reported by [@Westra2010].
+The climate drivers (at lag zero) accounted, on average, for 6.5\% of the rainfall variability in both the QLD region and the NSW/VIC region (see Figure \@ref(fig:rsq) for the distribution of adjusted $r^2$ in these two regions).
+These figures are within the upper bound of seasonal rainfall predictability by a SST anomaly field reported by @Westra2010.
 
@@
 
-Statistically significant long term trends were only observed in part of the NSW region (results not shown).
-However, this result might prove or disprove the existence a long term trend in rainfall.
+There were generally no statistically significant long term trends in both regions.
+However, this result might not prove or disprove the existence a long term trend in rainfall.
 The overall time period is fairly short [@Koutsoyiannis2007] and more pixels in NSW/VIC could indicate a significant step change if the long term trend effect is not removed by the model.
@@
 ```{r LCp, fig.cap = "The spatial distribution of significance of land cover step change variable in the GAM model predicting changes in rainfall in the both regions.
-The top row reflects NSW/VIC, while the bottom relates to the Qld study site.
-The left hand plots are the data up to 2009, while the right hand plots are for the full series.
+The left graph relates to the Qld region, while the right hand plot reflects the NSW/VIC region.
+The outlines of relevant Australian states and the Murray Darling Basin are indicated in grey.
 The p value reported is for the land cover variable in the model.", echo=F, out.width="90%"}
@@
 The land cover variable in the model (Equation \@ref(eq:model2)) aims to identify a step change in the rainfall before and after the observed change in land cover.
-The  variable was only significant ($p \leq 0.05$) for the rainfall estimates in some areas in NSW/VIC, as shown in Figure \@ref(fig:LCp).
-However, the number of pixels where the landcover variable was significant was much greater for the series up to 2009, compared to the series to 2015.
-There was no direct relationship between the areas of bushfires in Figure \@ref(fig:bushfire).
-No significant step change due to the land cover changes was found in rainfall in the QLD region in the series to 2015, but a small area of change was identified in the period to 2009 (left bottom panel).
-
-Figure \@ref(fig:LCp) for NSW/VIC suggests that a significant step change in rainfall related to land cover change was found in an area larger than where the bushfires has occurred.
-This might be showing a large scale effect that reaches beyond the vegetation cover change effect.
-
-More generally, the model suggests that the tree cover has a positive impact on rainfall in both regions.
+The  variable was mainly significant ($p \leq 0.05$) for the rainfall estimates in some areas in NSW/VIC, as shown in Figure \@ref(fig:LCp) (right panel).
+However, the number of pixels where the landcover variable was significant was much greater in NSW/VIC compared to the QLD region.
+There was also some relationship between the areas of bushfires in Figure \@ref(fig:bushfire).
+Only a very small area with a significant step change due to the land cover changes was found in rainfall in the QLD region (left panel)
+
+More generally, the model for NSW/VIC suggests that the land cover variable has a positive impact on rainfall in both regions.
 The fitted coefficients for the Land cover change variable were consistently positive for the "tree" part of the series.
@@
 ```{r steptest30, fig.cap="Spatial distribution of the step trend test Z' statistics in the two study sites.
-Panels on the top are for the NSW/Vic site, while panels of the bottom ar for Qld.
-The left panels are for the series to 2009, while the right panels are for the full data series.
-Warm colours (yellow, orange and red) are for positive Z' values which indicate decreasing rainfall trend due to the land surface intervention.
+The panel on the left is for the QLD region, while the panel of the right is for NSW/VIC.
+Warm colours (yellow, orange and red) are for positive Z' values which indicate decreasing rainfall trend due to the land surface change.
 Cold colours (light blue to blue) are for negative Z' values which indicate increasing rainfall trend.
@@
 
-The spatial step trend test Z' scores for the Qld region (top row) and NSWVIC region (bottom row)  are shown in Figure \@ref(fig:steptest30).
-The figure also contains the period to 2009 (left panels) and the period to 2015 (right panels).
+The spatial step trend test based on the regression residuals without the landcover variable generates Z' scores.
+The results for the Qld region (left) and NSWVIC region (right)  are shown in Figure \@ref(fig:steptest30).
 This figure provides two types of information: the sign and the significance level.
-The sign indicates the direction of the step change, as listed in Table \@ref(Zscore).
-In each region up to 2009, there is a broad area of positive Z' values which implies a decrease in rainfall.
-There appears to be little relationship between the locations where changes in tree cover are observed (Figure \@ref(fig:tc_trend)) and the patterns in the Z' score.
+The sign indicates the direction of the step change in the residuals of the regression, as listed in Table \@ref(Zscore).
+In both regions there are areas of positive Z' values which imply a decrease in rainfall, but this is stronger in the NSW/VIC area than in the QLD area.
+For the NSW/VIC area there appears to be a reasonable relationship between the locations where changes in tree cover are observed (Figure \@ref(fig:tc_trend)) and the patterns in the Z' score.
 However, there is not necessarily a direct relationship as movement of air masses could mean that actual changes of rainfall are observed close by, but not necessarily exactly at areas with changes of landcover.
-There is once again a difference between the left panels and the right panels, indicating that considering a longer time series (to 2015) reduces the pattern and significance of Z' scores.
-In the QLD region to 2009, 21\% of the pixels obtained a positive Z' score with p $<$ 0.1, and for the period to 2015 no pixels with positive Z' score at p $<$ 0.1 occurred.
-In the NSW/VIC region to 2009,  only 2.6\% of pixels in the alpine area having a positive Z' score with p $<$ 0.1.
+There is once again a difference between the left panel and the right panel, indicating only a few significant of Z' scores in the QLD region wich matches the earlier significance in the landcover variable in the full regresion.
+In the QLD region, only 0.2\% of the pixels obtained a positive Z' score with p $<$ 0.1.
+In the NSW/VIC region 3.7\% of pixels have a positive Z' score with p $<$ 0.1.
 In general it is only a small proportion of both study regions.
@@
 
-The rainfall residuals of the two periods (before-change since 1979 and after-change) were also compared (Figure \@ref(fig:meandiff)) using a simple t-test.
-From the boxplots it is difficult to see that there were significantly different ($p < 0.05$) mean values between the "before" and "after" periods in both regions.
-For the Queensland locations, there was no significant difference if measured to 2009 and slightly less rain ($p < 0.05$) before if measured to 2015 ($~ 2 mm/month$).
-In contrast, for the NSW/VIC locations there was also slightly more rain ($p < 0.05$) after the change if measured to 2015 ($~ 1 mm/month$), but significantly less rain ($p < 0.05$) after the change ($~ 9 mm/month$) if measured to 2009.
-
-```{r dif_mean, echo=F, eval=F}
-## calculate mean rainfall difference in two periods up to 2010
-## and up to 2015
-# exclude 2010 and 2006
-# First across all the data (repeat of earlier t-test)
-
-resQld <- resplot_df %>%
-  filter(Date >= ymd("1998-01-01")) %>%
-  mutate(Change = ifelse(year(Date) > 2003, "After", "Before")) %>%
-  filter(year(Date) < 2010 & year(Date) != 2006) %>%
-  filter(State == "QLD")
-
-  t.test(formula=residuals~Change,data=resQld)
-
-# no difference to 2015
-# data:  residuals by Change
-# t = 0.47233, df = 486360, p-value = 0.6367
-# alternative hypothesis: true difference in means is not equal to 0
-# 95 percent confidence interval:
-#  -0.1436579  0.2348813
-# sample estimates:
-#  mean in group After mean in group Before
-#            -1.287096            -1.332708
-
-# Qld to 2009 more rain after  (2 mm/month)
-# data:  residuals by Change
-# t = 24.139, df = 348950, p-value < 2.2e-16
-# alternative hypothesis: true difference in means is not equal to 0
-# 95 percent confidence interval:
-#  2.574463 3.029469
-# sample estimates:
-#  mean in group After mean in group Before
-#             1.469258            -1.332708
-
-resNSW <- resplot_df %>%
-  filter(Date >= ymd("1998-01-01")) %>%
-  mutate(Change = ifelse(year(Date) > 2003, "After", "Before")) %>%
-  filter(year(Date) < 2010 & year(Date) != 2006) %>%
-  filter(State != "QLD")
-
-  t.test(formula=residuals~Change,data=resNSW)
-# NSW to 2015 Before less rain (about 2 mm/month)
-#   data:  residuals by Change
-# t = 12.092, df = 217850, p-value < 2.2e-16
-# alternative hypothesis: true difference in means is not equal to 0
-# 95 percent confidence interval:
-#  1.776336 2.463545
-# sample estimates:
-#  mean in group After mean in group Before
-#            -2.034961            -4.154902
-
-# NSW to 2009 : Less rain After
-#  data:  residuals by Change
-# t = -28.481, df = 161480, p-value < 2.2e-16
-# alternative hypothesis: true difference in means is not equal to 0
-# 95 percent confidence interval:
-#  -5.698085 -4.964340
-# sample estimates:
-#  mean in group After mean in group Before
-#            -9.486115            -4.154902
-
-# Only need to do this for pos Z score (decrease in rainfall)
-
-# full period
-  sel_neg.qld <- residual_qld[which(values(tr_qld.step.new)>=1.645)]
-  sel_neg_df <- data_frame(Date = rep(ymd(paste(nsw_30[[1]],
-                                              "01", sep="-")),
-                        length(sel_neg.qld)),
-                          residuals = unlist(sel_neg.qld),
-                        pixel = rep(1:length(sel_neg.qld), each = 444))
-
-  data.qld <- sel_neg_df %>%
-    group_by(year = year(Date), pixel = pixel) %>%
-    summarise(an_residual = sum(residuals)) %>%
-  # rearrange
-    filter(year > 1997) %>%
-    mutate(group = ifelse(year < 2003,"1before","2after"))
-
-  # shorter period to 2010
-  sel_neg.qld_7y <- residual_qld_7y[which(values(tr_qld.step.new7)>=1.645)]
-  sel_neg_df_7y <- data_frame(Date = rep(ymd(paste(nsw_30[[1]][1:372],
-                                              "01", sep="-")),
-                        length(sel_neg.qld_7y)),
-                          residuals = unlist(sel_neg.qld_7y),
-                        pixel = rep(1:length(sel_neg.qld_7y), each = 372))
-
-  data.qld_7y <- sel_neg_df_7y %>%
-    group_by(year = year(Date), pixel=pixel) %>%
-    summarise(an_residual = sum(residuals)) %>%
-    filter(year > 1997) %>%
-    mutate(group = ifelse(year < 2003,"1before","2after"))
-
-# NSW/VIC full period no pixels at p < 0.1
-  # sel_neg.nswvic <-
-  #   residual_nswvic[which(values(tr_nswvic.step.new)>=1.645)]
-  # sel_neg_df <- data_frame(Date = rep(ymd(paste(nsw_30[[1]],
-  #                                             "01", sep="-")),
-  #                       length(sel_neg.nswvic)),
-  #                         residuals = unlist(sel_neg.nswvic),
-  #                       pixel = rep(1:length(sel_neg.nswvic), each = 444))
-  #
-  # data.nswvic <- sel_neg_df %>%
-  #   group_by(year = year(Date), pixel = pixel) %>%
-  #   summarise(an_residual = sum(residuals)) %>%
-  # # rearrange
-  #   filter(year > 1997) %>%
-  #   mutate(group = ifelse(year < 2003,"1before","2after"))
-
-# NSW/VIC shorter period to 2010
-  sel_neg.nswvic_7y <-
-    residual_nswvic7[which(values(tr_nswvic.step.new7)>=1.645)]
-  sel_neg_df_7y <- data_frame(Date = rep(ymd(paste(nsw_30[[1]][1:372],
-                                              "01", sep="-")),
-                        length(sel_neg.nswvic_7y)),
-                          residuals = unlist(sel_neg.nswvic_7y),
-                        pixel = rep(1:length(sel_neg.nswvic_7y), each = 372))
-
-  data.nswvic_7y <- sel_neg_df_7y %>%
-    group_by(year = year(Date), pixel = pixel) %>%
-    summarise(an_residual = sum(residuals)) %>%
-  # rearrange
-    filter(year > 1997) %>%
-    mutate(group = ifelse(year < 2003,"1before","2after"))
-
-# Qld full period
-## perform a 2 sample t test
-  var.test(unlist(data.qld[data.qld$group=="1before","an_residual"]),
-           unlist(data.qld[data.qld$group=="2after" &
-                      data.qld$year!=2006 &
-                      data.qld$year!=2010 ,3]),alternative="two.sided")
-## equal variance p = 0.1802
-  t.test(unlist(data.qld[data.qld$group=="1before","an_residual"]),
-           unlist(data.qld[data.qld$group=="2after" &
-                      data.qld$year!=2006 &
-                      data.qld$year!=2010  ,3]),alternative="greater")
-## p = 0.1702
-
-# Qld shorter period
-## perform a 2 sample t test
-  var.test(unlist(data.qld_7y[data.qld_7y$group=="1before","an_residual"]),
-           unlist(data.qld_7y[data.qld_7y$group=="2after" &
-                      data.qld_7y$year!=2006 &
-                      data.qld_7y$year!=2010 ,3]),alternative="two.sided")
-## unequal variance p < 0.05
-  t.test(unlist(data.qld_7y[data.qld_7y$group=="1before","an_residual"]),
-           unlist(data.qld_7y[data.qld_7y$group=="2after" &
-                      data.qld_7y$year!=2006 &
-                      data.qld_7y$year!=2010,3]),alternative="greater")
-## p < 0.05
-#   sample estimates:
-# mean of x mean of y
-#  25.38161  11.14015
-
-# NSW only shorter period
-## perform a 2 sample t test
-  var.test(unlist(data.nswvic_7y[data.nswvic_7y$group=="1before",
-                                 "an_residual"]),
-           unlist(data.nswvic_7y[data.nswvic_7y$group=="2after" &
-                      data.nswvic_7y$year!=2006 &
-                      data.nswvic_7y$year!=2010 ,3]),alternative="two.sided")
-## unequal variance p = 0.0543
-  t.test(unlist(data.nswvic_7y[data.nswvic_7y$group=="1before",
-                               "an_residual"]),
-           unlist(data.nswvic_7y[data.nswvic_7y$group=="2after" &
-                      data.nswvic_7y$year!=2006 &
-                      data.nswvic_7y$year!=2010,3]),alternative="greater")
-## p < 0.05
-# sample estimates:
-# mean of x mean of y
-#  49.92258 -51.18393
-
-### REPEAT FOR THE REVERSE ###----------------------------------
-  # full period
-  sel_neg.qld <- residual_qld[which(values(tr_qld.step.new)<1.645)]
-  sel_neg_df <- data_frame(Date = rep(ymd(paste(nsw_30[[1]],
-                                              "01", sep="-")),
-                        length(sel_neg.qld)),
-                          residuals = unlist(sel_neg.qld),
-                        pixel = rep(1:length(sel_neg.qld), each = 444))
-
-  data.qld <- sel_neg_df %>%
-    group_by(year = year(Date), pixel = pixel) %>%
-    summarise(an_residual = sum(residuals)) %>%
-  # rearrange
-    filter(year > 1997) %>%
-    mutate(group = ifelse(year < 2003,"1before","2after"))
-
-  # shorter period to 2010  ## repeat for reverse
-  sel_neg.qld_7y <- residual_qld_7y[which(values(tr_qld.step.new7)<1.645)]
-  sel_neg_df_7y <- data_frame(Date = rep(ymd(paste(nsw_30[[1]][1:372],
-                                              "01", sep="-")),
-                        length(sel_neg.qld_7y)),
-                          residuals = unlist(sel_neg.qld_7y),
-                        pixel = rep(1:length(sel_neg.qld_7y), each = 372))
-
-  data.qld_7y <- sel_neg_df_7y %>%
-    group_by(year = year(Date), pixel=pixel) %>%
-    summarise(an_residual = sum(residuals)) %>%
-    filter(year > 1997) %>%
-    mutate(group = ifelse(year < 2003,"1before","2after"))
-
-## REPEAT FOR REVERSE NSW -----------------------
-# NSW/VIC full period no pixels at p < 0.1
-  # sel_neg.nswvic <-
-  #   residual_nswvic[which(values(tr_nswvic.step.new)>=1.645)]
-  # sel_neg_df <- data_frame(Date = rep(ymd(paste(nsw_30[[1]],
-  #                                             "01", sep="-")),
-  #                       length(sel_neg.nswvic)),
-  #                         residuals = unlist(sel_neg.nswvic),
-  #                       pixel = rep(1:length(sel_neg.nswvic), each = 444))
-  #
-  # data.nswvic <- sel_neg_df %>%
-  #   group_by(year = year(Date), pixel = pixel) %>%
-  #   summarise(an_residual = sum(residuals)) %>%
-  # # rearrange
-  #   filter(year > 1997) %>%
-  #   mutate(group = ifelse(year < 2003,"1before","2after"))
-
-# NSW/VIC shorter period to 2010
-  sel_neg.nswvic_7y <-
-    residual_nswvic7[which(values(tr_nswvic.step.new7)<1.645)]
-  sel_neg_df_7y <- data_frame(Date = rep(ymd(paste(nsw_30[[1]][1:372],
-                                              "01", sep="-")),
-                        length(sel_neg.nswvic_7y)),
-                          residuals = unlist(sel_neg.nswvic_7y),
-                        pixel = rep(1:length(sel_neg.nswvic_7y), each = 372))
-
-  data.nswvic_7y <- sel_neg_df_7y %>%
-    group_by(year = year(Date), pixel = pixel) %>%
-    summarise(an_residual = sum(residuals)) %>%
-  # rearrange
-    filter(year > 1997) %>%
-    mutate(group = ifelse(year < 2003,"1before","2after"))
-
-# Qld full period Repeat for reverse
-## perform a 2 sample t test
-  var.test(unlist(data.qld[data.qld$group=="1before","an_residual"]),
-           unlist(data.qld[data.qld$group=="2after" &
-                      data.qld$year!=2006 &
-                      data.qld$year!=2010 ,3]),alternative="two.sided")
-## equal variance p = 0.1802
-  t.test(unlist(data.qld[data.qld$group=="1before","an_residual"]),
-           unlist(data.qld[data.qld$group=="2after" &
-                      data.qld$year!=2006 &
-                      data.qld$year!=2010  ,3]),alternative="greater")
-## p < 0.05
- #  sample estimates:
- # mean of x  mean of y
- # -8.978994 -18.078116
-
-# Qld shorter period repeat for reverse
-## perform a 2 sample t test
-  var.test(unlist(data.qld_7y[data.qld_7y$group=="1before","an_residual"]),
-           unlist(data.qld_7y[data.qld_7y$group=="2after" &
-                      data.qld_7y$year!=2006 &
-                      data.qld_7y$year!=2010 ,3]),alternative="two.sided")
-## unequal variance p < 0.05
-  t.test(unlist(data.qld_7y[data.qld_7y$group=="1before","an_residual"]),
-           unlist(data.qld_7y[data.qld_7y$group=="2after" &
-                      data.qld_7y$year!=2006 &
-                      data.qld_7y$year!=2010,3]),alternative="greater")
-## p = 1
-
-# NSW only shorter period
-## perform a 2 sample t test
-  var.test(unlist(data.nswvic_7y[data.nswvic_7y$group=="1before",
-                                 "an_residual"]),
-           unlist(data.nswvic_7y[data.nswvic_7y$group=="2after" &
-                      data.nswvic_7y$year!=2006 &
-                      data.nswvic_7y$year!=2010 ,3]),alternative="two.sided")
-## unequal variance p = 0.0543
-  t.test(unlist(data.nswvic_7y[data.nswvic_7y$group=="1before",
-                               "an_residual"]),
-           unlist(data.nswvic_7y[data.nswvic_7y$group=="2after" &
-                      data.nswvic_7y$year!=2006 &
-                      data.nswvic_7y$year!=2010,3]),alternative="greater")
-## p < 0.05
-# sample estimates:
-#  mean of x  mean of y
-#   8.792574 -15.508618
-
-```
-
-A closer look at the rainfall data indicates that rainfall was consistently very low in 2006 for both regions and consistently high for 2010 \@ref(fig:ts-mean).
-The low rainfall in 2006 was due to a weak El Niño and unrecovered conditions from the previous drought, in contrast, 2010 brough record drought breaking rains across Eastern Australia related to a La Niña [@vanDijk2013].
-While the regression model has removed most of the effect of ENSO, the effect of these anomolous years could still be visible in the residuals if the response of the rainfall to the ENSO effect is non-linear, and for example the memory of the past drought is persistent.
-This can also explain the significant results in the previous method.
-To remove the possible outlier effects from 2006 and 2010, the two periods were compared excluding the 2006 and 2010 rainfall.
-In addition, the group of pixels showing a negative step change in rainfall ($p < 0.10$) and the group of pixels indicating no change were analysed separately for the period to 2009.
-An unpaired unequal variance two sample t-test was applied to test whether the after period had lower mean rainfall than the before period, excluding the 2006 and 2010 years for the full period, and excluding 2006 for the period to 2009.
-
-For the Queensland locations using the regression model residuals and excluding 2006 and 2010, there was no difference between the period before and after the landuse change for the full data series.
-However for the period to 2009, there is slightly more rain ($p < 0.05, ~ 2mm/month$) before than after the change.
-For the NSW/VIC locations for the same analyses, there is slightly more rain ($p < 0.05$) after the change if measured to 2015 ($~ 2 mm/month$), and for the period to 2009 there is still significantly less rain ($p < 0.05, ~ 5 mm/month$).
-Focussing on the pixels with a negative step change (positive significant Z' score) and the monthly rainfall totals, only the period to 2009 is of interest, as the full period had too few significant pixels (\@ref(fig:steptest30)).
-For both regions and excluding 2006, the Qld area had significantly more rain before than after at $p < 0.05$.
-For the NSW/VIC area, there was similarly significantly more rain before than after the change ($p < 0.05 and 100mm/month$).
-For pixels where the Z' score was not significantly positive ($p > 0.1$ and Z' score negative), the Qld area showed no  significant difference in monthly rainfall totals before and after.
-For the NSW/VIC area there was still significantly less rain after the change ($p < 0.05$) but difference in monthly rainfall totals was smaller than for positive and signifcant Z' scores: 24 mm/month less.
-
-The choice of $ns$ has some impact on the test results [@Hirsch1985].
+The rainfall regression residuals without the landcover variable for the two regions (before-change since 1979 and after-change) were also compared (Figure \@ref(fig:meandiff)) using a simple t-test.
+From the boxplots it is difficult to see that there were significantly different ($p < 0.05$) mean rainfall residual values between the "before" and "after" periods in both regions.
+For the Queensland locations, the mean monthly rainfall residuals were slightly lower  ($p < 0.05$) after the change in landcover ($~ 0.45 mm/month$).
+For the NSW/VIC locations there was larger decrease in the mean monthly rainfall residual ($p < 0.05$) after the change($~ 1.5 mm/month$).
+
+The choice of the number of pixels averaged in the step trend test ($ns$) has some impact on the test results [@Hirsch1985].
 The cases of $ns = 1$ and $ns = 9$ were also tested.
-The results for $ns = 1$ showed a lower number of significant pixels(at p $<$ 0.10) compared to the $ns = 4$ test, with only 66 pixels in the Queensland area in the period to 2010 and a slightly higher 6 pixels in the period to 2015.
-No pixels were significant at p $<$ 0.1 in the NSW/VIC area.
+The results for $ns = 1$ showed a lower number of significant pixels(at p $<$ 0.10) compared to the $ns = 4$ test.
 The results from the $ns = 9$ test was not different from the $ns = 4$ test.
+We therefore decided to stay with $ns = 4$.
 The power of the test does not change much after $ns = 4$, as shown by @Hirsch1985.
 
-As part of the analysis, the "field significance" of the Z' score test was considered to better interpret the step change at regional scales from multiple local tests [@Wilks2006, @Westra2013].
-Here, the bootstrapping resampling method from @Westra2013 was used to evaluate the field significance for the period up to 2009 (as the period to 2015 essentially showed no significant positive Z' scores).
+As part of the analysis, the "field significance" of the Z' score test was considered to improve the interpretion of the step change at regional scales from multiple local tests [@Wilks2006, @Westra2013].
+Here, the bootstrapping resampling method from @Westra2013 was used to evaluate the field significance.
 This means the spatial structure of the pixels was maintained, but the order of the years and months was changed by random resampling.
-For each resampling the test statistics identifies the percentage of the pixels with significant positive step change for the step trend test.
-The test statistics on 1000 resampled replicates were used to develop the distribution of these percentage values under the local null hypothesis that there was no step change.
-The results showed that none of the resampled series indicated any significant step trend for both regions, suggesting that the actual values found are for a unique series of years.
+For each resampling, the test statistic identifies the percentage of the pixels with a significant positive or negative step change for the step trend test.
+The test statistics on 1000 resampled replicates were used to develop the distribution of these percentage values, and the observed fraction of significant Z' scores (at $p < 0.1$) is plotted on the distribution (Figure \@ref(fig:fieldsig)).
+The results support the earlier Z' score results.
+Only the percentage significant positive Z' scores for the NSW/VIC region falls on the tail of the field significance distribution, while all the others are well within the distribution.
+This suggests that the percentage of significant positive Z' scores in the NSW/VIC region is least likely due to a unique series of rainfall years, but is most likely due to changes in landcover.
+In other words, in the NSW/VIC region it is most likely that the change in landcover caused a decrease in the rainfall.
+
+```{r fieldsig, fig.cap ="Field significance results showing the distribution of % of significant Z' scores for the resampled series (blue bars), and the % significant Z'scores for the original series of rainfall years (red circle)", echo=F, out.width="90%"}
+
+include_graphics("figures/bt4_hist.jpg")
+```
 
@@
 table <- read_csv("TestSummaryTable2.csv")
-kable(table,caption="Summary table of all tests on the two regions and for the two time periods", format = "latex", booktabs =TRUE) %>%
+kable(table,caption="Summary table of all tests on the two regions ", format = "latex", booktabs =TRUE) %>%
   kable_styling(latex_options = "scale_down")#%>%
@@
 
-Overall the summary table of the results indicates that the effects of land clearing or bushfire on rainfall are not easily detected, even with a range of different statistical tests.
-In addition several of the tests appear to indicate contradictory results, in particular in the comparison between the two regions and between the two time periods.
+Overall the summary table of the results indicates that the effects of land clearing or bushfire on rainfall are consistent across the range of different statistical tests.
+In all cases, the data for the NSW/VIC region
 
@@
 
-## Misalignment rainfall effects and vegetation clearing
+## Misalignment of rainfall effects and vegetation clearing
 
@@
 
-<!-- To discuss:   -->
-
-<!-- * Post-fire revegetation and restoration activities in ACT could have reduced the significance of the difference in tree cover before and after the bushfires.
-It is an indication that deforestation might cause a step change in rainfall but reforestation could take a long time to show effects.
-Regrowth in Qld and increase in clearing again (Figure \@ref(fig:slat)).
--->
-<!-- * Rainfall variation and anomalous years 2006 and 2010.   -->
-<!-- * misalingment rainfall change and clearing  -->
-<!-- * gridded rainfal    -->
-<!-- * general approach   -->
-
 ## General approach
@@
 
-In this study, we present an approach to identify iffound some observationl data based evidence, although not strong, that vegetation cover change has changed local rainfall.
+In this study, we present an approach to identify if
+
+found some observationl data based evidence, although not strong, that vegetation cover change has changed local rainfall.
 The semi-parametric method and non-parametric method did totally agree on detecting a significant step change in rainfall in the hot dry QLD region where land clearing has occurred.
@@
 
-<!--  \begin{sidewaystable} -->
-<!--  \caption{Summary of data.} -->
-<!--   \label{tab:ch3Data} -->
-<!--  \begin{tabular}{lllll} -->
-<!--   \hline -->
-<!--   \textbf{Data} & \textbf{Source} & \multicolumn{2}{c}{\textbf{Resolution}} & \textbf{Analysis period} \\\cline{3-4} -->
-<!--   & & Temporal & Spatial & \\\hline -->
-<!--   Percent tree cover & MOD44B & Annual &	250m & 2000-2010\\ -->
-<!--   Trend of vegetation cover change  &	DLCD (2009)	& Onetime & 250m	& Trend of Apr 2000 - Apr 2008\\ -->
-<!--   Rainfall &	AWAP gridded rainfall data &	Monthly &	0.05\textdegree$\times$0.05\textdegree & Jan 1979- Dec 2008\\ -->
-<!--   SOI	& BoM	& Monthly &	N/A &	Jan 1979- Dec 2008\\ -->
-<!--   NINO 3, 3.4, 4 &	IRI/LDEO data library	& Monthly	& N/A	& Jan 1979- Dec 2008\\ -->
-<!--   PDO	& NOAA & Monthly	& N/A	& Jan 1979- Dec 2008\\ -->
-<!--   IOD	& POAMA-2 dataset	& Monthly	& N/A	& Jan 1979- Dec 2008\\ -->
-<!--   \hline -->
-<!--   \end{tabular} -->
-<!-- \end{sidewaystable} -->
-
 \newpage

### 2019-02-28 Manuscripts/PeerJRainfall_Landuse/Rainfall_Landuse.tex
update for PeerJ paper

--- 2018-09-04 Manuscripts/Rainfall_Landuse_paper/Rainfall_Landuse_paper.tex
@@
-\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
-%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
-\usepackage[hyphens]{url}
-
-  \journal{Journal of Hydrology Regional Studies} % Sets Journal name
-
-\usepackage{lineno} % add
-\providecommand{\tightlist}{%
+%% Submissions for peer-review must enable line-numbering
+%% using the lineno option in the \documentclass command.
+%%
+%% Preprints and camera-ready submissions do not need
+%% line numbers, and should have this option removed.
+%%
+%% Please note that the line numbering option requires
+%% version 1.1 or newer of the wlpeerj.cls file, and
+%% the corresponding author info requires v1.2
+
+\documentclass[fleqn,10pt,lineno]{wlpeerj} % for journal submissions
+
+% ZNK -- Adding headers for pandoc
+
+\setlength{\emergencystretch}{3em}
+\providecommand{\tightlist}{
 \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-\bibliographystyle{elsarticle-harv}
-\biboptions{sort&compress} % For natbib
-\usepackage{graphicx}
-\usepackage{booktabs} % book-quality tables
-%%%%%%%%%%%%%%%% end my additions to header
-
-\usepackage[T1]{fontenc}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \usepackage{fontspec}
-  \ifxetex
-    \usepackage{xltxtra,xunicode}
-  \fi
-  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
-  \newcommand{\euro}{€}
-\fi
-% use microtype if available
-\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
+\usepackage{lipsum}
+\usepackage[unicode=true]{hyperref}
 \usepackage{longtable}
-\ifxetex
-  \usepackage[setpagesize=false, % page size defined by xetex
-              unicode=false, % unicode breaks when used with xetex
-              xetex]{hyperref}
-\else
-  \usepackage[unicode=true]{hyperref}
-\fi
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={},
-            pdftitle={Detecting influences of changes in land cover on observed rainfall.},
-            colorlinks=true,
-            urlcolor=blue,
-            linkcolor=magenta,
-            pdfborder={0 0 0}}
-\urlstyle{same}  % don't use monospace font for urls
-
-\setcounter{secnumdepth}{5}
-% Pandoc toggle for numbering sections (defaults to be off)
-% Pandoc header
-
-\usepackage{amsthm}
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}{Lemma}[section]
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
-\newtheorem{corollary}{Corollary}[section]
-\newtheorem{proposition}{Proposition}[section]
-\theoremstyle{definition}
-\newtheorem{example}{Example}[section]
-\theoremstyle{definition}
-\newtheorem{exercise}{Exercise}[section]
-\theoremstyle{remark}
-\newtheorem*{remark}{Remark}
-\newtheorem*{solution}{Solution}
-\begin{document}
-\begin{frontmatter}
-
-  \title{Detecting influences of changes in land cover on observed rainfall.}
-    \author[a]{Chun X. Liang}
-   \ead{summer.chunliang@gmail.com}
-
-    \author[a]{Floris F. van Ogtrop}
-   \ead{floris.vanogtrop@sydney.edu.au}
-
-    \author[a]{R. Willem Vervoort\corref{c1}}
-   \ead{willem.vervoort@sydney.edu.au} 
-   \cortext[c1]{Corresponding Author}
-      \address[a]{Sydney Institute of Agriculture, The University of Sydney, NSW 2006}
+
+\usepackage{lipsum} \usepackage{textcomp}
+
+\title{Detecting the impact of land cover change on observed rainfall.}
+
+\author[1]{Chun X. Liang}
+
+\author[1]{Floris F. van Ogtrop}
+
+\author[1]{R. Willem Vervoort}
+
+\corrauthor[1]{R. Willem Vervoort}{\href{mailto:willem.vervoort@sydney.edu.au}{\nolinkurl{willem.vervoort@sydney.edu.au}}}
+
+\affil[1]{Sydney Institute of Agriculture, The University of Sydney, NSW 2006}
+
+%
+% \author[1]{First Author}
+% \author[2]{Second Author}
+% \affil[1]{Address of first author}
+% \affil[2]{Address of second author}
+% \corrauthor[1]{First Author}{f.author@email.com}
+
+%
 
@@
 changes were observed in alonger time sereies to 2015.
+% Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
 \end{abstract}
 
- \end{frontmatter}
-
-\section{Introduction}\label{introduction}
+\begin{document}
+
+\flushbottom
+\maketitle
+\thispagestyle{empty}
+
+\section*{Introduction}\label{introduction}
+\addcontentsline{toc}{section}{Introduction}
 
 Land use and land cover changes can lead to changes in the local
-climate. Empirical and modelling studies have found cloud types and
+climate. Empirical and modelling studies have found that cloud types and
 rainfall are correlated to large scale vegetation cover changes, such as
@@
 it take for land cover change effects to appear or disappear?). In
-addition, in many areas across the globe rainfall variability is related
-to complex set of interactions, of which land use change might only be a
-minor component.
+addition, in many areas across the globe, rainfall variability is
+related to a complex set of interactions, of which land use change might
+only be a minor component.
 
@@
 scale atmospheric dynamics and hence the amount of rainfall, including
-contribution from both main sources. Therefore land surface is suggested
-to play an important role in local rainfall.
+contribution from both main sources.
 
@@
 local rainfall (e.g. Ma et al. 2011; Zeng et al. 2012; Pitman and Lorenz
-2016; Saha, Dirmeyer, and Chase 2016). However, they are mostly based on
-modelling experiments and little evidence was reported from
-observations. However, Pitman et al. (2004) found a good match between
+2016; Saha, Dirmeyer, and Chase 2016). However, most are based on
+modelling experiments and few empirical observational studies have been
+reported. However, Pitman et al. (2004) found a good match between
 observations and simulated rainfall changes in southwest Western
-Australia ,forced by land cover change. Timbal and Arblaster (2006) were
+Australia, forced by land cover change. Timbal and Arblaster (2006) were
 able to reproduce the rainfall decline in south west Australia by
@@
 
-Therefore, the aim of this study is to expand the available methodology
-to use empirical evidence at regional scales to investigate the cause
-and effect relationship between land cover change and local rainfall.
-More specifically, we hypothesize that a step change on the land cover
-on the surface will cause a step change in the rainfall. To demonstrate
-the approach and this effect we applied the methodology to study the
-changes in rainfall at two locations in Queensland and NSW/Victoria
-where there are possible step changes in land cover change due to land
-clearing and bush fires. The methodology is mainly based on statistical
-approaches to identify changes in rainfall, which were subsequently
+Therefore, the aim of this study is to use a statistical approach on
+rainfall data at regional scales to investigate the cause and effect
+relationship between land cover change and local rainfall, which is
+demonstrated in many modelling studies. More specifically, we
+hypothesize that a step change on the land cover on the surface will
+cause a step change in the rainfall. To demonstrate this we study
+changes in observed rainfall over time at a Queensland and NSW/Victoria
+location where there are possible step changes in land cover change due
+to land clearing and bush fires. The methodology uses statistical
+approaches to identify changes in rainfall, which are subsequently
 associated with land cover change through spatial comparison.
@@
 
-We use two areas in Australia to demonstrate the approach. In Australia,
-significant tree cover change has mainly occurred in the north east of
-the continent and on the southeast coast, as well as in the southwest of
-Western Australia. According to the National Dynamic Land Cover Dataset
-(DLCD) (Lymburner et al. 2010), most of these areas have experienced
-decreasing EVI post 2000. As an index for vegetation greenness, the
-decreasing values indicate lower biomass over time in the tree cover
+In Australia, significant tree cover change has mainly occurred in the
+north east and south east of the continent, as well as in the southwest
+of Western Australia. According to the National Dynamic Land Cover
+Dataset (DLCD) (Lymburner et al. 2010), most of these areas have
+experienced decreases in the Enhanced Vegetation Index (EVI) post 2000
+aws derived from satellite data. As an index for vegetation greenness,
+the decreasing values indicate lower biomass over time in the tree cover
 regions. The possible EVI reduction might be due to land clearing, bush
@@
 In site 1 (the QLD region), the tree cover is mostly sparse.
-In site 2 (the NSW/VIC region), many areas have open or close forest in which tree cover is denser.}\label{fig:selreg}
+In site 2 (the NSW/VIC region), many areas have open or close forest where the tree cover is denser.}\label{fig:selreg}
 \end{figure}
 
-Two regions were selected where significant tree cover change was
-reported. The first region is located in south central Queensland to the
-north of the Murray Darling Basin (MDB) (site 1 in Figure
-\ref{fig:selreg}. High rates of land clearing have been reported in this
-region during the early 2000s (Department of Natural Resources and Water
-2007). The second study region is located at the border of New South
-Wales and Victoria, and includes the Snowy Mountain ranges (site 2 in
-Figure \ref{fig:selreg}. Severe bush fires occurred in this area and the
-surroundings in early 2003 (see Figure \ref{fig:figure2bushfire}). The
-2003 bush fires were the largest and the worst in this area for the last
-60 years (The State Government of Victoria 2011). Two thirds of
-Kosciuszko national park was heavily burned and regrowth was reported to
-be slow due to drought and cold conditions (ABC News 2003) and the type
-of species in this region. However, in the longer term, after an early
-high transpiration period a recovery of pre-fire evapotranspiration
-would be expected (Kuczera 1987). As a result , significant tree cover
-loss has happened in both study areas in the last decade, either
-permanently or temporarily.
+Two regions were selected where significant tree cover change since 2000
+was reported. The first region is located in south central Queensland
+partly covering the north of the Murray Darling Basin (MDB) (site 1 in
+Figure @ref(fig:selreg)). High rates of land clearing have been reported
+in this region during the early 2000s (Department of Natural Resources
+and Water 2007). The second study region is located at the border of New
+South Wales and Victoria, and includes the Snowy Mountain ranges (site 2
+in Figure @ref(fig:selreg)). Severe bush fires occurred in this area in
+early 2003 (see Figure @ref(fig:figure2bushfire)). The 2003 bush fires
+were the largest and the worst in the last 60 years (The State
+Government of Victoria 2011). Two thirds of Kosciuszko national park was
+heavily burned and regrowth was reported to be slow due to drought and
+cold conditions (ABC News 2003), and the type of species in this region.
+However, in the longer term, after an early high transpiration period a
+recovery of pre-fire evapotranspiration would be expected (Kuczera
+1987). For the purpose of this study, significant tree cover loss has
+happened in both study areas in the last decade, either permanently or
+temporarily.
 
@@
 \includegraphics[width=0.9\linewidth]{figures/bushfire_nswvic} \caption{Location of bushfires occurring in January 2003, in and around the NSW/VIC study region, as shown by the red pixels.
-The map shows large area in the Kosciuszko national park has been burned.
-Some locations in the southwest of ACT have also experienced intensive bushfires.}\label{fig:figure2bushfire}
+The map shows the large area in the Kosciuszko national park that has been burned.
+Some locations in the southwest of teh Australian Capital Territory (ACT) have also experienced intensive bushfires.}\label{fig:figure2bushfire}
 \end{figure}
 
-The two regions have different climate characteristics. The QLD region
-is partially grassland and partially subtropical, while the NSW/VIC
-region is mainly within the temperate zone, under the Köppen
-classification. According to Australian Bureau of Meteorology (BoM), the
-NSW/VIC region receives 1000 - 2000 mm rainfall annually, which is more
-than double of the rainfall in the QLD region. Evapotranspiration is
-similar in both regions. Marine moisture and orographic effects are
-likely to be the main contributors to rainfall in the southeast mountain
-areas of the NSW/VIC region.
+The two regions have different characteristics. The QLD region is
+partially grassland and subtropical, while the NSW/VIC region is mainly
+within the temperate zone, under the Köppen classification. According to
+Australian Bureau of Meteorology (BoM), the NSW/VIC region receives 1000
+- 2000 mm rainfall annually, which is more than double the annual
+rainfall in the QLD region. Evapotranspiration is similar in both
+regions. Marine moisture and orographic effects are likely to be the
+main contributors to rainfall in the southeast mountain areas of the
+NSW/VIC region.
 
@@
 been cleared at a massive scale over the last decade, especially during
-2002 - 2004. The Kosciuszko national park is within the NSW/VIC region.
-Here tree cover is denser with open or even closed forest (the tree
-cover distribution is bimodal at 10 - 20\% and 60 - 70\%). The dominant
+2002 - 2004. The reports from the Queensland Statewide Land Cover and
+Trees Study (SLATS) (e.g. Department of Natural Resources and Mines
+2005; Department of Science, Information Technology and Innovation 2017)
+were used to investigate the time and location of the land clearing in
+the QLD region.
+
+The Kosciuszko national park is within the NSW/VIC region. Here tree
+cover is denser with open or even closed forest (the tree cover
+distribution is bimodal at 10 - 20\% and 60 - 70\%). The dominant
 species in the alpine area are Snow Gum and large stand species such as
@@
 is not a major issue in this region, it is vulnerable to fires and
-drought.
-
-Therefore two types of land cover changes were studied. The reports from
-the Queensland Statewide Land Cover and Trees Study (SLATS) (e.g.
-Department of Natural Resources and Mines 2005; Department of Science,
-Information Technology and Innovation 2017) were used to investigate the
-time and location of land clearing in the QLD region. The MODIS burned
-area product, MCD45A1 (Roy, Lewis, and Justice 2002; Roy et al. 2005;
-Roy et al. 2008), was used to locate bush fires areas in the NSW/VIC
-region, with a grid resolution of 500 m. MCD45A1 provides monthly
-burning information on all pixels, which helps to pinpoint an abrupt
-event. Due to the nature of the different land cover change, the
-post-change vegetation status in the two regions is expected to be
-different (see Figure \ref{fig:figure3-tc-simple}).
+drought. The MODIS burned area product, MCD45A1 (Roy, Lewis, and Justice
+2002; Roy et al. 2005; Roy et al. 2008), was used to locate bush fires
+areas in the NSW/VIC region, with a grid resolution of 500 m. MCD45A1
+provides monthly burning information on all pixels, which helps to
+pinpoint an abrupt event.
+
+Due to the difference in nature of the land cover change in the two
+regions, the post-change vegetation status is hypothesized to be
+different as well (see Figure @ref(fig:figure3-tc-simple)).
 
@@
 
-The hypothesis in this study therefore is that the effect of 2003 - 2004
-land clearings in the QLD region and the 2003 bush fires in the NSW/VIC
-region cause a step change in the local rainfall. The actual tree cover
-change at the pixel level during this time was derived from the 15-year
-MODIS data (discussed below). The difference of tree cover before and
-after the land disturbance was tested using a Student's t-test. As the
-length of the tree cover data is shorter than available rainfall data,
-earlier land clearings in the QLD region cannot be identified spatially,
-hence they are excluded from the analysis.
+The overall hypothesis is that the effect of 2003 - 2004 land clearings
+in the QLD region and the 2003 bush fires in the NSW/VIC region cause a
+step change in the local rainfall. The actual tree cover change during
+this time at the pixel level was derived from the 15-year MODIS data
+(discussed below). The difference of tree cover before and after the
+land disturbance was tested using a Student's t-test. As the length of
+the tree cover data is shorter than available rainfall data, earlier
+land clearing in the QLD region cannot be identified spatially, hence
+they are excluded from the analysis.
 
@@
 
-Rainfall data for Australia (Jones, Wang, and Fawcett 2009) was obtained
-from the Bureau of Meteorology. The data has been projected onto a
-national 0.05\textdegree × 0.05\textdegree grid (approximately 5 km × 5
-km). This gridded data set was generated from station observations using
-an optimised Barnes successive correction technique. The Barnes
-technique combines a weighted averaging process and defined
-topographical information to estimate rainfall values between spatial
-points (BoM 2009). The resulting data set provides additional
-information for data-sparse areas like central Australia but reduces
-information in the data-rich areas, such as southeast Australia where
-station density is up to 20 per 100 \(\text{km}^2\). The data is
-available on a monthly basis from 1900 to current. Here a subset of 36
-years (1979 - 2015) was used. The study was conducted on monthly data,
-as a land cover change effect on annual rainfall might be negligible but
-can often found to be significant in particular months or seasons (e.g.
-Otterman et al. 1990; Gaertner et al. 2001; Semazzi and Song 2001;
-Oleson et al. 2004; Deo et al. 2009).
+The SILO Rainfall product data for Australia was used (Jeffrey et al.
+2001) (available online at \url{https://silo.longpaddock.qld.gov.au/}))
+. The data has been projected onto a national 0.05\textdegree by
+0.05\textdegree grid (approximately 5 km by 5 km). This gridded data set
+was generated from station observations using spline interpolation and
+kriging (Jeffrey et al. 2001). The data has been compared to other
+gridded products and observed data and is generally of high quality
+(Tozer, Kiem, and Verdon-Kidd 2009; Tozer, Kiem, and Verdon-Kidd 2012).
+The data is available on a daily and monthly basis from 1889 to current.
+Here a subset of 36 years (1979 - 2015) was used. The study was
+conducted on monthly data, as a land cover change effect on annual
+rainfall might be negligible, but can often be significant in particular
+months or seasons (e.g. Otterman et al. 1990; Gaertner et al. 2001;
+Semazzi and Song 2001; Oleson et al. 2004; Deo et al. 2009).
 
@@
 to be more important than ENSO in south Western Australia (Meneghini,
-Simmonds, and Smith 2007). The suitability of each index for the regions
-of interest was tested in section @ref(sec:reg\_model). The following
-climate indices were used as candidate predictors for local rainfall.
+Simmonds, and Smith 2007). The testing of the suitability of each index
+for the regions of interest is described in a later section. The
+following climate indices were used as candidate predictors for local
+rainfall.
 
@@
 
-The step changes on rainfall were analysed using two different
-statistical methods. A step change is not obvious in the time series
-data of the rainfall residuals (Figure \ref{fig:ts-mean}), even though
-the data is deseasonalised and detrended. Both methods in this paper
-make use of a regression model to remove variability in rainfall due to
-climatic factors to strengthen the tree cover change signal. In the
-first method, the tree cover change was implemented as a factor variable
-in the regression model. In the second method, a rank sum test (step
-trend test), was applied to the model residuals after effects of other
-major factors were removed. This assumes that after removal of all
-climate variation, the vegetation cover change is the only factor
-explaining the non-random pattern in the rainfall residuals.
+As an initial analysis a simple boxplot and t-test is used to analyse
+whether there is a significant change in tree cover in time, and before
+and after the suspected change in the regions.
+
+To assess the actual causal relationship between the tree cover and the
+rainfall a flexible regression model is applied (discussed in detail
+below). A step change is not directly obvious in the time series of the
+rainfall anomalies (Figure @ref(fig:ts-mean)) for both regions, even
+though the data is deseasonalised and detrended. In this study we apply
+different statistical methods to analyse the effect of tree cover change
+on rainfall. Both methods make use of a regression model to remove
+year-on-year variability in rainfall to strengthen the tree cover change
+signal.
+
+In the first method, the tree cover change is implemented as a factor
+variable in the regression model, and the significance of this variable
+is tested. In the second method, a rank sum test (step trend test), is
+applied to the regression model residuals after effects of other major
+factors were removed. This assumes that after removal of all climate,
+long term linear trend and seasonal variation, the vegetation cover
+change is the only factor explaining the non-random pattern in the
+rainfall residuals.
 
@@
 al. (2009) compared five large-scale drivers, including ENSO (measured
-by SOI and the Tropical Pacific SSTs), IOD, SAM, MJO (Madden-Julian
-oscillation) and blocking, in relation to Australian rainfall
-variability. The MJO is a large scale eastward-propagating wave-like
-disturbance in equatorial latitudes (Risbey et al. 2009). They
-identified SOI as the most important index among all indices tested for
-broad parts of Australia (including QLD and NSW/VIC) in almost any
-season. In this study, four indices were selected from the main climatic
-indicators (see section \protect\hyperlink{Data}{Data}) were used as the
-explanatory variables in the model for each study region.
+by SOI and the Tropical Pacific Sea surface temperatures (SSTs)), IOD,
+SAM, MJO (Madden-Julian oscillation) and blocking, in relation to
+Australian rainfall variability. The MJO is a large scale
+eastward-propagating wave-like disturbance located around equatorial
+latitudes (Risbey et al. 2009). They identified SOI as the most
+important index among all climate indices tested for broad parts of
+Australia (including QLD and NSW/VIC) in almost any season. In this
+study, four climate indices were selected from the main climatic
+indicators (see section \protect\hyperlink{Data}{Data}) and used as the
+explanatory variables in the model for each study region. A further
+complicating factor is the influence of the ``millenium drought'' over
+the study period and in particular the change to wet conditions in 2010
+- 2011 (Dijk and Viney 2013). Therefore, the spatially averaged monthly
+rainfall in the Murray Darling Basin (MDB, downloaded from
+\href{http://www.bom.gov.au/web01/ncc/www/cli_chg/timeseries/rain/allmonths/mdb/latest.txt}{the
+Bureau of Meteorology}) was used to explain the year-on-year variation
+in the rainfall in the regions. As climate alone are unlikely able to
+fully explain the temporal variation in rainfall in the regions,
+including the drought (Dijk and Viney 2013; Westra and Sharma 2010),
+this variable was included. Since both regions at least partly overlap
+with the MDB, the average rainfall across was assumed to be a useful
+explaining variable.
 
@@
 \texttt{deseasonalise} in R (R Core Team, 2018). Using detrended data
-gives a better indication of the underlying correlation rather than the
+gives a better indication of the underlying correlation compared to the
 correlation between trends in the data (Smith and Timbal 2012). The
@@
 
-Based on the correlation between the climatic indices and rainfall (as
-shown in Figure \ref{fig:cor-rain-qld} and Figure
-\ref{fig:cor-rain-nsw}), it can be concluded that:
+Based on the correlation between the climatic indices and rainfall in
+the regions (as shown in Figure @ref(fig:cor-rain-qld) and Figure
+@ref(fig:cor-rain-nsw)), it can be concluded that:
 
@@
   \setlength{\parskip}{0cm}
-  \item In QLD, the correlation between rainfall and SOI at zero time lags is the highest across all indices, outweighing the other ENSO indicators. IOD, had a weak influence in QLD.
-  \item In NSW/VIC, again the SOI has the highest correlation with rainfall, followed by the IOD. Both occur at the zero time lags. 
+  \item In QLD, the correlation between rainfall and SOI at zero time lags is the strongest across all indices, outweighing the other ENSO indicators. IOD and PDO have a weak influence in QLD.
+  \item In NSW/VIC, again the SOI has the strongest correlation with rainfall, followed by the IOD. For both, the strongest correlations occur at the zero time lags. 
   \item in both cases PDO had the weakest correlations, and this factor was not further considered as a predictor.
+  \item in general, for the better correlated indices, strongest correlations occured at zero time lags.
 \end{itemize}
@@
 indices are serially correlated with rainfall up to several months, the
-lag zero events have the most significant correlation coefficients.
-Concurrent climatic index series were generally found most useful in
+lag zero events have the greatest correlation coefficients. Using
+multiple climatic index series was generally found most useful in
 rainfall prediction (e.g. Risbey et al. 2009; Kamruzzaman, Beecham, and
-Metcalfe 2011). The correlations between the climatic indices and
-rainfall for each individual season have also been tested, and the
-results were similar.
+Metcalfe 2011).
 
@@
 We assumed all the factors are additive smooth components in determining
-rainfall following Kamruzzaman, Beecham, and Metcalfe (2011). Generally,
-monthly rainfall has a skewed distribution so the normality assumption
-of the residuals in a general linear model could be violated. In this
+rainfall following Kamruzzaman, Beecham, and Metcalfe (2011). In this
 case, the rainfall model is a generalised additive model (GAM) (Hastie
 and Tibshirani 1986) with a log link function \texttt{g()} and assuming
-the residuals are gamma distributed (see Figure \ref{fig:hist-rain}).
+the residuals are gamma distributed (see Figure @ref(fig:hist-rain)).
 This means all predictors are modelled as smooth functions, in this case
 using the shrinkage version of the cubic regression splines (Wood 2011).
-\vspace{0.5cm}
+All splines were limited to 3 knots in flexibility (Wood 2011), to
+reduce the risk of overfitting. \vspace{0.5cm}
 
@@
 \begin{array}{lll}
-g(E(\mathbf{R}_r)) = &\beta_0 + s_1(\mathbf{SOI}) + s_2(\mathbf{IOD}) + \\                &s_3(\mathbf{Nino3.4}) + s_4(\mathbf{Nino4}) + s_5(\mathbf{Season}) + \\
+g(E(\mathbf{R}_r)) = &\beta_0 + s_1(\mathbf{MDB_{monthlyRain}} + \\ &s_2(\mathbf{SOI}) + s_3(\mathbf{IOD}) + \\                &s_4(\mathbf{Nino3.4}) + s_5(\mathbf{Nino4}) + s_6(\mathbf{Season}) + \\
            &\beta_1\mathbf{Trend} + \boldsymbol{\epsilon}_r
 \end{array}
-\label{eq:model}
+(\#eq:model)
 \end{equation}
@@
 indicated by \(r\), while \(\beta_u\) (\(_u\)=0, 1) are the fitted
-coefficients in the model. \(s_v\) (\(_v\)=1, 2, 3) are the smooth
-penalized cubic regression spline functions on the climatic indices and
-the season. Apart from dropping the PDO as a predictor, all other
-climate indices were included, allowing the model to select the
-appropriate predictors.
+coefficients in the model. \(s_v\) (\(_v\)=1, 2, 3,\ldots{}) are the
+smooth penalized cubic regression spline functions on the climatic
+indices and the season. MDB\_\{monthlyRain\} is the spatially averaged
+monthly rainfall across the Murray Darling Basin. Apart from dropping
+PDO as a predictor, all other climate indices were included, allowing
+the model to select the appropriate predictors.
 
@@
 
-One of the main difficulties in empirical observation studies on the
-effect of land cover change on rainfall is the lack of continuous
-monitoring of land surface variables, or even, that no specific variable
-can be defined that can clearly represent the land surface process.
+One of the main difficulties in empirical observation studies related to
+the effect of land cover change on rainfall is the lack of continuous
+monitoring of land surface variables. Moreover, no specific variable can
+possibly be defined that can clearly represent the land surface process.
 Given the lack of a full picture of the land surface process, a factor
-variable was used in this study to represent the abrupt land surface
-change (see Equation \eqref{eq:model}). The change could be a result of
-either land clearing or bush fires as long as it is permanent or takes a
-long time to recover. Here we approached the problem with two different
-models.
+variable was used in the regression model to represent the abrupt land
+surface change (see Equation @ref(eq:model)). The change could be a
+result of either land clearing or bush fires as long as it is permanent
+or takes a long time to recover. As indicated we approached the problem
+in two different models ways.
 
@@
 significance of the coefficient of \textbf{LC}, denoted as \(\beta'_5\)
-in Equation \eqref{eq:model}, can be determined by a ratio test.
+in Equation @ref(eq:model), can be determined by a ratio test.
 
@@
   \end{array} \right.\\
-  \label{eq:dummyC}
+  (\#eq:dummyC)
 \end{equation}
@@
 \begin{array}{lll}
-g(E(\mathbf{R}_r)) = &\beta'_0 + s'_1(\mathbf{SOI}) + s'_2(\mathbf{IOD}) + \\
-  &s'_3(\mathbf{Nino3.4}) + s'_4(\mathbf{Nino4}) + s'_5\mathbf{Season} + \\
+g(E(\mathbf{R}_r)) = &\beta'_0 + s'_1(\mathbf{MDB_{monthlyRain}} + \\
+  &s'_2(\mathbf{SOI}) + s'_3(\mathbf{IOD}) + \\
+  &s'_4(\mathbf{Nino3.4}) + s'_5(\mathbf{Nino4}) + s'_6\mathbf{Season} + \\
   &\beta'_1\mathbf{Trend} + \beta'_2\mathbf{LC} + \boldsymbol{\epsilon'}_r
 \end{array}
-  \label{eq:model2}
+  (\#eq:model2)
 \end{equation}
@@
 period is missing. Therefore, four scenarios were initially tested in
-the analysis. In these scenarios the ``after change'' period started
-from: (1) June 2003, (2) January 2004, (3) June 2004 and (4) January
-2005. A further complicating factor is the influence of the ``millenium
-drought'' over the study period and in particular to change to wet
-conditions in 2010 - 2011 (Dijk and Viney 2013). In the NSW/VIC region,
-severe bush fires were reported in early January 2003. Hence the
+the analysis (\textbf{see supplementary material}). In the NSW/VIC
+region, severe bush fires were reported in early January 2003. Hence the
 ``tree'' cover state was up to December 2002 then it was changed to
-``removed'' state from January 2003. However, there is similar
-uncertainty about when the tree cover had recovered from the bushfire
-and how this interacted with climate variability. As a result several
-scenarios were tested for both locations: 1) analysis up to 7 years post
-change (assumed to be 2010 for both regions); and 2) analysis of the
-full period (2015). More detail about this is later in the methods. As a
-starting date, the regression model was run from 1979 for both regions.
+``removed'' state from January 2003. As a starting date, the regression
+model was run from 1979 for both regions.
 
@@
 The rainfall residuals from the regression model in Equation
-\eqref{eq:model2} were used for the step trend test. According to (Hirsch
+@ref(eq:model2) were used for the step trend test. According to (Hirsch
 and Gilroy 1985), to detect a step change, using deseasonalised and
@@
 year of the millenium drought (Dijk and Viney 2013), followed by two
-very wet years (also visible in Figure \ref{fig:ts-mean} in 2010 and
+very wet years (also visible in Figure @ref(fig:ts-mean) in 2010 and
 2011. As highlighted earlier, this means the the length of after period
@@
   W_{jk} = \sum_{i=1}^{n_1}R'_{ijk}.
-  \label{eq:Wj}
+  (\#eq:Wj)
 \end{equation}
@@
   Z'=\frac{\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk}-12\cdot ns\cdot\mu_w}{\sqrt{Var(\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk})}}.
-  \label{eq:Z}
+  (\#eq:Z)
 \end{equation}
@@
 trend test can be interpreted according to the sign of the Z' score (see
-Table \ref{tab:Zscore} (Chapter 23, P887 Hipel and McLeod 1994)). Z' is
+Table @ref(tab:Zscore) (Chapter 23, P887 Hipel and McLeod 1994)). Z' is
 normally distributed similar to the standard normal statistics Z. Hence
@@
 significant (\(p \leq 0.05\)) in each study region, are shown in Figure
-\ref{fig:tctrend} for the time series to 2009 (left panels) and the time
+@ref(fig:tctrend) for the time series to 2009 (left panels) and the time
 series to 2015 (right panels). In both cases it can be seen that the
@@
 of this region in the series to 2009. The tree cover change map
-(\ref{fig:tctrend} is consistent with the annual mean EVI trend map
+(@ref(fig:tctrend) is consistent with the annual mean EVI trend map
 (based on DLCD data, not shown here). However in the series to 2015 the
 tree cover loss seems to have dissappeared (top right Figure
-\ref{fig:tctrend}) and suggests an increase in tree cover over the area.
+@ref(fig:tctrend)) and suggests an increase in tree cover over the area.
 The implications of this will be discussed in more detail in the
@@
 Generally the regression model only explains a limited amount of the
-rainfall variability. The model in Equation \eqref{eq:model2} accounts for
+rainfall variability. The model in Equation @ref(eq:model2) accounts for
 around
@@
 that the assumptions of the regression model are generally met (Figure
-\ref{fig:residuals}. The standardised residual plots, however, show some
+@ref(fig:residuals). The standardised residual plots, however, show some
 funnelling for the NSW/VIC regions, suggesting non constant variance.
@@
 zero) accounted, on average, for 6.7\% of the rainfall variability in
-both the QLD region and the NSW/VIC region (see Figure \ref{fig:rsq} for
+both the QLD region and the NSW/VIC region (see Figure @ref(fig:rsq) for
 the distribution of \(R^2\) in these two regions). These figures are
@@
 
-The land cover variable in the model (Equation \eqref{eq:model2}) aims to
+The land cover variable in the model (Equation @ref(eq:model2)) aims to
 identify a step change in the rainfall before and after the observed
@@
 (\(p \leq 0.05\)) for the rainfall estimates in some areas in NSW/VIC,
-as shown in Figure \ref{fig:LCp}. However, the number of pixels where
+as shown in Figure @ref(fig:LCp). However, the number of pixels where
 the landcover variable was significant was much greater for the series
@@
 relationship between the areas of bushfires in Figure
-\ref{fig:bushfire}. No significant step change due to the land cover
+@ref(fig:bushfire). No significant step change due to the land cover
 changes was found in rainfall in the QLD region in the series to 2015,
@@
 
-Figure \ref{fig:LCp} for NSW/VIC suggests that a significant step change
+Figure @ref(fig:LCp) for NSW/VIC suggests that a significant step change
 in rainfall related to land cover change was found in an area larger
@@
 The spatial step trend test Z' scores for the Qld region (top row) and
-NSWVIC region (bottom row) are shown in Figure \ref{fig:steptest30}. The
+NSWVIC region (bottom row) are shown in Figure @ref(fig:steptest30). The
 figure also contains the period to 2009 (left panels) and the period to
@@
 sign and the significance level. The sign indicates the direction of the
-step change, as listed in Table \ref{Zscore}. In each region up to 2009,
+step change, as listed in Table @ref(Zscore). In each region up to 2009,
 there is a broad area of positive Z' values which implies a decrease in
@@
 The rainfall residuals of the two periods (before-change since 1979 and
-after-change) were also compared (Figure \ref{fig:meandiff}) using a
+after-change) were also compared (Figure @ref(fig:meandiff)) using a
 simple t-test. From the boxplots it is difficult to see that there were
@@
 consistently very low in 2006 for both regions and consistently high for
-2010 \ref{fig:ts-mean}. The low rainfall in 2006 was due to a weak El
+2010 @ref(fig:ts-mean). The low rainfall in 2006 was due to a weak El
 Niño and unrecovered conditions from the previous drought, in contrast,
@@
 totals, only the period to 2009 is of interest, as the full period had
-too few significant pixels (\ref{fig:steptest30}). For both regions and
+too few significant pixels (@ref(fig:steptest30)). For both regions and
 excluding 2006, the Qld area had significantly more rain before than
@@
 \section{Discussion}\label{discussion}
+
+\begin{verbatim}
+## Warning: package 'kableExtra' was built under R version 3.5.2
+\end{verbatim}
 
@@
 While there is some indication that the observed landuse changes (Figure
-\ref{fig:tctrend}) cause a decrease in the rainfall, our analysis has
+@ref(fig:tctrend)) cause a decrease in the rainfall, our analysis has
 not been able to give an unequivocal answer. There are several possible
@@
 only around 7\% on average is due to the climate drivers (Figure
-\ref{fig:rsq} . And while this is consistent with the literature e.g.
+@ref(fig:rsq) . And while this is consistent with the literature e.g.
 Westra and Sharma (2010), a large amount of variation is left in the
@@
 2009 was at the end of the drought, the series to 2015 includes the
-record breaking year 2010 (Figure \ref{fig:ts-mean}).
+record breaking year 2010 (Figure @ref(fig:ts-mean)).
 
@@
 clearings occurred in 1999 - 2000 and 2002 - 2004 (Figure
-\ref{fig:slat}). And even though there was a decrease in land clearing
+@ref(fig:slat)). And even though there was a decrease in land clearing
 post 2005, it is difficult to define a clear cut change in this region.
@@
 Bray 2016), and this could explain the change in vegetation cover
-between 2009 and 2015 in the Qld area (Figure \ref{fig:tctrend}). In
+between 2009 and 2015 in the Qld area (Figure @ref(fig:tctrend)). In
 particular the favourable rainfall years of 2010 and 2011 would have
@@
 QLD region than in the NSW/VIC region before land surface interventions.
-The significant bushfires (Figure \ref{fig:fig2bushfire}) would have
+The significant bushfires (Figure @ref(fig:fig2bushfire)) would have
 drastically reduced the vegetation cover and recovery was very slow in
@@
 Particularly in the NSW/VIC area, there is no direct overlap between the
-pixels with significant vegetation change (Figure \ref{fig:tctrend}) and
+pixels with significant vegetation change (Figure @ref(fig:tctrend)) and
 the pixels with significant land cover variables or positive Z'scores
-(Figure \ref{fig:LCp} and Figure \ref{fig:steptest30}). While this seems
+(Figure @ref(fig:LCp) and Figure @ref(fig:steptest30)). While this seems
 possibly surprising at first, a plausible explanation could be the
@@
 
-\hypertarget{ref-BoM2009}{}
-BoM. 2009. ``Gridded Monthly Rainfall Metadata.'' Australian Bureau of
-Meteorology.
-\url{http://www.bom.gov.au/climate/austmaps/metadata-monthly-rainfall.shtml}.
-
 \hypertarget{ref-BoM2012a}{}
----------. 2012a. ``Australia - Climate of Our Continent.''
+BoM. 2012a. ``Australia - Climate of Our Continent.''
 
@@
 
+\hypertarget{ref-SILO2001}{}
+Jeffrey, Stephen J., John O. Carter, Keith B. Moodie, and Alan R.
+Beswick. 2001. ``Using Spatial Interpolation to Construct a
+Comprehensive Archive of Australian Climate Data.'' Journal Article.
+\emph{Environmental Modelling \& Software} 16 (4 SU -): 309--30.
+\url{http://www.sciencedirect.com/science/article/B6VHC-42SXF8F-2/2/b5c0a24b85149f40ba5d2419ddc0999c}.
+
 \hypertarget{ref-Jones2009}{}
@@
 
+\hypertarget{ref-Beesley2009}{}
+Tozer, C. R., A. S. Kiem, and D. C. Verdon-Kidd. 2009. ``On the
+Uncertainties Associated with Using Gridded Rainfall Data as a Proxy for
+Observed.'' Conference Paper. In \emph{18th World Imacs Congress and
+Modsim09 International Congress on Modelling and Simulation.}, edited by
+R.S. Anderssen, R.D. Braddock, and L.T.H. Newham, 3886--92.
+\href{http://www.mssanz.org.au/modsim09/I13/beesley.pdf\%20}{http://www.mssanz.org.au/modsim09/I13/beesley.pdf}.
+
+\hypertarget{ref-Tozer2012}{}
+---------. 2012. ``On the Uncertainties Associated with Using Gridded
+Rainfall Data as a Proxy for Observed.'' Journal Article. \emph{Hydrol.
+Earth Syst. Sci.} 16 (5): 1481--99.
+doi:\href{https://doi.org/10.5194/hess-16-1481-2012}{10.5194/hess-16-1481-2012}.
+
 \hypertarget{ref-Trenberth1999}{}

### 2019-02-28 Scripts/CorrelationClimateRainfall2018.Rmd
update for PeerJ paper

---
title: "Correlation Analysis climate and rainfall"
author: "Willem Vervoort, Chun Liang, Floris van Ogtrop"
date: "17 December 2018"
output: 
    pdf_document:
      fig_width: 7
      fig_height: 6
      fig_caption: true
---
```{r setup, warning=F, message=F, echo=F}
# root dir
knitr::opts_knit$set(root.dir =  
"E:/Cloudstor/Chun/Chun")

library(knitr)
require(tidyverse)
require(lubridate)
require(raster)
require(lme4)
# require(deseasonalize)
## ----rcurl----------------------------------------------------------
require(curl)
require(ncdf4)

```

# Introduction  
This document only covers the analysis of the rainfall and climate in terms of correlation and basic statistics. This is a rewrite of the original paper analysis, but uses the SILO data rather than the BOM gridded data to check if this makes a difference.

# load the climate index data

All the older climate index data

```{r climaticIndex}

# climate indicator data
file1 <- "Data\\BoM\\SOI\\SOI.csv"
SOI <- read_csv(file1)
SOI <- gather(SOI,key="Month", value="SOI",Jan:Dec)
SOI <- SOI[order(as.Date(paste(SOI$Year,SOI$Month,
                                        "01",sep="-"), "%Y-%b-%d")),]
SOI_37Y <- SOI %>%
  filter(Year > 1978 & Year < 2016)
#SST34
file2 <- "Data\\BoM\\SST\\SST34_anomaly.nc"
SST34 <- nc_open(file2)
SST34_var <- ncvar_get(SST34,SST34$var[[1]])
nc_close(SST34)
SST34_df <- tibble(Date = seq.Date(as.Date("1856-01-01"),
                                   as.Date("2018-04-01"),
                                   by="month"),
                   SST34 = SST34_var)
SST34_37Y <- SST34_df %>%
  filter(Date >= "1979-01-01" & Date < "2016-01-01")

# SST3
file3 <- "Data\\BoM\\SST\\SST3_anomaly.nc"
SST3 <- nc_open(file3)
SST3_var <- ncvar_get(SST3,SST3$var[[1]])
nc_close(SST3)
SST3_df <- tibble(Date = seq.Date(as.Date("1856-01-01"),
                                   as.Date("2018-04-01"),
                                   by="month"),
                   SST3 = SST3_var)
SST3_37Y <- SST3_df %>%
  filter(Date >= "1979-01-01" & Date < "2016-01-01")

#SST4
file4 <- "Data\\BoM\\SST\\SST4_anomaly.nc"
SST4 <- nc_open(file4)
SST4_var <- ncvar_get(SST4,SST4$var[[1]])
nc_close(SST4)
SST4_df <- tibble(Date = seq.Date(as.Date("1856-01-01"),
                                   as.Date("2018-04-01"),
                                   by="month"),
                   SST4 = SST4_var)
SST4_37Y <- SST4_df %>%
  filter(Date >= "1979-01-01" & Date < "2016-01-01")

# PDO
file5 <- "Data\\BoM\\PDO\\PDO.latest_1900-.txt"
pdo_mth <- read_table(file5)
pdo_mth <- gather(pdo_mth,key="Month", value="PDO", JAN:DEC)
pdo_mth$YEAR <- substr(pdo_mth$YEAR,1,4)
pdo_mth <- pdo_mth[order(as.Date(paste(pdo_mth$YEAR,pdo_mth$Month,
                                        "01",sep="-"), "%Y-%b-%d")),]
PDO_37Y <- pdo_mth %>%
  filter(YEAR > 1978 & YEAR < 2016)
## deseasonalised PDO
PDO_100 <- pdo_mth

#IOD
file7 <- "Data\\BoM\\IOD\\iod_jamstec.txt"
IOD <- read_table2(file7)
IOD <- tibble(Date = seq.Date(as.Date("1870-01-16"),
                                   as.Date("2018-03-16"),
                                   by="month"),
              IOD = IOD$`DMI(=West-East)`)
IOD_37Y <- IOD %>%
  filter(Date > "1979-01-01" & Date < "2016-01-01")
# end of read data
# should write out useful data
#save.image("../../data/data_environment.Rdata")
```

# Download and manage the rainfall data

This deals with 100 years of data, deseasonalise and analyse correlations, subset in 37 years and 100 years

```{r Rainfall}

# SILO data

## ----extract_Silo-----------------------------------------------------
years <- seq(1901,2016,by=1)

SILOdata <- list()

# run a loop
for (i in 1:length(years)) {
   # url <- paste("https://s3-ap-southeast-2.amazonaws.com/silo-open-data/annual/monthly_rain/",
   #              years[i],
   #              ".monthly_rain.nc",sep="")
   # curl_download(url,
   #               paste("data/monthly_rainfall_",years[i],".nc",sep=""))
  SILOdata[[i]] <- brick(paste("data/monthly_rainfall_",years[i],".nc",
                          sep=""),varname="monthly_rain")
}

SILOdata__b <- do.call(stack, SILOdata)

#crop to extents
e111 <- extent(c(145.825, 148.625, -28.075, -25.575)) 
e222 <- extent(c(147.075, 148.925, -36.675, -35.075))

#  NSW
SILOdata.NSW <- crop(SILOdata__b, extent(e222)) 
#QLD
SILOdata.QLD <- crop(SILOdata__b, extent(e111))
# remove orig
rm(SILOdata)
rm(SILOdata__b)

# rewrite as a monthly dataframe
# NSW
rainNSW_t <- as.tibble(data.frame(Date =  seq.Date(as.Date("1901-01-01"),
                                   as.Date("2016-12-31"),by="month"),
                                  rain=t(SILOdata.NSW@data@values)))

# calculate mean by row
nsw_mean <- rainNSW_t %>%
  mutate(Means = rowMeans(.[,-1])) %>%
  dplyr::select(Date,Means)

nsw37_mean <- nsw_mean %>%
 filter(Date >= "1979-01-01")

# Qld
rainQld_t <- as.tibble(data.frame(Date =  seq.Date(as.Date("1901-01-01"),
                                   as.Date("2016-12-31"),by="month"),
                                  rain=t(SILOdata.QLD@data@values)))

# calculate mean by row
qld_mean <- rainQld_t %>%
  mutate(Means = rowMeans(.[,-1])) %>%
  dplyr::select(Date,Means)

qld37_mean <- qld_mean %>%
 filter(Date >= "1979-01-01")

rm(SILOdata.NSW)
rm(SILOdata.QLD)

# some simple stats graphs
#jpeg("../../figures/hist_rainfallSilo.jpg", width=960, height=480)
par(mfrow=c(1,2))
hist(qld_mean$Means,main="",xlab=expression(paste("rainfall (mm mont",h^-1,")",sep="")))
 hist(nsw_mean$Means,main="",xlab=expression(paste("rainfall (mm mont",h^-1,")",sep="")))
#dev.off()

# autocorrelation in rainfall
acf_df <- data_frame(lag = 1:12, 
                     acf_Qld = acf(qld_mean$Means,
                                   lag.max=12,main="QLD")$acf[2:13,1,1],
                     acf_NSW = acf(nsw_mean$Means,
                                   lag.max=12,main="NSW & VIC")$acf[2:13,1,1],
                     pacf_Qld = pacf(qld_mean$Means,
                                     lag.max=12,main="")$acf[,1,1],
                     pacf_NSW = pacf(nsw_mean$Means,
                                     lag.max=12,main="")$acf[,1,1])

#jpeg("../../figures/autocorrelationSilo.jpg")
acf_df %>%
  gather(key="acf", value="cor", acf_Qld:pacf_NSW) %>%
  ggplot(aes(lag, cor)) + geom_bar(stat="identity", fill = "blue") +
  facet_wrap(~acf) + xlab("Lag") + ylab("correlation")
#dev.off()

```

## deseasonalise and correlations with climate indices

We can now deseasonalise using the package deseasonalise

```{r deseasonalise}
require(deseasonalize)
## deseasonalise & detrend rainfall: assume the deseasonalised monthly zonal rainfall is normal distributed and can be used in pearson cross-correlation function

qld37_deseason <- ds(qld37_mean$Means,ic="AIC", type="monthly",
                     standardizeQ=F)$z
nsw37_deseason <- ds(nsw37_mean$Means,ic="AIC", type="monthly",
                     standardizeQ=F)$z
qld100_deseason <- ds(qld_mean$Means,ic="AIC", type="monthly",
                      standardizeQ=F)$z
nsw100_deseason <- ds(qld_mean$Means,ic="AIC", type="monthly",
                      standardizeQ=F)$z

# Show a plot
plot_df <- tibble(Dates = rep(qld37_mean$Date,2),
                  Resid = c(qld37_deseason,nsw37_deseason),
                  State = c(rep("Qld",length(qld37_deseason)),
                            rep("NSW",length(nsw37_deseason))))

#jpeg("../../figures/Rainfall_resid.jpg", width=960, height=960)
plot_df %>%
  ggplot(aes(Dates,Resid)) + geom_line() + 
  geom_vline(xintercept = as.Date("2003-01-01"), 
             col="red", size=1.5) +
  geom_text(aes(x=as.Date("2004-12-01"),y=150),
            label="2003", col="red", size=7) +
  facet_wrap(~State, ncol = 1) + xlab("Date") +
  ylab("Deseasonalised rainfall for the NSW and Qld locations") +
  theme(axis.text=element_text(size=rel(2)),
        axis.title=element_text(size=rel(2.5)),
        strip.text =element_text(size=rel(2), face="bold"))
#dev.off()

```

### trends in the deseasonalised rainfall

Investigate if any of the detrended rainfall has a trend

```{r}
# test whether any linear trends are significant
df_list <- list(qld37 = data.frame(z=qld37_deseason,
                           trend=1:length(qld37_deseason)),
                nsw37 = data.frame(z=nsw37_deseason,
                           trend=1:length(nsw37_deseason)),
                qld100 = data.frame(z=qld100_deseason,
                 trend=1:length(qld100_deseason)),
                nsw100 = data.frame(z=nsw100_deseason,
                 trend=1:length(nsw100_deseason)))
result <- lapply(df_list,
                 function(x) summary(lm(z~trend,
                                        data=x))$coefficients)
result
```

None have a significant trend

### correlations with climate indices

Correlate the deseasonalised rainfall with the climate indices

```{r}
# Figure 5
plot_df_qld <- tibble(Lag = ccf(as.vector(SOI_37Y$SOI),
                            as.vector(qld37_deseason),lag.max=12, 
                            plot=F)$lag[,,1],
                  SOI = ccf(as.vector(SOI_37Y$SOI),
                            as.vector(qld37_deseason),lag.max=12, 
                            plot=F)$acf[,,1],
                  NINO3.4 = ccf(as.vector(SST34_37Y$SST34),
                                as.vector(qld37_deseason),lag.max=12,
                                plot=F)$acf[,,1],
                  NINO4 = ccf(as.vector(SST4_37Y$SST4),
                              as.vector(qld37_deseason),lag.max=12,
                              plot=F)$acf[,,1],
                  PDO = ccf(as.vector(PDO_100$PDO),
                            as.vector(qld100_deseason),lag.max=12,
                            plot=F)$acf[,,1],
                  IOD = ccf(as.vector(IOD_37Y$IOD),
                            as.vector(qld37_deseason),lag.max=12, 
                            plot=F)$acf[,,1])

#jpeg("../../figures/cor_qld.jpg", width=960, height=960)
plot_df_qld %>%
  gather(key="SST", value="correlation", SOI:IOD) %>%
  ggplot(aes(Lag,correlation)) + 
  geom_bar(fill="blue", stat="identity") +
  facet_wrap(~SST) + xlab("Lag in months") + ylab("correlation") +
  xlim(c(-12.5,0.5)) +
  theme(axis.text=element_text(size=rel(2)),
        axis.title=element_text(size=rel(2.5)),
        strip.text =element_text(size=rel(2), face="bold"))
#dev.off()

# to add and calculate ci: qnorm((1 + 0.95)/2)/sqrt(x$n.used)

plot_df <- tibble(Lag = ccf(as.vector(SOI_37Y$SOI),
                            as.vector(nsw37_deseason),lag.max=12, 
                            plot=F)$lag[,,1],
                  SOI = ccf(as.vector(SOI_37Y$SOI),
                            as.vector(nsw37_deseason),lag.max=12, 
                            plot=F)$acf[,,1],
                  NINO3.4 = ccf(as.vector(SST34_37Y$SST34),
                                as.vector(nsw37_deseason),lag.max=12,
                                plot=F)$acf[,,1],
                  NINO4 = ccf(as.vector(SST4_37Y$SST4),
                              as.vector(nsw37_deseason),lag.max=12,
                              plot=F)$acf[,,1],
                  PDO = ccf(as.vector(PDO_100$PDO),
                            as.vector(nsw100_deseason),lag.max=12,
                            plot=F)$acf[,,1],
                  IOD = ccf(as.vector(IOD_37Y$IOD),
                            as.vector(nsw37_deseason),lag.max=12, 
                            plot=F)$acf[,,1])

#jpeg("../../figures/cor_nswvic.jpg", width=960, height=960)
plot_df %>%
  gather(key="SST", value="correlation", SOI:IOD) %>%
  ggplot(aes(Lag,correlation)) + 
  geom_bar(fill="blue", stat="identity") +
  facet_wrap(~SST) + xlab("Lag in months") + ylab("correlation") +
  xlim(c(-12.5,0.5)) +
  theme(axis.text=element_text(size=rel(2)),
        axis.title=element_text(size=rel(2.5)),
        strip.text =element_text(size=rel(2), face="bold"))
#dev.off()

```

From the paper:

Correlations between rainfall and each climate index were analysed. Rainfall in each study region was first deseasonalised and detrended using the seasonal decomposition function `ds` in the package `deseasonalise` in R [@Rstats2018]. Using detrended data gives a better indication of the underlying correlation rather than the correlation between trends in the data [@Smith2012]. 
The cross-correlations between the deseasonalised and detrended rainfall and the climatic indices were tested using the Pearson's product moment correlation method, assuming the relationships are linear. Although the optimal technique for exploring the correlation with each index could be different as described in @Risbey2009, the Pearson's method was applied to all indices for consistency. Because the PDO describes the multi-decadal SST with lower frequency [@MacDonald2005; @Zanchettin2008; @Kamruzzaman2011], instead of 37-year rainfall data, a longer period (115 years, from 1900 to 2015) was used to estimate the correlation with PDO, up to lag 24. For the other indices, the 37-year data was used.

### 2019-02-28 Scripts/ReanalyseData2018.Rmd
update for PeerJ paper

---
title: "Chun's paper restart analysis"
author: "Willem Vervoort, Chun Liang, Floris van Ogtrop"
date: "`r format(Sys.Date(),'%d-%m-%Y')`"
output: 
    pdf_document:
      fig_width: 7
      fig_height: 6
      fig_caption: true
---
```{r setup, warning=F, message=F, echo=F}
# root dir
knitr::opts_knit$set(root.dir =  
"E:/Cloudstor/Chun/Chun")

library(knitr)
require(tidyverse)
require(lubridate)
require(raster)
require(lme4)
## ----rcurl----------------------------------------------------------
require(curl)
require(ncdf4)
require(maptools)
require(mgcv)
require(fields)
require(gridExtra)
require(RColorBrewer)
require(rgeos)
```

```{r mapping, echo = F, eval=T}

## define the boundary of sub-regions
   e1 <- extent(c(1324304.091,1723901.773,-222556.668,80451.265))
   e2 <- extent(c(1346395.141,1534425.527,-1169402.246,-607644.171))
  # 
   e11 <- extent(c(145.425, 149.025, -28.475, -25.175))
   e22 <- extent(c(146.675, 149.325, -37.075, -34.675))

  e111 <- extent(c(145.825, 148.625, -28.075, -25.575)) 
  e222 <- extent(c(147.075, 148.925, -36.675, -35.075))

## the selected regions for mapping
root_dir <- "e:/cloudstor/chun/chun/data"
aus_reg <- readShapePoly(paste(root_dir,
                               "Geo\\aus_state2011\\STE11aAust.shp",sep="/"),
                         proj4string=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0"))
aus_reg2 <- gSimplify(aus_reg,0.01,topologyPreserve=T)

mdb_reg <- readShapePoly(paste(root_dir,
                               "Geo\\MDB\\mdb_boundary.shp", sep="/"),
                               proj4string=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0"))
mdb_reg2 <- gSimplify(mdb_reg,0.01,topologyPreserve=TRUE)

#remove(list=c("aus_reg","mdb_reg"))

```

# Introduction

This is a rewrite of the paper to tackle the issues that are still remaining.

Here is the general approach proposed:

1. Run a regression model on the tree cover data (mixed model): $Tree_cover ~ time(years) + AR1(error)$. The H_0 would be that the tree cover has not changed in time. This model would also more clearly identify those years where change in tree cover is likely (but cannot confirm this).  
2. Run a regression model that removes the variation in monthly rainfall and tests the significance of a "tree_cover" variable. This model also includes the monthly rainfall averaged over the Murray Darling Basin, which might mean that variables such as "season" is no longer needed. $Rainfall ~ Monthly_{MDBRain} +  SST + trend_{treecover} + trend_{time} + AR1(error)$    
3. Calculate monthly rainfall residuals using a regression model (based on the above model, but dropping the tree cover trend term): $RainfallResiduals ~ Rainfall - Monthly_{MDBRain} + SST + trend_{time}$  
4. Running Mann-Whitney or Pettitt tests on the rainfall residuals. However, Serinaldi et al argue that the MK and Pettitt tests are equivalent on timeseries data (give very similar results).

# Step 1, Tree cover change

First we load the data, using the older code
```{r treeCover,eval=T , echo = F}

rootdir <-"E:/cloudstor/chun2/data/satellite/MOD44/B"

# read in the file list from the directory
x <- dir(path=rootdir,pattern="Percent_Tree_Cover.tif")
# Identify the different states in the images
# NSW
NSW <- x[grep("NSW",x)]
# Qld
Qld <- x[grep("Qld",x)]

img <- list()
# raster and brick
for(i in 1:length(NSW)) {
  img[[i]] <- raster(paste(rootdir,NSW[i],sep="/"))
}

imgQ <- list()
# raster and brick
for(i in 1:length(Qld)) {
  imgQ[[i]] <- raster(paste(rootdir,Qld[i],sep="/"))
}

NSW_b <- do.call(brick,img)
Qld_b <- do.call(brick, imgQ)

# NSW
tm <- seq(as.Date("2000-03-06"), as.Date("2016-03-06"), "year")
mod_tc_r22 <- setZ(NSW_b, tm, "year")
# Qld
mod_tc_r11 <- setZ(Qld_b, tm, "year")

```

### plot for NSW

```{r tree_coverNSW}

NSW_data <- as.tibble(rasterToPoints(mod_tc_r22))
colnames(NSW_data)[3:19] <- as.character(tm)

NSWdata <- NSW_data %>%
  gather(key="Year", value="Tree_cover", `2000-03-06`:`2016-03-06`) %>%
  filter(Tree_cover < 200)
NSWdata %>%
  ggplot(aes(as.factor(Year),Tree_cover)) + geom_boxplot() +
  xlab("Year") + ylab("Tree cover %") +
  ggtitle("Percent tree cover change NSW") +
  theme(axis.text.x = element_text(angle=45, size=rel(1.5)),
        axis.text.y = element_text(size=rel(1.5)),
        axis.title = element_text(size=rel(2), face = "bold"))

mod_NSW <- lmer(Tree_cover ~ as.factor(Year) + (1|Year),
               data = NSWdata)
summary(mod_NSW)

```

This shows that in NSW the years 2003 and 2004 were likley to be different from the mean at p < 0.10, and they were less than the mean. all the other years had higher p values (and are thus less likely to be different from the mean). These would coincide with the bushfire.

### plot for QLD

```{r tree_coverQLD}

QLD_data <- as.tibble(rasterToPoints(mod_tc_r11))
colnames(QLD_data)[3:19] <- as.character(tm)

QLDdata <- QLD_data %>%
  gather(key="Year", value="Tree_cover", `2000-03-06`:`2016-03-06`) %>%
  mutate(Tree_cover = replace(Tree_cover,Tree_cover == 200, NA))
QLDdata %>%
  ggplot(aes(as.factor(Year),Tree_cover)) + geom_boxplot() +
  xlab("Year") + ylab("Tree cover %") +
  ggtitle("Percent tree cover change QLD") +
  theme(axis.text.x = element_text(angle=45, size=rel(1.5)),
        axis.text.y = element_text(size=rel(1.5)),
        axis.title = element_text(size=rel(2), face = "bold"))

mod_QLD <- lmer(Tree_cover ~ as.factor(Year) + (1|Year),
               data = QLDdata)
summary(mod_QLD)

```

For Qld only the years 2010 and 2012 are likely to be different from the mean (p < 0.10) and were consistently higher than the mean in % cover, they were also two very wet years.

# Step 2 Rainfall regression model

This mainly repeats the analysis that has been done in the earlier versions of the paper.  But there are some changes:

- the correlation analysis is done in another script "CorrelationClimateRainfall2018.Rmd"  
- We have decided to use SILO data as the ANUCLIM (finer grid data) only extends to 2014 and therefore we cannot capture 2016

## load the data

All the older climate index data

```{r climaticIndex,eval=T , echo = F}

# climate indicator data
file1 <- "Data\\BoM\\SOI\\SOI.csv"
SOI <- read_csv(file1)
SOI <- gather(SOI,key="Month", value="SOI",Jan:Dec)
SOI <- SOI[order(as.Date(paste(SOI$Year,SOI$Month,
                                        "01",sep="-"), "%Y-%b-%d")),]
SOI_37Y <- SOI %>%
  filter(Year > 1978 & Year <= 2016)
#SST34
file2 <- "Data\\BoM\\SST\\SST34_anomaly.nc"
SST34 <- nc_open(file2)
SST34_var <- ncvar_get(SST34,SST34$var[[1]])
nc_close(SST34)
SST34_df <- tibble(Date = seq.Date(as.Date("1856-01-01"),
                                   as.Date("2018-04-01"),
                                   by="month"),
                   SST34 = SST34_var)
SST34_37Y <- SST34_df %>%
  filter(Date >= "1979-01-01" & Date < "2017-01-01")

# SST3
file3 <- "Data\\BoM\\SST\\SST3_anomaly.nc"
SST3 <- nc_open(file3)
SST3_var <- ncvar_get(SST3,SST3$var[[1]])
nc_close(SST3)
SST3_df <- tibble(Date = seq.Date(as.Date("1856-01-01"),
                                   as.Date("2018-04-01"),
                                   by="month"),
                   SST3 = SST3_var)
SST3_37Y <- SST3_df %>%
  filter(Date >= "1979-01-01" & Date < "2017-01-01")

#SST4
file4 <- "Data\\BoM\\SST\\SST4_anomaly.nc"
SST4 <- nc_open(file4)
SST4_var <- ncvar_get(SST4,SST4$var[[1]])
nc_close(SST4)
SST4_df <- tibble(Date = seq.Date(as.Date("1856-01-01"),
                                   as.Date("2018-04-01"),
                                   by="month"),
                   SST4 = SST4_var)
SST4_37Y <- SST4_df %>%
  filter(Date >= "1979-01-01" & Date < "2017-01-01")

# PDO
file5 <- "Data\\BoM\\PDO\\PDO.latest_1900-.txt"
pdo_mth <- read_table(file5)
pdo_mth <- gather(pdo_mth,key="Month", value="PDO", JAN:DEC)
pdo_mth$YEAR <- substr(pdo_mth$YEAR,1,4)
pdo_mth <- pdo_mth[order(as.Date(paste(pdo_mth$YEAR,pdo_mth$Month,
                                        "01",sep="-"), "%Y-%b-%d")),]
PDO_37Y <- pdo_mth %>%
  filter(YEAR > 1978 & YEAR <= 2016)
## deseasonalised PDO
# PDO_100 <- ds(PDO_100a$PDO,ic = "AIC", type = "monthly",
#               standardizeQ=F)$z

#IOD
file7 <- "Data\\BoM\\IOD\\iod_jamstec.txt"
IOD <- read_table2(file7)
IOD <- tibble(Date = seq.Date(as.Date("1870-01-16"),
                                   as.Date("2018-03-16"),
                                   by="month"),
              IOD = IOD$`DMI(=West-East)`)
IOD_37Y <- IOD %>%
  filter(Date > "1979-01-01" & Date < "2017-01-01")
# end of read data
# should write out useful data
#save.image("../../data/data_environment.Rdata")
```

### Rainfall

Downloading from SILO. Can be replaced by ANUCLIM if the data to 2016 becomes available

```{r Rainfall}

# SILO data

## ----extract_Silo-----------------------------------------------------
years <- seq(1979,2016,by=1)

SILOdata <- list()

# run a loop
for (i in 1:length(years)) {
   # url <- paste("https://s3-ap-southeast-2.amazonaws.com/silo-open-data/annual/monthly_rain/",
   #              years[i],
   #              ".monthly_rain.nc",sep="")
   # curl_download(url,
   #               paste("data/monthly_rainfall_",years[i],".nc",sep=""))
  SILOdata[[i]] <- brick(paste("data/monthly_rainfall_",years[i],".nc",
                          sep=""),varname="monthly_rain")
}

SILOdata__b <- do.call(stack, SILOdata)

#crop to extents
e111 <- extent(c(145.825, 148.625, -28.075, -25.575)) 
e222 <- extent(c(147.075, 148.925, -36.675, -35.075))

#  NSW
SILOdata.NSW <- crop(SILOdata__b, extent(e222)) 
#QLD
SILOdata.QLD <- crop(SILOdata__b, extent(e111))
# MDB
mdb_reg <- readShapePoly("data/Geo\\MDB\\mdb_boundary.shp",
                               proj4string=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0"))
SILOdata.MDB <- crop(SILOdata__b, extent(mdb_reg))

aus_reg <- readShapePoly("data/Geo\\aus_state2011\\STE11aAust.shp",
                         proj4string=CRS("+proj=lcc +lat_1=-18 +lat_2=-36"))

# remove orig
rm(SILOdata)
rm(SILOdata__b)
# test -----------------------------
# require(rasterVis)
# gplot(SILOdata.NSW[[1]])  + geom_tile(aes(fill = value)) + coord_equal() +
#   facet_wrap(~variable) + scale_fill_gradient(low="darksalmon", high="blue",
#                                         guide="colorbar",na.value="white") +
#   xlab("Longitude") + ylab("Latitude")
# ------------------------

# rewrite as a monthly dataframe
# NSW
rainNSW_t <- as.tibble(data.frame(Date =  seq.Date(as.Date("1979-01-01"),
                                   as.Date("2016-12-31"),by="month"),
                                  rain=t(SILOdata.NSW@data@values)))

# calculate mean by row
nsw_mean <- rainNSW_t %>%
  mutate(Means = rowMeans(.[,-1], na.rm = T)) %>%
  dplyr::select(Date,Means)

rm(SILOdata.NSW)

# Qld
rainQld_t <- as.tibble(data.frame(Date =  seq.Date(as.Date("1979-01-01"),
                                   as.Date("2016-12-31"),by="month"),
                                  rain=t(SILOdata.QLD@data@values)))

# calculate mean by row
qld_mean <- rainQld_t %>%
  mutate(Means = rowMeans(.[,-1], na.rm = T)) %>%
  dplyr::select(Date,Means)

rm(SILOdata.QLD)

# MDB
rainMDB_t <- as.tibble(data.frame(Date =  seq.Date(as.Date("1979-01-01"),
                                   as.Date("2016-12-31"),by="month"),
                                  rain=t(SILOdata.MDB@data@values)))

# calculate mean by row
MDB_mean <- rainMDB_t %>%
  mutate(MDBrain = rowMeans(.[,-1], na.rm = T)) %>%
  dplyr::select(Date,MDBrain)

rm(SILOdata.MDB)
```

# Regression model

This regression model now includes the mean monthly MDB rainfall as an additional variable to take out the seasonal trend and the overall possible long term trends.

```{r fullRegression}
## predicting model: land cover is category variable
rain.trend <- function(x,y,var.data) {
  # add a very small number to take out 0 values
  var.data$rain <- as.vector(unlist(y[x]))[1:nrow(var.data)]+0.0001

      rain_fit <- gam(rain ~ MDB
                  + s(SOI, k=3, bs="cs")
                  + s(IOD, k=3, bs="cs")
                  #+ s(PDO, k=3, bs="cs")
                  + s(NINO3.4, k=3, bs="cs")
                  + s(NINO4, k=3, bs="cs")
                  + cover
                  + trd
                  , data=var.data,#family=Gamma(link=log),
                  na.action = na.omit)
    #
  return(rain_fit)

}

## generate variables for the model
trd <- seq(1:456)

LC <- c(rep("tree",288),rep("clear",168))
# different QLD scenarios
LC_qld1 <- c(rep("tree",294),rep("clear",162)) ## after-change: from Jun 2003
LC_qld2 <- c(rep("tree",300),rep("clear",156)) ## after- : from Jan 2004
LC_qld3 <- c(rep("tree",306),rep("clear",150)) ## after- : from Jun 2004
LC_qld4 <- c(rep("tree",312),rep("clear",144)) ## after- : from Jan 2005

# NSW
var.data <- data.frame(trd, MDB = MDB_mean$MDBrain, 
                       SOI = SOI_37Y$SOI,
                       IOD = IOD_37Y$IOD,
                       PDO = PDO_37Y$PDO,
                       NINO3.4 = SST34_37Y$SST34,
                       NINO4 = SST4_37Y$SST4,
                       cover = LC)

# QLD
# test different scenarios
#
var.qld2 <- data.frame(trd, MDB = MDB_mean$MDBrain, 
                       SOI = SOI_37Y$SOI,
                       IOD = IOD_37Y$IOD,
                       PDO = PDO_37Y$PDO,
                       NINO3.4 = SST34_37Y$SST34,
                       NINO4 = SST4_37Y$SST4,
                       cover = LC_qld2)

var.qld4 <- data.frame(trd, MDB = MDB_mean$MDBrain,
                       SOI = SOI_37Y$SOI,
                       IOD = IOD_37Y$IOD,
                       PDO = PDO_37Y$PDO,
                       NINO3.4 = SST34_37Y$SST34,
                       NINO4 = SST4_37Y$SST4,
                       cover = LC_qld4)

## run model for all periods
## QLD
coef.rain30.qld2 <- lapply(2:ncol(rainQld_t),
                    function(x) rain.trend(x,rainQld_t,var.qld2))
coef.rain30.qld4 <- lapply(2:ncol(rainQld_t),
                       function(x) rain.trend(x,rainQld_t,var.qld4))
# NSW/VIC
coef.rain30.nswvic <- lapply(2:ncol(rainNSW_t),
                   function(x) rain.trend(x,rainNSW_t,var.data))
```

Extract the sign and significance of the MDB variable. This is just to show that this variable is significant and has a consistent value across the area

```{r}
## extract sign of MDB, which is assumed to have a linear relationship with
## the rainfall in the area
# Qld
MDB_qld <- unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$p.coeff[2],NA)))
# NSW/VIC
MDB_nswvic <- unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$p.coeff[2],NA)))

par(mfrow=c(2,1))
hist(MDB_nswvic, xlab = "value of MDB rainfall estimate", main="NSW full data period")
hist(MDB_qld, xlab = "value of MDB rainfall estimate", main="Qld full data period")

```

The value of the linear effect of the MDB rainfall is all positive and ranges from 0.8 - 2. Interestingly, in both locations, the rainfall is generally above the average monthly MDB rainfall.

### Extract the sign of C, which is the tree cover effect.

```{r}
## extract sign of C (the tree cover effect)
C30_qld <- unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$p.coeff[3],NA)))
# NSW/VIC
C30_nswvic <- unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$p.coeff[3],NA)))
# NSW/VIC 7 year (includes 2009)
#  C30_nswvic7 <- unlist(lapply(1:length(coef.rain30.nswvic7),function(x) ifelse(summary(coef.rain30.nswvic7)[x,2]=="gam",summary(coef.rain30.nswvic7[[x]])$p.coeff[2],NA)))
# # NSW/VIC 8 year (includes 2010)
#  C30_nswvic8 <- unlist(lapply(1:length(coef.rain30.nswvic8),function(x) ifelse(summary(coef.rain30.nswvic8)[x,2]=="gam",summary(coef.rain30.nswvic8[[x]])$p.coeff[2],NA)))

par(mfrow=c(2,1))
hist(unlist(lapply(1:length(coef.rain30.nswvic),function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",summary(coef.rain30.nswvic[[x]])$p.coeff[3],NA))), xlab = "value of tree cover estimate", main="NSW full data period")
hist(unlist(lapply(1:length(coef.rain30.qld4),function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",summary(coef.rain30.qld4[[x]])$p.coeff[3],NA))), xlab = "value of tree cover estimate", main="Qld full data period")

```
Tree cover effect values are quite significant on a linear relationship, but they are not all significant (see below). For NSW/VIC they are all positive, but for QLD they are centered around 0. What does it mean when they are positive? It is categorical variable, which is ranked alphabetically, so "clear" = 1 and "tree" = 2. This means that a positive coefficient (on a linear scale) means there is more rainfall with tree cover and less when it is "clear".

```{r}

# p-value of the linear variables
pvalues_qld4 <- lapply(1:length(coef.rain30.qld4),
   function(x) summary(coef.rain30.qld4[[x]])$p.pv[2:4])

plinear_qld <- as.tibble(do.call(rbind,pvalues_qld4))
colnames(plinear_qld) = c("MDB", "LC", "trd")
plinear_qld$state <- "QLD"

# NSW all data
pvalues_nswvic <- lapply(1:length(coef.rain30.nswvic),
    function(x) summary(coef.rain30.nswvic[[x]])$p.pv[2:4])

plinear_nsw <- as.tibble(do.call(rbind,pvalues_nswvic))
colnames(plinear_nsw) = c("MDB", "LC", "trd")
plinear_nsw$state <- "NSW/VIC"

# plotting
p_plot_df <- rbind(plinear_qld,plinear_nsw)
p_plot_df %>%
  gather(key="Variable", value="p_value", MDB:trd) %>%
  ggplot(aes(p_value,fill=state)) + geom_histogram(alpha=0.5) +
  facet_wrap(~Variable) +
  scale_fill_manual(values = c("blue","red")) +
  theme(axis.title = element_text(size = rel(2)),
        axis.text = element_text(size=rel(1)),
        strip.text = element_text(face="bold", size= rel(2)),
        legend.text = element_text(face="bold", size= rel(1)),
        legend.title = element_text(face="bold", size= rel(2)))
# produce also as jpeg
jpeg("figures/pvalues_linear_fullmodel.jpg", height=960, width=960)
p_plot_df %>%
  gather(key="Variable", value="p_value", MDB:trd) %>%
  ggplot(aes(p_value,fill=state)) + geom_histogram(alpha=0.5) +
  facet_wrap(~Variable) +
  scale_fill_manual(values = c("blue","red")) +
  theme(axis.title = element_text(size = rel(2)),
        axis.text = element_text(size=rel(1)),
        strip.text = element_text(face="bold", size= rel(2)),
        legend.text = element_text(face="bold", size= rel(1)),
        legend.title = element_text(face="bold", size= rel(2)))
dev.off()

rm(p_plot_df)

## p value of seasons and climate indices
p_qld <- do.call(rbind,
                 lapply(1:length(coef.rain30.qld4), function(x) {
                   ifelse(summary(coef.rain30.qld4)[x,2]=="gam",
                      return(summary(coef.rain30.qld4[[x]])$s.pv),
                                       return(rep(NA,6)))}))

p_qld <- as.tibble(p_qld)
colnames(p_qld) = c("SOI", "IOD", "NINO3.4", "NINO4")
p_qld$state <- "QLD"

p_nsw <- do.call(rbind,
                 lapply(1:length(coef.rain30.nswvic), function(x) {
                ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",
                  return(summary(coef.rain30.nswvic[[x]])$s.pv),
                       return(rep(NA,6)))}))
p_nsw <- as.tibble(p_nsw)
colnames(p_nsw) = c("SOI", "IOD", "NINO3.4", "NINO4")
p_nsw$state <- "NSW/VIC"

p_plot_df <- rbind(p_qld,p_nsw)

p_plot_df %>%
  gather(key="Variable", value="p_value", SOI:NINO4) %>%
  ggplot(aes(p_value,fill=state)) + geom_histogram(alpha=0.5) +
  facet_wrap(~Variable) +
  scale_fill_manual(values = c("blue","red")) +
  theme(axis.title = element_text(size = rel(2)),
        axis.text = element_text(size=rel(1.5)),
        strip.text = element_text(face="bold", size= rel(2)),
        legend.text = element_text(face="bold", size= rel(1.5)),
        legend.title = element_text(face="bold", size= rel(2)))
# produce also as jpg
jpeg("figures/pvalues_fullmodel.jpg", height=960, width=960)
p_plot_df %>%
  gather(key="Variable", value="p_value", SOI:NINO4) %>%
  ggplot(aes(p_value,fill=state)) + geom_histogram(alpha=0.5) +
  facet_wrap(~Variable) +
  scale_fill_manual(values = c("blue","red")) +
  theme(axis.title = element_text(size = rel(2)),
        axis.text = element_text(size=rel(1.5)),
        strip.text = element_text(face="bold", size= rel(2)),
        legend.text = element_text(face="bold", size= rel(1.5)),
        legend.title = element_text(face="bold", size= rel(2)))
#dev.off()
rm(p_plot_df)

```

Clearly the MDB data is highly significant and takes out a lot of the variation in the model.This takes out quite a bit of variation out of the model for the climate indices, but at some locations these are still significant. It is also clear that the average MDB rainfall explains more of the variation in rainfall in QLD then it does in NSW/VIC. This is particularly clear in the significance of the other variables, which have much less significance in QLD.

```{r}
## extract r.sq of the model
rsq_qld <- unlist(lapply(1:length(coef.rain30.qld4),
          function(x) ifelse(summary(coef.rain30.qld4)[x,2]=="gam",
                    summary(coef.rain30.qld4[[x]])$r.sq,NA)))
rsq_nsw <- unlist(lapply(1:length(coef.rain30.nswvic),
    function(x) ifelse(summary(coef.rain30.nswvic)[x,2]=="gam",
                    summary(coef.rain30.nswvic[[x]])$r.sq,NA)))

# show histogram
plot_df <- data_frame(`r.squared` = c(rsq_qld,rsq_nsw),
                      state = c(rep("QLD", length(rsq_qld)),
                                rep("NSW/VIC", length(rsq_nsw))))

#jpeg("../../figures/rsq_fullmodel.jpg", height=480, width=480)
plot_df %>%
  ggplot(aes(`r.squared`*100, fill=state)) + geom_histogram() +
  scale_fill_manual(values = c("blue","red")) +
  xlab(expression(paste("% of variability explained (",r^2,")")))
#dev.off()

plot_df %>%
  group_by(state) %>%
  summarise(mean_rsq = mean(`r.squared`))

## frame of the two regions (100 years data)
qld_tif <- raster("Data\\BoM\\rainfall\\AWAP\\from1900_FVG\\r19000101_qld.tif")
nsw_tif <- raster("Data\\BoM\\rainfall\\AWAP\\from1900_FVG\\r19000101_nsw.tif")

  dist_Cp30_qld4 <- raster(qld_tif)
  dist_Cp30_qld4[] <- plinear_qld$LC

#  jpeg("../../figures/Cp_30yrs_qld.jpg", width=1440, height=720)
#  par(mfrow=c(1,2))
  par(mar=c(3,3,1,7),xpd=T)
  plot(e111,xlab="",ylab="", cex.lab = 2,
       cex.axis=2)
  image(dist_Cp30_qld4,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
   plot(aus_reg,border="grey",add=T)
  plot(mdb_reg,border="grey",lwd=3,add=T)
  plot(e111,add=T)
  image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
legend(148.7,-27.9,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=1)

  dist_Cp30_nsw <- raster(nsw_tif)
  dist_Cp30_nsw[] <- plinear_nsw$LC
#
#  jpeg("../../figures/Cp_30yrs_nswvic.jpg", width=1440, height=720)
#  par(mfrow=c(1,3))
  par(mar=c(3,3,1,7),xpd=T)
  plot(e222,xlab="",ylab="", cex.lab = 2,
       cex.axis=2) #,border="white")
  image(dist_Cp30_nsw,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
  plot(aus_reg,border="grey75",lwd=5,add=T)
  plot(mdb_reg,border="grey75",lwd=5,add=T)
  plot(e222,add=T)
  image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
  #dev.off()

# Now make the final plot for the paper with both sites all data
#  jpeg("../../figures/Cp_30yrs.jpg", width=1440, height=1440)
  par(mfrow=c(2,1))
  # NSW first
  par(mar=c(3,3,1,9),xpd=T)
  plot(e222,xlab="",ylab="", cex.lab = 2,
       cex.axis=2) #,border="white")
  image(dist_Cp30_nsw,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
  plot(aus_reg,border="grey75",lwd=5,add=T)
  plot(mdb_reg,border="grey75",lwd=5,add=T)
  plot(e222,add=T)
  image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
legend(149,-36.6,c("State","MDB"),col=c("grey","grey"),
       lwd=c(1,3),bty="n",title="Boundary",cex=1.8)
  #dev.off()

  # Then qld
  par(mar=c(3,3,1,9),xpd=T)
  plot(e111,xlab="",ylab="", cex.lab = 2,
       cex.axis=2)
  image(dist_Cp30_qld4,col=c("red","orange","yellow","light blue"),breaks=c(0,0.05,0.1,0.2,1),legend=F,add=T)
  plot(aus_reg,border="grey",add=T)
  plot(mdb_reg,border="grey",lwd=3,add=T)
  plot(e111,add=T)
  image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=0.7,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))

```
Overall, more of the variation in the rainfall is explained in QLD compared to NSW/VIC by the model. THe NSW/VIC model shows an almost bimodal distribution of the r-squared.

In terms of the spatial distribution of the significance of the tree cover (LC) effect, there is little evidence of a consistent change in the rainfall due to tree cover in QLD. However across the NSW/VIC area there is a significant area of change, and as indicated earlier, this change is all positive, suggesting more rainfall with tree cover.

### Rainfall model residuals

```{r}
#
resid_qld <- lapply(1:length(coef.rain30.qld4), function(x) if(summary(coef.rain30.qld4)[x,2]=="gam") {residuals(coef.rain30.qld4[[x]])} else{NA})

# ## calculate mean residual
  test <- do.call(cbind,resid_qld)
  qld.resid.mean <- apply(test,1, mean,na.rm=T)
  residuals <- data_frame(Dates = seq.Date(as.Date("1979-01-01"),
                             length = length(qld.resid.mean),
                                           by = "month"),
                          QLD = qld.resid.mean)
#
resid_nswvic <- lapply(1:length(coef.rain30.nswvic),
            function(x) if(summary(coef.rain30.nswvic)[x,2]=="gam")
              {residuals(coef.rain30.nswvic[[x]])} else{NA})

## calculate mean residual
test <- do.call(cbind,resid_nswvic)
nswvic.resid.mean <- apply(test,1, mean,na.rm=T)
residuals$NSWVIC <- nswvic.resid.mean

## plots
residuals %>%
  gather(key="State", value="residual", QLD, NSWVIC) %>%
  ggplot(aes(Dates,residual)) + geom_line(colour="blue") +
  geom_vline(xintercept=as.numeric(as.Date("2003-01-01")), colour="red",
             size=1.5) +
  facet_wrap(~State)  +
  theme(axis.text = element_text(size=rel(1.5)),
        axis.title = element_text(size=rel(2)),
        strip.text = element_text(size=rel(2), face = "bold"))

jpeg("figures/GAMresidualTime.jpeg", height=960, width=960)
residuals %>%
  gather(key="State", value="residual", QLD, NSWVIC) %>%
  ggplot(aes(Dates,residual)) + geom_line(colour="blue") +
  geom_vline(xintercept=as.numeric(as.Date("2003-01-01")), colour="red",
             size=1.5) +
  facet_wrap(~State)  +
  theme(axis.text = element_text(size=rel(1.5)),
        axis.title = element_text(size=rel(2)),
        strip.text = element_text(size=rel(2), face = "bold"))
dev.off()

#
#
## plot gam.check
jpeg("figures/gam_check.jpg", width=720, height=960)
  par(mfrow=c(4,2))
  gam.check(coef.rain30.qld4[[1]])
  gam.check(coef.rain30.nswvic[[1]])
dev.off()

```

# Step 3 Residual trend analysis for model without tree cover

This is the same model as above, but we have removed the tree cover variable.

```{r model_noLC}
## predicting model: land cover is category variable
rain.trend3 <- function(x,y,var.data) {
  # add a very small number to take out 0 values
  var.data$rain <- as.vector(unlist(y[x]))[1:nrow(var.data)]+0.0001

      rain_fit <- gam(rain ~ MDB
                  + s(SOI, k=3, bs="cs")
                  + s(IOD, k=3, bs="cs")
                  #+ s(PDO, k=3, bs="cs")
                  + s(NINO3.4, k=3, bs="cs")
                  + s(NINO4, k=3, bs="cs")
                  + trd
                  , data=var.data,#family=Gamma(link=log),
                  na.action = na.omit)
    #
  return(rain_fit)

}

## run model for 37 years
mod_rain_qld4 <- lapply(2:ncol(rainQld_t),
                       function(x) rain.trend3(x,rainQld_t,var.qld4))
mod_rain_qld2 <- lapply(2:ncol(rainQld_t),
                  function(x) rain.trend3(x,rainQld_t,var.qld2))
## QLD
rsq_qld <- unlist(lapply(1:length(mod_rain_qld4), 
                    function(x) summary(mod_rain_qld4[[x]])$r.sq))
#
#
## extract residual
residual_qld <- lapply(1:length(mod_rain_qld4), 
      function(x) {mod_rain_qld4[[x]]$y-fitted(mod_rain_qld4[[x]])})

residual_qld2 <- lapply(1:length(mod_rain_qld2), 
    function(x) {mod_rain_qld2[[x]]$y-fitted(mod_rain_qld2[[x]])})

#
# ## NSW/VIC
#
#NSW/VIC
mod_rain_nswvic <- lapply(2:ncol(rainNSW_t), 
                function(x) rain.trend3(x,rainNSW_t,var.data))

## explaining power of the partial model
rsq_nswvic <- unlist(lapply(1:length(mod_rain_nswvic), 
              function(x) summary(mod_rain_nswvic[[x]])$r.sq))
#

# extract residual
residual_nswvic <- lapply(1:length(mod_rain_nswvic),
  function(x) {mod_rain_nswvic[[x]]$y-fitted(mod_rain_nswvic[[x]])})
```

### partial r-sq

We can now analyse the residuals and the models. First we look at the partial explaining power of the climate indices in the model

```{r plotResiduals}
# histogram of partial model rsq
qld_df <-  data.frame(rsq = rsq_qld*100,
                      state = "Qld")
nsw_df <-  data.frame(rsq = rsq_nswvic*100,
                      state = "NSW/VIC")

plot_df <- rbind(qld_df, nsw_df)
#jpeg("../../figures/pmodel_explHistogram.jpg")
plot_df %>%
  ggplot(aes(rsq)) + geom_histogram(fill="blue") +
  facet_wrap(~state) +
  xlab(expression(paste("% of variability explained (",r^2,")"))) +
  theme(axis.text = element_text(size=rel(1.5)),
        strip.text = element_text(size=rel(2)),
        axis.title = element_text(size=rel(2)))
# dev.off()
```

Similar to earlier graph, shows that more of the variation is explained in Qld compared to NSW/VIC.

### Residual distributions and t-test

We can now analyse the residuals in more detail and look at t-tests

```{r residualModelAnalysis}
# stack the residuals

# create a boxplot with the annual residuals from 1998
resplot_df <- data_frame(Date = rep(ymd(rainNSW_t[[1]]),
          length(residual_nswvic)+length(residual_qld)),
          residuals = c(unlist(residual_nswvic),                                               unlist(residual_qld)),
          State = c(rep("NSW/VIC",
                        length(unlist(residual_nswvic))),
                        rep("QLD", length(unlist(residual_qld)))))
resplot_df %>%
  filter(Date >= ymd("1998-01-01")) %>%
  ggplot(aes(Date,residuals, fill=State)) +
  geom_boxplot(aes(group=year(Date))) +
  facet_wrap(~State, ncol=1)

jpeg("figures/ResidualBoxplot.jpg")
resplot_df %>%
  filter(Date >= ymd("1998-01-01")) %>%
  ggplot(aes(Date,residuals, fill=State)) +
  geom_boxplot(aes(group=year(Date))) +
  facet_wrap(~State, ncol=1)
dev.off()

# residuals before and after change
#par(mfrow=c(1,2))
jpeg("figures/ResidualBoxplotchange.jpg")
p1 <- resplot_df %>%
  filter(Date >= ymd("1998-01-01")) %>%
  mutate(Change = ifelse(year(Date) > 2003, "After", "Before")) %>%
  ggplot(aes(Change,residuals, fill= State)) +
  geom_boxplot(aes(group=Change)) + facet_wrap(~State, ncol=1) +
  guides(fill=FALSE) + ggtitle("whole period")
#dev.off()

#jpeg("../../figures/ResidualBoxplotchange2.jpg")
p2 <- resplot_df %>%
  filter(Date >= ymd("1998-01-01")) %>%
  mutate(Change = ifelse(year(Date) > 2003, "After", "Before")) %>%
  filter(year(Date) < 2010) %>%
  ggplot(aes(Change,residuals, fill = State)) +
  geom_boxplot(aes(group=Change)) + facet_wrap(~State, ncol=1) +
  ylab(label="") + ggtitle("1998 - 2009")
lay <- rbind(c(1,1,2,2,2))
grid.arrange(p1,p2, layout_matrix = lay)
dev.off()
p1 <- resplot_df %>%
  filter(Date >= ymd("1998-01-01")) %>%
  mutate(Change = ifelse(year(Date) > 2003, "After", "Before")) %>%
  ggplot(aes(Change,residuals, fill= State)) +
  geom_boxplot(aes(group=Change)) + facet_wrap(~State, ncol=1) +
  guides(fill=FALSE) + ggtitle("whole period")
#dev.off()

#jpeg("../../figures/ResidualBoxplotchange2.jpg")
p2 <- resplot_df %>%
  filter(Date >= ymd("1998-01-01")) %>%
  mutate(Change = ifelse(year(Date) > 2003, "After", "Before")) %>%
  filter(year(Date) < 2010) %>%
  ggplot(aes(Change,residuals, fill = State)) +
  geom_boxplot(aes(group=Change)) + facet_wrap(~State, ncol=1) +
  ylab(label="") + ggtitle("1998 - 2009")
lay <- rbind(c(1,1,2,2,2))
grid.arrange(p1,p2, layout_matrix = lay)

res2 <- resplot_df %>%
  filter(Date >= ymd("1998-01-01")) %>%
  mutate(Change = ifelse(year(Date) > 2003, "After", "Before")) %>%
  #filter(year(Date) < 2010) %>%
  filter(State == "QLD")

  summary(lm(residuals~Change-1,data=res2))
  t.test(formula=residuals~Change,data=res2)
# No significant difference if measured to 2010 and slightly less rain (p < 0.05) before if measured to 2015 (~ 2 mm/month)

res2 <- resplot_df %>%
  filter(Date >= ymd("1998-01-01")) %>%
  mutate(Change = ifelse(year(Date) > 2003, "After", "Before")) %>%
  #filter(year(Date) < 2010) %>%
  filter(State != "NSW/VIC")

  summary(lm(residuals~Change-1,data=res2))
  t.test(formula=residuals~Change,data=res2)

```

This does suggest that the residuals are significantly different before and after the change in landcover

# step 4 Analysis of the residuals Mann-Whitney rank-sum statistic

This mainly follows the original paper, except we will go straight for the averaging by 4 cells. The bootstrap analysis is also just run on the HPC, and we only pull in the results here

```{r steptrend_ns4}
## perform step test on the data using ns = 4
## function to calculate the step trend statistic (Hirsch, 1985)
stepTrend2 <- function(x,y,i, endyear=2015){
#browser()
    n1 <- 2003-1997

    if (i==1){

      z <- 56  # no. of column, related to the dim of the region
      # Qld 57 by 51
      # NSW 38 by 33

    } else {

    z <- 37 }

if (x%%z!=0){ #& x%%z!=(z-1)) {
    y.1 <- y[[x]]
    y.2 <- y[[x+1]]
    y.3 <- y[[x+z]]
    y.4 <- y[[x+z+1]]

    if(!is.na(y.1[1])&!is.na(y.2[1])&!is.na(y.3[1])&!is.na(y.4[1])) { 
    if (i == 1) {
      n2 <- endyear-2004
      v.1 <- ts(y.1[c(229:288,313:length(y.1))],
                start=c(1997,1),frequency=12)
      v.2 <- ts(y.2[c(229:288,313:length(y.1))],
                start=c(1997,1),frequency=12)
      v.3 <- ts(y.3[c(229:288,313:length(y.1))],
                start=c(1997,1),frequency=12)
      v.4 <- ts(y.4[c(229:288,313:length(y.1))],
                start=c(1997,1),frequency=12)

    } else {

      n2 <- endyear-2003

      v.1 <- ts(y.1[c(229:length(y.1))],
                start=c(1997,1),frequency=12)
      v.2 <- ts(y.2[c(229:length(y.2))],
                start=c(1997,1),frequency=12)
      v.3 <- ts(y.3[c(229:length(y.3))],
                start=c(1997,1),frequency=12)
      v.4 <- ts(y.4[c(229:length(y.4))],
                start=c(1997,1),frequency=12)
}

## rank by month
   for (j in 1:12) {
    assign(paste("r1",j,sep=""),rank(subset(v.1,cycle(v.1)==j)))
    assign(paste("r2",j,sep=""),rank(subset(v.2,cycle(v.2)==j)))
    assign(paste("r3",j,sep=""),rank(subset(v.3,cycle(v.3)==j)))
    assign(paste("r4",j,sep=""),rank(subset(v.4,cycle(v.4)==j)))

    }

## mann-whitney rank sum statistic
    rlist.1 <- list(r11,r12,r13,r14,r15,r16,r17,r18,r19,r110,r111,r112) ## full year
    rlist.2 <- list(r21,r22,r23,r24,r25,r26,r27,r28,r29,r210,r211,r212)
    rlist.3 <- list(r31,r32,r33,r34,r35,r36,r37,r38,r39,r310,r311,r312)
    rlist.4 <- list(r41,r42,r43,r44,r45,r46,r47,r48,r49,r410,r411,r412)
    rlist <- list(unlist(rlist.1),unlist(rlist.2),unlist(rlist.3),unlist(rlist.4))

    w.s1 <- 0
    w.s2 <- 0
    w.s3 <- 0
    w.s4 <- 0

for (k in 1:length(rlist.1)) {
    w.s1 <- sum(rlist.1[[k]][1:n1])+w.s1
    w.s2 <- sum(rlist.2[[k]][1:n1])+w.s2
    w.s3 <- sum(rlist.3[[k]][1:n1])+w.s3
    w.s4 <- sum(rlist.4[[k]][1:n1])+w.s4
    }
    w.s <- sum(w.s1,w.s2,w.s3,w.s4) #,w.s5,w.s6,w.s7,w.s8,w.s9)

## other statistic
# mean for full year
    mu <- n1*(n1+n2+1)/2
    mu.s <- 12*4*mu

## variance (standard variation). The autocorrelation of the residuals can be assumed to be null (checked, cor.test())
# assume only cross correlation
    ss <- n1*n2*(n1+n2+1)/12
    ss.s <- 0
    for (kk in 1:4) {
      for (kkk in 1:4) {
#         for (kkkk in 1:12) {
          ss.s <- 12*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
#         }
      }
    }

## step trend statistic
    z <- (w.s - mu.s)/sqrt(ss.s)

    return(z) }

     else {return(NA)} }

     else {return(NA)}

  }

## 30 years
## full year,
## central QLD
  qld.z.new <- unlist(lapply(1:(length(residual_qld)-57),function(x) stepTrend2(x,residual_qld,1)))
  qld.z.new <- c(qld.z.new,rep(NA,57))
  tr_qld.step.new <- raster(qld_tif)
  tr_qld.step.new[] <- qld.z.new

## NSW
  nswvic.z.new <- unlist(lapply(1:(length(residual_nswvic)-38),function(x) stepTrend2(x,residual_nswvic,2)))
  nswvic.z.new <- c(nswvic.z.new,rep(NA,38))

  tr_nswvic.step.new <- raster(nsw_tif)
  tr_nswvic.step.new[] <- nswvic.z.new
plotclt <- rev(brewer.pal(8,"RdYlBu"))

jpeg("figures/step_new.jpg", width=1240, height=960, quality=100)
  par(mar=c(3,3,1,9), mfrow=c(2,2), xpd=T)
  plot(e222,xlab="",ylab="", cex.axis=2)#,border="white")
  image(tr_nswvic.step.new,col=plotclt,breaks=c(-2.3,-1.645,-1.282,-0.842,0,0.842,1.282,1.645,3),legend=F,add=T)
  plot(mdb_reg2,border="grey",add=T,lwd=3)
  plot(aus_reg2,border="grey",add=T)
  plot(e222,add=T)
image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,0,1,2,3,4),lab.breaks=c(0.00,0.05,0.1,0.2,1,0.2,0.1,0.05,0.00),legend.only=T,horizontal=F,col=plotclt,legend.shrink=0.7,legend.width=0.9,legend.args=list(text=c(expression(paste("level of significance (",alpha,")")),"positive Z'","negative Z'"),side=c(2,3,1),line=0.7,cex=1.5,font=2),axis.args=list(cex.axis=1.5))
legend(149,-36.6,c("State","MDB"),col=c("grey","grey"),
       lwd=c(1,3),bty="n",title="Boundary",cex=1.8)

  # Qld
  par(mar=c(3,3,1,9),xpd=T)
  plot(e111,xlab="",ylab="", cex.axis=2)#,border="white")
  image(tr_qld.step.new,col=plotclt,breaks=c(-3,-1.645,-1.282,-0.842,0,0.842,1.282,1.645,3),legend=F,add=T)
  plot(mdb_reg2,border="grey",add=T,lwd=3)
  plot(e111,add=T)
image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,0,1,2,3,4),lab.breaks=c(0.00,0.05,0.1,0.2,1,0.2,0.1,0.05,0.00),legend.only=T,horizontal=F,col=plotclt,legend.shrink=0.7,legend.width=0.9,legend.args=list(text=c(expression(paste("level of significance (",alpha,")")),"positive Z'","negative Z'"),side=c(2,3,1),line=0.9,cex=1.5,font=2),axis.args=list(cex.axis=1.5))

dev.off()

#ns = 4 results
# positive
qld_origp.new <- length(subset(qld.z.new,qld.z.new>=1.645))/
  length(qld.z.new)
# 0.00034
# negative
qld_orign.new <- length(subset(qld.z.new,qld.z.new<=-1.645))/
  length(qld.z.new)
# 0

nsw_origp.z.new <- length(subset(nswvic.z.new,nswvic.z.new>=1.645))/
  length(nswvic.z.new)
# 0
nsw_orign.z.new <- length(subset(nswvic.z.new,nswvic.z.new<=-1.645))/
  length(nswvic.z.new)
# 0

```

Bootstrap analysis was moved the HPC in January 2019.

```{r bootstrap_ns4, eval=F, echo=F}

# moved to HPC 08/01/2019
# # bootstrap ns = 4
# ## multiple cells
# # Set up for the shorter period
# NstepTrend <- function(x,y,m,i){
# 
#     n1 <- 2003-1997
# 
#     if (i==1){
# 
#       z <- 56  # no. of column, related to the dim of the region
#       # Qld 57 by 51
#       # NSW 38 by 33
# 
#     } else {
# 
#     z <- 37 }
# 
# if (x%%z!=0){ #& x%%z!=(z-1)) {
#     y.1 <- m[[x]]
#     y.2 <- m[[x+1]]
#     y.3 <- m[[x+z]]
#     y.4 <- m[[x+z+1]]
# 
#     if(!is.na(y.1[1])&!is.na(y.2[1])&!is.na(y.3[1])&!is.na(y.4[1])) {
#     if (i == 1) {
#       n2 <- 2016-2004
#       v.1 <- ts(y.1[c(229:288,313:length(y.1))],
#                 start=c(1997,1),frequency=12)
#       v.2 <- ts(y.2[c(229:288,313:length(y.2))],
#                 start=c(1997,1),frequency=12)
#       v.3 <- ts(y.3[c(229:288,313:length(y.3))],
#                 start=c(1997,1),frequency=12)
#       v.4 <- ts(y.4[c(229:288,313:length(y.4))],
#                 start=c(1997,1),frequency=12)
# 
#     } else {
# 
#       n2 <- 2016-2002
# 
#       v.1 <- ts(y.1[c(229:length(y.1))],start=c(1997,1),frequency=12)
#       v.2 <- ts(y.2[c(229:length(y.2))],start=c(1997,1),frequency=12)
#       v.3 <- ts(y.3[c(229:length(y.3))],start=c(1997,1),frequency=12)
#       v.4 <- ts(y.4[c(229:length(y.4))],start=c(1997,1),frequency=12)
# }
# 
# ## rank by month
# #browser()
#    for (j in 1:12) {
#     y1.mth <- data.frame(mth=as.vector(window(v.1,
#                                               start=c(1997,i),
#                                               frequency=1)),
#                          o=y[[j]])[order(y[[j]]),]
#     y2.mth <- data.frame(mth=as.vector(window(v.2,
#                                               start=c(1997,i),
#                                               frequency=1)),
#                          o=y[[j]])[order(y[[j]]),]
#     y3.mth <- data.frame(mth=as.vector(window(v.3,
#                                               start=c(1997,i),
#                                               frequency=1)),
#                          o=y[[j]])[order(y[[j]]),]
#     y4.mth <- data.frame(mth=as.vector(window(v.4,
#                                               start=c(1997,i),
#                                               frequency=1)),
#                          o=y[[j]])[order(y[[j]]),]
# 
#     assign(paste("r1",j,sep=""),rank(y1.mth$mth))
#     assign(paste("r2",j,sep=""),rank(y2.mth$mth))
#     assign(paste("r3",j,sep=""),rank(y3.mth$mth))
#     assign(paste("r4",j,sep=""),rank(y4.mth$mth))
# 
#     }
# 
# ## mann-whitney rank sum statistic
#     rlist.1 <- list(r11,r12,r13,r14,r15,r16,r17,r18,r19,r110,r111,r112) ## full year
#     rlist.2 <- list(r21,r22,r23,r24,r25,r26,r27,r28,r29,r210,r211,r212)
#     rlist.3 <- list(r31,r32,r33,r34,r35,r36,r37,r38,r39,r310,r311,r312)
#     rlist.4 <- list(r41,r42,r43,r44,r45,r46,r47,r48,r49,r410,r411,r412)
#     rlist <- list(unlist(rlist.1),unlist(rlist.2),
#                   unlist(rlist.3),unlist(rlist.4))
# ## seasonal
#     w.s1 <- 0
#     w.s2 <- 0
#     w.s3 <- 0
#     w.s4 <- 0
#   for (k in 1:length(rlist.1)) {
# #   for (k in c(9:11)) {
#     w.s1 <- sum(rlist.1[[k]][1:n1])+w.s1
#     w.s2 <- sum(rlist.2[[k]][1:n1])+w.s2
#     w.s3 <- sum(rlist.3[[k]][1:n1])+w.s3
#     w.s4 <- sum(rlist.4[[k]][1:n1])+w.s4
#     }
#     w.s <- sum(w.s1,w.s2,w.s3,w.s4) #,w.s5,w.s6,w.s7,w.s8,w.s9)
# 
# ## other statistic
# # mean for full year
#     mu <- n1*(n1+n2+1)/2
#     mu.s <- 12*4*mu
# # ## mean for one season
# #     mu <- n1*(n1+n2+1)/2
# #     mu.s <- 3*4*mu
# 
# ## variance (standard variation). 
# # assume only cross correlation
#     ss <- n1*n2*(n1+n2+1)/12
#     ss.s <- 0
#     for (kk in 1:4) {
#       for (kkk in 1:4) {
# #         for (kkkk in 1:12) {
#           ss.s <- 12*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
# #         }
#       }
#     }
# ## variance for one season
# #     ss <- n1*n2*(n1+n2+1)/3
# #     ss.s <- 0
# #     for (kk in 1:4) {
# #       for (kkk in 1:4) {
# #           ss.s <- 3*ss*cor(rlist[[kk]],rlist[[kkk]]) + ss.s
# #       }
# #     }
# 
# ## step trend statistic
#     z <- (w.s - mu.s)/sqrt(ss.s)
# 
#     return(z) }
# 
#      else {return(NA)} }
# 
#      else {return(NA)}
# 
#   }
# 
# 
# R <- 1000
# new.order <- list()
# 
# zpq.values = numeric(R)
# znq.values = numeric(R)
# period1 <- 2002 - 1997
# period2 <- 2016 - 2004
# mth.l <- rep((period1 + period2),12)
# 
# #ptm <- proc.time()
# require(doParallel)
# cl <- makeCluster(4)
# registerDoParallel(cl)
# 
# QLD_zvalues <- foreach (i=1:R, .combine = rbind) %do% {
# 
#   for (j in 1:12) {
#     new.order[[j]]<-sample(1:mth.l[j])
#   }
# 
#   temp <- unlist(lapply(1:(length(residual_qld)-57),
#           function(x,y) NstepTrend(x,y,residual_qld,1),new.order))
#   # if(i==1) {write.table(temp,file="bt_zvalues_qld_7y.txt",row.names=F)
#   #           } else
#   #             {write.table(temp,file="bt_zvalues_qld_7y.txt",append=T,row.names=F,col.names=F)}
#   data.frame(zpq.values =
#                length(subset(temp,temp>=1.645))/length(temp),
#               znq.values =
#                length(subset(temp,temp<=-1.645))/length(temp))
# }
# stopCluster(cl)
# 
# registerDoParallel(cl)
# # NSW VIC
# NSW_zvalues <- foreach (i=1:R, .combine = rbind) %do% {
# 
#   for (j in 1:12) {
#     new.order[[j]]<-sample(1:mth.l[j])
#   }
# 
#   temp <- unlist(lapply(1:(length(residual_nswvic)-38),
#           function(x,y) NstepTrend(x,y,residual_nswvic,1),new.order))
#   # if(i==1) {write.table(temp,file="bt_zvalues_qld_7y.txt",row.names=F)
#   #           } else
#   #             {write.table(temp,file="bt_zvalues_qld_7y.txt",append=T,row.names=F,col.names=F)}
#   data.frame(zpq.values =
#                length(subset(temp,temp>=1.645))/length(temp),
#               znq.values =
#                length(subset(temp,temp<=-1.645))/length(temp))
# }
# stopCluster(cl)

```

Read in results from the HPC bootstrap analysis of the z values.

```{r}
NSW_zvalues <- readRDS("processed data/NSW_zvalues.RDS")
QLD_zvalues <- readRDS("processed data/QLD_zvalues.RDS")
```

Create bootstrap histograms of the results

```{r}

bt4_nswp.values <- NSW_zvalues$zpq.values*100
bt4_nswn.values <- NSW_zvalues$znq.values*100 
bt4_qldp.values <- QLD_zvalues$zpq.values*100
bt4_qldn.values <- QLD_zvalues$znq.values*100

#jpeg("figures/bt4_hist.jpg")
par(mfrow=c(2,2), mar = c(4,3,2,1)+0.1)
# nsw pos
hist(bt4_nswp.values,nclass=100,
     xlab="% positive Z'",main="NSW/VIC",cex.lab=1.4,cex.main=1.3) #xlim=c(0,20))
points(nsw_origp.z.new*100,0,pch=16,col="red",cex=1.8)
# nsw neg
hist(bt4_nswn.values,nclass=100,
     xlab="% negative Z'",main="",cex.lab=1.4,
     cex.main=1.3)#, xlim=c(0,20))
points(nsw_orign.z.new*100,0,pch=16,col="red",cex=1.8)
#qld pos
hist(bt4_qldp.values,nclass=100,
     xlab="% positive Z'",main="Qld",cex.lab=1.4,cex.main=1.3) #xlim=c(0,20))
points(qld_origp.new*100,0,pch=16,col="red",cex=1.8)
#qld neg
hist(bt4_qldn.values,nclass=100,
     xlab="% negative Z'",main="",cex.lab=1.4,
     cex.main=1.3)#, xlim=c(0,20))
points(qld_orign.new*100,0,pch=16,col="red",cex=1.8)
#dev.off()

```

### 2019-02-28 DATA/other/bushfires/read_code.r

@@ -1,3 +1,86 @@
-version https://git-lfs.github.com/spec/v1
-oid sha256:dd27652b8cc39e546805696a24053ed91a00527e0c6df292f4908f61da0ef33e
-size 3783
+library(RCurl)
+library(rgdal)
+library(gdata)
+library(raster)
+library(maptools)
+library(ncdf)
+library(fields)
+library(nnet)
+library(RNetCDF)
+
+#######################################################################
+#                 read data
+#######################################################################
+
+MRT <- "C:\\MRT\\MRT\\bin\\"
+
+### based on R:\\Data\\Satellite\\MOD44\\B, pixel size 250m
+setwd("R:\\Data\\Other\\bushfires\\Burnt_Area")
+
+mod <- dir("R:\\Data\\Other\\bushfires\\Burnt_Area") ## the last two are not .hdf file
+datad <- "R:\\Data\\Other\\bushfires\\Burnt_Area\\"
+
+mydirname <- unique(unlist(lapply(mod[grep(mod,pattern="MCD45A1.*.*.*.*.hdf")], function(x) unlist(strsplit(x, "\\."))[2])))
+
+for (i in 1:length(mydirname))  {
+  
+    dirname_y <- mod[grep(mod,pattern=paste("MCD45A1.",mydirname[i],".*.*.*.hdf",sep=""))]
+  
+    # mosaic the blocks:
+    mosaicname = file(paste(MRT, "TmpMosaic.prm", sep=""), open="wt")
+
+    for (j in 1:length(dirname_y)) {
+      if (j==1) write(paste(datad, dirname_y[1], sep=""), mosaicname) else
+      { write(paste(datad, dirname_y[j], sep=""), mosaicname, append=T)}
+    }
+    
+    close(mosaicname)
+
+
+shell(cmd=paste(MRT, 'mrtmosaic -i ', MRT, 'TmpMosaic.prm -s "1 0 0 0 0 0 0 " -o ', datad, 'TmpMosaic.hdf', sep=""))
+
+filename = file(paste(MRT, "mrt", mydirname[i], ".prm", sep=""), open="wt")
+    write(paste('INPUT_FILENAME = ', datad, 'TmpMosaic.hdf', sep=""), filename) 
[…]

### 2019-02-28 Scripts/ChunHPC_PBS.sh

@@ -1,0 +1,14 @@
+#!/bin/bash
+
+#PBS -l ncpus=12
+#PBS -l walltime=45:00:00
+#PBS -l mem=32GB
+#PBS -N RainChange
+#PBS -P RDS-FAE-HPWC-RW
+#PBS -m ae
+#PBS -M willem.vervoort@sydney.edu.au
+# #PBS -q defaultQ
+
+module load R/3.4.0
+
+R --vanilla </project/RDS-FAE-HPWC-RW/Script/HPC_gridtrendmodel.R> Shell_output.out

### 2019-02-28 Scripts/HPC_gridtrendmodel.R

@@ -1,0 +1,347 @@
+# script to run Chun's grid analysis on HPC
+require(tidyverse)
+require(lubridate)
+require(raster)
+require(lme4)
+## ----rcurl----------------------------------------------------------
+require(maptools)
+require(mgcv)
+
+setwd("/project/RDS-FAE-HPWC-RW")
+
+load("Data/Chundata.Rdata")
+
+## perform step test on the data using ns = 4
+## function to calculate the step trend statistic (Hirsch, 1985)
+stepTrend2 <- function(x,y,i, endyear=2015){
+  #browser()
+  n1 <- 2003-1997
+  
+  if (i==1){
+    
+    z <- 56  # no. of column, related to the dim of the region
+    # Qld 57 by 51
+    # NSW 38 by 33
+    
+  } else {
+    
+    z <- 37 }
+  
+  if (x%%z!=0){ #& x%%z!=(z-1)) {
+    y.1 <- y[[x]]
+    y.2 <- y[[x+1]]
+    y.3 <- y[[x+z]]
+    y.4 <- y[[x+z+1]]
+    
+    if(!is.na(y.1[1])&!is.na(y.2[1])&!is.na(y.3[1])&!is.na(y.4[1])) { 
+      if (i == 1) {
+        n2 <- endyear-2004
+        v.1 <- ts(y.1[c(229:288,313:length(y.1))],
+                  start=c(1997,1),frequency=12)
+        v.2 <- ts(y.2[c(229:288,313:length(y.1))],
+                  start=c(1997,1),frequency=12)
+        v.3 <- ts(y.3[c(229:288,313:length(y.1))],
+                  start=c(1997,1),frequency=12)
+        v.4 <- ts(y.4[c(229:288,313:length(y.1))],
+                  start=c(1997,1),frequency=12)
+        
+      } else {
+        
+        n2 <- endyear-2003
+        
+        v.1 <- ts(y.1[c(229:length(y.1))],
+                  start=c(1997,1),frequency=12)
+        v.2 <- ts(y.2[c(229:length(y.2))],
+                  start=c(1997,1),frequency=12)
+        v.3 <- ts(y.3[c(229:length(y.3))],
+                  start=c(1997,1),frequency=12)
+        v.4 <- ts(y.4[c(229:length(y.4))],
+                  start=c(1997,1),frequency=12)
+      }
+      
+      ## rank by month
[…]

### 2019-03-14 Manuscripts/PeerJRainfall_Landuse/Rainfall_Landuse.Rmd
paper for submission

@@
 abstract: >
- Analysis of observational data to identify relationships between rainfall and land cover change are difficult due to multiple environmental factors that cannot be strictly controlled.
-In this study we present a methodology to investigate the relationship using statistical methods on data from best available sources at two sites in Australia.
-Gridded data of rainfall and tree cover were used as spatially corresponding local conditions.
-Large scale effects were represented by climatic indicators, such as SOI and IOD.
-Regression analysis and step trend tests were used to assess the effect of abrupt land surface intervention.
-  At a Queensland site, significant tree cover change between 2002 - 2005 did not result in strong statistically significant precipitation changes.
-On the other hand, results from a bushfire affected NSW/VIC region suggests significant changes in the rainfall.
+ Analysis of observational data to pinpoint impact of land cover change on local rainfall are difficult due to multiple environmental factors that cannot be strictly controlled.
+In this study we use a statistical approach to identify the relationship between removal of tree cover and rainfall with data from best available sources for two large areas in Australia.
+Gridded rainfall data between 1979 and 2015 was used for the areas, while large scale (exogenous) effects were represented by mean rainfall across a much larger area and climatic indicators, such as SOI and IOD.
+Both general additive modelling and step trend tests were used for the analysis.
+  For a region in south central Queensland, the reported change in tree clearing between 2002 - 2005 did not result in strong statistically significant precipitation changes.
+On the other hand, results from a bushfire affected region on the border of New SOuth Wales and Victoria suggests significant changes in the rainfall due to changes in tree cover.
 This indicates the method works better when a abrupt change in the data can be clearly identified.
-The results from the step trend test implied a positive relationship between the tree cover and the rainfall at 0.1 significance level in both locations in data up to 2009.
-However, high rainfall variability and possible regrowth meant that no signifcant changes were observed in alonger time sereies to 2015.
+The results from the step trend test also mainly identified a positive relationship between the tree cover and the rainfall at p < 0.1 at the NSW/Victoria region.
+High rainfall variability and possible regrowth could have impacted the results in the Queensland region.
 header-includes: >
   \usepackage{lipsum}
-  \usepackage{textcomp}
+  \usepackage{textcomp, rotating}
 bibliography: reflist_chap3.bib
@@
 ```{r packages, echo=F, warnings=F, message=F, eval=T}
-require(knitr)
-require(bookdown)
-require(pander)
-library(fields)
-library(graphics)
-library(maptools)
-library(plotKML)
-library(raster)
-library(reshape)
-library(rgeos)
-library(rgdal)
-library(RColorBrewer)
-library(tidyverse)
-require(lubridate)
-library(deseasonalize)
-library(mgcv)
-library(ncdf4)
-require(gridExtra)
+suppressWarnings(require(knitr))
+suppressWarnings(require(bookdown))
+suppressWarnings(require(pander))
+suppressWarnings(library(fields))
+suppressWarnings(library(graphics))
+suppressWarnings(library(maptools))
+suppressWarnings(library(plotKML))
+suppressWarnings(library(raster))
+suppressWarnings(library(reshape))
+suppressWarnings(library(rgeos))
+suppressWarnings(library(rgdal))
+suppressWarnings(library(RColorBrewer))
+suppressWarnings(library(tidyverse))
+suppressWarnings(require(lubridate))
+suppressWarnings(library(deseasonalize))
+suppressWarnings(library(mgcv))
+suppressWarnings(library(ncdf4))
+suppressWarnings(require(gridExtra))
+suppressWarnings(require(kableExtra))
 #library(nnet)
@@
 Empirical and modelling studies have found that cloud types and rainfall are correlated to large scale vegetation cover changes, such as deforestation in the Amazon and in the Sahel [@Chagnon2005;@Pinto2009; @Wang2009; @Mei2010; @kucharski_further_2013; @pitman_scale_2016] and afforestation in south Israel [@Otterman1990; @Ben-Gai1998].
-Using airborne measurement in Western Australia, @Junkermann2009 showed a significantly higher level of aerosols over an agricultural area compared to an adjacent natural vegetation.
+Using airborne measurements in Western Australia, @Junkermann2009 showed a significantly higher level of aerosols over an agricultural area compared to an adjacent area with natural vegetation.
 They suggested that a modification of aerosol concentrations due to deforestation could have contributed to a reduction of local rainfall, as more, but smaller rain droplets were observed.
@@
 Therefore, the aim of this study is to use a statistical approach on rainfall data at regional scales to investigate the cause and effect relationship between land cover change and local rainfall, which is demonstrated in many modelling studies.
-More specifically, we hypothesize that a step change on the land cover on the surface will cause a step change in the rainfall.
+More specifically, we hypothesize that a step change in land cover on the surface will cause a step change in the rainfall.
 To demonstrate this we study changes in observed rainfall over time at a Queensland and NSW/Victoria location where there are possible step changes in land cover change due to land clearing and bush fires.
@@
 In Australia, significant tree cover change has mainly occurred in the north east  and south east of the continent,  as well as in the southwest of Western Australia.
-According to the National Dynamic Land Cover Dataset (DLCD) [@Lymburner2010], most of these areas have experienced decreases in the Enhanced Vegetation Index (EVI) post 2000 aws derived from satellite data.
-As an index for vegetation greenness, the decreasing values indicate lower biomass over time in the tree cover regions.
+According to the National Dynamic Land Cover Dataset (DLCD) [@Lymburner2010], most of these areas have experienced decreases in the Enhanced Vegetation Index (EVI) post 2000, as derived from satellite data.
+As an index for vegetation greenness, the decreasing values of EVI indicate lower biomass over time in the tree cover regions.
 The possible EVI reduction might be due to land clearing, bush fires or drought.
@@
 Two regions were selected where significant tree cover change since 2000 was reported.
-The first region is located in south central Queensland partly covering the north of the Murray Darling Basin (MDB) (site 1 in Figure \@ref(fig:selreg)).
+The first region is located in south central Queensland (QLD) partly covering the north of the Murray Darling Basin (MDB) (site 1 in Figure \@ref(fig:selreg)).
 High rates of land clearing have been reported in this region during the early 2000s [@SLATS2004].
-The second study region is located at the border of New South Wales and Victoria, and includes the Snowy Mountain ranges (site 2 in Figure \@ref(fig:selreg)).
-Severe bush fires occurred in this area  in early 2003 (see Figure \@ref(fig:figure2bushfire)).
+The second study region is located at the border of New South Wales and Victoria (NSW/VIC), and includes the Snowy Mountain ranges (site 2 in Figure \@ref(fig:selreg)).
+Severe bush fires occurred in this area  in early 2003 (see Figure \@ref(fig:bushfire)).
 The 2003 bush fires were the largest and the worst in the last 60 years [@Fire2011].
@@
 
-```{r figure2bushfire, fig.cap = "Location of bushfires occurring in January 2003, in and around the NSW/VIC study region, as shown by the red pixels.
+```{r bushfire, fig.cap = "Location of bushfires occurring in January 2003, in and around the NSW/VIC study region, as shown by the red pixels.
 The map shows the large area in the Kosciuszko national park that has been burned.
-Some locations in the southwest of teh Australian Capital Territory (ACT) have also experienced intensive bushfires.", out.width="90%", echo=F}
+Some locations in the southwest of the Australian Capital Territory (ACT) have also experienced intensive bushfires.", out.width="90%", echo=F}
 include_graphics("figures/bushfire_nswvic.jpg")
@@
 The main one was the MOD44B product Global Vegetation Continuous Field data set (version 5).
-This data set provides estimates of percent tree cover (percentage of ground surface covered by trees) at a grid resolution of 250 m [@Townshend2011], which is finer then the earlier mentioned burned product MCD45A1.
+This data set provides estimates of percent tree cover (percentage of ground surface covered by trees) at a grid resolution of 250 m [@Townshend2011], which is finer than the earlier mentioned burned product MCD45A1.
 The data set is available on an annual time interval for the study period of 2000 - 2015.
@@
 
-### Regression model {#reg_model}
+## Regression model {#reg_model}
 
@@
 As climate alone are unlikely able to fully explain the temporal variation in rainfall in the regions, including the drought [@vanDijk2013; @Westra2010], this variable was included.
-Since both regions at least partly overlap with the MDB, the average rainfall across was assumed to be a useful explaining variable.
+Since both regions at least partly overlap with the MDB, the average rainfall for the entire basin was assumed to be a useful explaining variable.
 
@@
 Rainfall in each study region was first deseasonalised and detrended using the seasonal decomposition function `ds` in the package `deseasonalise` in R [@Rstats2018].
-Using detrended data gives a better indication of the underlying correlation compared to the correlation between trends in the data [@Smith2012].
+Using detrended data gives a better indication of the underlying correlation by removing the underlying correlation in the data [@Smith2012].
 The cross-correlations between the deseasonalised and detrended rainfall and the climatic indices were tested using the Pearson's product moment correlation method, assuming the relationships are linear.
@@
   \item In NSW/VIC, again the SOI has the strongest correlation with rainfall, followed by the IOD. For both, the strongest correlations occur at the zero time lags. 
-  \item in both cases PDO had the weakest correlations, and this factor was not further considered as a predictor.
-  \item in general, for the better correlated indices, strongest correlations occured at zero time lags.
+  \item In both cases PDO had the weakest correlations, and this factor was not further considered as a predictor.
+  \item In general, for the better correlated indices, strongest correlations occured at zero time lags.
 \end{itemize}
@@
 \begin{array}{lll}
-g(E(\mathbf{R}_r)) = &\beta_0 + s_1(\mathbf{MDB_{monthlyRain}} + \\ &s_2(\mathbf{SOI}) + s_3(\mathbf{IOD}) + \\                &s_4(\mathbf{Nino3.4}) + s_5(\mathbf{Nino4}) + s_6(\mathbf{Season}) + \\
+g(E(\mathbf{R}_r)) = &\beta_0 + s_1(\mathbf{MDB_{monthlyRain}} + s_2(\mathbf{SOI}) + s_3(\mathbf{IOD}) + \\
+&s_4(\mathbf{Nino3.4}) + s_5(\mathbf{Nino4}) + s_6(\mathbf{Season}) + \\
            &\beta_1\mathbf{Trend} + \boldsymbol{\epsilon}_r
@@
 $s_v$ ($_v$=1, 2, 3,...) are the smooth penalized cubic regression spline functions on the climatic indices and the season.
-MDB_{monthlyRain} is the spatially averaged monthly rainfall across the Murray Darling Basin.
+MDB~monthlyRain~ is the spatially averaged monthly rainfall across the Murray Darling Basin.
 Apart from dropping PDO as a predictor, all other climate indices were included, allowing the model to select the appropriate predictors.
@@
 However, the information on the change in type of land cover during the time period is missing.
-Therefore, four scenarios were initially tested in the analysis (**see supplementary material**).
+Therefore, four scenarios were initially tested in the analysis, however there was no real difference between these scenarios.
 In the NSW/VIC region, severe bush fires were reported in early January 2003.
@@
 The advantages of using the Rank-Sum test are: (1) it does not depend on assumptions of the data distribution; (2) it is not restricted to datasets with no missing data; (3) it is robust and not as easily influenced by outliers and negative numbers [@Hirsch1985].
+However, the test has less power than parametric tests.
 
@@
 colnames(table)[3] <- paste("Rank","$R'_{1k}$")
-pander(table, caption="Example of ranking rainfall residuals",
-             style="rmarkdown")
+kable(table, caption="Example of ranking rainfall residuals",
+             booktabs=T)
 
@@
 Hence 2003 and 2004 were excluded from the analysis.
-As a result, the after-change period was 2005 - 2015 for the Queensland location.
+As a result, the post change period was 2005 - 2015 for the Queensland location.
 
@@
 The change was within a relatively short period of the year.
-Therefore the after change period in this region still started in January 2003.
-Following @Hirsch1985, the before period was set to five years (1998 - 2002) in both regions.
-The length of the after period is also difficult for the NSW/VIC region, as the regrowth would at some point have impacted the local effects.
+Therefore the post change period in this region still started in January 2003.
+Following @Hirsch1985, the pre change period was set to five years (1998 - 2002) in both regions.
+The length of the post change period is also difficult for the NSW/VIC region, as the regrowth would at some point have impacted the local effects.
 
@@
 Instead of completing the Mann-Whitney U-test, @Hirsch1985 applied a standard normal Z test to the rank-sum statistics.
-As highlighted, this modified test accounts for serial and cross correlation in the data.
+As highlighted, this modified test accounts for serial temporal autocorrelation and spatial cross correlation in the data.
 In the case here, the deseasonalised and detrended data shows little autocorrelation in the time series but possesses strong cross correlation between neighbouring pixels, i.e.
@@
 
-```{r (ref:Zscore), echo=F}
-table2 <- data_frame(" "= c("Z' $>$ 0 & rainfall decreases after change",
-    "Z' $<$ 0 & rainfall increases after change",
-    "Z' = 0 & rainfall does not change"))
-pander(table2, caption="The interpretation of Z' score in the step trend test", style="rmarkdown")
+```{r Zscore, echo=F}
+table2 <- data_frame(" "= c("$Z'>0$  and rainfall decreases after change",
+    "$Z'<0$ and rainfall increases post change",
+    "$Z'=0$ and rainfall does not change"))
+kable(table2, caption="The interpretation of Z' score in the step trend test",
+      booktabs=T)
 ```
@@
 
-Generally the regression model only explains a limited amount of the rainfall variability (Figure \@reg(fig:rsq_model).
+```{r rsqmodel, fig.cap = "The distribution of the variance explained (adjusted $r^2$ ) by the full regression model", echo=F, out.width="90%"}
+include_graphics("figures/rsq_fullmodel.jpg")
+
+```
+
+Generally the regression model only explains a limited amount of the rainfall variability (Figure \@reg(fig:rsqmodel)).
 The model in Equation \@ref(eq:model2) accounts for an average around 33\% of the variation in the NSW/VIC region and about 50% of the variation in the QLD region as indicated by the adjusted $r^2$.
 The adjusted $r^2$ is the coefficient of determination, a measurement of the amount of variability predicted by the model adjusting for the number of explanatory terms.
-The residual analysis shows that the assumptions of the regression model are generally met (Figure \@ref(fig:residuals).
+The residual analysis shows that the assumptions of the regression model are generally met (Figure \@ref(fig:residuals)).
 The standardised residual plots, however, show some funnelling for the NSW/VIC regions, suggesting non constant variance.
 The residual patterns are consistent for all pixels within each region.
-
-```{r rsq_model, fig.cap = "The distribution of the variance explained (adjusted $r^2$ ) by the full regression model", echo=F, out.width="90%"}
-include_graphics("figures/rsq_fullmodel.jpg")
-
-```
 
@@
 ```{r rsq, fig.cap = "The performance of the regression model if rainfall is only modelled by the climate drivers.
-It shows the percentage of rainfall variability that can be explained by the climate drivers for the Qld and NSW/VIC region", echo=F, out.width="90%"}
+It shows the percentage of rainfall variability that can be explained by the climate drivers for the Qld and NSW/VIC region", echo=F, out.width="70%", out.height="70%"}
 include_graphics("figures/soi_explhistogram.jpg")
@@
 This figure provides two types of information: the sign and the significance level.
-The sign indicates the direction of the step change in the residuals of the regression, as listed in Table \@ref(Zscore).
+The sign indicates the direction of the step change in the residuals of the regression, as listed in Table \@ref(tab:Zscore).
 In both regions there are areas of positive Z' values which imply a decrease in rainfall, but this is stronger in the NSW/VIC area than in the QLD area.
-For the NSW/VIC area there appears to be a reasonable relationship between the locations where changes in tree cover are observed (Figure \@ref(fig:tc_trend)) and the patterns in the Z' score.
+For the NSW/VIC area there appears to be a reasonable relationship between the locations where changes in tree cover are observed (Figure \@ref(fig:tctrend)) and the patterns in the Z' score.
 However, there is not necessarily a direct relationship as movement of air masses could mean that actual changes of rainfall are observed close by, but not necessarily exactly at areas with changes of landcover.
@@
 ```{r meandiff, fig.cap ="Boxplots of annual rainfall residuals (estimated based on Equation 2 before and after the land cover intervention during 1979 - 2015 in the study regions.
-On average, the after period has a significantly lower annual rainfall residual in NSW/VIC, but a signifcantly higher annual rainfall residual in the Qld study area", echo=F, out.width="90%"}
+On average, the after period has a significantly lower annual rainfall residual in NSW/VIC, but a signifcantly higher annual rainfall residual in the Qld study area", echo=F, out.width="70%", out.height="70%"}
 
@@
 For the Queensland locations, the mean monthly rainfall residuals were slightly lower  ($p < 0.05$) after the change in landcover ($~ 0.45 mm/month$).
-For the NSW/VIC locations there was larger decrease in the mean monthly rainfall residual ($p < 0.05$) after the change($~ 1.5 mm/month$).
+For the NSW/VIC locations there was larger decrease in the mean monthly rainfall residual ($p < 0.05$) post change ($~ 1.5 mm/month$).
 
@@
 The cases of $ns = 1$ and $ns = 9$ were also tested.
-The results for $ns = 1$ showed a lower number of significant pixels(at p $<$ 0.10) compared to the $ns = 4$ test.
+The results for $ns = 1$ showed a lower number of significant pixels (at $p< 0.10$) compared to the $ns = 4$ test.
 The results from the $ns = 9$ test was not different from the $ns = 4$ test.
@@
 
-As part of the analysis, the "field significance" of the Z' score test was considered to improve the interpretion of the step change at regional scales from multiple local tests [@Wilks2006, @Westra2013].
+As part of the analysis, the "field significance" of the Z' score test was considered to improve the interpretion of the step change at regional scales from multiple local tests [@Wilks2006; @Westra2013].
 Here, the bootstrapping resampling method from @Westra2013 was used to evaluate the field significance.
@@
 The results support the earlier Z' score results.
-Only the percentage significant positive Z' scores for the NSW/VIC region falls on the tail of the field significance distribution, while all the others are well within the distribution.
+Only the percentage significant positive Z' scores, indicating a decrease in rainfall for the NSW/VIC region as a result in a decrease in tree cover, falls on the tail of the field significance distribution, while all the others are well within the distribution.
 This suggests that the percentage of significant positive Z' scores in the NSW/VIC region is least likely due to a unique series of rainfall years, but is most likely due to changes in landcover.
-In other words, in the NSW/VIC region it is most likely that the change in landcover caused a decrease in the rainfall.
-
-```{r fieldsig, fig.cap ="Field significance results showing the distribution of % of significant Z' scores for the resampled series (blue bars), and the % significant Z'scores for the original series of rainfall years (red circle)", echo=F, out.width="90%"}
-
+In other words, in the NSW/VIC region it is most likely that the change in landcover (decrease in tree cover) caused a decrease in the rainfall.
+
+```{r fieldsig, fig.cap ="Field significance results showing the distribution of percentage of significant $Z'$ scores for the resampled series (blue bars), and the precentage significant $Z'$ scores for the original series of rainfall years (red circle)", echo=F, out.width="70%"}
 include_graphics("figures/bt4_hist.jpg")
@@
 
-```{r summarytable, echo=F, message=F}
-library(kableExtra)
-
-table <- read_csv("TestSummaryTable2.csv")
-kable(table,caption="Summary table of all tests on the two regions ", format = "latex", booktabs =TRUE) %>%
+```{r summarytable, echo=F, message=F, eval=T}
+
+table <- as.data.frame(matrix(c("LC variable","Small area of significant pixels in the centre","Large area possibly aligning with bushfire affected area.",
+"t-test regression model residuals","Slightly lower mean residual ($p<0.05$) after change (0.5 mm per month)","Lower mean residual ($p<0.05$) after change (1.5 mm per month)",
+"Positive score","0.2\\% of pixels at $p<0.1$","3.7\\% of pixel at $p<0.1$",
+"Field significance \\% of scores", "Both positive and negative \\% within random distribution", "Positive \\% outside random distribution."
+),nrow=4,ncol=3, byrow=T))
+kable(table,caption="Summary table of all tests on the two regions ", 'latex',
+      booktabs =TRUE, escape=F,col.names = c("Test","Qld" ,"NSW VIC")) %>%
+#column_spec(2:3, width = "5cm") %>%
   kable_styling(latex_options = "scale_down")#%>%
@@
 Overall the summary table of the results indicates that the effects of land clearing or bushfire on rainfall are consistent across the range of different statistical tests.
-In all cases, the data for the NSW/VIC region
+In all cases, the data for the NSW/VIC region indicates the strongest evidence that the reduction in tree cover (the land cover change) resulted in a local decrease in rainfall.
+In contrast the data for the QLD region only indicates weak evidence that the loss of trees due to land clearing resulted in reduction in the rainfall.
 
 Generally, empirical studies on LCC-precipitation interaction are conducted within an area with known land surface intervention [e.g.
-@Otterman1990,@Durieux2003,@Negri2004,@Sato2007].
+@Otterman1990; @Durieux2003; @Negri2004; @Sato2007].
 However, these locations are rare and difficult to isolate from real landscape change.
-Modelling studies are abundant, but are generally not linked to observed data.
-In this study we therefore tested the effect of land cover change across a broad region, rather than only for locations where changes were known to occur or have occured.
-The advantage of the suggested approach is that it does not require a long time series of land cover data as this is usually unavailable.
+Modelling studies are abundant [e.g.
+@Chagnon2005; @Pinto2009; @Wang2009], but these are generally not directly linked to observed data.
+In this study we tested the effect of land cover change across a broad region, which included locations where changes were known to occur or have occured.
+The advantage of the current approach is that long time series of land cover data are not required.
+This is advantageous as this type of that is currently not always available.
 Furthermore, it does not assume a specific relationship between vegetation cover change and rainfall but allows the data to show this relationship, by applying the analysis to a broader area outside the boundary of the vegetation cover change.
@@
 
-While there is some indication that the observed landuse changes (Figure \@ref(fig:tctrend)) cause a decrease in the rainfall, our analysis has not been able to give an unequivocal answer.
-There are several possible complicating fatcors in the data that could explain the differences between the two regions and the two time periods.
+Overall the results suggest that at least for the NSW/VIC region that a decrease in tree cover causes a decrease in rainfall (Figure \@ref(fig:tctrend)).
+However, there are several possible complicating factors in the data that require discussion.
 
@@
 
-The first and most significant effect on the results remains the variability in the rainfall.
-The regression model used here is a simple model.
-We only consider the important effects of a historical trend, seasonality and climate drivers.
+Rainfall in Australia is highly variable from year to year, and the time period in this study included a severe drought [@vanDijk2013], as well as the drought breaking years 2010 and 2011.
 The purpose of using the regression model is to remove the year to year and month to month variability in rainfall and therefore strengthen the land cover signal in the residuals.
-However, the model shows that trend, seasonality, ENSO and IOD together explain no more than 30\% of the rainfall variability, and only around 7\% on average is due to the climate drivers (Figure \@ref(fig:rsq) .
+We used the spatially averaged monthly rainfall timeseries for the larger Murray Darling Basin (Figure \@ref(fig:selreg)) to remove the general variability.
+However, overall the model does not explain more than 50\% (QLD) and 30\% (NSW/VIC) of the rainfall variability, and only around 7\% on average is due to the climate drivers (Figure \@ref(fig:rsq)) .
 And while this is consistent with the literature e.g.
-@Westra2010, a large amount of variation is left in the model residuals.
+@Westra2010, this means some variation due to external factors could still be left in the residuals.
 Rainfall is generally considered a stochastic process [e.g.
-@Fowler2005,@Cowpertwait2009,@Burton2010] and clearly some of the variability is either a different response to a combination of climate factors (as interactions were not tested in the model), or a strongly non-stationary response to the climate drivers.
-The remaining variability in the residuals increases the difficulty to detect a change in rainfall.
-This is clearly demonstrated in the difference in land cover variable significance and the Z' score between the series to 2009 and the series to 2015.
-While 2009 was at the end of the drought, the series to 2015 includes the record breaking year 2010 (Figure \@ref(fig:ts-mean)).
-
+@Fowler2005;@Cowpertwait2009;@Burton2010] and some of the variability could either be a different response to a combination of climate factors (as interactions were not tested in the model), or a non-stationary response to the climate drivers.
 The severe bushfires in 2003 were also triggered by the extreme drought conditions during the millenium drought [@vanDijk2013].
-Although the drought on rainfall has been accounted for in part by the model, a further delayed or cumulative impact of drought could be feeding into the local land-atmosphere interaction.
-As a result, the rainfall feedback to the vegetation cover change could be weak under the dry conditions, and this could have affected the result.
+Although the effect of the drought on the overall rainfall quantities has been accounted for in part by the model, a further delayed or cumulative effect of the drought could be feeding into the local land-atmosphere interaction.
+As a result, the rainfall feedback to the vegetation cover change could be weak under the dry conditions between 2001 and 2009, and this could have affected the result.
 
@@
 
-The second possible effect is the dynamic nature of the vegetation clearing and recovery.
+The second possible effect is the dynamic nature of the vegetation clearing and recovery, especially for the QLD region.
 Although land clearing has occurred at a high rate and broad scale in Queensland [@SLATS2017], the clearing does not have a clear start and end point.
@@
 Major broad scale and high rate clearings occurred in 1999 - 2000 and 2002 - 2004 (Figure \@ref(fig:slat)).
-And even though there was a decrease in land clearing post 2005, it is difficult to define a clear cut change in this region.
-The more continuous ongoing land clearing could have reduced the significance of a step change.
+And even though there was a decrease in land clearing post 2005, it is difficult to define a clear cut change in the tree cover in this region.
+The specific vegetation class in the Queensland area is also well-known for rapid regrowth and "thickening" in favourable conditions [@Gowen2016], and this could explain the increase in vegetation cover in 2015 in the Qld area (Figure \@ref(fig:tctrend)).
+In particular the favourable rainfall years of 2010 and 2011 would have boosted regrowth, increasing evapotranspiration and therefore decreasing the effect of the rainfall change.
 
@@
 
-The two approaches used in this study indicate quite different results on the level of change in the NSW/VIC region.
-The regression model showed that a large area in the NSW/VIC region has experienced a significant land cover effect ($p < 0.05$) on the rainfall after 2003.
-The effect in the step trend test (Z' scores) is much mroe subdued, even though it was able to detect significant changes ($p < 0.1$) within the area in the period to 2009, but almost no effect in the period to 2015.
-In contrast, the results are almost reversed in the Qld area.
-While some of this could be due to the difficulty in removing year on year rainfall variability, some of this might also be due to vegetation regrowth in both locations.
-
-The specific vegetation class in the Queensland area is well-known for rapid regrowth and "thickening" in favourable conditions [@Gowen2016], and this could explain the change in vegetation cover between 2009 and 2015 in the Qld area (Figure \@ref(fig:tctrend)).
-In particular the favourable rainfall years of 2010 and 2011 would have boosted regrowth, increasing evapotranspiration and therefore decreasing the effect of the rainfall change.
-
 The different causes of vegetation cover change in these two regions could lead to different post-change characteristics.
-The magnitude of EVI decreasing trends in the QLD region are less than in NSW region, as reported in the DLCD data.
-This is due to the lower tree density in the QLD region than in the NSW/VIC region before land surface interventions.
-The significant bushfires (Figure \@ref(fig:fig2bushfire)) would have drastically reduced the vegetation cover and recovery was very slow in some areas within the NSW/VIC area.
-The persistent drought in the 2000s [@Howden2012,@vanDijk2013] delayed the regrowth of trees.
-On the other hand, replacing tree cover with pasture and crops in the Qld area might have a relatively subtle impact on the EVI.
-
-## Misalignment of rainfall effects and vegetation clearing
-
-Particularly in the NSW/VIC area, there is no direct overlap between the pixels with significant vegetation change (Figure \@ref(fig:tctrend)) and the pixels with significant land cover variables or positive Z'scores (Figure \@ref(fig:LCp) and Figure \@ref(fig:steptest30)).
-While this seems possibly surprising at first, a plausible explanation could be the general movement of moisture and climate systems over the landscape, resulting in a possible shift of where the moisture is evaporated and where rainfall occurs.
-Specifically for this reason, this study selected rather large spatial boxes to capture the overall response rather than the pixel by pixel comparison/
+The trends in EVI in the QLD region are opposite to the NSW/VIC region, based on the DLCD data.
+This could also be due to the lower tree density in the QLD region compared to the NSW/VIC region before land surface interventions.
+In contrast, the significant bushfires (Figure \@ref(fig:bushfire)) would have drastically reduced the vegetation cover and recovery was very slow in some areas within the NSW/VIC area.
+The persistent drought in the 2000s [@Howden2012; @vanDijk2013] delayed the regrowth of trees.
+Conversely, replacing tree cover with pasture and crops in the Qld area might only have a relatively subtle impact on the EVI.
 
@@
 
-The rainfall data used in this study is a gridded data set.
-This data set is robust and consistent over a long time series (from 1900 to current) and has a broad national wide coverage which can provide more information spatially.
-However, high cross correlation between pixels, due to the interpolation method generating this data set, can also introduce spatial noise.
+The rainfall data used in this study is gridded.
+This data set is robust and consistent over a long time series (from 1900 to current) and has a broad national wide coverage which can provide more information spatially [@SILO2001; @Beesley2009].
+However, high cross correlation between pixels, due to the interpolation method in this data set [@SILO2001], can also introduce spatial noise.
 In the step trend test the cross correlation was accounted for.
-Some other methods are also available which can be used to perform a comparative trial.
-For example, [@Narisma2007] applied a spatial Gaussian filter on a similar data set and used wavelet analysis to detect step change in rainfall.
+However, some other methods are also available, which could be used to perform a comparative trial.
+For example, @Narisma2007 applied a spatial Gaussian filter on a similar data set and used wavelet analysis to detect a step change in rainfall.
 High quality station data is another option to test whether the observed spatial pattern in the step trend test results was not due to the gridded data itself.
-Resampling methods, such as bootstrapping and permutation [@Wilks1997,@Kundzewicz2004,@Westra2013], can also be used to further assess the strength of significance of results and incorporate spatial and temporal patterns in the analysis.
-While the gridded data set is most useful in regions with sparse rain gauge networks, it can actually reduces information where the rain gauge density is high [@Jones2009].
-In the NSW/VIC area, the coverage of rainfall stations is more intensive but they are mainly located in the valleys.
-The interpolated data might not be the best represention of the local rainfall.
+Resampling methods, such as bootstrapping and permutation (used in this study) [@Wilks1997; @Kundzewicz2004; @Westra2013], can also be used to further assess the strength of the significance of results .
+While the gridded data set is most useful in regions with sparse rain gauge networks, such as in this study, it can actually reduce information where the rain gauge density is high [@Jones2009].
+In the NSW/VIC region, the coverage of rainfall stations is more intensive, but they are mainly located in the valleys.
+As a result, the interpolated data might be a limited representation of the true local rainfall.
 
@@
 
-Parametric tests are generally more powerful than nonparametric test in detecting a trend, when the data is normally distributed [@Onoz2003,@Kundzewicz2004].
-As a non-parametric test, the step trend test has the advantages of distribution free and having no restriction on missing data [@Hirsch1985].
-This is particularly useful in rainfall analysis since rainfall data is usually skewed.
-On the other hand, the disadvantages of non-parametric tests, such as being limited to hypothesis testing and weaker in power, also hold for the step trend test [@Whitley2002].
-
-Overall, the current study provides a clear approach build on several lines of evidence to provide some evidence to reject the null hypothesis (no step change in rainfall occurs as a result of tree cover loss).
-Limited by the available data, the time frame under study was chosen to include a long lasting drought period [@Holper2011, @vanDijk2013].
-The strong impact of this prolonged drought might have suppressed the land-atmosphere interaction and confused the cause and effect relation between rainfall and vegetation cover change.
-This could be one of the reasons that the LCC effects on the local climate found in other studies [e.g.
-@Gorgen2006,@McAlpine2007] are not found to be significant here.
-Possible future work could focus on a non-drought period, or when a longer series of land cover data is available.
-Wile the power of the test can be improved with the longer length of the after-intervention period [@Hirsch1985], the dynamic nature of vegetation regrowth in this case study prevents this effect.
+Parametric tests are generally more powerful than nonparametric test in detecting a trend, when the data is normally distributed [@Onoz2003; @Kundzewicz2004].
+As a non-parametric test, the step trend test has the advantage of being distribution free and having no restriction on missing data [@Hirsch1985].
+This is particularly useful in rainfall analysis since rainfall data is usually highly skewed.
+On the other hand, the disadvantages of non-parametric tests, such as being limited to hypothesis testing and being weaker in power, also hold for the step trend test [@Whitley2002].
+
+Overall, the current study provides a statistical data based approach, building on several lines of evidence to  reject the null hypothesis (no step change in rainfall occurs as a result of tree cover loss) at least for the NSW/VIC region.
+Limited by the available data, the time frame under study included a long lasting drought period [@Holper2011; @vanDijk2013].
+The strong impact of this prolonged drought might have suppressed the land-atmosphere interaction and modified the cause and effect relationship between rainfall and vegetation cover change.
+This could be one of the reasons that the land cover change effects on the local climate found in other studies [e.g.
+@Gorgen2006; @McAlpine2007] do not appear in the QLD region.
+Possible future work could focus on a non-drought period, or using a different series of land cover data, possibly on another continent.
+The recent large bushfires on the US continent come to mind as a possible candidate study once sufficient post event data is collected.
+Wile the power of the test can be improved with the longer length of the post-intervention period [@Hirsch1985], the dynamic nature of vegetation regrowth in this case study also affects this effect.
 A better approach might be to build a global study that investigates multiple locations where drastic landcover changes have taken place, which would also remove some of the climate variability effects due to the larger sample size.
@@
 
-In this study, we present an approach to identify if
-
-found some observationl data based evidence, although not strong, that vegetation cover change has changed local rainfall.
-The semi-parametric method and non-parametric method did totally agree on detecting a significant step change in rainfall in the hot dry QLD region where land clearing has occurred.
-On the other hand, the bushfires in the humid, temperate mountain range in the NSW/VIC region has experienced reduced rainfall.
-But the dry spell also plays an important role in the results.
-
-Drought has had a pronounced impact on the land surface condition during the study period, leading to significant reduction in vegetation and extreme events such as bushfires.
-The associated lack of rainfall and high temperatures may mask the step change in the vegetation.
-Hence, the signal of LCC feedback on rainfall is probably weaker under such regional dry conditions, as the impact of LCC on rainfall is mainly through changes in moisture convergence [@Gorgen2006,@Pitman2007].
-
-acknowledgments
-===============
-CL was supported by an Australian Postgraduate Award for this work.
-
-\newpage
+In this study, we present a statistical approach to identify the impact of a change in land cover on local rainfall.The results, based on gridded rainfall data found,  that a reduction in vegetation cover is likely to have reduced local rainfall for a large area in NSW/VIC affected by bushfires.
+However, land clearing in QLD was unlikely to have reduced rainfall over the same time period.
+
+Drought may have had a pronounced impact on the land surface condition during the study period, also leading to significant reduction in vegetation cover and extreme events such as bushfires.
+The lack of rainfall and associated high temperatures may mask the impact on rainfall of a step change in the vegetation cover.
+Hence, the signal of Land Cover Change feedback on rainfall is probably weaker under such regional dry conditions, as the impact of Land Cover Change on rainfall is mainly through changes in moisture convergence [@Gorgen2006; @Pitman2007].
+
 appendix
 ==============
-
-## Summary of Data
+*Summary of Data*
+
+ \begin{sidewaystable}
+ \caption{Summary of data.}
+  \label{tab:ch3Data}
+ \begin{tabular}{lllll}
+  \hline
+  \textbf{Data} & \textbf{Source} & \multicolumn{2}{c}{\textbf{Resolution}} & \textbf{Analysis period} \\\cline{3-4}
+  & & Temporal & Spatial & \\\hline
+  Percent tree cover & MOD44B & Annual &	250m & 2000-2010\\
+  Trend of vegetation cover change  &	DLCD (2009)	& Onetime & 250m	& Trend of Apr 2000 - Apr 2008\\
+  Rainfall &	AWAP gridded rainfall data &	Monthly &	0.05\textdegree$\times$0.05\textdegree & Jan 1979- Dec 2008\\
+  SOI	& BoM	& Monthly &	N/A &	Jan 1979- Dec 2008\\
+  NINO 3, 3.4, 4 &	IRI/LDEO data library	& Monthly	& N/A	& Jan 1979- Dec 2008\\
+  PDO	& NOAA & Monthly	& N/A	& Jan 1979- Dec 2008\\
+  IOD	& POAMA-2 dataset	& Monthly	& N/A	& Jan 1979- Dec 2008\\
+  \hline
+  \end{tabular}
+\end{sidewaystable}
 

### 2019-03-14 Manuscripts/PeerJRainfall_Landuse/Rainfall_Landuse.tex
paper for submission

@@
 
-\usepackage{lipsum} \usepackage{textcomp}
+\usepackage{lipsum} \usepackage{textcomp, rotating}
 
@@
 \begin{abstract}
-Analysis of observational data to identify relationships between
-rainfall and land cover change are difficult due to multiple
-environmental factors that cannot be strictly controlled. In this study
-we present a methodology to investigate the relationship using
-statistical methods on data from best available sources at two sites in
-Australia. Gridded data of rainfall and tree cover were used as
-spatially corresponding local conditions. Large scale effects were
-represented by climatic indicators, such as SOI and IOD. Regression
-analysis and step trend tests were used to assess the effect of abrupt
-land surface intervention. At a Queensland site, significant tree cover
-change between 2002 - 2005 did not result in strong statistically
-significant precipitation changes. On the other hand, results from a
-bushfire affected NSW/VIC region suggests significant changes in the
-rainfall. This indicates the method works better when a abrupt change in
-the data can be clearly identified. The results from the step trend test
-implied a positive relationship between the tree cover and the rainfall
-at 0.1 significance level in both locations in data up to 2009. However,
-high rainfall variability and possible regrowth meant that no signifcant
-changes were observed in alonger time sereies to 2015.
+Analysis of observational data to pinpoint impact of land cover change
+on local rainfall are difficult due to multiple environmental factors
+that cannot be strictly controlled. In this study we use a statistical
+approach to identify the relationship between removal of tree cover and
+rainfall with data from best available sources for two large areas in
+Australia. Gridded rainfall data between 1979 and 2015 was used for the
+areas, while large scale (exogenous) effects were represented by mean
+rainfall across a much larger area and climatic indicators, such as SOI
+and IOD. Both general additive modelling and step trend tests were used
+for the analysis. For a region in south central Queensland, the reported
+change in tree clearing between 2002 - 2005 did not result in strong
+statistically significant precipitation changes. On the other hand,
+results from a bushfire affected region on the border of New SOuth Wales
+and Victoria suggests significant changes in the rainfall due to changes
+in tree cover. This indicates the method works better when a abrupt
+change in the data can be clearly identified. The results from the step
+trend test also mainly identified a positive relationship between the
+tree cover and the rainfall at p \textless{} 0.1 at the NSW/Victoria
+region. High rainfall variability and possible regrowth could have
+impacted the results in the Queensland region.
 % Dummy abstract text.
@@
 Israel (Otterman et al. 1990; Ben-Gai et al. 1998). Using airborne
-measurement in Western Australia, Junkermann et al. (2009) showed a
+measurements in Western Australia, Junkermann et al. (2009) showed a
 significantly higher level of aerosols over an agricultural area
-compared to an adjacent natural vegetation. They suggested that a
-modification of aerosol concentrations due to deforestation could have
-contributed to a reduction of local rainfall, as more, but smaller rain
-droplets were observed. Nair et al. (2011) reported from the Bunny Fence
-Experiment in Western Australia that local land use change altered the
-synoptic west coast trough dynamics and surface roughness, and this
+compared to an adjacent area with natural vegetation. They suggested
+that a modification of aerosol concentrations due to deforestation could
+have contributed to a reduction of local rainfall, as more, but smaller
+rain droplets were observed. Nair et al. (2011) reported from the Bunny
+Fence Experiment in Western Australia that local land use change altered
+the synoptic west coast trough dynamics and surface roughness, and this
 resulted in an observed rainfall decrease. Maximum temperatures were
@@
 demonstrated in many modelling studies. More specifically, we
-hypothesize that a step change on the land cover on the surface will
-cause a step change in the rainfall. To demonstrate this we study
-changes in observed rainfall over time at a Queensland and NSW/Victoria
-location where there are possible step changes in land cover change due
-to land clearing and bush fires. The methodology uses statistical
-approaches to identify changes in rainfall, which are subsequently
-associated with land cover change through spatial comparison.
+hypothesize that a step change in land cover on the surface will cause a
+step change in the rainfall. To demonstrate this we study changes in
+observed rainfall over time at a Queensland and NSW/Victoria location
+where there are possible step changes in land cover change due to land
+clearing and bush fires. The methodology uses statistical approaches to
+identify changes in rainfall, which are subsequently associated with
+land cover change through spatial comparison.
 
@@
 Dataset (DLCD) (Lymburner et al. 2010), most of these areas have
-experienced decreases in the Enhanced Vegetation Index (EVI) post 2000
-aws derived from satellite data. As an index for vegetation greenness,
-the decreasing values indicate lower biomass over time in the tree cover
-regions. The possible EVI reduction might be due to land clearing, bush
-fires or drought.
+experienced decreases in the Enhanced Vegetation Index (EVI) post 2000,
+as derived from satellite data. As an index for vegetation greenness,
+the decreasing values of EVI indicate lower biomass over time in the
+tree cover regions. The possible EVI reduction might be due to land
+clearing, bush fires or drought.
 
@@
 was reported. The first region is located in south central Queensland
-partly covering the north of the Murray Darling Basin (MDB) (site 1 in
-Figure @ref(fig:selreg)). High rates of land clearing have been reported
-in this region during the early 2000s (Department of Natural Resources
-and Water 2007). The second study region is located at the border of New
-South Wales and Victoria, and includes the Snowy Mountain ranges (site 2
-in Figure @ref(fig:selreg)). Severe bush fires occurred in this area in
-early 2003 (see Figure @ref(fig:figure2bushfire)). The 2003 bush fires
-were the largest and the worst in the last 60 years (The State
-Government of Victoria 2011). Two thirds of Kosciuszko national park was
-heavily burned and regrowth was reported to be slow due to drought and
-cold conditions (ABC News 2003), and the type of species in this region.
-However, in the longer term, after an early high transpiration period a
-recovery of pre-fire evapotranspiration would be expected (Kuczera
-1987). For the purpose of this study, significant tree cover loss has
-happened in both study areas in the last decade, either permanently or
-temporarily.
+(QLD) partly covering the north of the Murray Darling Basin (MDB) (site
+1 in Figure \ref{fig:selreg}). High rates of land clearing have been
+reported in this region during the early 2000s (Department of Natural
+Resources and Water 2007). The second study region is located at the
+border of New South Wales and Victoria (NSW/VIC), and includes the Snowy
+Mountain ranges (site 2 in Figure \ref{fig:selreg}). Severe bush fires
+occurred in this area in early 2003 (see Figure \ref{fig:bushfire}). The
+2003 bush fires were the largest and the worst in the last 60 years (The
+State Government of Victoria 2011). Two thirds of Kosciuszko national
+park was heavily burned and regrowth was reported to be slow due to
+drought and cold conditions (ABC News 2003), and the type of species in
+this region. However, in the longer term, after an early high
+transpiration period a recovery of pre-fire evapotranspiration would be
+expected (Kuczera 1987). For the purpose of this study, significant tree
+cover loss has happened in both study areas in the last decade, either
+permanently or temporarily.
 
@@
 The map shows the large area in the Kosciuszko national park that has been burned.
-Some locations in the southwest of teh Australian Capital Territory (ACT) have also experienced intensive bushfires.}\label{fig:figure2bushfire}
+Some locations in the southwest of the Australian Capital Territory (ACT) have also experienced intensive bushfires.}\label{fig:bushfire}
 \end{figure}
@@
 regions, the post-change vegetation status is hypothesized to be
-different as well (see Figure @ref(fig:figure3-tc-simple)).
+different as well (see Figure \ref{fig:figure3-tc-simple}).
 
@@
 
-\hypertarget{Data}{\subsection{Data}\label{Data}}
+\hypertarget{Data}{\section{Data}\label{Data}}
 
@@
 of ground surface covered by trees) at a grid resolution of 250 m
-(Townshend et al. 2011), which is finer then the earlier mentioned
+(Townshend et al. 2011), which is finer than the earlier mentioned
 burned product MCD45A1. The data set is available on an annual time
@@
 . The data has been projected onto a national 0.05\textdegree  by
-0.05\textdegree grid (approximately 5 km by 5 km). This gridded data set
-was generated from station observations using spline interpolation and
-kriging (Jeffrey et al. 2001). The data has been compared to other
+0.05\textdegree  grid (approximately 5 km by 5 km). This gridded data
+set was generated from station observations using spline interpolation
+and kriging (Jeffrey et al. 2001). The data has been compared to other
 gridded products and observed data and is generally of high quality
@@
 
-\subsection{Statistical method}\label{statistical-method}
+\section{Statistical method}\label{statistical-method}
 
@@
 below). A step change is not directly obvious in the time series of the
-rainfall anomalies (Figure @ref(fig:ts-mean)) for both regions, even
+rainfall anomalies (Figure \ref{fig:ts-mean}) for both regions, even
 though the data is deseasonalised and detrended. In this study we apply
@@
 
-\subsubsection{Regression model}\label{reg_model}
+\subsection{Regression model}\label{reg_model}
 
@@
 this variable was included. Since both regions at least partly overlap
-with the MDB, the average rainfall across was assumed to be a useful
-explaining variable.
+with the MDB, the average rainfall for the entire basin was assumed to
+be a useful explaining variable.
 
@@
 \texttt{deseasonalise} in R (R Core Team, 2018). Using detrended data
-gives a better indication of the underlying correlation compared to the
-correlation between trends in the data (Smith and Timbal 2012). The
+gives a better indication of the underlying correlation by removing the
+underlying correlation in the data (Smith and Timbal 2012). The
 cross-correlations between the deseasonalised and detrended rainfall and
@@
 Based on the correlation between the climatic indices and rainfall in
-the regions (as shown in Figure @ref(fig:cor-rain-qld) and Figure
-@ref(fig:cor-rain-nsw)), it can be concluded that:
+the regions (as shown in Figure \ref{fig:cor-rain-qld} and Figure
+\ref{fig:cor-rain-nsw}), it can be concluded that:
 
@@
 and Tibshirani 1986) with a log link function \texttt{g()} and assuming
-the residuals are gamma distributed (see Figure @ref(fig:hist-rain)).
+the residuals are gamma distributed (see Figure \ref{fig:hist-rain}).
 This means all predictors are modelled as smooth functions, in this case
@@
 \begin{array}{lll}
-g(E(\mathbf{R}_r)) = &\beta_0 + s_1(\mathbf{MDB_{monthlyRain}} + \\ &s_2(\mathbf{SOI}) + s_3(\mathbf{IOD}) + \\                &s_4(\mathbf{Nino3.4}) + s_5(\mathbf{Nino4}) + s_6(\mathbf{Season}) + \\
+g(E(\mathbf{R}_r)) = &\beta_0 + s_1(\mathbf{MDB_{monthlyRain}} + s_2(\mathbf{SOI}) + s_3(\mathbf{IOD}) + \\
+&s_4(\mathbf{Nino3.4}) + s_5(\mathbf{Nino4}) + s_6(\mathbf{Season}) + \\
            &\beta_1\mathbf{Trend} + \boldsymbol{\epsilon}_r
 \end{array}
-(\#eq:model)
+\label{eq:model}
 \end{equation}
@@
 smooth penalized cubic regression spline functions on the climatic
-indices and the season. MDB\_\{monthlyRain\} is the spatially averaged
-monthly rainfall across the Murray Darling Basin. Apart from dropping
-PDO as a predictor, all other climate indices were included, allowing
-the model to select the appropriate predictors.
+indices and the season. MDB\textsubscript{monthlyRain} is the spatially
+averaged monthly rainfall across the Murray Darling Basin. Apart from
+dropping PDO as a predictor, all other climate indices were included,
+allowing the model to select the appropriate predictors.
 
@@
 variable was used in the regression model to represent the abrupt land
-surface change (see Equation @ref(eq:model)). The change could be a
+surface change (see Equation \eqref{eq:model}). The change could be a
 result of either land clearing or bush fires as long as it is permanent
@@
 significance of the coefficient of \textbf{LC}, denoted as \(\beta'_5\)
-in Equation @ref(eq:model), can be determined by a ratio test.
+in Equation \eqref{eq:model}, can be determined by a ratio test.
 
@@
   \end{array} \right.\\
-  (\#eq:dummyC)
+\label{eq:dummyC}
 \end{equation}
@@
 \end{array}
-  (\#eq:model2)
+\label{eq:model2}
 \end{equation}
@@
 period is missing. Therefore, four scenarios were initially tested in
-the analysis (\textbf{see supplementary material}). In the NSW/VIC
-region, severe bush fires were reported in early January 2003. Hence the
-``tree'' cover state was up to December 2002 then it was changed to
-``removed'' state from January 2003. As a starting date, the regression
-model was run from 1979 for both regions.
+the analysis, however there was no real difference between these
+scenarios. In the NSW/VIC region, severe bush fires were reported in
+early January 2003. Hence the ``tree'' cover state was up to December
+2002 then it was changed to ``removed'' state from January 2003. As a
+starting date, the regression model was run from 1979 for both regions.
 
@@
 
-To support the regression analysis, a step trend test was used to detect
-changes in rainfall as a result of vegetation cover change. This
-nonparametric statistical test was modified from the Mann-Whitney
-Rank-Sum test by (Hirsch and Gilroy 1985) and can identify a step change
-in data which is cross correlated. The gridded rainfall dataset used in
-this study has a high spatial correlation between neighbouring pixels
-due to the underlying interpolation method. The advantages of using the
-Rank-Sum test therefore are: (1) it does not depend on assumptions of
-the data distribution; (2) it is not restricted to datasets with no
-missing data; (3) it is robust and not as easily influenced by outliers
-and negative numbers (Hirsch and Gilroy 1985).
+To support the regression analysis, a Mann-Whitney Rank-Sum step trend
+test was used to detect changes in rainfall as a result of vegetation
+cover change. This specific nonparametric statistical test was modified
+from the Mann-Whitney U test by Hirsch and Gilroy (1985) and can
+identify a step change in data which is cross-correlated. The gridded
+rainfall dataset used in this study has a high spatial correlation
+between neighbouring pixels due to the underlying interpolation method.
+The advantages of using the Rank-Sum test are: (1) it does not depend on
+assumptions of the data distribution; (2) it is not restricted to
+datasets with no missing data; (3) it is robust and not as easily
+influenced by outliers and negative numbers (Hirsch and Gilroy 1985).
+However, the test has less power than parametric tests.
 
 The rainfall residuals from the regression model in Equation
-@ref(eq:model2) were used for the step trend test. According to (Hirsch
-and Gilroy 1985), to detect a step change, using deseasonalised and
-detrended data is important. Furthermore, as rainfall can only be
-partially attributed to local sources and conditions, other effects are
-introduced by large scale dynamics and changes in other climatic
-factors. The assumption is that the regression model should remove these
-effects, and additionally deseasonalise and detrend the rainfall data.
-As a result, the local landuse effects are amplified in the variation of
-the remaining residuals. The test, described in the following sub
-section, subsequently associates trends in the rainfall residuals with
-tree cover changes.
+\eqref{eq:model2} were used for the step trend test. To detect a step
+change, using deseasonalised and detrended data is important (Hirsch and
+Gilroy 1985). Furthermore, as rainfall can only be partially attributed
+to local sources and conditions, other effects are introduced by large
+scale dynamics and changes in climatic factors. The assumption is that
+the regression model should remove these effects, and additionally
+deseasonalise and detrend the rainfall data. As a result, the local
+landuse effects are amplified in the remaining residuals. The test,
+described in the following sub section, subsequently associates trends
+in the rainfall residuals with tree cover changes.
 
@@
 As indicated, the step trend test is a modified version of the
-Mann-Whitney rank-sum statistic (Hirsch and Gilroy 1985). As a
-nonparametric rank-based test, the Mann-Whitney test does not use the
-exact values of rainfall but depends on the ranks of the data. For each
-month, rainfall residuals of each year were ranked in an ascending
-order. The ranking of January rainfall in a sample pixel k in QLD is
-illustrated below:
-
-\begin{longtable}[]{@{}ccc@{}}
-\caption{Example of ranking rainfall residuals}\tabularnewline
+Mann-Whitney U statistic (Hirsch and Gilroy 1985). As a nonparametric
+rank-based test, the Mann-Whitney test does not use the exact values of
+rainfall but depends on the ranks of the data. For each month, rainfall
+residuals of each year were ranked in an ascending order. The ranking of
+January rainfall in a sample pixel k in QLD is illustrated below:
+
+\begin{table}[t]
+
+\caption{\label{tab:exampleRank}Example of ranking rainfall residuals}
+\centering
+\begin{tabular}{rrr}
 \toprule
-Year & Rainfall residuals & Rank \(R'_{1k}\)\tabularnewline
+Year & Rainfall residuals & Rank \$R'\_\{1k\}\$\\
 \midrule
-\endfirsthead
-\toprule
-Year & Rainfall residuals & Rank \(R'_{1k}\)\tabularnewline
-\midrule
-\endhead
-1998 & -0.3 & 6\tabularnewline
-1999 & -60.9 & 2\tabularnewline
-2000 & -16.1 & 4\tabularnewline
-2001 & -71.7 & 1\tabularnewline
-2002 & 111.1 & 7\tabularnewline
-2005 & -7.2 & 5\tabularnewline
-2006 & -60.5 & 3\tabularnewline
+1998 & -0.3 & 6\\
+1999 & -60.9 & 2\\
+2000 & -16.1 & 4\\
+2001 & -71.7 & 1\\
+2002 & 111.1 & 7\\
+\addlinespace
+2005 & -7.2 & 5\\
+2006 & -60.5 & 3\\
 \bottomrule
-\end{longtable}
+\end{tabular}
+\end{table}
 
@@
 ignored. Hence 2003 and 2004 were excluded from the analysis. As a
-result, the after-change period was 2005 - 2015 for the Queensland
+result, the post change period was 2005 - 2015 for the Queensland
 location.
@@
 The change was within a relatively short period of the year. Therefore
-the after change period in this region still started in January 2003.
-Following Hirsch and Gilroy (1985), the before period was set to five
-years (1998 - 2002) in both regions. The length of the after period is
-difficult for the NSW/VIC as the regrowth would at some point have
-impacted the local effects.
-
-There is a further complication in the data. The year 2009 was the final
-year of the millenium drought (Dijk and Viney 2013), followed by two
-very wet years (also visible in Figure @ref(fig:ts-mean) in 2010 and
-2011. As highlighted earlier, this means the the length of after period
-could influence the analysis by the inclusion of more or fewer years. as
-a result the period up to (and including) 2009 was compared to the full
-period up to 2015. Further analysis up to 2010 is included in the
-supplementary material.
+the post change period in this region still started in January 2003.
+Following Hirsch and Gilroy (1985), the pre change period was set to
+five years (1998 - 2002) in both regions. The length of the post change
+period is also difficult for the NSW/VIC region, as the regrowth would
+at some point have impacted the local effects.
 
@@
   W_{jk} = \sum_{i=1}^{n_1}R'_{ijk}.
-  (\#eq:Wj)
+  \label{eq:Wj}
 \end{equation}
@@
 applied a standard normal Z test to the rank-sum statistics. As
-highlighted, this modified test accounts for serial and cross
-correlation in the data. In the case here, the deseasonalised and
-detrended data shows little autocorrelation in the time series but
-possesses strong cross correlation between neighbouring pixels, i.e.
-\(R>0.99\).
+highlighted, this modified test accounts for serial temporal
+autocorrelation and spatial cross correlation in the data. In the case
+here, the deseasonalised and detrended data shows little autocorrelation
+in the time series but possesses strong cross correlation between
+neighbouring pixels, i.e. \(R>0.99\).
 
@@
   Z'=\frac{\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk}-12\cdot ns\cdot\mu_w}{\sqrt{Var(\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk})}}.
-  (\#eq:Z)
+  \label{eq:Z}
 \end{equation}
@@
 
+\begin{table}[t]
+
+\caption{\label{tab:Zscore}The interpretation of Z' score in the step trend test}
+\centering
+\begin{tabular}{l}
+\toprule
+ \\
+\midrule
+\$Z'>0\$  and rainfall decreases after change\\
+\$Z'<0\$ and rainfall increases post change\\
+\$Z'=0\$ and rainfall does not change\\
+\bottomrule
+\end{tabular}
+\end{table}
+
 The null hypothesis (\(H_0\)) in this study is that there was no change
@@
 trend test can be interpreted according to the sign of the Z' score (see
-Table @ref(tab:Zscore) (Chapter 23, P887 Hipel and McLeod 1994)). Z' is
+Table \ref{tab:Zscore}, Chapter 23, P887 (Hipel and McLeod 1994)). Z' is
 normally distributed similar to the standard normal statistics Z. Hence
@@
 
-\begin{longtable}[]{@{}c@{}}
-\caption{The interpretation of Z' score in the step trend
-test}\tabularnewline
-\toprule
-Z' \(>\) 0 \& rainfall decreases after change\tabularnewline
-Z' \(<\) 0 \& rainfall increases after change\tabularnewline
-Z' = 0 \& rainfall does not change\tabularnewline
-\bottomrule
-\end{longtable}
-
 \section{Results}\label{results}
@@
 significant (\(p \leq 0.05\)) in each study region, are shown in Figure
-@ref(fig:tctrend) for the time series to 2009 (left panels) and the time
-series to 2015 (right panels). In both cases it can be seen that the
-area of negative cover change was greater in the series to 2009 than in
-the series to 2015. But even the longer series indicates a large change
-in the NSW/VIC region (bottom row). In the NSW/VIC region, much of the
-tree loss between 2002 and 2003 was concentrated in the Snowy Mountains
-close to the border of NSW and VIC. Tree cover loss occurred in large
+\ref{fig:tctrend} for the NSW/VIC region (left panel) and the QLD region
+(right panel). In the QLD region there actually has generally been an
+increase in tree cover after the clearing of native vegetation stopped.
+In the NSW/VIC region, much of the tree loss between 2002 and 2003 was
+concentrated in the Snowy Mountains close to the border of NSW and VIC,
+which is also evident in the figure. Tree cover loss occurred in large
 parts of the QLD region between 2002 and 2005, but this tree loss was
-spatially less concentrated. Most of the clearings appear in the centre
-of this region in the series to 2009. The tree cover change map
-(@ref(fig:tctrend) is consistent with the annual mean EVI trend map
-(based on DLCD data, not shown here). However in the series to 2015 the
-tree cover loss seems to have dissappeared (top right Figure
-@ref(fig:tctrend)) and suggests an increase in tree cover over the area.
-The implications of this will be discussed in more detail in the
-discussion.
+spatially less concentrated.
 
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{figures/tc_figs} \caption{The maps show signifcant changes in tree cover identified from the MOD44B data between 2003 and 2009 (top left) and 2003 and 2015 (top right) in the NSW/VIC region and the Qld region (bottom left to 2009 and bottom right to 2015).
+\includegraphics[width=0.9\linewidth]{figures/tc_figs} \caption{The maps show significant changes in tree cover identified from the MOD44B data between 2003 and 2015 in the NSW/VIC region (right) and the Qld region (left).
 The amount of change was calculated as the difference in tree cover before and after the specified land cover intervention and it is shown as the percentage of the ground area.
@@
 
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{figures/rsq_fullmodel} \caption{The distribution of the variance explained (adjusted $r^2$ ) by the full regression model}\label{fig:rsqmodel}
+\end{figure}
+
 Generally the regression model only explains a limited amount of the
-rainfall variability. The model in Equation @ref(eq:model2) accounts for
-around
-15\%\footnote{Here the adjusted $R^2$ was reported.
-Adjusted $R^2$ is the coefficient of determination, a measurement of the amount of variability predicted by the model adjusting for the number of explanatory terms}
-of the rainfall variation in both regions. The residual analysis shows
-that the assumptions of the regression model are generally met (Figure
-@ref(fig:residuals). The standardised residual plots, however, show some
-funnelling for the NSW/VIC regions, suggesting non constant variance.
-The residual patterns are consistent for all pixels within each region.
-
-The model, however, confirms the importance of the climate drivers and
-the seasonality in Australian rainfall. Even at the grid level, the
-seasons and several of the climatic indices were significant
-(\(p \leq 0.05\)) everywhere in both regions. The explaining power of
-the model is mostly due to these variables. The climate drivers (at lag
-zero) accounted, on average, for 6.7\% of the rainfall variability in
-both the QLD region and the NSW/VIC region (see Figure @ref(fig:rsq) for
-the distribution of \(R^2\) in these two regions). These figures are
-within the upper bound of seasonal rainfall predictability by a SST
-anomaly field reported by (Westra and Sharma 2010).
+rainfall variability (Figure @reg(fig:rsqmodel)). The model in Equation
+\eqref{eq:model2} accounts for an average around 33\% of the variation in
+the NSW/VIC region and about 50\% of the variation in the QLD region as
+indicated by the adjusted \(r^2\). The adjusted \(r^2\) is the
+coefficient of determination, a measurement of the amount of variability
+predicted by the model adjusting for the number of explanatory terms.
+The residual analysis shows that the assumptions of the regression model
+are generally met (Figure \ref{fig:residuals}). The standardised
+residual plots, however, show some funnelling for the NSW/VIC regions,
+suggesting non constant variance. The residual patterns are consistent
+for all pixels within each region.
+
+In terms of predictors of the local rainfall in the model, logically,
+the average rainfall across the larger Murray Darling Basin is highly
+significant. This confirms that this variable is a good reflection of
+the year on year variability in the rainfall. The model also confirms
+the importance of the climate drivers and the seasonality in Australian
+rainfall, as many of these variables were significant. Even at the grid
+level, the seasons and several of the climatic indices were significant
+(\(p \leq 0.05\)) everywhere in both regions. The climate drivers (at
+lag zero) accounted, on average, for 6.5\% of the rainfall variability
+in both the QLD region and the NSW/VIC region (see Figure \ref{fig:rsq}
+for the distribution of adjusted \(r^2\) in these two regions). These
+figures are within the upper bound of seasonal rainfall predictability
+by a SST anomaly field reported by Westra and Sharma (2010).
 
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{figures/soi_explhistogram} \caption{The performance of the regression model if rainfall is only modelled by the climate drivers.
+\includegraphics[width=0.7\linewidth,height=0.7\textheight]{figures/soi_explhistogram} \caption{The performance of the regression model if rainfall is only modelled by the climate drivers.
 It shows the percentage of rainfall variability that can be explained by the climate drivers for the Qld and NSW/VIC region}\label{fig:rsq}
@@
 
-Statistically significant long term trends were only observed in part of
-the NSW region (results not shown). However, this result might prove or
-disprove the existence a long term trend in rainfall. The overall time
-period is fairly short (Koutsoyiannis 2006) and more pixels in NSW/VIC
-could indicate a significant step change if the long term trend effect
-is not removed by the model. As trend free data is an important
-requirement for the step trend test, the trend term was kept in the
-regression model to ensure the detection of step change was not due to a
-possible long term trend.
+There were generally no statistically significant long term trends in
+both regions. However, this result might not prove or disprove the
+existence a long term trend in rainfall. The overall time period is
+fairly short (Koutsoyiannis 2006) and more pixels in NSW/VIC could
+indicate a significant step change if the long term trend effect is not
+removed by the model. As trend free data is an important requirement for
+the step trend test, the trend term was kept in the regression model to
+ensure the detection of step change was not due to a possible long term
+trend.
 
@@
 \includegraphics[width=0.9\linewidth]{figures/Cp_30yrs} \caption{The spatial distribution of significance of land cover step change variable in the GAM model predicting changes in rainfall in the both regions.
-The top row reflects NSW/VIC, while the bottom relates to the Qld study site.
-The left hand plots are the data up to 2009, while the right hand plots are for the full series.
+The left graph relates to the Qld region, while the right hand plot reflects the NSW/VIC region.
+The outlines of relevant Australian states and the Murray Darling Basin are indicated in grey.
 The p value reported is for the land cover variable in the model.}\label{fig:LCp}
@@
 
-The land cover variable in the model (Equation @ref(eq:model2)) aims to
+The land cover variable in the model (Equation \eqref{eq:model2}) aims to
 identify a step change in the rainfall before and after the observed
-change in land cover. The variable was only significant
+change in land cover. The variable was mainly significant
 (\(p \leq 0.05\)) for the rainfall estimates in some areas in NSW/VIC,
-as shown in Figure @ref(fig:LCp). However, the number of pixels where
-the landcover variable was significant was much greater for the series
-up to 2009, compared to the series to 2015. There was no direct
-relationship between the areas of bushfires in Figure
-@ref(fig:bushfire). No significant step change due to the land cover
-changes was found in rainfall in the QLD region in the series to 2015,
-but a small area of change was identified in the period to 2009 (left
-bottom panel).
-
-Figure @ref(fig:LCp) for NSW/VIC suggests that a significant step change
-in rainfall related to land cover change was found in an area larger
-than where the bushfires has occurred. This might be showing a large
-scale effect that reaches beyond the vegetation cover change effect.
-
-More generally, the model suggests that the tree cover has a positive
-impact on rainfall in both regions. The fitted coefficients for the Land
-cover change variable were consistently positive for the ``tree'' part
-of the series. It implies that rainfall was higher when the surface was
-covered by trees.
+as shown in Figure \ref{fig:LCp} (right panel). However, the number of
+pixels where the landcover variable was significant was much greater in
+NSW/VIC compared to the QLD region. There was also some relationship
+between the areas of bushfires in Figure \ref{fig:bushfire}. Only a very
+small area with a significant step change due to the land cover changes
+was found in rainfall in the QLD region (left panel)
+
+More generally, the model for NSW/VIC suggests that the land cover
+variable has a positive impact on rainfall in both regions. The fitted
+coefficients for the Land cover change variable were consistently
+positive for the ``tree'' part of the series. It implies that rainfall
+was higher when the surface was covered by trees.
 
@@
 \includegraphics[width=0.9\linewidth]{figures/step_new} \caption{Spatial distribution of the step trend test Z' statistics in the two study sites.
-Panels on the top are for the NSW/Vic site, while panels of the bottom ar for Qld.
-The left panels are for the series to 2009, while the right panels are for the full data series.
-Warm colours (yellow, orange and red) are for positive Z' values which indicate decreasing rainfall trend due to the land surface intervention.
+The panel on the left is for the QLD region, while the panel of the right is for NSW/VIC.
+Warm colours (yellow, orange and red) are for positive Z' values which indicate decreasing rainfall trend due to the land surface change.
 Cold colours (light blue to blue) are for negative Z' values which indicate increasing rainfall trend.
@@
 
-The spatial step trend test Z' scores for the Qld region (top row) and
-NSWVIC region (bottom row) are shown in Figure @ref(fig:steptest30). The
-figure also contains the period to 2009 (left panels) and the period to
-2015 (right panels). This figure provides two types of information: the
+The spatial step trend test based on the regression residuals without
+the landcover variable generates Z' scores. The results for the Qld
+region (left) and NSWVIC region (right) are shown in Figure
+\ref{fig:steptest30}. This figure provides two types of information: the
 sign and the significance level. The sign indicates the direction of the
-step change, as listed in Table @ref(Zscore). In each region up to 2009,
-there is a broad area of positive Z' values which implies a decrease in
-rainfall. There appears to be little relationship between the locations
-where changes in tree cover are observed (Figure @ref(fig:tc\_trend))
-and the patterns in the Z' score. However, there is not necessarily a
-direct relationship as movement of air masses could mean that actual
-changes of rainfall are observed close by, but not necessarily exactly
-at areas with changes of landcover. There is once again a difference
-between the left panels and the right panels, indicating that
-considering a longer time series (to 2015) reduces the pattern and
-significance of Z' scores. In the QLD region to 2009, 21\% of the pixels
-obtained a positive Z' score with p \(<\) 0.1, and for the period to
-2015 no pixels with positive Z' score at p \(<\) 0.1 occurred. In the
-NSW/VIC region to 2009, only 2.6\% of pixels in the alpine area having a
-positive Z' score with p \(<\) 0.1. In general it is only a small
-proportion of both study regions.
+step change in the residuals of the regression, as listed in Table
+\ref{tab:Zscore}. In both regions there are areas of positive Z' values
+which imply a decrease in rainfall, but this is stronger in the NSW/VIC
+area than in the QLD area. For the NSW/VIC area there appears to be a
+reasonable relationship between the locations where changes in tree
+cover are observed (Figure \ref{fig:tctrend}) and the patterns in the Z'
+score. However, there is not necessarily a direct relationship as
+movement of air masses could mean that actual changes of rainfall are
+observed close by, but not necessarily exactly at areas with changes of
+landcover. There is once again a difference between the left panel and
+the right panel, indicating only a few significant of Z' scores in the
+QLD region wich matches the earlier significance in the landcover
+variable in the full regresion. In the QLD region, only 0.2\% of the
+pixels obtained a positive Z' score with p \(<\) 0.1. In the NSW/VIC
+region 3.7\% of pixels have a positive Z' score with p \(<\) 0.1. In
+general it is only a small proportion of both study regions.
 
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{figures/ResidualBoxplotchange} \caption{Boxplots of annual rainfall residuals (estimated based on Equation 2 before and after the land cover intervention during 1979 - 2015 in the study regions.
+\includegraphics[width=0.7\linewidth,height=0.7\textheight]{figures/ResidualBoxplotchange} \caption{Boxplots of annual rainfall residuals (estimated based on Equation 2 before and after the land cover intervention during 1979 - 2015 in the study regions.
 On average, the after period has a significantly lower annual rainfall residual in NSW/VIC, but a signifcantly higher annual rainfall residual in the Qld study area}\label{fig:meandiff}
@@
 
-The rainfall residuals of the two periods (before-change since 1979 and
-after-change) were also compared (Figure @ref(fig:meandiff)) using a
-simple t-test. From the boxplots it is difficult to see that there were
-significantly different (\(p < 0.05\)) mean values between the
-``before'' and ``after'' periods in both regions. For the Queensland
-locations, there was no significant difference if measured to 2009 and
-slightly less rain (\(p < 0.05\)) before if measured to 2015
-(\(~ 2 mm/month\)). In contrast, for the NSW/VIC locations there was
-also slightly more rain (\(p < 0.05\)) after the change if measured to
-2015 (\(~ 1 mm/month\)), but significantly less rain (\(p < 0.05\))
-after the change (\(~ 9 mm/month\)) if measured to 2009.
-
-A closer look at the rainfall data indicates that rainfall was
-consistently very low in 2006 for both regions and consistently high for
-2010 @ref(fig:ts-mean). The low rainfall in 2006 was due to a weak El
-Niño and unrecovered conditions from the previous drought, in contrast,
-2010 brough record drought breaking rains across Eastern Australia
-related to a La Niña (Dijk and Viney 2013). While the regression model
-has removed most of the effect of ENSO, the effect of these anomolous
-years could still be visible in the residuals if the response of the
-rainfall to the ENSO effect is non-linear, and for example the memory of
-the past drought is persistent. This can also explain the significant
-results in the previous method. To remove the possible outlier effects
-from 2006 and 2010, the two periods were compared excluding the 2006 and
-2010 rainfall. In addition, the group of pixels showing a negative step
-change in rainfall (\(p < 0.10\)) and the group of pixels indicating no
-change were analysed separately for the period to 2009. An unpaired
-unequal variance two sample t-test was applied to test whether the after
-period had lower mean rainfall than the before period, excluding the
-2006 and 2010 years for the full period, and excluding 2006 for the
-period to 2009.
-
-For the Queensland locations using the regression model residuals and
-excluding 2006 and 2010, there was no difference between the period
-before and after the landuse change for the full data series. However
-for the period to 2009, there is slightly more rain
-(\(p < 0.05, ~ 2mm/month\)) before than after the change. For the
-NSW/VIC locations for the same analyses, there is slightly more rain
-(\(p < 0.05\)) after the change if measured to 2015 (\(~ 2 mm/month\)),
-and for the period to 2009 there is still significantly less rain
-(\(p < 0.05, ~ 5 mm/month\)). Focussing on the pixels with a negative
-step change (positive significant Z' score) and the monthly rainfall
-totals, only the period to 2009 is of interest, as the full period had
-too few significant pixels (@ref(fig:steptest30)). For both regions and
-excluding 2006, the Qld area had significantly more rain before than
-after at \(p < 0.05\). For the NSW/VIC area, there was similarly
-significantly more rain before than after the change
-(\(p < 0.05 and 100mm/month\)). For pixels where the Z' score was not
-significantly positive (\(p > 0.1\) and Z' score negative), the Qld area
-showed no significant difference in monthly rainfall totals before and
-after. For the NSW/VIC area there was still significantly less rain
-after the change (\(p < 0.05\)) but difference in monthly rainfall
-totals was smaller than for positive and signifcant Z' scores: 24
-mm/month less.
-
-The choice of \(ns\) has some impact on the test results (Hirsch and
-Gilroy 1985). The cases of \(ns = 1\) and \(ns = 9\) were also tested.
-The results for \(ns = 1\) showed a lower number of significant
-pixels(at p \(<\) 0.10) compared to the \(ns = 4\) test, with only 66
-pixels in the Queensland area in the period to 2010 and a slightly
-higher 6 pixels in the period to 2015. No pixels were significant at p
-\(<\) 0.1 in the NSW/VIC area. The results from the \(ns = 9\) test was
-not different from the \(ns = 4\) test. The power of the test does not
-change much after \(ns = 4\), as shown by Hirsch and Gilroy (1985).
+The rainfall regression residuals without the landcover variable for the
+two regions (before-change since 1979 and after-change) were also
+compared (Figure \ref{fig:meandiff}) using a simple t-test. From the
+boxplots it is difficult to see that there were significantly different
+(\(p < 0.05\)) mean rainfall residual values between the ``before'' and
+``after'' periods in both regions. For the Queensland locations, the
+mean monthly rainfall residuals were slightly lower (\(p < 0.05\)) after
+the change in landcover (\(~ 0.45 mm/month\)). For the NSW/VIC locations
+there was larger decrease in the mean monthly rainfall residual
+(\(p < 0.05\)) post change (\(~ 1.5 mm/month\)).
+
+The choice of the number of pixels averaged in the step trend test
+(\(ns\)) has some impact on the test results (Hirsch and Gilroy 1985).
+The cases of \(ns = 1\) and \(ns = 9\) were also tested. The results for
+\(ns = 1\) showed a lower number of significant pixels (at \(p< 0.10\))
+compared to the \(ns = 4\) test. The results from the \(ns = 9\) test
+was not different from the \(ns = 4\) test. We therefore decided to stay
+with \(ns = 4\). The power of the test does not change much after
+\(ns = 4\), as shown by Hirsch and Gilroy (1985).
 
 As part of the analysis, the ``field significance'' of the Z' score test
-was considered to better interpret the step change at regional scales
-from multiple local tests (Wilks 2006, Westra, Alexander, and Zwiers
-(2013)). Here, the bootstrapping resampling method from Westra,
-Alexander, and Zwiers (2013) was used to evaluate the field significance
-for the period up to 2009 (as the period to 2015 essentially showed no
-significant positive Z' scores). This means the spatial structure of the
-pixels was maintained, but the order of the years and months was changed
-by random resampling. For each resampling the test statistics identifies
-the percentage of the pixels with significant positive step change for
-the step trend test. The test statistics on 1000 resampled replicates
-were used to develop the distribution of these percentage values under
-the local null hypothesis that there was no step change. The results
-showed that none of the resampled series indicated any significant step
-trend for both regions, suggesting that the actual values found are for
-a unique series of years.
+was considered to improve the interpretion of the step change at
+regional scales from multiple local tests (Wilks 2006; Westra,
+Alexander, and Zwiers 2013). Here, the bootstrapping resampling method
+from Westra, Alexander, and Zwiers (2013) was used to evaluate the field
+significance. This means the spatial structure of the pixels was
+maintained, but the order of the years and months was changed by random
+resampling. For each resampling, the test statistic identifies the
+percentage of the pixels with a significant positive or negative step
+change for the step trend test. The test statistics on 1000 resampled
+replicates were used to develop the distribution of these percentage
+values, and the observed fraction of significant Z' scores (at
+\(p < 0.1\)) is plotted on the distribution (Figure \ref{fig:fieldsig}).
+The results support the earlier Z' score results. Only the percentage
+significant positive Z' scores, indicating a decrease in rainfall for
+the NSW/VIC region as a result in a decrease in tree cover, falls on the
+tail of the field significance distribution, while all the others are
+well within the distribution. This suggests that the percentage of
+significant positive Z' scores in the NSW/VIC region is least likely due
+to a unique series of rainfall years, but is most likely due to changes
+in landcover. In other words, in the NSW/VIC region it is most likely
+that the change in landcover (decrease in tree cover) caused a decrease
+in the rainfall.
+
+\begin{figure}
+\includegraphics[width=0.7\linewidth]{figures/bt4_hist} \caption{Field significance results showing the distribution of percentage of significant $Z'$ scores for the resampled series (blue bars), and the precentage significant $Z'$ scores for the original series of rainfall years (red circle)}\label{fig:fieldsig}
+\end{figure}
 
@@
 
-\begin{verbatim}
-## Warning: package 'kableExtra' was built under R version 3.5.2
-\end{verbatim}
-
-\begin{table}
-
-\caption{\label{tab:summarytable}Summary table of all tests on the two regions and for the two time periods}
+\begin{table}[t]
+
+\caption{\label{tab:summarytable}Summary table of all tests on the two regions }
 \centering
 \resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lllll}
+\begin{tabular}{lll}
 \toprule
-Test & Qld 2009 & Qld 2015 & NSW/VIC 2009 & NSW/VIC 2015\\
+Test & Qld & NSW VIC\\
 \midrule
-LC variable & small area on west side & no pixels significant & Large area & Smaller area than 2009\\
-t-test regression \textbackslash{}\& model residuals & not significant & less rain before & less rain after & slightly less rain after\\
-t-test residuals \textbackslash{}\& excluding 2006 and 2010 & slightly more rain after & no significant difference & significantly less rain after & slightly more rain after\\
-Z' score & 21\% of pixels at p < 0.1 & 1 pixel at p < 0.1 & 2.6\% of pixels at p < 0.1 & no pixels at p < 0.1\\
-t-test pixels positive Z' score & significantly less rain after & NA & 100 mm/month less rain after (p < 0.05) & NA\\
-t-test other pixels & no difference & NA & 24 mm/month less rain after (p < 0.05) & NA\\
+LC variable & Small area of significant pixels in the centre & Large area possibly aligning with bushfire affected area.\\
+t-test regression model residuals & Slightly lower mean residual ($p<0.05$) after change (0.5 mm per month) & Lower mean residual ($p<0.05$) after change (1.5 mm per month)\\
+Positive score & 0.2\% of pixels at $p<0.1$ & 3.7\% of pixel at $p<0.1$\\
+Field significance \% of scores & Both positive and negative \% within random distribution & Positive \% outside random distribution.\\
 \bottomrule
@@
 Overall the summary table of the results indicates that the effects of
-land clearing or bushfire on rainfall are not easily detected, even with
-a range of different statistical tests. In addition several of the tests
-appear to indicate contradictory results, in particular in the
-comparison between the two regions and between the two time periods.
+land clearing or bushfire on rainfall are consistent across the range of
+different statistical tests. In all cases, the data for the NSW/VIC
+region indicates the strongest evidence that the reduction in tree cover
+(the land cover change) resulted in a local decrease in rainfall. In
+contrast the data for the QLD region only indicates weak evidence that
+the loss of trees due to land clearing resulted in reduction in the
+rainfall.
 
@@
 conducted within an area with known land surface intervention (e.g.
-Otterman et al. 1990,Durieux, Machado, and Laurent (2003),Negri et al.
-(2004),Sato, Kimura, and Hasegawa (2007)). However, these locations are
+Otterman et al. 1990; Durieux, Machado, and Laurent 2003; Negri et al.
+2004; Sato, Kimura, and Hasegawa 2007). However, these locations are
 rare and difficult to isolate from real landscape change. Modelling
-studies are abundant, but are generally not linked to observed data. In
-this study we therefore tested the effect of land cover change across a
-broad region, rather than only for locations where changes were known to
-occur or have occured. The advantage of the suggested approach is that
-it does not require a long time series of land cover data as this is
-usually unavailable. Furthermore, it does not assume a specific
-relationship between vegetation cover change and rainfall but allows the
-data to show this relationship, by applying the analysis to a broader
-area outside the boundary of the vegetation cover change. This approach
-is expected to provide a way to reduce the risk of a false positive
-paradox, by comparing results between areas with and without vegetation
-cover change.
-
-While there is some indication that the observed landuse changes (Figure
-@ref(fig:tctrend)) cause a decrease in the rainfall, our analysis has
-not been able to give an unequivocal answer. There are several possible
-complicating fatcors in the data that could explain the differences
-between the two regions and the two time periods.
+studies are abundant (e.g. Chagnon and Bras 2005; Pinto et al. 2009;
+Wang et al. 2009), but these are generally not directly linked to
+observed data. In this study we tested the effect of land cover change
+across a broad region, which included locations where changes were known
+to occur or have occured. The advantage of the current approach is that
+long time series of land cover data are not required. This is
+advantageous as this type of that is currently not always available.
+Furthermore, it does not assume a specific relationship between
+vegetation cover change and rainfall but allows the data to show this
+relationship, by applying the analysis to a broader area outside the
+boundary of the vegetation cover change. This approach is expected to
+provide a way to reduce the risk of a false positive paradox, by
+comparing results between areas with and without vegetation cover
+change.
+
+Overall the results suggest that at least for the NSW/VIC region that a
+decrease in tree cover causes a decrease in rainfall (Figure
+\ref{fig:tctrend}). However, there are several possible complicating
+factors in the data that require discussion.
 
@@
 
-The first and most significant effect on the results remains the
-variability in the rainfall. The regression model used here is a simple
-model. We only consider the important effects of a historical trend,
-seasonality and climate drivers. The purpose of using the regression
-model is to remove the year to year and month to month variability in
-rainfall and therefore strengthen the land cover signal in the
-residuals. However, the model shows that trend, seasonality, ENSO and
-IOD together explain no more than 30\% of the rainfall variability, and
-only around 7\% on average is due to the climate drivers (Figure
-@ref(fig:rsq) . And while this is consistent with the literature e.g.
-Westra and Sharma (2010), a large amount of variation is left in the
-model residuals. Rainfall is generally considered a stochastic process
-(e.g. Fowler et al. 2005,Cowpertwait, Salinger, and Mullen (2009),Burton
-et al. (2010)) and clearly some of the variability is either a different
-response to a combination of climate factors (as interactions were not
-tested in the model), or a strongly non-stationary response to the
-climate drivers. The remaining variability in the residuals increases
-the difficulty to detect a change in rainfall. This is clearly
-demonstrated in the difference in land cover variable significance and
-the Z' score between the series to 2009 and the series to 2015. While
-2009 was at the end of the drought, the series to 2015 includes the
-record breaking year 2010 (Figure @ref(fig:ts-mean)).
-
-The severe bushfires in 2003 were also triggered by the extreme drought
-conditions during the millenium drought (Dijk and Viney 2013). Although
-the drought on rainfall has been accounted for in part by the model, a
-further delayed or cumulative impact of drought could be feeding into
-the local land-atmosphere interaction. As a result, the rainfall
-feedback to the vegetation cover change could be weak under the dry
-conditions, and this could have affected the result.
+Rainfall in Australia is highly variable from year to year, and the time
+period in this study included a severe drought (Dijk and Viney 2013), as
+well as the drought breaking years 2010 and 2011. The purpose of using
+the regression model is to remove the year to year and month to month
+variability in rainfall and therefore strengthen the land cover signal
+in the residuals. We used the spatially averaged monthly rainfall
+timeseries for the larger Murray Darling Basin (Figure \ref{fig:selreg})
+to remove the general variability. However, overall the model does not
+explain more than 50\% (QLD) and 30\% (NSW/VIC) of the rainfall
+variability, and only around 7\% on average is due to the climate
+drivers (Figure \ref{fig:rsq}) . And while this is consistent with the
+literature e.g. Westra and Sharma (2010), this means some variation due
+to external factors could still be left in the residuals. Rainfall is
+generally considered a stochastic process (e.g. Fowler et al. 2005;
+Cowpertwait, Salinger, and Mullen 2009; Burton et al. 2010) and some of
+the variability could either be a different response to a combination of
+climate factors (as interactions were not tested in the model), or a
+non-stationary response to the climate drivers. The severe bushfires in
+2003 were also triggered by the extreme drought conditions during the
+millenium drought (Dijk and Viney 2013). Although the effect of the
+drought on the overall rainfall quantities has been accounted for in
+part by the model, a further delayed or cumulative effect of the drought
+could be feeding into the local land-atmosphere interaction. As a
+result, the rainfall feedback to the vegetation cover change could be
+weak under the dry conditions between 2001 and 2009, and this could have
+affected the result.
 
@@
 The second possible effect is the dynamic nature of the vegetation
-clearing and recovery. Although land clearing has occurred at a high
-rate and broad scale in Queensland (Department of Science, Information
-Technology and Innovation 2017), the clearing does not have a clear
-start and end point. QLD has a long history of land clearing. According
-to the series of SLATS reports on land cover changes in QLD released by
-the Queensland government, land clearing continued in and around the
-study region between 1988 - 2008. Major broad scale and high rate
-clearings occurred in 1999 - 2000 and 2002 - 2004 (Figure
-@ref(fig:slat)). And even though there was a decrease in land clearing
-post 2005, it is difficult to define a clear cut change in this region.
-The more continuous ongoing land clearing could have reduced the
-significance of a step change.
+clearing and recovery, especially for the QLD region. Although land
+clearing has occurred at a high rate and broad scale in Queensland
+(Department of Science, Information Technology and Innovation 2017), the
+clearing does not have a clear start and end point. QLD has a long
+history of land clearing. According to the series of SLATS reports on
+land cover changes in QLD released by the Queensland government, land
+clearing continued in and around the study region between 1988 - 2008.
+Major broad scale and high rate clearings occurred in 1999 - 2000 and
+2002 - 2004 (Figure \ref{fig:slat}). And even though there was a
+decrease in land clearing post 2005, it is difficult to define a clear
+cut change in the tree cover in this region. The specific vegetation
+class in the Queensland area is also well-known for rapid regrowth and
+``thickening'' in favourable conditions (Gowen and Bray 2016), and this
+could explain the increase in vegetation cover in 2015 in the Qld area
+(Figure \ref{fig:tctrend}). In particular the favourable rainfall years
+of 2010 and 2011 would have boosted regrowth, increasing
+evapotranspiration and therefore decreasing the effect of the rainfall
+change.
 
@@
 
-The two approaches used in this study indicate quite different results
-on the level of change in the NSW/VIC region. The regression model
-showed that a large area in the NSW/VIC region has experienced a
-significant land cover effect (\(p < 0.05\)) on the rainfall after 2003.
-The effect in the step trend test (Z' scores) is much mroe subdued, even
-though it was able to detect significant changes (\(p < 0.1\)) within
-the area in the period to 2009, but almost no effect in the period to
-2015. In contrast, the results are almost reversed in the Qld area.
-While some of this could be due to the difficulty in removing year on
-year rainfall variability, some of this might also be due to vegetation
-regrowth in both locations.
-
-The specific vegetation class in the Queensland area is well-known for
-rapid regrowth and ``thickening'' in favourable conditions (Gowen and
-Bray 2016), and this could explain the change in vegetation cover
-between 2009 and 2015 in the Qld area (Figure @ref(fig:tctrend)). In
-particular the favourable rainfall years of 2010 and 2011 would have
-boosted regrowth, increasing evapotranspiration and therefore decreasing
-the effect of the rainfall change.
-
 The different causes of vegetation cover change in these two regions
-could lead to different post-change characteristics. The magnitude of
-EVI decreasing trends in the QLD region are less than in NSW region, as
-reported in the DLCD data. This is due to the lower tree density in the
-QLD region than in the NSW/VIC region before land surface interventions.
-The significant bushfires (Figure @ref(fig:fig2bushfire)) would have
-drastically reduced the vegetation cover and recovery was very slow in
-some areas within the NSW/VIC area. The persistent drought in the 2000s
-(Howden 2012,Dijk and Viney (2013)) delayed the regrowth of trees. On
-the other hand, replacing tree cover with pasture and crops in the Qld
-area might have a relatively subtle impact on the EVI.
-
-\subsection{Misalignment rainfall effects and vegetation
-clearing}\label{misalignment-rainfall-effects-and-vegetation-clearing}
-
-Particularly in the NSW/VIC area, there is no direct overlap between the
-pixels with significant vegetation change (Figure @ref(fig:tctrend)) and
-the pixels with significant land cover variables or positive Z'scores
-(Figure @ref(fig:LCp) and Figure @ref(fig:steptest30)). While this seems
-possibly surprising at first, a plausible explanation could be the
-general movement of moisture and climate systems over the landscape,
-resulting in a possible shift of where the moisture is evaporated and
-where rainfall occurs. Specifically for this reason, this study selected
-rather large spatial boxes to capture the overall response rather than
-the pixel by pixel comparison/
+could lead to different post-change characteristics. The trends in EVI
+in the QLD region are opposite to the NSW/VIC region, based on the DLCD
+data. This could also be due to the lower tree density in the QLD region
+compared to the NSW/VIC region before land surface interventions. In
+contrast, the significant bushfires (Figure \ref{fig:bushfire}) would
+have drastically reduced the vegetation cover and recovery was very slow
+in some areas within the NSW/VIC area. The persistent drought in the
+2000s (Howden 2012; Dijk and Viney 2013) delayed the regrowth of trees.
+Conversely, replacing tree cover with pasture and crops in the Qld area
+might only have a relatively subtle impact on the EVI.
 
@@
 
-The rainfall data used in this study is a gridded data set. This data
-set is robust and consistent over a long time series (from 1900 to
-current) and has a broad national wide coverage which can provide more
-information spatially. However, high cross correlation between pixels,
-due to the interpolation method generating this data set, can also
-introduce spatial noise. In the step trend test the cross correlation
-was accounted for. Some other methods are also available which can be
-used to perform a comparative trial. For example, (Narisma et al. 2007)
-applied a spatial Gaussian filter on a similar data set and used wavelet
-analysis to detect step change in rainfall. High quality station data is
-another option to test whether the observed spatial pattern in the step
-trend test results was not due to the gridded data itself. Resampling
-methods, such as bootstrapping and permutation (Wilks 1997,Kundzewicz
-and Robson (2004),Westra, Alexander, and Zwiers (2013)), can also be
-used to further assess the strength of significance of results and
-incorporate spatial and temporal patterns in the analysis. While the
-gridded data set is most useful in regions with sparse rain gauge
-networks, it can actually reduces information where the rain gauge
-density is high (Jones, Wang, and Fawcett 2009). In the NSW/VIC area,
-the coverage of rainfall stations is more intensive but they are mainly
-located in the valleys. The interpolated data might not be the best
-represention of the local rainfall.
+The rainfall data used in this study is gridded. This data set is robust
+and consistent over a long time series (from 1900 to current) and has a
+broad national wide coverage which can provide more information
+spatially (Jeffrey et al. 2001; Tozer, Kiem, and Verdon-Kidd 2009).
+However, high cross correlation between pixels, due to the interpolation
+method in this data set (Jeffrey et al. 2001), can also introduce
+spatial noise. In the step trend test the cross correlation was
+accounted for. However, some other methods are also available, which
+could be used to perform a comparative trial. For example, Narisma et
+al. (2007) applied a spatial Gaussian filter on a similar data set and
+used wavelet analysis to detect a step change in rainfall. High quality
+station data is another option to test whether the observed spatial
+pattern in the step trend test results was not due to the gridded data
+itself. Resampling methods, such as bootstrapping and permutation (used
+in this study) (Wilks 1997; Kundzewicz and Robson 2004; Westra,
+Alexander, and Zwiers 2013), can also be used to further assess the
+strength of the significance of results . While the gridded data set is
+most useful in regions with sparse rain gauge networks, such as in this
+study, it can actually reduce information where the rain gauge density
+is high (Jones, Wang, and Fawcett 2009). In the NSW/VIC region, the
+coverage of rainfall stations is more intensive, but they are mainly
+located in the valleys. As a result, the interpolated data might be a
+limited representation of the true local rainfall.
 
@@
 detecting a trend, when the data is normally distributed (Onoz and
-Bayazit 2003,Kundzewicz and Robson (2004)). As a non-parametric test,
-the step trend test has the advantages of distribution free and having
+Bayazit 2003; Kundzewicz and Robson 2004). As a non-parametric test, the
+step trend test has the advantage of being distribution free and having
 no restriction on missing data (Hirsch and Gilroy 1985). This is
 particularly useful in rainfall analysis since rainfall data is usually
-skewed. On the other hand, the disadvantages of non-parametric tests,
-such as being limited to hypothesis testing and weaker in power, also
-hold for the step trend test (Whitley and Ball 2002).
-
-Overall, the current study provides a clear approach build on several
-lines of evidence to provide some evidence to reject the null hypothesis
-(no step change in rainfall occurs as a result of tree cover loss).
-Limited by the available data, the time frame under study was chosen to
-include a long lasting drought period (Holper 2011, Dijk and Viney
-(2013)). The strong impact of this prolonged drought might have
-suppressed the land-atmosphere interaction and confused the cause and
-effect relation between rainfall and vegetation cover change. This could
-be one of the reasons that the LCC effects on the local climate found in
-other studies (e.g. Görgen et al. 2006,McAlpine et al. (2007)) are not
-found to be significant here. Possible future work could focus on a
-non-drought period, or when a longer series of land cover data is
-available. Wile the power of the test can be improved with the longer
-length of the after-intervention period (Hirsch and Gilroy 1985), the
-dynamic nature of vegetation regrowth in this case study prevents this
-effect. A better approach might be to build a global study that
-investigates multiple locations where drastic landcover changes have
-taken place, which would also remove some of the climate variability
-effects due to the larger sample size.
+highly skewed. On the other hand, the disadvantages of non-parametric
+tests, such as being limited to hypothesis testing and being weaker in
+power, also hold for the step trend test (Whitley and Ball 2002).
+
+Overall, the current study provides a statistical data based approach,
+building on several lines of evidence to reject the null hypothesis (no
+step change in rainfall occurs as a result of tree cover loss) at least
+for the NSW/VIC region. Limited by the available data, the time frame
+under study included a long lasting drought period (Holper 2011; Dijk
+and Viney 2013). The strong impact of this prolonged drought might have
+suppressed the land-atmosphere interaction and modified the cause and
+effect relationship between rainfall and vegetation cover change. This
+could be one of the reasons that the land cover change effects on the
+local climate found in other studies (e.g. Görgen et al. 2006; McAlpine
+et al. 2007) do not appear in the QLD region. Possible future work could
+focus on a non-drought period, or using a different series of land cover
+data, possibly on another continent. The recent large bushfires on the
+US continent come to mind as a possible candidate study once sufficient
+post event data is collected. Wile the power of the test can be improved
+with the longer length of the post-intervention period (Hirsch and
+Gilroy 1985), the dynamic nature of vegetation regrowth in this case
+study also affects this effect. A better approach might be to build a
+global study that investigates multiple locations where drastic
+landcover changes have taken place, which would also remove some of the
+climate variability effects due to the larger sample size.
 
@@
 
-In this study, we present an approach to identify iffound some
-observationl data based evidence, although not strong, that vegetation
-cover change has changed local rainfall. The semi-parametric method and
-non-parametric method did totally agree on detecting a significant step
-change in rainfall in the hot dry QLD region where land clearing has
-occurred. On the other hand, the bushfires in the humid, temperate
-mountain range in the NSW/VIC region has experienced reduced rainfall.
-But the dry spell also plays an important role in the results.
-
-Drought has had a pronounced impact on the land surface condition during
-the study period, leading to significant reduction in vegetation and
-extreme events such as bushfires. The associated lack of rainfall and
-high temperatures may mask the step change in the vegetation. Hence, the
-signal of LCC feedback on rainfall is probably weaker under such
-regional dry conditions, as the impact of LCC on rainfall is mainly
-through changes in moisture convergence (Görgen et al. 2006,Pitman and
-Hesse (2007)).
-
-\section{acknowledgments}\label{acknowledgments}
-
-CL was supported by an Australian Postgraduate Award for this work.
-
-\newpage
+In this study, we present a statistical approach to identify the impact
+of a change in land cover on local rainfall.The results, based on
+gridded rainfall data found, that a reduction in vegetation cover is
+likely to have reduced local rainfall for a large area in NSW/VIC
+affected by bushfires. However, land clearing in QLD was unlikely to
+have reduced rainfall over the same time period.
+
+Drought may have had a pronounced impact on the land surface condition
+during the study period, also leading to significant reduction in
+vegetation cover and extreme events such as bushfires. The lack of
+rainfall and associated high temperatures may mask the impact on
+rainfall of a step change in the vegetation cover. Hence, the signal of
+Land Cover Change feedback on rainfall is probably weaker under such
+regional dry conditions, as the impact of Land Cover Change on rainfall
+is mainly through changes in moisture convergence (Görgen et al. 2006;
+Pitman and Hesse 2007).
 
@@
 
-\subsection{Summary of Data}\label{summary-of-data}
+\emph{Summary of Data}
+
+\begin{sidewaystable}
+ \caption{Summary of data.}
+  \label{tab:ch3Data}
+ \begin{tabular}{lllll}
+  \hline
+  \textbf{Data} & \textbf{Source} & \multicolumn{2}{c}{\textbf{Resolution}} & \textbf{Analysis period} \\\cline{3-4}
+  & & Temporal & Spatial & \\\hline
+  Percent tree cover & MOD44B & Annual &    250m & 2000-2010\\
+  Trend of vegetation cover change  &   DLCD (2009) & Onetime & 250m    & Trend of Apr 2000 - Apr 2008\\
+  Rainfall &    AWAP gridded rainfall data &    Monthly &   0.05\textdegree$\times$0.05\textdegree & Jan 1979- Dec 2008\\
+  SOI   & BoM   & Monthly & N/A &   Jan 1979- Dec 2008\\
+  NINO 3, 3.4, 4 &  IRI/LDEO data library   & Monthly   & N/A   & Jan 1979- Dec 2008\\
+  PDO   & NOAA & Monthly    & N/A   & Jan 1979- Dec 2008\\
+  IOD   & POAMA-2 dataset   & Monthly   & N/A   & Jan 1979- Dec 2008\\
+  \hline
+  \end{tabular}
+\end{sidewaystable}
 
@@
 in the Amazon.'' \emph{Geophysical Research Letters} 32 (13).
-\href{\%3CGo\%20to\%20ISI\%3E://WOS:000230640100004\%20http://www.agu.org/journals/gl/gl0513/2005GL022722/2005GL022722.pdf}{\textless{}Go to ISI\textgreater{}://WOS:000230640100004 http://www.agu.org/journals/gl/gl0513/2005GL022722/2005GL022722.pdf}.
 
@@
 Deforestation.'' \emph{Remote Sensing of Environment} 86 (1): 132--40.
-\url{http://www.sciencedirect.com/science/article/pii/S0034425703000956}.
 
@@
 Climate} 17 (6): 1306--19.
-\href{\%3CGo\%20to\%20ISI\%3E://WOS:000220450100013\%20http://journals.ametsoc.org/doi/pdf/10.1175/1520-0442\%282004\%29017\%3C1306\%3ATIOADO\%3E2.0.CO\%3B2}{\textless{}Go to ISI\textgreater{}://WOS:000220450100013 http://journals.ametsoc.org/doi/pdf/10.1175/1520-0442\%282004\%29017\%3C1306\%3ATIOADO\%3E2.0.CO\%3B2}.
 
@@
 Science Reviews} 26 (1-2): 189--200.
-\url{http://www.sciencedirect.com/science/article/B6VBC-4MC71SN-1/2/0e1d4dff6e15a1de79cefbfa612f9b37}.
 
@@
 \emph{Journal of Geophysical Research-Atmospheres} 112 (D24).
-\href{\%3CGo\%20to\%20ISI\%3E://WOS:000252013700002\%20http://www.agu.org/journals/jd/jd0724/2006JD008129/2006JD008129.pdf}{\textless{}Go to ISI\textgreater{}://WOS:000252013700002 http://www.agu.org/journals/jd/jd0724/2006JD008129/2006JD008129.pdf}.
 
@@
 States of America} 106 (10): 3670.
-\url{http://ezproxy.library.usyd.edu.au/login?url=http://search.proquest.com/docview/201417978?accountid=14757}.
 
@@
 Meteorological Society: 3332--51.
-\href{http://ezproxy.library.usyd.edu.au/login?url=http://search.proquest.com/docview/814127501?accountid=14757\%20http://DD8GH5YX7K.search.serialssolutions.com/?SS_Source=3\&genre=article\&sid=ProQ:\&atitle=An+Upper+Limit+to+Seasonal+Rainfall+Predictability\%3F\&title=Journal+of+Climate\&issn=0894-8755\&date=2010-06-01\&volume=23\&issue=12\&spage=3332\&SS_docid=814127501\&author=Westra\%2C+Seth\%3BSharma\%2C+Ashish}{http://ezproxy.library.usyd.edu.au/login?url=http://search.proquest.com/docview/814127501?accountid=14757 http://DD8GH5YX7K.search.serialssolutions.com/?SS\_Source=3\&genre=article\&sid=ProQ:\&atitle=An+Upper+Limit+to+Seasonal+Rainfall+Predictability\%3F\&title=Journal+of+Climate\&issn=0894-8755\&date=2010-06-01\&volume=23\&issue=12\&spage=3332\&SS\_docid=814127501\&author=Westra\%2C+Seth\%3BSharma\%2C+Ashish}.
 
@@
 American Meteorological Society: 1181--9.
-\href{http://ezproxy.library.usyd.edu.au/login?url=http://search.proquest.com/docview/224582143?accountid=14757\%20http://DD8GH5YX7K.search.serialssolutions.com/?ctx_ver=Z39.88-2004\&ctx_enc=info:ofi/enc:UTF-8\&rfr_id=info:sid/ProQ\%3Amilitary\&rft_val_fmt=info:ofi/fmt:kev:mtx:journal\&rft.genre=article\&rft.jtitle=Journal+of+Applied+Meteorology+and+Climatology\&rft.atitle=On+\%22Field+Significance\%22+and+the+False+Discovery+Rate\&rft.au=Wilks\%2C+D+S\&rft.aulast=Wilks\&rft.aufirst=D\&rft.date=2006-09-01\&rft.volume=45\&rft.issue=9\&rft.spage=1181\&rft.isbn=\&rft.btitle=\&rft.title=Journal+of+Applied+Meteorology+and+Climatology\&rft.issn=15588424}{http://ezproxy.library.usyd.edu.au/login?url=http://search.proquest.com/docview/224582143?accountid=14757 http://DD8GH5YX7K.search.serialssolutions.com/?ctx\_ver=Z39.88-2004\&ctx\_enc=info:ofi/enc:UTF-8\&rfr\_id=info:sid/ProQ\%3Amilitary\&rft\_val\_fmt=info:ofi/fmt:kev:mtx:journal\&rft.genre=article\&rft.jtitle=Journal+of+Applied+Meteorology+and+Climatology\&rft.atitle=On+\%22Field+Significance\%22+and+the+False+Discovery+Rate\&rft.au=Wilks\%2C+D+S\&rft.aulast=Wilks\&rft.aufirst=D\&rft.date=2006-09-01\&rft.volume=45\&rft.issue=9\&rft.spage=1181\&rft.isbn=\&rft.btitle=\&rft.title=Journal+of+Applied+Meteorology+and+Climatology\&rft.issn=15588424}.
 

### 2019-03-14 Manuscripts/test.Rmd
paper for submission

---
title: "Untitled"
author: "Willem"
date: "13 March 2019"
output: pdf_document
---
```{r packages, echo=F, warnings=F, message=F, eval=T}
require(knitr)
require(kableExtra)
```

Try out the table

```{r summarytable, echo=F, message=F, eval=T}

table <- as.data.frame(matrix(c("LC variable","Small area of significant pixels in the centre","Large area possibly aligning with bushfire affected area.",
"t-test regression model residuals","Slightly lower mean residual ($p<0.05$) after change (0.5 mm per month)","Lower mean residual ($p<0.05$) after change (1.5 mm per month)",
"Positive score","0.2\\% of pixels at $p<0.1$","3.7\\% of pixel at $p<0.1$",
"Field significance \\% of scores", "Both positive and negative \\% within random distribution", "Positive \\% outside random distribution."
),nrow=4,ncol=3, byrow=T))
kable(table,caption="Summary table of all tests on the two regions ", 'latex',
      booktabs =TRUE, escape=F,col.names = c("Test","Qld" ,"NSW VIC")) %>%
column_spec(2:3, width = "7cm") %>%
  kable_styling(latex_options = "scale_down")#%>%
  #kableExtra::landscape()
```

### 2019-03-14 Manuscripts/PeerJRainfall_Landuse/Rainfall_Landuse.Rmd
Update files for change to png
Upload since push does not work?

@@
 In this study we use a statistical approach to identify the relationship between removal of tree cover and rainfall with data from best available sources for two large areas in Australia.
-Gridded rainfall data between 1979 and 2015 was used for the areas, while large scale (exogenous) effects were represented by mean rainfall across a much larger area and climatic indicators, such as SOI and IOD.
+Gridded rainfall data between 1979 and 2015 was used for the areas, while large scale (exogenous) effects were represented by mean rainfall across a much larger area and climatic indicators, such as Southern Oscillation Index and Indian Ocean Dipole.
 Both general additive modelling and step trend tests were used for the analysis.
   For a region in south central Queensland, the reported change in tree clearing between 2002 - 2005 did not result in strong statistically significant precipitation changes.
-On the other hand, results from a bushfire affected region on the border of New SOuth Wales and Victoria suggests significant changes in the rainfall due to changes in tree cover.
+On the other hand, results from a bushfire affected region on the border of New South Wales and Victoria suggests significant changes in the rainfall due to changes in tree cover.
 This indicates the method works better when a abrupt change in the data can be clearly identified.
@@
 
-include_graphics("figures/map_selreg.jpg")
+include_graphics("figures/map_selreg.png")
 ```
@@
 Some locations in the southwest of the Australian Capital Territory (ACT) have also experienced intensive bushfires.", out.width="90%", echo=F}
-include_graphics("figures/bushfire_nswvic.jpg")
+include_graphics("figures/bushfire_nswvic.png")
 ```
@@
 ```{r figure3-tc-simple, fig.cap= "The expected evolution of the land surface after trees have been removed in (a) the QLD region and (b) the NSW/VIC region.", out.width="90%", echo=F}
-include_graphics("figures/tc_simple.jpg")
+include_graphics("figures/tc_simple.png")
 ```
@@
 
-jpeg("../../figures/bushfire_nswvic.jpg", width=960, height=960)
-  par(mar=c(3,3,1,6),xpd=T)
-  plot(e222,border="white")  
+png("figures/bushfire_nswvic.png", width=960, height=960)
+  par(mar=c(3,3,1,7),xpd=T)
+  plot(e222,border="white", xlab="",ylab="", cex.axis=2)  
   image(fire_loc,add=T, breaks=c(0,0.99,366.5,10000),col=c("white","red","white"))
@@
 #   text(1430000,-1000000,"study region",cex=1.2)
-  text(149.1, -35.3,"ACT",font=2,cex=1.2)
-  text(148.4, -36,"Kosciuszko",font=2,cex=0.9)#,srt=90)
-  text(147.5, -36.4,"VIC",font=2,cex=1.2)
-  text(148, -35.5,"NSW",font=2,cex=1.2)
-  image.plot(zlim=c(0,1),breaks=c(0,0.5,1),lab.breaks=c("","burned \narea",""),legend.only=T,horizontal=F,col=c("red","red"),legend.shrink=0.03,legend.width=1.3,legend.args=list(text="",line=0),lwd.poly=0.1)
-legend(149,-36.1,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=0.9)
+  text(149.1, -35.3,"ACT",font=2,cex=2)
+  text(148.4, -36,"Kosciuszko",font=2,cex=2)#,srt=90)
+  text(147.5, -36.4,"VIC",font=2,cex=2)
+  text(148, -35.5,"NSW",font=2,cex=2)
+  image.plot(zlim=c(0,1),breaks=c(0,0.5,1),lab.breaks=c("","burned \narea",""),axis.args=list(cex.axis=1.5),legend.only=T,horizontal=F,col=c("red","red"),legend.shrink=0.07,legend.width=1.0,lwd.poly=0.1, legend.mar=7.1)
+legend(148.7,-36.2,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=1.5)
  dev.off()
@@
 
-jpeg("../../figures/slats.jpg", width=1024, height = 760)
+png("figures/slats.png", width=1200, height = 900)
 Data %>% 
@@
 
-jpeg("../../figures/tc_figs.jpg", width=1280, height=640, quality=100)
+png("figures/tc_figs.png", width=1500, height=900)
 par(mfrow=c(1,2), mar=c(5,3,5,5))
@@
 
-include_graphics("figures/Rainfall_resid.jpg")
+include_graphics("figures/Rainfall_resid.png")
 ```
@@
 
-jpeg("../../figures/hist_rainfall.jpg")
+png("figures/hist_rainfall.png", width=900, height=900)
 rain %>%
@@
 
-jpeg("../../figures/autocorrelation.jpg")
+png("figures/autocorrelation.png", width =900, height=900)
 acf_df %>%
@@
 
-jpeg("../../figures/Rainfall_resid.jpg", width=960, height=960)
+png("figures/Rainfall_resid.png", width=960, height=960)
 plot_df %>%
@@
 
-jpeg("../../figures/cor_qld.jpg", width=960, height=960)
+png("figures/cor_qld.png", width=960, height=960)
 plot_df_qld %>%
@@
 
-jpeg("../../figures/cor_nswvic.jpg", width=960, height=960)
+png("figures/cor_nswvic.png", width=960, height=960)
 plot_df %>%
@@
 The correlation with NINO 3 is not shown as it is very similar to but weaker than for NINO 3.4.", out.width="90%", echo=F}
-include_graphics("figures/cor_qld.tif")
+include_graphics("figures/cor_qld.png")
 ```
@@
 ```{r cor-rain-nsw, fig.cap="Cross-correlation of six climate indices and rainfall in NSW/VIC study region.", out.width="90%", echo=F}
-include_graphics("figures/cor_nswvic.tif")
+include_graphics("figures/cor_nswvic.png")
 ```
@@
 Using a Kolmogorov-Smirnov test with shape = 1 and 2.4  for QLD and NSW/VIC respectively, rainfall in both regions can be modelled as a gamma distribution.", echo=F, out.width="90%"}
-include_graphics("figures/hist_rainfall.jpg")
+include_graphics("figures/hist_rainfall.png")
 ```
@@
 
-jpeg("../figures/tc_simple.jpg", width=720, height=480)
-par(mar=c(5,6,1,1), mfrow=c(1,2))
+png("figures/tc_simple.png", width=1200, height=900)
+par(mar=c(5,8,4,3), mfrow=c(1,2))
 plot(data1$year,data1$tc,type="l",ylim=c(0.5,2.5),xlab="",ylab="",xaxt="n",yaxt="n")
-axis(1,at=c(1,2,3),labels=c("before","LCC","after"),cex.axis=1.3)
-axis(2,at=c(1,2),label=c("removed","tree"),cex.axis=1.3,las=2)
-par(mar=c(5,6,1,1))
+axis(1,at=c(1,2,3),labels=c("before","LCC","after"),cex.axis=2, cex.lab=2)
+axis(2,at=c(1,2),label=c("removed","tree"),cex.axis=2,las=2, cex.lab=1.5)
+par(mar=c(5,8,4,3))
 plot(data2$year,data2$tc,type="l",ylim=c(0.5,2.5),xlab="",ylab="",xaxt="n",yaxt="n")
-axis(1,at=c(1,2,4),labels=c("before","LCC","after"),cex.axis=1.3)
-axis(2,at=c(1,2),label=c("removed","tree"),cex.axis=1.3,las=2)
+axis(1,at=c(1,2,4),labels=c("before","LCC","after"),cex.axis=2, cex.lab=2)
+axis(2,at=c(1,2),label=c("removed","tree"),cex.axis=2, cex.lab=2,las=1.5)
 dev.off()
@@
 
-jpeg("../../figures/hist_sign_LC.jpg")
+png("figures/hist_sign_LC.png", width=900, height = 900)
 par(mfrow=c(2,1))
@@
 # Now make the final plot for the paper with both sites and 2009 and all data
-  jpeg("../../figures/Cp_30yrs.jpg", width=1440, height=720)
+  png("figures/Cp_30yrs.png", width=1800, height=900)
   par(mfrow=c(1,2))
@@
 # produce also as jpeg
-jpeg("../../figures/pvalues_linear_fullmodel.jpg", height=960, width=960)
+png("figures/pvalues_linear_fullmodel.png", height=960, width=960)
 p_plot_df %>%
@@
 # produce also as jpg
-jpeg("../../figures/pvalues_fullmodel.jpg", height=960, width=960)
+png("figures/pvalues_fullmodel.png", height=960, width=960)
 p_plot_df %>%
@@
 
-jpeg("../../figures/rsq_fullmodel.jpg", height=480, width=480)
+png("figures/rsq_fullmodel.png", height=900, width=900)
 plot_df %>%
@@
   scale_fill_manual(values = c("blue","red")) +
-  xlab(expression(paste("% of variability explained (",r^2,")")))
+  xlab(expression(paste("% of variability explained (",r^2,")"))) +
+    theme(axis.text = element_text(size=rel(1.5)),
+        strip.text = element_text(size=rel(2)),
+        axis.title = element_text(size=rel(2)))
 dev.off()
@@
 ## plots
-jpeg("../../figures/GAMresidualTime.jpeg", height=960, width=960)
+png("figures/GAMresidualTime.png", height=960, width=960)
 residuals %>%
@@
 ## plot gam.check
-jpeg("../../figures/gam_check.jpg", width=720, height=960)
+png("figures/gam_check.png", width=960, height=960)
   par(mfrow=c(4,2))
-  gam.check(coef.rain30.qld4[[1]])
-  gam.check(coef.rain30.nswvic[[1]])
+  gam.check(coef.rain30.qld4[[1]], cex=1.5, lwd=2,cex.lab=2, cex.axis = 1.5)
+  gam.check(coef.rain30.nswvic[[1]], cex=1.5, lwd=2,cex.lab=2, cex.axis = 1.5)
 dev.off()
@@
 plot_df <- rbind(qld_df, nsw_df)
-jpeg("../../figures/pmodel_explHistogram.jpg")
+png("figures/pmodel_explHistogram.png", width=960, height=960)
 plot_df %>%
@@
 
-jpeg("../../figures/ResidualBoxplot.jpg")
+png("figures/ResidualBoxplot.png", width=960, height=960)
 resplot_df %>%
@@
   geom_boxplot(aes(group=year(Date))) +
-  facet_wrap(~State, ncol=1)
+  facet_wrap(~State, ncol=1) +
+    theme(axis.text = element_text(size=rel(1.5)),
+        strip.text = element_text(size=rel(2)),
+        axis.title = element_text(size=rel(2)))
 dev.off()
@@
 #par(mfrow=c(1,2))
-jpeg("figures/ResidualBoxplotchange.jpg")
+png("figures/ResidualBoxplotchange.png", width=960, height=960)
 resplot_df %>%
@@
   geom_boxplot(aes(group=Change)) + facet_wrap(~State, ncol=1) +
-  guides(fill=FALSE) + ggtitle("whole period")
+  guides(fill=FALSE) +
+  theme(axis.text = element_text(size=rel(1.5)),
+        strip.text = element_text(size=rel(2)),
+        axis.title = element_text(size=rel(2)))
 dev.off()
@@
 plot_df <- rbind(qld_df, nsw_df)
-jpeg("../../figures/soi_explHistogram.jpg", width=960, height=960)
+png("figures/soi_explHistogram.png", width=960, height=960)
 plot_df %>%
@@
     "$Z'=0$ and rainfall does not change"))
-kable(table2, caption="The interpretation of Z' score in the step trend test",
-      booktabs=T)
+kable(table2, 'latex',
+      caption="The interpretation of Z' score in the step trend test",
+      booktabs=T, escape=F)
 ```
@@
 
-jpeg("figures/step_new.jpg", width=1440, height=720, quality=100)
+png("figures/step_new.png", width=1800, height=900)
   par(mar=c(5,3,3,7), mfrow=c(1,2), xpd=T)
@@
 
-jpeg("figures/bt4_hist.jpg", width=740,height=740, quality = 100)
+png("figures/bt4_hist.png", width=1000,height=1000)
 # A histogram of taus
@@
 Green colour indicates an increase in tree cover, while red colour indicates a decrease in tree cover.", echo=F, out.width="90%"}
-include_graphics("figures/tc_figs.jpg")
+include_graphics("figures/tc_figs.png")
 ```
@@
 ```{r rsqmodel, fig.cap = "The distribution of the variance explained (adjusted $r^2$ ) by the full regression model", echo=F, out.width="90%"}
-include_graphics("figures/rsq_fullmodel.jpg")
+include_graphics("figures/rsq_fullmodel.png")
 
@@
 ```{r residuals, fig.cap ="The residual analysis of a sample pixel in the QLD region (top) and NSW/VIC region (bottom).", echo=F, out.width="90%"}
-include_graphics("figures/gam_check.jpg")
+include_graphics("figures/gam_check.png")
 ```
@@
 ```{r rsq, fig.cap = "The performance of the regression model if rainfall is only modelled by the climate drivers.
-It shows the percentage of rainfall variability that can be explained by the climate drivers for the Qld and NSW/VIC region", echo=F, out.width="70%", out.height="70%"}
-include_graphics("figures/soi_explhistogram.jpg")
+It shows the percentage of rainfall variability that can be explained by the climate drivers for the Qld and NSW/VIC region", echo=F, out.width="70%"}
+include_graphics("figures/soi_explhistogram.png")
 
@@
 The p value reported is for the land cover variable in the model.", echo=F, out.width="90%"}
-include_graphics("figures/Cp_30yrs.jpg")
+include_graphics("figures/Cp_30yrs.png")
 ```
@@
 The deeper the colour, the more significant the statistic.", echo = F, out.width="90%"}
-include_graphics("figures/step_new.jpg")
+include_graphics("figures/step_new.png")
 ```
@@
 ```{r meandiff, fig.cap ="Boxplots of annual rainfall residuals (estimated based on Equation 2 before and after the land cover intervention during 1979 - 2015 in the study regions.
-On average, the after period has a significantly lower annual rainfall residual in NSW/VIC, but a signifcantly higher annual rainfall residual in the Qld study area", echo=F, out.width="70%", out.height="70%"}
-
-include_graphics("figures/ResidualBoxplotchange.jpg")
+On average, the after period has a significantly lower annual rainfall residual in NSW/VIC, but a signifcantly higher annual rainfall residual in the Qld study area", echo=F, out.width="70%"}
+
+include_graphics("figures/ResidualBoxplotchange.png")
 ```
@@
 
-```{r fieldsig, fig.cap ="Field significance results showing the distribution of percentage of significant $Z'$ scores for the resampled series (blue bars), and the precentage significant $Z'$ scores for the original series of rainfall years (red circle)", echo=F, out.width="70%"}
-include_graphics("figures/bt4_hist.jpg")
+```{r fieldsig, fig.cap ="Field significance results showing the distribution of percentage of significant Z' scores for the resampled series (blue bars), and the precentage significant Z' scores for the original series of rainfall years (red circle)", echo=F, out.width="70%"}
+include_graphics("figures/bt4_hist.png")
 ```
@@
 The data were obtained from SLATS (2017), and clearly indicate a sharp decrease in the clearing rate after 2005.", echo=F, out.width="90%"}
-include_graphics("figures/slats.jpg")
+include_graphics("figures/slats.png")
 ```

### 2019-06-17 Manuscripts/PeerJRainfall_Landuse/ReplyToReviewers.Rmd
changed files after review and resubmitted PeerJ

---
title: Detecting the impact of land cover change on observed rainfall.
preprint: false
author:
  - name: Chun X. Liang
    affiliation: 1
  - name: Floris F. van Ogtrop
    affiliation: 1
  - name: R. Willem Vervoort
    affiliation: 1
    corresponding: true
    email: willem.vervoort@sydney.edu.au
affiliation:
  - code: 1
    address: Sydney Institute of Agriculture, The University of Sydney, NSW 2006
abstract: >
 This is the detailed reply to reviewers for the paper" Detecting the impact of land cover change on observed rainfall." Peerj article number 35847 
header-includes: >
 \usepackage{lipsum}
 \usepackage{textcomp, rotating}
bibliography: reflist_chap3.bib
date: "`r Sys.Date()`"
output:
  bookdown::pdf_book:
    base_format: rticles::peerj_article # for using bookdown features like \@ref()
  rticles::peerj_article: default
---
```{r, echo=F}
knitr::opts_knit$set(root.dir =
       #"E:/cloudstor/chun/chun/manuscripts/PeerJRainfall_Landuse")
      "C:/users/rver4657/owncloud/chun/chun/manuscripts/PeerJRainfall_Landuse")
```

# Response to reviewers

## Reviewer 1 (John Boland) 
Comments for the Author 
I think the authors should continue to investigate this topic as one of the case studies - the Queensland one - proved probably inconclusive. They should also try and see if they can find instances of increase of land cover and increased rainfall.

**reply from the authors**
We would love to expand this study with more case studies, but we have come to realise that this is not that easy. This study also shows this, with the Qeensland location being inconclusive, despite reported significant land use change. There are several requirements for a good study area:

1. It needs to be large enough to capture the effect of landuse on rainfall within the area;  
2. It needs to have a drastic enough landuse change to be observable above the rainfall variation;  
3. It needs to have a long enough climate data time series both prior and post landuse change.

Such locations with land use changes are not easy to find. The obvious locations have always been Amazonia and sub-Saharan Africa, where also most of the modelling studies have taken place. The problem with the sub-Saharan location is that the land use change is quite long ago and local data difficult to obtain. The problem with the Amazonian location is that this area is well-known for the feedback and we felt that we could not add much to this work. The recent bushfires in California and Colorado could potentially be a good location, but at the moment there is insufficient "post event data". We have added some additional sentences at the end of the discussion (line 485ff)

Minor points:

Line 50 - problem with El Nin ~o

**reply** fixed by using UTF-8 encoding

Line 97 - problem with Koppen

**reply** fixed by using UTF-8 encoding

Queensland study area - did not have much tree cover initially.

**reply** Indeed that is one of the problems with this location.

Line 216 - I would have thought Spearman would be more appropriate

**reply** We are working with continuous data which has been centered, detrended and deaseasonalised and is therefore normally distributed and therefore Pearson's correlation is appropriate. However, in the analysis we simply used the cross correlation analysis (which is based on Pearson's). We have rewritten this part of the paper to explain that we simply focus on the cross correlation at lag 0.:
"The cross-correlations between the deseasonalised and detrended rainfall and the climatic indices were tested, and the strongests indicators at lag 0 were identified for the model."

Line 306 - below??

**reply** Agree that was a mistake, it was supposed to refer to Table 1. We have fixed this. Table 1 simply shows an example calculation.

Lines 360-361 - something wrong with figure reference

**reply** Indeed, error in the coding. It is supposed to refer to Figure 9

## Reviewer 2 (Anonymous)

Basic reporting 
Line 34, maximum temperature at what height? surface temperature?

**reply** It was modelled surface temperature [p2, L22711, @McAlpine2007].

Line 39, "in both space and time", omit "in"

**reply** We think "in" is appropriate, we are talking about variation in both space and time.

Line 41, better to add more details/examples about "those complex set of interactions"

**reply** We added "outlined below" as this sentence relates to the text in the next paragraph.

Line 48, "On the longer time scale", compare to daily which is not mentioned in previous sentence. This paragraph is comparing local vs. larger scale, not on the time scale

**reply** We agree, but it is in both time and space. We added into line 43: "Locally, and on a shorter, daily time scale, there are two main sources that generate rainfall:"

Line 54, should this topic sentence and this paragraph not restricted to "predict Australian rainfall"? but rainfall at any locations?

**reply** The sentence and comment about "Australian rainfall" relates to the previous paragraph and to the Westra and Sharma paper. We have reworded the subsequent sentence: "...annual precipitation variance. More generally, some of..."

Line 90, either use "the largest" or "the worst", they mean the same thing

**reply** accepted

Figure 3, no (a) and (b) were labeled in the graph, or use (left and right).

**reply** accepted

Table 1, extra space between row "2002: and row "2005"

**reply** This is automatically inserted in the article via the Latex and PeerJ article template as part of the knitr package as there is a jump in the years. It is to indicate the gap left by leaving out the years 2003 and 2004.

Experimental design 
Line 92, "..due to the drought and cold condition", the reference for that scientific conclusion is from News, a reference from a scientific journal would be more appropriate

**reply** We agree that this would be preferavle, however, there is no scientific journal paper to help us out here. We have changed the word "reported" into "suggested" to de-emphasise the importance of the finding.

Line 113, please include scientific name (Latin) for those tree species.

**reply** accepted included *Eucalyptus pauciflora*, *Eucalyptus dalrympleana* and *Eucalyptus delegatensis*

Line 115, previous sentence authors mentioned about great height, so here better to add details of height. "...20 years to mature (~? m)"

**reply** accepted and added a reference [@buckley2012].

Line 115-116, changing subjects. "Land clearing" is the subject, but not for the rest of the sentence.

**reply** Agreed We changed the sentence to read: "This region is vulnerable to fires and drought, however land clearing is not a major issue."

Line 125, add more details, t test for the averaged precipitation before (1979-2003) and after (2004-2015) the disturbance? Or move this sentence to the section of "statistical analysis" in below

**reply** Agreed, sentence removed, as it is repeated in the "statistical analysis".

Line 208, ", this variable was included" what variable? subject changes within a sentence, rewrite

**reply**Correct, the actual sentence is an error from an earlier version. The new text in the section now reads: "A further complicating factor is the influence of the "millenium drought" over the study period and in particular the change to wet conditions in 2010 - 2011 [@vanDijk2013]. Therefore, the spatially averaged monthly rainfall in the Murray Darling Basin (MDB, downloaded from [the Bureau of Meteorology](http://www.bom.gov.au/web01/ncc/www/cli_chg/timeseries/rain/allmonths/mdb/latest.txt)) was used to explain the year-on-year variation in the rainfall in the regions. Since both regions at least partly overlap with the MDB, the average rainfall for the entire basin was assumed to be a useful explaining variable."

Line 225-231, those paragraphs are results, do authors need them to be here for explaining a further step analysis? or should be moved down to Result section

**reply** We agree that these figures are a bit much at this point and we have moved this to the "supplementary data" section.

Line 236-246, those two paragraphs sound like a discussion? should it be moved down to Discussion section?

**reply** Not really, in these two paragraphs we argue why we should include seasonal and long term trends in the model. Following @serinaldi2018, we believe that it is important to base statistical trend detection on a clear theory about all the trends included in the model. We therefore believe that these two paragraphs should stay in the methodology section of the paper. We have however shortened the text to:
"Rainfall in Australia shows strong seasonal patterns [@Holper2011; @ABS2012]. As a result a seasonal component of rainfall has a periodic pattern which should be included in the model. In addition, long term trends in the regional rainfall in some parts of Australia are significant [@Hughes2003; @Gallant2007; @Chowdhury2010]. The presence of long term trends can be confused with the outcome of a step change in rainfall. As a result a linear trend term was implemented in the model to remove any long term effects."

Line 287-301, should these two paragraphs be combined with each individual analysis below? I don't see any reason to give a summary first and then introducing each analysis

**reply** Agreed, on reflection, it is better to integrate these paragraphs and to drop the subheadings in the section.

Validity of the findings  
Line 363-367, those are not key findings and should be moved to Method section.

**reply** Agreed, we removed the section from the results, but also did not insert in the methods, because we also agree that this is more valid in supplementary material. However, as we supply the full script of the analysis with the paper and the data via the guithub repository we believe that this is not relevant to include.

Line 377-382, except the first sentence, the rest should be moved down to the discussion section

**reply** Agreed, however, we wanted to highlight that the long term trend was maintained. The part was modified to: "There were generally no statistically significant long term trends in both regions. However, the trend term was kept in the regression model to ensure the detection of step change was not due to a possible long term trend (even if this was not significant)."

Line 386, "..compared to the number in QLD region"

*reply** agreed

Line 394, put the key findings as the subheading text instead of "step trend test" would be better

**reply** agreed subheading text changed to: "Stronger step trend in NSW/VIC compared to QLD"

Line 395-398, those sentences are illustrating how to read the figure, should be put in the figure caption, not here in the Result section

**reply** agreed, text removed and added to the figure caption.

Line 400, what is a "reasonable relationship"? authors should use more statistically sound words to describe the results.

**reply** agreed, we have used the word qualitatively to emphasise that this is simply a visual comparison. The new text now reads:"Qualitatively the locations where changes in tree cover occurred in the NSW/VIC area (Figure 5) seem to agree with the patterns in the Z' scores. However, a direct relationship would not necesssarily be expected as movement of air masses could mean that actual changes of rainfall are observed close by, but not necessarily exactly at areas with changes of landcover."

Line 404, a difference of what?

**reply** agreed, the sentence was rewritten: "For the QLD region (left panel) there are only a few significant Z' scores which matches the earlier significance in the landcover variable in the full regresion."

Line 411, if authors knew it was difficult to see, why not regraph the data to make it easy to see?

**reply** It is difficult to make the figure more clear as there are a lot of data and there is a wide spread in the distribution. As a result, we dropped the figure from the paper and only described the statistical analysis in the text. The new text now reads: "The rainfall regression residuals without the landcover variable for the two regions (before-change since 1979 and after-change) were also compared using a simple t-test. The mean rainfall residuals were significantly different ($p < 0.05$). between the "before" and "after" periods in both regions."

Line 421-429, All those sentences read like Method to me

**reply** Agreed, moved to the methods section.

Line 439 and 440, "strongest evidence" "weak evidence"? could it be more statistically described? like p value or with how much variance explained?

**reply** Not really, these are summary statements related to the overview presented in the summary Table 3. However, to make this link clearer, we have inserted references to this table in the text and also reworded the text slightly.

Line 476, the discussion point of "recovery" is worth to expand. Not only the evolution model from Figure 3 which only illustrates the mass/area, the species composition might be changed too. And ties into the evapotranspiration rate.

**reply** Agreed, we added a sentence to address this issue: "Not only does this refer to a change in the total biomass, but this could also include a change in the species composition as a result of the disturbance."

Line 526, "Possible future work could focus on a non-drought period.." I think research on lands with disturbance(e.g. fire, drought, etc) is also important even though it is hard to study as reported from this study.

**reply** We are not sure about this comment. The study specifically covered a fire affected area (NSW/VIC) and a following drought period. The problem was that here those two effects were combined, so we needed to disentangle the effects. Our suggestion for a future study is to look at a disturbance where the following period is not a drought, but a wet period, as we might be able to see a greater effect on the rainfall. The hypothesis in that case would be that if there is increased moisture in the atmosphere and rootzone, this might result in increased feedback.

Comments for the Author 
This paper uses sophisticated statistical approaches to identify impacts of changes in land cover on rainfall from 1979 to 2015 in two regions in Australia. Results show lands under historical bushfire had significant changes in precipitation, but not for lands with tree clearing. I am sure this paper adds an important value to the community. However, this paper has a very detailed Method section which includes some results and discussion. Some part of the Method section also includes a summary test first and then introducing each analysis. This would cause repetitive sentences. For example, line 303 "As indicated, the step trend test is a modified version of the Mann-Whitney U statistic (Hirsch and Gilroy 1985).", this sentence is repeating line 289 in the summary. I think the summary is not necessary and should be merged to each analysis.
I suggested authors to reorganize the whole paper and keep the sections simple and clear (not overwritten with details not necessary for readers to understand the analysis/paper). Or if authors think those sentences in the Method section are necessary, put them in the appendix, so the main body of this paper can be simplified. I understand authors trying to describe all their trails, thoughts and steps in analyzing those data in this paper. This is my personal opinion. I will let the editor decide the format of this paper.

**reply** Thank you for your detailed comments which helped improve the paper. The paper is indeed quite "heavy" on the methods, but this is because we wanted to use several methods to make sure we were identifying an effect and not just a "chance" occurrence. We agree with most of your comments and as indicated have made several changes to hopefully improve the flow of the paper. In addition, based on your comments we removed three of the figures: the rainfall histogram figure (figure 7 in the reviewed manuscript), the gam resiudal analysis (Figure 10 in the reviewed paper) and removed the rainfall residual boxplots (figure 14 in the reviewed paper). In addition, the correlation plots with the climate indices were moved to the supplementary data including the table of data sources and the appendix was removed.

## Reviewer 3 (Hongkai Gao)

Basic reporting 
No comments 
Experimental design 
No comments 
Validity of the findings 
No comments 
Comments for the Author 
In this paper, Vervoort and his colleagues reported their finding about the impact of land cover change on observed rainfall. The analysis is very comprehensive. And I learned a lot from this manuscript. The authors used statistic approach to analyze observed data, to detect the change of precipitation after forest cleaning and bushfires. They found different influences after two changes. The methods are quite reliable, and conclusions can be trusted. I think this paper has great potential to be a high-cited paper. But the paper still needs to be further improved before publication.

1. Not only the statistically analysis, but also the physically meaning should be further discussed. Theoretically, the change of vegetation will impact on ET, especially in dry seasons, because land cover change will impact on the root zone storage capacity, which provides an important buffer to increase resilience to drought (Gao et al., 2014). This should impact on downwind precipitation.

2. Several important papers, on the impact of land cover on precipitation, are still missed in the reference list (e.g. Keys et al., 20142).

Gao, H. , Hrachowitz, M. , Schymanski, S. J. , Fenicia, F. , Sriwongsitanon, N. , & Savenije, H. H. G. . (2014). Climate controls how ecosystems size the root zone storage capacity at catchment scale. Geophysical Research Letters, 41(22), 7916-7923.
Keys, P. W. , Van, d. E. R. J. , Gordon, L. J. , Hoff, H. , Nikoli, R. , & Savenije, H. H. G. . (2012). Analyzing precipitationsheds to understand the vulnerability of rainfall dependent regions. Biogeosciences, 9(2), 733-746.

**reply** Thank you for the suggested papers which we have read with interest. The Keys et al (2012) paper is an interesting modelling study that builds on earlier modelling studies and globally focussed. In relation to our current study, this paper is not as relevant. While we mention in our discussion that there will be a spatial distance between the source of moisture from evaporation and the rainfall. For individual storms we can do moisture backtracking, but for the monthly data we used this is difficult unless we do extensive modelling based on a whole new set of assumptions. In our study we have focussed on data and the analysis of actual data rather than using a modelling approach. We tried to choose our regions large enough to be able to capture the spatial distance, but as indicated this is not quantified (as this cannot be simply quantified) from the data.

We also read the Gao et al. (2014) paper with interest. This paper deals with simulation of the rootzone depth and suggests that rootzone storage capacity is dynamically adapted ecologically to rainfall availability and other climate factors.  
We are unsure how this relates to our current research which deals with the impact of landuse on local rainfall. We understand that a change in landuse impacts the storage and this might impact ET. However, while this might be a further explanation of the decrease in rainfall due to landuse change, it cannot be directly derived from the rainfall data without further modelling.

# References

### 2019-06-17 Manuscripts/PeerJRainfall_Landuse/ReplyToReviewers.tex
changed files after review and resubmitted PeerJ

%% Submissions for peer-review must enable line-numbering
%% using the lineno option in the \documentclass command.
%%
%% Preprints and camera-ready submissions do not need
%% line numbers, and should have this option removed.
%%
%% Please note that the line numbering option requires
%% version 1.1 or newer of the wlpeerj.cls file, and
%% the corresponding author info requires v1.2

\documentclass[fleqn,10pt,lineno]{wlpeerj} % for journal submissions

% ZNK -- Adding headers for pandoc

\setlength{\emergencystretch}{3em}
\providecommand{\tightlist}{
\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage{lipsum}
\usepackage[unicode=true]{hyperref}
\usepackage{longtable}

\usepackage{lipsum} \usepackage{textcomp, rotating}

\title{Detecting the impact of land cover change on observed rainfall.}

\author[1]{Chun X. Liang}

\author[1]{Floris F. van Ogtrop}

\author[1]{R. Willem Vervoort}

\corrauthor[1]{R. Willem Vervoort}{\href{mailto:willem.vervoort@sydney.edu.au}{\nolinkurl{willem.vervoort@sydney.edu.au}}}

\affil[1]{Sydney Institute of Agriculture, The University of Sydney, NSW 2006}

%
% \author[1]{First Author}
% \author[2]{Second Author}
% \affil[1]{Address of first author}
% \affil[2]{Address of second author}
% \corrauthor[1]{First Author}{f.author@email.com}

%

\begin{abstract}
This is the detailed reply to reviewers for the paper" Detecting the
impact of land cover change on observed rainfall." Peerj article number
35847
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}

\flushbottom
\maketitle
\thispagestyle{empty}

\section{Response to reviewers}\label{response-to-reviewers}

\subsection{Reviewer 1 (John Boland)}\label{reviewer-1-john-boland}

Comments for the Author I think the authors should continue to
investigate this topic as one of the case studies - the Queensland one -
proved probably inconclusive. They should also try and see if they can
find instances of increase of land cover and increased rainfall.

\textbf{reply from the authors} We would love to expand this study with
more case studies, but we have come to realise that this is not that
easy. This study also shows this, with the Qeensland location being
inconclusive, despite reported significant land use change. There are
several requirements for a good study area:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  It needs to be large enough to capture the effect of landuse on
  rainfall within the area;\\
\item
  It needs to have a drastic enough landuse change to be observable
  above the rainfall variation;\\
\item
  It needs to have a long enough climate data time series both prior and
  post landuse change.
\end{enumerate}

Such locations with land use changes are not easy to find. The obvious
locations have always been Amazonia and sub-Saharan Africa, where also
most of the modelling studies have taken place. The problem with the
sub-Saharan location is that the land use change is quite long ago and
local data difficult to obtain. The problem with the Amazonian location
is that this area is well-known for the feedback and we felt that we
could not add much to this work. The recent bushfires in California and
Colorado could potentially be a good location, but at the moment there
is insufficient ``post event data''. We have added some additional
sentences at the end of the discussion (line 485ff)

Minor points:

Line 50 - problem with El Nin \textasciitilde{}o

\textbf{reply} fixed by using UTF-8 encoding

Line 97 - problem with Koppen

\textbf{reply} fixed by using UTF-8 encoding

Queensland study area - did not have much tree cover initially.

\textbf{reply} Indeed that is one of the problems with this location.

Line 216 - I would have thought Spearman would be more appropriate

\textbf{reply} We are working with continuous data which has been
centered, detrended and deaseasonalised and is therefore normally
distributed and therefore Pearson's correlation is appropriate. However,
in the analysis we simply used the cross correlation analysis (which is
based on Pearson's). We have rewritten this part of the paper to explain
that we simply focus on the cross correlation at lag 0.: ``The
cross-correlations between the deseasonalised and detrended rainfall and
the climatic indices were tested, and the strongests indicators at lag 0
were identified for the model.''

Line 306 - below??

\textbf{reply} Agree that was a mistake, it was supposed to refer to
Table 1. We have fixed this. Table 1 simply shows an example
calculation.

Lines 360-361 - something wrong with figure reference

\textbf{reply} Indeed, error in the coding. It is supposed to refer to
Figure 9

\subsection{Reviewer 2 (Anonymous)}\label{reviewer-2-anonymous}

Basic reporting Line 34, maximum temperature at what height? surface
temperature?

\textbf{reply} It was modelled surface temperature (p2, L22711, McAlpine
et al. 2007).

Line 39, ``in both space and time'', omit ``in''

\textbf{reply} We think ``in'' is appropriate, we are talking about
variation in both space and time.

Line 41, better to add more details/examples about ``those complex set
of interactions''

\textbf{reply} We added ``outlined below'' as this sentence relates to
the text in the next paragraph.

Line 48, ``On the longer time scale'', compare to daily which is not
mentioned in previous sentence. This paragraph is comparing local
vs.~larger scale, not on the time scale

\textbf{reply} We agree, but it is in both time and space. We added into
line 43: ``Locally, and on a shorter, daily time scale, there are two
main sources that generate rainfall:''

Line 54, should this topic sentence and this paragraph not restricted to
``predict Australian rainfall''? but rainfall at any locations?

\textbf{reply} The sentence and comment about ``Australian rainfall''
relates to the previous paragraph and to the Westra and Sharma paper. We
have reworded the subsequent sentence: ``\ldots{}annual precipitation
variance. More generally, some of\ldots{}''

Line 90, either use ``the largest'' or ``the worst'', they mean the same
thing

\textbf{reply} accepted

Figure 3, no (a) and (b) were labeled in the graph, or use (left and
right).

\textbf{reply} accepted

Table 1, extra space between row ``2002: and row''2005"

\textbf{reply} This is automatically inserted in the article via the
Latex and PeerJ article template as part of the knitr package as there
is a jump in the years. It is to indicate the gap left by leaving out
the years 2003 and 2004.

Experimental design Line 92, ``..due to the drought and cold
condition'', the reference for that scientific conclusion is from News,
a reference from a scientific journal would be more appropriate

\textbf{reply} We agree that this would be preferavle, however, there is
no scientific journal paper to help us out here. We have changed the
word ``reported'' into ``suggested'' to de-emphasise the importance of
the finding.

Line 113, please include scientific name (Latin) for those tree species.

\textbf{reply} accepted included \emph{Eucalyptus pauciflora},
\emph{Eucalyptus dalrympleana} and \emph{Eucalyptus delegatensis}

Line 115, previous sentence authors mentioned about great height, so
here better to add details of height. ``\ldots{}20 years to mature
(\textasciitilde{}? m)''

\textbf{reply} accepted and added a reference (Buckley et al. 2012).

Line 115-116, changing subjects. ``Land clearing'' is the subject, but
not for the rest of the sentence.

\textbf{reply} Agreed We changed the sentence to read: ``This region is
vulnerable to fires and drought, however land clearing is not a major
issue.''

Line 125, add more details, t test for the averaged precipitation before
(1979-2003) and after (2004-2015) the disturbance? Or move this sentence
to the section of ``statistical analysis'' in below

\textbf{reply} Agreed, sentence removed, as it is repeated in the
``statistical analysis''.

Line 208, ``, this variable was included'' what variable? subject
changes within a sentence, rewrite

\textbf{reply}Correct, the actual sentence is an error from an earlier
version. The new text in the section now reads: ``A further complicating
factor is the influence of the''millenium drought" over the study period
and in particular the change to wet conditions in 2010 - 2011 (Dijk and
Viney 2013). Therefore, the spatially averaged monthly rainfall in the
Murray Darling Basin (MDB, downloaded from
\href{http://www.bom.gov.au/web01/ncc/www/cli_chg/timeseries/rain/allmonths/mdb/latest.txt}{the
Bureau of Meteorology}) was used to explain the year-on-year variation
in the rainfall in the regions. Since both regions at least partly
overlap with the MDB, the average rainfall for the entire basin was
assumed to be a useful explaining variable."

Line 225-231, those paragraphs are results, do authors need them to be
here for explaining a further step analysis? or should be moved down to
Result section

\textbf{reply} We agree that these figures are a bit much at this point
and we have moved this to the ``supplementary data'' section.

Line 236-246, those two paragraphs sound like a discussion? should it be
moved down to Discussion section?

\textbf{reply} Not really, in these two paragraphs we argue why we
should include seasonal and long term trends in the model. Following
Serinaldi, Kilsby, and Lombardo (2018), we believe that it is important
to base statistical trend detection on a clear theory about all the
trends included in the model. We therefore believe that these two
paragraphs should stay in the methodology section of the paper. We have
however shortened the text to: ``Rainfall in Australia shows strong
seasonal patterns (Holper 2011; Australian Bureau of Statistics 2012).
As a result a seasonal component of rainfall has a periodic pattern
which should be included in the model. In addition, long term trends in
the regional rainfall in some parts of Australia are significant (Hughes
2003; Gallant, Hennessy, and Risbey 2007; Chowdhury and Beecham 2010).
The presence of long term trends can be confused with the outcome of a
step change in rainfall. As a result a linear trend term was implemented
in the model to remove any long term effects.''

Line 287-301, should these two paragraphs be combined with each
individual analysis below? I don't see any reason to give a summary
first and then introducing each analysis

\textbf{reply} Agreed, on reflection, it is better to integrate these
paragraphs and to drop the subheadings in the section.

Validity of the findings\\
Line 363-367, those are not key findings and should be moved to Method
section.

\textbf{reply} Agreed, we removed the section from the results, but also
did not insert in the methods, because we also agree that this is more
valid in supplementary material. However, as we supply the full script
of the analysis with the paper and the data via the guithub repository
we believe that this is not relevant to include.

Line 377-382, except the first sentence, the rest should be moved down
to the discussion section

\textbf{reply} Agreed, however, we wanted to highlight that the long
term trend was maintained. The part was modified to: ``There were
generally no statistically significant long term trends in both regions.
However, the trend term was kept in the regression model to ensure the
detection of step change was not due to a possible long term trend (even
if this was not significant).''

Line 386, ``..compared to the number in QLD region''

*reply** agreed

Line 394, put the key findings as the subheading text instead of ``step
trend test'' would be better

\textbf{reply} agreed subheading text changed to: ``Stronger step trend
in NSW/VIC compared to QLD''

Line 395-398, those sentences are illustrating how to read the figure,
should be put in the figure caption, not here in the Result section

\textbf{reply} agreed, text removed and added to the figure caption.

Line 400, what is a ``reasonable relationship''? authors should use more
statistically sound words to describe the results.

\textbf{reply} agreed, we have used the word qualitatively to emphasise
that this is simply a visual comparison. The new text now
reads:``Qualitatively the locations where changes in tree cover occurred
in the NSW/VIC area (Figure 5) seem to agree with the patterns in the Z'
scores. However, a direct relationship would not necesssarily be
expected as movement of air masses could mean that actual changes of
rainfall are observed close by, but not necessarily exactly at areas
with changes of landcover.''

Line 404, a difference of what?

\textbf{reply} agreed, the sentence was rewritten: ``For the QLD region
(left panel) there are only a few significant Z' scores which matches
the earlier significance in the landcover variable in the full
regresion.''

Line 411, if authors knew it was difficult to see, why not regraph the
data to make it easy to see?

\textbf{reply} It is difficult to make the figure more clear as there
are a lot of data and there is a wide spread in the distribution. As a
result, we dropped the figure from the paper and only described the
statistical analysis in the text. The new text now reads: ``The rainfall
regression residuals without the landcover variable for the two regions
(before-change since 1979 and after-change) were also compared using a
simple t-test. The mean rainfall residuals were significantly different
(\(p < 0.05\)). between the''before" and ``after'' periods in both
regions."

Line 421-429, All those sentences read like Method to me

\textbf{reply} Agreed, moved to the methods section.

Line 439 and 440, ``strongest evidence'' ``weak evidence''? could it be
more statistically described? like p value or with how much variance
explained?

\textbf{reply} Not really, these are summary statements related to the
overview presented in the summary Table 3. However, to make this link
clearer, we have inserted references to this table in the text and also
reworded the text slightly.

Line 476, the discussion point of ``recovery'' is worth to expand. Not
only the evolution model from Figure 3 which only illustrates the
mass/area, the species composition might be changed too. And ties into
the evapotranspiration rate.

\textbf{reply} Agreed, we added a sentence to address this issue: ``Not
only does this refer to a change in the total biomass, but this could
also include a change in the species composition as a result of the
disturbance.''

Line 526, ``Possible future work could focus on a non-drought period..''
I think research on lands with disturbance(e.g.~fire, drought, etc) is
also important even though it is hard to study as reported from this
study.

\textbf{reply} We are not sure about this comment. The study
specifically covered a fire affected area (NSW/VIC) and a following
drought period. The problem was that here those two effects were
combined, so we needed to disentangle the effects. Our suggestion for a
future study is to look at a disturbance where the following period is
not a drought, but a wet period, as we might be able to see a greater
effect on the rainfall. The hypothesis in that case would be that if
there is increased moisture in the atmosphere and rootzone, this might
result in increased feedback.

Comments for the Author This paper uses sophisticated statistical
approaches to identify impacts of changes in land cover on rainfall from
1979 to 2015 in two regions in Australia. Results show lands under
historical bushfire had significant changes in precipitation, but not
for lands with tree clearing. I am sure this paper adds an important
value to the community. However, this paper has a very detailed Method
section which includes some results and discussion. Some part of the
Method section also includes a summary test first and then introducing
each analysis. This would cause repetitive sentences. For example, line
303 ``As indicated, the step trend test is a modified version of the
Mann-Whitney U statistic (Hirsch and Gilroy 1985).'', this sentence is
repeating line 289 in the summary. I think the summary is not necessary
and should be merged to each analysis. I suggested authors to reorganize
the whole paper and keep the sections simple and clear (not overwritten
with details not necessary for readers to understand the
analysis/paper). Or if authors think those sentences in the Method
section are necessary, put them in the appendix, so the main body of
this paper can be simplified. I understand authors trying to describe
all their trails, thoughts and steps in analyzing those data in this
paper. This is my personal opinion. I will let the editor decide the
format of this paper.

\textbf{reply} Thank you for your detailed comments which helped improve
the paper. The paper is indeed quite ``heavy'' on the methods, but this
is because we wanted to use several methods to make sure we were
identifying an effect and not just a ``chance'' occurrence. We agree
with most of your comments and as indicated have made several changes to
hopefully improve the flow of the paper. In addition, based on your
comments we removed three of the figures: the rainfall histogram figure
(figure 7 in the reviewed manuscript), the gam resiudal analysis (Figure
10 in the reviewed paper) and removed the rainfall residual boxplots
(figure 14 in the reviewed paper). In addition, the correlation plots
with the climate indices were moved to the supplementary data including
the table of data sources and the appendix was removed.

\subsection{Reviewer 3 (Hongkai Gao)}\label{reviewer-3-hongkai-gao}

Basic reporting No comments Experimental design No comments Validity of
the findings No comments Comments for the Author In this paper, Vervoort
and his colleagues reported their finding about the impact of land cover
change on observed rainfall. The analysis is very comprehensive. And I
learned a lot from this manuscript. The authors used statistic approach
to analyze observed data, to detect the change of precipitation after
forest cleaning and bushfires. They found different influences after two
changes. The methods are quite reliable, and conclusions can be trusted.
I think this paper has great potential to be a high-cited paper. But the
paper still needs to be further improved before publication.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Not only the statistically analysis, but also the physically meaning
  should be further discussed. Theoretically, the change of vegetation
  will impact on ET, especially in dry seasons, because land cover
  change will impact on the root zone storage capacity, which provides
  an important buffer to increase resilience to drought (Gao et al.,
  2014). This should impact on downwind precipitation.
\item
  Several important papers, on the impact of land cover on
  precipitation, are still missed in the reference list (e.g.~Keys et
  al., 20142).
\end{enumerate}

Gao, H. , Hrachowitz, M. , Schymanski, S. J. , Fenicia, F. ,
Sriwongsitanon, N. , \& Savenije, H. H. G. . (2014). Climate controls
how ecosystems size the root zone storage capacity at catchment scale.
Geophysical Research Letters, 41(22), 7916-7923. Keys, P. W. , Van, d.
E. R. J. , Gordon, L. J. , Hoff, H. , Nikoli, R. , \& Savenije, H. H. G.
. (2012). Analyzing precipitationsheds to understand the vulnerability
of rainfall dependent regions. Biogeosciences, 9(2), 733-746.

\textbf{reply} Thank you for the suggested papers which we have read
with interest. The Keys et al (2012) paper is an interesting modelling
study that builds on earlier modelling studies and globally focussed. In
relation to our current study, this paper is not as relevant. While we
mention in our discussion that there will be a spatial distance between
the source of moisture from evaporation and the rainfall. For individual
storms we can do moisture backtracking, but for the monthly data we used
this is difficult unless we do extensive modelling based on a whole new
set of assumptions. In our study we have focussed on data and the
analysis of actual data rather than using a modelling approach. We tried
to choose our regions large enough to be able to capture the spatial
distance, but as indicated this is not quantified (as this cannot be
simply quantified) from the data.

We also read the Gao et al. (2014) paper with interest. This paper deals
with simulation of the rootzone depth and suggests that rootzone storage
capacity is dynamically adapted ecologically to rainfall availability
and other climate factors.\\
We are unsure how this relates to our current research which deals with
the impact of landuse on local rainfall. We understand that a change in
landuse impacts the storage and this might impact ET. However, while
this might be a further explanation of the decrease in rainfall due to
landuse change, it cannot be directly derived from the rainfall data
without further modelling.

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\hypertarget{ref-ABS2012}{}
Australian Bureau of Statistics. 2012. ``Geography and Climate -
Australia's Climate.''

\hypertarget{ref-buckley2012}{}
Buckley, ThomasN, TarrynL Turnbull, Sebastian Pfautsch, Mana Gharun, and
MarkA Adams. 2012. ``Differences in Water Use Between Mature and
Post-Fire Regrowth Stands of Subalpine Eucalyptus Delegatensis R.
Baker.'' Journal Article. \emph{Forest Ecology and Management} 270:
1--10.
doi:\href{https://doi.org/10.1016/j.foreco.2012.01.008}{10.1016/j.foreco.2012.01.008}.

\hypertarget{ref-Chowdhury2010}{}
Chowdhury, R. K., and S. Beecham. 2010. ``Australian Rainfall Trends and
Their Relation to the Southern Oscillation Index.'' \emph{Hydrological
Processes} 24 (4). John Wiley \& Sons, Ltd.: 504--14.
\url{http://dx.doi.org/10.1002/hyp.7504}.

\hypertarget{ref-vanDijk2013}{}
Dijk, Beck van, Albert I. J. M., and Neil R. Viney. 2013. ``The
Millennium Drought in Southeast Australia (2001--2009): Natural and
Human Causes and Implications for Water Resources, Ecosystems, Economy,
and Society.'' \emph{Water Resources Research} 49 (2): 1040--57.
doi:\href{https://doi.org/10.1002/wrcr.20123}{10.1002/wrcr.20123}.

\hypertarget{ref-Gallant2007}{}
Gallant, Ailie J. E., Kevin J. Hennessy, and James Risbey. 2007.
``Trends in Rainfall Indices for Six Australian Regions: 1910 - 2005.''
\emph{Australian Meteorological Magazine} 56: 223--39.

\hypertarget{ref-Holper2011}{}
Holper, Paul N. 2011. \emph{Climate Change, Science Information Paper:
Australian Rainfall-Past, Present and Future}. Canberra: CSIRO.

\hypertarget{ref-Hughes2003}{}
Hughes, Lesley. 2003. ``Climate Change and Australia: Trends,
Projections and Impacts.'' \emph{Austral Ecology} 28 (4). Blackwell
Science Pty: 423--43.
\url{http://dx.doi.org/10.1046/j.1442-9993.2003.01300.x}.

\hypertarget{ref-McAlpine2007}{}
McAlpine, C. A., J. Syktus, R. C. Deo, P. J. Lawrence, H. A. McGowan, I.
G. Watterson, and S. R. Phinn. 2007. ``Modeling the Impact of Historical
Land Cover Change on Australia's Regional Climate.'' \emph{Geophysical
Research Letters} 34 (22).
\href{\%3CGo\%20to\%20ISI\%3E://000251345100005}{\textless{}Go to ISI\textgreater{}://000251345100005}.

\hypertarget{ref-serinaldi2018}{}
Serinaldi, Francesco, Chris G. Kilsby, and Federico Lombardo. 2018.
``Untenable Nonstationarity: An Assessment of the Fitness for Purpose of
Trend Tests in Hydrology.'' Journal Article. \emph{Advances in Water
Resources} 111: 132--55.
doi:\href{https://doi.org/https://doi.org/10.1016/j.advwatres.2017.10.015}{https://doi.org/10.1016/j.advwatres.2017.10.015}.

\end{document}

### 2019-06-17 Manuscripts/PeerJRainfall_Landuse/Rainfall_Landuse.Rmd
changed files after review and resubmitted PeerJ

@@
 abstract: > 
- Analysis of observational data to pinpoint impact of land cover change on local rainfall are difficult due to multiple environmental factors that cannot be strictly controlled.
+ Analysis of observational data to pinpoint impact of land cover change on local rainfall is difficult due to multiple environmental factors that cannot be strictly controlled.
 In this study we use a statistical approach to identify the relationship between removal of tree cover and rainfall with data from best available sources for two large areas in Australia.
 Gridded rainfall data between 1979 and 2015 was used for the areas, while large scale (exogenous) effects were represented by mean rainfall across a much larger area and climatic indicators, such as Southern Oscillation Index and Indian Ocean Dipole.
-Both general additive modelling and step trend tests were used for the analysis.
+Both generalised additive modelling and step trend tests were used for the analysis.
   For a region in south central Queensland, the reported change in tree clearing between 2002 - 2005 did not result in strong statistically significant precipitation changes.
 On the other hand, results from a bushfire affected region on the border of New South Wales and Victoria suggests significant changes in the rainfall due to changes in tree cover.
-This indicates the method works better when a abrupt change in the data can be clearly identified.
+This indicates the method works better when an abrupt change in the data can be clearly identified.
 The results from the step trend test also mainly identified a positive relationship between the tree cover and the rainfall at p < 0.1 at the NSW/Victoria region.
@@
  \usepackage{textcomp, rotating}
+ \usepackage[normalem]{ulem}
 bibliography: reflist_chap3.bib
+csl: peerj.csl
 date: "`r Sys.Date()`"
 output:
-  rticles::peerj_article:
-    base_format: bookdown::pdf_document2 # bookdown::pdf_document2 # for using \@ref()
+  bookdown::pdf_book:
+    base_format: rticles::peerj_article # for using bookdown features like \@ref()
+  rticles::peerj_article: default
 ---
@@
 knitr::opts_knit$set(root.dir =
-                       "E:/cloudstor/chun/chun/manuscripts/PeerJRainfall_Landuse")
-```
-
-# Introduction {-}
-
+      # "E:/cloudstor/chun/chun/manuscripts/PeerJRainfall_Landuse")
+      "C:/users/rver4657/owncloud/chun/chun/manuscripts/PeerJRainfall_Landuse")
+```
+
+Introduction {-}
+========================
 Land use and land cover changes can lead to changes in the local climate.
@@
 Most of the studies mentioned previously were either model simulations, or comparisons of modelled data with observations.
-This is because there are some fundamental experimental difficulties in both space (where does evaporated water reappear as rainfall?) and in time (how much time does it take for land cover change effects to appear or disappear?).
-In addition, in many areas across the globe, rainfall variability is related to a complex set of interactions, of which land use change might only be a minor component.
-
-Locally, there are two main sources that generate rainfall: moisture from advective atmospheric transport; and local evapotranspiration [@Eltahir1996; @Bosilovich2006; @Dirmeyer2009; @Gimeno2010].
+This is because there are some fundamental experimental difficulties in both space (where does evaporated water reappear as rainfall?) and time (how much time does it take for land cover change effects to appear or disappear?).
+In addition, in many areas across the globe, rainfall variability is related to a complex set of interactions (outlined below), of which land use change might only be a minor component.
+
+Locally, and on a shorter, daily time scale, there are two main sources that generate rainfall: moisture from advective atmospheric transport; and local evapotranspiration [@Eltahir1996; @Bosilovich2006; @Dirmeyer2009; @Gimeno2010].
 The local evapotranspiration component is the component considered to be affected by land use change [@Eltahir1996].
@@
 @Westra2010 pointed out that models based on global sea surface temperature anomalies can only predict up to 14.7\% of annual precipitation variance.
-Some of the remaining variance could be due to land surface processes as suggested in studies predicting local rainfall [e.g.
+More generally, some of the remaining variance could be due to land surface processes as suggested in studies predicting local rainfall [e.g.
 @Ma2011; @Zeng2012; @pitman_scale_2016; @saha_investigating_2016].
@@
 Severe bush fires occurred in this area  in early 2003 (see Figure \@ref(fig:bushfire)).
-The 2003 bush fires were the largest and the worst in the last 60 years [@Fire2011].
-Two thirds of Kosciuszko national park was heavily burned and regrowth was reported to be slow due to drought and cold conditions [@ABC2003], and the type of species in this region.
+The 2003 bush fires were the largest in the last 60 years [@Fire2011].
+Two thirds of Kosciuszko national park was heavily burned and regrowth was suggested to be slow due to drought and cold conditions [@ABC2003], and the type of species in this region.
 However, in the longer term, after an early high transpiration period a recovery of pre-fire evapotranspiration would be expected [@Kuzcera1987].
@@
 Here tree cover is denser with open or even closed forest (the tree cover distribution is bimodal at 10 - 20\% and 60 - 70\%).
-The dominant species in the alpine area are Snow Gum and large stand species such as Alpine Ash and Mountain Gum in the sub-alpine area.
+The dominant species in the alpine area are Snow Gum (*Eucalyptus Pauciflora*)and large stand species such as Alpine Ash (*Eucalyptus delegatensis*) and Mountain Gum (*Eucalyptus dalrympleana*) in the sub-alpine area.
 These trees can reach a great height but they take long time to grow.
-For example, Alpine Ash would need about 20 years to mature.
-Although land clearing is not a major issue in this region, it is vulnerable to fires and drought.
+For example, Alpine Ash (*Eucalyptus delegatensis*) would need about 20 years to mature (40 - 45 m, @buckley2012).
+This region is vulnerable to fires and drought, however land clearing is not a major issue.
 The MODIS burned area product, MCD45A1 [@Roy2002; @Roy2005; @Roy2008], was used to locate bush fires areas in the NSW/VIC region, with a grid resolution of 500 m.
@@
 
-```{r figure3-tc-simple, fig.cap= "The expected evolution of the land surface after trees have been removed in (a) the QLD region and (b) the NSW/VIC region.", out.width="90%", echo=F}
+```{r figure3-tc-simple, fig.cap= "The expected evolution of the land surface after trees have been removed in (left) the QLD region and (right) the NSW/VIC region.", out.width="90%", echo=F}
 include_graphics("figures/tc_simple.png")
@@
 The actual tree cover change during this time at the pixel level was derived from the 15-year MODIS data (discussed below).
-The difference of tree cover before and after the land disturbance was tested using a Student's t-test.
 As the length of the tree cover data is shorter than available rainfall data, earlier land clearing in the QLD region cannot be identified spatially, hence they are excluded from the analysis.
@@
 Therefore, the spatially averaged monthly rainfall in the Murray Darling Basin (MDB, downloaded from [the Bureau of Meteorology](http://www.bom.gov.au/web01/ncc/www/cli_chg/timeseries/rain/allmonths/mdb/latest.txt)) was used to explain the year-on-year variation in the rainfall in the regions.
-As climate alone are unlikely able to fully explain the temporal variation in rainfall in the regions, including the drought [@vanDijk2013; @Westra2010], this variable was included.
 Since both regions at least partly overlap with the MDB, the average rainfall for the entire basin was assumed to be a useful explaining variable.
@@
 # autocorrelation in rainfall
-acf_df <- data_frame(lag = 1:12, 
+acf_df <- tibble(lag = 1:12, 
                      acf_Qld = acf(qld_mean$Means,
@@
 Using detrended data gives a better indication of the underlying correlation by removing the underlying correlation in the data [@Smith2012].
-The cross-correlations between the deseasonalised and detrended rainfall and the climatic indices were tested using the Pearson's product moment correlation method, assuming the relationships are linear.
-Although the optimal technique for exploring the correlation with each index could be different as described in @Risbey2009, the Pearson's method was applied to all indices for consistency.
+The cross-correlations between the deseasonalised and detrended rainfall and the climatic indices were tested, and the strongests indicators at lag 0 were identified for the model.
 Because the PDO describes the multi-decadal SST with lower frequency [@MacDonald2005; @Zanchettin2008; @Kamruzzaman2011], instead of 37-year rainfall data, a longer period (115 years, from 1900 to 2015) was used to estimate the correlation with PDO, up to lag 24.
@@
 
-```{r cor-rain-qld, fig.cap="Cross-correlation of six climate indices and rainfall in QLD study region.
-For the PDO analysis, 108-year rainfall data (1900 - 2008) are used.
-Otherwise, 36-year rainfall data are used.
-The correlation with NINO 3 is not shown as it is very similar to but weaker than for NINO 3.4.", out.width="90%", echo=F}
-include_graphics("figures/cor_qld.png")
-```
-
-```{r cor-rain-nsw, fig.cap="Cross-correlation of six climate indices and rainfall in NSW/VIC study region.", out.width="90%", echo=F}
-include_graphics("figures/cor_nswvic.png")
-```
-
-Based on the correlation between the climatic indices and rainfall in the regions (as shown in Figure \@ref(fig:cor-rain-qld) and Figure \@ref(fig:cor-rain-nsw)), it can be concluded that:
-\begin{itemize}
-  \setlength{\itemsep}{0cm}
-  \setlength{\parskip}{0cm}
-  \item In QLD, the correlation between rainfall and SOI at zero time lags is the strongest across all indices, outweighing the other ENSO indicators. IOD and PDO have a weak influence in QLD.
-  \item In NSW/VIC, again the SOI has the strongest correlation with rainfall, followed by the IOD. For both, the strongest correlations occur at the zero time lags. 
-  \item In both cases PDO had the weakest correlations, and this factor was not further considered as a predictor.
-  \item In general, for the better correlated indices, strongest correlations occured at zero time lags.
-\end{itemize}
-The above findings are consistent with previous studies.
-Although some indices are serially correlated with rainfall up to several months, the lag zero events have the greatest correlation coefficients.
-Using multiple climatic index series was generally found most useful in rainfall prediction [e.g.
-@Risbey2009; @Kamruzzaman2011].
-
 Rainfall in Australia shows strong seasonal patterns [@Holper2011; @ABS2012].
-For example, the north part of the country is summer rainfall dominant with a dry winter, while most of the southern part has a winter rainfall regime.
-This character is driven by the movement of subtropical high pressure systems which dominate the Australian climate [@BoM2012a].
-These characteristic summer and winter rainfall patterns mean that the seasonal component of rainfall has a periodic pattern which should be included in the model.
-
-Long term trends in the regional rainfall in some parts of Australia are significant [@Hughes2003; @Gallant2007; @Chowdhury2010].
-In the northern and eastern parts of the continent, increasing rainfall is reported over the last century [@Hughes2003].
-The presence of such long term trends may be confused with the outcome of a step change in rainfall.
-As a result a linear trend term was implemented in the model to remove any long term effect.
+As a result a seasonal component of rainfall has a periodic pattern which should be included in the model.
+In addition, long term trends in the regional rainfall in some parts of Australia are significant [@Hughes2003; @Gallant2007; @Chowdhury2010].
+The presence of long term trends can be confused with the outcome of a step change in rainfall.
+As a result a linear trend term was implemented in the model to remove any long term effects.
 
 We assumed all the factors are additive smooth components in determining rainfall following @Kamruzzaman2011.
-In this case, the rainfall model is a generalised additive model (GAM) [@Hastie1986] with a log link function `g()` and assuming the residuals are gamma distributed (see Figure \@ref(fig:hist-rain)).
-This means all predictors are modelled as smooth functions, in this case using the shrinkage version of the cubic regression splines [@Wood2011].
+In this case, the rainfall model is a generalised additive model (GAM) [@Hastie1986] with a log link function `g()` and assuming the residuals are gamma distributed.
+We used the shrinkage version of the cubic regression splines [@Wood2011] as smooth function.
 All splines were limited to 3 knots in flexibility [@Wood2011], to reduce the risk of overfitting.
@@
 \begin{array}{lll}
-g(E(\mathbf{R}_r)) = &\beta_0 + s_1(\mathbf{MDB_{monthlyRain}} + s_2(\mathbf{SOI}) + s_3(\mathbf{IOD}) + \\
-&s_4(\mathbf{Nino3.4}) + s_5(\mathbf{Nino4}) + s_6(\mathbf{Season}) + \\
-           &\beta_1\mathbf{Trend} + \boldsymbol{\epsilon}_r
+g(E(\mathbf{R}_r)) = &\beta_0 + s_1(\mathbf{MDB_{monthlyRain}}) + s_2(\mathbf{SOI}) + s_3(\mathbf{IOD}) + \\
+&s_4(\mathbf{Nino3.4}) + s_5(\mathbf{Nino4}) + s_6(\mathbf{PDO}) + \\
+&s_7(\mathbf{Season}) + \beta_1\mathbf{Trend} + \boldsymbol{\epsilon}_r
 \end{array}
@@
 MDB~monthlyRain~ is the spatially averaged monthly rainfall across the Murray Darling Basin.
-Apart from dropping PDO as a predictor, all other climate indices were included, allowing the model to select the appropriate predictors.
-
-A possible linear long term trend in the rainfall data is modelled by \textbf{Trend} = 1,2,3...n, where n is the total number of months in the time series. \textbf{Season} is the seasonal component.
-The climatic terms are also modelled with smooth functions.
-The effect of large scale drivers on Australian rainfall is more likely to be seasonal [@Murphy2008; @Schepen2012], and the smooth spline function is more flexible in reproducing the variability in impacts of the climatic indices.
+
+The variable \textbf{Trend} = 1,2,3...n, where n is the total number of months in the time series is the possible long term trend in the data, while \textbf{Season} is the seasonal component.
+The climatic terms are also modelled with smooth functions as the effect of large scale drivers on Australian rainfall can be seasonal [@Murphy2008; @Schepen2012].
 
 ```{r hist-rain, fig.cap="Distribution of monthly rainfall in (a) QLD and (b) NSW/VIC.
-Using a Kolmogorov-Smirnov test with shape = 1 and 2.4  for QLD and NSW/VIC respectively, rainfall in both regions can be modelled as a gamma distribution.", echo=F, out.width="90%"}
+Using a Kolmogorov-Smirnov test with shape = 1 and 2.4  for QLD and NSW/VIC respectively, rainfall in both regions can be modelled as a gamma distribution.", echo=F, eval=F, out.width="90%"}
+# removed after review
 include_graphics("figures/hist_rainfall.png")
@@
 \begin{array}{lll}
-g(E(\mathbf{R}_r)) = &\beta'_0 + s'_1(\mathbf{MDB_{monthlyRain}} + \\
+g(E(\mathbf{R}_r)) = &\beta'_0 + s'_1(\mathbf{MDB_{monthlyRain}}) + \\
   &s'_2(\mathbf{SOI}) + s'_3(\mathbf{IOD}) + \\
-  &s'_4(\mathbf{Nino3.4}) + s'_5(\mathbf{Nino4}) + s'_6\mathbf{Season} + \\
+  &s'_4(\mathbf{Nino3.4}) + s'_5(\mathbf{Nino4}) + s'_6\mathbf{PDO} + s'_7\mathbf{Season} + \\
   &\beta'_1\mathbf{Trend} + \beta'_2\mathbf{LC} + \boldsymbol{\epsilon'}_r
@@
 # show histogram
-plot_df <- data_frame(`r.squared` = c(rsq_qld,rsq_nsw),
+plot_df <- tibble(`r.squared` = c(rsq_qld,rsq_nsw),
                       state = c(rep("QLD", length(rsq_qld)),
@@
   qld.resid.mean <- apply(test,1, mean,na.rm=T)
-  residuals <- data_frame(Dates = seq.Date(as.Date("1979-01-01"),
+  residuals <- tibble(Dates = seq.Date(as.Date("1979-01-01"),
                              length = length(qld.resid.mean),
@@
 This specific nonparametric statistical test was modified from the Mann-Whitney U test by @Hirsch1985 and can identify a step change in data which is cross-correlated.
-The gridded rainfall dataset used in this study has a high spatial correlation between neighbouring pixels due to the underlying interpolation method.
+In this case, this is important as the gridded rainfall dataset has a high spatial correlation between neighbouring pixels.
 The advantages of using the Rank-Sum test are: (1) it does not depend on assumptions of the data distribution; (2) it is not restricted to datasets with no missing data; (3) it is robust and not as easily influenced by outliers and negative numbers [@Hirsch1985].
 However, the test has less power than parametric tests.
-
-The rainfall residuals from the regression model in Equation \@ref(eq:model2) were used for the step trend test.
-To detect a step change, using deseasonalised and detrended data is important [@Hirsch1985].
-Furthermore, as rainfall can only be partially attributed to local sources and conditions, other effects are introduced by large scale dynamics and changes in climatic factors.
-The assumption is that the regression model should remove these effects, and additionally deseasonalise and detrend the rainfall data.
-As a result, the local landuse effects are amplified in the remaining residuals.
-The  test, described in the following sub section, subsequently associates trends in the rainfall residuals with tree cover changes.
+As a nonparametric rank-based test, it  depends on the ranks of the data.
+
+The rainfall residuals from the regression model in Equation \@ref(eq:model2) were used.
+The assumption is that the regression model deseasonalises and detrends the rainfall data  [@Hirsch1985], amplifying the local landuse effects.
+For each month, rainfall residuals of each year were ranked in an ascending order.
+The ranking of January rainfall in a sample pixel k in QLD is illustrated in Table \@ref(tab:exampleRank).
 
@@
 # # create a boxplot with the annual residuals from 1998
-# resplot_df <- data_frame(Date = rep(ymd(paste(nsw_30[[1]],
+# resplot_df <- tibble(Date = rep(ymd(paste(nsw_30[[1]],
 #                                               "01", sep="-")),
@@
 # create a boxplot with the annual residuals from 1998
-resplot_df <- data_frame(Date = rep(ymd(rainNSW_t[[1]][1:444]),
+resplot_df <- tibble(Date = rep(ymd(rainNSW_t[[1]][1:444]),
           length(residual_nswvic) + length(residual_qld)),
@@
 
-### Mann-Whitney rank-sum statistic
-
-As indicated, the step trend test is a modified version of the Mann-Whitney U statistic [@Hirsch1985].
-As a nonparametric rank-based test, the Mann-Whitney test does not use the exact values of rainfall but depends on the ranks of the data.
-For each month, rainfall residuals of each year were ranked in an ascending order.
-The ranking of January rainfall in a sample pixel k in QLD is illustrated below:
-
-```{r exampleRank, echo=F}
-table <- data_frame(Year = c(1998 , 1999 , 2000 , 2001 ,
-                             2002 , 2005 , 2006),
+```{r exampleRank, echo=F, warning = F}
+table <- tibble(Year = c(1998, 1999, 2000, 2001,
+                             2002, 2005, 2006),
                     `Rainfall residuals` =c(-0.3, -60.9, -16.1,
-                                                      -71.7, 111.1,
-                                             -7.2, -60.5),
-                    Rank= c(6, 2, 4,
-                                                               1, 7, 5, 3))
+                                            -71.7, 111.1,-7.2,
+                                            -60.5),
+                    Rank= c(6, 2, 4, 1, 7, 5, 3))
 colnames(table)[3] <- paste("Rank","$R'_{1k}$")
 kable(table, caption="Example of ranking rainfall residuals",
-             booktabs=T)
-
-```
-
-\noindent Therefore, the smallest or most negative value has rank 1 and the largest value has the maximum rank.
+             booktabs=T, escape=F)
+
+```
 
@@
 In the QLD region, changes occurred anytime during 2003 and 2004.
-In contrast to the previous method, the time period covering the land cover change  was excluded, as the nonparametric test allows missing data.
-@Hirsch1985 also pointed out that the power of the test is higher if the data of the change period is ignored.
-Hence 2003 and 2004 were excluded from the analysis.
+In contrast to the previous method, the time period covering the land cover change  was excluded, as the nonparametric test allows missing data and the power of the test is greater if data of the change period is ignored [@Hirsch1985].
 As a result, the post change period was 2005 - 2015 for the Queensland location.
@@
 Therefore the post change period in this region still started in January 2003.
-Following @Hirsch1985, the pre change period was set to five years (1998 - 2002) in both regions.
-The length of the post change period is also difficult for the NSW/VIC region, as the regrowth would at some point have impacted the local effects.
+The pre change period was set to five years (1998 - 2002) in both regions.
 
@@
 
-### Step trend test
-
-Instead of completing the Mann-Whitney U-test, @Hirsch1985 applied a standard normal Z test to the rank-sum statistics.
-As highlighted, this modified test accounts for serial temporal autocorrelation and spatial cross correlation in the data.
-In the case here, the deseasonalised and detrended data shows little autocorrelation in the time series but possesses strong cross correlation between neighbouring pixels, i.e.
-$R>0.99$.
+Here the deseasonalised and detrended data shows little autocorrelation in time, but possesses strong cross correlation between neighbouring pixels, i.e. $R>0.99$.
 
@@
 
-```{r Zscore, echo=F}
-table2 <- data_frame(" "= c("$Z'>0$  and rainfall decreases after change",
+```{r Zscore, echo=F, warning = F}
+table2 <- tibble(" "= c("$Z'>0$  and rainfall decreases after change",
     "$Z'<0$ and rainfall increases post change",
@@
 The null hypothesis ($H_0$) in this study is that there was no change in rainfall due to land surface intervention.
-The results of the step trend test can be interpreted according to the sign of the Z' score (see Table \@ref(tab:Zscore), Chapter 23, P887 [@Hipel1994]).
-Z' is normally distributed similar to the standard normal statistics Z.
-Hence it can be compared to a standard normal distribution to determine the p value.
-
-```{r steptrend_ns4, echo=F, eval=F}
+The results of the step trend test can be interpreted according to the sign of the Z' score (see Table \@ref(tab:Zscore), Chapter 23, P887 [@Hipel1994]), and is normally distributed similar to the standard normal statistics Z.
+
+As part of the analysis, the "field significance" of the Z' score test was considered to improve the interpretion of the step change at regional scales from multiple local tests [@Wilks2006; @Westra2013].
+Here, the bootstrapping resampling method from @Westra2013 was used to evaluate the field significance.
+This means the spatial structure of the pixels was maintained, but the order of the years and months was changed by random resampling.
+For each resampling, the test statistic identifies the percentage of the pixels with a significant positive or negative step change for the step trend test.
+The test statistics on 1000 resampled replicates were used to develop the distribution of these percentage values.
+
+```{r steptrend_ns4, echo=F, eval=F, warning = F}
 ## perform step test on the data using ns = 4
@@
 
-```{r bootstrap_ns1, echo=F, eval=F}
+```{r bootstrap_ns1, echo=F, eval=F, warning = F}
 ## 4 cell step trend
@@
 Tree cover loss occurred in large parts of the QLD region between 2002 and 2005, but this tree loss was spatially less concentrated.
+
 ```{r tctrend, fig.cap ="The maps show significant changes in tree cover identified from the MOD44B data between 2003 and 2015 in the NSW/VIC region (right) and the Qld region (left).
@@
 
-## Regression Model and significance of Vegetation Cover Changes
-
-```{r rsqmodel, fig.cap = "The distribution of the variance explained (adjusted $r^2$ ) by the full regression model", echo=F, out.width="90%"}
+## Regression Model
+
+The correlation between the climatic indices and rainfall in the regions (See the supplementary material with this paper) indicated that for both regions the PDO had the weakest correlations, and this factor was therefore dropped as a predictor.
+Although some indices are serially correlated with rainfall up to several months, the lag zero events have the greatest correlation coefficients.
+Furthermore, using multiple climatic index series was generally found most useful in rainfall prediction [e.g.
+@Risbey2009; @Kamruzzaman2011].
+
+```{r rsqmodel, fig.cap = "The distribution of the variance explained (adjusted $r^2$ ) by the full regression model", echo=F, out.width="70%"}
 include_graphics("figures/rsq_fullmodel.png")
@@
 
-Generally the regression model only explains a limited amount of the rainfall variability (Figure \@reg(fig:rsqmodel)).
+Generally the regression model only explains a limited amount of the rainfall variability (Figure \@ref(fig:rsqmodel)).
 The model in Equation \@ref(eq:model2) accounts for an average around 33\% of the variation in the NSW/VIC region and about 50% of the variation in the QLD region as indicated by the adjusted $r^2$.
 The adjusted $r^2$ is the coefficient of determination, a measurement of the amount of variability predicted by the model adjusting for the number of explanatory terms.
-The residual analysis shows that the assumptions of the regression model are generally met (Figure \@ref(fig:residuals)).
-The standardised residual plots, however, show some funnelling for the NSW/VIC regions, suggesting non constant variance.
-The residual patterns are consistent for all pixels within each region.
 
@@
 
-```{r residuals, fig.cap ="The residual analysis of a sample pixel in the QLD region (top) and NSW/VIC region (bottom).", echo=F, out.width="90%"}
-include_graphics("figures/gam_check.png")
-```
-
 ```{r rsq, fig.cap = "The performance of the regression model if rainfall is only modelled by the climate drivers.
@@
 There were generally no statistically significant long term trends in both regions.
-However, this result might not prove or disprove the existence a long term trend in rainfall.
-The overall time period is fairly short [@Koutsoyiannis2007] and more pixels in NSW/VIC could indicate a significant step change if the long term trend effect is not removed by the model.
-As trend free data is an important requirement for the step trend test, the trend term was kept in the regression model to ensure the detection of step change was not due to a possible long term trend.
+However, the trend term was kept in the regression model to ensure the detection of step change was not due to a possible long term trend (even if this was not significant).
 
@@
 The  variable was mainly significant ($p \leq 0.05$) for the rainfall estimates in some areas in NSW/VIC, as shown in Figure \@ref(fig:LCp) (right panel).
-However, the number of pixels where the landcover variable was significant was much greater in NSW/VIC compared to the QLD region.
+However, the number of pixels where the landcover variable was significant was much greater in NSW/VIC compared to the number in the QLD region.
 There was also some relationship between the areas of bushfires in Figure \@ref(fig:bushfire).
@@
 
-## Step Trend Test
+## Stronger step trend in NSW/VIC compared to QLD
 
@@
 
-The spatial step trend test based on the regression residuals without the landcover variable generates Z' scores.
-The results for the Qld region (left) and NSWVIC region (right)  are shown in Figure \@ref(fig:steptest30).
-This figure provides two types of information: the sign and the significance level.
-The sign indicates the direction of the step change in the residuals of the regression, as listed in Table \@ref(tab:Zscore).
-In both regions there are areas of positive Z' values which imply a decrease in rainfall, but this is stronger in the NSW/VIC area than in the QLD area.
-For the NSW/VIC area there appears to be a reasonable relationship between the locations where changes in tree cover are observed (Figure \@ref(fig:tctrend)) and the patterns in the Z' score.
-However, there is not necessarily a direct relationship as movement of air masses could mean that actual changes of rainfall are observed close by, but not necessarily exactly at areas with changes of landcover.
-There is once again a difference between the left panel and the right panel, indicating only a few significant of Z' scores in the QLD region wich matches the earlier significance in the landcover variable in the full regresion.
+In both regions there are areas of positive Z' values which imply a decrease in rainfall (Figure \@ref(fig:steptest30)), but this is stronger in the NSW/VIC area than in the QLD area.
+Qualitatively the locations where changes in tree cover occurred in the NSW/VIC area (Figure \@ref(fig:tctrend)) seem to agree with the patterns in the Z' scores.
+However, a direct relationship would not necesssarily be expected as movement of air masses could mean that actual changes of rainfall are observed close by, but not necessarily exactly at areas with changes of landcover.
+For the QLD region (left panel) there are only a few significant Z' scores which matches the earlier significance in the landcover variable in the full regresion.
 In the QLD region, only 0.2\% of the pixels obtained a positive Z' score with p $<$ 0.1.
@@
 
-```{r meandiff, fig.cap ="Boxplots of annual rainfall residuals (estimated based on Equation 2 before and after the land cover intervention during 1979 - 2015 in the study regions.
-On average, the after period has a significantly lower annual rainfall residual in NSW/VIC, but a signifcantly higher annual rainfall residual in the Qld study area", echo=F, out.width="70%"}
-
-include_graphics("figures/ResidualBoxplotchange.png")
-```
-
-The rainfall regression residuals without the landcover variable for the two regions (before-change since 1979 and after-change) were also compared (Figure \@ref(fig:meandiff)) using a simple t-test.
-From the boxplots it is difficult to see that there were significantly different ($p < 0.05$) mean rainfall residual values between the "before" and "after" periods in both regions.
+```{r fieldsig, fig.cap ="Field significance results showing the distribution of percentage of significant Z' scores for the resampled series (blue bars), and the precentage significant Z' scores for the original series of rainfall years (red circle)", echo=F, out.width="70%"}
+include_graphics("figures/bt4_hist.png")
+```
+
+The rainfall regression residuals without the landcover variable for the two regions (before-change since 1979 and after-change) were also compared using a simple t-test.
+The mean rainfall residuals were significantly different ($p < 0.05$) between the "before" and "after" periods in both regions.
 For the Queensland locations, the mean monthly rainfall residuals were slightly lower  ($p < 0.05$) after the change in landcover ($~ 0.45 mm/month$).
@@
 
-The choice of the number of pixels averaged in the step trend test ($ns$) has some impact on the test results [@Hirsch1985].
-The cases of $ns = 1$ and $ns = 9$ were also tested.
-The results for $ns = 1$ showed a lower number of significant pixels (at $p< 0.10$) compared to the $ns = 4$ test.
-The results from the $ns = 9$ test was not different from the $ns = 4$ test.
-We therefore decided to stay with $ns = 4$.
-The power of the test does not change much after $ns = 4$, as shown by @Hirsch1985.
-
-As part of the analysis, the "field significance" of the Z' score test was considered to improve the interpretion of the step change at regional scales from multiple local tests [@Wilks2006; @Westra2013].
-Here, the bootstrapping resampling method from @Westra2013 was used to evaluate the field significance.
-This means the spatial structure of the pixels was maintained, but the order of the years and months was changed by random resampling.
-For each resampling, the test statistic identifies the percentage of the pixels with a significant positive or negative step change for the step trend test.
-The test statistics on 1000 resampled replicates were used to develop the distribution of these percentage values, and the observed fraction of significant Z' scores (at $p < 0.1$) is plotted on the distribution (Figure \@ref(fig:fieldsig)).
+The observed fraction of significant Z' scores (at $p < 0.1$) is plotted on the results of the bootstrap distribution (Figure \@ref(fig:fieldsig)).
 The results support the earlier Z' score results.
@@
 In other words, in the NSW/VIC region it is most likely that the change in landcover (decrease in tree cover) caused a decrease in the rainfall.
-
-```{r fieldsig, fig.cap ="Field significance results showing the distribution of percentage of significant Z' scores for the resampled series (blue bars), and the precentage significant Z' scores for the original series of rainfall years (red circle)", echo=F, out.width="70%"}
-include_graphics("figures/bt4_hist.png")
-```
 
@@
 
-Overall the summary table of the results indicates that the effects of land clearing or bushfire on rainfall are consistent across the range of different statistical tests.
-In all cases, the data for the NSW/VIC region indicates the strongest evidence that the reduction in tree cover (the land cover change) resulted in a local decrease in rainfall.
-In contrast the data for the QLD region only indicates weak evidence that the loss of trees due to land clearing resulted in reduction in the rainfall.
+Overall the summary table (Table \@ref(tab:summarytable)) of the results indicates that the effects of land clearing or bushfire on rainfall are consistent across the range of different statistical tests.
+In all cases, the data for the NSW/VIC region indicates statistical evidence that the reduction in tree cover (the land cover change) resulted in a local decrease in rainfall.
+In contrast, the data for the QLD region indicates lower statistical evidence that the loss of trees due to land clearing resulted in reduction in the rainfall (Table \@ref(tab:summarytable)).
 
@@
 The advantage of the current approach is that long time series of land cover data are not required.
-This is advantageous as this type of that is currently not always available.
 Furthermore, it does not assume a specific relationship between vegetation cover change and rainfall but allows the data to show this relationship, by applying the analysis to a broader area outside the boundary of the vegetation cover change.
@@
 
-Overall the results suggest that at least for the NSW/VIC region that a decrease in tree cover causes a decrease in rainfall (Figure \@ref(fig:tctrend)).
+Overall the results suggest that at least for the NSW/VIC region, a decrease in tree cover causes a decrease in rainfall.
 However, there are several possible complicating factors in the data that require discussion.
@@
 We used the spatially averaged monthly rainfall timeseries for the larger Murray Darling Basin (Figure \@ref(fig:selreg)) to remove the general variability.
-However, overall the model does not explain more than 50\% (QLD) and 30\% (NSW/VIC) of the rainfall variability, and only around 7\% on average is due to the climate drivers (Figure \@ref(fig:rsq)) .
+However, overall the model does not explain more than 50\% (QLD) and 30\% (NSW/VIC) of the rainfall variability (Figure \@ref(fig:rsqmodel)), and only around 7\% on average is due to the climate drivers (Figure \@ref(fig:rsq)).
 And while this is consistent with the literature e.g.
@@
 
+The fact that no significant long trend was identified, may not disprove a long term trend in rainfall, as overall time period is fairly short [@Koutsoyiannis2007].
+Removing the long term trend could mean more pixels in NSW/VIC would indicate a significant step change.
+
 ## Vegetation dynamics
@@
 The second possible effect is the dynamic nature of the vegetation clearing and recovery, especially for the QLD region.
+Not only does this refer to a change in the total biomass, but this could also include a change in the species composition as a result of the disturbance.
+
 Although land clearing has occurred at a high rate and broad scale in Queensland [@SLATS2017], the clearing does not have a clear start and end point.
@@
 And even though there was a decrease in land clearing post 2005, it is difficult to define a clear cut change in the tree cover in this region.
+
 The specific vegetation class in the Queensland area is also well-known for rapid regrowth and "thickening" in favourable conditions [@Gowen2016], and this could explain the increase in vegetation cover in 2015 in the Qld area (Figure \@ref(fig:tctrend)).
+This again, could also be related to a change in the vegetation composition further complicating the analysis.
 In particular the favourable rainfall years of 2010 and 2011 would have boosted regrowth, increasing evapotranspiration and therefore decreasing the effect of the rainfall change.
@@
 As a non-parametric test, the step trend test has the advantage of being distribution free and having no restriction on missing data [@Hirsch1985].
-This is particularly useful in rainfall analysis since rainfall data is usually highly skewed.
 On the other hand, the disadvantages of non-parametric tests, such as being limited to hypothesis testing and being weaker in power, also hold for the step trend test [@Whitley2002].
@@
 Overall, the current study provides a statistical data based approach, building on several lines of evidence to  reject the null hypothesis (no step change in rainfall occurs as a result of tree cover loss) at least for the NSW/VIC region.
-Limited by the available data, the time frame under study included a long lasting drought period [@Holper2011; @vanDijk2013].
+Limited by the available data, the time frame under study had to include a long lasting drought period [@Holper2011; @vanDijk2013].
 The strong impact of this prolonged drought might have suppressed the land-atmosphere interaction and modified the cause and effect relationship between rainfall and vegetation cover change.
@@
 @Gorgen2006; @McAlpine2007] do not appear in the QLD region.
-Possible future work could focus on a non-drought period, or using a different series of land cover data, possibly on another continent.
-The recent large bushfires on the US continent come to mind as a possible candidate study once sufficient post event data is collected.
-Wile the power of the test can be improved with the longer length of the post-intervention period [@Hirsch1985], the dynamic nature of vegetation regrowth in this case study also affects this effect.
+
+Future research will include additional areas to further test the approach in this study. However, such areas are not easy to find. There are several requirements for a good study area:
+
+1. It needs to be large enough to capture the effect of landuse on rainfall within the area;  
+2. It needs to have a drastic enough landuse change to be observable above the rainfall variation;  
+3. It needs to have a long enough climate data time series both prior and post landuse change.
+
+The obvious locations have always been Amazonia and sub-Saharan Africa, where also most of the modelling studies have taken place [@kucharski_further_2013; @pitman_scale_2016].
+The problem with the sub-Saharan location is that the land use change is quite long ago and local data is difficult to obtain.
+The problem with the Amazonian location is that this area is well-known for the feedback and we felt that we could not add much to this work.
+We propose that it would better to focus on areas affected by bushfires, as we did with the NSW/VIC location.
+The recent bushfires in California and Colorado could potentially be good locations, but at the moment there is insufficient "post event data".
+Further work could also focus on identifying an area that has been impacted by land cover change, but does not include a significant drought period.
+
+While the power of the test can be improved with the longer length of the post-intervention period [@Hirsch1985], the dynamic nature of vegetation regrowth in this case study also affects this effect.
 A better approach might be to build a global study that investigates multiple locations where drastic landcover changes have taken place, which would also remove some of the climate variability effects due to the larger sample size.
@@
 
-Drought may have had a pronounced impact on the land surface condition during the study period, also leading to significant reduction in vegetation cover and extreme events such as bushfires.
+Drought may have had a pronounced impact on the land surface condition during the study period, such asleading to significant reduction in vegetation cover and extreme events such as bushfires.
 The lack of rainfall and associated high temperatures may mask the impact on rainfall of a step change in the vegetation cover.
@@
 
-appendix
-==============
-*Summary of Data*
-
- \begin{sidewaystable}
- \caption{Summary of data.}
-  \label{tab:ch3Data}
- \begin{tabular}{lllll}
-  \hline
-  \textbf{Data} & \textbf{Source} & \multicolumn{2}{c}{\textbf{Resolution}} & \textbf{Analysis period} \\\cline{3-4}
-  & & Temporal & Spatial & \\\hline
-  Percent tree cover & MOD44B & Annual &	250m & 2000-2010\\
-  Trend of vegetation cover change  &	DLCD (2009)	& Onetime & 250m	& Trend of Apr 2000 - Apr 2008\\
-  Rainfall &	AWAP gridded rainfall data &	Monthly &	0.05\textdegree$\times$0.05\textdegree & Jan 1979- Dec 2008\\
-  SOI	& BoM	& Monthly &	N/A &	Jan 1979- Dec 2008\\
-  NINO 3, 3.4, 4 &	IRI/LDEO data library	& Monthly	& N/A	& Jan 1979- Dec 2008\\
-  PDO	& NOAA & Monthly	& N/A	& Jan 1979- Dec 2008\\
-  IOD	& POAMA-2 dataset	& Monthly	& N/A	& Jan 1979- Dec 2008\\
-  \hline
-  \end{tabular}
-\end{sidewaystable}
-
-\newpage
-
 References {#references .unnumbered}

### 2019-06-17 Manuscripts/PeerJRainfall_Landuse/SupplementaryMaterialRainfall_Landuse.Rmd
changed files after review and resubmitted PeerJ

---
title: Detecting the impact of land cover change on observed rainfall.
preprint: false
author:
  - name: Chun X. Liang
    affiliation: 1
  - name: Floris F. van Ogtrop
    affiliation: 1
  - name: R. Willem Vervoort
    affiliation: 1
    corresponding: true
    email: willem.vervoort@sydney.edu.au
affiliation:
  - code: 1
    address: Sydney Institute of Agriculture, The University of Sydney, NSW 2006
abstract: >
  This is the supplementary data for the paper "Detecting the impact of land cover change on observed rainfall." Peerj article number 35847 
header-includes: >
 \usepackage{lipsum}
 \usepackage{textcomp, rotating}
 \usepackage[normalem]{ulem}
bibliography: reflist_chap3.bib
date: "`r Sys.Date()`"
output:
  bookdown::pdf_book:
    base_format: rticles::peerj_article # for using bookdown features like \@ref()
  rticles::peerj_article: default
---
```{r setup, warning=F, message=F, echo =F}
# root dir
knitr::opts_knit$set(root.dir =
      # "E:/cloudstor/chun/chun/manuscripts/PeerJRainfall_Landuse")
      "C:/users/rver4657/owncloud/chun/chun/manuscripts/PeerJRainfall_Landuse")
```

```{r packages, echo=F, warnings=F, message=F, eval=T}
suppressWarnings(require(knitr))
suppressWarnings(require(bookdown))
suppressWarnings(require(pander))
suppressWarnings(library(fields))
suppressWarnings(library(graphics))
suppressWarnings(library(maptools))
suppressWarnings(library(plotKML))
suppressWarnings(library(raster))
suppressWarnings(library(reshape))
suppressWarnings(library(rgeos))
suppressWarnings(library(rgdal))
suppressWarnings(library(RColorBrewer))
suppressWarnings(library(tidyverse))
suppressWarnings(require(lubridate))
suppressWarnings(library(deseasonalize))
suppressWarnings(library(mgcv))
suppressWarnings(library(ncdf4))
suppressWarnings(require(gridExtra))
suppressWarnings(require(kableExtra))
#library(nnet)
```

Code from the paper on Github
===================
The Rmarkdown documents contain the code used in the analysis. These documents including most of the data and additional scripts are accessible via Github: \url{https://github.com/WillemVervoort/RainfallLandcover/releases}.  
Please cite our paper if you plan to use any of the code.

Summary of Data
================

This is a summary table of all the data used in the paper.

 \begin{sidewaystable}
 \caption{Summary of data.}
  \label{tab:ch3Data}
 \begin{tabular}{lllll}
  \hline
  \textbf{Data} & \textbf{Source} & \multicolumn{2}{c}{\textbf{Resolution}} & \textbf{Analysis period} \\\cline{3-4}
  & & Temporal & Spatial & \\\hline
  Percent tree cover & MOD44B & Annual &	250m & 2000-2010\\
  Trend of vegetation cover change  &	DLCD (2009)	& Onetime & 250m	& Trend of Apr 2000 - Apr 2008\\
  Rainfall &	AWAP gridded rainfall data &	Monthly &	0.05\textdegree$\times$0.05\textdegree & Jan 1979- Dec 2008\\
  SOI	& BoM	& Monthly &	N/A &	Jan 1979- Dec 2008\\
  NINO 3, 3.4, 4 &	IRI/LDEO data library	& Monthly	& N/A	& Jan 1979- Dec 2008\\
  PDO	& NOAA & Monthly	& N/A	& Jan 1979- Dec 2008\\
  IOD	& POAMA-2 dataset	& Monthly	& N/A	& Jan 1979- Dec 2008\\
  \hline
  \end{tabular}
\end{sidewaystable}

\newpage

Cross correlations Rainfall and Climate Indices
===============================================

These figures were moved to the supplementary data in the review process.

```{r cor-rain-qld, fig.cap="Cross-correlation of six climate indices and rainfall in QLD study region. For the PDO analysis, 108-year rainfall data (1900 - 2008) are used. Otherwise, 36-year rainfall data are used. The correlation with NINO 3 is not shown as it is very similar to but weaker than for NINO 3.4.", out.width="90%", echo=F}
include_graphics("figures/cor_qld.png")
```

```{r cor-rain-nsw, fig.cap="Cross-correlation of six climate indices and rainfall in NSW/VIC study region.", out.width="90%", echo=F}
include_graphics("figures/cor_nswvic.png")
```

Based on the correlation between the climatic indices and rainfall in the regions (as shown in Figure \@ref(fig:cor-rain-qld) and Figure \@ref(fig:cor-rain-nsw)), it can be concluded that:

- In QLD, the correlation between rainfall and SOI at zero time lags is the strongest across all indices, outweighing the other ENSO indicators. IOD and PDO have a weak influence in QLD.  
- In NSW/VIC, again the SOI has the strongest correlation with rainfall, followed by the IOD. For both, the strongest correlations occur at the zero time lags.  
- In both cases PDO had the weakest correlations, and this factor was therefore dropped as a predictor.  
- In general, for the better correlated indices, strongest correlations occured at zero time lags.

Although some indices are serially correlated with rainfall up to several months, the lag zero events have the greatest correlation coefficients. Furthermore, using multiple climatic index series was generally found most useful in rainfall prediction [e.g. @Risbey2009; @Kamruzzaman2011].

\newpage

Further figures
===================

These two figures highlight the GAM residual analysis for a sample pixel. Residuals for other pixels were similar across the regions analysed.

```{r residuals, fig.cap ="The residual analysis of a sample pixel in the QLD region (top) and NSW/VIC region (bottom).", echo=F, eval=T, out.width="90%"}
# removed after review
include_graphics("figures/gam_check.png")
```

This figure indicates a boxplot of the annual rainfall residuals across the regions.

```{r meandiff, fig.cap ="Boxplots of annual rainfall residuals (estimated based on Equation 2 before and after the land cover intervention during 1979 - 2015 in the study regions. On average, the after period has a significantly lower annual rainfall residual in NSW/VIC, but a significantly higher annual rainfall residual in the Qld study area", echo=F, eval=T, out.width="70%"}
# figure dropped after review
include_graphics("figures/ResidualBoxplotchange.png")
```

# References

### 2019-06-17 Manuscripts/PeerJRainfall_Landuse/SupplementaryMaterialRainfall_Landuse.tex
changed files after review and resubmitted PeerJ

--- 2019-03-14 Manuscripts/PeerJRainfall_Landuse/Rainfall_Landuse.tex
@@
 \usepackage{lipsum} \usepackage{textcomp, rotating}
+\usepackage[normalem]{ulem}
 
@@
 \begin{abstract}
-Analysis of observational data to pinpoint impact of land cover change
-on local rainfall are difficult due to multiple environmental factors
-that cannot be strictly controlled. In this study we use a statistical
-approach to identify the relationship between removal of tree cover and
-rainfall with data from best available sources for two large areas in
-Australia. Gridded rainfall data between 1979 and 2015 was used for the
-areas, while large scale (exogenous) effects were represented by mean
-rainfall across a much larger area and climatic indicators, such as SOI
-and IOD. Both general additive modelling and step trend tests were used
-for the analysis. For a region in south central Queensland, the reported
-change in tree clearing between 2002 - 2005 did not result in strong
-statistically significant precipitation changes. On the other hand,
-results from a bushfire affected region on the border of New SOuth Wales
-and Victoria suggests significant changes in the rainfall due to changes
-in tree cover. This indicates the method works better when a abrupt
-change in the data can be clearly identified. The results from the step
-trend test also mainly identified a positive relationship between the
-tree cover and the rainfall at p \textless{} 0.1 at the NSW/Victoria
-region. High rainfall variability and possible regrowth could have
-impacted the results in the Queensland region.
+This is the supplementary data for the paper ``Detecting the impact of
+land cover change on observed rainfall.'' Peerj article number 35847
 % Dummy abstract text.
@@
 
+\usepackage{amsthm}
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{lemma}{Lemma}[section]
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
+\newtheorem{corollary}{Corollary}[section]
+\newtheorem{proposition}{Proposition}[section]
+\theoremstyle{definition}
+\newtheorem{example}{Example}[section]
+\theoremstyle{definition}
+\newtheorem{exercise}{Exercise}[section]
+\theoremstyle{remark}
+\newtheorem*{remark}{Remark}
+\newtheorem*{solution}{Solution}
 \begin{document}
@@
 
-\section*{Introduction}\label{introduction}
-\addcontentsline{toc}{section}{Introduction}
+\section{Code from the paper on
+Github}\label{code-from-the-paper-on-github}
 
-Land use and land cover changes can lead to changes in the local
-climate. Empirical and modelling studies have found that cloud types and
-rainfall are correlated to large scale vegetation cover changes, such as
-deforestation in the Amazon and in the Sahel (Chagnon and Bras 2005;
-Pinto et al. 2009; Wang et al. 2009; Mei and Wang 2010; Kucharski, Zeng,
-and Kalnay 2013; Pitman and Lorenz 2016) and afforestation in south
-Israel (Otterman et al. 1990; Ben-Gai et al. 1998). Using airborne
-measurements in Western Australia, Junkermann et al. (2009) showed a
-significantly higher level of aerosols over an agricultural area
-compared to an adjacent area with natural vegetation. They suggested
-that a modification of aerosol concentrations due to deforestation could
-have contributed to a reduction of local rainfall, as more, but smaller
-rain droplets were observed. Nair et al. (2011) reported from the Bunny
-Fence Experiment in Western Australia that local land use change altered
-the synoptic west coast trough dynamics and surface roughness, and this
-resulted in an observed rainfall decrease. Maximum temperatures were
-also found to be sensitive to land cover change in eastern Australia
-(McAlpine et al. 2007).
+The Rmarkdown documents contain the code used in the analysis. These
+documents including most of the data and additional scripts are
+accessible via Github:
+\url{https://github.com/WillemVervoort/RainfallLandcover/releases}.\\
+Please cite our paper if you plan to use any of the code.
 
-Overall the number of empirical studies analyzing changes to rainfall
-due to land cover change from observational data is limited. Most of the
-studies mentioned previously were either model simulations, or
-comparisons of modelled data with observations. This is because there
-are some fundamental experimental difficulties in both space (where does
-evaporated water reappear as rainfall?) and in time (how much time does
-it take for land cover change effects to appear or disappear?). In
-addition, in many areas across the globe, rainfall variability is
-related to a complex set of interactions, of which land use change might
-only be a minor component.
+\section{Summary of Data}\label{summary-of-data}
 
-Locally, there are two main sources that generate rainfall: moisture
-from advective atmospheric transport; and local evapotranspiration
-(Eltahir and Bras 1996; Bosilovich and Chern 2006; Dirmeyer, Brubaker,
-and DelSole 2009; Gimeno et al. 2010). The local evapotranspiration
-component is the component considered to be affected by land use change
-(Eltahir and Bras 1996). According to Trenberth (1999), the contribution
-of advective moisture partially depends on the availability of external
-moisture and atmospheric transport. On the longer time scale, such as
-monthly and annually, large scale atmospheric dynamics are affected by
-large scale climate drivers. For example, many studies have reported
-significant relationships between rainfall in large parts of Australia
-and the El Niño-Southern Oscillation (ENSO) (Verdon et al. 2004; Risbey
-et al. 2009; Speer, Leslie, and Fierro 2011). In contrast, local ET is
-determined by local land surface characteristics, which influence local
-scale atmospheric dynamics and hence the amount of rainfall, including
-contribution from both main sources.
-
-Although climate drivers demonstrate some capability to predict
-Australian rainfall, there is still a large amount of unexplained
-variance. Westra and Sharma (2010) pointed out that models based on
-global sea surface temperature anomalies can only predict up to 14.7\%
-of annual precipitation variance. Some of the remaining variance could
-be due to land surface processes as suggested in studies predicting
-local rainfall (e.g. Ma et al. 2011; Zeng et al. 2012; Pitman and Lorenz
-2016; Saha, Dirmeyer, and Chase 2016). However, most are based on
-modelling experiments and few empirical observational studies have been
-reported. However, Pitman et al. (2004) found a good match between
-observations and simulated rainfall changes in southwest Western
-Australia, forced by land cover change. Timbal and Arblaster (2006) were
-able to reproduce the rainfall decline in south west Australia by
-including land cover influence. In addition, local land use change might
-not be a primary, but is likely to be a secondary cause of rainfall
-change (Nicholls 2006).
-
-Therefore, the aim of this study is to use a statistical approach on
-rainfall data at regional scales to investigate the cause and effect
-relationship between land cover change and local rainfall, which is
-demonstrated in many modelling studies. More specifically, we
-hypothesize that a step change in land cover on the surface will cause a
-step change in the rainfall. To demonstrate this we study changes in
-observed rainfall over time at a Queensland and NSW/Victoria location
-where there are possible step changes in land cover change due to land
-clearing and bush fires. The methodology uses statistical approaches to
-identify changes in rainfall, which are subsequently associated with
-land cover change through spatial comparison.
-
-In this paper, after this section (the introduction), section 2 covers
-the case study areas and the observed land use change. Section 3
-describes the data used in the study in more detail. Section 4 details
-the statistical methods and the underlying assumptions related to the
-modelling approach, Section 5 gives the results, which are further
-discussed in section 6 and finally section 7 offers the conclusions.
-
-\section{Study regions and tree cover
-change}\label{study-regions-and-tree-cover-change}
-
-In Australia, significant tree cover change has mainly occurred in the
-north east and south east of the continent, as well as in the southwest
-of Western Australia. According to the National Dynamic Land Cover
-Dataset (DLCD) (Lymburner et al. 2010), most of these areas have
-experienced decreases in the Enhanced Vegetation Index (EVI) post 2000,
-as derived from satellite data. As an index for vegetation greenness,
-the decreasing values of EVI indicate lower biomass over time in the
-tree cover regions. The possible EVI reduction might be due to land
-clearing, bush fires or drought.
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{figures/map_selreg} \caption{Selected study regions are highlighted by red rectangles in the main map (the red rectangle in the insert indicates the location of the main map).
-The types of tree cover in 2008 from the DLCD product is shown at the background.
-In site 1 (the QLD region), the tree cover is mostly sparse.
-In site 2 (the NSW/VIC region), many areas have open or close forest where the tree cover is denser.}\label{fig:selreg}
-\end{figure}
-
-Two regions were selected where significant tree cover change since 2000
-was reported. The first region is located in south central Queensland
-(QLD) partly covering the north of the Murray Darling Basin (MDB) (site
-1 in Figure \ref{fig:selreg}). High rates of land clearing have been
-reported in this region during the early 2000s (Department of Natural
-Resources and Water 2007). The second study region is located at the
-border of New South Wales and Victoria (NSW/VIC), and includes the Snowy
-Mountain ranges (site 2 in Figure \ref{fig:selreg}). Severe bush fires
-occurred in this area in early 2003 (see Figure \ref{fig:bushfire}). The
-2003 bush fires were the largest and the worst in the last 60 years (The
-State Government of Victoria 2011). Two thirds of Kosciuszko national
-park was heavily burned and regrowth was reported to be slow due to
-drought and cold conditions (ABC News 2003), and the type of species in
-this region. However, in the longer term, after an early high
-transpiration period a recovery of pre-fire evapotranspiration would be
-expected (Kuczera 1987). For the purpose of this study, significant tree
-cover loss has happened in both study areas in the last decade, either
-permanently or temporarily.
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{figures/bushfire_nswvic} \caption{Location of bushfires occurring in January 2003, in and around the NSW/VIC study region, as shown by the red pixels.
-The map shows the large area in the Kosciuszko national park that has been burned.
-Some locations in the southwest of the Australian Capital Territory (ACT) have also experienced intensive bushfires.}\label{fig:bushfire}
-\end{figure}
-
-The two regions have different characteristics. The QLD region is
-partially grassland and subtropical, while the NSW/VIC region is mainly
-within the temperate zone, under the Köppen classification. According to
-Australian Bureau of Meteorology (BoM), the NSW/VIC region receives 1000
-- 2000 mm rainfall annually, which is more than double the annual
-rainfall in the QLD region. Evapotranspiration is similar in both
-regions. Marine moisture and orographic effects are likely to be the
-main contributors to rainfall in the southeast mountain areas of the
-NSW/VIC region.
-
-The land use and land cover characteristics in the two regions are also
-different. In the Queensland region, the tree cover is sparse over most
-of the area. The MODIS satellite tree cover data (discussed in more
-detail in section 3) shows that tree cover in this region is generally
-below 20\% of total ground area. Grazing is the main activity in this
-region, with over 90\% of land used by the grazing industry (ABARES
-2010). Our starting assumption is that the main cause of the EVI decline
-over large part of the region is due to land clearing. Tree cover has
-been cleared at a massive scale over the last decade, especially during
-2002 - 2004. The reports from the Queensland Statewide Land Cover and
-Trees Study (SLATS) (e.g. Department of Natural Resources and Mines
-2005; Department of Science, Information Technology and Innovation 2017)
-were used to investigate the time and location of the land clearing in
-the QLD region.
-
-The Kosciuszko national park is within the NSW/VIC region. Here tree
-cover is denser with open or even closed forest (the tree cover
-distribution is bimodal at 10 - 20\% and 60 - 70\%). The dominant
-species in the alpine area are Snow Gum and large stand species such as
-Alpine Ash and Mountain Gum in the sub-alpine area. These trees can
-reach a great height but they take long time to grow. For example,
-Alpine Ash would need about 20 years to mature. Although land clearing
-is not a major issue in this region, it is vulnerable to fires and
-drought. The MODIS burned area product, MCD45A1 (Roy, Lewis, and Justice
-2002; Roy et al. 2005; Roy et al. 2008), was used to locate bush fires
-areas in the NSW/VIC region, with a grid resolution of 500 m. MCD45A1
-provides monthly burning information on all pixels, which helps to
-pinpoint an abrupt event.
-
-Due to the difference in nature of the land cover change in the two
-regions, the post-change vegetation status is hypothesized to be
-different as well (see Figure \ref{fig:figure3-tc-simple}).
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{figures/tc_simple} \caption{The expected evolution of the land surface after trees have been removed in (a) the QLD region and (b) the NSW/VIC region.}\label{fig:figure3-tc-simple}
-\end{figure}
-
-The overall hypothesis is that the effect of 2003 - 2004 land clearings
-in the QLD region and the 2003 bush fires in the NSW/VIC region cause a
-step change in the local rainfall. The actual tree cover change during
-this time at the pixel level was derived from the 15-year MODIS data
-(discussed below). The difference of tree cover before and after the
-land disturbance was tested using a Student's t-test. As the length of
-the tree cover data is shorter than available rainfall data, earlier
-land clearing in the QLD region cannot be identified spatially, hence
-they are excluded from the analysis.
-
-\hypertarget{Data}{\section{Data}\label{Data}}
-
-Several land surface data sets were used in this study. The main one was
-the MOD44B product Global Vegetation Continuous Field data set (version
-5). This data set provides estimates of percent tree cover (percentage
-of ground surface covered by trees) at a grid resolution of 250 m
-(Townshend et al. 2011), which is finer than the earlier mentioned
-burned product MCD45A1. The data set is available on an annual time
-interval for the study period of 2000 - 2015. The tree cover data was
-produced from 16-day Terra MODIS Land Surface Reflectance data and Land
-Surface Temperature (Townshend et al. 2011). The National Dynamic Land
-Cover Dataset (DLCD) (Lymburner et al. 2010) from the Australian
-Collaborative Land Use Mapping Program (ACLUMP) was used to verify the
-trend of vegetation cover change calculated from the previous data set.
-This data set, developed by Geoscience Australia and Australian Bureau
-of Agricultural and Resource Economics and Sciences (ABARES), is the
-first nationally consistent and thematically comprehensive land cover
-reference for Australia. The DLCD is based on the 16-day Enhanced
-Vegetation Index (EVI), again from the MODIS satellite, between April
-2000 and April 2015. It also has a grid resolution of 250 m. The data
-set provides information on the final land cover types (as in 2015) and
-estimated trend of EVI statistics (annual mean, maximum and minimum).
-
-The SILO Rainfall product data for Australia was used (Jeffrey et al.
-2001) (available online at \url{https://silo.longpaddock.qld.gov.au/}))
-. The data has been projected onto a national 0.05\textdegree  by
-0.05\textdegree  grid (approximately 5 km by 5 km). This gridded data
-set was generated from station observations using spline interpolation
-and kriging (Jeffrey et al. 2001). The data has been compared to other
-gridded products and observed data and is generally of high quality
-(Tozer, Kiem, and Verdon-Kidd 2009; Tozer, Kiem, and Verdon-Kidd 2012).
-The data is available on a daily and monthly basis from 1889 to current.
-Here a subset of 36 years (1979 - 2015) was used. The study was
-conducted on monthly data, as a land cover change effect on annual
-rainfall might be negligible, but can often be significant in particular
-months or seasons (e.g. Otterman et al. 1990; Gaertner et al. 2001;
-Semazzi and Song 2001; Oleson et al. 2004; Deo et al. 2009).
-
-Large scale climate drivers are represented by various climatic indices.
-The Southern Oscillation Index (SOI) is generally regarded as a good
-predictor of Australian rainfall (Risbey et al. 2009; Chowdhury and
-Beecham 2010; Westra and Sharma 2010), but its skill is weaker in some
-parts of Australia. For example the Southern Annular Mode (SAM) is found
-to be more important than ENSO in south Western Australia (Meneghini,
-Simmonds, and Smith 2007). The testing of the suitability of each index
-for the regions of interest is described in a later section. The
-following climate indices were used as candidate predictors for local
-rainfall.
-
-\begin{itemize}
-\tightlist
-\item
-  Southern Oscillation Index (SOI). The Troup version of the monthly SOI
-  series used in this study was obtained from BoM (available online at
-  \url{http://www.bom.gov.au/climate/current/soihtm1.shtml}).\\
-\item
-  Eastern, East Central and Central Tropical Pacific Sea Surface
-  Temperatures (NINO 3, NINO 3.4 and NINO 4). Monthly SST anomalies are
-  available from IRI/LDEO data library and the extended NINO data set is
-  used (available online at
-  \url{http://iridl.ldeo.columbia.edu/SOURCES/.Indices/.nino/.EXTENDED/}).\\
-\item
-  Pacific Decadal Oscillation (PDO). The Pacific Decadal Oscillation is
-  the leading principal component of monthly SST anomaly in the North
-  Pacific Ocean.. The monthly PDO series was provided by JISAO (Joint
-  Institute for the Study of the Atmosphere and Ocean, University of
-  Washington) (available online at
-  \url{http://jisao.washington.edu/pdo/PDO.latest}).\\
-\item
-  Indian Ocean Dipole (IOD). The Indian Ocean dipole is commonly
-  measured by the difference between SST anomaly in the western (50 -
-  70\textdegree E and 10\textdegree S-10\textdegree N) and eastern (90 -
-  110\textdegree E and 0 - 10\textdegree S) equatorial India Ocean (Saji
-  et al. 1999). Monthly IOD was obtained from JAMSTEC (the Japan Agency
-  for Marine-Earth Science and Technology) (available online at
-  \url{http://www.jamstec.go.jp/frcgc/research/d1/iod/DATA/dmi.monthly.txt}).
-\end{itemize}
-
-\section{Statistical method}\label{statistical-method}
-
-As an initial analysis a simple boxplot and t-test is used to analyse
-whether there is a significant change in tree cover in time, and before
-and after the suspected change in the regions.
-
-To assess the actual causal relationship between the tree cover and the
-rainfall a flexible regression model is applied (discussed in detail
-below). A step change is not directly obvious in the time series of the
-rainfall anomalies (Figure \ref{fig:ts-mean}) for both regions, even
-though the data is deseasonalised and detrended. In this study we apply
-different statistical methods to analyse the effect of tree cover change
-on rainfall. Both methods make use of a regression model to remove
-year-on-year variability in rainfall to strengthen the tree cover change
-signal.
-
-In the first method, the tree cover change is implemented as a factor
-variable in the regression model, and the significance of this variable
-is tested. In the second method, a rank sum test (step trend test), is
-applied to the regression model residuals after effects of other major
-factors were removed. This assumes that after removal of all climate,
-long term linear trend and seasonal variation, the vegetation cover
-change is the only factor explaining the non-random pattern in the
-rainfall residuals.
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{figures/Rainfall_resid} \caption{The deseasonalised and detrended rainfall over the 30 years period in (a) the QLD region and (b) the NSW/VIC region.
-The vertical red lines indicate the year of 2003, in which the studied land cover changes occurred.
-A change in the time series data is not obvious before and after the land cover changes.}\label{fig:ts-mean}
-\end{figure}
-
-\subsection{Regression model}\label{reg_model}
-
-As highlighted in the introduction, the Australian climate is influenced
-by sea surface temperatures in the tropical Pacific and Indian Oceans,
-as well as pressure systems in the Southern Ocean (BoM 2012b). Risbey et
-al. (2009) compared five large-scale drivers, including ENSO (measured
-by SOI and the Tropical Pacific Sea surface temperatures (SSTs)), IOD,
-SAM, MJO (Madden-Julian oscillation) and blocking, in relation to
-Australian rainfall variability. The MJO is a large scale
-eastward-propagating wave-like disturbance located around equatorial
-latitudes (Risbey et al. 2009). They identified SOI as the most
-important index among all climate indices tested for broad parts of
-Australia (including QLD and NSW/VIC) in almost any season. In this
-study, four climate indices were selected from the main climatic
-indicators (see section \protect\hyperlink{Data}{Data}) and used as the
-explanatory variables in the model for each study region. A further
-complicating factor is the influence of the ``millenium drought'' over
-the study period and in particular the change to wet conditions in 2010
-- 2011 (Dijk and Viney 2013). Therefore, the spatially averaged monthly
-rainfall in the Murray Darling Basin (MDB, downloaded from
-\href{http://www.bom.gov.au/web01/ncc/www/cli_chg/timeseries/rain/allmonths/mdb/latest.txt}{the
-Bureau of Meteorology}) was used to explain the year-on-year variation
-in the rainfall in the regions. As climate alone are unlikely able to
-fully explain the temporal variation in rainfall in the regions,
-including the drought (Dijk and Viney 2013; Westra and Sharma 2010),
-this variable was included. Since both regions at least partly overlap
-with the MDB, the average rainfall for the entire basin was assumed to
-be a useful explaining variable.
-
-Correlations between rainfall and each climate index were analysed.
-Rainfall in each study region was first deseasonalised and detrended
-using the seasonal decomposition function \texttt{ds} in the package
-\texttt{deseasonalise} in R (R Core Team, 2018). Using detrended data
-gives a better indication of the underlying correlation by removing the
-underlying correlation in the data (Smith and Timbal 2012). The
-cross-correlations between the deseasonalised and detrended rainfall and
-the climatic indices were tested using the Pearson's product moment
-correlation method, assuming the relationships are linear. Although the
-optimal technique for exploring the correlation with each index could be
-different as described in Risbey et al. (2009), the Pearson's method was
-applied to all indices for consistency. Because the PDO describes the
-multi-decadal SST with lower frequency (MacDonald and Case 2005;
-Zanchettin et al. 2008; Kamruzzaman, Beecham, and Metcalfe 2011),
-instead of 37-year rainfall data, a longer period (115 years, from 1900
-to 2015) was used to estimate the correlation with PDO, up to lag 24.
-For the other indices, the 37-year data was used.
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{figures/cor_qld} \caption{Cross-correlation of six climate indices and rainfall in QLD study region.
-For the PDO analysis, 108-year rainfall data (1900 - 2008) are used.
-Otherwise, 36-year rainfall data are used.
-The correlation with NINO 3 is not shown as it is very similar to but weaker than for NINO 3.4.}\label{fig:cor-rain-qld}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{figures/cor_nswvic} \caption{Cross-correlation of six climate indices and rainfall in NSW/VIC study region.}\label{fig:cor-rain-nsw}
-\end{figure}
-
-Based on the correlation between the climatic indices and rainfall in
-the regions (as shown in Figure \ref{fig:cor-rain-qld} and Figure
-\ref{fig:cor-rain-nsw}), it can be concluded that:
-
-\begin{itemize}
-  \setlength{\itemsep}{0cm}
-  \setlength{\parskip}{0cm}
-  \item In QLD, the correlation between rainfall and SOI at zero time lags is the strongest across all indices, outweighing the other ENSO indicators. IOD and PDO have a weak influence in QLD.
-  \item In NSW/VIC, again the SOI has the strongest correlation with rainfall, followed by the IOD. For both, the strongest correlations occur at the zero time lags. 
-  \item In both cases PDO had the weakest correlations, and this factor was not further considered as a predictor.
-  \item In general, for the better correlated indices, strongest correlations occured at zero time lags.
-\end{itemize}
-
-The above findings are consistent with previous studies. Although some
-indices are serially correlated with rainfall up to several months, the
-lag zero events have the greatest correlation coefficients. Using
-multiple climatic index series was generally found most useful in
-rainfall prediction (e.g. Risbey et al. 2009; Kamruzzaman, Beecham, and
-Metcalfe 2011).
-
-Rainfall in Australia shows strong seasonal patterns (Holper 2011;
-Australian Bureau of Statistics 2012). For example, the north part of
-the country is summer rainfall dominant with a dry winter, while most of
-the southern part has a winter rainfall regime. This character is driven
-by the movement of subtropical high pressure systems which dominate the
-Australian climate (BoM 2012a). These characteristic summer and winter
-rainfall patterns mean that the seasonal component of rainfall has a
-periodic pattern which should be included in the model.
-
-Long term trends in the regional rainfall in some parts of Australia are
-significant (Hughes 2003; Gallant, Hennessy, and Risbey 2007; Chowdhury
-and Beecham 2010). In the northern and eastern parts of the continent,
-increasing rainfall is reported over the last century (Hughes 2003). The
-presence of such long term trends may be confused with the outcome of a
-step change in rainfall. As a result a linear trend term was implemented
-in the model to remove any long term effect.
-
-We assumed all the factors are additive smooth components in determining
-rainfall following Kamruzzaman, Beecham, and Metcalfe (2011). In this
-case, the rainfall model is a generalised additive model (GAM) (Hastie
-and Tibshirani 1986) with a log link function \texttt{g()} and assuming
-the residuals are gamma distributed (see Figure \ref{fig:hist-rain}).
-This means all predictors are modelled as smooth functions, in this case
-using the shrinkage version of the cubic regression splines (Wood 2011).
-All splines were limited to 3 knots in flexibility (Wood 2011), to
-reduce the risk of overfitting. \vspace{0.5cm}
-
-\begin{equation}
-\begin{array}{lll}
-g(E(\mathbf{R}_r)) = &\beta_0 + s_1(\mathbf{MDB_{monthlyRain}} + s_2(\mathbf{SOI}) + s_3(\mathbf{IOD}) + \\
-&s_4(\mathbf{Nino3.4}) + s_5(\mathbf{Nino4}) + s_6(\mathbf{Season}) + \\
-           &\beta_1\mathbf{Trend} + \boldsymbol{\epsilon}_r
-\end{array}
-\label{eq:model}
-\end{equation}
-
-The bold letters represent the time series vectors. The region is
-indicated by \(r\), while \(\beta_u\) (\(_u\)=0, 1) are the fitted
-coefficients in the model. \(s_v\) (\(_v\)=1, 2, 3,\ldots{}) are the
-smooth penalized cubic regression spline functions on the climatic
-indices and the season. MDB\textsubscript{monthlyRain} is the spatially
-averaged monthly rainfall across the Murray Darling Basin. Apart from
-dropping PDO as a predictor, all other climate indices were included,
-allowing the model to select the appropriate predictors.
-
-A possible linear long term trend in the rainfall data is modelled by
-\textbf{Trend} = 1,2,3\ldots{}n, where n is the total number of months
-in the time series. \textbf{Season} is the seasonal component. The
-climatic terms are also modelled with smooth functions. The effect of
-large scale drivers on Australian rainfall is more likely to be seasonal
-(Murphy and Timbal 2008; Schepen, Wang, and Robertson 2012), and the
-smooth spline function is more flexible in reproducing the variability
-in impacts of the climatic indices.
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{figures/hist_rainfall} \caption{Distribution of monthly rainfall in (a) QLD and (b) NSW/VIC.
-Using a Kolmogorov-Smirnov test with shape = 1 and 2.4  for QLD and NSW/VIC respectively, rainfall in both regions can be modelled as a gamma distribution.}\label{fig:hist-rain}
-\end{figure}
-
-\subsection{Tree cover change as factor variable}
-
-One of the main difficulties in empirical observation studies related to
-the effect of land cover change on rainfall is the lack of continuous
-monitoring of land surface variables. Moreover, no specific variable can
-possibly be defined that can clearly represent the land surface process.
-Given the lack of a full picture of the land surface process, a factor
-variable was used in the regression model to represent the abrupt land
-surface change (see Equation \eqref{eq:model}). The change could be a
-result of either land clearing or bush fires as long as it is permanent
-or takes a long time to recover. As indicated we approached the problem
-in two different models ways.
-
-In the first method, the tree cover change was used as a predictor in
-the regression model, represented by a factor variable \textbf{LC}. The
-significance of the coefficient of \textbf{LC}, denoted as \(\beta'_5\)
-in Equation \eqref{eq:model}, can be determined by a ratio test.
-
-\begin{equation}
-  \mbox{LC} = \left\{
-  \begin{array}{ll}
-     \mbox{Trees} \\
-     \mbox{Removed}
-  \end{array} \right.\\
-\label{eq:dummyC}
-\end{equation}
-
-Therefore in both regions, land cover is ``trees'' for the period before
-land cover change and ``removed'' for the period after the change. Here
-we simply assumed that vegetation cover change has occurred on every
-pixel. The remaining term \(\epsilon_r\) is the amount of rainfall that
-is attributed to other unspecified factors and random errors. Hence the
-regression model becomes \vspace{0.5cm}
-
-\begin{equation}
-\begin{array}{lll}
-g(E(\mathbf{R}_r)) = &\beta'_0 + s'_1(\mathbf{MDB_{monthlyRain}} + \\
-  &s'_2(\mathbf{SOI}) + s'_3(\mathbf{IOD}) + \\
-  &s'_4(\mathbf{Nino3.4}) + s'_5(\mathbf{Nino4}) + s'_6\mathbf{Season} + \\
-  &\beta'_1\mathbf{Trend} + \beta'_2\mathbf{LC} + \boldsymbol{\epsilon'}_r
-\end{array}
-\label{eq:model2}
-\end{equation}
-
-One of the difficulties is to point an exact time to the changes in the
-vegetation cover in the two regions. In the QLD region, no exact time
-can be assigned to the land clearing. According to the SLATS reports,
-the most substantial clearing occurred between 2003 - 2004 . However,
-the information on the change in type of land cover during the time
-period is missing. Therefore, four scenarios were initially tested in
-the analysis, however there was no real difference between these
-scenarios. In the NSW/VIC region, severe bush fires were reported in
-early January 2003. Hence the ``tree'' cover state was up to December
-2002 then it was changed to ``removed'' state from January 2003. As a
-starting date, the regression model was run from 1979 for both regions.
-
-\subsection{Step trend test}\label{step-trend-test}
-
-To support the regression analysis, a Mann-Whitney Rank-Sum step trend
-test was used to detect changes in rainfall as a result of vegetation
-cover change. This specific nonparametric statistical test was modified
-from the Mann-Whitney U test by Hirsch and Gilroy (1985) and can
-identify a step change in data which is cross-correlated. The gridded
-rainfall dataset used in this study has a high spatial correlation
-between neighbouring pixels due to the underlying interpolation method.
-The advantages of using the Rank-Sum test are: (1) it does not depend on
-assumptions of the data distribution; (2) it is not restricted to
-datasets with no missing data; (3) it is robust and not as easily
-influenced by outliers and negative numbers (Hirsch and Gilroy 1985).
-However, the test has less power than parametric tests.
-
-The rainfall residuals from the regression model in Equation
-\eqref{eq:model2} were used for the step trend test. To detect a step
-change, using deseasonalised and detrended data is important (Hirsch and
-Gilroy 1985). Furthermore, as rainfall can only be partially attributed
-to local sources and conditions, other effects are introduced by large
-scale dynamics and changes in climatic factors. The assumption is that
-the regression model should remove these effects, and additionally
-deseasonalise and detrend the rainfall data. As a result, the local
-landuse effects are amplified in the remaining residuals. The test,
-described in the following sub section, subsequently associates trends
-in the rainfall residuals with tree cover changes.
-
-\subsubsection{Mann-Whitney rank-sum
-statistic}\label{mann-whitney-rank-sum-statistic}
-
-As indicated, the step trend test is a modified version of the
-Mann-Whitney U statistic (Hirsch and Gilroy 1985). As a nonparametric
-rank-based test, the Mann-Whitney test does not use the exact values of
-rainfall but depends on the ranks of the data. For each month, rainfall
-residuals of each year were ranked in an ascending order. The ranking of
-January rainfall in a sample pixel k in QLD is illustrated below:
-
-\begin{table}[t]
-
-\caption{\label{tab:exampleRank}Example of ranking rainfall residuals}
-\centering
-\begin{tabular}{rrr}
-\toprule
-Year & Rainfall residuals & Rank \$R'\_\{1k\}\$\\
-\midrule
-1998 & -0.3 & 6\\
-1999 & -60.9 & 2\\
-2000 & -16.1 & 4\\
-2001 & -71.7 & 1\\
-2002 & 111.1 & 7\\
-\addlinespace
-2005 & -7.2 & 5\\
-2006 & -60.5 & 3\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\noindent Therefore, the smallest or most negative value has rank 1 and
-the largest value has the maximum rank.
-
-The before and after period in the data formed two groups of samples.
-The split point of the two periods was based on the timing of the
-vegetation cover changes. In the QLD region, changes occurred anytime
-during 2003 and 2004. In contrast to the previous method, the time
-period covering the land cover change was excluded, as the nonparametric
-test allows missing data. Hirsch and Gilroy (1985) also pointed out that
-the power of the test is higher if the data of the change period is
-ignored. Hence 2003 and 2004 were excluded from the analysis. As a
-result, the post change period was 2005 - 2015 for the Queensland
-location.
-
-In the case of NSW/VIC, the bushfires broke out in early January 2003.
-The change was within a relatively short period of the year. Therefore
-the post change period in this region still started in January 2003.
-Following Hirsch and Gilroy (1985), the pre change period was set to
-five years (1998 - 2002) in both regions. The length of the post change
-period is also difficult for the NSW/VIC region, as the regrowth would
-at some point have impacted the local effects.
-
-The rank of rainfall in month j year i in pixel k is denoted as
-\(R'_{ijk}\). The sum of ranks of rainfall in month j in pixel k before
-the known intervention is:
-
-\begin{equation}
-  W_{jk} = \sum_{i=1}^{n_1}R'_{ijk}.
-  \label{eq:Wj}
-\end{equation}
-
-\(n_1\) is the number of years before the land cover change. The
-expected value of \(W_{jk}\) is
-
-\begin{equation}
-  \mu_w=n_1(n_1+n_2+1)/2
-\end{equation}
-
-\(n_2\) is the number of years after the change. Hence the expected
-value of the rank sum before the intervention is the same for all months
-and all pixels. The sum of ranks for the whole time period is fixed, as
-\((n_1+n_2)(n_1+n_2+1)/2\). In this study, since there are only two
-groups (before and after), knowing the rank-sum of one group is the same
-as knowing the rank-sum of the other group. If the rainfall data is
-temporally and spatially independent, the variance of \(W_{jk}\) is
-
-\begin{equation}
-  \sigma^2_w = n_1\cdot n_2(n_1+n_2+1)/m
-\end{equation}
-
-where m is the number of months which is 12 in the case of a full year.
-
-\subsubsection{Step trend test}\label{step-trend-test-1}
-
-Instead of completing the Mann-Whitney U-test, Hirsch and Gilroy (1985)
-applied a standard normal Z test to the rank-sum statistics. As
-highlighted, this modified test accounts for serial temporal
-autocorrelation and spatial cross correlation in the data. In the case
-here, the deseasonalised and detrended data shows little autocorrelation
-in the time series but possesses strong cross correlation between
-neighbouring pixels, i.e. \(R>0.99\).
-
-The sum of \(W_{jk}\) for a block of \(ns\) pixels over the whole year,
-\(\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk}\), has mean
-
-\begin{equation}
-  E(\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk})=12\cdot ns\cdot\mu_W
-\end{equation}
-
-and variance
-
-\begin{equation}
-  Var(\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk})=\sum_{j=1}^{12}\sum_{k=1}^{ns}\sum_{h=1}^{ns}C(W_{jk},W_{jh}).
-\end{equation}
-
-\(C(W_{jk},W_{jh})\) is the covariance of the W statistics between pixel
-k and pixel h in month j. When \(k=h\), \(C(W_{jk},W_{jh})=\sigma^2_w\).
-When \(k\neq h\),
-
-\begin{equation}
-  C(W_{jk},W_{jh})=\sigma^2_w r(R_k,R_h)
-\end{equation}
-
-where \(r(R_k,R_h)\) is the product moment correlation coefficient of
-the concurrent ranks in pixel k and h. Here \(r\) is calculated on the
-full time series in each pixel. In the analysis, the test was applied to
-a square block of four pixels each time. As argued by Hirsch and Gilroy
-(1985), \(ns=4\) is the most optimal solution to balance the cost and
-the gain in the test power.
-
-The statistic of the step trend test is then defined as
-
-\begin{equation}
-  Z'=\frac{\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk}-12\cdot ns\cdot\mu_w}{\sqrt{Var(\sum_{j=1}^{12}\sum_{k=1}^{ns}W_{jk})}}.
-  \label{eq:Z}
-\end{equation}
-
-The above statistic is written for a 12 month period. By changing the
-value 12, it can also be used to test seasonal rainfall change or for
-other customized periods.
-
-\begin{table}[t]
-
-\caption{\label{tab:Zscore}The interpretation of Z' score in the step trend test}
-\centering
-\begin{tabular}{l}
-\toprule
- \\
-\midrule
-\$Z'>0\$  and rainfall decreases after change\\
-\$Z'<0\$ and rainfall increases post change\\
-\$Z'=0\$ and rainfall does not change\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-The null hypothesis (\(H_0\)) in this study is that there was no change
-in rainfall due to land surface intervention. The results of the step
-trend test can be interpreted according to the sign of the Z' score (see
-Table \ref{tab:Zscore}, Chapter 23, P887 (Hipel and McLeod 1994)). Z' is
-normally distributed similar to the standard normal statistics Z. Hence
-it can be compared to a standard normal distribution to determine the p
-value.
-
-\section{Results}\label{results}
-
-\subsection{Tree cover change}\label{tree-cover-change}
-
-The pixels, where the tree cover change based on the MOD44B data was
-significant (\(p \leq 0.05\)) in each study region, are shown in Figure
-\ref{fig:tctrend} for the NSW/VIC region (left panel) and the QLD region
-(right panel). In the QLD region there actually has generally been an
-increase in tree cover after the clearing of native vegetation stopped.
-In the NSW/VIC region, much of the tree loss between 2002 and 2003 was
-concentrated in the Snowy Mountains close to the border of NSW and VIC,
-which is also evident in the figure. Tree cover loss occurred in large
-parts of the QLD region between 2002 and 2005, but this tree loss was
-spatially less concentrated.
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{figures/tc_figs} \caption{The maps show significant changes in tree cover identified from the MOD44B data between 2003 and 2015 in the NSW/VIC region (right) and the Qld region (left).
-The amount of change was calculated as the difference in tree cover before and after the specified land cover intervention and it is shown as the percentage of the ground area.
-Green colour indicates an increase in tree cover, while red colour indicates a decrease in tree cover.}\label{fig:tctrend}
-\end{figure}
-
-\subsection{Regression Model and significance of Vegetation Cover
-Changes}\label{regression-model-and-significance-of-vegetation-cover-changes}
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{figures/rsq_fullmodel} \caption{The distribution of the variance explained (adjusted $r^2$ ) by the full regression model}\label{fig:rsqmodel}
-\end{figure}
-
-Generally the regression model only explains a limited amount of the
-rainfall variability (Figure @reg(fig:rsqmodel)). The model in Equation
-\eqref{eq:model2} accounts for an average around 33\% of the variation in
-the NSW/VIC region and about 50\% of the variation in the QLD region as
-indicated by the adjusted \(r^2\). The adjusted \(r^2\) is the
-coefficient of determination, a measurement of the amount of variability
-predicted by the model adjusting for the number of explanatory terms.
-The residual analysis shows that the assumptions of the regression model
-are generally met (Figure \ref{fig:residuals}). The standardised
-residual plots, however, show some funnelling for the NSW/VIC regions,
-suggesting non constant variance. The residual patterns are consistent
-for all pixels within each region.
-
-In terms of predictors of the local rainfall in the model, logically,
-the average rainfall across the larger Murray Darling Basin is highly
-significant. This confirms that this variable is a good reflection of
-the year on year variability in the rainfall. The model also confirms
-the importance of the climate drivers and the seasonality in Australian
-rainfall, as many of these variables were significant. Even at the grid
-level, the seasons and several of the climatic indices were significant
-(\(p \leq 0.05\)) everywhere in both regions. The climate drivers (at
-lag zero) accounted, on average, for 6.5\% of the rainfall variability
-in both the QLD region and the NSW/VIC region (see Figure \ref{fig:rsq}
-for the distribution of adjusted \(r^2\) in these two regions). These
-figures are within the upper bound of seasonal rainfall predictability
-by a SST anomaly field reported by Westra and Sharma (2010).
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{figures/gam_check} \caption{The residual analysis of a sample pixel in the QLD region (top) and NSW/VIC region (bottom).}\label{fig:residuals}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=0.7\linewidth,height=0.7\textheight]{figures/soi_explhistogram} \caption{The performance of the regression model if rainfall is only modelled by the climate drivers.
-It shows the percentage of rainfall variability that can be explained by the climate drivers for the Qld and NSW/VIC region}\label{fig:rsq}
-\end{figure}
-
-There were generally no statistically significant long term trends in
-both regions. However, this result might not prove or disprove the
-existence a long term trend in rainfall. The overall time period is
-fairly short (Koutsoyiannis 2006) and more pixels in NSW/VIC could
-indicate a significant step change if the long term trend effect is not
-removed by the model. As trend free data is an important requirement for
-the step trend test, the trend term was kept in the regression model to
-ensure the detection of step change was not due to a possible long term
-trend.
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{figures/Cp_30yrs} \caption{The spatial distribution of significance of land cover step change variable in the GAM model predicting changes in rainfall in the both regions.
-The left graph relates to the Qld region, while the right hand plot reflects the NSW/VIC region.
-The outlines of relevant Australian states and the Murray Darling Basin are indicated in grey.
-The p value reported is for the land cover variable in the model.}\label{fig:LCp}
-\end{figure}
-
-The land cover variable in the model (Equation \eqref{eq:model2}) aims to
-identify a step change in the rainfall before and after the observed
-change in land cover. The variable was mainly significant
-(\(p \leq 0.05\)) for the rainfall estimates in some areas in NSW/VIC,
-as shown in Figure \ref{fig:LCp} (right panel). However, the number of
-pixels where the landcover variable was significant was much greater in
-NSW/VIC compared to the QLD region. There was also some relationship
-between the areas of bushfires in Figure \ref{fig:bushfire}. Only a very
-small area with a significant step change due to the land cover changes
-was found in rainfall in the QLD region (left panel)
-
-More generally, the model for NSW/VIC suggests that the land cover
-variable has a positive impact on rainfall in both regions. The fitted
-coefficients for the Land cover change variable were consistently
-positive for the ``tree'' part of the series. It implies that rainfall
-was higher when the surface was covered by trees.
-
-\subsection{Step Trend Test}\label{step-trend-test-2}
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{figures/step_new} \caption{Spatial distribution of the step trend test Z' statistics in the two study sites.
-The panel on the left is for the QLD region, while the panel of the right is for NSW/VIC.
-Warm colours (yellow, orange and red) are for positive Z' values which indicate decreasing rainfall trend due to the land surface change.
-Cold colours (light blue to blue) are for negative Z' values which indicate increasing rainfall trend.
-The deeper the colour, the more significant the statistic.}\label{fig:steptest30}
-\end{figure}
-
-The spatial step trend test based on the regression residuals without
-the landcover variable generates Z' scores. The results for the Qld
-region (left) and NSWVIC region (right) are shown in Figure
-\ref{fig:steptest30}. This figure provides two types of information: the
-sign and the significance level. The sign indicates the direction of the
-step change in the residuals of the regression, as listed in Table
-\ref{tab:Zscore}. In both regions there are areas of positive Z' values
-which imply a decrease in rainfall, but this is stronger in the NSW/VIC
-area than in the QLD area. For the NSW/VIC area there appears to be a
-reasonable relationship between the locations where changes in tree
-cover are observed (Figure \ref{fig:tctrend}) and the patterns in the Z'
-score. However, there is not necessarily a direct relationship as
-movement of air masses could mean that actual changes of rainfall are
-observed close by, but not necessarily exactly at areas with changes of
-landcover. There is once again a difference between the left panel and
-the right panel, indicating only a few significant of Z' scores in the
-QLD region wich matches the earlier significance in the landcover
-variable in the full regresion. In the QLD region, only 0.2\% of the
-pixels obtained a positive Z' score with p \(<\) 0.1. In the NSW/VIC
-region 3.7\% of pixels have a positive Z' score with p \(<\) 0.1. In
-general it is only a small proportion of both study regions.
-
-\begin{figure}
-\includegraphics[width=0.7\linewidth,height=0.7\textheight]{figures/ResidualBoxplotchange} \caption{Boxplots of annual rainfall residuals (estimated based on Equation 2 before and after the land cover intervention during 1979 - 2015 in the study regions.
-On average, the after period has a significantly lower annual rainfall residual in NSW/VIC, but a signifcantly higher annual rainfall residual in the Qld study area}\label{fig:meandiff}
-\end{figure}
-
-The rainfall regression residuals without the landcover variable for the
-two regions (before-change since 1979 and after-change) were also
-compared (Figure \ref{fig:meandiff}) using a simple t-test. From the
-boxplots it is difficult to see that there were significantly different
-(\(p < 0.05\)) mean rainfall residual values between the ``before'' and
-``after'' periods in both regions. For the Queensland locations, the
-mean monthly rainfall residuals were slightly lower (\(p < 0.05\)) after
-the change in landcover (\(~ 0.45 mm/month\)). For the NSW/VIC locations
-there was larger decrease in the mean monthly rainfall residual
-(\(p < 0.05\)) post change (\(~ 1.5 mm/month\)).
-
-The choice of the number of pixels averaged in the step trend test
-(\(ns\)) has some impact on the test results (Hirsch and Gilroy 1985).
-The cases of \(ns = 1\) and \(ns = 9\) were also tested. The results for
-\(ns = 1\) showed a lower number of significant pixels (at \(p< 0.10\))
-compared to the \(ns = 4\) test. The results from the \(ns = 9\) test
-was not different from the \(ns = 4\) test. We therefore decided to stay
-with \(ns = 4\). The power of the test does not change much after
-\(ns = 4\), as shown by Hirsch and Gilroy (1985).
-
-As part of the analysis, the ``field significance'' of the Z' score test
-was considered to improve the interpretion of the step change at
-regional scales from multiple local tests (Wilks 2006; Westra,
-Alexander, and Zwiers 2013). Here, the bootstrapping resampling method
-from Westra, Alexander, and Zwiers (2013) was used to evaluate the field
-significance. This means the spatial structure of the pixels was
-maintained, but the order of the years and months was changed by random
-resampling. For each resampling, the test statistic identifies the
-percentage of the pixels with a significant positive or negative step
-change for the step trend test. The test statistics on 1000 resampled
-replicates were used to develop the distribution of these percentage
-values, and the observed fraction of significant Z' scores (at
-\(p < 0.1\)) is plotted on the distribution (Figure \ref{fig:fieldsig}).
-The results support the earlier Z' score results. Only the percentage
-significant positive Z' scores, indicating a decrease in rainfall for
-the NSW/VIC region as a result in a decrease in tree cover, falls on the
-tail of the field significance distribution, while all the others are
-well within the distribution. This suggests that the percentage of
-significant positive Z' scores in the NSW/VIC region is least likely due
-to a unique series of rainfall years, but is most likely due to changes
-in landcover. In other words, in the NSW/VIC region it is most likely
-that the change in landcover (decrease in tree cover) caused a decrease
-in the rainfall.
-
-\begin{figure}
-\includegraphics[width=0.7\linewidth]{figures/bt4_hist} \caption{Field significance results showing the distribution of percentage of significant $Z'$ scores for the resampled series (blue bars), and the precentage significant $Z'$ scores for the original series of rainfall years (red circle)}\label{fig:fieldsig}
-\end{figure}
-
-\section{Discussion}\label{discussion}
-
-\begin{table}[t]
-
-\caption{\label{tab:summarytable}Summary table of all tests on the two regions }
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}{lll}
-\toprule
-Test & Qld & NSW VIC\\
-\midrule
-LC variable & Small area of significant pixels in the centre & Large area possibly aligning with bushfire affected area.\\
-t-test regression model residuals & Slightly lower mean residual ($p<0.05$) after change (0.5 mm per month) & Lower mean residual ($p<0.05$) after change (1.5 mm per month)\\
-Positive score & 0.2\% of pixels at $p<0.1$ & 3.7\% of pixel at $p<0.1$\\
-Field significance \% of scores & Both positive and negative \% within random distribution & Positive \% outside random distribution.\\
-\bottomrule
-\end{tabular}}
-\end{table}
-
-Overall the summary table of the results indicates that the effects of
-land clearing or bushfire on rainfall are consistent across the range of
-different statistical tests. In all cases, the data for the NSW/VIC
-region indicates the strongest evidence that the reduction in tree cover
-(the land cover change) resulted in a local decrease in rainfall. In
-contrast the data for the QLD region only indicates weak evidence that
-the loss of trees due to land clearing resulted in reduction in the
-rainfall.
-
-Generally, empirical studies on LCC-precipitation interaction are
-conducted within an area with known land surface intervention (e.g.
-Otterman et al. 1990; Durieux, Machado, and Laurent 2003; Negri et al.
-2004; Sato, Kimura, and Hasegawa 2007). However, these locations are
-rare and difficult to isolate from real landscape change. Modelling
-studies are abundant (e.g. Chagnon and Bras 2005; Pinto et al. 2009;
-Wang et al. 2009), but these are generally not directly linked to
-observed data. In this study we tested the effect of land cover change
-across a broad region, which included locations where changes were known
-to occur or have occured. The advantage of the current approach is that
-long time series of land cover data are not required. This is
-advantageous as this type of that is currently not always available.
-Furthermore, it does not assume a specific relationship between
-vegetation cover change and rainfall but allows the data to show this
-relationship, by applying the analysis to a broader area outside the
-boundary of the vegetation cover change. This approach is expected to
-provide a way to reduce the risk of a false positive paradox, by
-comparing results between areas with and without vegetation cover
-change.
-
-Overall the results suggest that at least for the NSW/VIC region that a
-decrease in tree cover causes a decrease in rainfall (Figure
-\ref{fig:tctrend}). However, there are several possible complicating
-factors in the data that require discussion.
-
-\subsection{Rainfall variability}\label{rainfall-variability}
-
-Rainfall in Australia is highly variable from year to year, and the time
-period in this study included a severe drought (Dijk and Viney 2013), as
-well as the drought breaking years 2010 and 2011. The purpose of using
-the regression model is to remove the year to year and month to month
-variability in rainfall and therefore strengthen the land cover signal
-in the residuals. We used the spatially averaged monthly rainfall
-timeseries for the larger Murray Darling Basin (Figure \ref{fig:selreg})
-to remove the general variability. However, overall the model does not
-explain more than 50\% (QLD) and 30\% (NSW/VIC) of the rainfall
-variability, and only around 7\% on average is due to the climate
-drivers (Figure \ref{fig:rsq}) . And while this is consistent with the
-literature e.g. Westra and Sharma (2010), this means some variation due
-to external factors could still be left in the residuals. Rainfall is
-generally considered a stochastic process (e.g. Fowler et al. 2005;
-Cowpertwait, Salinger, and Mullen 2009; Burton et al. 2010) and some of
-the variability could either be a different response to a combination of
-climate factors (as interactions were not tested in the model), or a
-non-stationary response to the climate drivers. The severe bushfires in
-2003 were also triggered by the extreme drought conditions during the
-millenium drought (Dijk and Viney 2013). Although the effect of the
-drought on the overall rainfall quantities has been accounted for in
-part by the model, a further delayed or cumulative effect of the drought
-could be feeding into the local land-atmosphere interaction. As a
-result, the rainfall feedback to the vegetation cover change could be
-weak under the dry conditions between 2001 and 2009, and this could have
-affected the result.
-
-\subsection{Vegetation dynamics}\label{vegetation-dynamics}
-
-The second possible effect is the dynamic nature of the vegetation
-clearing and recovery, especially for the QLD region. Although land
-clearing has occurred at a high rate and broad scale in Queensland
-(Department of Science, Information Technology and Innovation 2017), the
-clearing does not have a clear start and end point. QLD has a long
-history of land clearing. According to the series of SLATS reports on
-land cover changes in QLD released by the Queensland government, land
-clearing continued in and around the study region between 1988 - 2008.
-Major broad scale and high rate clearings occurred in 1999 - 2000 and
-2002 - 2004 (Figure \ref{fig:slat}). And even though there was a
-decrease in land clearing post 2005, it is difficult to define a clear
-cut change in the tree cover in this region. The specific vegetation
-class in the Queensland area is also well-known for rapid regrowth and
-``thickening'' in favourable conditions (Gowen and Bray 2016), and this
-could explain the increase in vegetation cover in 2015 in the Qld area
-(Figure \ref{fig:tctrend}). In particular the favourable rainfall years
-of 2010 and 2011 would have boosted regrowth, increasing
-evapotranspiration and therefore decreasing the effect of the rainfall
-change.
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{figures/slats} \caption{Woody vegetation clearing rate in QLD for the two major bioregions in the study area.
-The data were obtained from SLATS (2017), and clearly indicate a sharp decrease in the clearing rate after 2005.}\label{fig:slat}
-\end{figure}
-
-The different causes of vegetation cover change in these two regions
-could lead to different post-change characteristics. The trends in EVI
-in the QLD region are opposite to the NSW/VIC region, based on the DLCD
-data. This could also be due to the lower tree density in the QLD region
-compared to the NSW/VIC region before land surface interventions. In
-contrast, the significant bushfires (Figure \ref{fig:bushfire}) would
-have drastically reduced the vegetation cover and recovery was very slow
-in some areas within the NSW/VIC area. The persistent drought in the
-2000s (Howden 2012; Dijk and Viney 2013) delayed the regrowth of trees.
-Conversely, replacing tree cover with pasture and crops in the Qld area
-might only have a relatively subtle impact on the EVI.
-
-\subsection{Gridded monthly rainfall
-data}\label{gridded-monthly-rainfall-data}
-
-The rainfall data used in this study is gridded. This data set is robust
-and consistent over a long time series (from 1900 to current) and has a
-broad national wide coverage which can provide more information
-spatially (Jeffrey et al. 2001; Tozer, Kiem, and Verdon-Kidd 2009).
-However, high cross correlation between pixels, due to the interpolation
-method in this data set (Jeffrey et al. 2001), can also introduce
-spatial noise. In the step trend test the cross correlation was
-accounted for. However, some other methods are also available, which
-could be used to perform a comparative trial. For example, Narisma et
-al. (2007) applied a spatial Gaussian filter on a similar data set and
-used wavelet analysis to detect a step change in rainfall. High quality
-station data is another option to test whether the observed spatial
-pattern in the step trend test results was not due to the gridded data
-itself. Resampling methods, such as bootstrapping and permutation (used
-in this study) (Wilks 1997; Kundzewicz and Robson 2004; Westra,
-Alexander, and Zwiers 2013), can also be used to further assess the
-strength of the significance of results . While the gridded data set is
-most useful in regions with sparse rain gauge networks, such as in this
-study, it can actually reduce information where the rain gauge density
-is high (Jones, Wang, and Fawcett 2009). In the NSW/VIC region, the
-coverage of rainfall stations is more intensive, but they are mainly
-located in the valleys. As a result, the interpolated data might be a
-limited representation of the true local rainfall.
-
-\subsection{General approach}\label{general-approach}
-
-Parametric tests are generally more powerful than nonparametric test in
-detecting a trend, when the data is normally distributed (Onoz and
-Bayazit 2003; Kundzewicz and Robson 2004). As a non-parametric test, the
-step trend test has the advantage of being distribution free and having
-no restriction on missing data (Hirsch and Gilroy 1985). This is
-particularly useful in rainfall analysis since rainfall data is usually
-highly skewed. On the other hand, the disadvantages of non-parametric
-tests, such as being limited to hypothesis testing and being weaker in
-power, also hold for the step trend test (Whitley and Ball 2002).
-
-Overall, the current study provides a statistical data based approach,
-building on several lines of evidence to reject the null hypothesis (no
-step change in rainfall occurs as a result of tree cover loss) at least
-for the NSW/VIC region. Limited by the available data, the time frame
-under study included a long lasting drought period (Holper 2011; Dijk
-and Viney 2013). The strong impact of this prolonged drought might have
-suppressed the land-atmosphere interaction and modified the cause and
-effect relationship between rainfall and vegetation cover change. This
-could be one of the reasons that the land cover change effects on the
-local climate found in other studies (e.g. Görgen et al. 2006; McAlpine
-et al. 2007) do not appear in the QLD region. Possible future work could
-focus on a non-drought period, or using a different series of land cover
-data, possibly on another continent. The recent large bushfires on the
-US continent come to mind as a possible candidate study once sufficient
-post event data is collected. Wile the power of the test can be improved
-with the longer length of the post-intervention period (Hirsch and
-Gilroy 1985), the dynamic nature of vegetation regrowth in this case
-study also affects this effect. A better approach might be to build a
-global study that investigates multiple locations where drastic
-landcover changes have taken place, which would also remove some of the
-climate variability effects due to the larger sample size.
-
-\section{Conclusions}\label{conclusions}
-
-In this study, we present a statistical approach to identify the impact
-of a change in land cover on local rainfall.The results, based on
-gridded rainfall data found, that a reduction in vegetation cover is
-likely to have reduced local rainfall for a large area in NSW/VIC
-affected by bushfires. However, land clearing in QLD was unlikely to
-have reduced rainfall over the same time period.
-
-Drought may have had a pronounced impact on the land surface condition
-during the study period, also leading to significant reduction in
-vegetation cover and extreme events such as bushfires. The lack of
-rainfall and associated high temperatures may mask the impact on
-rainfall of a step change in the vegetation cover. Hence, the signal of
-Land Cover Change feedback on rainfall is probably weaker under such
-regional dry conditions, as the impact of Land Cover Change on rainfall
-is mainly through changes in moisture convergence (Görgen et al. 2006;
-Pitman and Hesse 2007).
-
-\section{appendix}\label{appendix}
-
-\emph{Summary of Data}
+This is a summary table of all the data used in the paper.
 
@@
 
+\section{Cross correlations Rainfall and Climate
+Indices}\label{cross-correlations-rainfall-and-climate-indices}
+
+These figures were moved to the supplementary data in the review
+process.
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{figures/cor_qld} \caption{Cross-correlation of six climate indices and rainfall in QLD study region.
+For the PDO analysis, 108-year rainfall data (1900 - 2008) are used.
+Otherwise, 36-year rainfall data are used.
+The correlation with NINO 3 is not shown as it is very similar to but weaker than for NINO 3.4.}\label{fig:cor-rain-qld}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{figures/cor_nswvic} \caption{Cross-correlation of six climate indices and rainfall in NSW/VIC study region.}\label{fig:cor-rain-nsw}
+\end{figure}
+
+Based on the correlation between the climatic indices and rainfall in
+the regions (as shown in Figure \ref{fig:cor-rain-qld} and Figure
+\ref{fig:cor-rain-nsw}), it can be concluded that:
+
+\begin{itemize}
+\tightlist
+\item
+  In QLD, the correlation between rainfall and SOI at zero time lags is
+  the strongest across all indices, outweighing the other ENSO
+  indicators. IOD and PDO have a weak influence in QLD.\\
+\item
+  In NSW/VIC, again the SOI has the strongest correlation with rainfall,
+  followed by the IOD. For both, the strongest correlations occur at the
+  zero time lags.\\
+\item
+  In both cases PDO had the weakest correlations, and this factor was
+  therefore dropped as a predictor.\\
+\item
+  In general, for the better correlated indices, strongest correlations
+  occured at zero time lags.
+\end{itemize}
+
+Although some indices are serially correlated with rainfall up to
+several months, the lag zero events have the greatest correlation
+coefficients. Furthermore, using multiple climatic index series was
+generally found most useful in rainfall prediction (e.g. Risbey et al.
+2009; Kamruzzaman, Beecham, and Metcalfe 2011).
+
+\newpage
+
+\section{Further figures}\label{further-figures}
+
+These two figures highlight the GAM residual analysis for a sample
+pixel. Residuals for other pixels were similar across the regions
+analysed.
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{figures/gam_check} \caption{The residual analysis of a sample pixel in the QLD region (top) and NSW/VIC region (bottom).}\label{fig:residuals}
+\end{figure}
+
+This figure indicates a boxplot of the annual rainfall residuals across
+the regions.
+
+\begin{figure}
+\includegraphics[width=0.7\linewidth]{figures/ResidualBoxplotchange} \caption{Boxplots of annual rainfall residuals (estimated based on Equation 2 before and after the land cover intervention during 1979 - 2015 in the study regions.
+On average, the after period has a significantly lower annual rainfall residual in NSW/VIC, but a significantly higher annual rainfall residual in the Qld study area}\label{fig:meandiff}
+\end{figure}
+
 \section*{References}\label{references}
@@
 \hypertarget{refs}{}
-\hypertarget{ref-ABARES2010}{}
-ABARES. 2010. ``Land Use of Australia, Version 4, 2005/2006.''
-
-\hypertarget{ref-ABC2003}{}
-ABC News. 2003. ``Kosciuszko Slow to Recover from Bushfires.''
-
-\hypertarget{ref-ABS2012}{}
-Australian Bureau of Statistics. 2012. ``Geography and Climate -
-Australia's Climate.''
-
-\hypertarget{ref-Ben-Gai1998}{}
-Ben-Gai, T., A. Bitan, A. Manes, P. Alpert, and S. Rubin. 1998.
-``Spatial and Temporal Changes in Rainfall Frequency Distribution
-Patterns in Israel.'' \emph{Theoretical and Applied Climatology} 61
-(3-4): 177--90.
-\href{\%3CGo\%20to\%20ISI\%3E://000078025000005}{\textless{}Go to ISI\textgreater{}://000078025000005}.
-
-\hypertarget{ref-BoM2012a}{}
-BoM. 2012a. ``Australia - Climate of Our Continent.''
-
-\hypertarget{ref-BoM2012}{}
----------. 2012b. ``Australian Climate Influence.''
-
-\hypertarget{ref-Bosilovich2006}{}
-Bosilovich, Michael G., and Jiun-Dar Chern. 2006. ``Simulation of Water
-Sources and Precipitation Recycling for the MacKenzie, Mississippi, and
-Amazon River Basins.'' \emph{Journal of Hydrometeorology} 7 (3):
-312--29. \url{http://dx.doi.org/10.1175\%2FJHM501.1}.
-
-\hypertarget{ref-Burton2010}{}
-Burton, A, HJ Fowler, S Blenkinsop, and CG Kilsby. 2010. ``Downscaling
-Transient Climate Change Using a Neyman-Scott Rectangular Pulses
-Stochastic Rainfall Model.'' \emph{Journal of Hydrology} 381 (1):
-18--32.
-
-\hypertarget{ref-Chagnon2005}{}
-Chagnon, F. J. F., and R. L. Bras. 2005. ``Contemporary Climate Change
-in the Amazon.'' \emph{Geophysical Research Letters} 32 (13).
-
-\hypertarget{ref-Chowdhury2010}{}
-Chowdhury, R. K., and S. Beecham. 2010. ``Australian Rainfall Trends and
-Their Relation to the Southern Oscillation Index.'' \emph{Hydrological
-Processes} 24 (4). John Wiley \& Sons, Ltd.: 504--14.
-\url{http://dx.doi.org/10.1002/hyp.7504}.
-
-\hypertarget{ref-Cowpertwait2009}{}
-Cowpertwait, PSP, Jim Salinger, and B Mullen. 2009. ``A Spatial-Temporal
-Stochastic Rainfall Model for Auckland City: Scenarios for Current and
-Future Climates.'' \emph{Journal of Hydrology (NZ)} 48: 95--109.
-
-\hypertarget{ref-Deo2009}{}
-Deo, RC, JI Syktus, CA McAlpine, and KK Wong. 2009. ``The Simulated
-Impact of Land Cover Change on Climate Extremes in Eastern Australia.''
-In \emph{Proceedings of the 18th World Imacs Congress and Modsim09
-International Congress on Modelling and Simulation}, 2035--41.
-Modelling; Simulation Society of Australia; New Zealand Inc.;
-International Association for Mathematics; Computers in Simulation.
-
-\hypertarget{ref-SLATS2001}{}
-Department of Natural Resources and Mines. 2005. ``Land Cover Change in
-Queensland 2001-2003, Incorporating 2001-2002 Adn 2002-2003 Change
-Periods: A Statewide Landcover and Trees Study (SLATS) Report.''
-Brisbane: Department of Natural Resources; Mines.
-
-\hypertarget{ref-SLATS2004}{}
-Department of Natural Resources and Water. 2007. ``Land Cover Change in
-Queensland 2004-2005: A Statewide Landcover and Trees Study (SLATS)
-Report.'' Brisbane: Department of Natural Resources; Water.
-
-\hypertarget{ref-SLATS2017}{}
-Department of Science, Information Technology and Innovation. 2017.
-``Land Cover Change in Queensland 2015-2016: A Statewide Landcover and
-Trees Study (SLATS) Report.'' Brisbane: Department of Science,
-Information Technology; Innovation.
-
-\hypertarget{ref-vanDijk2013}{}
-Dijk, Beck van, Albert I. J. M., and Neil R. Viney. 2013. ``The
-Millennium Drought in Southeast Australia (2001--2009): Natural and
-Human Causes and Implications for Water Resources, Ecosystems, Economy,
-and Society.'' \emph{Water Resources Research} 49 (2): 1040--57.
-doi:\href{https://doi.org/10.1002/wrcr.20123}{10.1002/wrcr.20123}.
-
-\hypertarget{ref-Dirmeyer2009}{}
-Dirmeyer, P. A., K. L. Brubaker, and T. DelSole. 2009. ``Import and
-Export of Atmospheric Water Vapor Between Nations.'' \emph{Journal of
-Hydrology} 365 (1-2): 11--22.
-\href{\%3CGo\%20to\%20ISI\%3E://000263575600003}{\textless{}Go to ISI\textgreater{}://000263575600003}.
-
-\hypertarget{ref-Durieux2003}{}
-Durieux, Laurent, Luiz Augusto Toledo Machado, and Henri Laurent. 2003.
-``The Impact of Deforestation on Cloud Cover over the Amazon Arc of
-Deforestation.'' \emph{Remote Sensing of Environment} 86 (1): 132--40.
-
-\hypertarget{ref-Eltahir1996}{}
-Eltahir, E. A. B., and R. L. Bras. 1996. ``Precipitation Recycling.''
-\emph{Reviews of Geophysics} 34 (3): 367--78.
-\href{\%3CGo\%20to\%20ISI\%3E://A1996VD58700003}{\textless{}Go to ISI\textgreater{}://A1996VD58700003}.
-
-\hypertarget{ref-Fowler2005}{}
-Fowler, HJ, CG Kilsby, PE O'connell, and A Burton. 2005. ``A
-Weather-Type Conditioned Multi-Site Stochastic Rainfall Model for the
-Generation of Scenarios of Climatic Variability and Change.''
-\emph{Journal of Hydrology} 308 (1): 50--66.
-
-\hypertarget{ref-Gaertner2001}{}
-Gaertner, M. A., O. B. Christensen, J. A. Prego, J. Polcher, C.
-Gallardo, and M. Castro. 2001. ``The Impact of Deforestation on the
-Hydrological Cycle in the Western Mediterranean: An Ensemble Study with
-Two Regional Climate Models.'' \emph{Climate Dynamics} 17 (11): 857--73.
-
-\hypertarget{ref-Gallant2007}{}
-Gallant, Ailie J. E., Kevin J. Hennessy, and James Risbey. 2007.
-``Trends in Rainfall Indices for Six Australian Regions: 1910 - 2005.''
-\emph{Australian Meteorological Magazine} 56: 223--39.
-
-\hypertarget{ref-Gimeno2010}{}
-Gimeno, Luis, Anita Drumond, Raquel Nieto, Ricardo M. Trigo, and Andreas
-Stohl. 2010. ``On the Origin of Continental Precipitation.''
-\emph{Geophysical Research Letters} 37 (13). AGU.
-\url{http://dx.doi.org/10.1029/2010GL043712}.
-
-\hypertarget{ref-Gowen2016}{}
-Gowen, Rebecca, and Steven G. Bray. 2016. ``Bioeconomic Modelling of
-Woody Regrowth Carbon Offset Options in Productive Grazing Systems.''
-\emph{The Rangeland Journal} 38 (3): 307--17.
-doi:\href{https://doi.org/https://doi.org/10.1071/RJ15084}{https://doi.org/10.1071/RJ15084}.
-
-\hypertarget{ref-Gorgen2006}{}
-Görgen, K., A. H. Lynch, A. G. Marshall, and J. Beringer. 2006. ``Impact
-of Abrupt Land Cover Changes by Savanna Fire on Northern Australian
-Climate.'' \emph{Journal of Geophysical Research-Atmospheres} 111 (D19):
-D19106. \url{http://dx.doi.org/10.1029/2005JD006860}.
-
-\hypertarget{ref-Hastie1986}{}
-Hastie, Trevor, and Robert Tibshirani. 1986. ``Generalized Additive
-Models.'' \emph{Statistical Science}, 297--310.
-
-\hypertarget{ref-Hipel1994}{}
-Hipel, K.W., and A.I. McLeod. 1994. \emph{Time Series Modelling of Water
-Resources and Environmental Systems}. Vol. 45. Elsevier Science Ltd.
-
-\hypertarget{ref-Hirsch1985}{}
-Hirsch, Robert M., and Edward J. Gilroy. 1985. ``Detectability of Step
-Trends in the Rate of Atmospheric Deposition of Sulfate.'' \emph{JAWRA
-Journal of the American Water Resources Association} 21 (5): 773--84.
-\url{http://dx.doi.org/10.1111/j.1752-1688.1985.tb00171.x}.
-
-\hypertarget{ref-Holper2011}{}
-Holper, Paul N. 2011. \emph{Climate Change, Science Information Paper:
-Australian Rainfall-Past, Present and Future}. Canberra: CSIRO.
-
-\hypertarget{ref-Howden2012}{}
-Howden, Saffron. 2012. ``It's Official: Australia No Longer in
-Drought.'' \emph{Brisbane Times}.
-
-\hypertarget{ref-Hughes2003}{}
-Hughes, Lesley. 2003. ``Climate Change and Australia: Trends,
-Projections and Impacts.'' \emph{Austral Ecology} 28 (4). Blackwell
-Science Pty: 423--43.
-\url{http://dx.doi.org/10.1046/j.1442-9993.2003.01300.x}.
-
-\hypertarget{ref-SILO2001}{}
-Jeffrey, Stephen J., John O. Carter, Keith B. Moodie, and Alan R.
-Beswick. 2001. ``Using Spatial Interpolation to Construct a
-Comprehensive Archive of Australian Climate Data.'' Journal Article.
-\emph{Environmental Modelling \& Software} 16 (4 SU -): 309--30.
-\url{http://www.sciencedirect.com/science/article/B6VHC-42SXF8F-2/2/b5c0a24b85149f40ba5d2419ddc0999c}.
-
-\hypertarget{ref-Jones2009}{}
-Jones, D.A., W. Wang, and R. Fawcett. 2009. ``High-Quality Spatial
-Climate Data-Sets for Australia.'' \emph{Australian Meteorological and
-Oceanographic Journal} 58 (4): 233--48.
-
-\hypertarget{ref-Junkermann2009}{}
-Junkermann, W., J. Hacker, T. Lyons, and U. Nair. 2009. ``Land Use
-Change Suppresses Precipitation.'' \emph{Atmospheric Chemistry and
-Physics} 9 (17): 6531--9.
-\href{\%3CGo\%20to\%20ISI\%3E://WOS:000269778500018}{\textless{}Go to ISI\textgreater{}://WOS:000269778500018}.
-
 \hypertarget{ref-Kamruzzaman2011}{}
@@
 Ltd.: 1659--75. \url{http://dx.doi.org/10.1002/hyp.7928}.
-
-\hypertarget{ref-Koutsoyiannis2007}{}
-Koutsoyiannis, Demetris. 2006. ``Nonstationarity Versus Scaling in
-Hydrology.'' Journal Article. \emph{Journal of Hydrology} 324 (1-4):
-239--54.
-\href{http://www.sciencedirect.com/science/article/B6V6C-4HP6GJ9-1/2/67dcc2e8f8e75b7abb918617f5a79a37\%20}{http://www.sciencedirect.com/science/article/B6V6C-4HP6GJ9-1/2/67dcc2e8f8e75b7abb918617f5a79a37}.
-
-\hypertarget{ref-kucharski_further_2013}{}
-Kucharski, Fred, Ning Zeng, and Eugenia Kalnay. 2013. ``A Further
-Assessment of Vegetation Feedback on Decadal Sahel Rainfall
-Variability.'' \emph{Climate Dynamics} 40 (5-6): 1453--66.
-\url{http://link.springer.com/article/10.1007/s00382-012-1397-x}.
-
-\hypertarget{ref-Kuzcera1987}{}
-Kuczera, George. 1987. ``Prediction of Water Yield Reductions Following
-a Bushfire in Ash-Mixed Species Eucalypt Forest.'' \emph{Journal of
-Hydrology} 94 (3-4): 215--36.
-doi:\href{https://doi.org/https://10.1016/0022-1694(87)90054-0}{https://10.1016/0022-1694(87)90054-0}.
-
-\hypertarget{ref-Kundzewicz2004}{}
-Kundzewicz, Z.W., and A.J. Robson. 2004. ``Change Detection in
-Hydrological Records-a Review of the Methodology/Revue Méthodologique de
-La Détection de Changements Dans Les Chroniques Hydrologiques.''
-\emph{Hydrological Sciences Journal} 49 (1).
-
-\hypertarget{ref-Lymburner2010}{}
-Lymburner, Leo, Peter Tan, Norman Mueller, Richard Thackway, Adam Lewis,
-Medhavy Thankappan, Lucy Randall, Anisul Islam, and Udaya Senarath.
-2010. ``250 Metre Dynamic Land Cover Dataset.'' Geoscience Australia,
-Canberra.
-
-\hypertarget{ref-Ma2011}{}
-Ma, H. Y., C. R. Mechoso, Y. Xue, H. Xiao, C. M. Wu, J. L. Li, and F. De
-Sales. 2011. ``Impact of Land Surface Processes on the South American
-Warm Season Climate.'' \emph{Climate Dynamics} 37 (1-2): 187--203.
-\href{\%3CGo\%20to\%20ISI\%3E://WOS:000293403000012}{\textless{}Go to ISI\textgreater{}://WOS:000293403000012}.
-
-\hypertarget{ref-MacDonald2005}{}
-MacDonald, Glen M., and Roslyn A. Case. 2005. ``Variations in the
-Pacific Decadal Oscillation over the Past Millennium.''
-\emph{Geophysical Research Letters} 32 (8): L08703.
-\url{http://dx.doi.org/10.1029/2005GL022478}.
-
-\hypertarget{ref-McAlpine2007}{}
-McAlpine, C. A., J. Syktus, R. C. Deo, P. J. Lawrence, H. A. McGowan, I.
-G. Watterson, and S. R. Phinn. 2007. ``Modeling the Impact of Historical
-Land Cover Change on Australia's Regional Climate.'' \emph{Geophysical
-Research Letters} 34 (22).
-\href{\%3CGo\%20to\%20ISI\%3E://000251345100005}{\textless{}Go to ISI\textgreater{}://000251345100005}.
-
-\hypertarget{ref-Mei2010}{}
-Mei, R., and G. L. Wang. 2010. ``Rain Follows Logging in the Amazon?
-Results from CAM3-CLM3.'' \emph{Climate Dynamics} 34 (7-8): 983--96.
-\href{\%3CGo\%20to\%20ISI\%3E://WOS:000278088400005}{\textless{}Go to ISI\textgreater{}://WOS:000278088400005}.
-
-\hypertarget{ref-Meneghini2007}{}
-Meneghini, Belinda, Ian Simmonds, and Ian N. Smith. 2007. ``Association
-Between Australian Rainfall and the Southern Annular Mode.''
-\emph{International Journal of Climatology} 27 (1): 109--21.
-\url{http://dx.doi.org/10.1002/joc.1370}.
-
-\hypertarget{ref-Murphy2008}{}
-Murphy, Bradley F., and Bertrand Timbal. 2008. ``A Review of Recent
-Climate Variability and Climate Change in Southeastern Australia.''
-\emph{International Journal of Climatology} 28 (7): 859--79.
-\url{http://dx.doi.org/10.1002/joc.1627}.
-
-\hypertarget{ref-Nair2011}{}
-Nair, Udaysankar S., Y. Wu, J. Kala, T. J. Lyons, Sr. Pielke R. A., and
-J. M. Hacker. 2011. ``The Role of Land Use Change on the Development and
-Evolution of the West Coast Trough, Convective Clouds, and Precipitation
-in Southwest Australia.'' \emph{Journal of Geophysical
-Research-Atmospheres} 116 (D7). AGU: D07103.
-\url{http://dx.doi.org/10.1029/2010JD014950}.
-
-\hypertarget{ref-Narisma2007}{}
-Narisma, G.T., J.A. Foley, R. Licker, and N. Ramankutty. 2007. ``Abrupt
-Changes in Rainfall During the Twentieth Century.'' \emph{Geophysical
-Research Letters} 34 (6): L06710.
-
-\hypertarget{ref-Negri2004}{}
-Negri, A. J., R. F. Adler, L. M. Xu, and J. Surratt. 2004. ``The Impact
-of Amazonian Deforestation on Dry Season Rainfall.'' \emph{Journal of
-Climate} 17 (6): 1306--19.
-
-\hypertarget{ref-Nicholls2006}{}
-Nicholls, N. 2006. ``Detecting and Attributing Australian Climate
-Change: A Review.'' \emph{Australian Meteorological Magazine} 55 (3).
-Bureau of Meteorology.: 199--211.
-
-\hypertarget{ref-Oleson2004}{}
-Oleson, K. W., G. B. Bonan, S. Levis, and M. Vertenstein. 2004.
-``Effects of Land Use Change on North American Climate: Impact of
-Surface Datasets and Model Biogeophysics.'' \emph{Climate Dynamics} 23
-(2): 117--32.
-
-\hypertarget{ref-Onoz2003}{}
-Onoz, B., and M. Bayazit. 2003. ``The Power of Statistical Tests for
-Trend Detection.'' \emph{Turkish Journal of Engineering and
-Environmental Sciences} 27 (4): 247--51.
-
-\hypertarget{ref-Otterman1990}{}
-Otterman, J., A. Manes, S. Rubin, P. Alpert, and D. Starr. 1990. ``An
-Increase of Early Rains in Southern Israel Following Land-Use Change?''
-\emph{Boundary-Layer Meteorology} 53 (4): 333--51.
-
-\hypertarget{ref-Pinto2009}{}
-Pinto, E., Y. Shin, S. A. Cowling, and C. D. Jones. 2009. ``Past,
-Present and Future Vegetation-Cloud Feedbacks in the Amazon Basin.''
-\emph{Climate Dynamics} 32 (6): 741--51.
-\href{\%3CGo\%20to\%20ISI\%3E://WOS:000264118400001}{\textless{}Go to ISI\textgreater{}://WOS:000264118400001}.
-
-\hypertarget{ref-Pitman2007}{}
-Pitman, A. J., and P. P. Hesse. 2007. ``The Significance of Large-Scale
-Land Cover Change on the Australian Palaeomonsoon.'' \emph{Quaternary
-Science Reviews} 26 (1-2): 189--200.
-
-\hypertarget{ref-pitman_scale_2016}{}
-Pitman, A. J., and R. Lorenz. 2016. ``Scale Dependence of the Simulated
-Impact of Amazonian Deforestation on Regional Climate.''
-\emph{Environmental Research Letters} 11 (9): 094025.
-doi:\href{https://doi.org/10.1088/1748-9326/11/9/094025}{10.1088/1748-9326/11/9/094025}.
-
-\hypertarget{ref-Pitman2004}{}
-Pitman, A. J., G. T. Narisma, R. A. Pielke, and N. J. Holbrook. 2004.
-``Impact of Land Cover Change on the Climate of Southwest Western
-Australia.'' \emph{Journal of Geophysical Research-Atmospheres} 109
-(D18).
-\href{\%3CGo\%20to\%20ISI\%3E://WOS:000224126500001}{\textless{}Go to ISI\textgreater{}://WOS:000224126500001}.
-
-\hypertarget{ref-Rstats2018}{}
-R Core Team, 2018. \emph{R: A Language and Environment for Statistical
-Computing}. Vienna, Austria: R Foundation for Statistical Computing.
-\url{http://www.R-project.org/}.
 
@@
 
-\hypertarget{ref-Roy2008}{}
-Roy, David P, Luigi Boschetti, Christopher O Justice, and J Ju. 2008.
-``The Collection 5 Modis Burned Area Product---Global Evaluation by
-Comparison with the Modis Active Fire Product.'' \emph{Remote Sensing of
-Environment} 112 (9). Elsevier: 3690--3707.
-
-\hypertarget{ref-Roy2005}{}
-Roy, DP, Y Jin, PE Lewis, and CO Justice. 2005. ``Prototyping a Global
-Algorithm for Systematic Fire-Affected Area Mapping Using Modis Time
-Series Data.'' \emph{Remote Sensing of Environment} 97 (2). Elsevier:
-137--62.
-
-\hypertarget{ref-Roy2002}{}
-Roy, DP, PE Lewis, and CO Justice. 2002. ``Burned Area Mapping Using
-Multi-Temporal Moderate Spatial Resolution Data---A Bi-Directional
-Reflectance Model-Based Expectation Approach.'' \emph{Remote Sensing of
-Environment} 83 (1). Elsevier: 263--86.
-
-\hypertarget{ref-saha_investigating_2016}{}
-Saha, Subodh K., Paul A. Dirmeyer, and Thomas N. Chase. 2016.
-``Investigating the Impact of Land-Use Land-Cover Change on Indian
-Summer Monsoon Daily Rainfall and Temperature During 1951-2005 Using a
-Regional Climate Model.'' \emph{Hydrology and Earth System Sciences} 20
-(5): 1765.
-\url{http://search.proquest.com/openview/c4242c9ecce075e3545390b1d00f2751/1?pq-origsite=gscholar\&cbl=105724}.
-
-\hypertarget{ref-Saji1999}{}
-Saji, NH, B.N. Goswami, PN Vinayachandran, and T. Yamagata. 1999. ``A
-Dipole Mode in the Tropical Indian Ocean.'' \emph{Nature} 401 (6751):
-360--63.
-
-\hypertarget{ref-Sato2007}{}
-Sato, T., F. Kimura, and A. S. Hasegawa. 2007. ``Vegetation and
-Topographic Control of Cloud Activity over Arid/Semiarid Asia.''
-\emph{Journal of Geophysical Research-Atmospheres} 112 (D24).
-
-\hypertarget{ref-Schepen2012}{}
-Schepen, Andrew, Q. J. Wang, and David Robertson. 2012. ``Evidence for
-Using Lagged Climate Indices to Forecast Australian Seasonal Rainfall.''
-\emph{Journal of Climate} 25 (4). American Meteorological Society:
-1230--46.
-
-\hypertarget{ref-Semazzi2001}{}
-Semazzi, F. H. M., and Y. Song. 2001. ``A GCM Study of Climate Change
-Induced by Deforestation in Africa.'' \emph{Climate Research} 17 (2):
-169--82.
-
-\hypertarget{ref-Smith2012}{}
-Smith, Ian N., and Bertrand Timbal. 2012. ``Links Between Tropical
-Indices and Southern Australian Rainfall.'' \emph{International Journal
-of Climatology} 32 (1): 33--40.
-
-\hypertarget{ref-Speer2011}{}
-Speer, Milton, Lance Leslie, and Alexandre Fierro. 2011. ``Australian
-East Coast Rainfall Decline Related to Large Scale Climate Drivers.''
-\emph{Climate Dynamics} 36 (7): 1419--29.
-\url{http://dx.doi.org/10.1007/s00382-009-0726-1}.
-
-\hypertarget{ref-Fire2011}{}
-The State Government of Victoria. 2011. ``Bushfire History.''
-
-\hypertarget{ref-Timbal2006}{}
-Timbal, B., and J.M. Arblaster. 2006. ``Land Cover Change as an
-Additional Forcing to Explain the Rainfall Decline in the South West of
-Australia.'' \emph{Geophysical Research Letters} 33 (7): L07717.
-
-\hypertarget{ref-Townshend2011}{}
-Townshend, J.R.G., M. Carroll, C. Dimiceli, R. Sohlberg, M. Hansen, and
-R. DeFries. 2011. ``Vegetation Continuous Fields MOD44B, 2001 Percent
-Tree Cover, Collection 5.'' University of Maryland, College Park,
-Maryland, 2001.
-
-\hypertarget{ref-Beesley2009}{}
-Tozer, C. R., A. S. Kiem, and D. C. Verdon-Kidd. 2009. ``On the
-Uncertainties Associated with Using Gridded Rainfall Data as a Proxy for
-Observed.'' Conference Paper. In \emph{18th World Imacs Congress and
-Modsim09 International Congress on Modelling and Simulation.}, edited by
-R.S. Anderssen, R.D. Braddock, and L.T.H. Newham, 3886--92.
-\href{http://www.mssanz.org.au/modsim09/I13/beesley.pdf\%20}{http://www.mssanz.org.au/modsim09/I13/beesley.pdf}.
-
-\hypertarget{ref-Tozer2012}{}
----------. 2012. ``On the Uncertainties Associated with Using Gridded
-Rainfall Data as a Proxy for Observed.'' Journal Article. \emph{Hydrol.
-Earth Syst. Sci.} 16 (5): 1481--99.
-doi:\href{https://doi.org/10.5194/hess-16-1481-2012}{10.5194/hess-16-1481-2012}.
-
-\hypertarget{ref-Trenberth1999}{}
-Trenberth, K. E. 1999. ``Atmospheric Moisture Recycling: Role of
-Advection and Local Evaporation.'' \emph{Journal of Climate} 12 (5):
-1368--81.
-\href{\%3CGo\%20to\%20ISI\%3E://000080145700002}{\textless{}Go to ISI\textgreater{}://000080145700002}.
-
-\hypertarget{ref-Verdon2004}{}
-Verdon, Danielle C., Adam M. Wyatt, Anthony S. Kiem, and Stewart W.
-Franks. 2004. ``Multidecadal Variability of Rainfall and Streamflow:
-Eastern Australia.'' \emph{Water Resources Research} 40 (10): W10201.
-\url{http://dx.doi.org/10.1029/2004WR003234}.
-
-\hypertarget{ref-Wang2009}{}
-Wang, Jingfeng, Frederic Chagnon, Earle Williams, Alan Betts, Nilton
-Renno, Luiz Machado, Gautam Bisht, Ryan Knox, and Rafael Bras. 2009.
-``Impact of Deforestation in the Amazon Basin on Cloud Climatology.''
-\emph{Proceedings of the National Academy of Sciences of the United
-States of America} 106 (10): 3670.
-
-\hypertarget{ref-Westra2013}{}
-Westra, Seth, Lisa V. Alexander, and Francis W. Zwiers. 2013. ``Global
-Increasing Trends in Annual Maximum Daily Precipitation.'' \emph{Journal
-of Climate} 26 (11). American Meteorological Society: 3904--18.
-\url{http://dx.doi.org/10.1175/JCLI-D-12-00502.1}.
-
-\hypertarget{ref-Westra2010}{}
-Westra, Seth, and Ashish Sharma. 2010. ``An Upper Limit to Seasonal
-Rainfall Predictability?'' \emph{Journal of Climate} 23 (12). American
-Meteorological Society: 3332--51.
-
-\hypertarget{ref-Whitley2002}{}
-Whitley, Elise, and Jonathan Ball. 2002. ``Statistics Review 6:
-Nonparametric Methods.'' \emph{Critical Care, London} 6 (6): 509--13.
-
-\hypertarget{ref-Wilks2006}{}
-Wilks, D. S. 2006. ``On `Field Significance' and the False Discovery
-Rate.'' \emph{Journal of Applied Meteorology and Climatology} 45 (9).
-American Meteorological Society: 1181--9.
-
-\hypertarget{ref-Wilks1997}{}
-Wilks, DS. 1997. ``Resampling Hypothesis Tests for Autocorrelated
-Fields.'' \emph{Journal of Climate} 10 (1): 65--82.
-
-\hypertarget{ref-Wood2011}{}
-Wood, S. N. 2011. ``Fast Stable Restricted Maximum Likelihood and
-Marginal Likelihood Estimation of Semiparametric Generalized Linear
-Models.'' \emph{Journal of the Royal Statistical Society (B)} 73 (1):
-3--36.
-
-\hypertarget{ref-Zanchettin2008}{}
-Zanchettin, Davide, Stewart W. Franks, Pietro Traverso, and Mario
-Tomasino. 2008. ``On Enso Impacts on European Wintertime Rainfalls and
-Their Modulation by the Nao and the Pacific Multi-Decadal Variability
-Described Through the Pdo Index.'' \emph{International Journal of
-Climatology} 28 (8). John Wiley \& Sons, Ltd.: 995--1006.
-\url{http://dx.doi.org/10.1002/joc.1601}.
-
-\hypertarget{ref-Zeng2012}{}
-Zeng, X. M., Z. H. Wu, S. Song, S. Y. Xiong, Y. Q. Zheng, Z. G. Zhou,
-and H. Q. Liu. 2012. ``Effects of Land Surface Schemes on the Simulation
-of a Heavy Rainfall Event by WRF.'' \emph{Chinese Journal of
-Geophysics-Chinese Edition} 55 (1): 16--28.
-\href{\%3CGo\%20to\%20ISI\%3E://WOS:000300129300002}{\textless{}Go to ISI\textgreater{}://WOS:000300129300002}.
-
 \end{document}

### 2019-07-05 Manuscripts/PeerJRainfall_Landuse/Rainfall_Landuse.Rmd
changes requested by PeerJ submission figures

@@
   - code: 1
-    address: Sydney Institute of Agriculture, The University of Sydney, NSW 2006
+    address: School of Life and Environmental Sciences, Sydney Institute of Agriculture, The University of Sydney, New South Wales, Australia.
 abstract: > 
@@
 #library(nnet)
+
 ```
@@
 ```{r setup, warning=F, message=F, echo =F}
+computer <- "laptop"
+
 # root dir
+if (computer == "laptop") {
   knitr::opts_knit$set(root.dir =
-      # "E:/cloudstor/chun/chun/manuscripts/PeerJRainfall_Landuse")
       "C:/users/rver4657/owncloud/chun/chun/manuscripts/PeerJRainfall_Landuse")
+} else {
+  knitr::opts_knit$set(root.dir =
+       "E:/cloudstor/chun/chun/manuscripts/PeerJRainfall_Landuse")
+}
+
 ```
@@
 
-include_graphics("figures/map_selreg.png")
+include_graphics("figures/Fig1.png")
 ```
@@
 Some locations in the southwest of the Australian Capital Territory (ACT) have also experienced intensive bushfires.", out.width="90%", echo=F}
-include_graphics("figures/bushfire_nswvic.png")
+include_graphics("figures/Fig2.png")
 ```
@@
 
-```{r figure3-tc-simple, fig.cap= "The expected evolution of the land surface after trees have been removed in (left) the QLD region and (right) the NSW/VIC region.", out.width="90%", echo=F}
-include_graphics("figures/tc_simple.png")
+```{r figure3-tc-simple, fig.cap= "The expected evolution of the land surface after trees have been removed in (A) the QLD region and (B) the NSW/VIC region.", out.width="70%", echo=F}
+include_graphics("figures/Fig3.png")
 ```
@@
 ## the selected regions for mapping
+if (computer == "laptop") {
+  root_dir <- "c:/users/rver4657/owncloud/chun/chun/data"
+} else {
   root_dir <- "e:/cloudstor/chun/chun/data"
+}
 aus_reg <- readShapePoly(paste(root_dir,
@@
 
-png("figures/bushfire_nswvic.png", width=960, height=960)
+# bushfire nsw and vic figure
+png("figures/Fig2.png", width=1000, height=1000, units="px")
   par(mar=c(3,3,1,7),xpd=T)
@@
 
-png("figures/slats.png", width=1200, height = 900)
+# SLATS figure at end of the paper
+png("figures/Fig11.png", width=1200, height = 900, units = "px")
 Data %>% 
@@
   scale_fill_grey() +
+  theme_bw() +
   theme(axis.text.x = element_text(size = rel(2),
@@
         legend.text = element_text(size = rel(2)),
-        legend.title = element_text(size = rel(2), face = "bold"))
+        legend.title = element_text(size = rel(2), face = "bold"),
+  legend.position=c(0.9,0.7))
 dev.off()
@@
 ```{r treeCover,eval=F , echo = F}
-
-rootdir <-"E:/cloudstor/chun2/data/satellite/MOD44/B"
+if (computer == "laptop") {
+  root_dir <- "c:/users/rver4657/owncloud/chun2/data/satellite/MOD44/B"
+} else {
+  root_dir <- "e:/cloudstor/chun2/data/satellite/MOD44/B"
+}
 
 # read in the file list from the directory
-x <- dir(path=rootdir,pattern="Percent_Tree_Cover.tif")
+x <- dir(path=root_dir,pattern="Percent_Tree_Cover.tif")
 # Identify the different states in the images
@@
 for(i in 1:length(NSW)) {
-  img[[i]] <- raster(paste(rootdir,NSW[i],sep="/"))
+  img[[i]] <- raster(paste(root_dir,NSW[i],sep="/"))
 }
@@
 for(i in 1:length(Qld)) {
-  imgQ[[i]] <- raster(paste(rootdir,Qld[i],sep="/"))
+  imgQ[[i]] <- raster(paste(root_dir,Qld[i],sep="/"))
 }
@@
 plotclr <- brewer.pal(8,"PiYG")
-
-png("figures/tc_figs.png", width=1500, height=900)
+# Fig 5 in the paper was tc_figs.png
+png("figures/Fig5.png", width=1500, height=900, unit = "px")
 par(mfrow=c(1,2), mar=c(5,3,5,5))
@@
 plot(mdb_reg2,border="grey",lwd=5,add=T,ylab="")
+mtext("A",3, font=2, cex=2.5, line = -5, at=c(146,-25.6))
 # FULL PERIOD NSW
@@
 plot(mdb_reg2,border="grey",lwd=5,add=T,ylab="")
+mtext("B",3, font=2, cex=2.5, line = -5, at=c(147.2,-35.1))
 image.plot(zlim=c(-4,4), breaks=c(-4,-3,-2,-1,-0.2,0.2,1,2,3,4),
@@
 The vertical red lines indicate the year of 2003, in which the studied land cover changes occurred.
-A change in the time series data is not obvious before and after the land cover changes.", out.width="90%", echo=F}
-
-include_graphics("figures/Rainfall_resid.png")
+A change in the time series data is not obvious before and after the land cover changes.", out.width="90%"}
+
+include_graphics("figures/Fig4.png")
 ```
@@
 
-png("figures/hist_rainfall.png", width=900, height=900)
+png("figures/hist_rainfall.png", width=900, height=900, 
+    units = "px")
 rain %>%
@@
 #
-plot_df <- tibble(Dates = rep(qld_mean$Date,2),
-                  Resid = c(qld_deseason,nsw_deseason),
-                  State = c(rep("Qld",length(qld_deseason)),
-                            rep("NSW",length(nsw_deseason))))
-
-png("figures/Rainfall_resid.png", width=960, height=960)
-plot_df %>%
-  ggplot(aes(Dates,Resid)) + geom_line() + 
-  geom_vline(xintercept = as.Date("2003-01-01"), 
-             col="red", size=1.5) +
-  geom_text(aes(x=as.Date("2004-12-01"),y=150),
-            label="2003", col="red", size=7) +
-  facet_wrap(~State, ncol = 1) + xlab("Date") +
-  ylab("Deseasonalised rainfall for the NSW and Qld locations") +
-  theme(axis.text=element_text(size=rel(2)),
-        axis.title=element_text(size=rel(2.5)),
-        strip.text =element_text(size=rel(2), face="bold"))
-dev.off()
 
@@
 
-png("figures/tc_simple.png", width=1200, height=900)
+png("figures/Fig3.png", width=1200, height=900)
 par(mar=c(5,8,4,3), mfrow=c(1,2))
-plot(data1$year,data1$tc,type="l",ylim=c(0.5,2.5),xlab="",ylab="",xaxt="n",yaxt="n")
-axis(1,at=c(1,2,3),labels=c("before","LCC","after"),cex.axis=2, cex.lab=2)
-axis(2,at=c(1,2),label=c("removed","tree"),cex.axis=2,las=2, cex.lab=1.5)
+plot(data1$year,data1$tc,type="l",ylim=c(0.5,2.5),xlab="",ylab="",xaxt="n",yaxt="n", lwd=2, col = "blue")
+axis(1,at=c(1,2,3),labels=c("before","LCC","after"),cex.axis=2, cex.lab=2, lwd=2)
+axis(2,at=c(1,2),labels=c("removed","tree"),cex.axis=2,las=2, cex.lab=1.5, lwd=2)
+mtext("A",3, font=2, cex=2.5, line = -3, at=1)
 par(mar=c(5,8,4,3))
-plot(data2$year,data2$tc,type="l",ylim=c(0.5,2.5),xlab="",ylab="",xaxt="n",yaxt="n")
-axis(1,at=c(1,2,4),labels=c("before","LCC","after"),cex.axis=2, cex.lab=2)
-axis(2,at=c(1,2),label=c("removed","tree"),cex.axis=2, cex.lab=2,las=1.5)
+plot(data2$year,data2$tc,type="l",ylim=c(0.5,2.5),xlab="",ylab="",xaxt="n",yaxt="n", lwd=2, col = "blue")
+axis(1,at=c(1,2,4),labels=c("before","LCC","after"),cex.axis=2, cex.lab=2, lwd=2)
+axis(2,at=c(1,2),labels=c("removed","tree"),cex.axis=2, cex.lab=2,las=1.5, lwd=2)
+mtext("B",3, font=2, cex=2.5, line = -3, at=1)
 dev.off()
@@
 
-png("figures/hist_sign_LC.png", width=900, height = 900)
+png("figures/hist_sign_LC.png", width=900, height = 900, units = "px")
 par(mfrow=c(2,1))
@@
 # Now make the final plot for the paper with both sites and 2009 and all data
-  png("figures/Cp_30yrs.png", width=1800, height=900)
+  # Figure 8
+  png("figures/Fig8.png", width=1800, height=900, units="px")
   par(mfrow=c(1,2))
@@
   plot(e111,add=T)
+  mtext("A",3, font=2, cex=2.5, line = -5, at=c(146,-25.6))
 legend(148,-27.5,c("State","MDB"),col=c("grey","grey"),lwd=c(2,5),bty="n",title="Boundary",cex=1.5)
@@
   plot(e222,add=T)
+  mtext("B",3, font=2, cex=2.5, line = -5, at=c(147.2,-35.7))
   image.plot(zlim=c(0,4),breaks=c(0,1,2,3,4),lab.breaks=c(0,0.05,0.1,0.2,1),legend.only=T,horizontal=F,col=c("red","orange","yellow","light blue"),legend.shrink=0.6,legend.width=1.2,legend.args=list(text="p value",line=1,cex=2,font=2),axis.args=list(cex.axis=1.8))
@@
                                 rep("NSW/VIC", length(rsq_nsw))))
-
-png("figures/rsq_fullmodel.png", height=900, width=900)
+# Histogram of rsquared of full model
+png("figures/Fig6.png", height=1000, width=1000, units = "px")
 plot_df %>%
@@
   xlab(expression(paste("% of variability explained (",r^2,")"))) +
-    theme(axis.text = element_text(size=rel(1.5)),
-        strip.text = element_text(size=rel(2)),
-        axis.title = element_text(size=rel(2)))
+  ylab("Frequency") +
+   theme_bw() +
+    theme(axis.text = element_text(size=rel(2.5)),
+        strip.text = element_text(size=rel(3)),
+        axis.title = element_text(size=rel(3)),
+        legend.text = element_text(size=rel(2.5)),
+        legend.title = element_text(size=rel(2.5)),
+        legend.position = c(0.15, 0.9)) 
 dev.off()
@@
 plot_df <- rbind(qld_df, nsw_df)
-png("figures/soi_explHistogram.png", width=960, height=960)
+# SOI explaining histogram
+dat_text <- data.frame(
+  label = c("A", "B"),
+  state   = c("Qld","NSW"),
+  x     = c(2.5,2.5),
+  y     = c(285,285)
+)
+
+png("figures/Fig7.png", width=960, height=960, units="px")
 plot_df %>%
   ggplot(aes(rsq)) + geom_histogram(fill="blue") +
+  xlab(expression(paste("% of variability explained (adj. ",r^2,
+                        ")"))) +  
+  ylab("frequency") +
+  geom_text(data=dat_text,aes(x=x,y=y, label=label), 
+            col="black", size=10, fontface="bold") +
   facet_wrap(~state) +
-  xlab(expression(paste("% of variability explained (adj",r^2,")"))) +
-  theme(axis.text = element_text(size=rel(1.5)),
-        strip.text = element_text(size=rel(2)),
-        axis.title = element_text(size=rel(2)))
+  theme_bw() +
+  theme(axis.text = element_text(size=rel(3)),
+        strip.text = element_text(size=rel(3)),
+        axis.title = element_text(size=rel(3)))
  dev.off()
@@
 plotclt <- rev(brewer.pal(8,"RdYlBu"))
-
-png("figures/step_new.png", width=1800, height=900)
+# new step trend Z scores
+png("figures/Fig9.png", width=1800, height=900, units = "px")
   par(mar=c(5,3,3,7), mfrow=c(1,2), xpd=T)
@@
   plot(e111,add=T)
+  mtext("A",3, font=2, cex=2.5, line = -5, at=c(146,-25.6))
+
 # image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,0,1,2,3,4),lab.breaks=c(0.00,0.05,0.1,0.2,1,0.2,0.1,0.05,0.00),legend.only=T,horizontal=F,col=plotclt,legend.shrink=0.7,legend.width=0.9,legend.args=list(text=c(expression(paste("level of significance (",alpha,")")),"positive Z'","negative Z'"),side=c(2,3,1),line=0.7,cex=1.5,font=2),axis.args=list(cex.axis=1.5))
@@
   plot(e222,add=T)
+  mtext("B",3, font=2, cex=2.5, line = -5, at=c(147.2,-35.7))
   par(mar=c(3,3,1,9),xpd=T)
@@
                                  qld_origp.new, qld_orign.new)*100, 
-                      sign = rep(c("postive","negative"),2),
+                      sign = rep(c("positive","negative"),2),
                       State=c("NSW/VIC","NSW/VIC","QLD","QLD"),
@@
                                  bt4_qldp.values, bt4_qldp.values),
-                    sign =c(rep("postive",length(bt4_nswp.values)),
+                    sign =c(rep("positive",length(bt4_nswp.values)),
                             rep("negative",length(bt4_nswn.values)),
-                            rep("postive",length(bt4_qldp.values)),
+                            rep("positive",length(bt4_qldp.values)),
                             rep("negative",length(bt4_qldn.values))),
@@
 
-png("figures/bt4_hist.png", width=1000,height=1000)
+# histogram for bootstrao z scores
+dat_text <- data.frame(
+  label = c("A", "B", "C", "D"),
+  State   = c("NSW/VIC","NSW/VIC", "QLD", "QLD"),
+  sign = c("negative","positive", "negative","positive"),
+  x     =rep(100,4),
+  y     = c(275,900,900,900)
+)
+
+png("figures/Fig10.png", width=1000,height=1000, units="px")
 # A histogram of taus
@@
                  shape=16,size=8) + 
+    geom_text(data=dat_text,aes(x=x,y=y, label=label), 
+            col="black", size=10, fontface="bold") +
   facet_wrap(~ State*sign,ncol=2, scales="free") #+
 hp <- hp + scale_colour_grey(start = 0, end = 0.6)
-hp <- hp + theme(axis.title = element_text(face ="bold", size=rel(2)),
-                 strip.text = element_text(face ="bold", size=rel(2)),
-                 axis.text = element_text(face ="bold", size=rel(1.5)))
+hp <- hp + theme_bw() + 
+  theme(axis.title = element_text(face ="bold", size=rel(3)),
+                 strip.text = element_text(face ="bold",
+                                           size=rel(3)),
+                 axis.text = element_text(face ="bold",
+                                          size=rel(2.5)))
 hp
@@
 
-```{r tctrend, fig.cap ="The maps show significant changes in tree cover identified from the MOD44B data between 2003 and 2015 in the NSW/VIC region (right) and the Qld region (left).
+```{r tctrend, fig.cap ="The maps show significant changes in tree cover identified from the MOD44B data between 2003 and 2015 in the Qld region (A) and the NSW/VIC region (B).
 The amount of change was calculated as the difference in tree cover before and after the specified land cover intervention and it is shown as the percentage of the ground area.
 Green colour indicates an increase in tree cover, while red colour indicates a decrease in tree cover.", echo=F, out.width="90%"}
-include_graphics("figures/tc_figs.png")
+include_graphics("figures/Fig5.png")
 ```
@@
 ```{r rsqmodel, fig.cap = "The distribution of the variance explained (adjusted $r^2$ ) by the full regression model", echo=F, out.width="70%"}
-include_graphics("figures/rsq_fullmodel.png")
+include_graphics("figures/Fig6.png")
 
@@
 ```{r rsq, fig.cap = "The performance of the regression model if rainfall is only modelled by the climate drivers.
-It shows the percentage of rainfall variability that can be explained by the climate drivers for the Qld and NSW/VIC region", echo=F, out.width="70%"}
-include_graphics("figures/soi_explhistogram.png")
+It shows the percentage of rainfall variability that can be explained by the climate drivers for the Qld (A) and NSW/VIC (B) region", echo=F, out.width="70%"}
+include_graphics("figures/Fig7.png")
 
@@
 
-```{r LCp, fig.cap = "The spatial distribution of significance of land cover step change variable in the GAM model predicting changes in rainfall in the both regions.
-The left graph relates to the Qld region, while the right hand plot reflects the NSW/VIC region.
+```{r LCp, fig.cap = "The spatial distribution of significance of land cover step change variable in the GAM model predicting changes in rainfall in the both regions, the Qld region (A) and the NSW/VIC region (B).
 The outlines of relevant Australian states and the Murray Darling Basin are indicated in grey.
-The p value reported is for the land cover variable in the model.", echo=F, out.width="90%"}
-include_graphics("figures/Cp_30yrs.png")
+The p-value reported is for the land cover variable in the model.", echo=F, out.width="90%"}
+include_graphics("figures/Fig8.png")
 ```
@@
 The land cover variable in the model (Equation \@ref(eq:model2)) aims to identify a step change in the rainfall before and after the observed change in land cover.
-The  variable was mainly significant ($p \leq 0.05$) for the rainfall estimates in some areas in NSW/VIC, as shown in Figure \@ref(fig:LCp) (right panel).
-However, the number of pixels where the landcover variable was significant was much greater in NSW/VIC compared to the number in the QLD region.
+The  variable was mainly significant ($p \leq 0.05$) for the rainfall estimates in some areas in NSW/VIC, as shown in Figure \@ref(fig:LCp) (B).
+However, the number of pixels where the landcover variable was significant was much greater in NSW/VIC compared to the number in the QLD region (Figure \@ref(fig:LCp) (A)).
 There was also some relationship between the areas of bushfires in Figure \@ref(fig:bushfire).
-Only a very small area with a significant step change due to the land cover changes was found in rainfall in the QLD region (left panel)
+Only a very small area with a significant step change due to the land cover changes was found in rainfall in the QLD region (A)
 
@@
 
-```{r steptest30, fig.cap="Spatial distribution of the step trend test Z' statistics in the two study sites.
-The panel on the left is for the QLD region, while the panel of the right is for NSW/VIC.
+```{r steptest30, fig.cap="Spatial distribution of the step trend test Z' statistics in the two study sites, the QLD region (A) and the NSW/VIC region (B).
 Warm colours (yellow, orange and red) are for positive Z' values which indicate decreasing rainfall trend due to the land surface change.
@@
 The deeper the colour, the more significant the statistic.", echo = F, out.width="90%"}
-include_graphics("figures/step_new.png")
-```
-
-In both regions there are areas of positive Z' values which imply a decrease in rainfall (Figure \@ref(fig:steptest30)), but this is stronger in the NSW/VIC area than in the QLD area.
-Qualitatively the locations where changes in tree cover occurred in the NSW/VIC area (Figure \@ref(fig:tctrend)) seem to agree with the patterns in the Z' scores.
+include_graphics("figures/Fig9.png")
+```
+
+In both regions there are areas of positive Z' values which imply a decrease in rainfall (Figure \@ref(fig:steptest30)), but this is stronger in the NSW/VIC area (B) than in the QLD area (A).
+Qualitatively the locations where changes in tree cover occurred in the NSW/VIC area (Figure \@ref(fig:tctrend) (B)) seem to agree with the patterns in the Z' scores.
 However, a direct relationship would not necesssarily be expected as movement of air masses could mean that actual changes of rainfall are observed close by, but not necessarily exactly at areas with changes of landcover.
-For the QLD region (left panel) there are only a few significant Z' scores which matches the earlier significance in the landcover variable in the full regresion.
+For the QLD region (A) there are only a few significant Z' scores which matches the earlier significance in the landcover variable in the full regresion.
 In the QLD region, only 0.2\% of the pixels obtained a positive Z' score with p $<$ 0.1.
-In the NSW/VIC region 3.7\% of pixels have a positive Z' score with p $<$ 0.1.
+In the NSW/VIC region (B) 3.7\% of pixels have a positive Z' score with p $<$ 0.1.
 In general it is only a small proportion of both study regions.
@@
 ```{r fieldsig, fig.cap ="Field significance results showing the distribution of percentage of significant Z' scores for the resampled series (blue bars), and the precentage significant Z' scores for the original series of rainfall years (red circle)", echo=F, out.width="70%"}
-include_graphics("figures/bt4_hist.png")
+include_graphics("figures/Fig10.png")
 ```
@@
 The data were obtained from SLATS (2017), and clearly indicate a sharp decrease in the clearing rate after 2005.", echo=F, out.width="90%"}
-include_graphics("figures/slats.png")
+include_graphics("figures/Fig11.png")
 ```

### 2019-07-05 Manuscripts/PeerJRainfall_Landuse/SupplementaryMaterialRainfall_Landuse.Rmd
changes requested by PeerJ submission figures

@@
       "C:/users/rver4657/owncloud/chun/chun/manuscripts/PeerJRainfall_Landuse")
+knitr::opts_knit$set('figcap.prefix','Figure S')
+knitr::opts_knit$set('tabcap.prefix','Table S')
 ```
@@
 ```{r cor-rain-qld, fig.cap="Cross-correlation of six climate indices and rainfall in QLD study region.
+Bars in the plot indicate the strength of the cross-correlation at different lags.
 For the PDO analysis, 108-year rainfall data (1900 - 2008) are used.
@@
 The correlation with NINO 3 is not shown as it is very similar to but weaker than for NINO 3.4.", out.width="90%", echo=F}
-include_graphics("figures/cor_qld.png")
+include_graphics("figures/FigS1.png")
 ```
 
-```{r cor-rain-nsw, fig.cap="Cross-correlation of six climate indices and rainfall in NSW/VIC study region.", out.width="90%", echo=F}
-include_graphics("figures/cor_nswvic.png")
+```{r cor-rain-nsw, fig.cap="Cross-correlation of six climate indices and rainfall in NSW/VIC study region.
+Bars in the plot indicate the strength of the cross-correlation at different lags.
+For the PDO analysis, 108-year rainfall data (1900 - 2008) are used.
+Otherwise, 36-year rainfall data are used.
+The correlation with NINO 3 is not shown as it is very similar to but weaker than for NINO 3.4.", out.width="90%", echo=F}
+include_graphics("figures/FigS2.png")
 ```
@@
 
-```{r residuals, fig.cap ="The residual analysis of a sample pixel in the QLD region (top) and NSW/VIC region (bottom).", echo=F, eval=T, out.width="90%"}
+```{r residuals, fig.cap ="The residual analysis of the Generalised additive modelling  indicating how well the residuals follow the regression assumptions.
+Results are shown for a sample pixel in the QLD region (top) and NSW/VIC region (bottom).
+Residuals are fairly normal as seen from the histogram and the qqplot, but show some scattering in the variance.", echo=F, eval=T, out.width="90%"}
 # removed after review
-include_graphics("figures/gam_check.png")
+include_graphics("figures/FigS3.png")
 ```
@@
 # figure dropped after review
-include_graphics("figures/ResidualBoxplotchange.png")
+include_graphics("figures/FigS4.png")
 ```

### 2019-07-05 Manuscripts/PeerJRainfall_Landuse/SupplementaryMaterialRainfall_Landuse.tex
changes requested by PeerJ submission figures

@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{figures/cor_qld} \caption{Cross-correlation of six climate indices and rainfall in QLD study region.
+\includegraphics[width=0.9\linewidth]{figures/FigS1} \caption{Cross-correlation of six climate indices and rainfall in QLD study region.
+Bars in the plot indicate the strength of the cross-correlation at different lags.
 For the PDO analysis, 108-year rainfall data (1900 - 2008) are used.
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{figures/cor_nswvic} \caption{Cross-correlation of six climate indices and rainfall in NSW/VIC study region.}\label{fig:cor-rain-nsw}
+\includegraphics[width=0.9\linewidth]{figures/FigS2} \caption{Cross-correlation of six climate indices and rainfall in NSW/VIC study region.
+Bars in the plot indicate the strength of the cross-correlation at different lags.
+For the PDO analysis, 108-year rainfall data (1900 - 2008) are used.
+Otherwise, 36-year rainfall data are used.
+The correlation with NINO 3 is not shown as it is very similar to but weaker than for NINO 3.4.}\label{fig:cor-rain-nsw}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{figures/gam_check} \caption{The residual analysis of a sample pixel in the QLD region (top) and NSW/VIC region (bottom).}\label{fig:residuals}
+\includegraphics[width=0.9\linewidth]{figures/FigS3} \caption{The residual analysis of the Generalised additive modelling  indicating how well the residuals follow the regression assumptions.
+Results are shown for a sample pixel in the QLD region (top) and NSW/VIC region (bottom).
+Residuals are fairly normal as seen from the histogram and the qqplot, but show some scattering in the variance.}\label{fig:residuals}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=0.7\linewidth]{figures/ResidualBoxplotchange} \caption{Boxplots of annual rainfall residuals (estimated based on Equation 2 before and after the land cover intervention during 1979 - 2015 in the study regions.
+\includegraphics[width=0.7\linewidth]{figures/FigS4} \caption{Boxplots of annual rainfall residuals (estimated based on Equation 2 before and after the land cover intervention during 1979 - 2015 in the study regions.
 On average, the after period has a significantly lower annual rainfall residual in NSW/VIC, but a significantly higher annual rainfall residual in the Qld study area}\label{fig:meandiff}

### 2019-07-29 Manuscripts/PeerJRainfall_Landuse/Rainfall_Landuse.Rmd
small changes before production

@@
 
-The pixels, where the tree cover change based on the MOD44B data was significant ($p \leq 0.05$) in each study region, are shown in Figure \@ref(fig:tctrend) for the NSW/VIC region  (left panel) and the QLD region (right panel).
+The pixels, where the tree cover change based on the MOD44B data was significant ($p \leq 0.05$) in each study region, are shown in Figure \@ref(fig:tctrend) for the NSW/VIC region  (A) and the QLD region (B).
 In the QLD region there actually has generally been an increase in tree cover after the clearing of native vegetation stopped.

### 2019-07-29 Manuscripts/PeerJRainfall_Landuse/SupplementaryMaterialRainfall_Landuse.Rmd
small changes before production

@@
 ---
+\renewcommand{\figurename}{Figure S}
 ```{r setup, warning=F, message=F, echo =F}
@@
       "C:/users/rver4657/owncloud/chun/chun/manuscripts/PeerJRainfall_Landuse")
-knitr::opts_knit$set('figcap.prefix','Figure S')
-knitr::opts_knit$set('tabcap.prefix','Table S')
 ```
